### 2014-06-03 R/Mort_models.Rmd
Draft JAGS mortality  models based on Daniel's 2011 paper.

Mortality models - currently in draft form
========================================================

Below outlines the skeleton structure of potential mortality
models that can be run through JAGS.
These models incorporate changes in hazard rate with time.

Currently only including models based on Daniel's 2011 paper.
Will add other models later.

Also will convert these models to run in stan and 
vectorise where possible.
Models are likely to converge alot faster then using
the gibb sampler in JAGS.

Data will need to be supplied in long format.
i.e. individual measurements in all years occur as rows.
This allows makes it easier to sample individuals that
drop out during 1990 - 2010 while also accounting for new 
recruits. Data will contain only single stemmed plants that
are alive for at least 2 censuses and contain required trait data.

```{r Daniel's model - no random effects for species}  
model {
  for(it in 1:n.obs) {
    # Loop over observations (seedling at site at time).
    # Seedling status, y, is bernoulii with prob=p.
    # Complementary log-log link.
    # Intercept (expected mortality rate).
    y[it] ~ dbern(p[it])
    cloglog(p[it]) <-
      IndMort * exp( - beta[1] * WD[it])  + GRMort * exp(- beta[2]* basal.rgr[it])) +
      logT[it] - # log of time since last observation
      (p[PREV1[it]] +   # Probability of death at last timestep.
         p[PREV2[it]] + # Probability of death at last 2 timesteps.
         p[PREV3[it]] + # Probability of death at last 3 timesteps.
         p[PREV4[it]] + # Probability of death at last 4 timesteps.
         p[PREV5[it]])  # Probability of death at last 5 timesteps.
  }

#Priors

IndMort[j] ~ dnorm(0, 0.0001) # C0 in Daniels model (growth independent)
GRMort[j] ~ dnorm(0, 0.0001) # C2 in Daniels model (growth dependent)

beta[1] ~ dnorm(0, .0001) # Effect of woody density on intrinsic mortality (c1)
beta[2] ~ dnorm(0, .0001) # Effect of growth rate on intrinsic mortality (c3)

p[P0] <- 0 # P0 is a supplied object defining the n.obs + 1
```

You can also embed plots, for example:

```{r Daniel's model with species random effect}
model {
  for(it in 1:n.obs) {
    # Loop over observations (seedling at site at time).
    # Seedling status, y, is bernoulii with prob=p.
    # Complementary log-log link.
    # Intercept (expected mortality rate).
    y[it] ~ dbern(p[it])
    cloglog(p[it]) <-
      IndMort[SPP[it]] * exp( - beta[1] * WD[it])  + GRMort[SPP[it]] * exp(- beta[2]* basal.rgr[it])) +
      logT[it] - # log of time since last observation
      (p[PREV1[it]] +   # Probability of death at last timestep.
         p[PREV2[it]] + # Probability of death at last 2 timesteps.
         p[PREV3[it]] + # Probability of death at last 3 timesteps.
         p[PREV4[it]] + # Probability of death at last 4 timesteps.
         p[PREV5[it]])  # Probability of death at last 5 timesteps.
  }

# Species random effects on C0 and C2 of Daniels model
for(j in 1:n.spp){
IntMort[j] ~ dnorm(IntGM.SPP, IntSD.SPP^-2)
GRMort[j] ~ dnorm(GRMort.SPP, GRSD.SPP^-2)
}
# Coefficient priors on C1 and C3

beta[1] ~ dnorm(0, .0001) # Effect of woody density on intrinsic mortality (C1)
beta[2] ~ dnorm(0, .0001) # Effect of growth rate on intrinsic mortality (C3)
IndGM.SPP ~ dnorm(0, 0.0001) # Mean independent mortality rate (across species)
GRMort.SPP ~ dnorm(0, 0.0001) # Mean Growth related mortality rate across species
IntSD.SPP ~ dunif(0, 100) # Variation in SPP independent intrinic mortality rate
GRSD.SPP ~ dunif(0, 100) # Variation in SPP growth dependent intrinsic rate

p[P0] <- 0 # This gets added as value 2012 for p in Will's model.

```

### 2014-07-27 R/JAGS.Rmd
JAGS code for mortality models

---
title: "Mortality Models"
author: "J Camac"
date: "27 July 2014"
output: pdf_document
---

Mortality model
========================================================

**Summary**
Below outlines the code used for a linear cloglog mortality model that estimates the probability an individual will die by the next census (t+1) as a function of wood density (measured at species level) and either a size (e.g. dbh or basal area) or their associated growth rates (either incremental growth or relative growth rate - measured at individual level).

*Data*
This model uses BCI data for single stemmed plants that are alive for at least two censuses - which is required to estimate a growth rate.
Furthermore, individuals that return from the dead have also been removed from the dataset as well as individuals that are missed in one or more censuses. Note the latter two cases involves few individuals. 
Observations in 1990 are treated as baseline. Thus mortality is not estimated in this year because there is no record of growth rate and only individuals alive in that year are used. 
Observed deaths in 2010 are included. However, model does not make predictions based on 2010 growth rates for 2015 mortality. *This can be done at a later stage.*
Because this model predicts future mortality, the model only uses observations of when the individual is alive. Thus census observations of deaths become redunant and are removed from the final dataset (as this information is incoporated into the column dead.next.census).
Lastly only species that contain trait values are used within this model.

In total there are:
240 species
164,472 individuals
391,267 observations
For complete break down of observations removed see make.bci.mainstem.R.

*Covariate transformations*
In this model the covariates have been centered on their means. Thus the intercept is interpreted as the average probability an individual of average wood density and growth rate will die in the next census. The covariate effects are the additional effects when either covariate is above or below their mean. *Note* signs of covariates invert when estimating covariate values below the mean.
Covariates have also been scaled by two standard deviations so that effect sizes can be compared on standardised scale - see Gelman & Hill 2007. 
*Potential issue* some of the covariate data is skewed and thus centering and scaling will be biased. dbh, basal.area and their incremental growth rates cannot be log or sqrt transfromed as negative growth is observed. Will consider other transformations.

*Model assumptions*
* Hazard rate is constant - i.e. Probability of dying is not adjusted by the probability of dying in the previous census.
* The model assumes a linear, multiplicative relationship between traits and growth rate covariates.
* Models incoporating RGR implicilty include size effects.

```{r Data prep, echo=FALSE}
library('dplyr')
library('R2jags')
load('../data_BCI/output/bci.mainstem.Rdata')
bci.traits <- read.csv('../data_BCI/rawdata/traits/BCI_traits_20101220.csv')
names(bci.traits) <- tolower(names(bci.traits)) # lowers trait column names for merging
bci.traits$sp <- tolower(bci.traits$sp) # lowers species code names for merging

bci.mainstem <- merge(bci.mainstem,bci.traits[,c('sp','sg100c_avg')],by = 'sp') #only uses species trait data exists for.
rm(bci.traits) # no longer needed

bci.mainstem<- bci.mainstem %.%
  filter(!is.na(census.interval) & !is.na(sg100c_avg) & !is.na(dead.next.census) & census.interval< 6) %.%
  mutate(sp.id = as.numeric(factor(sp), as.character(unique(sp))))
```

```{r Linear cloglog mortality model, echo=FALSE}
  data <- list(
  n.obs = nrow(bci.mainstem),
  n.spp = length(unique(bci.mainstem$sp)),
  SPP = as.numeric(factor(bci.mainstem$sp), as.character(unique(bci.mainstem$sp))),
  wd = scale(unique(bci.mainstem[c('sp', 'sg100c_avg')])[,'sg100c_avg'])[,1]/2,
  dbh = scale(bci.mainstem$dbh)[,1]/2,
  dbh.gr = scale(bci.mainstem$dbh.gr)[,1]/2,
  dbh.rgr = scale(bci.mainstem$dbh.rgr)[,1]/2,
  basal.area = scale(bci.mainstem$basal.area)[,1]/2,
  basal.gr = scale(bci.mainstem$basal.area.gr)[,1]/2,
  basal.rgr = scale(bci.mainstem$basal.area.rgr)[,1]/2,
  ln.census = scale(log(bci.mainstem$census.interval))[,1]/2,
  died = bci.mainstem$dead.next.census)

# Note this runs multiple models with each loop iteration changing the covariates used
models <- c('dbh', 'dbh.gr', 'dbh.rgr', 'basal.area', 'basal.gr','basal.rgr')
nmodels <- length(models)
for (i in 1:nmodels) {
cat(sprintf('
  model {
    for(it in 1:n.obs) {

      died[it] ~ dbern(p[it])
      cloglog(p[it]) <- Alpha[SPP[it]] + b.2 * %1$s[it] + b.3 * (wd[SPP[it]] * %1$s[it]) + ln.census[it]
      }

    for (j in 1:n.spp) {
      Alpha[j] <- GM + b.1 * wd[j]
      }

    #Priors

    GM ~ dnorm(0, 0.0001) # Intercept
    b.1 ~ dnorm(0, 0.0001) # Wood Density effect
    b.2 ~ dnorm(0, 0.0001) # Growth/size effect
    b.3 ~ dnorm(0, 0.0001) # Wood density x growth/size effect

    }', names(data)[which(names(data) %in% models)][i]),
    file=(tmp <- tempfile()))

  assign(sprintf('%s.mod', models[i]), 
         jags.parallel(model.file = tmp,
                 n.iter = 4000,
                 data= names(data), 
                 envir=list2env(data),
                 inits = NULL,
                 parameters.to.save = c('GM','b.1','b.2','b.3')))
  }
```

### 2014-07-27 analysis.R
Adds cod eto build coefficient and 3d plots of JAGS models

@@ -1,9 +1,66 @@
-#!/usr/bin/env Rscript
-
 source("R/utils.R")
+library(R2jags)
+load('output/JAGS_mort.rda')
+# Prep for simulations
+sim <- sim.coeff(dbh.mod, c('b.3', 'b.2', 'b.1', 'GM'))
+cov.info <- cbind(data.summary(bci.mainstem, c('dbh','dbh.gr','dbh.rgr','basal.area','basal.area.gr','basal.area.rgr')), # individual level 
+                  data.summary(unique(bci.mainstem[,c('sp','sg100c_avg')])['sg100c_avg'], c('sg100c_avg'))) # species level
 
-fig1 <- function(){
-	plot(1:10, 1:10, type='b')
+
+dbh.figs <- function(){
+  par(mfrow=c(3,2))
+  
+  #dbh
+  coeff.plot(dbh.mod, c('b.3', 'b.2','b.1','GM'), labels = c('WD x DBH','DBH','WD','GM'), 
+             xlab = 'cloglog (effect size)')
+  plot.3d(jagsfit = dbh.mod, x = seq(0.2,0.8,0.05), y = seq(11,230, 10), xmean =
+            cov.info['mean', 'sg100c_avg'], xstd = cov.info['std', 'sg100c_avg'], 
+          ymean = cov.info['mean', 'dbh'], ystd = cov.info['std', 'dbh'], 
+          xlab = 'WD', ylab= 'DBH', zlab = 'Prob dying')
+  
+  #dbh.gr
+  coeff.plot(dbh.gr.mod, c('b.3', 'b.2','b.1','GM'), labels = c('WD x DBH.gr','DBH.gr','WD','GM'), 
+             xlab = 'cloglog (effect size)')
+  plot.3d(jagsfit = dbh.gr.mod, x = seq(0.2,0.8,0.05), y = seq(0,5, 0.5), xmean =
+            cov.info['mean', 'sg100c_avg'], xstd = cov.info['std', 'sg100c_avg'], 
+          ymean = cov.info['mean', 'dbh.gr'], ystd = cov.info['std', 'dbh.gr'], 
+          xlab = 'WD', ylab= 'DBH.gr', zlab = 'Prob dying')
+  
+  #dbh.rgr
+  coeff.plot(dbh.rgr.mod, c('b.3', 'b.2','b.1','GM'), labels = c('WD x DBH.gr','DBH.rgr','WD','GM'), 
+             xlab = 'cloglog (effect size)')
[…]

### 2014-07-27 R/utils.R
Adds several functions to explore JAGS model outputs

@@ -1,4 +1,3 @@
-
 to.dev <- function(expr, dev, filename, ..., verbose=TRUE) {
   if(!file.exists(dirname(filename)))
     dir.create(dirname(filename), recursive=TRUE)
@@ -12,6 +11,85 @@ to.dev <- function(expr, dev, filename, ..., verbose=TRUE) {
 to.pdf <- function(expr, filename, ..., verbose=TRUE) {
   to.dev(expr, pdf, filename, ..., verbose=verbose)
 }
+
+# Simulates coefficient values based on its mean and standard deviation
+sim.coeff <- function(jagsfit, params) {
+  if(!require(jagstools)) {
+    require(devtools)
+    install_github(repo='jagstools', username='johnbaums')
+  }
+  as.data.frame(apply(jagsresults(jagsfit, params), 1, function(x) rnorm(1000, x['mean'], x['sd'])))}
+
+# Provides the mean, standard deviation, 95%,80% and 50% quantiles of desired covariates.
+data.summary <- function(x, covariates){
+  as.data.frame(x)
+  if(ncol(x) ==1) {
+    as.data.frame(apply(x,2, function(x) c(mean= mean(x), std= sd(x), 
+                                           quantile(x,c(0.025, 0.1, 0.25, 0.5, 0.75, 0.9,0.975)))))}
+  else {
+    df <- x[, c(covariates)]
+    as.data.frame(apply(df,2, function(x) c(mean= mean(x), std= sd(x),
+                                            quantile(x,c(0.025, 0.1, 0.25, 0.5, 0.75, 0.9,0.975)))))}}
+
+# Produces a JAGS coefficient plot with credible intervals
+coeff.plot <- function(jagsfit, params, labels=NULL, conf=c(95, 50), xlab='Effect size') {
+  
+  if(!require(jagstools)) {
+    require(devtools)
+    install_github(repo='jagstools', username='johnbaums')
+  }
+  
+  conf <- unique(as.numeric(conf))
+  if (!all(conf %in% c(95, 50))) 
+    stop('conf must be a vector of one or more of 95 and 50.')
+  dat <- jagsresults(jagsfit, params)
+  dat <- dat[match(params, row.names(dat)), ]
+  if (is.null(labels)) labels <- row.names(dat)
[…]

### 2014-11-11 ms/Mortality Proposal.Rmd
Preliminary draft of proposals - without maths

---
title: "Mortality Notes"
author: "J Camac"
date: "9 July 2014"
output: pdf_document
---

###Background

Mortality rates vary both spatially and temporally, and as such, are critical to accurately predict vegetation dynamics. However, few models of vegetation dynamics explicitly account for changes in mortality rates (e.g. REFS). Instead, most models used simplifed assumptions such as constant mortality (REF), no mortality (REF) or that mortality varies as a function of some arbitary set of rules (REF). These simplifcations are predominately a consequence of our limited understanding of what governs mortality rates and how it varies temporally, spatially and within and between species. This limited knowledge is partly due to ecologists placing greater emphasis on examining plant growth and how it affects community (REF) and population dynamics (REF). But also because of difficulties associated with modelling mortality. Mortality is highly variable and episodic (REF). As such, it requires individuals to be monitoried for sufficiently long periods of time, ideally from birth to death, which is problematic for long-lived species. Mortality rates are also affected by multiple biotic and abiotic factors, as well as the frequency (REF) and severity (REF) of multiple hazards (e.g. fire, wind, drought), which may or may not be dependent on one another (REF).

This complexity has led to the widespread use of phenomological models to describe patterns in plant mortality. In a review of more than 150 woody plant mortality models, Hawke (2000) found that 93% were phenomological and only 7% were mechanistic; further highlighting the paucity of process-based information about mortality. The mechanistic models all assumed that mortality occured when carbon supply was insufficient for growth (e.g. Friend et al. 1997, Bossel 1986, Weinstein et al 1991, Bugmann et al 1997). As such, none incorporated differential mortality rates among species, or how different hazards may these rates. Phenomological models also

How many data are needed on mortality? When and where should they be ob- tained? This is a general criticism of Friend et al. (1997), who claim these problems render valida- tion of mechanistic models useless. Also, data sources for mortality testing are often flawed. Bugmann (1996) points out that monitoring pro- grammes are usually shorter than many plants’ lifespans, while palaeo data does not provide the necessary resolution. And the potential variation between sizes, classes, ages etc. is so high that even a large sample will not provide the statistical significance required.

Mechanistic models are constructed with al- gorithms that describe biological processes, as opposed to an empirical data set or predefined relationship. Thus mechanistic algorithms require no calibration, and are inherently deterministic.

Furthermore, ecologists must also asertain when an individual died and what caused its death, the determination of which is inherently difficult in plants.

#How have people attempted modelling mortality?
- In order to capture the U-shaped mortality researchers have used a variety of models to capture broad patterns. These include running seperate models for small and large plants, the use of polynomials, or the use of non-parameteric approaches that do not assume an underlying curve.
- They have often constructed these models to determine the effects of competition, size or ...
- However, the problem with such approaches is that they do not attempt to explain the underlying mechanisms that result in U-shaped mortality.
- Some have suggested that certain hazards influence different parts of the U-shape curve. For example, competitive process influence mortality when individuals are small and less so when they are large. Whereas the oppose is true for events such as windthrow.
- Furthermore, the vast majority of these empirical models have been conducted for individual species, and thus are of limited use for more applying to applying to other species.

#What do we know about mortality in plants?
Evidence suggests that mortality is U-shaped for most plants.
We also know that plant survival is influenced by multiple hazards (fire, herbivory, climate, drought, competition and windthrow)

#Traits
More recently ecologists have used traits in an attempt to obtain more generality. The vast majority of this research has focused on how traits influence growth rates, with very few studies examining trait effects on mortality.
Those that have, have reported that traits such as wood density can significantly influence rates of mortality. The logic here is that wood density conveys additional resources into structural properties and thus makes plants more resistant to breakage and invertebrate attack. However, as far as I am aware, all these studies have assumed that the effect of traits remains constant and does not change with size.

#Solution

Many modellers have called for more mechanistic al- gorithms in plant growth models and see the path
towards improvement as a reformulation of mor- tality algorithms to better reflect the physiological base of a plant (e.g. Loehle and LeBlanc, 1996). (Hawke)

Based on the information above I have three primary objectives:

1) Scour the literature on how traits and size influence mortality rates under common hazards
2) Build an integrated process-based model that incorporates multiple hazards that can be used to determine how the combination of size and traits governs instantenous mortality rates.
3) Apply model to multiple datasets.

#Objective 1

As far as I am aware, there does not appear to be an integrated review of how traits influence various hazards, or how they vary with size.
Most trait analysis of mortality has either ignored the cause of death or has been related to very specific forms of mortality. For example, there are several studies that highlight suite of traits that minimise death by fire, but less so on other hazards such as wind throw and climatatic extremes.
In order to remedy this. I am to write a brief literature review highlighting the primary hazards plants are exposed to and how various traits have been linked to decreased or increased susceptibility.

In order to achieve this I plan to use google scholar and...

#Objective 2
Include maths here

#Objective 3
Preliminary thoughts

Modelling survival and mortality within a population is of fundamental importance in ecology, epidemiology, demography, engineering and economics. It is particularly

Burgman et al. (1994) points out the significance of tree mortality for forest yield and growth rates, and Pacala et al. (1996) shows the importance of species mortality rates in forest succession.What is clear is that mortality is very difficult to model owing to its variable and episodic nature (Burgman et al., 1994) (Hawke)

#Difficulty in modelling mortality
The complexity of environmental stresses and lack of process information of woody plant mortality has however led to the wide use of empirical algorithms.These have many problems, including data intensity and spatial and temporal specificity. Such problems render mortality simulation prone to error and weak in response to environmental change.  (Hawke)

Describing size-specific mortality patterns can be especially difficult, due to few trees in larger size classes. (Hurst)

###Aim:
1. Determine traits influence the likelihood of a particular hazard resulting in death.
2. Determine whether instantaneous mortality rates are commonly U-shaped as a function of dbh.
3. Examine whether trait effects on instantaneous mortality rates vary with individual size and growth rate
4. Build a process-based mortality model that encapsulates how the combination of various hazards and traits can result in various mortality curves.

###Hypotheses
1. The most common hazards that can result in plant death are all size dependent and include:
**Competition**: This hazard will most strongly affect smaller individuals because they are more at risk of being shaded compared to larger plants. Traits affecting the likelihood of competition resulting in death will be correlated with shade tolerance such as *LMA* and *wood density*.

**Herbivory**: This hazard will most strongly affect small individuals more susceptible to complete foliage loss or being consumed entirely. Traits affecting the liklihood of herbivory resulting in death will be correlated with herbivory preference (e.g. *leaf nitrogen*), resistance (e.g. *wood density*) and persistance (*resprouting capabilities*).

**Fire**: This hazard is likely to be size dependent with small individuals more suspectible to fire compared to larger plants. This is because for small/juvenile plants will are less likely to have developed fire-resistant traits. They are also lower in stature making their leaves exposed to both low severity and high severity fires. Traits affecting the likelihood of fire resulting in death will be correlated with fire-resistance (e.g. bark thickness) and persistance (e.g. resprouting)

**Drought/Extreme heat/ Extreme cold**: These hazard is likely t
  - Extreme temperatures
  - Windthrow
  - Neighbour falling
  - Disease

Below are some excerpts from William Morris's thesis on mortality of trees.

The **finite mortality rate**, $M$, ignoring recruitment, is the complement of the ratio of final populatio size, $N_T$, to initial population size, $N_0$, divided by the length of the observation period, $T$,

\begin{equation} \label{eq:FiniteMort}
M = 1 -\frac{N_T/N_0}{T}
\end{equation}

However, in order to infer an unbiased expected proportion of dying individuals for periods shorter or longer than that used to calculate finite mortality one must calculate the **instantaneous mortality**, $M_T$:

\begin{equation} \label{eq:InstMort}
M_T = 1-\textrm{exp}(-\lambda T)
\end{equation}

where

\begin{equation} \label{eq:lambda}
\lambda = - \frac{\textrm{log}_e(1-M)}{T}.
\end{equation}

This function is widely used to describe the loss of individuals from a population over time. Here, when $M_T$ and $\lambda$ are more or less equivilent mortality rates are low.

Calculating mortality rate is made more complicated when multiple censuses (a period of time in which mortality risk is being assessed, where census interval = $T$) are considered, partlicalrly when census interval lengths vary. In such cases equations \ref{eq:FiniteMort} and \ref{eq:InstMort} cannot be used because $T$ is not dixed and substituting $T$ with average interval length has been shown to underestimate mortality (Kubo et al., 2000). This is a particular problem with large-scale monitoring programs such as BCI where the period of time between revisits of individual trees varies within a signle census because there are too many individuals for observations to be made simultaneously. Datasets such as BCI can take more than a year to complete an entire census.

When $T$ varies within a mortality dataset, estimates of population mortality rate will be biased if this variation is not accounted for (e.g. by assuming a mean value for $T$ and using equation \ref{eq:InstMort}. In order to achieve an unbiased estimate of $M_T$ and $\lambda$ when census length varies, maximum likelihood (Kubo et al, 2000) or Bayesian (He, 2003) methods can be used, which allow $T$ to be considered a continous variable.

The methods so far assume $\lambda$ is fixed in space and time, between individuals and over the lifespan of an individual. Often these assumptions are not met because individual hazard rates vary. Heterogeneous hazard biases estimates of population mortality rates (Sheil & May 1996) and this bias is greater for longer census intervals (Zens & Peart 2003). The need for unbiased estimates of mortality, despite the hazard heterogeneity present in data from forests and woodlands, has lead to the development and application of methods that move beyond considering mortality at the population level and directly modelling hazard of individual trees (Zens & Peart, 2003).

The logistic regression family of methods are the most common individual-based approach used to analyse tree mortality data (see Breece, Kolb & Dickson, 2008, Carus 2010, Flewelling & Monserud, 2002). Logistic regression models consider tree death, $\textrm{Pr}(y_i = 1)$, as a series of binomially distributed data with the probability of observing a death as function of a set of covariates, $X_i$, and associated coefficients, $\beta$ related to data via the logit link function.

\begin{equation} \label{eq:logit}
\textrm{Pr}(y_{i} = 1) = \textrm{logit}^{-1}(\beta X_i)
\end{equation}

The logit function, $\textrm{log}_e(p/(1 - p))$, transforms a quantity bound within the unit interval, such as probability, $p$, to cary between negative infinity and infinity. By relating mortality to covariates that can explain differences in death rate, individual based methods such as logistic regression can reduce the bias found in population rate estimates when individuals have variable hazard (Zens & Peart, 2003).

The complementary log-log (cloglog) link, $\textrm{log}_e(-\textrm{log}_e(1-p))$, is another link that can be used to estimate mortality rates.
The advantage of the cloglog link is that it is equivalent to standard survival analysis... NEED MORE INFO. It also reduces bias due to long and variable census intervals (BY?).
By adding the census interval $T$ to equation \ref{eq:logit} and substituting the logit link with a cloglog link yields

\begin{equation} \label{eq:cloglog}
\textrm{Pr}(y_{i} = 1) = \textrm{cloglog}^{-1}(\beta X_{i} + \textrm{log}(T))
\end{equation}

y = log(-log(1-P))

y = log(e) + log(1/(1-P)))
e^y = (1/(1-P))
e^y*(1-P) = 1
e^y - Pe^y = 1

-e(e^y) + 1

1 - e(e^y)

### MULTI-LEVEL NOTES:

####From Hedeker et al 2014

Nice summary of why multi-level models are essential for accounting for non-independence in nested datasets.

"An important question is then to determine the degree to which covariates are related to substance use initiation. In these studies it is often of interest to model the student outcomes while controlling for the nesting of students in classrooms and/or schools. In analysis of such grouped-time initiation (or survival) data, use of grouped-time regression models that assume independence of observations [Thompson, 1977; Prentice & Gloeckler, 1978; Allison, 1982] is therefore problematic because of this clustering of students. More generally, this same issue arises for other types of clustered datasets in which subjects are observed nested within various types of clusters (e.g., hospitals, firms, clinics, counties), and thus cannot be assumed to be independent. To account for the data clustering, multilevel models (also called hierarchical linear or mixed models) provide a useful approach for simultaneously estimating the parameters of the regression model and the variance components that account for the data clustering [Goldstein, 1995; Raudenbush & Bryk, 2002]."

## TWO REVIEW PAPERS TO READ Pickles et al 1995 & Hougaard 1995 ###

Lee et al 1992 has developed continuous0time survival models. However, the application of these models to grouped or discrete-time survival data is generally not recommended because of the large number of ties that result.

Instead, models specifically developed for grouped or discrete-time survival data have been proposed.

Both Han & Hausman (1990) and Scheike & Jensen (1997) have described proportional hazards models incorporating a log-gamma distribution specification of heterogeneity. Furthermore, Ten Have (1996) developed a discrete-time proportional hazards survival model incorporating a log-gamma random effects distribution, additionally allowing for ordinal survival and failure categories. Ten Have & Uttal (1994) used Gibbs samplng to fit a continuation ratio logit model with multiple normally distributed random effects.

## IMPORTANT NOTE ON link function

Doksum & Gasko (1990) note, that large amounts of high quality data are often necessary for link function selections to be relevant. McCullagh (1980) notes that the link function choice should be based primarily on ease of interpretation.

## Multilevel model that treats each individual's survival time as a set of dichotomous observations indicating whether or not an individual failed in each time unti until a person either experiences the event or is censored. This type of model is extensively described in Singer & Willett (2003) & is illustrated in Reardon et al (2002).

This approach is particularly useful for handling time-dependent covariates and fitting non-proportional hazards models because the covariate values can change across each individual's Tij timepoints.

### 2014-11-11 ms/Mortality Proposal.Rmd
added some maths and figures - needs substantial clean up

@@
 
-**Drought/Extreme heat/ Extreme cold**: These hazard is likely t
-  - Extreme temperatures
-  - Windthrow
-  - Neighbour falling
-  - Disease
+**Drought/Extreme heat/ Extreme cold**: These hazard is likely to..
 
+###Factors influencing small individuals
+```{r, echo=FALSE}
+par(xaxs='i', yaxs='i', mfrow= c(3,2))
+curve(1/(1+((1/0.99)-1)*exp(5*x)), ylim=c(0,1), xlim=c(0, 1.2), xaxt='n', ylab= 'Mortality rate', xlab='Size', main='Death by fire', lwd=2)
+axis(1, at=c(0,1.2), labels=c('Small', 'Large'))
+
+curve(1/(1+((1/0.99)-1)*exp(7*x)), ylim=c(0,1), xlim=c(0, 1.2), xaxt='n', ylab= 'Mortality rate', xlab='Size', main='Death by drought',lwd=2)
+axis(1, at=c(0,1.2), labels=c('Small', 'Large'))
+
+curve(1/(1+((1/0.99)-1)*exp(7*x)), ylim=c(0,1), xlim=c(0, 1.2), xaxt='n', ylab= 'Mortality rate', xlab='Size', main='Death by high temperatures',lwd=2)
+axis(1, at=c(0,1.2), labels=c('Small', 'Large'))
+
+curve(1/(1+((1/0.99)-1)*exp(7*x)), ylim=c(0,1), xlim=c(0, 1.2), xaxt='n', ylab= 'Mortality rate', xlab='Size', main='Death by low temperatures',lwd=2)
+axis(1, at=c(0,1.2), labels=c('Small', 'Large'))
+
+curve(1/(1+((1/0.99)-1)*exp(7*x)), ylim=c(0,1), xlim=c(0, 1.2), xaxt='n', ylab= 'Mortality rate', xlab='Size', main='Death by herbivory',lwd=2)
+axis(1, at=c(0,1.2), labels=c('Small', 'Large'))
+
+curve(1/(1+((1/0.99)-1)*exp(15*x)), ylim=c(0,1), xlim=c(0, 1.2), xaxt='n', ylab= 'Mortality rate', xlab='Size', main='Death by light competition', lwd=2)
+axis(1, at=c(0,1.2), labels=c('Small', 'Large'))
+```
+
+###Factors influencing large individuals
+```{r, echo=FALSE}
+par(xaxs='i', yaxs='i',mfrow=c(1,3))
+curve(1/(1+((1/0.02)-1)*exp(-10*x)), ylim=c(0,1), xlim=c(0, 1.2), xaxt='n', ylab= 'Mortality rate', xlab='Size', main='Death by wind fall', lwd=2)
+axis(1, at=c(0,1.2), labels=c('Small', 'Large'))
+
+curve(0.9/(1+((0.9/0.2)-1)*exp(-7*x)), ylim=c(0,1), xlim=c(0, 1.2), xaxt='n', ylab= 'Mortality rate', xlab='Size', main='Death by falling neighbour',lwd=2)
+axis(1, at=c(0,1.2), labels=c('Small', 'Large'))
+
+curve(1/(1+((1/0.001)-1)*exp(-8*x)), ylim=c(0,1), xlim=c(0, 1.2), xaxt='n', ylab= 'Mortality rate', xlab='Time', main='Death by disease',lwd=2)
+axis(1, at=c(0,1.2), labels=c('0', 'infinity'))
+```
+
+###Traits are likely to affect shape of these hazard rates
+```{r, echo=FALSE}
+par(xaxs='i', yaxs='i', mfrow=c(1,2))
+curve(1/(1+((1/0.02)-1)*exp(-10*x)), ylim=c(0,1), xlim=c(0, 2), xaxt='n', ylab= 'Mortality rate', xlab='Size', main='Death by wind fall')
+curve(1/(1+((1/0.02)-1)*exp(-5*x)), ylim=c(0,1), xlim=c(0, 2), xaxt='n', ylab= 'Mortality rate', xlab='Size', main='Death by wind fall', add=T, col='red', lty=2, lwd=2)
+axis(1, at=c(0,2), labels=c('Small', 'Large'))
+legend(0.7,0.2, legend = c('low wd', 'high wd'), col=c('black','red'), lty=c(1,2), bty = 'n', lwd=2, cex=2,)
+
+curve(1/(1+((1/0.02)-1)*exp(-10*x)), ylim=c(0,1), xlim=c(0, 2), xaxt='n', ylab= 'Mortality rate', xlab='Size', main='Death by wind fall')
+curve(1/(1+((1/0.02)-1)*exp(-5*x)), ylim=c(0,1), xlim=c(0, 2), xaxt='n', ylab= 'Mortality rate', xlab='Size', main='Death by wind fall', add=T, col='red', lty=2, lwd=2)
+axis(1, at=c(0,2), labels=c('Small', 'Large'))
+legend(0.7,0.2, legend = c('low severity', 'high severity'), col=c('black','red'), lty=c(1,2), bty = 'n', lwd=2, cex=2,)
+```
+
+## Survival analysis
+
+- Most people model survival curves
+- However, here we are interested in determining how trait alter hazards.
+
+The survival function is intimately related to the **hazard** or **instantaneous mortality function**,  denoted $\lambda(t)$, which gives the event rate at time $t$ conditional on survival until time $t$:
+
+\begin{equation} \label{eq:lambda} \lambda(t) = \lim_{dt \rightarrow 0} \frac{\Pr(t \leq T < t+dt)}{dt\cdot S(t)} = -\frac{S'(t)}{S(t)}.\end{equation}
+
+This means that:
+\begin{equation}
+Survival(t) = \lambda_1(t) \times \lambda_2(t) \times \lambda_3(t)... \times \lambda_n
+\end{equation}
+
+### Here I plan to model hazards by...
+####EG Fire
+
+\begin{equation}
+\lambda(t) = f(age)^{-1} \times \int_{0}^{\infty} p(intensity|age) \times \Pr(\textrm{Death in fire of intensity}\, i) \, \textrm{d}i
+\end{equation}
+
+\begin{equation}
+\Pr(Death)  = \alpha + \beta_1 \times Size + \beta_2 \times Resprouter + \beta_3 \times Bark thickness
+\end{equation}
+
+##OTHER STUFF
 Below are some excerpts from William Morris's thesis on mortality of trees.

### 2014-11-11 Junk/kmcurves.R
Quick attempt at KM-Curves

@@ -1,0 +1,142 @@
+
+test <-subset(bci.mainstem, !is.na(bci.mainstem$sg100c_avg) & !is.na(bci.mainstem$dead.next.census & !is.na(bci.mainstem$rel.dbh)) & !is.na(bci.mainstem$dbh.gr))
+test$dead.next.census[is.na(test$dead.next.census)]<-0
+library(survival)
+test <-subset(bci.mainstem, !is.na(bci.mainstem$sg100c_avg) & !is.na(bci.mainstem$dead.next.census & !is.na(rel.dbh)) & !is.na(dbh.gr))
+test$sizeclass <- 0
+test$sizeclass[test$rel.dbh > 0.33]<- 1
+test$sizeclass[test$rel.dbh > 0.63]<- 2
+test$dead.next.census[is.na(test$dead.next.census)]<-0
+lfit <- survfit(Surv(rel.dbh, dead.next.census) ~ 1, data=test)
+
+
+plot(lfit, xlab="time", ylab="Survival Probability")
+lines(lfit, col='pink')
+
+
+
+for(i in unique(test$sp)) {
+  lfit[[i]] <- survfit(Surv(rel.dbh, dead.next.census) ~ 1, data= subset(bci.mainstem, 
+                                                                       bci.mainstem$sp== i & !is.na(bci.mainstem$sg100c_avg) & 
+                                                                         !is.na(bci.mainstem$dead.next.census & !is.na(rel.dbh)) & !is.na(dbh.gr)))
+}
+
+name <- unique(test$sp)
+par(mfrow=c(4,4), mar=c(2,2,1,1), oma=c(2.5,2.5,0,0))
+for(i in 1:16){
+plot(lfit[[name[i]]], main = name[i])
+}
+mtext(side = 1, text = 'Relative DBH', outer=TRUE, line = 1)
+mtext(side = 2, text = 'Survival Probability', outer=TRUE, line=1)
+
+par(mfrow=c(4,4), mar=c(2,2,1,1), oma=c(2.5,2.5,0,0))
+
+for(i in 17:32){
+  plot(lfit[[name[i]]], main = name[i])
+}
+mtext(side = 1, text = 'Relative DBH', outer=TRUE, line = 1)
+mtext(side = 2, text = 'Survival Probability', outer=TRUE, line=1)
+
+
+for(i in 33:48){
+  plot(lfit[[name[i]]], main = name[i])
+}
[…]

### 2014-11-11 Junk/literature notes.Rmd
Literature notes prior to review

---
title: "Literature notes"
author: "J Camac"
date: "22 October 2014"
output: html_document
---
## DROUGHT
**Reference**
Sterck, F. J., Martínez-Vilalta, J., Mencuccini, M., Cochard, H., Gerrits, P., Zweifel, R., et al. (2012). Understanding trait interactions and their impacts on growth in Scots pine branches across Europe. Functional Ecology, 26(2), 541–549. doi:10.1111/j.1365-2435.2012.01963.x

*Findings*
- Tracheid hydraulic diameter was the best predictor of cavitation vulnerability
- Leaf area to sapwood area ratio was positively correlated with branch growth and tracheid hydraulic diameter (thus caviatation vulnerability)
- Wood density decreased cavitation vulnerability, while it increased with double cell wall thickness & tracheid diameter
- Cause & effect analysis (pathway analysis) suggested the causal link affecting caviation vulnerability and specific conductivity was tracheid hydraulic diameter and not other anatomical traits such as wood density and double cell wall thickness.
- Pathway analysis also suggested that growth was only driven by leaf area to sapwood ratio.

*Other*
-intraspecific variation to demonstrate functional trade-offs is superior to, or at least as valuable as, the use of interspecific variation.

**Reference**
Poorter, L., & Markesteijn, L. (2008). Seedling traits determine drought tolerance of tropical tree species. Biotropica, 40(3), 321–331. doi:10.1111/j.1744-74.29007.00380.x

*Findings*
- Dry forest species tended to have compound leaves,high stem dry matter content(stem dry mass/fresh mass),and low leaf area ratio,suggesting that reduction of transpiration and avoidance of xylem cavitation are important for their success
- Deciduous species had longer survival times than evergreens (because they could effectively shut down)
- When assessing only evergreens Wood density (stem density) was postively correlated with drought tolerance.
- Drought survival was mainly determined by deciduousness & presence of taproot. Both factors explained 74% of the survival variation. wood density.
- Trait relationships with drought tolerance: Leaf are ratio (-), Leaf mass fraction(-), Root mass fraction (+; i.e. large roots can act as storage), Stem dry matter content (+).

- Three functional groups appear to exist:
  - Drought avoiders (deciduous plants, taproots)
  - Drought resisters (high dry matter content in leaves & stem)
  - light-demanding moist forest species with high belowground foraging capacity

*Other*
- "The seedling stage is generally considered to be the most important bottleneck for successful regeneration in dry areas, as seedlings, with their limited root system are most vulnerable to drought."

- "Dry adapted species have lower tranpiring leaf area per unit plant mass and have compound leaves."
- "Small leaflets have a lower boundary layer resistance, which allows for better convective cooling of leaves."
- Rooting depth could also be associated with increased tolerance as deeper soil layers have more water.
- "Xylem caviation is generally considered to be the most important cause of tree mortality in dry habitats"

## Light

**Reference**

Poorter, L., Bongers, L., & Bongers, F. (2006). Architecture of 54 moist-forest tree species: traits, trade-offs, and functional groups. Ecology, 87(5), 1289–1301.

*Findings*
1) Crown area increased strongly, and nonlinearly, with tree height; Crown length increased monotonically with tree height
2) Light demand was a stronger predictor of architecture than Hmax. The strength of these relationships changed with ontogeny with light strength increasing with size.

- Stem height was positively related to Hmax at larger reference diameters. 
- Wood density was not related to Hmax (contrary to expectations)
- Height was positively correlated with crown area (non-linear) and crown length (linear)
- Asymptotic height growth was not found in all species (these tended to be small understorey species)

## Drought & Shade tolerance

**Reference**
Sterck, F., Markesteijn, L., & Schieving, F. (2011). Functional traits determine trade-offs and niches in a tropical forest community. Proceedings of the National Academy of Sciences of the United States of America.

*Findings*
- Aquisitive species would do better then conservative species under high light & water conditions but do worse under low light & water conditions. (supported)

*Suggested mechanism*
Acquisitive species are characterized by higher levels of photosynthetic proteins (such as Rubisco) than conservative species; have higher carbon gain rates than conservative species when light is not limiting, have protein-driven high respiration rate at low light levels.

- Under dry conditions aquisitive species are expected to close their stomata and reduce transpiration to avoid hydraulic failure, which creates a risk to starve from carbon deficiency when photosynthesis goes too far down. This is because such species require a higher hydraulic conductivity to sustain high transpiration rates. Such highly conductive wood, however, involves a higher risk of hydraulic failure.

*Other information*
"neutral theory advocates that coexisting species tend to be selected for the most common environmental conditions and, therefore, that most coexisting species are functionally equivalent and share similar niches"

**Reference**
Cavaleri, M. A., Oberbauer, S. F., Clark, D. B., Clark, D. A., & Ryan, M. G. (2010). Height is more important than light in determining leaf morphology in a tropical forest. Ecology, 91(6), 1730–1739.

*Findings*
- LMA linearly increased with height for all species pooled and for each functional group individually (except epiphytes), possibly through the influence of gravity on leaf water potential and turgor pressure.
- Light was also positively correlated with LMA... however this was mainly at light levels < 20% TRANS

*Suggested mechanisms*
- Xylem water potential decreases by 0.01 MPa per m of height simply because of gravity, as such leaf turgor pressure is lower at higher heights. This means that cells are small and thus can be packed more denesly causing an increase in cell wall per given area.

*Other*
"A kg of leaves at the bottom of a canopy generally has three to five times the surface area of the same mass of leaves at the top of the canopy"

**Reference**
Markesteijn, L., Poorter, L., Bongers, F., Paz, H., & Sack, L. (2011). Hydraulics and life history of tropical dry forest tree species: coordination of species’ drought and shade tolerance. New Phytologist, 191(2), 480–495. doi:10.1111/j.1469-8137.2011.03708.x

*Findings*
- Leaf hydraulic conductivity was negatively correlated with drought tolerance
- Pioneer species and deciduous species have higher hydraulic efficiency than shade tolerant and evergreen species.
- Stem hydraulic conductivity was negatively associated with wood density and positively correlated with maximum vessel length (reflecting a trade-off between hydraulic saftey vs efficiency)
- drought and shade tolerance scaled similarly with hydraulic properties (-), wood density (+) & leaf dry matter content (+).

*Suggested mechanism*
- Stem hydraulic conducitivity was be negatively associated with both shade and drought tolerances, because light-demanding species have higher photosynthetic capacity, increased transpiration and higher leaf water demand and thus are more sensitive to caviatation than shade-tolerant species.

**Reference**
Markesteijn, L., & Poorter, L. (2009). Seedling root morphology and biomass allocation of 62 tropical tree species in relation to drought- and shade-tolerance. Journal of Ecology, 97(2), 311–325. doi:10.1111/j.1365-2745.2008.01466.x

*Findings*
- Seedlings of dry forest species improve water foraging capacity in deep soil layers by an increased below-ground biomass allocation and by having deep roots. They minimize the risk of cavitation by making dense stems, and reduce transpiration by producing less leaf tissue. 
- Moist forest seedlings have large leaf areas and a greater above-ground biomass, to maximize light interception, and long, cheap, branched root systems, to increase water and nutrient capture.
- Found no trade-off between shade and drought tolerance as they are thought to depend on different morphological adaptations. High biomass investment into root system for drought tolerance and shade tolerance by low growth rate and low SLA (this latter point is odd... as more persistent organs will result in slower growth rates... so if anything they are correlated.)

-three major drought strategies: 
(i) evergreen drought-tolerant species have high biomass investment in enduring organs, minimize cavitation and minimize transpiration to persist under dry conditions; 
(ii) drought-avoiding species maximize resource capture during a limited growing season and then avoid stress with a deciduous leaf habit in the dry season; 
(iii) drought-intolerant species maximize both below- and above-ground resource capture to increase competitiveness for light, but are consequently precluded from dry habitats.

*Other*
"Species are most susceptible to drought in the early phase of their life cycle when seedlings have limited access to water as their small root systems is confined to the upper soil layers"
"Seedling mortality rates increase exponentially with a decrease in rainfall, as such seedling size at the onset of the dry season is therefore important"

"Meta-analysis of 806 woody species from the northern hemisphere suggested a negative relationship between drought- and shade- tolerance (Niinemets & Valladares 2006)."

"The functional equilibrium hypothesis suggests under a given regime of stress, plants maximise their suface area for intake of the most limiting resource. (Brouwer 1963)"

### 2014-11-11 Junk/basic_lmermodels.R
Basic LMER models to examine BCI mortality data - key finding is that mortality tends to be higher with very large dbh.

@@ -1,0 +1,60 @@
+library('dplyr')
+library('R2jags')
+load('../data_BCI/output/bci.mainstem.Rdata')
+bci.traits <- read.csv('../data_BCI/rawdata/traits/BCI_traits_20101220.csv')
+names(bci.traits) <- tolower(names(bci.traits)) # lowers trait column names for merging
+bci.traits$sp <- tolower(bci.traits$sp) # lowers species code names for merging
+
+bci.mainstem <- merge(bci.mainstem,bci.traits[,c('sp','sg100c_avg')],by = 'sp') #only uses species trait data exists for.
+rm(bci.traits) # no longer needed
+
+bci.mainstem <- bci.mainstem %.%
+  group_by('sp') %.% 
+  mutate(rel.dbh = dbh/max(dbh, na.rm=T),
+         max.dbh = max(dbh, na.rm=T),
+         bins = bins.greedy(rel.dbh, target.bins = 100, max.breaks = 200)$binct)
+
+test<- bci.mainstem %.%
+  mutate(
+         bins = bins.greedy(rel.dbh, nbins = 10)$binct
+
+bci.mainstem <-subset(bci.mainstem, bci.mainstem$year==2000 & bci.mainstem$sp=='hybapr' & !is.na(bci.mainstem$sg100c_avg) & !is.na(bci.mainstem$census.interval))
+
+data <- list(
+  n.obs = nrow(bci.mainstem),
+  n.spp = length(unique(bci.mainstem$sp)),
+  SPP = as.numeric(factor(bci.mainstem$sp), as.character(unique(bci.mainstem$sp))),
+  wd = scale(unique(bci.mainstem[c('sp', 'sg100c_avg')])[,'sg100c_avg'])[,1]/2,
+  dbh = scale(log(bci.mainstem$rel.dbh))[,1]/2,
+  ln.census = scale(log(bci.mainstem$census.interval))[,1]/2,
+  died = bci.mainstem$dead.next.census)
+
+# Note this runs multiple models with each loop iteration changing the covariates used
+
+  model <- function(x) {
+    for(it in 1:n.obs) {
+      
+      died[it] ~ dbern(p[it])
+      cloglog(p[it]) <- alpha[SPP[it]] + b.2 * dbh[it] + b.3 * dbh[it] * dbh[it] + b.4 * dbh[it] * wd[SPP[it]]
+      }
+    
+    for (j in 1:n.spp) {
+      alpha[j] <- GM +  b.1 * wd[j]
+      }
+    
+    #Priors
+    
[…]

### 2014-11-12 ms/Mortality Proposal.Rmd
additional edits

@@
 In a review of more than 150 woody plant mortality models, Hawke (2000) found that 93% were phenomological and only 7% were mechanistic; further highlighting the paucity of process-based information about mortality.
-The mechanistic models all assumed that mortality occured when carbon supply was insufficient for growth (e.g.
+These phenological models....
+
+These mechanistic models all assumed that mortality occured when carbon supply was insufficient for growth (e.g.
 Friend et al.
 1997, Bossel 1986, Weinstein et al 1991, Bugmann et al 1997).
-As such, none incorporated differential mortality rates among species, or how different hazards may these rates.
-Phenomological models also
+As such, none incorporated differential mortality rates among species, or how different hazards may alter these rates.
+Phenomological model on the other hand, have attempted to explain mortality as a function of competitive interactions, abiotic conditions as well as size.
+One finding that appears to
 

### 2014-11-13 R/figures.R
Shift to maker workflow

@@ -1,0 +1,66 @@
+# source("R/utils.R")
+# library(R2jags)
+# load('output/JAGS_mort.rda')
+# # Prep for simulations
+# sim <- sim.coeff(dbh.mod, c('b.3', 'b.2', 'b.1', 'GM'))
+# cov.info <- cbind(data.summary(bci.mainstem, c('dbh','dbh.gr','dbh.rgr','basal.area','basal.area.gr','basal.area.rgr')), # individual level
+#                   data.summary(unique(bci.mainstem[,c('sp','sg100c_avg')])['sg100c_avg'], c('sg100c_avg'))) # species level
+
+
+dbh.figs <- function(){
+  par(mfrow=c(3,2))
+
+  #dbh
+  coeff.plot(dbh.mod, c('b.3', 'b.2','b.1','GM'), labels = c('WD x DBH','DBH','WD','GM'),
+             xlab = 'cloglog (effect size)')
+  plot.3d(jagsfit = dbh.mod, x = seq(0.2,0.8,0.05), y = seq(11,230, 10), xmean =
+            cov.info['mean', 'sg100c_avg'], xstd = cov.info['std', 'sg100c_avg'],
+          ymean = cov.info['mean', 'dbh'], ystd = cov.info['std', 'dbh'],
+          xlab = 'WD', ylab= 'DBH', zlab = 'Prob dying')
+
+  #dbh.gr
+  coeff.plot(dbh.gr.mod, c('b.3', 'b.2','b.1','GM'), labels = c('WD x DBH.gr','DBH.gr','WD','GM'),
+             xlab = 'cloglog (effect size)')
+  plot.3d(jagsfit = dbh.gr.mod, x = seq(0.2,0.8,0.05), y = seq(0,5, 0.5), xmean =
+            cov.info['mean', 'sg100c_avg'], xstd = cov.info['std', 'sg100c_avg'],
+          ymean = cov.info['mean', 'dbh.gr'], ystd = cov.info['std', 'dbh.gr'],
+          xlab = 'WD', ylab= 'DBH.gr', zlab = 'Prob dying')
+
+  #dbh.rgr
+  coeff.plot(dbh.rgr.mod, c('b.3', 'b.2','b.1','GM'), labels = c('WD x DBH.gr','DBH.rgr','WD','GM'),
+             xlab = 'cloglog (effect size)')
+  plot.3d(jagsfit = dbh.rgr.mod, x = seq(0.2,0.8,0.05), y = seq(0,0.1, 0.01), xmean =
[…]

### 2014-11-13 R/ms_fun.R

@@ -1,0 +1,13 @@
+
+pandoc_build <- function(file){
+  args <- list('--template=data/include.tex', '--latex-engine=xelatex')
+  pandoc_convert(file, output= paste0(tools::file_path_sans_ext(file), ".pdf"), citeproc = TRUE, options = args, verbose = TRUE, wd = ".")
+}
+
+get_nature_csl <- function(dest, url="https://raw.githubusercontent.com/citation-style-language/styles/master/nature.csl"){
+	cat(getURL(url), file=dest)
+}
+
+get_amnat_csl <- function(dest, url="https://raw.githubusercontent.com/citation-style-language/styles/master/the-american-naturalist.csl"){
+	cat(getURL(url), file=dest)
+}

### 2014-11-13 R/utils.R

@@ -1,17 +1,3 @@
-to.dev <- function(expr, dev, filename, ..., verbose=TRUE) {
-  if(!file.exists(dirname(filename)))
-    dir.create(dirname(filename), recursive=TRUE)
-  if ( verbose )
-    cat(sprintf("Creating %s\n", filename))
-  dev(filename, ...)
-  on.exit(dev.off())
-  eval.parent(substitute(expr))
-}
-
-to.pdf <- function(expr, filename, ..., verbose=TRUE) {
-  to.dev(expr, pdf, filename, ..., verbose=verbose)
-}
-
 # Simulates coefficient values based on its mean and standard deviation
 sim.coeff <- function(jagsfit, params) {
   if(!require(jagstools)) {
@@ -24,7 +10,7 @@ sim.coeff <- function(jagsfit, params) {
 data.summary <- function(x, covariates){
   as.data.frame(x)
   if(ncol(x) ==1) {
-    as.data.frame(apply(x,2, function(x) c(mean= mean(x), std= sd(x), 
+    as.data.frame(apply(x,2, function(x) c(mean= mean(x), std= sd(x),
                                            quantile(x,c(0.025, 0.1, 0.25, 0.5, 0.75, 0.9,0.975)))))}
   else {
     df <- x[, c(covariates)]
@@ -33,14 +19,14 @@ data.summary <- function(x, covariates){
 
 # Produces a JAGS coefficient plot with credible intervals
 coeff.plot <- function(jagsfit, params, labels=NULL, conf=c(95, 50), xlab='Effect size') {
-  
+
   if(!require(jagstools)) {
     require(devtools)
     install_github(repo='jagstools', username='johnbaums')
   }
-  
+
   conf <- unique(as.numeric(conf))
-  if (!all(conf %in% c(95, 50))) 
+  if (!all(conf %in% c(95, 50)))
     stop('conf must be a vector of one or more of 95 and 50.')
   dat <- jagsresults(jagsfit, params)
   dat <- dat[match(params, row.names(dat)), ]
@@ -48,9 +34,9 @@ coeff.plot <- function(jagsfit, params, labels=NULL, conf=c(95, 50), xlab='Effec
   opar <- par(mai=c(1, max(strwidth(labels, 'inches')) + 0.3, 0.5, 0.5))
   on.exit(par(opar))
[…]

### 2014-11-13 analysis.R

@@ -1,66 +1,0 @@
-source("R/utils.R")
-library(R2jags)
-load('output/JAGS_mort.rda')
-# Prep for simulations
-sim <- sim.coeff(dbh.mod, c('b.3', 'b.2', 'b.1', 'GM'))
-cov.info <- cbind(data.summary(bci.mainstem, c('dbh','dbh.gr','dbh.rgr','basal.area','basal.area.gr','basal.area.rgr')), # individual level 
-                  data.summary(unique(bci.mainstem[,c('sp','sg100c_avg')])['sg100c_avg'], c('sg100c_avg'))) # species level
-
-
-dbh.figs <- function(){
-  par(mfrow=c(3,2))
-  
-  #dbh
-  coeff.plot(dbh.mod, c('b.3', 'b.2','b.1','GM'), labels = c('WD x DBH','DBH','WD','GM'), 
-             xlab = 'cloglog (effect size)')
-  plot.3d(jagsfit = dbh.mod, x = seq(0.2,0.8,0.05), y = seq(11,230, 10), xmean =
-            cov.info['mean', 'sg100c_avg'], xstd = cov.info['std', 'sg100c_avg'], 
-          ymean = cov.info['mean', 'dbh'], ystd = cov.info['std', 'dbh'], 
-          xlab = 'WD', ylab= 'DBH', zlab = 'Prob dying')
-  
-  #dbh.gr
-  coeff.plot(dbh.gr.mod, c('b.3', 'b.2','b.1','GM'), labels = c('WD x DBH.gr','DBH.gr','WD','GM'), 
-             xlab = 'cloglog (effect size)')
-  plot.3d(jagsfit = dbh.gr.mod, x = seq(0.2,0.8,0.05), y = seq(0,5, 0.5), xmean =
-            cov.info['mean', 'sg100c_avg'], xstd = cov.info['std', 'sg100c_avg'], 
-          ymean = cov.info['mean', 'dbh.gr'], ystd = cov.info['std', 'dbh.gr'], 
-          xlab = 'WD', ylab= 'DBH.gr', zlab = 'Prob dying')
-  
-  #dbh.rgr
-  coeff.plot(dbh.rgr.mod, c('b.3', 'b.2','b.1','GM'), labels = c('WD x DBH.gr','DBH.rgr','WD','GM'), 
-             xlab = 'cloglog (effect size)')
[…]

### 2014-11-16 proposal.Rmd
Shift to maker workflow
Added changes

--- 2014-11-12 ms/Mortality Proposal.Rmd
@@
 output: pdf_document
+bibliography: data/refs.bib
+csl: downloads/style.csl
 ---
@@
 
+[@kleinbaum_survival_2005]
+
 Mortality rates vary both spatially and temporally, and as such, are critical to accurately predict vegetation dynamics.
-However, few models of vegetation dynamics explicitly account for changes in mortality rates (e.g.
-REFS).
-Instead, most models used simplifed assumptions such as constant mortality (REF), no mortality (REF) or that mortality varies as a function of some arbitary set of rules (REF).
-These simplifcations are predominately a consequence of our limited understanding of what governs mortality rates and how it varies temporally, spatially and within and between species.
-This limited knowledge is partly due to ecologists placing greater emphasis on examining plant growth and how it affects community (REF) and population dynamics (REF).
-But also because of difficulties associated with modelling mortality.
-Mortality is highly variable and episodic (REF).
-As such, it requires individuals to be monitoried for sufficiently long periods of time, ideally from birth to death, which is problematic for long-lived species.
-Mortality rates are also affected by multiple biotic and abiotic factors, as well as the frequency (REF) and severity (REF) of multiple hazards (e.g.
-fire, wind, drought), which may or may not be dependent on one another (REF).
+However, few models of vegetation dynamics explicitly account for changes in mortality rates.
+Instead, most models used simplifed assumptions such as constant mortality rates (Thorton 2007m White 2000, Delbart et al 2010) or that mortality varies as a function of some arbitary set of rules or thresholds (Sitch et al 2003, Sato et al 2007).
+In many cases, mortality is not simulated at all (Hawkes 2000).
+These simplifications are predominately a consequence of our limited understanding of what governs mortality rates and how it varies temporally, spatially and within and between species (Hawkes 2000, McDowell et al 2011).
+This limited knowledge is partly due to ecologists placing greater emphasis on examining plant growth and how it affects community and population dynamics (Hawke 2000).
+But also because of difficulties associated with modelling mortality (Hawke 2000).
+Mortality is highly variable and episodic (Burgmann 1994).
+As such, it requires individuals to be monitoried for sufficiently long periods of time, ideally from birth to death, which is problematic for long-lived species (maybe Bugman 1996).
+Mortality rates are also affected by multiple biotic and abiotic factors, as well as the frequency and severity of multiple hazards (e.g.
+fire, wind, drought), which may or may not be dependent on one another (McDowell 2011).
 
-This complexity has led to the widespread use of phenomological models to describe patterns in plant mortality.
-In a review of more than 150 woody plant mortality models, Hawke (2000) found that 93% were phenomological and only 7% were mechanistic; further highlighting the paucity of process-based information about mortality.
-These phenological models....
+This complexity has led to the widespread use of phenomological models that attempt to describe patterns in plant mortality as a function of correlations with various covariates.
+In a review of more than 150 woody plant mortality models, Hawke (2000) found that 93% were phenomological and only 7% were mechanistic.
+Such phemenological models have been used to document how mortality varies as a function of competition (REF), climate (REF), age (REF) and size (REF).
+The effect of size has been one
 
@@
 Friend et al.
-1997, Bossel 1986, Weinstein et al 1991, Bugmann et al 1997).
+1997, Bossel 1986, Weinstein et al 1991, Bugmann et al 1997; Fisher et al 2010).
 As such, none incorporated differential mortality rates among species, or how different hazards may alter these rates.
@@
 
-How many data are needed on mortality?
-When and where should they be ob- tained?
-This is a general criticism of Friend et al.
-(1997), who claim these problems render valida- tion of mechanistic models useless.
-Also, data sources for mortality testing are often flawed.
-Bugmann (1996) points out that monitoring pro- grammes are usually shorter than many plants’ lifespans, while palaeo data does not provide the necessary resolution.
-And the potential variation between sizes, classes, ages etc.
-is so high that even a large sample will not provide the statistical significance required.
+And the potential variation between sizes, classes, ages etc. is so high that even a large sample will not provide the statistical significance required.
 
@@
 This approach is particularly useful for handling time-dependent covariates and fitting non-proportional hazards models because the covariate values can change across each individual's Tij timepoints.
+
+# REFERENCES

### 2014-11-18 proposal.Rmd
Wrote more intro, added template code for JAGS and Stan implementation of growth dependent and independent effects of wood density

@@
 
-Mortality rates vary both spatially and temporally, and as such, are critical to accurately predict vegetation dynamics.
-However, few models of vegetation dynamics explicitly account for changes in mortality rates.
-Instead, most models used simplifed assumptions such as constant mortality rates (Thorton 2007m White 2000, Delbart et al 2010) or that mortality varies as a function of some arbitary set of rules or thresholds (Sitch et al 2003, Sato et al 2007).
-In many cases, mortality is not simulated at all (Hawkes 2000).
-These simplifications are predominately a consequence of our limited understanding of what governs mortality rates and how it varies temporally, spatially and within and between species (Hawkes 2000, McDowell et al 2011).
-This limited knowledge is partly due to ecologists placing greater emphasis on examining plant growth and how it affects community and population dynamics (Hawke 2000).
-But also because of difficulties associated with modelling mortality (Hawke 2000).
-Mortality is highly variable and episodic (Burgmann 1994).
-As such, it requires individuals to be monitoried for sufficiently long periods of time, ideally from birth to death, which is problematic for long-lived species (maybe Bugman 1996).
-Mortality rates are also affected by multiple biotic and abiotic factors, as well as the frequency and severity of multiple hazards (e.g.
-fire, wind, drought), which may or may not be dependent on one another (McDowell 2011).
+Plant mortality and the processes affecting it are integral to understanding vegetation dynamics.
+However, few vegetation dynamic models explicitly simulate mortality (Hawkes 2000), and those that have, do so using little process-based information.
+For example, many models that account for plant mortality by assuming mortality rates are constant (Thorton 2007m White 2000, Delbart et al 2010) or that they change when a threshold has been surpassed (Sitch et al 2003, Sato et al 2007).
+These simplifications are a result of a lack of process-based information on what causes mortality rates to vary temporally, spatially and within and between species (Hawkes 2000, McDowell et al 2011).
+In part, this lack of process-based knowledge is a consequence of researchers placing greater emphasis on delimiting what influences plant growth (Hawkes 2000).
+But it is also due to the difficulties associated with modelling and estimating mortality rates.
 
-This complexity has led to the widespread use of phenomological models that attempt to describe patterns in plant mortality as a function of correlations with various covariates.
-In a review of more than 150 woody plant mortality models, Hawke (2000) found that 93% were phenomological and only 7% were mechanistic.
-Such phemenological models have been used to document how mortality varies as a function of competition (REF), climate (REF), age (REF) and size (REF).
-The effect of size has been one
+Temporal patterns in plant mortality are highly variable and episodic (Burgmann 1994).
+Mortality rates also vary between species and are strongly influenced by multiple biotic and abiotic factors that may interact with one another (e.g.
+drought and herbivory: McDowell 2011).
+Because of this complexity, improving process-based understanding of mortality requires monitoring large numbers of species and individuals across entire life-cycles and across multiple biotic and abiotic gradients.
+More importantly, it requires the cause of death to be known for each individual.
+However, few, if any, data meet all these requirements.
+Rather, most available data comes from monitoring programs that are usually shorter than many plants lifespans (Bugmann 1996), and that focus on single species at single sites.
+Furthermore, because of the inherent difficulty of defining the cause of death, most monitoring programs do not delimit the cause of death for each individual.
 
-These mechanistic models all assumed that mortality occured when carbon supply was insufficient for growth (e.g.
+Consequently, there are very few mechanistic mortality models.
+In a review of 61 woody plant mortality models, Hawke (2000) found that only 4 were mechanistic and all these were concerned with carbon budgets.
+In these models mortality occurred when carbon supply was insufficient for growth (e.g.
 Friend et al.
 1997, Bossel 1986, Weinstein et al 1991, Bugmann et al 1997; Fisher et al 2010).
-As such, none incorporated differential mortality rates among species, or how different hazards may alter these rates.
-Phenomological model on the other hand, have attempted to explain mortality as a function of competitive interactions, abiotic conditions as well as size.
-One finding that appears to
+By contrast, the remaining 57 models were empirical and included forest yield models (Guan and Gertner 1991), gap models (Dale et al 1985), plant-environment models (King and Grant 1996) and spatial models (Jeltsch et al 1997).
+These empirical mortality models could be either deterministic (REFS) or stochastic (REFS) and typically included correlations of one or multiple, abiotic and biotic processes such as competition (REF), climate (REF), age (REF), size (REF) and growth rate (REF).
 
-And the potential variation between sizes, classes, ages etc. is so high that even a large sample will not provide the statistical significance required.
+More recently, researchers have used plant traits to determine why some species are more predispositioned to die under particular conditions than others.
 
-Mechanistic models are constructed with al- gorithms that describe biological processes, as opposed to an empirical data set or predefined relationship.
-Thus mechanistic algorithms require no calibration, and are inherently deterministic.
-
-Furthermore, ecologists must also asertain when an individual died and what caused its death, the determination of which is inherently difficult in plants.
+One consistent finding among several phenomological models is that mortality rates tend to be U-shaped.
+That is, small and large plants are more prone to mortality relative to intermediate sized plants.
+However, the problem with these models is that most have been fitted for single species and thus may not accurately predict mortality in other species.
 
@@
 These include running seperate models for small and large plants, the use of polynomials, or the use of non-parameteric approaches that do not assume an underlying curve.
-- They have often constructed these models to determine the effects of competition, size or ...
 - However, the problem with such approaches is that they do not attempt to explain the underlying mechanisms that result in U-shaped mortality.
@@
 
-1) Scour the literature on how traits and size influence mortality rates under common hazards
-2) Build an integrated process-based model that incorporates multiple hazards that can be used to determine how the combination of size and traits governs instantenous mortality rates.
-3) Apply model to multiple datasets.
+1) Examine whether the effects of wood density are growth dependent or growth independent.
+#Current JAGS model
+```{r, eval=FALSE}
+model <- function(x) {
+  for(it in 1:n.obs) {
+
+		died[it] ~ dbern(p[it])
+		cloglog(p[it]) <- max(-9999, min(9999,  ln.census + log(cd0*exp(-cd1 * wd[spp[it]]) + cd2 * exp(-cd3 * dbh.gr[it]))))
+	}
+
+	#Priors
+  cd0 ~ dlnorm(0, 0.0001)
+  cd1 ~ dlnorm(0, 0.0001)
+	cd2 ~ dlnorm(0, 0.0001)
+	cd3 ~ dlnorm(0, 0.0001)
+
+}
+    test2<-     jags.parallel(model.file = model,
+                 n.iter = 4000,
+                 data= names(data), 
+                 envir=list2env(data),
+                 inits = NULL,
+                 parameters.to.save = c('cd0','cd1','cd2','cd3'))
+```
+
+#Potential Stan Model
+```{r, eval=FALSE}
+library(parallel)
+library(rstan)
+
+cat('
+  data {
+		int<lower=0> N;
+		int<lower=0, upper=1> M[N];
+		vector[N] ln_T;
+		vector[N] wd;
+		vector[N] gr;
+
+	}
+	parameters {
+		real<lower=0> c0;
+		real c1;
+		real<lower=0> c2;
+		real<lower=0> c3;
+
+	} 
+	model {
+		for (n in 1:N)
+			M[n] ~ bernoulli(inv_cloglog(ln_T + log(c0 * exp(-c1 * wd[n]) + c2 * exp(-c3 * gr[n]))));
+	}
+', file=(StanModel <- tempfile())
+
+BCImodel <- sflist2stanfit(
+	mclapply(
+		1:4,
+		function(i) {
+			stan(file=StanModel, data=BCIdata, seed=123, chains=1, chain_id=i, refresh=-1)
+			}, 
+		mc.cores=2
+	)
+)
+```
+
+2) Examine whether the effects of wood density varies with plant size.
+3) Write a literature review summarising how plant traits influence mortality across multiple hazards (e.g. wind throw, drought, fire, disease, herbivory)
+4) Build a process-based model based on what we have learnt from the above objectives.
 

### 2014-11-19 proposal.Rmd
Tweaked JAGS and Stan models of growth dependent and independent wood density effects. Previous model failed to converge do to incorrect priors on coefficients. Hopefully this fixes the issue

@@
 
-[@kleinbaum_survival_2005]
-
 Plant mortality and the processes affecting it are integral to understanding vegetation dynamics.
-However, few vegetation dynamic models explicitly simulate mortality (Hawkes 2000), and those that have, do so using little process-based information.
-For example, many models that account for plant mortality by assuming mortality rates are constant (Thorton 2007m White 2000, Delbart et al 2010) or that they change when a threshold has been surpassed (Sitch et al 2003, Sato et al 2007).
-These simplifications are a result of a lack of process-based information on what causes mortality rates to vary temporally, spatially and within and between species (Hawkes 2000, McDowell et al 2011).
-In part, this lack of process-based knowledge is a consequence of researchers placing greater emphasis on delimiting what influences plant growth (Hawkes 2000).
-But it is also due to the difficulties associated with modelling and estimating mortality rates.
+However, few mechanistic models of vegetation dynamic explicitly simulate mortality [@Hawkes_2000], and those that have, do so using little process-based information.
+For example, mechanistic models have accounted for plant mortality by assuming mortality rates are constant through time and space (Thorton 2007m White 2000, Delbart et al 2010) or that they change when a threshold has been surpassed (e.g.
+when Sitch et al 2003, Sato et al 2007).
+These simplifications are a result of a lack of process-based information on what causes mortality rates to vary temporally, spatially and between species (Hawkes 2000, McDowell et al 2011).
+In part, this is because researchers have placed a greater emphasis on modelling plant growth (Hawkes 2000).
+But it is also due to the difficulties associated with modelling mortality rates.
 
@@
 drought and herbivory: McDowell 2011).
-Because of this complexity, improving process-based understanding of mortality requires monitoring large numbers of species and individuals across entire life-cycles and across multiple biotic and abiotic gradients.
+Because of this complexity, calibrating empirical models or validating mechanistic models of plant mortality requires datasets that monitor large numbers of species and individuals across their lifespan and across multiple biotic and abiotic gradients.
 More importantly, it requires the cause of death to be known for each individual.
@@
 Consequently, there are very few mechanistic mortality models.
-In a review of 61 woody plant mortality models, Hawke (2000) found that only 4 were mechanistic and all these were concerned with carbon budgets.
-In these models mortality occurred when carbon supply was insufficient for growth (e.g.
+In a review of 61 woody plant mortality models, Hawke (2000) found that only 4 were mechanistic.
+In these models, mortality occurred when carbon supply was insufficient for growth (e.g.
 Friend et al.
-1997, Bossel 1986, Weinstein et al 1991, Bugmann et al 1997; Fisher et al 2010).
-By contrast, the remaining 57 models were empirical and included forest yield models (Guan and Gertner 1991), gap models (Dale et al 1985), plant-environment models (King and Grant 1996) and spatial models (Jeltsch et al 1997).
-These empirical mortality models could be either deterministic (REFS) or stochastic (REFS) and typically included correlations of one or multiple, abiotic and biotic processes such as competition (REF), climate (REF), age (REF), size (REF) and growth rate (REF).
+1997, Bossel 1986, Weinstein et al 1991, Bugmann et al 1997).
+By contrast, the majority of mortality models (57) were empirical and included forest yield models (Guan and Gertner 1991), gap models (Dale et al 1985), plant-environment models (King and Grant 1996) and spatial models (Jeltsch et al 1997).
+These empirical mortality models spanned the range from deterministic (REFS) to stochastic (REFS) and typically included correlations of one or multiple, abiotic and biotic factors such as competition (REF) and climate (REF) as well as state variables such as age (REF), size (REF) and growth rate (REF).
 
-More recently, researchers have used plant traits to determine why some species are more predispositioned to die under particular conditions than others.
+More recently, researchers have used plant traits in an attempt to explain inter-specific variability in mortality rates. For example, ....
+However, to date, most of these studies have assumed that these trait effects are constant through time and do not interact with changing environmental conditions or individual state variables such as growth rate or size.
 
-One consistent finding among several phenomological models is that mortality rates tend to be U-shaped.
-That is, small and large plants are more prone to mortality relative to intermediate sized plants.
-However, the problem with these models is that most have been fitted for single species and thus may not accurately predict mortality in other species.
+###Aims
+Based on these knowledge gaps I have three primary objectives:
+1) Examine whether the effects of wood density (and other traits) on mortality are growth & size dependent or independent.
+2) Conduct a literature review summarising how plant traits influence mortality rates under common hazards (e.g. wind throw, drought, fire, disease, herbivory).
+3) Build a process-based mortality model that incorporating the common hazard experienced by plants and how traits mediate their effects.
 
-#How have people attempted modelling mortality?
-- In order to capture the U-shaped mortality researchers have used a variety of models to capture broad patterns.
-These include running seperate models for small and large plants, the use of polynomials, or the use of non-parameteric approaches that do not assume an underlying curve.
-- However, the problem with such approaches is that they do not attempt to explain the underlying mechanisms that result in U-shaped mortality.
-- Some have suggested that certain hazards influence different parts of the U-shape curve.
-For example, competitive process influence mortality when individuals are small and less so when they are large.
-Whereas the oppose is true for events such as windthrow.
-- Furthermore, the vast majority of these empirical models have been conducted for individual species, and thus are of limited use for more applying to applying to other species.
+##Growth dependent & independent effects of wood density
 
-#What do we know about mortality in plants?
-Evidence suggests that mortality is U-shaped for most plants.
-We also know that plant survival is influenced by multiple hazards (fire, herbivory, climate, drought, competition and windthrow)
+###Example JAGS model with no random effects
 
-#Traits
-More recently ecologists have used traits in an attempt to obtain more generality.
-The vast majority of this research has focused on how traits influence growth rates, with very few studies examining trait effects on mortality.
-Those that have, have reported that traits such as wood density can significantly influence rates of mortality.
-The logic here is that wood density conveys additional resources into structural properties and thus makes plants more resistant to breakage and invertebrate attack.
-However, as far as I am aware, all these studies have assumed that the effect of traits remains constant and does not change with size.
+```{r, eval=FALSE}
+library('dplyr')
+library('R2jags')
+#Loads BCI data and attaches trait data
+load('../data_BCI/output/bci.mainstem.Rdata')
+bci.traits <- read.csv('../data_BCI/rawdata/traits/BCI_traits_20101220.csv')
+names(bci.traits) <- tolower(names(bci.traits)) # lowers trait column names for merging
+bci.traits$sp <- tolower(bci.traits$sp) # lowers species code names for merging
+bci.mainstem <- merge(bci.mainstem,bci.traits[,c('sp','sg100c_avg')],by = 'sp') #only uses species trait data exists for.
+rm(bci.traits) # trait db no longer needed
 
-#Solution
+#Subsets data to only include species with traits and has a status record.
+#Also removes couple of rows that have 10 yr census intervals.
+#Then adds a sp.id column
+bci.mainstem<- bci.mainstem %>%
+  filter(!is.na(census.interval) & !is.na(sg100c_avg) & !is.na(dead.next.census) & census.interval< 6) %>%
+  mutate(sp.id = as.numeric(factor(sp)))
 
-Many modellers have called for more mechanistic al- gorithms in plant growth models and see the path
-towards improvement as a reformulation of mor- tality algorithms to better reflect the physiological base of a plant (e.g. Loehle and LeBlanc, 1996). (Hawke)
+  #Preps data for model by centering and scaling predictors and time
+  data <- list(
+  n_obs = nrow(bci.mainstem),
+  n_spp = length(unique(bci.mainstem$sp)),
+  spp = as.numeric(factor(bci.mainstem$sp), as.character(unique(bci.mainstem$sp))),
+  wd = scale(unique(bci.mainstem[c('sp', 'sg100c_avg')])[,'sg100c_avg'])[,1]/2,
+  dbh = scale(bci.mainstem$dbh)[,1]/2,
+  dbh_gr = scale(bci.mainstem$dbh.gr)[,1]/2,
+  dbh_rgr = scale(bci.mainstem$dbh.rgr)[,1]/2,
+  basal_area = scale(bci.mainstem$basal.area)[,1]/2,
+  basal_gr = scale(bci.mainstem$basal.area.gr)[,1]/2,
+  basal_rgr = scale(bci.mainstem$basal.area.rgr)[,1]/2,
+  ln_census = scale(log(bci.mainstem$census.interval))[,1]/2, #logged for model
+  died = bci.mainstem$dead.next.census)
 
-Based on the information above I have three primary objectives:
-
-1) Examine whether the effects of wood density are growth dependent or growth independent.
-#Current JAGS model
-```{r, eval=FALSE}
+#Actual model
 model <- function(x) {
-  for(it in 1:n.obs) {
+  for(it in 1:n_obs) {
 
     died[it] ~ dbern(p[it])
-		cloglog(p[it]) <- max(-9999, min(9999,  ln.census + log(cd0*exp(-cd1 * wd[spp[it]]) + cd2 * exp(-cd3 * dbh.gr[it]))))
+    cloglog(p[it]) <- max(-9999, min(9999,  ln_census[it] + log(cd0*exp(-cd1 * wd[spp[it]]) + cd2 * exp(-cd3 * wd[spp[it]] * dbh_gr[it]))))
     }
@@
   #Priors
-  cd0 ~ dlnorm(0, 0.0001)
-  cd1 ~ dlnorm(0, 0.0001)
-	cd2 ~ dlnorm(0, 0.0001)
-	cd3 ~ dlnorm(0, 0.0001)
+  cd0 ~ dlnorm(0, 0.0001) # Must be positive or large growth rates are undefined.
+  cd1 ~ dnorm(0, 0.0001) # Can be positive or negative (but theory would suggest positive)
+  cd2 ~ dlnorm(0, 0.0001) # Has to be positive or small growth rates not defined
+  cd3 ~ dnorm(0, 0.0001) # Can be positive or negative
 
   }
-    test2<-     jags.parallel(model.file = model,
+mod<-     jags.parallel(model.file = model,
                         n.iter = 4000,
@@
   data {
-		int<lower=0> N;
-		int<lower=0, upper=1> M[N];
-		vector[N] ln_T;
-		vector[N] wd;
-		vector[N] gr;
+		int<lower=0> n_obs;
+		int<lower=0, upper=1> died[n_obs];
+		vector[n_obs] ln_T;
+		vector[n_obs] wd;
+		vector[n_obs] gr;
 
@@
 		real<lower=0> c2;
-		real<lower=0> c3;
+		real c3;
 
@@
 	model {
-		for (n in 1:N)
-			M[n] ~ bernoulli(inv_cloglog(ln_T + log(c0 * exp(-c1 * wd[n]) + c2 * exp(-c3 * gr[n]))));
+		for (it in 1:n_obs)
+			died[it] ~ bernoulli(inv_cloglog(ln_census[it] + log(c0 * exp(-c1 * wd[it]) + c2 * exp(-c3 * dbh_gr[it]))));
 	}
@@
 
-#Objective 1
-
-As far as I am aware, there does not appear to be an integrated review of how traits influence various hazards, or how they vary with size.
-Most trait analysis of mortality has either ignored the cause of death or has been related to very specific forms of mortality.
-For example, there are several studies that highlight suite of traits that minimise death by fire, but less so on other hazards such as wind throw and climatatic extremes.
-In order to remedy this.
-I am to write a brief literature review highlighting the primary hazards plants are exposed to and how various traits have been linked to decreased or increased susceptibility.
-
-In order to achieve this I plan to use google scholar and...
-
-#Objective 2
-Include maths here
-
-#Objective 3
-Preliminary thoughts
-
-Modelling survival and mortality within a population is of fundamental importance in ecology, epidemiology, demography, engineering and economics. It is particularly
-
-Burgman et al.
-(1994) points out the significance of tree mortality for forest yield and growth rates, and Pacala et al.
-(1996) shows the importance of species mortality rates in forest succession.What is clear is that mortality is very difficult to model owing to its variable and episodic nature (Burgman et al., 1994) (Hawke)
-
-#Difficulty in modelling mortality
-The complexity of environmental stresses and lack of process information of woody plant mortality has however led to the wide use of empirical algorithms.These have many problems, including data intensity and spatial and temporal specificity.
-Such problems render mortality simulation prone to error and weak in response to environmental change.
-(Hawke)
-
-Describing size-specific mortality patterns can be especially difficult, due to few trees in larger size classes. (Hurst)
-
-###Aim:
-1. Determine traits influence the likelihood of a particular hazard resulting in death.
-2. Determine whether instantaneous mortality rates are commonly U-shaped as a function of dbh.
-3. Examine whether trait effects on instantaneous mortality rates vary with individual size and growth rate
-4. Build a process-based mortality model that encapsulates how the combination of various hazards and traits can result in various mortality curves.
-
-###Hypotheses
+### Hazard Hypotheses
 1. The most common hazards that can result in plant death are all size dependent and include:
@@
 
-**Drought/Extreme heat/ Extreme cold**: These hazard is likely to..
+**Drought/Extreme heat/ Extreme cold**: These hazard is likely to affect smaller plants moreso than larger plants.
 
@@
 
-y = log(-log(1-P))
-
-y = log(e) + log(1/(1-P)))
-e^y = (1/(1-P))
-e^y*(1-P) = 1
-e^y - Pe^y = 1
-
--e(e^y) + 1
-
-1 - e(e^y)
+#Random notes
 

### 2014-11-20 proposal.Rmd
Added parallel chains to stan model.

@@
 By contrast, the majority of mortality models (57) were empirical and included forest yield models (Guan and Gertner 1991), gap models (Dale et al 1985), plant-environment models (King and Grant 1996) and spatial models (Jeltsch et al 1997).
-These empirical mortality models spanned the range from deterministic (REFS) to stochastic (REFS) and typically included correlations of one or multiple, abiotic and biotic factors such as competition (REF) and climate (REF) as well as state variables such as age (REF), size (REF) and growth rate (REF).
+These empirical mortality models spanned the range from deterministic (REFS) to stochastic (REFS) and typically included correlations of one or multiple, abiotic and biotic factors such as competition (REF) and climate (REF) as well as state variables such as age (REF), size (Hurst et al 2011, , V) and growth rate (REF).
 
-More recently, researchers have used plant traits in an attempt to explain inter-specific variability in mortality rates. For example, ....
-However, to date, most of these studies have assumed that these trait effects are constant through time and do not interact with changing environmental conditions or individual state variables such as growth rate or size.
+More recently, plant traits have been incorporated into empirical models in order to examine inter-specific variability in mortality rates.
+These studies have suggest that a several traits may reduce mortality rates.
+For example, several studies have found that species with higher wood density exhibit reduced baseline mortality rates.
+This is thought to be due to higher wood density resulting in increased structural support, which in turn reduces the chance of breakage but also herbivory by wood boring herbivores.
+Other traits have been found to convey mortality advantages under particular hazards.
+For example, bark thickness has been correlated with greater survival after fire due to the bark insulating the sapwood from extreme temperatures.
+Or having a deciduous life form often allows for greater survival during prolonged drought by allowing plants to essentially hibernate during periods of high water stress, relative to evergreen species.
+However, the vast majority of these studies assume that the effects of traits remain do not interact with changing environmental conditions or individual state variables such as growth rate or size.
+This is despite theory suggesting otherwise.
+For example wood density is likely to matter more for an adult plant then a juvenile plant because adults are taller and are more prone to being knocked over by windthrow, whereas juveniles are likely to be more sheltered by surrounding taller vegetation.
 
 ###Aims
-Based on these knowledge gaps I have three primary objectives:
+In this project I aim to address three knowledge gaps in relation to traits and mortality rates.
+
 1) Examine whether the effects of wood density (and other traits) on mortality are growth & size dependent or independent.
 2) Conduct a literature review summarising how plant traits influence mortality rates under common hazards (e.g. wind throw, drought, fire, disease, herbivory).
-3) Build a process-based mortality model that incorporating the common hazard experienced by plants and how traits mediate their effects.
+3) Build a process-based mortality model that allows trait effects to interact with the type of hazard as well as growth and size.
+
+##1 - Growth dependent & independent effects of wood density
+
+The effect of wood density on plant survival is often fitted to empiricial models under the assumption that it is independent of state variables such as growth rate and size (i.e.
+models with no interaction terms).
+Here, the theory is that higher wood density equates to lower mortality rates regardless of state variables such as growth rate and size.
+However, wood density may actually have both dependent and independent effects on plant mortality.
+For example, higher wood density may reduce mortality rates via greater stem structural support and decreased wood herbivory regardless of other state and climatic variables.
+However, it may also interact with growth rates with high wood density and low or negative growth rates traits such as wood density may actually density may have both the effect of wood density is unl
 
 ##Growth dependent & independent effects of wood density
+# Current model may not be ideal. Partly because WD can never be zero and thus the model is unrealistic. Centering WD also won't work as it changes the form of the model by making it multiplicative.
+
+Potential solution is to use a different form for wood density (e.g. csWD^c1)
+Explore other model forms.
+Sub-questions:
+1) What is the best growth rate metric to use? dbh, basal area?, mass?
+2) Can we partition the effects of size and growth?
+3) What model best explains the effect mortality?
+4) Think about size related models that encapsulate increases at small and large sizes.
+RGR decreases with size. Incr.GR increases with size. By including both growth rate and size in the model we can determine whether size has additional affects that are not explained by growth.
 
@@
   n_obs = nrow(bci.mainstem),
-  n_spp = length(unique(bci.mainstem$sp)),
-  spp = as.numeric(factor(bci.mainstem$sp), as.character(unique(bci.mainstem$sp))),
-  wd = scale(unique(bci.mainstem[c('sp', 'sg100c_avg')])[,'sg100c_avg'])[,1]/2,
-  dbh = scale(bci.mainstem$dbh)[,1]/2,
-  dbh_gr = scale(bci.mainstem$dbh.gr)[,1]/2,
-  dbh_rgr = scale(bci.mainstem$dbh.rgr)[,1]/2,
-  basal_area = scale(bci.mainstem$basal.area)[,1]/2,
+  #n_spp = length(unique(bci.mainstem$sp)),
+  #pp = as.numeric(factor(bci.mainstem$sp), as.character(unique(bci.mainstem$sp))),
+  wd = unique(bci.mainstem[c('sp', 'sg100c_avg')])[,'sg100c_avg'],
+  #dbh = scale(bci.mainstem$dbh)[,1]/2,
+  #dbh_gr = bci.mainstem$dbh.gr,
+  #dbh_rgr = scale(bci.mainstem$dbh.rgr)[,1]/2,
+  #basal_area = scale(bci.mainstem$basal.area)[,1]/2,
   basal_gr = scale(bci.mainstem$basal.area.gr)[,1]/2,
-  basal_rgr = scale(bci.mainstem$basal.area.rgr)[,1]/2,
-  ln_census = scale(log(bci.mainstem$census.interval))[,1]/2, #logged for model
+  #basal_rgr = scale(bci.mainstem$basal.area.rgr)[,1]/2,
+  ln_census = log(bci.mainstem$census.interval), #logged for model
   died = bci.mainstem$dead.next.census)
@@
 library(rstan)
-
+data <- list(
+  n_obs = nrow(bci.mainstem),
+  #n_spp = length(unique(bci.mainstem$sp)),
+  #pp = as.numeric(factor(bci.mainstem$sp), as.character(unique(bci.mainstem$sp))),
+  wd = bci.mainstem$sg100c_avg,
+  #dbh = scale(bci.mainstem$dbh)[,1]/2,
+  #dbh_gr = bci.mainstem$dbh.gr,
+  #dbh_rgr = bci.mainstem$dbh.rgr,
+  #basal_area = scale(bci.mainstem$basal.area)[,1]/2,
+  basal_gr = scale(bci.mainstem$basal.area.gr,center = FALSE)[,1]/2,
+  #basal_rgr = scale(bci.mainstem$basal.area.rgr)[,1]/2,
+  census_length = bci.mainstem$census.interval, #logged for model
+  died = bci.mainstem$dead.next.census)
 cat('
@@
     int<lower=0, upper=1> died[n_obs];
-		vector[n_obs] ln_T;
+    vector[n_obs] census_length;
     vector[n_obs] wd;
-		vector[n_obs] gr;
+    vector[n_obs] basal_gr;
 
@@
     for (it in 1:n_obs)
-			died[it] ~ bernoulli(inv_cloglog(ln_census[it] + log(c0 * exp(-c1 * wd[it]) + c2 * exp(-c3 * dbh_gr[it]))));
+    died[it] ~ bernoulli(inv_cloglog(log(census_length[it]) + log(c0 * exp(-c1 * wd[it]) + c2 * exp(-c3 * basal_gr[it]))));
     }
-', file=(StanModel <- tempfile())
+    ', file=(StanModel <- tempfile()))
 
-BCImodel <- sflist2stanfit(
-	mclapply(
-		1:4,
-		function(i) {
-			stan(file=StanModel, data=BCIdata, seed=123, chains=1, chain_id=i, refresh=-1)
-			}, 
-		mc.cores=2
-	)
-)
+system.time(BCImodelbasal.gr <- sflist2stanfit(mclapply(1:1,mc.cores=1,
+		function(i) stan(file=StanModel, data=data, inter=500,seed=123, chains=1, chain_id=i, refresh=-1))))
 ```

### 2014-11-28 proposal.Rmd
Added potential state dependent and independent equations. Also added more text. Still requires fleshing out some of the biological interpretation behind the maths.

--- 2014-11-19 proposal.Rmd
@@
 ---
-title: "Mortality Notes"
+title: "Mortality Project"
 author: "J Camac"
@@
 Plant mortality and the processes affecting it are integral to understanding vegetation dynamics.
-However, few mechanistic models of vegetation dynamic explicitly simulate mortality [@Hawkes_2000], and those that have, do so using little process-based information.
+However, few mechanistic models of vegetation dynamic explicitly simulate mortality (Hawkes_2000), and those that have, do so using little process-based information.
 For example, mechanistic models have accounted for plant mortality by assuming mortality rates are constant through time and space (Thorton 2007m White 2000, Delbart et al 2010) or that they change when a threshold has been surpassed (e.g.
@@
 By contrast, the majority of mortality models (57) were empirical and included forest yield models (Guan and Gertner 1991), gap models (Dale et al 1985), plant-environment models (King and Grant 1996) and spatial models (Jeltsch et al 1997).
-These empirical mortality models spanned the range from deterministic (REFS) to stochastic (REFS) and typically included correlations of one or multiple, abiotic and biotic factors such as competition (REF) and climate (REF) as well as state variables such as age (REF), size (REF) and growth rate (REF).
+These empirical mortality models spanned the range from deterministic (REFS) to stochastic (REFS) and typically included correlations of one or multiple, abiotic and biotic factors such as competition (REF) and climate (REF) as well as state variables such as age (REF), size (Hurst et al 2011, , Vieilledent et al.
+2009, Yang et al.
+2003) and growth rate (Vielledent et al.
+2010, Wright et al 2010).
 
-More recently, researchers have used plant traits in an attempt to explain inter-specific variability in mortality rates. For example, ....
-However, to date, most of these studies have assumed that these trait effects are constant through time and do not interact with changing environmental conditions or individual state variables such as growth rate or size.
+More recently, plant traits have been incorporated into empirical models in order to examine inter-specific variability in mortality rates (e.g.
+Kitajima et al 2013, Wright et al.
+2010).
+Such studies have found several traits that are correlated with mortality rates.
+For example, several studies (Wright et al 2010) have found that species with higher wood density exhibit lower mortality rates relative to species with low wood density.
+This is thought to be due to higher construction costs resulting in stronger stems and branches that are less susceptible to breakage and wood herbivory (REFS).
+Bark thickness has also been correlated with greater survival after fire due via greater insulating properties (Pausas et al in press).
+Or having a deciduous life form often allows for greater survival during prolonged drought by allowing plants to essentially hibernate during periods of high water stress, relative to evergreen species.
+However, the vast majority of these studies assume that the effects of traits remain do not interact with changing environmental conditions or individual state variables such as growth rate or size.
+This is despite theory suggesting otherwise.
+For example wood density is likely to matter more for an adult plant then a juvenile plant because adults are taller and are more prone to being knocked over by windthrow, whereas juveniles are likely to be more sheltered by surrounding taller vegetation.
 
-###Aims
-Based on these knowledge gaps I have three primary objectives:
-1) Examine whether the effects of wood density (and other traits) on mortality are growth & size dependent or independent.
-2) Conduct a literature review summarising how plant traits influence mortality rates under common hazards (e.g. wind throw, drought, fire, disease, herbivory).
-3) Build a process-based mortality model that incorporating the common hazard experienced by plants and how traits mediate their effects.
+Below I briefly outline three projects that attempt to improve our understanding of how plant traits can influence mortality rates.
 
-##Growth dependent & independent effects of wood density
+**PROJECT 1.**
+Examine whether the effects of wood density (and other traits) on mortality are growth or size dependent
+
+**PROJECT 2.**
+Examine which plant traits influence mortality rates under common hazards (e.g. wind throw, drought, fire, disease, herbivory) experienced by vegetation.
+
+**PROJECT 3.**
+Build a process-based mortality model that allows trait effects to interact with the type of hazard as well as growth and size.
+
+##PROJECT 1 - Growth-Size independent and interactive effects of wood density
+
+###Brief summary
+Plant traits are often fitted to empirical mortality models without the examination of how they may interact with state variables such as size and growth rate.
+However, traits such as wood density are likely to influence mortality rates by both state-independent and state-dependent effects.
+For example, wood density may reduce wood herbivory irrespective of plant size.
+But it may also have additional effects on mortality rates that only become biologically significant as a plant gets larger.
+Higher wood density, for example, may reduce stem breakage; a hazard more likely to occur as a plant gets taller and more exposed to wind.
+Similarly, wood density may also interact with growth rates, whereby individuals with low growth rates and low wood density may be more susceptible to pathogens relative to individuals that have higher wood density.
+
+Here, I examine how plant traits affect mortality rates by partitioning effects into growth or size-independent effects and growth or size-dependent effects.
+
+Sub-questions:
+1) What is the best growth rate metric to use? dbh, basal area?, mass?
+2) Are trait effects predominately growth dependent or independent?
+3) Out of these series of hazards, which combination best predicts plant mortality in BCI?
+RGR decreases with size. Incr.GR increases with size. By including both growth rate and size in the model we can determine whether size has additional affects that are not explained by growth.
+
+####
+# Series of hazards
+GROWTH DEPENDENT & INDEPENDENT EFFECTS
+
+\begin{equation}\label{eqn:wd_grmod_inter}
+\overbrace{\log(-\log(S(i,t)))}^{\text{Survival Probability}} =
+\overbrace{\log(\Delta{t})}^{\text{census interval}} + \log\left(\lambda_0 + \lambda_1 + \lambda_2 ... \lambda_n\right). 
+\end{equation}
+
+\begin{equation}\label{eqn:indhaz}\begin{array}{ll}
+\lambda_0 = &
+\overbrace{\alpha_0}^{\text{baseline/unexplained}} 
+\\
+\\
+
+\lambda_1 = &
+\overbrace{\alpha_1 \times \exp{(-\beta_1 \times gr)}}^{\text{growth dependent}} 
+\\
+\\
+
+\lambda_2 = &
+\overbrace{\alpha_2 \times \exp{(-\beta_2 \times \log(wd))}}^{\text{wd dependent}} 
+\\
+\\
+
+\lambda_3 = &
+\overbrace{\alpha_3 \times \exp{(-\beta_3 \times \log(wd) + -\beta_4 \times gr)}}^{\text{wd independent and dependent}} 
+\\
+\\
+
+SIZE DEPENDENCY
+
+\begin{equation}\label{eqn:wd_grmod_inter}
+\overbrace{\log(-\log(S(i,t)))}^{\text{Survival Probability}} =
+\overbrace{\log(\Delta{t})}^{\text{census interval}} + \log\left(\lambda_1 + \lambda_2 ... \lambda_n\right). 
+\end{equation}
+
+\begin{equation}\label{eqn:indhaz}\begin{array}{ll}
+\lambda_1 = &
+\overbrace{\alpha_0}^{\text{baseline/unexplained}} 
+\\
+\\
+
+\lambda_2 = &
+\overbrace{\alpha_1 \times \exp{(-\beta_1 \times \log(size))}}^{\text{negative effect of size}} 
+\\
+\\
+
+\lambda_3 = &
+\overbrace{\alpha_2 \times \exp{(\beta_2 \times \log(wd))}}^{\text{positive effect of size}} 
+\\
+\\
+\lambda_4 = &
+\overbrace{\alpha_3 \times \exp{(-\beta_3 \times \log(wd))}}^{\text{wd effect}} 
+\\
+\\
+\lambda_5 = &
+\overbrace{\alpha_4 \times \exp{(-\beta_4 \times \log(size) + -beta_5 \times \log(wd))}}^{\text{wd-size positive effect}} 
+\\
+\\
+\lambda_6 = &
+\overbrace{\alpha_5 \times \exp{(\beta_6 \times \log(size) + \beta_7 \times \log(wd))}}^{\text{wd-size negative effect}} 
+\end{array} \end{equation}
+
+**Things to consider:** 
+How to include interactive effects... and what are their meaning?
+- Is it reasonable to assume that growth a size dependent and independent hazards have different baseline hazards?
+- Preliminary analysis suggests that the alpha's are non identifiable when multiple hazards are included.
+- Do not assume a functional form (in this case we assume an negative exponential effect), but instead use some non-parametric approach to give us ideas of about the curves.
+- This is partly done using the K-M plots but are difficult to interpret as typically associated with categorical predictors, whereas we have continuous.
 
@@
   n_obs = nrow(bci.mainstem),
-  n_spp = length(unique(bci.mainstem$sp)),
+  #n_spp = length(unique(bci.mainstem$sp)),
   spp = as.numeric(factor(bci.mainstem$sp), as.character(unique(bci.mainstem$sp))),
-  wd = scale(unique(bci.mainstem[c('sp', 'sg100c_avg')])[,'sg100c_avg'])[,1]/2,
-  dbh = scale(bci.mainstem$dbh)[,1]/2,
+  ln_cswd = scale(log(unique(bci.mainstem[c('sp', 'sg100c_avg')])[,'sg100c_avg']),center=FALSE)[,1]/2,
+  #dbh = scale(bci.mainstem$dbh)[,1]/2,
   dbh_gr = scale(bci.mainstem$dbh.gr)[,1]/2,
-  dbh_rgr = scale(bci.mainstem$dbh.rgr)[,1]/2,
-  basal_area = scale(bci.mainstem$basal.area)[,1]/2,
-  basal_gr = scale(bci.mainstem$basal.area.gr)[,1]/2,
-  basal_rgr = scale(bci.mainstem$basal.area.rgr)[,1]/2,
-  ln_census = scale(log(bci.mainstem$census.interval))[,1]/2, #logged for model
+  #dbh_rgr = scale(bci.mainstem$dbh.rgr)[,1]/2,
+  #basal_area = scale(bci.mainstem$basal.area)[,1]/2,
+  #basal_gr = scale(bci.mainstem$basal.area.gr)[,1]/2,
+  #basal_rgr = scale(bci.mainstem$basal.area.rgr)[,1]/2,
+  census_length = bci.mainstem$census.interval, #logged for model
   died = bci.mainstem$dead.next.census)
@@
     died[it] ~ dbern(p[it])
-    cloglog(p[it]) <- max(-9999, min(9999,  ln_census[it] + log(cd0*exp(-cd1 * wd[spp[it]]) + cd2 * exp(-cd3 * wd[spp[it]] * dbh_gr[it]))))
+    cloglog(p[it]) <- log(census_length[it]) + log(alpha0 + alpha1 * exp(-beta1 * ln_cswd[spp[it]] + alpha2 * exp(-beta2 * dbh_gr[it])))
     }
@@
   #Priors
-  cd0 ~ dlnorm(0, 0.0001) # Must be positive or large growth rates are undefined.
-  cd1 ~ dnorm(0, 0.0001) # Can be positive or negative (but theory would suggest positive)
-  cd2 ~ dlnorm(0, 0.0001) # Has to be positive or small growth rates not defined
-  cd3 ~ dnorm(0, 0.0001) # Can be positive or negative
+  alpha0 ~ dlnorm(0, 0.0001) # Must be positive or large growth rates are undefined.
+  alpha1 ~ dlnorm(0, 0.0001) # Can be positive or negative (but theory would suggest positive)
+  alpha2 ~ dlnorm(0, 0.0001) # Has to be positive or small growth rates not defined
+  beta1 ~ dnorm(0, 0.0001) # Can be positive or negative
+  beta2 ~ dnorm(0, 0.0001)
 
   }
-mod<-     jags.parallel(model.file = model,
-                        n.iter = 4000,
+mod3<-     jags.parallel(model.file = model,
+                        n.iter = 20000,
                         data= names(data), 
@@
                         inits = NULL,
-                        parameters.to.save = c('cd0','cd1','cd2','cd3'))
+                        parameters.to.save = c('alpha0','alpha1','alpha2','beta1','beta2'))
 ```
@@
 library(rstan)
+library(dplyr)
+#Loads BCI data and attaches trait data
+load('../data_BCI/output/bci.mainstem.Rdata')
+bci.traits <- read.csv('../data_BCI/rawdata/traits/BCI_traits_20101220.csv')
+names(bci.traits) <- tolower(names(bci.traits)) # lowers trait column names for merging
+bci.traits$sp <- tolower(bci.traits$sp) # lowers species code names for merging
+bci.mainstem <- merge(bci.mainstem,bci.traits[,c('sp','sg100c_avg')],by = 'sp') #only uses species trait data exists for.
+rm(bci.traits)
 
+#Subsets data to only include species with traits and has a status record.
+#Also removes couple of rows that have 10 yr census intervals.
+#Then adds a sp.id column
+bci.mainstem<- bci.mainstem %>%
+  filter(!is.na(census.interval) & !is.na(sg100c_avg) & !is.na(dead.next.census) & census.interval< 6) %>%
+  mutate(sp.id = as.numeric(factor(sp)))
+
+data <- list(
+  n_obs = nrow(bci.mainstem),
+  #n_spp = length(unique(bci.mainstem$sp)),
+  #pp = as.numeric(factor(bci.mainstem$sp), as.character(unique(bci.mainstem$sp))),
+  wd = scale(log(bci.mainstem$sg100c_avg*1000),center = FALSE)[,1]/2,
+  #dbh = scale(bci.mainstem$dbh)[,1]/2,
+  dbh_gr = scale(bci.mainstem$dbh.gr, center = FALSE)[,1]/2,
+  #dbh_rgr = bci.mainstem$dbh.rgr,
+  #basal_area = scale(bci.mainstem$basal.area)[,1]/2,
+  #basal_gr = scale(bci.mainstem$basal.area.gr,center = FALSE)[,1]/2,
+  #basal_rgr = scale(bci.mainstem$basal.area.rgr)[,1]/2,
+  census_length = bci.mainstem$census.interval, #logged for model
+  died = bci.mainstem$dead.next.census)
 cat('
@@
     int<lower=0, upper=1> died[n_obs];
-		vector[n_obs] ln_T;
+    vector[n_obs] census_length;
     vector[n_obs] wd;
-		vector[n_obs] gr;
+    vector[n_obs] dbh_gr;
 
@@
     parameters {
-		real<lower=0> c0;
-		real c1;
-		real<lower=0> c2;
-		real c3;
-
+    real<lower=0> alpha0;
+    real<lower=0> alpha1;
+    real<lower=0> alpha2;
+    real<lower=0> beta1;
+    real<lower=0> beta2;
     } 
@@
     for (it in 1:n_obs)
-			died[it] ~ bernoulli(inv_cloglog(ln_census[it] + log(c0 * exp(-c1 * wd[it]) + c2 * exp(-c3 * dbh_gr[it]))));
+    died[it] ~ bernoulli(inv_cloglog(log(census_length[it]) + 
+    log(alpha0 + alpha1 * exp(-beta1 * wd[it]) + alpha2 * exp(-beta2 * dbh_gr[it]))));
     }
-', file=(StanModel <- tempfile())
+    ', file=(StanModel <- tempfile()))
 
-BCImodel <- sflist2stanfit(
-	mclapply(
-		1:4,
-		function(i) {
-			stan(file=StanModel, data=BCIdata, seed=123, chains=1, chain_id=i, refresh=-1)
-			}, 
-		mc.cores=2
-	)
-)
+system.time(BCImodel <- sflist2stanfit(mclapply(1:2,mc.cores=2,
+		function(i) stan(file=StanModel, inter=4000, data=data,seed=123, chains=1, chain_id=i, refresh=-1))))
 ```
 
-2) Examine whether the effects of wood density varies with plant size.
-3) Write a literature review summarising how plant traits influence mortality across multiple hazards (e.g. wind throw, drought, fire, disease, herbivory)
-4) Build a process-based model based on what we have learnt from the above objectives.
-
+##PROJECT 2
 ### Hazard Hypotheses
@@
 
-## Survival analysis
-
-- Most people model survival curves
-- However, here we are interested in determining how trait alter hazards.
-
-The survival function is intimately related to the **hazard** or **instantaneous mortality function**,  denoted $\lambda(t)$, which gives the event rate at time $t$ conditional on survival until time $t$:
-
-\begin{equation} \label{eq:lambda} \lambda(t) = \lim_{dt \rightarrow 0} \frac{\Pr(t \leq T < t+dt)}{dt\cdot S(t)} = -\frac{S'(t)}{S(t)}.\end{equation}
-
-This means that:
-\begin{equation}
-Survival(t) = \lambda_1(t) \times \lambda_2(t) \times \lambda_3(t)... \times \lambda_n
-\end{equation}
-
-### Here I plan to model hazards by...
-####EG Fire
-
-\begin{equation}
-\lambda(t) = f(age)^{-1} \times \int_{0}^{\infty} p(intensity|age) \times \Pr(\textrm{Death in fire of intensity}\, i) \, \textrm{d}i
-\end{equation}
-
-\begin{equation}
-\Pr(Death)  = \alpha + \beta_1 \times Size + \beta_2 \times Resprouter + \beta_3 \times Bark thickness
-\end{equation}
-
-##OTHER STUFF
-Below are some excerpts from William Morris's thesis on mortality of trees.
-
-The **finite mortality rate**, $M$, ignoring recruitment, is the complement of the ratio of final populatio size, $N_T$, to initial population size, $N_0$, divided by the length of the observation period, $T$,
-
-\begin{equation} \label{eq:FiniteMort}
-M = 1 -\frac{N_T/N_0}{T}
-\end{equation}
-
-However, in order to infer an unbiased expected proportion of dying individuals for periods shorter or longer than that used to calculate finite mortality one must calculate the **instantaneous mortality**, $M_T$:
-
-\begin{equation} \label{eq:InstMort}
-M_T = 1-\textrm{exp}(-\lambda T)
-\end{equation}
-
-where
-
-\begin{equation} \label{eq:lambda}
-\lambda = - \frac{\textrm{log}_e(1-M)}{T}.
-\end{equation}
-
-This function is widely used to describe the loss of individuals from a population over time. Here, when $M_T$ and $\lambda$ are more or less equivilent mortality rates are low.
-
-Calculating mortality rate is made more complicated when multiple censuses (a period of time in which mortality risk is being assessed, where census interval = $T$) are considered, partlicalrly when census interval lengths vary.
-In such cases equations \ref{eq:FiniteMort} and \ref{eq:InstMort} cannot be used because $T$ is not dixed and substituting $T$ with average interval length has been shown to underestimate mortality (Kubo et al., 2000).
-This is a particular problem with large-scale monitoring programs such as BCI where the period of time between revisits of individual trees varies within a signle census because there are too many individuals for observations to be made simultaneously.
-Datasets such as BCI can take more than a year to complete an entire census.
-
-When $T$ varies within a mortality dataset, estimates of population mortality rate will be biased if this variation is not accounted for (e.g.
-by assuming a mean value for $T$ and using equation \ref{eq:InstMort}.
-In order to achieve an unbiased estimate of $M_T$ and $\lambda$ when census length varies, maximum likelihood (Kubo et al, 2000) or Bayesian (He, 2003) methods can be used, which allow $T$ to be considered a continous variable.
-
-The methods so far assume $\lambda$ is fixed in space and time, between individuals and over the lifespan of an individual.
-Often these assumptions are not met because individual hazard rates vary.
-Heterogeneous hazard biases estimates of population mortality rates (Sheil & May 1996) and this bias is greater for longer census intervals (Zens & Peart 2003).
-The need for unbiased estimates of mortality, despite the hazard heterogeneity present in data from forests and woodlands, has lead to the development and application of methods that move beyond considering mortality at the population level and directly modelling hazard of individual trees (Zens & Peart, 2003).
-
-The logistic regression family of methods are the most common individual-based approach used to analyse tree mortality data (see Breece, Kolb & Dickson, 2008, Carus 2010, Flewelling & Monserud, 2002).
-Logistic regression models consider tree death, $\textrm{Pr}(y_i = 1)$, as a series of binomially distributed data with the probability of observing a death as function of a set of covariates, $X_i$, and associated coefficients, $\beta$ related to data via the logit link function.
-
-\begin{equation} \label{eq:logit}
-\textrm{Pr}(y_{i} = 1) = \textrm{logit}^{-1}(\beta X_i)
-\end{equation}
-
-The logit function, $\textrm{log}_e(p/(1 - p))$, transforms a quantity bound within the unit interval, such as probability, $p$, to cary between negative infinity and infinity.
-By relating mortality to covariates that can explain differences in death rate, individual based methods such as logistic regression can reduce the bias found in population rate estimates when individuals have variable hazard (Zens & Peart, 2003).
-
-The complementary log-log (cloglog) link, $\textrm{log}_e(-\textrm{log}_e(1-p))$, is another link that can be used to estimate mortality rates.
-The advantage of the cloglog link is that it is equivalent to standard survival analysis... NEED MORE INFO. It also reduces bias due to long and variable census intervals (BY?).
-By adding the census interval $T$ to equation \ref{eq:logit} and substituting the logit link with a cloglog link yields
-
-\begin{equation} \label{eq:cloglog}
-\textrm{Pr}(y_{i} = 1) = \textrm{cloglog}^{-1}(\beta X_{i} + \textrm{log}(T))
-\end{equation}
-
 #Random notes

### 2015-01-22 R/Competing_risk_models.Rmd
Competing risks JAGS model

---
title: "Mortality models"
author: "J Camac"
output: html_document
---
```{r}
library('dplyr')
library('R2jags')
#Loads BCI data and attaches trait data
load('../data_BCI/output/bci.mainstem.Rdata')
bci.traits <- read.csv('../data_BCI/rawdata/traits/BCI_traits_20101220.csv')
names(bci.traits) <- tolower(names(bci.traits)) # lowers trait column names for merging
bci.traits$sp <- tolower(bci.traits$sp) # lowers species code names for merging
bci.mainstem <- merge(bci.mainstem,bci.traits[,c('sp','sg100c_avg')],by = 'sp') #only uses species trait data exists for.
rm(bci.traits) # trait db no longer needed

#Subsets data to only include species with traits and has a status record.
#Also removes couple of rows that have 10 yr census intervals.
#Then adds a sp.id column
bci.mainstem<- bci.mainstem %>%
  filter(!is.na(census.interval) & !is.na(sg100c_avg) & !is.na(dead.next.census) & census.interval< 6) %>%
  mutate(sp.id = as.numeric(factor(sp)))
bci.mainstem <- bci.mainstem[bci.mainstem$year==2005,]

  #Preps data for model by centering and scaling predictors and time
  data <- list(
  n_obs = nrow(bci.mainstem),
  n_spp = length(unique(bci.mainstem$sp)),
  spp = as.numeric(factor(bci.mainstem$sp), as.character(unique(bci.mainstem$sp))),
  wd =  scale(log(unique(bci.mainstem$sg100c_avg*1000)), 
              center = log(600), scale=2 *sd(log(unique(bci.mainstem$sg100c_avg*1000))))[,1],
  dbh = scale(bci.mainstem$dbh, center = FALSE, scale = 2 * sd(bci.mainstem$dbh))[,1],
  dbh_gr = scale(bci.mainstem$dbh.gr, center = FALSE, scale = 2 * sd(bci.mainstem$dbh.gr))[,1],
  dbh_rgr = scale(bci.mainstem$dbh.rgr,center = FALSE, scale = 2 * sd(bci.mainstem$dbh.rgr))[,1],
  basal_area = scale(bci.mainstem$basal.area, center = FALSE, scale = 2 * sd(bci.mainstem$basal.area))[,1],
  basal_gr = scale(bci.mainstem$basal.area.gr, center = FALSE, scale = 2 * sd(bci.mainstem$basal.area.gr))[,1],
  basal_rgr = scale(bci.mainstem$basal.area.rgr, center = FALSE, scale = 2 * sd(bci.mainstem$basal.area.rgr))[,1],
  census_length = bci.mainstem$census.interval,
  died = bci.mainstem$dead.next.census)

models <- c('basal_gr', 'basal_rgr')
nmodels <-length(models)
output <- list()
inits <- function() {
              list(    
              alpha0 = rlnorm(1,-3, 0.01),
              alpha1 = rlnorm(1,-3, 0.01),
              alpha2 = rlnorm(1,-3, 0.01),
              beta1 = rlnorm(1,-0.5, 0.01),
              beta2 = rlnorm(1,0, 0.01))
              }

for (i in 1:nmodels) {
  cat(sprintf('
              model {
              for(it in 1:n_obs) {
              died[it] ~ dbern(p[it])
              cloglog(p[it]) <- log(census_length[it]) + 
              log(alpha0 + alpha1 * exp(-beta1 * wd[spp[it]]) + alpha2 * exp(-beta2 * %s[it]))
              }

              #Priors
              alpha0 ~ dlnorm(0, 0.01) 
              alpha1 ~ dlnorm(0, 0.01) 
              alpha2 ~ dlnorm(0, 0.01)
              beta1 ~ dlnorm(0, 0.0001)
              beta2 ~ dlnorm(0, 0.0001)

              }',models[i]),file=(tmp <- tempfile()))

  system.time(output[[models[i]]] <- 
                      jags.parallel(
                        model.file = tmp,
                        n.iter = 30000,
                        data= names(data), 
                        envir=list2env(data),
                        inits = inits,
                        parameters.to.save = c('alpha0','alpha1','alpha2','beta1','beta2')))
  }
  beep('mario')
```

### 2015-01-22 proposal.Rmd
Draft proposal. No preliminary results and figures. References still to be added

--- 2014-11-19 proposal.Rmd
@@
 ---
-title: "Mortality Notes"
+title: "Hazard rates, growth, size and plant traits"
 author: "J Camac"
-date: "9 July 2014"
+date: "10 Dec 2014"
 output: pdf_document
@@
 ---
+##Background
 
-###Background
+Understanding who lives or dies, and why, is an integral component of natural selection, and is crucial for building better individual-based models of population and community dynamics.
+However, ecologists have barely begun to tap the information available in survival data.
+For plants, this untaped information has resulted in few mechanistic models of plant mortality (Hawkes_2000).
+The reason for this may be due to past and current work primarily focusing on modelling plant growth as opposed to mortality (Hawkes 2000).
+But it also may be a consequence of difficulties associated with modelling mortality.
 
-Plant mortality and the processes affecting it are integral to understanding vegetation dynamics.
-However, few mechanistic models of vegetation dynamic explicitly simulate mortality [@Hawkes_2000], and those that have, do so using little process-based information.
-For example, mechanistic models have accounted for plant mortality by assuming mortality rates are constant through time and space (Thorton 2007m White 2000, Delbart et al 2010) or that they change when a threshold has been surpassed (e.g.
-when Sitch et al 2003, Sato et al 2007).
-These simplifications are a result of a lack of process-based information on what causes mortality rates to vary temporally, spatially and between species (Hawkes 2000, McDowell et al 2011).
-In part, this is because researchers have placed a greater emphasis on modelling plant growth (Hawkes 2000).
-But it is also due to the difficulties associated with modelling mortality rates.
+Plant mortality is highly variable and episodic in both space and time (Burgmann 1994).
+It varies between species and between individuals within a species.
+This variability is thought to be affected by age (REF), size (REF) and genetics (REF) but also by multiple biotic and abiotic factors which may interact with one another (e.g.
+drought and herbivory: McDowell 2011).
+Because of this complexity, calibrating empirical models or validating mechanistic models of plant mortality requires datasets that monitor large numbers of individuals across entire lifespans and across multiple biotic and abiotic gradients.
+Furthermore, it also requires that causes of death to be identified for each individual.
+However, few, if any, field datasets meet all these requirements.
+Instead, most plant survival data come from monitoring programs conducted at few sites, that monitor survival as a binary event, with no delimintation of cause of death, often for periods shorter than that required to capture an individual or species entire lifespan (Bugmann 1996).
 
-Temporal patterns in plant mortality are highly variable and episodic (Burgmann 1994).
-Mortality rates also vary between species and are strongly influenced by multiple biotic and abiotic factors that may interact with one another (e.g.
-drought and herbivory: McDowell 2011).
-Because of this complexity, calibrating empirical models or validating mechanistic models of plant mortality requires datasets that monitor large numbers of species and individuals across their lifespan and across multiple biotic and abiotic gradients.
-More importantly, it requires the cause of death to be known for each individual.
-However, few, if any, data meet all these requirements.
-Rather, most available data comes from monitoring programs that are usually shorter than many plants lifespans (Bugmann 1996), and that focus on single species at single sites.
-Furthermore, because of the inherent difficulty of defining the cause of death, most monitoring programs do not delimit the cause of death for each individual.
+Currently, most models examining plant mortality have been correlative.
+In a review of 61 woody plant mortality models, Hawke (2000) found that only 4 were mechanistic and 57 were correlative.
+These models typically include correlations with one or multiple, abiotic and biotic factors such as competition (REF) and climate (REF) as well as state variables such as age (REF), size (Hurst et al 2011, Vieilledent et al.
+2009, Yang et al.
+2003) and growth rate (Vielledent et al.
+2010, Wright et al 2010).
+More recently, plant traits have also been incorporated into such correlative models to examine inter-specific variability in mortality rates (e.g.
+Kitajima et al 2013, Wright et al.
+2010) and improve model generality so that predictions can be made to new species outside the fitted data (REF).
+Trait-based mortality analysis have found that several traits are correlated with plant mortality rates.
+For example, wood density has been shown within BCI to be negatively correlated with mortality rates.
+Bark thickness (Pausas et al 2014) and being decidious (???) relative to evergreen has also been shown to reduce mortality after events such as fire and drought, respectively.
 
-Consequently, there are very few mechanistic mortality models.
-In a review of 61 woody plant mortality models, Hawke (2000) found that only 4 were mechanistic.
-In these models, mortality occurred when carbon supply was insufficient for growth (e.g.
-Friend et al.
-1997, Bossel 1986, Weinstein et al 1991, Bugmann et al 1997).
-By contrast, the majority of mortality models (57) were empirical and included forest yield models (Guan and Gertner 1991), gap models (Dale et al 1985), plant-environment models (King and Grant 1996) and spatial models (Jeltsch et al 1997).
-These empirical mortality models spanned the range from deterministic (REFS) to stochastic (REFS) and typically included correlations of one or multiple, abiotic and biotic factors such as competition (REF) and climate (REF) as well as state variables such as age (REF), size (REF) and growth rate (REF).
+While measures of fitness such as reproductive success have typically been quantified at the individual level, mortality is commonly quantified at the population level (Zan et al 2010, Sheil et al 1995).
+The finite population mortality rate, $M_{finite}$, a time discrete measure of mortality, is one common way that ecologists have examined mortality.
+Here, $M_{finite}$ is estimated as the complement of the ratio of final population size, $N_T$, to initial population size, $N_0$, divided by the length of the observation period, $T$,
 
-More recently, researchers have used plant traits in an attempt to explain inter-specific variability in mortality rates. For example, ....
-However, to date, most of these studies have assumed that these trait effects are constant through time and do not interact with changing environmental conditions or individual state variables such as growth rate or size.
+\begin{equation}\label{eqn:finitepopmortrate}
+M_{finite} = 1 - \frac{N_T/N_0}{T}.
+\end{equation}
 
-###Aims
-Based on these knowledge gaps I have three primary objectives:
-1) Examine whether the effects of wood density (and other traits) on mortality are growth & size dependent or independent.
-2) Conduct a literature review summarising how plant traits influence mortality rates under common hazards (e.g. wind throw, drought, fire, disease, herbivory).
-3) Build a process-based mortality model that incorporating the common hazard experienced by plants and how traits mediate their effects.
+However, in order to estimate the expected proportion of individuals surviving for periods shorter or longer than that used to calculate the $M_{finite}$, time must be treated as continuous, and the instantaneous population mortality rate (also known as the population hazard rate), $\lambda_{pop}$ needs to be calculated:
 
-##Growth dependent & independent effects of wood density
+\begin{equation}\label{eqn:poplambda}
+\lambda_{pop} = -\frac{\log_e(1 - S))}{T}.
+\end{equation}
+and
 
-###Example JAGS model with no random effects
+\begin{equation}\label{eqn:S}
+S = 1 - \exp{(-\lambda T)},
+\end{equation}
+
+where $S$ is the proportion of individuals that survive.
+
+In plant ecology, these functions have become common place in describing the loss of individuals from a population (e.g.
+Zens et al 2003, Condit, Hubbell & Fores).
+They have also been correlated with other covariates.
+For example, Wright et al (2010) quantified the finite population mortality rate for 103 tree species and then correlated these values with species level growth and wood density estimates.
+However, a fundamental problem with these population estimates of mortality is that they ignore individual variability and as such implicitly assume that all individuals are the same and experience the same likelihood of dying (i.e.
+the same hazard rate).
+Such an assumption violates the basic pretense of natural selection (REF).
+Individuals within a population are not identical, they vary genetically, typically span a range of sizes and ages, and are exposed to different levels of resources and abiotic and biotic pressures.
+The consequence of ignoring this important source of variability is that population estimates of mortality rates are likely to be biased and may severely limit the predictive capacity of population and community dynamic models (Zens et al 2003, Shiel et al 1995).
+Regardless of the method used, bias is an inevitable product of any heterogenous sample.
+It can, however, be reduced if the levels of heterogeneity are recognized and explicitly incorporated into analyses.
+
+Individual-based models of mortality or survival are one way to account for variable hazards and minimise associated biases at the population level.
+These models have been extensively applied in other disciplines (REFS), but are only now becoming regularly used in ecology and evolution (e.g.
+REFs, Zens et al 2003).
+At the core of all survival models is a hazard function which defines the instantaneous potential per unit time, $\lambda(t)$ for the event (in this case death) to occur, given that the individual has survived up to time $t$:
+
+\begin{equation} \label{eq:lambda} \lambda(t) = \lim_{dt \rightarrow 0} \frac{\Pr(t \leq T < t+dt)}{dt\cdot S(t)} = -\frac{S'(t)}{S(t)}.\end{equation}
+
+The hazard function is intimately related to the survival function, which gives the probability $S$ that an event does not occur before a given time $T$:
+
+$$ S(t) = \Pr(T > t)$$
+where "Pr" is the probability and $T$ is a random variable denoting the time of event. For a particular individual, $S(t)$ gives the expected probability that it will survive from $0 \rightarrow t$.
+
+In ecology, individual-based survival models that have been used include mark-recapture (Pollock et al 1995), survival analysis (Cox & Oakes 1984) and logistic regression (Allison 1982).
+Mark-recapture methods are typically applied to animal populations when it is uncertain whether marked individuals will be reencountered and mortality rates are estimated in light of this uncertainty.
+By contrast, survival analysis is used when accurate estimates of time of event (i.e.
+death) can be obtained, which commonly is not the case for most plant census data.
+However, survival analysis has been applied to tree-death data, with incremement growth being used as a surrogate for the passage of time (Woodall, Grambsch & Thomas 2005).
+
+The logistic regression family of methods are the most common individual-based models used to analyse plant mortality data (e.g.
+Breece, Kolb & Dickson 2008, Carus 2010, Flewelling & Monserud 2002).
+This is because most plant-death datasets are interval-censored; that is they only record the binary state of an individual between two time points $t_1$ and $t_2$.
+These models consider tree death $\Pr(y_i =1)$, as a series of binomially distributed data with the probablity of observing death as function of a covariates $X_i$, associated coefficients, $\beta$, and a baseline hazard $\lambda_0$ related via a link function, typically a logit $\log_e(p/1-p)$.
+Here, the coefficients explain differences in individual hazards and as such reduce the bias found in population estimates (Zens & Peart 2003).
+
+\begin{equation} \label{eq:logit}
+\log(\frac{\Pr(y_i = 1)}{1 - Pr(y_i = 1)} =  \sum_{i=1}^n \beta X_i\ + \lambda_0. 
+\end{equation}
+
+Although the logit (see\ref{eq:logit}) is often used, the complimentary-log-log, $\log_e(-\log_e(1-p))$ link is more advantageous in survival analysis.
+This is because it makes the regression model equivalent to proportional hazards model (see below) in that coefficients are interpreted as hazard ratios.
+That is, the hazard for one individual relative to another(Kleinbaum et al 2012).
+It also reduces bias due to long and variable census intervals (Allison, 1982), which is particularly important with large-scale monitoring programs such as BCI, where the period of time between revisits of individual trees varies within a single census because there are so many individuals that they cannot be observed simulataneously (Conduit 1995).
+
+\begin{equation} \label{eq:cloglog}
+\log_e\left(\log_e\left(1-\Pr(y_i = 1)\right)\right) =  \sum_{i=1}^n \beta_i X_i\ + \lambda_0 t. 
+\end{equation}.
+
+The proportional hazards model assumes a hazard function with form:
+
+\begin{equation} \label{eq:prop_hazards} \lambda(t) =  \lambda_0(t) \exp\left(\sum_{i=1}^n \beta_i x_i\right) =   \lambda_0(t)\prod_{i=1}^n \exp\left( \beta_i x_i\right),\end{equation}
+
+where $X_1, \ldots, X_n$ are various risk factors and $\beta_1, \ldots, \beta_n$ are associated coefficients. This leads (via eq. \ref{eq:HazS2}) to the following model for survival:
+
+\begin{equation} \label{eq:prop_hazards_S}
+S(t) =  \exp\left(\exp\left(-\sum_{i=1}^n \beta_i x_i\right)\right) S_0(t), \end{equation}
+
+where $S_0(t) = \exp\left(-\int_0^t \lambda_0(t)\right)$ is the baseline survival function (i.e.
+
+A key assumption of the proportional hazards model is that the relative risk from different predictors remains constant over time.
+For example, if being a particular size halved your hazard at time 0, it also halves your hazard at time 1 or any value of t (Kleinbaum et al 2012).
+However, these models can be extended to allow for individual covariates to vary over time (Kleinbaum et al 2012) and also include additive smoothing terms or hierarchical model structures.
+
+In general, these models are applied to data where a single event is being considered.
+However, when there are at least two possible pathways leading to any given outcome (e.g., death from any of several causes), the problem is characterized as a competing risk problem (Kleinbaum et al 2012).
+For example, if there are two alternative and independent mechanisms leading to plant death such as death by insufficient carbon and wind throw.
+The hazard function then has form:
+
+\begin{equation} \label{eq:additive_hazards} 
+\lambda(t) =  \lambda_{carbon}(t) + \lambda_{windthrow}(t) ,
+\end{equation}
+
+leading to the survival function
+
+\begin{equation} \label{eq:additive_hazards_S} S(t) =  \exp\left( - \int_0^t \lambda_{carbon}(t) + \lambda_{windthrow}(t) \, \textrm{d} t\right) =  \exp\left( - \int_0^t \lambda_{carbon}(t) \textrm{d} t\right)\exp\left( - \int_0^t \lambda_{windthrow}(t) \textrm{d} t\right).
+\end{equation}
+
+In the same way as we did for single event, each event may  be related to a number of covariates through a proportional hazards model, such that
+
+\begin{equation} 
+\lambda_{carbon}(t) =  \lambda_{carbon,0}(t) \exp\left(\sum_{i=1}^n \beta_{i,1} x_i\right),
+\end{equation}
+and
+\begin{equation} 
+\lambda_{windthrow}(t) =  \lambda_{windthrow,0}(t) \exp\left(\sum_{i=1}^n \beta_{i,2} x_i\right).
+\end{equation}
+
+We then have a model of the form
+
+\begin{equation} \label{eq:additive_hazards_S2} 
+S(t) =  S_{0}(t) \times \exp\left(\exp\left(-\sum_{i=1}^n \beta_{i,2} x_i\right)\right) \times \exp\left(\exp\left(-\sum_{i=1}^n \beta_{i,1} x_i\right)\right),
+\end{equation}
+
+where $S_{0}(t)=\exp\left(-\int_0^t \lambda_{carbon,0}(t)\right) \exp\left(-\int_0^t \lambda_{windthrow,0}(t)\right)$ is combined baseline hazard function.
+
+If the different time-of-event is known for each event type, a competing risk analysis can be used to estimate the underlying covariates (essentially this involves partitioning the dataset into two, one for each type of event, and adjusting the population sizes to account for deaths due to the other event)(kleinbaum 2005).
+
+The challenge here is to see if we can estimate the parameters of eq.  \ref{eq:additive_hazards_S2} using interval-census data where we do not know what tree died of.
+
+### PROJECT 1: A competing risks problem
+Plants can be killed by 1) not assimulating enough carbon for cell maintenance (i.e.
+carbon starvation), or 2) by some event that fatally damages the plant (e.g.
+fire, herbivory, windthrow).
+Ecologists have attempted to model the effects of insufficient carbon on hazard rates by using various measures of plant growth as a proxy for the carbon budget (e.g.
+Wright et al 2010).
+They have also used plant size and plant traits such as wood density as predictors of mortality.
+
+However, as far as I am aware, no study has attempted to partition these effects into a competiting risk problem, whereby death by carbon starvation is treated independently to the effect of wood density which may influence the susceptiblitly to other causes of death.
+Here, I attempt to do build a multiplicative and additive hazards model using interval-censored data in a logistic regression framework.
+
+This project is made up of a series of questions examining how wood density, growth rates and size influence plant mortality rates.
+
+-1 Can we use a competing risks framework to attempt to estimate cause of death? (i.e. by treating the effects of growth, size and wood density as independent causes of mortality)
+-2 How do the findings of a competing risks approach differ from standard logistic regression? Are there any additional insights that can be gleamed? 
+-3 How does wood density, growth and size influence mortality rates and how do the magnitude of effects compare?
+-4 Can interactive terms be included in a competing risk framework? Is the presence of a interaction nullify the independence assumption of competing risks?
+-5 What measure of growth or size (i.e. dbh, basal or mass - absolute or relative) best predicts mortality rates?
+
+#### Competing hazard models
+Here I attempt to fit a competing hazards model using interval-censored data where we don the following Bayesian logistic regression form:
+
+\begin{equation}\label{eqn:wd_grmod_inter}
+\overbrace{\log(-\log(1 -\Pr(y_{i,t} = 1)))}^{\text{Survival Probability}} =
+\overbrace{\log(\Delta{t})}^{\text{census interval}} + \log\left(\lambda_0 + \lambda_1 + \lambda_2 \ldots \lambda_n\right). 
+\end{equation}
+
+**Hazards currently fitted to data**
+Below I assume negative exponential relationships between mortality and growth rate, wood density and size.
+In these models the $\alpha$'s refer to the intercept and the $\beta$'s refer to the rate of change.
+In a competing risks framework each $\lambda$ is assumed to be independent of eachother.
+I've ran these models with different variants of growth (e.g.
+incremental dbh and basal growth and relative growth rates of each) in an attempt to determine which is the best predictor of mortality.
+
+\begin{equation}\label{eqn:indhaz}\begin{array}{ll}
+\lambda_0 = &
+\overbrace{\alpha_0}^{\text{baseline/unexplained}} 
+\\
+\\
+
+\lambda_1 = &
+\overbrace{\alpha_1 \times \exp{(-\beta_1 \times gr)}}^{\text{growth dependent hazard}} 
+\\
+\\
+
+\lambda_2 = &
+\overbrace{\alpha_2 \times \exp{(-\beta_2 \times \log(wd))}}^{\text{wd dependent hazard}} 
+\\
+\\
+\end{array} \end{equation}
+
+**Example JAGS competing risk model with no random effects**
 
 ```{r, eval=FALSE}
-library('dplyr')
-library('R2jags')
-#Loads BCI data and attaches trait data
-load('../data_BCI/output/bci.mainstem.Rdata')
-bci.traits <- read.csv('../data_BCI/rawdata/traits/BCI_traits_20101220.csv')
-names(bci.traits) <- tolower(names(bci.traits)) # lowers trait column names for merging
-bci.traits$sp <- tolower(bci.traits$sp) # lowers species code names for merging
-bci.mainstem <- merge(bci.mainstem,bci.traits[,c('sp','sg100c_avg')],by = 'sp') #only uses species trait data exists for.
-rm(bci.traits) # trait db no longer needed
+models <- c('dbh_gr','dbh_rgr','basal_gr', 'basal_rgr')
+nmodels <-length(models)
+output <- list()
+inits <- function() {
+              list(    
+              alpha0 = rlnorm(1,-3, 0.01),
+              alpha1 = rlnorm(1,-3, 0.01),
+              alpha2 = rlnorm(1,-3, 0.01),
+              beta1 = rlnorm(1,-0.5, 0.01),
+              beta2 = rlnorm(1,0, 0.01))
+              }
 
-#Subsets data to only include species with traits and has a status record.
-#Also removes couple of rows that have 10 yr census intervals.
-#Then adds a sp.id column
-bci.mainstem<- bci.mainstem %>%
-  filter(!is.na(census.interval) & !is.na(sg100c_avg) & !is.na(dead.next.census) & census.interval< 6) %>%
-  mutate(sp.id = as.numeric(factor(sp)))
-
-  #Preps data for model by centering and scaling predictors and time
-  data <- list(
-  n_obs = nrow(bci.mainstem),
-  n_spp = length(unique(bci.mainstem$sp)),
-  spp = as.numeric(factor(bci.mainstem$sp), as.character(unique(bci.mainstem$sp))),
-  wd = scale(unique(bci.mainstem[c('sp', 'sg100c_avg')])[,'sg100c_avg'])[,1]/2,
-  dbh = scale(bci.mainstem$dbh)[,1]/2,
-  dbh_gr = scale(bci.mainstem$dbh.gr)[,1]/2,
-  dbh_rgr = scale(bci.mainstem$dbh.rgr)[,1]/2,
-  basal_area = scale(bci.mainstem$basal.area)[,1]/2,
-  basal_gr = scale(bci.mainstem$basal.area.gr)[,1]/2,
-  basal_rgr = scale(bci.mainstem$basal.area.rgr)[,1]/2,
-  ln_census = scale(log(bci.mainstem$census.interval))[,1]/2, #logged for model
-  died = bci.mainstem$dead.next.census)
-
-#Actual model
-model <- function(x) {
+for (i in 1:nmodels) {
+  cat(sprintf('
+              model {
               for(it in 1:n_obs) {
-
               died[it] ~ dbern(p[it])
-    cloglog(p[it]) <- max(-9999, min(9999,  ln_census[it] + log(cd0*exp(-cd1 * wd[spp[it]]) + cd2 * exp(-cd3 * wd[spp[it]] * dbh_gr[it]))))
+              cloglog(p[it]) <- log(census_length[it]) + 
+              log(alpha0 + alpha1 * exp(-beta1 * log(wd[spp[it]])) + alpha2 * exp(-beta2 * %s[it]))
               }
@@
               #Priors
-  cd0 ~ dlnorm(0, 0.0001) # Must be positive or large growth rates are undefined.
-  cd1 ~ dnorm(0, 0.0001) # Can be positive or negative (but theory would suggest positive)
-  cd2 ~ dlnorm(0, 0.0001) # Has to be positive or small growth rates not defined
-  cd3 ~ dnorm(0, 0.0001) # Can be positive or negative
+              alpha0 ~ dlnorm(0, 0.01) 
+              alpha1 ~ dlnorm(0, 0.01) 
+              alpha2 ~ dlnorm(0, 0.01)
+              beta1 ~ dlnorm(0, 0.0001)
+              beta2 ~ dlnorm(0, 0.0001)
 
-  }
-mod<-     jags.parallel(model.file = model,
-                        n.iter = 4000,
+              }',models[i]),file=(tmp <- tempfile()))
+
+  system.time(output[[models[i]]] <- 
+                      jags.parallel(
+                        model.file = tmp,
+                        n.iter = 30000,
                         data= names(data), 
                         envir=list2env(data),
-                        inits = NULL,
-                        parameters.to.save = c('cd0','cd1','cd2','cd3'))
+                        inits = inits,
+                        parameters.to.save = c('alpha0','alpha1','alpha2','beta1','beta2')))
+  }
 ```
 
-#Potential Stan Model
-```{r, eval=FALSE}
-library(parallel)
-library(rstan)
+**Preliminary results of competing risk models**
 
-cat('
-  data {
-		int<lower=0> n_obs;
-		int<lower=0, upper=1> died[n_obs];
-		vector[n_obs] ln_T;
-		vector[n_obs] wd;
-		vector[n_obs] gr;
+TO BE ADDED
 
-	}
-	parameters {
-		real<lower=0> c0;
-		real c1;
-		real<lower=0> c2;
-		real c3;
+**Further extensions still to do**
+- Add a hazard effect of size? What would the shape of this be? Preliminary analysis of mortality vs dbh suggests that most species exhibit a U-shape mortality curve.
+- Validate that a negative exponential is a reasonable assumption for the effects of growth and wood density. This could be done via partial dependency plots and non-parametric approaches.
+- Cross validate models
 
-	} 
-	model {
-		for (it in 1:n_obs)
-			died[it] ~ bernoulli(inv_cloglog(ln_census[it] + log(c0 * exp(-c1 * wd[it]) + c2 * exp(-c3 * dbh_gr[it]))));
-	}
-', file=(StanModel <- tempfile())
+**Size mortality relationships**
 
-BCImodel <- sflist2stanfit(
-	mclapply(
-		1:4,
-		function(i) {
-			stan(file=StanModel, data=BCIdata, seed=123, chains=1, chain_id=i, refresh=-1)
-			}, 
-		mc.cores=2
-	)
-)
-```
+TO BE ADDED
 
-2) Examine whether the effects of wood density varies with plant size.
-3) Write a literature review summarising how plant traits influence mortality across multiple hazards (e.g. wind throw, drought, fire, disease, herbivory)
-4) Build a process-based model based on what we have learnt from the above objectives.
-
+### PROJECT 2: LITERATURE REVIEW. Examine which plant traits influence mortality rates under common hazards (e.g. wind throw, drought, fire, disease, herbivory) experienced by vegetation.
 ### Hazard Hypotheses
@@
 **Drought/Extreme heat/ Extreme cold**: These hazard is likely to affect smaller plants moreso than larger plants.
+
+### PROJECT 3: Process-based model - coming from first principles
+Build a process-based mortality model that allows trait effects to interact with the type of hazard as well as growth and size.
 
@@
 
-## Survival analysis
-
-- Most people model survival curves
-- However, here we are interested in determining how trait alter hazards.
-
-The survival function is intimately related to the **hazard** or **instantaneous mortality function**,  denoted $\lambda(t)$, which gives the event rate at time $t$ conditional on survival until time $t$:
-
-\begin{equation} \label{eq:lambda} \lambda(t) = \lim_{dt \rightarrow 0} \frac{\Pr(t \leq T < t+dt)}{dt\cdot S(t)} = -\frac{S'(t)}{S(t)}.\end{equation}
-
-This means that:
-\begin{equation}
-Survival(t) = \lambda_1(t) \times \lambda_2(t) \times \lambda_3(t)... \times \lambda_n
-\end{equation}
-
-### Here I plan to model hazards by...
-####EG Fire
-
-\begin{equation}
-\lambda(t) = f(age)^{-1} \times \int_{0}^{\infty} p(intensity|age) \times \Pr(\textrm{Death in fire of intensity}\, i) \, \textrm{d}i
-\end{equation}
-
-\begin{equation}
-\Pr(Death)  = \alpha + \beta_1 \times Size + \beta_2 \times Resprouter + \beta_3 \times Bark thickness
-\end{equation}
-
-##OTHER STUFF
-Below are some excerpts from William Morris's thesis on mortality of trees.
-
-The **finite mortality rate**, $M$, ignoring recruitment, is the complement of the ratio of final populatio size, $N_T$, to initial population size, $N_0$, divided by the length of the observation period, $T$,
-
-\begin{equation} \label{eq:FiniteMort}
-M = 1 -\frac{N_T/N_0}{T}
-\end{equation}
-
-However, in order to infer an unbiased expected proportion of dying individuals for periods shorter or longer than that used to calculate finite mortality one must calculate the **instantaneous mortality**, $M_T$:
-
-\begin{equation} \label{eq:InstMort}
-M_T = 1-\textrm{exp}(-\lambda T)
-\end{equation}
-
-where
-
-\begin{equation} \label{eq:lambda}
-\lambda = - \frac{\textrm{log}_e(1-M)}{T}.
-\end{equation}
-
-This function is widely used to describe the loss of individuals from a population over time. Here, when $M_T$ and $\lambda$ are more or less equivilent mortality rates are low.
-
-Calculating mortality rate is made more complicated when multiple censuses (a period of time in which mortality risk is being assessed, where census interval = $T$) are considered, partlicalrly when census interval lengths vary.
-In such cases equations \ref{eq:FiniteMort} and \ref{eq:InstMort} cannot be used because $T$ is not dixed and substituting $T$ with average interval length has been shown to underestimate mortality (Kubo et al., 2000).
-This is a particular problem with large-scale monitoring programs such as BCI where the period of time between revisits of individual trees varies within a signle census because there are too many individuals for observations to be made simultaneously.
-Datasets such as BCI can take more than a year to complete an entire census.
-
-When $T$ varies within a mortality dataset, estimates of population mortality rate will be biased if this variation is not accounted for (e.g.
-by assuming a mean value for $T$ and using equation \ref{eq:InstMort}.
-In order to achieve an unbiased estimate of $M_T$ and $\lambda$ when census length varies, maximum likelihood (Kubo et al, 2000) or Bayesian (He, 2003) methods can be used, which allow $T$ to be considered a continous variable.
-
-The methods so far assume $\lambda$ is fixed in space and time, between individuals and over the lifespan of an individual.
-Often these assumptions are not met because individual hazard rates vary.
-Heterogeneous hazard biases estimates of population mortality rates (Sheil & May 1996) and this bias is greater for longer census intervals (Zens & Peart 2003).
-The need for unbiased estimates of mortality, despite the hazard heterogeneity present in data from forests and woodlands, has lead to the development and application of methods that move beyond considering mortality at the population level and directly modelling hazard of individual trees (Zens & Peart, 2003).
-
-The logistic regression family of methods are the most common individual-based approach used to analyse tree mortality data (see Breece, Kolb & Dickson, 2008, Carus 2010, Flewelling & Monserud, 2002).
-Logistic regression models consider tree death, $\textrm{Pr}(y_i = 1)$, as a series of binomially distributed data with the probability of observing a death as function of a set of covariates, $X_i$, and associated coefficients, $\beta$ related to data via the logit link function.
-
-\begin{equation} \label{eq:logit}
-\textrm{Pr}(y_{i} = 1) = \textrm{logit}^{-1}(\beta X_i)
-\end{equation}
-
-The logit function, $\textrm{log}_e(p/(1 - p))$, transforms a quantity bound within the unit interval, such as probability, $p$, to cary between negative infinity and infinity.
-By relating mortality to covariates that can explain differences in death rate, individual based methods such as logistic regression can reduce the bias found in population rate estimates when individuals have variable hazard (Zens & Peart, 2003).
-
-The complementary log-log (cloglog) link, $\textrm{log}_e(-\textrm{log}_e(1-p))$, is another link that can be used to estimate mortality rates.
-The advantage of the cloglog link is that it is equivalent to standard survival analysis... NEED MORE INFO. It also reduces bias due to long and variable census intervals (BY?).
-By adding the census interval $T$ to equation \ref{eq:logit} and substituting the logit link with a cloglog link yields
-
-\begin{equation} \label{eq:cloglog}
-\textrm{Pr}(y_{i} = 1) = \textrm{cloglog}^{-1}(\beta X_{i} + \textrm{log}(T))
-\end{equation}
-
-#Random notes
-
-### MULTI-LEVEL NOTES:
-
-####From Hedeker et al 2014
-
-Nice summary of why multi-level models are essential for accounting for non-independence in nested datasets.
-
-"An important question is then to determine the degree to which covariates are related to substance use initiation.
-In these studies it is often of interest to model the student outcomes while controlling for the nesting of students in classrooms and/or schools.
-In analysis of such grouped-time initiation (or survival) data, use of grouped-time regression models that assume independence of observations [Thompson, 1977; Prentice & Gloeckler, 1978; Allison, 1982] is therefore problematic because of this clustering of students.
-More generally, this same issue arises for other types of clustered datasets in which subjects are observed nested within various types of clusters (e.g., hospitals, firms, clinics, counties), and thus cannot be assumed to be independent.
-To account for the data clustering, multilevel models (also called hierarchical linear or mixed models) provide a useful approach for simultaneously estimating the parameters of the regression model and the variance components that account for the data clustering [Goldstein, 1995; Raudenbush & Bryk, 2002]."
-
-## TWO REVIEW PAPERS TO READ Pickles et al 1995 & Hougaard 1995 ###
-
-Lee et al 1992 has developed continuous0time survival models.
-However, the application of these models to grouped or discrete-time survival data is generally not recommended because of the large number of ties that result.
-
-Instead, models specifically developed for grouped or discrete-time survival data have been proposed.
-
-Both Han & Hausman (1990) and Scheike & Jensen (1997) have described proportional hazards models incorporating a log-gamma distribution specification of heterogeneity.
-Furthermore, Ten Have (1996) developed a discrete-time proportional hazards survival model incorporating a log-gamma random effects distribution, additionally allowing for ordinal survival and failure categories.
-Ten Have & Uttal (1994) used Gibbs samplng to fit a continuation ratio logit model with multiple normally distributed random effects.
-
-## IMPORTANT NOTE ON link function
-
-Doksum & Gasko (1990) note, that large amounts of high quality data are often necessary for link function selections to be relevant.
-McCullagh (1980) notes that the link function choice should be based primarily on ease of interpretation.
-
-## Multilevel model that treats each individual's survival time as a set of dichotomous observations indicating whether or not an individual failed in each time unti until a person either experiences the event or is censored.
-This type of model is extensively described in Singer & Willett (2003) & is illustrated in Reardon et al (2002).
-
-This approach is particularly useful for handling time-dependent covariates and fitting non-proportional hazards models because the covariate values can change across each individual's Tij timepoints.
-
 # REFERENCES

### 2015-03-30 ms.Rmd
Updated mortality proposal with preliminary results

---
title: "Wood density and the growth-dependent & independent mortality"
author: "J S Camac"
date: "30 Mar 2015"
output: pdf_document
csl: downloads/style.csl
bibliography: data/refs.bib
---

```{r dependencies, echo=FALSE, message=FALSE,warning=FALSE}
source('R/utils.R')
load('output/rho_stan_model.rda')
library(rstan)
```

##Introduction
Understanding who lives or dies, and why, is critical to accurately predict population, community and ecosystem dynamics [@Hawkes:2000ib; @McDowell:2011dr]. In plants, the carbon budget is considered a fundamental determinant of mortality, such that if rates of carbon assimilation are low or become insufficient for growth, an individual is more likely to die [@Peet:1987va; @Hawkes:2000ib; @Keane:2001db; @McDowell:2011dr]. Empirical studies have broadly supported this assumption, with many studies showing that the probability of dying decreasing exponentially with growth rate [Fig 1; @Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw], and consequently most dynamic vegetation models, including tree gap models [@Botkin:1972ii; @Keane:2001db] and Dynamic Global Vegetation Models (DGVM's; @Woodward:2004ft) use this growth-mortality relationship as the predominant, and in many cases, only cause of plant mortality [@Hawkes:2000ib; @McDowell:2011dr].

![Fig1](figures/wyckoff2002.png)

**Fig 1:** The relationship between growth and mortality for seven co-occurring tree species in the southern Appalachian Mountains. From @Wyckoff:2002ul.

However, empirical evidence has highlighted that this growth-mortality relationship varies between species with those that grow intrinsically faster exhibiting higher mortality rates relative to those which are slower growing (Fig 2; @Poorter:2008iu; @Wright:2010fl). Plant functional traits – attributes of species that influence vital rates (i.e. survival, growth and reproduction) and ultimately fitness [@Ackerly:2003eb; @Poorter:2008iu] – have been used to examine this inter-specific variability.

![Fig2](figures/wright2010.jpg)

**Fig 2:** The growth–mortality trade-off for saplings expressed as the 95th percentile relative growth rate (RGR95) vs. the mortality rate of the slowest growing 25% of individuals (M25) for 103 tree species on Barro Colorado Island, Panama. From @Wright:2010fl.

In tropical rain forests, wood density, the biomass invested per unit wood volume, is one such functional trait that has been repeatedly identified as being negatively correlated with tree mortality [Fig. 3; @Nascimento:2005wx; @King:2006he; @AlvarezClare:2007gv; @Chao:2008hg; @Poorter:2008iu; @Kraft:2010kq; @Wright:2010fl]. This relationship is thought to be due to dense wood conveying enhanced physical resistance to growth-independent causes (Fig 4) of mortality such as stem breakage [@Putz:1983wu; @vanGelder:2006ex; @Chave:2009iy], wood herbivory (NEED REF), embolism [@Hacke:2001kj, @Jacobsen:2005fx, @Preston:2006jn] and fungal and pathogen attack [@Augspurger:1984wx]. However, wood density may also increase survival associated with growth-dependent mortality. For example, studies [e.g. @vanGelder:2006ex] have highlighted that species with high wood density are often more shade tolerant, and thus exhibit a lower probability of dying at low growth rates. Studies have also shown that wood density may reduce drought induced carbon starvation [@McDowell:2011dr] and the susceptibility of ‘stressed’ individuals to opportunistic invertebrate, pathogen or fungal attack [@Augspurger:1984wx; @McDowell:2011dr].

![Fig3](figures/kraft2010.png)

**Fig 3:** The relationship between wood density and mortality in a global tropical forest dataset. From @Kraft:2010kq

![Fig4](figures/chave2009.png)

**Fig 4:** Wood mechanical traits plotted against wood density. From @Chave:2009iy

\pagebreak

In this study we build a hierarchical Bayesian logistic regression that attempts to partitions 15-years of tropical tree mortality into growth-dependent and growth-independent hazards (i.e. death not related to growth). We then use this model to address three questions: 1) Does wood density affect the growth-mortality relationship? 2) Where does wood density have the greatest effect growth-dependent or growth-independent hazards? And 3) Do the observed relationships dependent on the choice of growth measure?

We hypothesize that wood density will decrease both growth-dependent and growth-independent sources of mortality. Specifically, we expect wood density to affect the growth-mortality relationship by decreasing its intercept, (i.e. mortality rates at zero growth) and steepening its slope (i.e. more negative, Fig 5). We also expect that the effect of wood density will be greatest for growth-independent mortality, and that these relationships will be broadly consistent across growth measures.

```{r hypotheses,fig.height=6,fig.width=6, echo=FALSE}
par(mfrow=c(2,1), mar=c(2,4,1,1), oma=c(3,0,0,0), xaxs='i')
curve(exp(1 -exp(-0.3) * x) + exp(-2), xlim=c(0,12), ylim=c(0,5),
xlab=NA, ylab='Hazard rate', xaxt='n')
curve(exp(0.5 -exp(1) * x) + exp(-4), add=TRUE, col='red')
axis(1, at=0)

legend('topright', c('low rho', 'high rho'), col=c('black','red'), lty=1, bty='n')
curve((1-exp(-(exp(1 - exp(-0.3) * x) + exp(-1.5))))*100, xlim=c(0,12), ylim=c(0,100),
xlab=NA, ylab='Pr(Mortality)/year', xaxt='n')
curve((1-exp(-(exp(0.5 - exp(1) * x) + exp(-4))))*100, add=TRUE, col='red')
abline(v=0, lty=2, col='blue')
axis(1, at=0)
mtext(side=1, 'Growth rate',line = 1, outer=TRUE)

```

**Fig 5:** Hypotheses of how wood density will affect growth-mortality curve. Top = Hazard curves (i.e. instantaneous mortality rates); Bottom = Mortality probability/year based on estimated hazard rates. Where curve asymptotes is dependent on wood density effects on growth-independent hazard. Note changes in intercept, slope and asymptote between both lines.

\pagebreak

##Data

We are fitting our model to growth and mortality data collected from a tropical rainforest on Barro Colorado Island, Panama @BarroColoradofores:2012up. This rainforest plot has been surveyed every 5 years since 1985 and includes size, growth and status (alive, dead) observations from more than 200,000 individuals across 286 species.

In this study we only use data recorded between 1990 and 2010 for single stemmed plants. Furthermore, we only used species that had an estimate for wood density. We removed multistemmed plants because of difficulties discerning mainstem.

For more details about the data, or to download it, please click [here](https://repository.si.edu/handle/10088/20925).
Currently species trait data is not publically available.

##Model Description

We used a hierarchical generalized linear model to examine how wood density affected growth-dependent and growth-independent tree mortality. These models are well suited to data sets that have a hierarchical structure (e.g. individuals within species) because they can account for observation error and partition both explained and unexplained variation at multiple levels of a dataset [@Gelman:2007te; @Kery:2010tp]. We used Bayesian inference and fitted models to data using Markov chain Monte Carlo (MCMC) sampling in R 3.1.3 using package rstan version 2.6 [@StanDevelopmentTeam:2015uz]. Before commencing model fit, we first scaled growth by dividing by two standard deviations. We then centered and scaled wood density, $rho$, by first log-transforming it and then substracting its mean and dividing by two standard deviations. The advantage of centering and standardizing is that it allows for better mixing of chains and easier interpretation, with intercepts interpreted as average responses and slope terms as partial dependencies conditional on other continuous variables at their mean. Furthermore, standardizing by two standard deviations allows the magnitude of effects to be directly compared between binary and continuous covariates (Gelman and Hill 2007). Three independent chains were executed and in all cases modeled parameters converged within 1000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic [@Gelman:1992ts; @Brooks:1998ju]. After discarding the first 1000 iterations as ‘burn in’, a further 1000 iterations were taken from the joint posterior. The residuals of the model were checked graphically against predictions with none showing any systematic pattern or severe heteroscedasticity. To ensure the effect of wood density was identifiable, and to examine it's effect on all three parameters (i.e growth-mortality intercept and slope, and growth-independent hazard) we selectively removed it from each and visually inspected whether the direction and magnitude of effects were severely altered. Below we describe the full model structure, parameters and priors. We also supply the stan code below.

We were interested in examining how wood density affected growth-dependent and growth-independent individual tree mortality.
However, our observations, $y_i$, were binary (0 = alive, 1 = dead) for each individual $i$. Consequently, we modeled these observations as a random realization from a Bernoulli distribution based on the model-fitted probability that individual $i$ would die, $p_i$:

\begin{equation} \label{eq:obs_model}
p_i \sim \textrm{bernoulli}(p_i) \end{equation}

We modelled the probability of drying, $p_i$, using a complimentary-log-log link, $\log_e(-\log_e(1-p))$. Unlike the traditional logit link, $\log_e(p/1-p)$, the complimentary-log-log allows regression coefficients to be interpreted as hazard ratios. That is, the hazard for one individual relative to another [@Kleinbaum:2012tn]. It is also asymmetrical and can explicitly account for variable census length [@Allison:1982vd; @Kleinbaum:2012tn], which is particularly important with large-scale monitoring programs such as BCI, where the period of time between revisits of individual trees varies within a single census because there are so many individuals that they cannot be observed simulataneously [@Condit:1995ux].

$p_i$ is modelled as a function of census length (ranging from 3.8 to 5.8 years) and the sum of two independent hazards:

\begin{equation} \label{eq:pred_model}
\log_e(-\log_e(1-p_i)) = \log_e\left(\textrm{census-length}_i \times 
\left(
\overbrace{\exp\left(alpha_{s[i]} - \exp(beta_{s[i]}) \times 
\frac{\textrm{growth}_i}{2 \times \sigma(\textrm{growth})}
\right)
}^{\text{Growth dependent hazard}} + 
\overbrace{\exp(ho_{s[i]})
}^{\text{Growth independent hazard}}\right)\right) 
\end{equation}

The growth-dependent hazard encapsulates the growth-mortality relationship by allowing the probability of dying to decrease exponentially with growth rate [@Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw]. This relationship consists of two species varying parameters alpha, the intercept term that defines the instantaneous hazard rate at zero growth, and beta, the slope term that defines the decrease in hazard rate per scaled unit growth (Fig.6). By contrast, the growth independent hazard is assumed to be linear and is defined by one species varying parameter, ho. Here, growth was estimated as $\frac{dbh_t - dbh_{t-1}}{\Delta{t}}$.

Instead of calculating the probability of death for a given time period, we can also use this model form to calculcate the instantaneous mortality rate, $\lambda_i$ (also known as the hazard rate; @Kleinbaum:2012tn), for each individual $i$:

\begin{equation} \label{eq:lambda}
\lambda_i = 
\left(
\overbrace{\exp\left(alpha_{s[i]} - \exp(beta_{s[i]}) \times 
\frac{\textrm{growth}_i}{2 \times \sigma(\textrm{growth})}
\right)
}^{\text{Growth dependent hazard}} + 
\overbrace{\exp(ho_{s[i]})
}^{\text{Growth independent hazard}}\right) 
\end{equation}

```{r model description, fig.height=6,fig.width=6, echo=FALSE}
par(mfrow=c(2,1), mar=c(3,4,1,1), xaxs='i')
       curve(exp(1 -exp(-0.3) * x) + exp(-4), xlim=c(0,12), ylim=c(0,5),
             xlab=NA, ylab='Hazard rate',xaxt='n', col='red')
       arrows(x0 = 2, y0 = 2.7, x1 = 0 ,y1 = 2.7)
       text(3.1,2.7, 'alpha')
       arrows(x0 = 4, y0 = 1.5, x1 = 1.5 ,y1 = 1)
       text(4.5,1.7, 'beta')
       arrows(x0 = 11, y0 = 0.8, x1 = 11 ,y1 = 0.018)
       text(11,1, 'ho')
       axis(1, at=0, labels= 0)

       curve((1-exp(-(exp(1 - exp(-0.3) * x) + exp(-4))))*100, xlim=c(0,12), ylim=c(0,100),
             xlab=NA, ylab='Pr(Mortality)', xaxt='n', col='red')
       text(3.1,93, 'alpha')
       arrows(x0 = 2, y0 = 93, x1 = 0 ,y1 = 93)
       text(6,30, 'beta')
       arrows(x0 = 5, y0 = 30, x1 = 3 ,y1 = 30)
       text(11,15, 'ho')
       arrows(x0 = 11, y0 = 10, x1 = 11 ,y1 = exp(-4)*100)
       axis(1, at=0, labels= 0)
       mtext(side=1, 'Growth rate',line = 2)
```

**Fig 6:** Interpreting model parameters. Top = Hazard curves (i.e. instantaneous mortality rates); Bottom = Mortality probability/year based on estimated hazard rates. The intercept at zero growth is defined by $alpha$. The slope is defined by $beta$. The asymptote is defined by $ho$ and represents the growth-independent hazard.

We modelled $alpha$, $beta$ and $ho$, by allowing their intercepts to vary by species and by including wood density as a species level predictor:

\begin{equation} \label{eq:alpha}
alpha_s = \alpha0_s + \alpha1 \times \frac{(\log_e(\rho_s) - \mu(\log_e(\rho)))}{2 \times(\sigma(\log(\rho)))}
\end{equation}

\begin{equation} \label{eq:beta}
beta_s = \beta0_s + \beta1 \times \frac{(\log_e(\rho_s) - \mu(\log_e(\rho)))}{2 \times(\sigma(\log(\rho)))}
\end{equation}

\begin{equation} \label{eq:ho}
ho_s = \gamma0_s + \gamma1 \times \frac{(\log_e(\rho_s) - \mu(\log_e(\rho)))}{2 \times(\sigma(\log(\rho)))}
\end{equation}

Here, the terms $\alpha1$, $\beta1$ and $\gamma1$ represent the effect of wood density on the intercept, slope and growth-independent hazard, respectively. The terms, $\alpha0_s$, $\beta0_s$ and $\gamma0_s$, respectively represent the growth-dependent intercept and slope and growth-independent intercept for species $s$ within which individual $i$ belongs. We estimates these varying terms from a normal distribution centred on the average species intercept (i.e. $\alpha_{\mu}$,$\gamma_{\mu}$) or slope ($\beta_{\mu}$) and the standard deviation of the species variation ($\alpha spp$,$\beta spp$, $\gamma spp$).

\begin{equation} \label{eq:alpha0}
\alpha_s \sim \textrm{Normal}(\alpha_{\mu}, \sigma_{\alpha spp})
\end{equation}

\begin{equation} \label{eq:beta0}
\beta_s \sim \textrm{Normal}(\beta_{\mu}, \sigma_{\beta spp})
\end{equation}

\begin{equation} \label{eq:gamma0}
\gamma_s \sim \textrm{Normal}(\gamma_{\mu}, \sigma_{\gamma spp})
\end{equation}

Because we used Bayesian inference we needed to specify prior distributions for all model parameters. For wood density effects ($\alpha1$, $\beta1$ and $\gamma1$) and mean hyperparamters ($\alpha_{\mu}$, $\beta_{mu}$ and $\gamma_{\mu}$) we used stan's default uninformative prior, $\textrm{Uniform(-2,2)}. For hyperparamater standard deviations we used $\textrm{Uniform(0,2)}$.

\pagebreak

##Results
**Take home message:**
Wood density decreases intercept (log_a) & further increases the steepness of the growth-mortality slope (log_b). It also decreases the growth-independent hazard log_ho

```{r Model output, echo=FALSE}
print(stan_model,probs = c(0.025,0.975), digits_summary = 3)
```

```{r Coefficient plot, fig.height=6,fig.width=6, echo=FALSE}
coeff_plot(stan_model, params=c('c1','c0_mu','b1','b0_mu','a1','a0_mu'),
           labels = c('ho x wd','ho','beta x wd','beta','alpha x wd','alpha'),
           xlab = 'cloglog effect size', transform = NULL)
```

**Fig 7** log centered and scaled coefficients ± 95 (thin line) and 80% (thick line) Bayesian Credible Intervals. log_a = intercept; log_b = slope, log_ho = asymptote.

```{r Hazard rates,fig.height=6,fig.width=6, echo=FALSE}
par(mfrow=c(2,1),mar=c(4,3,1,1), xaxs='i')
plot_mortality(model_fit = stan_model ,model_data = stan_data, cov_name = 'rho',
               wd_values = c(200,800), growth_name = 'dbh_dt', mortality_curve = FALSE, xaxis_on = FALSE, haz_lim=c(0,2))

plot_mortality(model_fit = stan_model ,model_data = stan_data, cov_name = 'rho',
               wd_values = c(200,800), growth_name = 'dbh_dt', mortality_curve = TRUE, xaxis_on = TRUE, legend = FALSE)
```

**Fig 8** Predicted hazard rates (top) and mortality probability (bottom) for an average 200 (black) and 800 kg/m2 (red) kg/m2 wood density species. Errors are 95% credible intervals. rho = wood density.

```{r 3d plots, fig.height=10,fig.width=10, echo=FALSE}
par(mfrow=c(2,1))
plot3d(model_fit = stan_model, 
        model_data = stan_data, cov_name = 'rho', 
        wd_values = seq(200,800, by=10), 
        theta=145, phi=20,mortality_curve = FALSE)

plot3d(model_fit = stan_model, 
       model_data = stan_data, cov_name = 'rho', 
       wd_values = seq(200,800, by=10), 
       theta=145, phi=20,mortality_curve = TRUE)
```

**Fig 9** Mean predicted 3D surface for hazard rates (top) and mortality probability (bottom). Note that there is no 'significant' difference at low growth rate between low and high wood density. rho = wood density,This can be better seen in Fig. 8.

\pagebreak

##Notes/Questions

1 - How much data should we fit to this model? All of it would require the inclusion of a individual random effect to account for non-independence of repeatedly sampled individuals. We could subsample individuals in entire dataset... but again this would require non-independence to be account for. The advantage of this approach is we can also cross validate. Another way to cross validate is to run the model for each census on a subset of individuals spanning all species. Question is, how can we best stratify sampling to ensure low abundance species are also represented.

2 - Growth is a predominant predictor in mortality models, however, it is not the only one. Size is also commonly used as a predictor of mortality. A typical finding from these size-mortality models is that mortality appears to be U-shaped. Could we account for this by fitting dbh and dbh x wd interaction on the growth independent hazard (ho)? My concern is whether this makes sense as dbh and growth are highly correlated (~50%) consequently, these two predictors may not be identifiable.

3 - Possibly not a major concern with size not in the model, but currently the data only utilises data from individuals that are at least 1.3 m high and have a diameter of > 1 cm. This means that we are missing alot of the mortality at the young age. Is it worth investing time in cleaning and preparing the BCI seedling data for inclusion? A difficulty to face is that the seedling data measures height while the main BCI dataset measures dbh. Maybe some type of allometric model using BCI?

4 - Current model form at very low growth rate forces the hazard for high wood density species to become larger than low wood density species. Though this effect is not significant (95% credible intervals overlap)... it appears to be the result of the model form we are using. Discuss with Daniel.

5 - Set model up to run on cluster. I've managed to dramatically speed up convergence time with these models. Approximately 5 hours. However, if we plan to run models containing multiple combinations of 4 growth rate types x 3 parameter combinations. We will need to set this up on the cluster.

6 - Need to integrate code into remaker

\pagebreak

##Model Code

```{r stan model,eval=FALSE}

StanModel <- '
  data {
    int<lower=0> n_obs;
    int<lower=0, upper=1> y[n_obs];
    int<lower=0> n_spp;
    int<lower=1> spp[n_obs];
    vector[n_obs] census_length;
    vector[n_obs] dbh_dt;
    vector[n_spp] rho;
  }
transformed data { // centers and standardizes predictors
  vector[n_spp] cs_lnrho;
  vector[n_obs] s_dbh_dt;
  cs_lnrho <- (log(rho) - mean(log(rho)))/ (2*sd(log(rho)));
  s_dbh_dt <- dbh_dt/ (2*sd(dbh_dt)); 
}
parameters { // assumes uniform priors on all parameters
  real a0_raw[n_spp];
  real b0_raw[n_spp];
  real c0_raw[n_spp];
  real a1; // effect of rho on a_log
  real b1; // effect of rho on b_log
  real c1; // effect of rho on ho_log
  real a0_mu; // a_log effect for average species
  real<lower=0> a0_sigma; // a_log effect species variation
  real b0_mu;  // b_log effect for average species
  real<lower=0> b0_sigma; // b_log effect species variation
  real c0_mu;  // ho_log effect for average species
  real<lower=0> c0_sigma; // ho_log effect species variation
}

transformed parameters {
  real<lower=0, upper=1> p[n_obs];
  real a_log[n_spp];
  real b_log[n_spp];
  real ho_log[n_spp];
  real a0[n_spp];
  real b0[n_spp];
  real c0[n_spp];

  for (s in 1:n_spp){ # Add species random effect & effects of rho
    a0[s] <- a0_raw[s] * a0_sigma + a0_mu;
    b0[s] <- b0_raw[s] * b0_sigma + b0_mu;
    c0[s] <- c0_raw[s] * c0_sigma + c0_mu;
    log_a[s] <- a0[s] + a1 * cs_lnrho[s];
    log_b[s] <- b0[s] + b1 * cs_lnrho[s];
    log_ho[s] <- c0[s] + c1 * cs_lnrho[s];
  }

  for (i in 1:n_obs) { // Estimate p
    p[i] <- inv_cloglog(log(census_length[i] * (exp(log_a[spp[i]] - exp(log_b[spp[i]]) * s_dbh_dt[i]) + exp(log_ho[spp[i]]))));
  }
}

model {
  for (s in 1:n_spp) { // non-centered parameterization Papaspiliopoulos et al. (2007).
    a0_raw[s] ~ normal(0,1); // implies normal(a0_mu, a0_sigma);
    b0_raw[s] ~ normal(0,1); // implies normal(b0_mu, b0_sigma); 
    c0_raw[s] ~ normal(0,1); // implies normal(c0_mu, c0_sigma); 
  }
  // Sample Pr(Dying) from bernoulli
  y ~ bernoulli(p);
}
'
```
\pagebreak

##References

### 2015-03-30 R/utils.R
Made a few corrections to some of the plot functions

@@ -39,10 +39,10 @@ sim_mortality <- function(model_fit, model_data, cov_name='rho',
   new_cs_lnwd <- (log(wd_values) - mean(log(model_data[[cov_name]])))/ (2*sd(log(model_data[[cov_name]])))
   
   if (is.null(growth_range)){
-    new_s_dbh_dt <- seq(min(model_data[[growth_name]]),max(model_data[[growth_name]]), by=0.001)/ (2*sd(model_data[[growth_name]]))
+    new_s_dbh_dt <- seq(min(model_data[[growth_name]]),max(model_data[[growth_name]]), by=0.0001)/ (2*sd(model_data[[growth_name]]))
   }
   else {
-    new_s_dbh_dt <- seq(min(growth_range), max(growth_range), by=0.00001)/ (2*sd(model_data[[growth_name]]))}
+    new_s_dbh_dt <- seq(min(growth_range), max(growth_range), by=0.0001)/ (2*sd(model_data[[growth_name]]))}
   
   preds <- sapply(new_cs_lnwd, function(wd) {
     with(model_fit, 
@@ -68,25 +68,25 @@ sim_mortality <- function(model_fit, model_data, cov_name='rho',
 # Plot hazard and mortality curves
 plot_mortality <- function(model_fit, model_data, cov_name='rho', wd_values=c(200,800), 
                            growth_name='dbh_dt', growth_range=NULL, 
-                           mortality_curve = FALSE, xaxis_on=TRUE,legend=TRUE, haz_lim=NULL) {
+                           mortality_curve = FALSE, xaxis_on=TRUE,legend=TRUE, haz_lim=NULL,xlim=xlim) {
   
   output <- sim_mortality(model_fit = model_fit, model_data = model_data, cov_name= cov_name, wd_values=wd_values, growth_name=growth_name, growth_range=growth_range, mortality_curve = mortality_curve)
   
   if (mortality_curve == TRUE){
-    plot(mn[,1]~ new_s_dbh_dt, type='n', data=output, ylim=c(0,1),
-         xaxt='n', ylab='Mortality probability/year', xlab='DBH growth (m)')
+    plot(mn[,1]*100~ new_s_dbh_dt, type='n', data=output, ylim=c(0,100),
+         xaxt='n', ylab='Pr(Mortality)/year', xlab=NA)
     
[…]

### 2015-04-07 ms.Rmd
Option to use differnt growth measures

@@
 
-```{r dependencies, echo=FALSE, message=FALSE,warning=FALSE}
-source('R/utils.R')
-load('output/rho_stan_model.rda')
-library(rstan)
-```
-
 ##Introduction
@@
 
-![Fig1](figures/wyckoff2002.png)
+<!-- ![Fig1](figures/wyckoff2002.png) -->
 
@@
 
-![Fig2](figures/wright2010.jpg)
+<!-- ![Fig2](figures/wright2010.jpg) -->
 
@@
 
-![Fig3](figures/kraft2010.png)
+<!-- ![Fig3](figures/kraft2010.png) -->
 
@@
 
-![Fig4](figures/chave2009.png)
+<!-- ![Fig4](figures/chave2009.png) -->
 
@@
 ```{r Model output, echo=FALSE}
-print(stan_model,probs = c(0.025,0.975), digits_summary = 3)
+print(stan_output_combined,probs = c(0.025,0.975), digits_summary = 3)
 ```
@@
 ```{r Coefficient plot, fig.height=6,fig.width=6, echo=FALSE}
-coeff_plot(stan_model, params=c('c1','c0_mu','b1','b0_mu','a1','a0_mu'),
+coeff_plot(stan_output_combined, params=c('c1','c0_mu','b1','b0_mu','a1','a0_mu'),
            labels = c('ho x wd','ho','beta x wd','beta','alpha x wd','alpha'),
@@
 par(mfrow=c(2,1),mar=c(4,3,1,1), xaxs='i')
-plot_mortality(model_fit = stan_model ,model_data = stan_data, cov_name = 'rho',
-               wd_values = c(200,800), growth_name = 'dbh_dt', mortality_curve = FALSE, xaxis_on = FALSE, haz_lim=c(0,2))
+plot_mortality(model_fit = stan_output_combined ,model_data = stan_data, cov_name = 'rho',
+               wd_values = c(200,800), growth_name = 'growth_dt', mortality_curve = FALSE, xaxis_on = FALSE, haz_lim=c(0,2))
 
-plot_mortality(model_fit = stan_model ,model_data = stan_data, cov_name = 'rho',
-               wd_values = c(200,800), growth_name = 'dbh_dt', mortality_curve = TRUE, xaxis_on = TRUE, legend = FALSE)
+plot_mortality(model_fit = stan_output_combined ,model_data = stan_data, cov_name = 'rho',
+               wd_values = c(200,800), growth_name = 'growth_dt', mortality_curve = TRUE, xaxis_on = TRUE, legend = FALSE)
 ```
@@
 par(mfrow=c(2,1))
-plot3d(model_fit = stan_model, 
+plot3d(model_fit = stan_output_combined,
         model_data = stan_data, cov_name = 'rho',
@@
 
-plot3d(model_fit = stan_model, 
+plot3d(model_fit = stan_output_combined,
        model_data = stan_data, cov_name = 'rho',
@@
     vector[n_obs] census_length;
-    vector[n_obs] dbh_dt;
+    vector[n_obs] growth_dt;
     vector[n_spp] rho;
@@
   vector[n_spp] cs_lnrho;
-  vector[n_obs] s_dbh_dt;
+  vector[n_obs] s_growth_dt;
   cs_lnrho <- (log(rho) - mean(log(rho)))/ (2*sd(log(rho)));
-  s_dbh_dt <- dbh_dt/ (2*sd(dbh_dt)); 
+  s_growth_dt <- growth_dt/ (2*sd(growth_dt));
 }
@@
   for (i in 1:n_obs) { // Estimate p
-    p[i] <- inv_cloglog(log(census_length[i] * (exp(log_a[spp[i]] - exp(log_b[spp[i]]) * s_dbh_dt[i]) + exp(log_ho[spp[i]]))));
+    p[i] <- inv_cloglog(log(census_length[i] * (exp(log_a[spp[i]] - exp(log_b[spp[i]]) * s_growth_dt[i]) + exp(log_ho[spp[i]]))));
   }

### 2015-04-07 R/process_output.R

@@ -32,17 +32,17 @@ coeff_plot <- function(stanfit, params, labels=NULL, quantiles =c(0.025,0.975,0.
 
 # Model simulation
 sim_mortality <- function(model_fit, model_data, cov_name='rho',
-                          wd_values=c(200,800), growth_name='dbh_dt',
+                          wd_values=c(200,800), growth_name='growth_dt',
                           growth_range=NULL, mortality_curve = FALSE){
   model_fit <- extract(model_fit)
 
   new_cs_lnwd <- (log(wd_values) - mean(log(model_data[[cov_name]])))/ (2*sd(log(model_data[[cov_name]])))
 
   if (is.null(growth_range)){
-    new_s_dbh_dt <- seq(min(model_data[[growth_name]]),max(model_data[[growth_name]]), by=0.0001)/ (2*sd(model_data[[growth_name]]))
+    new_s_growth_dt <- seq(min(model_data[[growth_name]]),max(model_data[[growth_name]]), by=0.0001)/ (2*sd(model_data[[growth_name]]))
   }
   else {
-    new_s_dbh_dt <- seq(min(growth_range), max(growth_range), by=0.0001)/ (2*sd(model_data[[growth_name]]))}
+    new_s_growth_dt <- seq(min(growth_range), max(growth_range), by=0.0001)/ (2*sd(model_data[[growth_name]]))}
 
   preds <- sapply(new_cs_lnwd, function(wd) {
     with(model_fit,
@@ -51,62 +51,62 @@ sim_mortality <- function(model_fit, model_data, cov_name='rho',
            b_log = b0_mu + b1 * wd
            ho_log = c0_mu + c1 * wd
            if (mortality_curve ==TRUE){
-             preds <-1 - exp(-(exp(a_log - exp(b_log) * new_s_dbh_dt) + exp(ho_log)))
+             preds <-1 - exp(-(exp(a_log - exp(b_log) * new_s_growth_dt) + exp(ho_log)))
            }
            else {
-             preds <- exp(a_log - exp(b_log) * new_s_dbh_dt) + exp(ho_log)
+             preds <- exp(a_log - exp(b_log) * new_s_growth_dt) + exp(ho_log)
            }
          }, a0_mu, b0_mu, c0_mu, a1, b1, c1))}, simplify = 'array')
   output <- list(mn = apply(preds,c(1,3), mean),
[…]

### 2015-04-07 R/stan_models.R

@@ -9,13 +9,13 @@ run_single_stan_chain <- function(model, data, chain_id, iter=1000, seed=123){
        iter = iter, seed=seed, chains=1, chain_id=chain_id, refresh=-1)
 }
 
-stan_data_BCI <- function(data) {
+stan_data_BCI <- function(data, growth_measure = "dbh_dt") {
   list(
     n_obs = nrow(data),
     n_spp = length(unique(data$sp)),
     spp = as.numeric(factor(data$sp), as.character(unique(data$sp))),
     rho =  unique(data$rho),
-    dbh_dt = data$dbh_dt,
+    growth_dt = data[[growth_measure]],
     census_length = data$census_interval,
     y = as.integer(data$dead_next_census)
     )
@@ -31,14 +31,14 @@ data {
   int<lower=0> n_spp;
   int<lower=1> spp[n_obs];
   vector[n_obs] census_length;
-  vector[n_obs] dbh_dt;
+  vector[n_obs] growth_dt;
   vector[n_spp] rho;
 }
 transformed data { // centers and standardizes predictors
   vector[n_spp] cs_lnrho;
-  vector[n_obs] s_dbh_dt;
+  vector[n_obs] s_growth_dt;
   cs_lnrho <- (log(rho) - mean(log(rho)))/ (2*sd(log(rho)));
-  s_dbh_dt <- dbh_dt/ (2*sd(dbh_dt));
+  s_growth_dt <- growth_dt/ (2*sd(growth_dt));
 }
 parameters { // assumes uniform priors on all parameters
   real a0_raw[n_spp];
@@ -74,7 +74,7 @@ transformed parameters {
   }
 
   for (i in 1:n_obs) { // Estimate p
-    p[i] <- inv_cloglog(log(census_length[i] * (exp(a_log[spp[i]] - exp(b_log[spp[i]]) * s_dbh_dt[i]) + exp(ho_log[spp[i]]))));
+    p[i] <- inv_cloglog(log(census_length[i] * (exp(a_log[spp[i]] - exp(b_log[spp[i]]) * s_growth_dt[i]) + exp(ho_log[spp[i]]))));
   }
 }
 

### 2015-04-07 ms.Rmd
rename variable names in model

@@
 
-```{r stan model,eval=FALSE}
-
-StanModel <- '
-  data {
-    int<lower=0> n_obs;
-    int<lower=0, upper=1> y[n_obs];
-    int<lower=0> n_spp;
-    int<lower=1> spp[n_obs];
-    vector[n_obs] census_length;
-    vector[n_obs] growth_dt;
-    vector[n_spp] rho;
-  }
-transformed data { // centers and standardizes predictors
-  vector[n_spp] cs_lnrho;
-  vector[n_obs] s_growth_dt;
-  cs_lnrho <- (log(rho) - mean(log(rho)))/ (2*sd(log(rho)));
-  s_growth_dt <- growth_dt/ (2*sd(growth_dt));
-}
-parameters { // assumes uniform priors on all parameters
-  real a0_raw[n_spp];
-  real b0_raw[n_spp];
-  real c0_raw[n_spp];
-  real a1; // effect of rho on a_log
-  real b1; // effect of rho on b_log
-  real c1; // effect of rho on ho_log
-  real a0_mu; // a_log effect for average species
-  real<lower=0> a0_sigma; // a_log effect species variation
-  real b0_mu;  // b_log effect for average species
-  real<lower=0> b0_sigma; // b_log effect species variation
-  real c0_mu;  // ho_log effect for average species
-  real<lower=0> c0_sigma; // ho_log effect species variation
-}
-
-transformed parameters {
-  real<lower=0, upper=1> p[n_obs];
-  real a_log[n_spp];
-  real b_log[n_spp];
-  real ho_log[n_spp];
-  real a0[n_spp];
-  real b0[n_spp];
-  real c0[n_spp];
-
-  for (s in 1:n_spp){ # Add species random effect & effects of rho
-    a0[s] <- a0_raw[s] * a0_sigma + a0_mu;
-    b0[s] <- b0_raw[s] * b0_sigma + b0_mu;
-    c0[s] <- c0_raw[s] * c0_sigma + c0_mu;
-    log_a[s] <- a0[s] + a1 * cs_lnrho[s];
-    log_b[s] <- b0[s] + b1 * cs_lnrho[s];
-    log_ho[s] <- c0[s] + c1 * cs_lnrho[s];
-  }
-
-  for (i in 1:n_obs) { // Estimate p
-    p[i] <- inv_cloglog(log(census_length[i] * (exp(log_a[spp[i]] - exp(log_b[spp[i]]) * s_growth_dt[i]) + exp(log_ho[spp[i]]))));
-  }
-}
-
-model {
-  for (s in 1:n_spp) { // non-centered parameterization Papaspiliopoulos et al. (2007).
-    a0_raw[s] ~ normal(0,1); // implies normal(a0_mu, a0_sigma);
-    b0_raw[s] ~ normal(0,1); // implies normal(b0_mu, b0_sigma);
-    c0_raw[s] ~ normal(0,1); // implies normal(c0_mu, c0_sigma);
-  }
-  // Sample Pr(Dying) from bernoulli
-  y ~ bernoulli(p);
-}
-'
-```
 \pagebreak

### 2015-04-07 R/process_output.R

@@ -36,31 +36,31 @@ sim_mortality <- function(model_fit, model_data, cov_name='rho',
                           growth_range=NULL, mortality_curve = FALSE){
   model_fit <- extract(model_fit)
 
-  new_cs_lnwd <- (log(wd_values) - mean(log(model_data[[cov_name]])))/ (2*sd(log(model_data[[cov_name]])))
+  new_log_rho_cs <- (log(wd_values) - mean(log(model_data[[cov_name]])))/ (2*sd(log(model_data[[cov_name]])))
 
   if (is.null(growth_range)){
-    new_s_growth_dt <- seq(min(model_data[[growth_name]]),max(model_data[[growth_name]]), by=0.0001)/ (2*sd(model_data[[growth_name]]))
+    new_growth_dt_s <- seq(min(model_data[[growth_name]]),max(model_data[[growth_name]]), by=0.0001)/ (2*sd(model_data[[growth_name]]))
   }
   else {
-    new_s_growth_dt <- seq(min(growth_range), max(growth_range), by=0.0001)/ (2*sd(model_data[[growth_name]]))}
+    new_growth_dt_s <- seq(min(growth_range), max(growth_range), by=0.0001)/ (2*sd(model_data[[growth_name]]))}
 
-  preds <- sapply(new_cs_lnwd, function(wd) {
+  preds <- sapply(new_log_rho_cs, function(wd) {
     with(model_fit,
          mapply(function(a0_mu, b0_mu, c0_mu, a1, b1, c1) {
            a_log = a0_mu + a1 * wd
            b_log = b0_mu + b1 * wd
-           ho_log = c0_mu + c1 * wd
+           h_log = c0_mu + c1 * wd
            if (mortality_curve ==TRUE){
-             preds <-1 - exp(-(exp(a_log - exp(b_log) * new_s_growth_dt) + exp(ho_log)))
+             preds <-1 - exp(-(exp(a_log - exp(b_log) * new_growth_dt_s) + exp(h_log)))
            }
            else {
-             preds <- exp(a_log - exp(b_log) * new_s_growth_dt) + exp(ho_log)
+             preds <- exp(a_log - exp(b_log) * new_growth_dt_s) + exp(h_log)
            }
          }, a0_mu, b0_mu, c0_mu, a1, b1, c1))}, simplify = 'array')
   output <- list(mn = apply(preds,c(1,3), mean),
[…]

### 2015-04-07 R/stan_models.R

@@ -35,10 +35,10 @@ data {
   vector[n_spp] rho;
 }
 transformed data { // centers and standardizes predictors
-  vector[n_spp] cs_lnrho;
-  vector[n_obs] s_growth_dt;
-  cs_lnrho <- (log(rho) - mean(log(rho)))/ (2*sd(log(rho)));
-  s_growth_dt <- growth_dt/ (2*sd(growth_dt));
+  vector[n_spp] log_rho_cs;
+  vector[n_obs] growth_dt_s;
+  log_rho_cs  <- (log(rho) - mean(log(rho)))/ (2*sd(log(rho)));
+  growth_dt_s <- growth_dt / (2*sd(growth_dt));
 }
 parameters { // assumes uniform priors on all parameters
   real a0_raw[n_spp];
@@ -46,20 +46,20 @@ parameters { // assumes uniform priors on all parameters
   real c0_raw[n_spp];
   real a1; // effect of rho on a_log
   real b1; // effect of rho on b_log
-  real c1; // effect of rho on ho_log
+  real c1; // effect of rho on h_log
   real a0_mu; // a_log effect for average species
   real<lower=0> a0_sigma; // a_log effect species variation
   real b0_mu;  // b_log effect for average species
   real<lower=0> b0_sigma; // b_log effect species variation
-  real c0_mu;  // ho_log effect for average species
-  real<lower=0> c0_sigma; // ho_log effect species variation
+  real c0_mu;  // h_log effect for average species
+  real<lower=0> c0_sigma; // h_log effect species variation
 }
 
 transformed parameters {
   real<lower=0, upper=1> p[n_obs];
   real a_log[n_spp];
   real b_log[n_spp];
-  real ho_log[n_spp];
+  real h_log[n_spp];
   real a0[n_spp];
   real b0[n_spp];
   real c0[n_spp];
@@ -68,13 +68,13 @@ transformed parameters {
     a0[s] <- a0_raw[s] * a0_sigma + a0_mu;
     b0[s] <- b0_raw[s] * b0_sigma + b0_mu;
     c0[s] <- c0_raw[s] * c0_sigma + c0_mu;
-    a_log[s] <- a0[s] + a1 * cs_lnrho[s];
-    b_log[s] <- b0[s] + b1 * cs_lnrho[s];
[…]

### 2015-04-07 ms.Rmd
Simplify equations

@@
 
-$p_i$ is modelled as a function of census length (ranging from 3.8 to 5.8 years) and the sum of two independent hazards:
+$p_i$ is modelled as a function of census length (ranging from 3.8 to 5.8 years) and the cumulative hazard $\lambda_i$ for each individual:
 
 \begin{equation} \label{eq:pred_model}
-\log_e(-\log_e(1-p_i)) = \log_e\left(\textrm{census-length}_i \times
-\left(
-\overbrace{\exp\left(alpha_{s[i]} - \exp(beta_{s[i]}) \times
-\frac{\textrm{growth}_i}{2 \times \sigma(\textrm{growth})}
-\right)
-}^{\text{Growth dependent hazard}} +
-\overbrace{\exp(ho_{s[i]})
-}^{\text{Growth independent hazard}}\right)\right)
+\log_e(-\log_e(1-p_i)) = \log_e\left(\textrm{census-length}_i \times \lambda_i\right) 
 \end{equation}
 
-The growth-dependent hazard encapsulates the growth-mortality relationship by allowing the probability of dying to decrease exponentially with growth rate [@Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw].
-This relationship consists of two species varying parameters alpha, the intercept term that defines the instantaneous hazard rate at zero growth, and beta, the slope term that defines the decrease in hazard rate per scaled unit growth (Fig.6).
+Here, the cumulative hazard, $\lambda_i$, is the sum of two independent hazards - growth-dependent hazard (i.e.
+death via carbon starvation) and growth-independent (death by all other causes).
+Growth-dependent hazard encapsulates the growth-mortality relationship by allowing the probability of dying to decrease exponentially with growth rate [@Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw].
+This relationship consists of two species varying parameters $alpha$, the intercept term that defines the instantaneous hazard rate at zero growth, and $beta$, the slope term that defines the decrease in hazard rate per scaled unit growth (Fig.6).
 By contrast, the growth independent hazard is assumed to be linear and is defined by one species varying parameter, ho.
-Here, growth was estimated as $\frac{dbh_t - dbh_{t-1}}{\Delta{t}}$.
-
-Instead of calculating the probability of death for a given time period, we can also use this model form to calculcate the instantaneous mortality rate, $\lambda_i$ (also known as the hazard rate; @Kleinbaum:2012tn), for each individual $i$:
+Note: we estimate growth as $\frac{dbh_t - dbh_{t-1}}{\Delta{t}}$.
 

### 2015-04-08 ms.Rmd
Added example rstan code to implement kfold cross validation

@@
 
-##Model Code
+## K-fold Cross validation example excerp from [Vehtari & Gelman 2014](http://www.stat.columbia.edu/~gelman/research/unpublished/waic_stan.pdf)
+
+To implement K-fold cross-validation in Stan we need to repeatedly fit the model to one subset of the data and use it to predict the held-out set.
+The way we recommend setting this up is to do the partitioning in R (or Python, or whatever data-processing environment is being used) and then pass the training data and held-out data to Stan in two pieces.
+With linear regression, for example, we start with the Stan model on page 4, passing in the training data as N, y, X and the held-out set as N holdout, y holdout, X holdout, with the data block augmented accordingly.
+We then keep the data block and model as is, and just alter the generated quantities block to operate on the held-out data.
+The code looks like this:
+
+```{r eval=FALSE}
+􏰋
+￼￼￼data { int N;
+  int J;
+vector[N] y;
+matrix[N,J] X;
+int N_holdout;
+vector[N_holdout] y_holdout;
+matrix[N_holdout,J] X_holdout;
+}
+  parameters {
+    vector[J] b;
+real<lower=0> sigma;
+  }
+  model {
+    y ~ normal(X*b, sigma);
+increment_log_prob(-log(sigma));
+  }
+  generated quantities {
+    vector[N_holdout] log_lik;
+for (n in 1:N_holdout){
+  // data model
+// log prior for p(sigma) propto 1/sigma
+log_lik[n] <- normal_log(y_holdout[n], X_holdout[n]*b, sigma);
+9
+} }
+
+```
 

### 2015-04-09 ms.Rmd
Remove old results, add notes on cross-val

@@
 
-##Introduction
+## Introduction
+
 Understanding who lives or dies, and why, is critical to accurately predict population, community and ecosystem dynamics [@Hawkes:2000ib; @McDowell:2011dr].
@@
 
-##Data
+## Data
 
@@
 
-##Model Description
+## Model Description
 
@@
 
-\pagebreak
+## Analysis plan
 
-##Results
-**Take home message:**
-Wood density decreases intercept (log_a) & further increases the steepness of the growth-mortality slope (log_b). It also decreases the growth-independent hazard log_ho
-
-```{r Model output, echo=FALSE}
-print(stan_output_combined,probs = c(0.025,0.975), digits_summary = 3)
-```
-
-```{r Coefficient plot, fig.height=6,fig.width=6, echo=FALSE}
-coeff_plot(stan_output_combined, params=c('c1','c0_mu','b1','b0_mu','a1','a0_mu'),
-           labels = c('ho x wd','ho','beta x wd','beta','alpha x wd','alpha'),
-           xlab = 'cloglog effect size', transform = NULL)
-```
-
-**Fig 7** log centered and scaled coefficients ± 95 (thin line) and 80% (thick line) Bayesian Credible Intervals. log_a = intercept; log_b = slope, log_ho = asymptote.
-
-```{r Hazard rates,fig.height=6,fig.width=6, echo=FALSE}
-par(mfrow=c(2,1),mar=c(4,3,1,1), xaxs='i')
-plot_mortality(model_fit = stan_output_combined ,model_data = stan_data, cov_name = 'rho',
-               wd_values = c(200,800), growth_name = 'growth_dt', mortality_curve = FALSE, xaxis_on = FALSE, haz_lim=c(0,2))
-
-plot_mortality(model_fit = stan_output_combined ,model_data = stan_data, cov_name = 'rho',
-               wd_values = c(200,800), growth_name = 'growth_dt', mortality_curve = TRUE, xaxis_on = TRUE, legend = FALSE)
-```
-
-**Fig 8** Predicted hazard rates (top) and mortality probability (bottom) for an average 200 (black) and 800 kg/m2 (red) kg/m2 wood density species.
-Errors are 95% credible intervals.
-rho = wood density.
-
-```{r 3d plots, fig.height=10,fig.width=10, echo=FALSE}
-par(mfrow=c(2,1))
-plot3d(model_fit = stan_output_combined,
-        model_data = stan_data, cov_name = 'rho',
-        wd_values = seq(200,800, by=10),
-        theta=145, phi=20,mortality_curve = FALSE)
-
-plot3d(model_fit = stan_output_combined,
-       model_data = stan_data, cov_name = 'rho',
-       wd_values = seq(200,800, by=10),
-       theta=145, phi=20,mortality_curve = TRUE)
-```
-
-**Fig 9** Mean predicted 3D surface for hazard rates (top) and mortality probability (bottom).
-Note that there is no 'significant' difference at low growth rate between low and high wood density.
-rho = wood density,This can be better seen in Fig.
-8.
+Step 1 - define a general structure of model we think will capture data, on basis of prior knowledge
+Step 2 - assuming species effects on all pars, ask which growth meausre best predicts data, compare 4 growth rates using cross validation
+Step 3 - fit full model with trait effects, look at effect sizes of WD on a, b and c
+Step 4 - based on 3, choose some subsets of models, compare these using both bayes factor (most parsimonious) and predictive capacity
 
@@
 
-##Notes/Questions
+## Results
+**Take home message:**
+Wood density decreases intercept (log_a) & further increases the steepness of the growth-mortality slope (log_b). It also decreases the growth-independent hazard log_ho
+\pagebreak
+
+## Notes/Questions
 
@@
 
-## K-fold Cross validation example excerp from [Vehtari & Gelman 2014](http://www.stat.columbia.edu/~gelman/research/unpublished/waic_stan.pdf)
+## Cross validation
+
+We will implement *K-fold cross validation*.
+
+According to [Vehtari & Gelman 2014](http://www.stat.columbia.edu/~gelman/research/unpublished/waic_stan.pdf):
 
@@
 We then keep the data block and model as is, and just alter the generated quantities block to operate on the held-out data.
-The code looks like this:
 
-```{r eval=FALSE}
-􏰋
-￼￼￼data { int N;
-  int J;
-vector[N] y;
-matrix[N,J] X;
-int N_holdout;
-vector[N_holdout] y_holdout;
-matrix[N_holdout,J] X_holdout;
-}
-  parameters {
-    vector[J] b;
-real<lower=0> sigma;
-  }
-  model {
-    y ~ normal(X*b, sigma);
-increment_log_prob(-log(sigma));
-  }
-  generated quantities {
-    vector[N_holdout] log_lik;
-for (n in 1:N_holdout){
-  // data model
-// log prior for p(sigma) propto 1/sigma
-log_lik[n] <- normal_log(y_holdout[n], X_holdout[n]*b, sigma);
-9
-} }
+For a model $M$, training data $D$, and test data $D^*$, we want to calculate
 
-```
+$$p(D^* | D,M) = \int p(\theta | D, M) \, p(D^* | \theta, M) \, d \theta$$
+
+The distribution $p(\theta | D, M)$ gives the likelihood of parameter values $\theta$ given the training data (obtained from fitting the model) and is calculated (via Bayes Rule) as
+
+$$p(\theta | D, M) = \frac{p(\theta) \, p(D|\theta)}{p(D)}.$$
+
+$p(D^* | \theta, M)$ is then the probability of observing the test data given a particular parameter set (i.e the likelihood of the data) and is calculated as
+$$p(D^* | \theta, M) = \prod_i p(D^*_i | \theta, M).$$
+
+Log-likelihoods are easier to deal with and so we have
+
+$$\log p(D^* | \theta, M) = \sum_i \log p(D^*_i | \theta, M).$$
+
+Aki \& Gelman 2014 provide methods for estimating "an importance-sampling approximated LOO directly using the log-likelihood evaluated at the posterior simulations of the parameter values"
 
@@
 
-##References
+## References

### 2015-06-10 R/model.R
Changed generated quanties to extract likelihood for each observation

@@ -8,7 +8,7 @@ get_model_chunks <- function(pars) {
     pars = c("a0_raw","a0_mu","a0_sigma","a0","a1_raw","a1_mu","a1_sigma","a1",if("a" %in% rho_combo) "a2",
              "b0_raw","b0_mu","b0_sigma","b0","b1_raw","b1_mu","b1_sigma","b1",if("b" %in% rho_combo) "b2",
              "c0_raw","c0_mu","c0_sigma","c0","c1_raw","c1_mu","c1_sigma","c1",if("c" %in% rho_combo) "c2",
-             "log_lik_fit_total","log_lik_tilde_total"),
+             "log_lik_fit", "log_lik_tilde","log_lik_fit_total","log_lik_tilde_total"),
     parameters = sprintf("
     real a0_raw[n_spp];
     real a0_mu;
@@ -120,7 +120,7 @@ get_model_chunks <- function(pars) {
       real b_log_fit;
       real c_log_fit;
       real cumulative_hazard_fit;
-      real log_lik_fit;
+      real log_lik_fit[n_obs];
       real log_lik_fit_total;
       
       // Heldout objects
@@ -128,7 +128,7 @@ get_model_chunks <- function(pars) {
       real b_log_tilde;
       real c_log_tilde;
       real cumulative_hazard_tilde;
-      real log_lik_tilde;
+      real log_lik_tilde[n_obs_tilde];
       real log_lik_tilde_total;
       
       // Initialization of summed log likelihoods
@@ -142,12 +142,12 @@ get_model_chunks <- function(pars) {
         c_log_fit <- c0[spp[i]] + c1[census[i]]%s;
         cumulative_hazard_fit <- -census_length[i] * (exp(a_log_fit - exp(b_log_fit) * growth_dt[i]) + exp(c_log_fit));
         if (y[i] == 0) {
-          log_lik_fit <- cumulative_hazard_fit;
+          log_lik_fit[i] <- cumulative_hazard_fit;
         }
         else {
-          log_lik_fit <- log1m_exp(cumulative_hazard_fit);
+          log_lik_fit[i] <- log1m_exp(cumulative_hazard_fit);
         }
-        log_lik_fit_total <- log_lik_fit_total + log_lik_fit;
+        log_lik_fit_total <- log_lik_fit_total + log_lik_fit[i];
       }
       
       // log likelihood for held out data
[…]

### 2015-06-10 ms.Rmd
Edited text in methods

@@
 
-In this study we build a hierarchical Bayesian logistic regression that attempts to partitions 15-years of tropical tree mortality into growth-dependent and growth-independent hazards (i.e.
+In this study we build a hierarchical Bayesian cloglog regression that attempts to partition 15-years of tropical tree mortality into growth-dependent and growth-independent hazards (i.e.
 death not related to growth).
@@
 ## Data
+We evaluate how wood density affects both the growth–mortality trade-off and growth-indepenent mortality using survival and growth data of 179 940 trees spanning 203 species from within the 50 ha tropical rainforest plot on Barro Colorado Island (BCI), Panama.
+Within this plot, the diameter and survival status of all free-standing woody plants with a diameter at breast height (dbh) greater or equal to 1 cm were recorded in 1982, 1990, 1995, 2000, 2005 and 2010 @BarroColoradofores:2012up.
+For the purpose of this study, we excluded data collected in 1982 and 1985 censuses because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
+We also excluded palms and ferns because they lack dbh growth @Wright:2010fl, multistemmed plants, species that did not contain a wood density estimate or contained fewer than 10 individuals and individuals that were not consistently measured at 1.3 m or that 'returned from the dead'.
+We also removed individuals that were deemed to contain erroneous observations using the CTFS error checker.
+This removed individuals that had annual dbh growth rates greater than 75 mm/yr and individuals that had excessively negative growth rates (> - 4 mm/yr).
 
-We are fitting our model to growth and mortality data collected from a tropical rainforest on Barro Colorado Island, Panama @BarroColoradofores:2012up.
-This rainforest plot has been surveyed every 5 years since 1985 and includes size, growth and status (alive, dead) observations from more than 200,000 individuals across 286 species.
+Because we were interested in determining how growth influenced annual mortality rates and whether the type of growth rate affected the  predictive capacity or the effects of wood density we calculated four commonly used growth rate measures: absolute dbh and basal growth, and relative dbh and basal growth.
+Absolute dbh growth rate was calculated as: $\frac{dbh_t - dbh_{t-1}}{\Delta{days/365.25}}$ and relative dbh growth was calculated as: $\frac{ \log_e(dbh_t) - \log_e(dbh_{t-1})}{\Delta{days/365.25}}$.
+Basal area for each individual in each census was calculated as: $0.25 \times \pi \times dbh_t^2$.
+Absolute and relative basal area growth were then calculated the same way as dbh.
+All growth rates were measured in metres/yr.
 
-In this study we only use data recorded between 1990 and 2010 for single stemmed plants.
-Furthermore, we only used species that had an estimate for wood density.
-We removed multistemmed plants because of difficulties discerning mainstem.
+Wood density for each species was estimated by coring trees located within 15 km of the BCI plot because increment borers are prohibited within the plot @Wright:2010fl.
+Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100oc) and dividing by the fresh volume (as measured by water displacement).
+Wood density was then calculated as an area-weighted average, where area refers to the annulus represented by each piece, assuming a circular trunk.
+These measurements were then converted to kg/m3.
 
-For more details about the data, or to download it, please click [here](https://repository.si.edu/handle/10088/20925).
-Currently species trait data is not publically available.
+##Model Description
 
-## Model Description
+We extend a common functional form, $M= \alpha \times \exp(-\beta \times growth)$, used to examine the growth-mortality relationship [@Kobe:1997vy] to include mortality associated with growth independent processes.
+Specifically, we extend this negative exponential to include an asymptote parameter,$\gamma$, such that the functional form now becomes $M= \alpha \times \exp(-\beta \times growth) + \gamma$.
+The advantage of adding $\gamma$ is that when growth rates are high the negative exponential asymptotes at a mortality rate above zero.
+This allows for deaths to occur at high growth rates that would otherwise not be accounted for in the original functional form.
+Another advantage of this functional form is that all parameters are biologically interpretable.
+$\alpha$ and $\beta$ are interpreted similarly to how they have been elsewhere [kobe1997]; $\alpha$ represents the mortality rate at zero growth (and controls where the function crosses the y-axis) and $\beta$ represents the sensitivity of mortality rates to changes in growth rate (corresponding to the function's decay).
+We interpret $\gamma$ as the growth-independent mortality rate (corresponding to where the function asymptotes).
+This parameter accounts for mortality not related to growth (e.g.
+windfall, fire).
 
-We used a hierarchical generalized linear model to examine how wood density affected growth-dependent and growth-independent tree mortality.
-These models are well suited to data sets that have a hierarchical structure (e.g.
-individuals within species) because they can account for observation error and partition both explained and unexplained variation at multiple levels of a dataset [@Gelman:2007te; @Kery:2010tp].
-We used Bayesian inference and fitted models to data using Markov chain Monte Carlo (MCMC) sampling in R 3.1.3 using package rstan version 2.6 [@StanDevelopmentTeam:2015uz].
-Before commencing model fit, we first scaled growth by dividing by two standard deviations.
-We then centered and scaled wood density, $rho$, by first log-transforming it and then substracting its mean and dividing by two standard deviations.
-The advantage of centering and standardizing is that it allows for better mixing of chains and easier interpretation, with intercepts interpreted as average responses and slope terms as partial dependencies conditional on other continuous variables at their mean.
-Furthermore, standardizing by two standard deviations allows the magnitude of effects to be directly compared between binary and continuous covariates (Gelman and Hill 2007).
+While this model can be used to examine mortality for individual species, it has limited capacity to provide general insights into the mechanisms affecting multiple species.
+One way add this capacity is to allow $\alpha$, $\beta$ and $\gamma$ to vary by traits (measures of species strategy).
+Furthermore, we can also include a species level random effect on these parameters that will effectively account for observation error and partition both explained and unexplained variation at multiple levels of a dataset (e.g.
+individuals, species, census)[@Gelman:2007te; @Kery:2010tp].
+
+Using this model, coupled with k-fold cross validation we address four questions:
+1) Which growth measure is most predictive of plant mortality?
+2) Does wood density influence both the growth-mortality trade-off and growth independent mortality?
+3) Does the effects of wood density vary with growth measure?
+and 4) What is the most parismonous mortality model in terms of wood density effects?
+
+##Model implementation
+
+We used Bayesian inference and fit models to data using Markov chain Monte Carlo (MCMC) sampling in R 3.1.3 using package rstan version 2.6 [@StanDevelopmentTeam:2015uz].
+Before commencing model fit, we first log transformed wood density, $\rho$, and then centered it on 600 kg m3 (i.e.
+$\log_e(\rho_i) - \log_e(600)).
+This was done for easier interpretation, with intercepts interpreted as average woody species responses (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant) (Gelman and Hill 2007).
 Three independent chains were executed and in all cases modeled parameters converged within 1000 iterations.
@@
 The residuals of the model were checked graphically against predictions with none showing any systematic pattern or severe heteroscedasticity.
-To ensure the effect of wood density was identifiable, and to examine it's effect on all three parameters (i.e growth-mortality intercept and slope, and growth-independent hazard) we selectively removed it from each and visually inspected whether the direction and magnitude of effects were severely altered.
 Below we describe the full model structure, parameters and priors.
-We also supply the stan code below.
+We also supply the stan code (see here).
 
-We were interested in examining how wood density affected growth-dependent and growth-independent individual tree mortality.
-However, our observations, $y_i$, were binary (0 = alive, 1 = dead) for each individual $i$.
+Our observations, $y_i$, were binary (0 = alive, 1 = dead) for each individual $i$.
 Consequently, we modeled these observations as a random realization from a Bernoulli distribution based on the model-fitted probability that individual $i$ would die, $p_i$:
@@
 
-We modelled the probability of drying, $p_i$, using a complimentary-log-log link, $\log_e(-\log_e(1-p))$.
+The probability of drying, $p_i$, was modelled using a complimentary-log-log link, $\log_e(-\log_e(1-p))$.
 Unlike the traditional logit link, $\log_e(p/1-p)$, the complimentary-log-log allows regression coefficients to be interpreted as hazard ratios.
@@
 
-$p_i$ is modelled as a function of census length (ranging from 3.8 to 5.8 years) and the cumulative hazard $\lambda_i$ for each individual:
+$p_i$ is modelled as a function of census length (ranging from 3.8 to 5.8 years) and the instantaneous hazard rate $\lambda_i$ for each individual:
 
@@
 
-Here, the cumulative hazard, $\lambda_i$, is the sum of two independent hazards - growth-dependent hazard (i.e.
-death via carbon starvation) and growth-independent (death by all other causes).
-Growth-dependent hazard encapsulates the growth-mortality relationship by allowing the probability of dying to decrease exponentially with growth rate [@Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw].
-This relationship consists of two species varying parameters $alpha$, the intercept term that defines the instantaneous hazard rate at zero growth, and $beta$, the slope term that defines the decrease in hazard rate per scaled unit growth (Fig.6).
-By contrast, the growth independent hazard is assumed to be linear and is defined by one species varying parameter, ho.
-Note: we estimate growth as $\frac{dbh_t - dbh_{t-1}}{\Delta{t}}$.
+Here, the instantaneous hazard, $\lambda_i$, is the sum of two independent hazards.
+The first is the growth-dependent hazard (i.e.
+death via carbon starvation, as described by $\alpha \times \exp(-\beta \times growth)$).
+This encapsulates the growth-mortality relationship by allowing the instantaneous mortality rate to decrease exponentially with growth rate [@Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw].
+The second hazard is the growth-independent (death by all other causes, as described by $\gamma$).
 
@@
 }^{\text{Growth dependent hazard}} +
-\overbrace{\exp(ho_{s[i]})
+\overbrace{\exp(gamma_{s[i]})
 }^{\text{Growth independent hazard}}\right)
@@
 
-We modelled $alpha$, $beta$ and $ho$, by allowing their intercepts to vary by species and by including wood density as a species level predictor:
+To account for species strategy and partition variance into explained and unexplained at the observation, species and sampling period level we modelled $alpha$, $beta$ and $ho$, by allowing their intercepts to vary by species and sampling period and by including wood density as a species level predictor:
 
 \begin{equation} \label{eq:alpha}
-alpha_s = \alpha0_s + \alpha1 \times \frac{(\log_e(\rho_s) - \mu(\log_e(\rho)))}{2 \times(\sigma(\log(\rho)))}
+alpha_s = \alpha0_s + \alpha1_t + \alpha2 \times (\log_e(\rho_s) - \log_e(600))
 \end{equation}
@@
 \begin{equation} \label{eq:beta}
-beta_s = \beta0_s + \beta1 \times \frac{(\log_e(\rho_s) - \mu(\log_e(\rho)))}{2 \times(\sigma(\log(\rho)))}
+beta_s = \beta0_s + \beta1_t + \beta2 \times (\log_e(\rho_s) - \log_e(600))
 \end{equation}
 
-\begin{equation} \label{eq:ho}
-ho_s = \gamma0_s + \gamma1 \times \frac{(\log_e(\rho_s) - \mu(\log_e(\rho)))}{2 \times(\sigma(\log(\rho)))}
+\begin{equation} \label{eq:gamma}
+ho_s = \gamma0_s + \gamma1_t + \gamma2 \times (\log_e(\rho_s) - \log_e(600))
 \end{equation}
 
-Here, the terms $\alpha1$, $\beta1$ and $\gamma1$ represent the effect of wood density on the intercept, slope and growth-independent hazard, respectively.
-The terms, $\alpha0_s$, $\beta0_s$ and $\gamma0_s$, respectively represent the growth-dependent intercept and slope and growth-independent intercept for species $s$ within which individual $i$ belongs.
-We estimates these varying terms from a normal distribution centred on the average species intercept (i.e.
-$\alpha_{\mu}$,$\gamma_{\mu}$) or slope ($\beta_{\mu}$) and the standard deviation of the species variation ($\alpha spp$,$\beta spp$, $\gamma spp$).
+Here, the terms $\alpha2$, $\beta2$ and $\gamma2$ represent the effect of log transformed wood density on $\alpha$, $\beta$ and $\gamma$.
+The terms, $\alpha0_s$, $\beta0_s$ and $\gamma0_s$, represent the species-level random effect for species $s$ within which individual $i$ belongs.
+$\alpha1_t$, $\beta1_t$ and $\gamma1_t$, represents the sample-level random effect for which time $t$ an individual was sampled.
+Both random effects were estimated from a normal distribution centred on a species or sample time mean (i.e.
+$\alpha0_{\mu}$, $\beta0_{\mu}$,$\gamma0_{\mu}$) and $\alpha1_{\mu}$,$\beta1_{\mu}$,$\gamma1_{\mu}$, respectively) and an associated standard deviation (i.e.
+$\alpha0_{\mu}$, $\beta0_{\sigma}$,$\gamma0_{\sigma}$) and $\alpha1_{\sigma}$,$\beta1_{\sigma}$,$\gamma1_{\sigma}$, respectively).
 
 \begin{equation} \label{eq:alpha0}
-\alpha_s \sim \textrm{Normal}(\alpha_{\mu}, \sigma_{\alpha spp})
+\alpha_s \sim \textrm{Normal}(\alpha_{\mu}, \sigma_{\alpha0 spp})
 \end{equation}
@@
 \begin{equation} \label{eq:beta0}
-\beta_s \sim \textrm{Normal}(\beta_{\mu}, \sigma_{\beta spp})
+\beta_s \sim \textrm{Normal}(\beta_{\mu}, \sigma_{\beta0 spp})
 \end{equation}
@@
 \begin{equation} \label{eq:gamma0}
-\gamma_s \sim \textrm{Normal}(\gamma_{\mu}, \sigma_{\gamma spp})
+\gamma_s \sim \textrm{Normal}(\gamma_{\mu}, \sigma_{\gamma0 spp})
 \end{equation}
 
+#Priors - THIS NEEDS UPDATING
 Because we used Bayesian inference we needed to specify prior distributions for all model parameters.
-For wood density effects ($\alpha1$, $\beta1$ and $\gamma1$) and mean hyperparamters ($\alpha_{\mu}$, $\beta_{mu}$ and $\gamma_{\mu}$) we used stan's default uninformative prior, $\textrm{Uniform(-2,2)}.
+The effects of wood density on all three parameters ($\alpha2$, $\beta2$ and $\gamma2$), were sampled from weakly informative normal priors with a mean of 0 and a standard deviation of 5.
+
+census random effects: we assumed little variation so centred on zero with a standard deviation of 0.1
+alpha species mu - we extracted from Kobe et al 1995 and are the parameter estimates currently used in SORTIE
+gamma species mu - we extracted from chambers 2013
+beta species mu - we had no idea about, especially as this parameter is likely to change with different growth rates. As such we fitted a uniform prior bounded between -20 and 20. 
+All sigma's were drawn from a half cauchy distribution with a mean of zero and a standard deviation of 2.5.
+
+and mean hyperparamters ($\alpha_{\mu}$, $\beta_{mu}$ and $\gamma_{\mu}$) we used stan's default uninformative prior, $\textrm{Uniform(-2,2)}.
 For hyperparamater standard deviations we used $\textrm{Uniform(0,2)}$.

### 2015-06-10 R/bci_data_processing.R
filtered out individuals that had census intervals >8 yrs suggesting that a census period in between was missed

@@ -168,7 +168,8 @@ BCI_calculate_individual_growth <- function(BCI_data, spp_table) {
       dead_next_census = mortality_in_next_census(dfstatus),
       dbh_prev = c(NA, drop_last(dbh))) %>%
     # Only keep alive stems
-    filter(dfstatus=="alive") %>%
+    filter(dfstatus=="alive" &
+           census_interval < 8) %>%
     filter(CTFS_sanity_check(dbh, dbh_increment, dbh_dt) & 
              !is.na(dbh_dt) & 
              !is.na(dead_next_census)) %>%

### 2015-07-29 ms.Rmd
More edits

@@
 
-## Introduction
+# Introduction
 
 Understanding who lives or dies, and why, is critical to accurately predict population, community and ecosystem dynamics [@Hawkes:2000ib; @McDowell:2011dr].
-In plants, the carbon budget is considered a fundamental determinant of mortality, such that if rates of carbon assimilation are low or become insufficient for growth, an individual is more likely to die [@Peet:1987va; @Hawkes:2000ib; @Keane:2001db; @McDowell:2011dr].
-Empirical studies have broadly supported this assumption, with many studies showing that the probability of dying decreasing exponentially with growth rate [Fig 1; @Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw], and consequently most dynamic vegetation models, including tree gap models [@Botkin:1972ii; @Keane:2001db] and Dynamic Global Vegetation Models (DGVM's; @Woodward:2004ft) use this growth-mortality relationship as the predominant, and in many cases, only cause of plant mortality [@Hawkes:2000ib; @McDowell:2011dr].
+In plants, the carbon budget is considered a fundamental determinant of mortality, such that if rates of carbon assimilation are low or become insufficient for growth, an individual is more susceptible to dying [@Peet:1987va; @Hawkes:2000ib; @Keane:2001db; @McDowell:2011dr].
+Empirical studies examining intraspecific mortality rates have broadly supported this assumption, with many finding that the probability a plant dies decreases exponentially with increasing growth rates [@Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw].
+Consequently, most dynamic vegetation models, including tree gap models [@Botkin:1972ii; @Keane:2001db], SORTIE (REF) and Dynamic Global Vegetation Models (DGVM's; @Woodward:2004ft) use the growth-mortality relationship as the predominant, and in some cases, only mechanism describing plant mortality [@Hawkes:2000ib; @McDowell:2011dr].
+Furthermore, the vast majority of these models assume the shape of this relationship is constant across species.
 
-<!-- ![Fig1](figures/wyckoff2002.png) -->
-
-**Fig 1:** The relationship between growth and mortality for seven co-occurring tree species in the southern Appalachian Mountains. From @Wyckoff:2002ul.
-
-However, empirical evidence has highlighted that this growth-mortality relationship varies between species with those that grow intrinsically faster exhibiting higher mortality rates relative to those which are slower growing (Fig 2; @Poorter:2008iu; @Wright:2010fl).
-Plant functional traits – attributes of species that influence vital rates (i.e.
-survival, growth and reproduction) and ultimately fitness [@Ackerly:2003eb; @Poorter:2008iu] – have been used to examine this inter-specific variability.
-
-<!-- ![Fig2](figures/wright2010.jpg) -->
-
-**Fig 2:** The growth–mortality trade-off for saplings expressed as the 95th percentile relative growth rate (RGR95) vs.
-the mortality rate of the slowest growing 25% of individuals (M25) for 103 tree species on Barro Colorado Island, Panama.
-From @Wright:2010fl.
-
-In tropical rain forests, wood density, the biomass invested per unit wood volume, is one such functional trait that has been repeatedly identified as being negatively correlated with tree mortality [Fig.
-3; @Nascimento:2005wx; @King:2006he; @AlvarezClare:2007gv; @Chao:2008hg; @Poorter:2008iu; @Kraft:2010kq; @Wright:2010fl].
-This relationship is thought to be due to dense wood conveying enhanced physical resistance to growth-independent causes (Fig 4) of mortality such as stem breakage [@Putz:1983wu; @vanGelder:2006ex; @Chave:2009iy], wood herbivory (NEED REF), embolism [@Hacke:2001kj, @Jacobsen:2005fx, @Preston:2006jn] and fungal and pathogen attack [@Augspurger:1984wx].
-However, wood density may also increase survival associated with growth-dependent mortality.
+This is despite empirical evidence highlighting that mortality rates vary considerably between species.
+One common finding from these multispecies studies that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species [@Poorter:2008iu; @Wright:2010fl].
+Ecologists have attempted to explain this variation between species by using functional traits - attributes of species that are thought to influence vital rates (i.e.
+survival, growth and reproduction) and ultimately fittness [@Ackerly:2003eb; @Poorter:2008iu].
+Wood density, the biomass invested per unit wood volume, is one such functional trait that has been shown to be negatively correlated with tropical tree mortality [@Chave et al 2004; @Nascimento:2005wx; @King:2006he; @AlvarezClare:2007gv; @Chao:2008hg; @Poorter:2008iu; @Kraft:2010kq; @Wright:2010fl].
+The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to stem breakage [@Putz:1983wu; @vanGelder:2006ex; @Chave:2009iy], wood herbivory (NEED REF), embolism [@Hacke:2001kj, @Jacobsen:2005fx, @Preston:2006jn] and fungal and pathogen attack [@Augspurger:1984wx].
+However, some research has also suggested that wood density may increase survival associated with growth-dependent mortality.
 For example, studies [e.g.
@@
 
-<!-- ![Fig3](figures/kraft2010.png) -->
+Most plant mortality research can be broadly seperated into two research focuses: 1) those that focus on estimating individual mortality within species; and 2) those that use suites of species to examine correlations between functional traits and population level estimates of tree mortality.
+However, there appears to be little cross over between these two research focuses.
+This is despite both approaches likely being required to develop a general model of tree mortality.
+A model that accounts for variation both within and between species.
 
-**Fig 3:** The relationship between wood density and mortality in a global tropical forest dataset. From @Kraft:2010kq
-
-<!-- ![Fig4](figures/chave2009.png) -->
-
-**Fig 4:** Wood mechanical traits plotted against wood density. From @Chave:2009iy
+In this study, we reconcile both approaches by fitting 15-years of growth and mortality data from 203 tropical rainforest species encompassing more than 180,000 individuals in a Bayesian hierarchical model that explicitly includes variability observed both within and between species.
+Specifically, our model extends the negative exponential functional form commonly observed between growth and mortality in intraspecific studies by adding a growth-independent baseline hazard and allowing each parameter to vary as a function of wood density.
+We then use this model to determine if wood density only influence growth-independent mortality (i.e.
+the baseline hazard/asymptote) or whether it can also influence growth-dependent paramters such as the intercept (i.e.
+the rate of mortality at zero growth) or the rate (i.e.
+the sensitivity of mortality rates to changes with growth rate)?
 
@@
 
-In this study we build a hierarchical Bayesian cloglog regression that attempts to partition 15-years of tropical tree mortality into growth-dependent and growth-independent hazards (i.e.
-death not related to growth).
-We then use this model to address three questions: 1) Does wood density affect the growth-mortality relationship?
-2) Where does wood density have the greatest effect growth-dependent or growth-independent hazards?
-And 3) Do the observed relationships dependent on the choice of growth measure?
+# Methods
 
-We hypothesize that wood density will decrease both growth-dependent and growth-independent sources of mortality.
-Specifically, we expect wood density to affect the growth-mortality relationship by decreasing its intercept, (i.e.
-mortality rates at zero growth) and steepening its slope (i.e.
-more negative, Fig 5).
-We also expect that the effect of wood density will be greatest for growth-independent mortality, and that these relationships will be broadly consistent across growth measures.
+## Study area
 
-```{r hypotheses,fig.height=6,fig.width=6, echo=FALSE}
-par(mfrow=c(2,1), mar=c(2,4,1,1), oma=c(3,0,0,0), xaxs='i')
-curve(exp(1 -exp(-0.3) * x) + exp(-2), xlim=c(0,12), ylim=c(0,5),
-xlab=NA, ylab='Hazard rate', xaxt='n')
-curve(exp(0.5 -exp(1) * x) + exp(-4), add=TRUE, col='red')
-axis(1, at=0)
-
-legend('topright', c('low rho', 'high rho'), col=c('black','red'), lty=1, bty='n')
-curve((1-exp(-(exp(1 - exp(-0.3) * x) + exp(-1.5))))*100, xlim=c(0,12), ylim=c(0,100),
-xlab=NA, ylab='Pr(Mortality)/year', xaxt='n')
-curve((1-exp(-(exp(0.5 - exp(1) * x) + exp(-4))))*100, add=TRUE, col='red')
-abline(v=0, lty=2, col='blue')
-axis(1, at=0)
-mtext(side=1, 'Growth rate',line = 1, outer=TRUE)
-
-```
-
-**Fig 5:** Hypotheses of how wood density will affect growth-mortality curve.
-Top = Hazard curves (i.e.
-instantaneous mortality rates); Bottom = Mortality probability/year based on estimated hazard rates.
-Where curve asymptotes is dependent on wood density effects on growth-independent hazard.
-Note changes in intercept, slope and asymptote between both lines.
-
-\pagebreak
+We analysed growth and survival data collected from a 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15&deg;N, 79.85&deg;W).
+The plot is predominately undisturbed old-growth rainforest (48-ha), but 2 ha is of 100 year old secondary rainforest (REF).
+The climate at Barro Colorado Island is warm throughout the year with a mean annual temperature of 26.1&deg;.
+Rainfall is seasonal with most of the 2500 mm falling between April and November (REF).
+The island is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI), which was granted long-term custodianship over the island by Panama's Environmental Authority.
+STRI have permission to establish the 50-ha plot as a permanent census in 1980.
+Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in (REFS).
 
 ## Data
-We evaluate how wood density affects both the growth–mortality trade-off and growth-indepenent mortality using survival and growth data of 179 940 trees spanning 203 species from within the 50 ha tropical rainforest plot on Barro Colorado Island (BCI), Panama.
-Within this plot, the diameter and survival status of all free-standing woody plants with a diameter at breast height (dbh) greater or equal to 1 cm were recorded in 1982, 1990, 1995, 2000, 2005 and 2010 @BarroColoradofores:2012up.
-For the purpose of this study, we excluded data collected in 1982 and 1985 censuses because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
-We also excluded palms and ferns because they lack dbh growth @Wright:2010fl, multistemmed plants, species that did not contain a wood density estimate or contained fewer than 10 individuals and individuals that were not consistently measured at 1.3 m or that 'returned from the dead'.
-We also removed individuals that were deemed to contain erroneous observations using the CTFS error checker.
-This removed individuals that had annual dbh growth rates greater than 75 mm/yr and individuals that had excessively negative growth rates (> - 4 mm/yr).
-
-Because we were interested in determining how growth influenced annual mortality rates and whether the type of growth rate affected the  predictive capacity or the effects of wood density we calculated four commonly used growth rate measures: absolute dbh and basal growth, and relative dbh and basal growth.
-Absolute dbh growth rate was calculated as: $\frac{dbh_t - dbh_{t-1}}{\Delta{days/365.25}}$ and relative dbh growth was calculated as: $\frac{ \log_e(dbh_t) - \log_e(dbh_{t-1})}{\Delta{days/365.25}}$.
-Basal area for each individual in each census was calculated as: $0.25 \times \pi \times dbh_t^2$.
-Absolute and relative basal area growth were then calculated the same way as dbh.
-All growth rates were measured in metres/yr.
+Within the 50-ha BCI plot the diameter at breast height and survival status of all free-standing woody plants with a diameter greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter @BarroColoradofores:2012up.
+For the purpose of this study, we discarded data collected in 1982 and 1985 censuses because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
+We excluded individual trees that did not survive at least two censuses (two being required to estimate growth rates), were not consistently measured at 1.3 m, were multistemmed, or had resprouted or 'returned from the dead'.
+We also discarded species that do not exhibit secondary growth (e.g.
+palms and ferns), contained fewer than 10 individuals or did not contain a wood density estimate were also discarded.
+Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25% of their initial dbh were also removed.
+Lastly, in order to reduce computational requirements we randomly selected one observation per individual.
+In total, 180,500 individuals of 203 species were included in our analysis.
 
@@
 
-##Model Description
+## Measurment error and calculating true growth
+Observed dbh measurements are likely to include measurement error that can manifest into biologically unrealistic scenarios of negative growth and ultimately erroneous conclusions.
+We estimate this measurement error using another BCI dataset consisting of 1562 randomly selected trees remeasured within 30 days (see XXXX).
+Assuming that no growth occurred between observations, we estimated the standard deviation of measurement error (and its uncertainty) by sampling the discrepencies for each individual, $i$, as random realisations from a normal distribution centered on zero with a standard deviation, $\sigma_{error}$, which describes the variation in error:
 
-We extend a common functional form, $M= \alpha \times \exp(-\beta \times growth)$, used to examine the growth-mortality relationship [@Kobe:1997vy] to include mortality associated with growth independent processes.
-Specifically, we extend this negative exponential to include an asymptote parameter,$\gamma$, such that the functional form now becomes $M= \alpha \times \exp(-\beta \times growth) + \gamma$.
-The advantage of adding $\gamma$ is that when growth rates are high the negative exponential asymptotes at a mortality rate above zero.
-This allows for deaths to occur at high growth rates that would otherwise not be accounted for in the original functional form.
-Another advantage of this functional form is that all parameters are biologically interpretable.
-$\alpha$ and $\beta$ are interpreted similarly to how they have been elsewhere [kobe1997]; $\alpha$ represents the mortality rate at zero growth (and controls where the function crosses the y-axis) and $\beta$ represents the sensitivity of mortality rates to changes in growth rate (corresponding to the function's decay).
-We interpret $\gamma$ as the growth-independent mortality rate (corresponding to where the function asymptotes).
-This parameter accounts for mortality not related to growth (e.g.
-windfall, fire).
+\begin{equation} \label{eq:error_model}
+discepency_i \sim \textrm{N}(0, \sigma_{error}). \end{equation}
 
-While this model can be used to examine mortality for individual species, it has limited capacity to provide general insights into the mechanisms affecting multiple species.
-One way add this capacity is to allow $\alpha$, $\beta$ and $\gamma$ to vary by traits (measures of species strategy).
-Furthermore, we can also include a species level random effect on these parameters that will effectively account for observation error and partition both explained and unexplained variation at multiple levels of a dataset (e.g.
-individuals, species, census)[@Gelman:2007te; @Kery:2010tp].
+$\sigma_{error}$ was estimated to be 0.75 (sd = 0.01), meaning that 95% of the measurement errors were within +/- 1.47 cm of the true diameter.
+We then used this standard deviation of 0.75 in another Bayesian model to estimate the initial and final true dbh (i.e.
+observation - measurement error) for each individual.
+This was done by first modelling the observed initial (dbh1) and final dbh (dbh2) of each individual, $i$, as a random realisation from a normal distribution centered on their respective true estimates (i.e.
+$\widehat{dbh1_i}$ and $\widehat{dbh2_i}$), with a standard deviation of 0.75 (i.e.
+$\sigma_{error}$).
 
+\begin{equation} \label{eq:obs dbh1}
+dbh1_i \sim \textrm{N}(\widehat{dbh1_i}, 0.75). \end{equation}
+
+\begin{equation} \label{eq:obs dbh2}
+dbh2_i \sim \textrm{N}(\widehat{dbh2_i}, 0.75). \end{equation}
+
+$\widehat{dbh1_i}$ was estimated from a lognormal distribution:
+
+\begin{equation} \label{eq:true dbh1}
+\hat{dbh1_i} \sim \textrm{lognormal}(\mu_{dbh1}, \sigma_{dbh1}), \end{equation}
+
+where, $\mu_{dbh1}$ defines the log of the grand mean initial dbh, and $\sigma_{dbh1}$ defines that variation around that mean.
+$\mu_{dbh1}$ was estimated using a normal prior $\textrm{N}(0, 2.5)$, which represents our prior expectation that the mean initial dbh would likely occur between 0 and 148 cm.
+$\sigma_{dbh1}$ was estimated using a weakly informative half cauchy prior centered on zero and a scale parameter of 2.5.
+
+$\widehat{dbh2_i}$ was estimated as:
+
+\begin{equation} \label{eq:true dbh2}
+\widehat{dbh2_i} + R_i \times T_i, \end{equation}
+
+where, $R_i$ represents the true annual increase in dbh (cm) and $T_i$ represents the number of days between measurements divided by 365.25.
+We modelled $R_i$ as a random realisation from a lognormal distribution centered on zero and a standard deviation of 1, conveying our prior expectation that annual increases in dbh would be between 0.13 and 7.4 cm/yr.
+We used a lognormal as it ensured dbh growth cannot be negative.
+We then used the mean posterior estimate of $R_i$ as our estimate of true dbh growth for the subsequent mortality analysis.
+
+## Mortality model
+
+We extend a common functional form used to model the growth-mortality relationship, $M= \alpha \times \exp(-\beta \times growth)$ [e.g.
+@Kobe:1997vy] to include a growth-independent baseline hazard, such that the functional form becomes $M= \alpha \times \exp(-\beta \times growth) + \gamma$.
+By adding $\gamma$ we allow the functional form to asymptote at a mortality rate above zero, which allows for deaths to occur at high growth rates that would otherwise not be accounted for.
+The advantage of this functional form is that all parameters are biologically interpretable.
+$\alpha$ represents the intercept and biologically defines the sum of mortality rate at zero growth and baseline hazard.
+$\beta$ represents the function's decay, or in otherwords, the sensitivity of mortality rates to changes with growth rate.
+$\gamma$, as stated before, represents asymptote and is referred to as the baseline hazard.
+$\alpha$ and $\beta$ are intrinsically linked to the growth-mortality relationship, and as such are growth-dependent parameters.
+By contrast, $\gamma$ accounts for mortality not related to growth (e.g.
+windfall, fire) and consquently is a growth-independent parameter.
+
+While this model can be used to examine and predict mortality for individual species, it has limited capacity to provide general insights into the mechanisms affecting multiple species.
+We add this capacity by allowing $\alpha$, $\beta$ and $\gamma$ to vary by wood denisty - a species trait thought to convey a species growth strategy (REFs).
+Furthermore, we also allow the three parameters to vary by species ID and census period in which the data were collected so that the model could partition explained and unexplained variance between census, species and individual levels of the dataset [@Gelman:2007te; @Kery:2010tp].
+
+## Predictive capacity
 Using this model, coupled with k-fold cross validation we address four questions:
@@
 
-##Model implementation
+## Model implementation
 

### 2015-07-31 ms.Rmd
Rework decription of model, algorithm for efficient numerical calculation

@@
 survival, growth and reproduction) and ultimately fittness [@Ackerly:2003eb; @Poorter:2008iu].
-Wood density, the biomass invested per unit wood volume, is one such functional trait that has been shown to be negatively correlated with tropical tree mortality [@Chave et al 2004; @Nascimento:2005wx; @King:2006he; @AlvarezClare:2007gv; @Chao:2008hg; @Poorter:2008iu; @Kraft:2010kq; @Wright:2010fl].
+Wood density, the biomass invested per unit wood volume, is one such functional trait that has been shown to be negatively correlated with tropical tree mortality [@Chave:2009iy; @Nascimento:2005wx; @King:2006he; @AlvarezClare:2007gv; @Chao:2008hg; @Poorter:2008iu; @Kraft:2010kq; @Wright:2010fl].
 The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to stem breakage [@Putz:1983wu; @vanGelder:2006ex; @Chave:2009iy], wood herbivory (NEED REF), embolism [@Hacke:2001kj, @Jacobsen:2005fx, @Preston:2006jn] and fungal and pathogen attack [@Augspurger:1984wx].
@@
 
+## Mortality model
+
+We extend a common functional form used to model the growth-mortality relationship [e.g. @Kobe:1997vy] to include a growth-independent baseline hazard, such that the functional form becomes
+\begin{equation} \label{eq:lambda}
+\lambda = \overbrace{\alpha \, \exp\left(-\beta \, G \right)}^{\text{Growth dependent}} + \overbrace{\gamma}^{\text{Growth independent}}
+\end{equation}
+
+By adding $\gamma$ we allow the functional form to asymptote at a mortality rate above zero, which allows for deaths to occur at high growth rates that would otherwise not be accounted for.
+The advantage of this functional form is that all parameters are biologically interpretable.
+$\alpha$ represents the intercept and biologically defines the sum of mortality rate at zero growth and baseline hazard.
+$\beta$ represents the function's decay, or in other words, the sensitivity of mortality rates to changes with growth rate.
+$\gamma$, as stated before, represents asymptote and is referred to as the baseline hazard.
+$\alpha$ and $\beta$ are intrinsically linked to the growth-mortality relationship, and as such are growth-dependent parameters.
+By contrast, $\gamma$ accounts for mortality not related to growth (e.g.
+windfall, fire) and consequently is a growth-independent parameter.
+
+While this model can be used to examine and predict mortality for individual species, it has limited capacity to provide general insights into the mechanisms affecting multiple species.
+We add this capacity by allowing $\alpha$, $\beta$ and $\gamma$ to vary by wood density - a species trait thought to convey a species growth strategy (REFs).
+Furthermore, we also allow the three parameters to vary by species ID and census period in which the data were collected so that the model could partition explained and unexplained variance between census, species and individual levels of the dataset [@Gelman:2007te; @Kery:2010tp].
+
+Using this model, coupled with k-fold cross validation we address four questions:
+
+1. Which growth measure is most predictive of plant mortality?
+2. Does wood density influence both the growth-mortality trade-off and growth independent mortality?
+3. Does the effects of wood density vary with growth measure? and
+4. What is the most parsimonious mortality model in terms of wood density effects?
+
 \pagebreak
@@
 
-## Study area
+## Data
 
@@
 
-## Data
-Within the 50-ha BCI plot the diameter at breast height and survival status of all free-standing woody plants with a diameter greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter @BarroColoradofores:2012up.
+Within the 50-ha BCI plot the diameter at breast height and survival status of all free-standing woody plants with a diameter greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter [@BarroColoradofores:2012up].
 For the purpose of this study, we discarded data collected in 1982 and 1985 censuses because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
-We excluded individual trees that did not survive at least two censuses (two being required to estimate growth rates), were not consistently measured at 1.3 m, were multistemmed, or had resprouted or 'returned from the dead'.
+We excluded individual trees that did not survive at least two censuses (two being required to estimate growth rates), were not consistently measured at 1.3 m, were multi-stemmed, or had re-sprouted or 'returned from the dead'.
 We also discarded species that do not exhibit secondary growth (e.g.
 palms and ferns), contained fewer than 10 individuals or did not contain a wood density estimate were also discarded.
-Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25% of their initial dbh were also removed.
+Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial dbh were also removed.
 Lastly, in order to reduce computational requirements we randomly selected one observation per individual.
@@
 
-Wood density for each species was estimated by coring trees located within 15 km of the BCI plot because increment borers are prohibited within the plot @Wright:2010fl.
-Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100oc) and dividing by the fresh volume (as measured by water displacement).
+Wood density for each species was estimated by coring trees located within 15 km of the BCI plot because increment borers are prohibited within the plot [@Wright:2010fl].
+Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100&deg;C) and dividing by the fresh volume (as measured by water displacement).
 Wood density was then calculated as an area-weighted average, where area refers to the annulus represented by each piece, assuming a circular trunk.
-These measurements were then converted to kg/m3.
+These measurements were then converted to kg m$^-3$.
 
-## Measurment error and calculating true growth
+## Model description
+
+We have binary observations $y_i$ (0 = alive, 1 = dead) for each individual $i$, and a hypothesised the underlying mortality model giving an instantaneous rate of mortality (or hazard) per unit time (eq.
+\ref{eq:lambda}).
+Based on standard demographic theory, the probability of death for individual $i$ across the census interval $t_1\rightarrow t_2$ is
+\begin{equation} \label{eq:pred_model}
+p_{i,t_1\rightarrow t_2} = 1 - \exp\left(\Lambda_{i,t_1\rightarrow t_2} \right),
+\end{equation}
+where $\Lambda_{i,t} = \int_{t1}^{t2} \lambda(t) \, dt$ is the cumulative hazard for that individual.
+Given suitable estimates of $\lambda$, one can model the observations as a random realization from a Bernoulli distribution based on the probability that individual $i$ would die, $p_{i,t_1\rightarrow t_2}$:
+\begin{equation} \label{eq:obs_model}
+y_{i,t_1\rightarrow t_2} \sim \textrm{bernoulli}(p_{i,t_1\rightarrow t_2}).\end{equation}
+
+Our first hypothesis is that the cumulative hazard rate within the focal census interval is reasonably approximated by the growth rate in the previous interval
+\begin{equation} \label{eq:hazard2}
+\Lambda_{i,t} \approx (t_2-t_1) \lambda (t_1).
+\end{equation}
+Growth rates are estimated .... (see below).
+
+Second, we assume that the key parameters of eq.
+\ref{eq:lambda} are influenced by 3 factors: census year, species identify, and trait.
+Because traits are measured at species-level, the trait based effect is logically a subset of species level effect.
+To account for species strategy and partition variance into explained and unexplained at the observation, species and sampling period level we modelled $\alpha$, $\beta$ and $\gamma$ via an equation of the form:
+\begin{equation} \label{eq:alpha}
+\alpha_s = \alpha_{1,t} \, \alpha_{0,s} \, \rho_s^{\alpha_{2}}.
+\end{equation}
+Here $\alpha_{1,t}$ represents the sample-level random effect for which time $t$ an individual was sampled, $\alpha_{0,s}$ represents the species-level random effect of species $s$ within which individual $i$ belongs, and $\alpha_{2}$ represents the effect of standardised wood density $\rho_s$ on $\alpha_s$.
+The form of eq \ref{eq:alpha} ensures that parameters remain positive.
+On a log scale equation \ref{eq:alpha} becomes an additive model,$\log \alpha_s = \log \alpha_{0,s} + \log \alpha_{1,t} + \alpha_{2} \, \log \rho_s$, equivalent to a classic proportional hazards  model widely used in survival analysis.
+We standardised wood density via the transformation: $\rho_s = \rho/600$, an operation which is equivalent to centring a log-transformed variable at a density of 600 kg m$^-3$.
+This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 600 (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant) [@Gelman:2007te].
+
+The species and census-level random effects $\alpha_{0,s}, \alpha_{1,t}$ were assumed to follow a log-normal distribution:
+\begin{equation} \label{eq:alpha0}
+\log \alpha_i \sim \mathcal{N}\left(\mu_{\log \alpha_i}, \sigma_{\log \alpha_i}\right),\\
+\end{equation}
+where $\mu_{\log \alpha_i},  \sigma_{\log \alpha_i}$ are the mean and standard deviation of log-transformed variables.
+
+Similar forms used for $\beta$ and $\gamma$.
+
+## Model fitting
+
+We used Bayesian inference and fit models to data using Markov chain Monte Carlo (MCMC) sampling in R 3.1.3 using package rstan version 2.6 [@StanDevelopmentTeam:2015uz].
+Three independent chains were executed and in all cases modelled parameters converged within 1000 iterations.
+Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic [@Gelman:1992ts; @Brooks:1998ju].
+After discarding the first 1000 iterations as ‘burn in’, a further 1000 iterations were taken from the joint posterior.
+The residuals of the model were checked graphically against predictions with none showing any systematic pattern or severe heteroscedasticity.
+
+To optimise calculations in stan, we tried to minimise the number of logarithmic and exponential transformations, resulting in the following algorithm:
+
+1. Estimate $\alpha, \beta, \gamma$
+  - Draw a number from standard lognormal distribution with location ($\mu$) and scale ($\sigma$) parameters being the mean and sd of logged variable.
+Note that this draw returns variable on non-logged axes, even though location and scale are specified on logged axes.
+It is equivalent to drawing a normal number on log-transformed axes and back transforming.
+$$z \sim \textrm{lognormal}\left(\mu_{\log \alpha_i}, \sigma_{\log \alpha_i}\right)$$
+  - model for $\mu_{\log \alpha_i}$ and $\sigma_{\log \alpha_i}$ stays the same, with priors being mean and variance of log-transformed values.
+2.  Estimate cumulative hazard $\Lambda$ as
+$$\Lambda = -T \, \left(\alpha_0\, \alpha_1 , \rho_s^{\alpha_2} \exp(\beta_0\, \beta_1 , \rho_s^{\beta_2}  * G) + \gamma_0\, \gamma_1 , \rho_s^{\gamma_2}\right)$$
+3. Increment log probability density by appropriate amount.
+  - If $y=0$ we have $$\Pr(y=0) = 1 - \Pr(y=1) = 1 - (1- \exp(-\Lambda) = \exp(-\Lambda),$$ so the log probability density is $$\log \Pr(y=0) = \log(\exp(-\Lambda)) = -\Lambda$$.
+ - If $y=1$ we have $\Pr(y=1) = 1 - \exp(-\Lambda)$, so the log probability density is $$\log \Pr(y=1) = \log(1-\exp(-\Lambda)) = \textrm{log1m\_exp}(-\Lambda)$$.
+
+### Measurement error and calculating true growth
+
 Observed dbh measurements are likely to include measurement error that can manifest into biologically unrealistic scenarios of negative growth and ultimately erroneous conclusions.
@@
 \begin{equation} \label{eq:true dbh2}
-\widehat{dbh2_i} + R_i \times T_i, \end{equation}
+\widehat{dbh2_i} + R_i \, T_i, \end{equation}
 
@@
 
-## Mortality model
+### Priors (NEEDS UPDATING)
 
-We extend a common functional form used to model the growth-mortality relationship, $M= \alpha \times \exp(-\beta \times growth)$ [e.g.
-@Kobe:1997vy] to include a growth-independent baseline hazard, such that the functional form becomes $M= \alpha \times \exp(-\beta \times growth) + \gamma$.
-By adding $\gamma$ we allow the functional form to asymptote at a mortality rate above zero, which allows for deaths to occur at high growth rates that would otherwise not be accounted for.
-The advantage of this functional form is that all parameters are biologically interpretable.
-$\alpha$ represents the intercept and biologically defines the sum of mortality rate at zero growth and baseline hazard.
-$\beta$ represents the function's decay, or in otherwords, the sensitivity of mortality rates to changes with growth rate.
-$\gamma$, as stated before, represents asymptote and is referred to as the baseline hazard.
-$\alpha$ and $\beta$ are intrinsically linked to the growth-mortality relationship, and as such are growth-dependent parameters.
-By contrast, $\gamma$ accounts for mortality not related to growth (e.g.
-windfall, fire) and consquently is a growth-independent parameter.
-
-While this model can be used to examine and predict mortality for individual species, it has limited capacity to provide general insights into the mechanisms affecting multiple species.
-We add this capacity by allowing $\alpha$, $\beta$ and $\gamma$ to vary by wood denisty - a species trait thought to convey a species growth strategy (REFs).
-Furthermore, we also allow the three parameters to vary by species ID and census period in which the data were collected so that the model could partition explained and unexplained variance between census, species and individual levels of the dataset [@Gelman:2007te; @Kery:2010tp].
-
-## Predictive capacity
-Using this model, coupled with k-fold cross validation we address four questions:
-1) Which growth measure is most predictive of plant mortality?
-2) Does wood density influence both the growth-mortality trade-off and growth independent mortality?
-3) Does the effects of wood density vary with growth measure?
-and 4) What is the most parismonous mortality model in terms of wood density effects?
-
-## Model implementation
-
-We used Bayesian inference and fit models to data using Markov chain Monte Carlo (MCMC) sampling in R 3.1.3 using package rstan version 2.6 [@StanDevelopmentTeam:2015uz].
-Before commencing model fit, we first log transformed wood density, $\rho$, and then centered it on 600 kg m3 (i.e.
-$\log_e(\rho_i) - \log_e(600)).
-This was done for easier interpretation, with intercepts interpreted as average woody species responses (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant) (Gelman and Hill 2007).
-Three independent chains were executed and in all cases modeled parameters converged within 1000 iterations.
-Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic [@Gelman:1992ts; @Brooks:1998ju].
-After discarding the first 1000 iterations as ‘burn in’, a further 1000 iterations were taken from the joint posterior.
-The residuals of the model were checked graphically against predictions with none showing any systematic pattern or severe heteroscedasticity.
-Below we describe the full model structure, parameters and priors.
-We also supply the stan code (see here).
-
-Our observations, $y_i$, were binary (0 = alive, 1 = dead) for each individual $i$.
-Consequently, we modeled these observations as a random realization from a Bernoulli distribution based on the model-fitted probability that individual $i$ would die, $p_i$:
-
-\begin{equation} \label{eq:obs_model}
-p_i \sim \textrm{bernoulli}(p_i) \end{equation}
-
-The probability of drying, $p_i$, was modelled using a complimentary-log-log link, $\log_e(-\log_e(1-p))$.
-Unlike the traditional logit link, $\log_e(p/1-p)$, the complimentary-log-log allows regression coefficients to be interpreted as hazard ratios.
-That is, the hazard for one individual relative to another [@Kleinbaum:2012tn].
-It is also asymmetrical and can explicitly account for variable census length [@Allison:1982vd; @Kleinbaum:2012tn], which is particularly important with large-scale monitoring programs such as BCI, where the period of time between revisits of individual trees varies within a single census because there are so many individuals that they cannot be observed simulataneously [@Condit:1995ux].
-
-$p_i$ is modelled as a function of census length (ranging from 3.8 to 5.8 years) and the instantaneous hazard rate $\lambda_i$ for each individual:
-
-\begin{equation} \label{eq:pred_model}
-\log_e(-\log_e(1-p_i)) = \log_e\left(\textrm{census-length}_i \times \lambda_i\right)
-\end{equation}
-
-Here, the instantaneous hazard, $\lambda_i$, is the sum of two independent hazards.
-The first is the growth-dependent hazard (i.e.
-death via carbon starvation, as described by $\alpha \times \exp(-\beta \times growth)$).
-This encapsulates the growth-mortality relationship by allowing the instantaneous mortality rate to decrease exponentially with growth rate [@Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw].
-The second hazard is the growth-independent (death by all other causes, as described by $\gamma$).
-
-\begin{equation} \label{eq:lambda}
-\lambda_i =
-\left(
-\overbrace{\exp\left(alpha_{s[i]} - \exp(beta_{s[i]}) \times
-\frac{\textrm{growth}_i}{2 \times \sigma(\textrm{growth})}
-\right)
-}^{\text{Growth dependent hazard}} +
-\overbrace{\exp(gamma_{s[i]})
-}^{\text{Growth independent hazard}}\right)
-\end{equation}
-
-```{r model description, fig.height=6,fig.width=6, echo=FALSE}
-par(mfrow=c(2,1), mar=c(3,4,1,1), xaxs='i')
-       curve(exp(1 -exp(-0.3) * x) + exp(-4), xlim=c(0,12), ylim=c(0,5),
-             xlab=NA, ylab='Hazard rate',xaxt='n', col='red')
-       arrows(x0 = 2, y0 = 2.7, x1 = 0 ,y1 = 2.7)
-       text(3.1,2.7, 'alpha')
-       arrows(x0 = 4, y0 = 1.5, x1 = 1.5 ,y1 = 1)
-       text(4.5,1.7, 'beta')
-       arrows(x0 = 11, y0 = 0.8, x1 = 11 ,y1 = 0.018)
-       text(11,1, 'ho')
-       axis(1, at=0, labels= 0)
-
-       curve((1-exp(-(exp(1 - exp(-0.3) * x) + exp(-4))))*100, xlim=c(0,12), ylim=c(0,100),
-             xlab=NA, ylab='Pr(Mortality)', xaxt='n', col='red')
-       text(3.1,93, 'alpha')
-       arrows(x0 = 2, y0 = 93, x1 = 0 ,y1 = 93)
-       text(6,30, 'beta')
-       arrows(x0 = 5, y0 = 30, x1 = 3 ,y1 = 30)
-       text(11,15, 'ho')
-       arrows(x0 = 11, y0 = 10, x1 = 11 ,y1 = exp(-4)*100)
-       axis(1, at=0, labels= 0)
-       mtext(side=1, 'Growth rate',line = 2)
-```
-
-**Fig 6:** Interpreting model parameters.
-Top = Hazard curves (i.e.
-instantaneous mortality rates); Bottom = Mortality probability/year based on estimated hazard rates.
-The intercept at zero growth is defined by $alpha$.
-The slope is defined by $beta$.
-The asymptote is defined by $ho$ and represents the growth-independent hazard.
-
-To account for species strategy and partition variance into explained and unexplained at the observation, species and sampling period level we modelled $alpha$, $beta$ and $ho$, by allowing their intercepts to vary by species and sampling period and by including wood density as a species level predictor:
-
-\begin{equation} \label{eq:alpha}
-alpha_s = \alpha0_s + \alpha1_t + \alpha2 \times (\log_e(\rho_s) - \log_e(600))
-\end{equation}
-
-\begin{equation} \label{eq:beta}
-beta_s = \beta0_s + \beta1_t + \beta2 \times (\log_e(\rho_s) - \log_e(600))
-\end{equation}
-
-\begin{equation} \label{eq:gamma}
-ho_s = \gamma0_s + \gamma1_t + \gamma2 \times (\log_e(\rho_s) - \log_e(600))
-\end{equation}
-
-Here, the terms $\alpha2$, $\beta2$ and $\gamma2$ represent the effect of log transformed wood density on $\alpha$, $\beta$ and $\gamma$.
-The terms, $\alpha0_s$, $\beta0_s$ and $\gamma0_s$, represent the species-level random effect for species $s$ within which individual $i$ belongs.
-$\alpha1_t$, $\beta1_t$ and $\gamma1_t$, represents the sample-level random effect for which time $t$ an individual was sampled.
-Both random effects were estimated from a normal distribution centred on a species or sample time mean (i.e.
-$\alpha0_{\mu}$, $\beta0_{\mu}$,$\gamma0_{\mu}$) and $\alpha1_{\mu}$,$\beta1_{\mu}$,$\gamma1_{\mu}$, respectively) and an associated standard deviation (i.e.
-$\alpha0_{\mu}$, $\beta0_{\sigma}$,$\gamma0_{\sigma}$) and $\alpha1_{\sigma}$,$\beta1_{\sigma}$,$\gamma1_{\sigma}$, respectively).
-
-\begin{equation} \label{eq:alpha0}
-\alpha_s \sim \textrm{Normal}(\alpha_{\mu}, \sigma_{\alpha0 spp})
-\end{equation}
-
-\begin{equation} \label{eq:beta0}
-\beta_s \sim \textrm{Normal}(\beta_{\mu}, \sigma_{\beta0 spp})
-\end{equation}
-
-\begin{equation} \label{eq:gamma0}
-\gamma_s \sim \textrm{Normal}(\gamma_{\mu}, \sigma_{\gamma0 spp})
-\end{equation}
-
-#Priors - THIS NEEDS UPDATING
 Because we used Bayesian inference we needed to specify prior distributions for all model parameters.
-The effects of wood density on all three parameters ($\alpha2$, $\beta2$ and $\gamma2$), were sampled from weakly informative normal priors with a mean of 0 and a standard deviation of 5.
+The effects of wood density on all three parameters ($\alpha_{2}$, $\beta_{2}$ and $\gamma_{2}$), were sampled from weakly informative normal priors with a mean of 0 and a standard deviation of 5.
 
@@
 
-and mean hyperparamters ($\alpha_{\mu}$, $\beta_{mu}$ and $\gamma_{\mu}$) we used stan's default uninformative prior, $\textrm{Uniform(-2,2)}.
+and mean hyperparamters ($\mu_{\alpha_0}$, $\beta_{mu}$ and $\gamma_{\mu}$) we used stan's default uninformative prior, $\textrm{Uniform(-2,2)}.
 For hyperparamater standard deviations we used $\textrm{Uniform(0,2)}$.
+
+## To do
+
+Change so that baseline effect is census year, not species
 
@@
 A typical finding from these size-mortality models is that mortality appears to be U-shaped.
-Could we account for this by fitting dbh and dbh x wd interaction on the growth independent hazard (ho)?
+Could we account for this by fitting dbh and dbh \, wd interaction on the growth independent hazard (ho)?
 My concern is whether this makes sense as dbh and growth are highly correlated (~50%) consequently, these two predictors may not be identifiable.
@@
 Approximately 5 hours.
-However, if we plan to run models containing multiple combinations of 4 growth rate types x 3 parameter combinations.
+However, if we plan to run models containing multiple combinations of 4 growth rate types \, 3 parameter combinations.
 We will need to set this up on the cluster.
@@
 
+# Figures
+
+```{r model description, fig.height=6,fig.width=6, echo=FALSE}
+par(mfrow=c(2,1), mar=c(3,4,1,1), xaxs='i')
+       curve(exp(1 -exp(-0.3) * x) + exp(-4), xlim=c(0,12), ylim=c(0,5),
+             xlab=NA, ylab='Hazard rate',xaxt='n', col='red')
+       arrows(x0 = 2, y0 = 2.7, x1 = 0 ,y1 = 2.7)
+       text(3.1,2.7, 'alpha')
+       arrows(x0 = 4, y0 = 1.5, x1 = 1.5 ,y1 = 1)
+       text(4.5,1.7, 'beta')
+       arrows(x0 = 11, y0 = 0.8, x1 = 11 ,y1 = 0.018)
+       text(11,1, 'ho')
+       axis(1, at=0, labels= 0)
+
+       curve((1-exp(-(exp(1 - exp(-0.3) * x) + exp(-4))))*100, xlim=c(0,12), ylim=c(0,100),
+             xlab=NA, ylab='Pr(Mortality)', xaxt='n', col='red')
+       text(3.1,93, 'alpha')
+       arrows(x0 = 2, y0 = 93, x1 = 0 ,y1 = 93)
+       text(6,30, 'beta')
+       arrows(x0 = 5, y0 = 30, x1 = 3 ,y1 = 30)
+       text(11,15, 'ho')
+       arrows(x0 = 11, y0 = 10, x1 = 11 ,y1 = exp(-4)*100)
+       axis(1, at=0, labels= 0)
+       mtext(side=1, 'Growth rate',line = 2)
+```
+
+**Fig 6:** Interpreting model parameters.
+Top = Hazard curves (i.e.
+instantaneous mortality rates); Bottom = Mortality probability/year based on estimated hazard rates.
+The intercept at zero growth is defined by $alpha$.
+The slope is defined by $beta$.
+The asymptote is defined by $\gamma$ and represents the growth-independent hazard.
+
 ## References

### 2015-07-31 R/model.R
Rename variables in model
log_a2 should be a2

log_mu_a should be mu_log_a
The first implies the logged value of the mean, whereas really should be the mean of logged value.

@@ -4,71 +4,71 @@ get_model_chunks <- function(tasks) {
   if(nchar(rho_combo) > 0) {
     rho_combo <- sapply(seq_len(nchar(rho_combo)), function(i) substr(rho_combo, i, i))
   }
-  
+
   list(
-    pars = c("log_mu_a0","log_sigma_a0","log_a0","log_sigma_a1","log_a1",if("a" %in% rho_combo) "log_a2",
-             "log_mu_b0","log_sigma_b0","log_b0","log_sigma_b1","log_b1",if("b" %in% rho_combo) "log_b2",
-             "log_mu_c0","log_sigma_c0","log_c0","log_sigma_c1","log_c1",if("c" %in% rho_combo) "log_c2",
+    pars = c("mu_log_a0","sigma_log_a0","log_a0","sigma_log_a1","log_a1",if("a" %in% rho_combo) "a2",
+             "mu_log_b0","sigma_log_b0","log_b0","sigma_log_b1","log_b1",if("b" %in% rho_combo) "b2",
+             "mu_log_c0","sigma_log_c0","log_c0","sigma_log_c1","log_c1",if("c" %in% rho_combo) "c2",
              "sum_log_lik_fit","sum_log_lik_heldout"),
     parameters = sprintf("
       // Mortality model parameters
       real log_raw_a0[n_spp];
-      real log_mu_a0;
-      real<lower=0> log_sigma_a0;
-      
+      real mu_log_a0;
+      real<lower=0> sigma_log_a0;
+
       real log_raw_a1[n_census];
-      real<lower=0> log_sigma_a1;
-      
+      real<lower=0> sigma_log_a1;
+
       real log_raw_b0[n_spp];
-      real log_mu_b0;  
-      real<lower=0> log_sigma_b0;
-      
+      real mu_log_b0;
+      real<lower=0> sigma_log_b0;
+
       real log_raw_b1[n_census];
-      real<lower=0> log_sigma_b1;
-      
+      real<lower=0> sigma_log_b1;
+
       real log_raw_c0[n_spp];
-      real log_mu_c0;  
-      real<lower=0> log_sigma_c0;
-      
+      real mu_log_c0;
+      real<lower=0> sigma_log_c0;
+
       real log_raw_c1[n_census];
-      real<lower=0> log_sigma_c1;
-      
[…]

### 2015-08-17 ms.Rmd
Updates to priors and maths

@@
 author: "J S Camac"
-date: "30 Mar 2015"
+date: "17 August 2015"
 output: pdf_document
@@
 In plants, the carbon budget is considered a fundamental determinant of mortality, such that if rates of carbon assimilation are low or become insufficient for growth, an individual is more susceptible to dying [@Peet:1987va; @Hawkes:2000ib; @Keane:2001db; @McDowell:2011dr].
-Empirical studies examining intraspecific mortality rates have broadly supported this assumption, with many finding that the probability a plant dies decreases exponentially with increasing growth rates [@Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw].
+Empirical studies examining intraspecific mortality have broadly supported this assumption, with many finding that the probability a plant dies decreases exponentially with increasing growth rates [@Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw].
 Consequently, most dynamic vegetation models, including tree gap models [@Botkin:1972ii; @Keane:2001db], SORTIE (REF) and Dynamic Global Vegetation Models (DGVM's; @Woodward:2004ft) use the growth-mortality relationship as the predominant, and in some cases, only mechanism describing plant mortality [@Hawkes:2000ib; @McDowell:2011dr].
-Furthermore, the vast majority of these models assume the shape of this relationship is constant across species.
+However, despite empirical evidence highlighting that mortality rates vary considerably between species, many of  dynamic model assume the shape of the growth-mortality relationship is constant across species.
 
-This is despite empirical evidence highlighting that mortality rates vary considerably between species.
-One common finding from these multispecies studies that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species [@Poorter:2008iu; @Wright:2010fl].
+Multispecies studies have shown that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species [@Poorter:2008iu; @Wright:2010fl], and thus, species have different mortality rates.
 Ecologists have attempted to explain this variation between species by using functional traits - attributes of species that are thought to influence vital rates (i.e.
@@
 The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to stem breakage [@Putz:1983wu; @vanGelder:2006ex; @Chave:2009iy], wood herbivory (NEED REF), embolism [@Hacke:2001kj, @Jacobsen:2005fx, @Preston:2006jn] and fungal and pathogen attack [@Augspurger:1984wx].
-However, some research has also suggested that wood density may increase survival associated with growth-dependent mortality.
-For example, studies [e.g.
-@vanGelder:2006ex] have highlighted that species with high wood density are often more shade tolerant, and thus exhibit a lower probability of dying at low growth rates.
+However, research has also indicated that wood density may increase survival associated with growth-dependent mortality.
+For example, species with high wood density are often more shade tolerant, and thus exhibit a lower probability of dying at low growth rates [e.g.
+@vanGelder:2006ex].
 Studies have also shown that wood density may reduce drought induced carbon starvation [@McDowell:2011dr] and the susceptibility of ‘stressed’ individuals to opportunistic invertebrate, pathogen or fungal attack [@Augspurger:1984wx; @McDowell:2011dr].
 
-Most plant mortality research can be broadly seperated into two research focuses: 1) those that focus on estimating individual mortality within species; and 2) those that use suites of species to examine correlations between functional traits and population level estimates of tree mortality.
-However, there appears to be little cross over between these two research focuses.
-This is despite both approaches likely being required to develop a general model of tree mortality.
-A model that accounts for variation both within and between species.
+To date, most plant mortality research can be broadly seperated into two research focuses: 1) those that focus on estimating individual mortality within species [e.g.
+@Kobe:1997vy]; and 2) those that use suites of species to examine correlations between functional traits and population level estimates of plant mortality (e.g.
+Wright et al 2010, Chave et al 2004).
+However, little crossover of ideas exists between these two research focuses.
+This is despite both individual and species level variation likely being required to build a general model of plant mortality.
+Furthermore, in both research focuses, the measure of growth rate used to examine the growth-mortality relationship appears to be based on researcher preference with little consideration given as to whether an area or diameter based growth rate is a better predictor of plant mortality.
 
-In this study, we reconcile both approaches by fitting 15-years of growth and mortality data from 203 tropical rainforest species encompassing more than 180,000 individuals in a Bayesian hierarchical model that explicitly includes variability observed both within and between species.
-Specifically, our model extends the negative exponential functional form commonly observed between growth and mortality in intraspecific studies by adding a growth-independent baseline hazard and allowing each parameter to vary as a function of wood density.
-We then use this model to determine if wood density only influence growth-independent mortality (i.e.
-the baseline hazard/asymptote) or whether it can also influence growth-dependent paramters such as the intercept (i.e.
-the rate of mortality at zero growth) or the rate (i.e.
-the sensitivity of mortality rates to changes with growth rate)?
+In this study, we reconcile both research focuses by fitting 15-years of growth and mortality data from 203 tropical rainforest species encompassing more than 180,000 individuals in a Bayesian hierarchical model that explicitly accounts for variability observed both within and between species.
+Specifically, our model uses the negative exponential growth-mortalty relationship commonly observed and fitted in intraspecific studies and extends this by both adding a growth-independent baseline hazard, and by allowing each parameter to vary as a function of species level traits (in this case wood density).
 
-## Mortality model
-
-We extend a common functional form used to model the growth-mortality relationship [e.g. @Kobe:1997vy] to include a growth-independent baseline hazard, such that the functional form becomes
-\begin{equation} \label{eq:lambda}
-\lambda = \overbrace{\alpha \, \exp\left(-\beta \, G \right)}^{\text{Growth dependent}} + \overbrace{\gamma}^{\text{Growth independent}}
-\end{equation}
-
-By adding $\gamma$ we allow the functional form to asymptote at a mortality rate above zero, which allows for deaths to occur at high growth rates that would otherwise not be accounted for.
-The advantage of this functional form is that all parameters are biologically interpretable.
-$\alpha$ represents the intercept and biologically defines the sum of mortality rate at zero growth and baseline hazard.
-$\beta$ represents the function's decay, or in other words, the sensitivity of mortality rates to changes with growth rate.
-$\gamma$, as stated before, represents asymptote and is referred to as the baseline hazard.
-$\alpha$ and $\beta$ are intrinsically linked to the growth-mortality relationship, and as such are growth-dependent parameters.
-By contrast, $\gamma$ accounts for mortality not related to growth (e.g.
-windfall, fire) and consequently is a growth-independent parameter.
-
-While this model can be used to examine and predict mortality for individual species, it has limited capacity to provide general insights into the mechanisms affecting multiple species.
-We add this capacity by allowing $\alpha$, $\beta$ and $\gamma$ to vary by wood density - a species trait thought to convey a species growth strategy (REFs).
-Furthermore, we also allow the three parameters to vary by species ID and census period in which the data were collected so that the model could partition explained and unexplained variance between census, species and individual levels of the dataset [@Gelman:2007te; @Kery:2010tp].
-
-Using this model, coupled with k-fold cross validation we address four questions:
+We then use this model, coupled with k-fold cross validation to address four questions:
 
@@
 
-We analysed growth and survival data collected from a 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15&deg;N, 79.85&deg;W).
+We fit our model using growth and survival data collected from a 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15&deg;N, 79.85&deg;W).
 The plot is predominately undisturbed old-growth rainforest (48-ha), but 2 ha is of 100 year old secondary rainforest (REF).
@@
 Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100&deg;C) and dividing by the fresh volume (as measured by water displacement).
-Wood density was then calculated as an area-weighted average, where area refers to the annulus represented by each piece, assuming a circular trunk.
-These measurements were then converted to kg m$^-3$.
+Wood density was then calculated as an area-weighted average (g cm$^-3$), where area refers to the annulus represented by each piece, assuming a circular trunk.
 
-## Model description
+## Measurement error and calculating true growth
 
-We have binary observations $y_i$ (0 = alive, 1 = dead) for each individual $i$, and a hypothesised the underlying mortality model giving an instantaneous rate of mortality (or hazard) per unit time (eq.
+Observed dbh measurements are likely to include measurement error that can manifest into biologically unrealistic scenarios of negative growth and ultimately erroneous conclusions.
+We estimate this measurement error using another BCI dataset consisting of 1562 randomly selected trees remeasured within 30 days (see XXXX).
+Assuming that no growth occurred between observations, we estimated the standard deviation of measurement error (and its uncertainty) by sampling the discrepencies for each individual, $i$, as random realisations from a normal distribution centered on zero with a standard deviation, $\sigma_{error}$, which describes the variation in error:
+
+\begin{equation} \label{eq:error_model}
+discepency_i \sim \textrm{N}(0, \sigma_{error}). \end{equation}
+
+$\sigma_{error}$ was estimated to be 0.75 (sd = 0.01), meaning that 95% of the measurement errors were within +/- 1.47 cm of the true diameter.
+We then used this standard deviation of 0.75 in another Bayesian model to estimate the initial and final true dbh (i.e.
+observation - measurement error) for each individual.
+This was done by first modelling the observed initial ($dbh_{t-1}$) and final dbh ($dbh_{t}$) of each individual, $i$, as random realisations from a normal distribution centered on their respective true estimates (i.e.
+$\widehat{dbh1_i}$ and $\widehat{dbh2_i}$), with a standard deviation of 0.75 (i.e.
+$\sigma_{error}$).
+
+\begin{equation} \label{eq:obs dbh1}
+dbh_{i,t-1} \sim \textrm{N}(\widehat{dbh_{i,t-1}}, 0.75). \end{equation}
+
+\begin{equation} \label{eq:obs dbh2}
+dbh_{i,t} \sim \textrm{N}(\widehat{dbh_{i,t}}, 0.75). \end{equation}
+
+$\widehat{dbh_{i,t-1}}$ was estimated from a lognormal distribution:
+
+\begin{equation} \label{eq:true dbh1}
+\widehat{dbh_{i,t-1}} \sim \textrm{lognormal}(\mu_{dbh_{i,t-1}}, \sigma_{dbh_{t-1}}), \end{equation}
+
+where, $\mu_{dbh_{t-1}}$ defines the log of the grand mean initial dbh, and $\sigma_{dbh_{t-1}}$ defines that variation around that mean.
+$\mu_{dbh_{t-1}}$ was estimated using a normal prior $\textrm{N}(0, 2.5)$, which represents our prior expectation that the mean initial dbh would likely occur between 0 and 148 cm.
+$\sigma_{dbh{t-1}}$ was estimated using a weakly informative half cauchy prior centered on zero and a scale parameter of 2.5.
+
+$\widehat{dbh_{i,t}}$ was estimated as:
+
+\begin{equation} \label{eq:true dbh2}
+\widehat{dbh_{i,t-1}} + R_i \, T_i, \end{equation}
+
+where, $R_i$ represents the true annual increase in dbh (cm) and $T_i$ represents the number of days between measurements divided by 365.25.
+We modelled $R_i$ as a random realisation from a lognormal distribution centered on zero and a standard deviation of 1, conveying our prior expectation that annual increases in dbh would be between 0.13 and 7.4 cm/yr.
+We used a lognormal as it ensures dbh growth cannot be negative.
+
+We calculated basal area growth rate using: $\frac{(\pi \, 0.25 \,\widehat{dbh_{i,t}}) - (\pi \, 0.25 \, \widehat{dbh_{i,t-1}})}{T_i}$.
+
+We then calculated the mean posterior growth rate estimate for each individual and used these as our estimates of true dbh growth for the subsequent mortality analysis.
+
+## Mortality model description
+
+We extend a common functional form used to model the growth-mortality relationship [e.g. @Kobe:1997vy] to include a growth-independent baseline hazard, such that the functional form becomes
+\begin{equation} \label{eq:lambda}
+\lambda = \overbrace{\alpha \, \exp\left(-\beta \, G \right)}^{\text{Growth dependent}} + \overbrace{\gamma}^{\text{Growth independent}}
+\end{equation}
+
+By adding $\gamma$ we allow the functional form to asymptote at a mortality rate above zero, which allows for deaths to occur at high growth rates that would otherwise not be accounted for.
+The advantage of this functional form is that all parameters are biologically interpretable.
+$\alpha$ represents the intercept and biologically defines the sum of mortality rate at zero growth and baseline hazard.
+$\beta$ represents the function's decay, or in other words, the sensitivity of mortality rates to change with growth rate.
+$\gamma$, as stated before, represents asymptote and is referred to as the baseline hazard.
+$\alpha$ and $\beta$ are intrinsically linked to growth rate, and as such are growth-dependent parameters.
+By contrast, $\gamma$ accounts for mortality not related to growth caused by stachastic events (e.g.
+windfall, fire) and consequently is a growth-independent parameter.
+
+To incorporate varability in species mortality rates, we add allowed $\alpha$, $\beta$ and $\gamma$ to vary by wood density - a species trait thought to encapsulate a species growth strategy (REFs).
+In addition to this, we also allowed the three parameters to vary by species ID and census period, by doing this, we effectively allow the model to partition variation explained by wood density from the residual variation explained by species, census and individual levels [@Gelman:2007te; @Kery:2010tp].
+
+## Bayesian Hierarchical Model implementation
+
+We have binary observations $y_i$ (0 = alive, 1 = dead) for each individual $i$.
+We also have a hypothesised underlying mortality model giving an instantaneous rate of mortality (or hazard) per unit time (eq.
 \ref{eq:lambda}).
@@
 where $\Lambda_{i,t} = \int_{t1}^{t2} \lambda(t) \, dt$ is the cumulative hazard for that individual.
-Given suitable estimates of $\lambda$, one can model the observations as a random realization from a Bernoulli distribution based on the probability that individual $i$ would die, $p_{i,t_1\rightarrow t_2}$:
+Given estimates of $\lambda$, one can model the observations as a random realization from a Bernoulli distribution based on the probability that individual $i$ would die, $p_{i,t_1\rightarrow t_2}$:
 \begin{equation} \label{eq:obs_model}
@@
 
-Our first hypothesis is that the cumulative hazard rate within the focal census interval is reasonably approximated by the growth rate in the previous interval
+Our first hypothesis is that the cumulative hazard rate within a census interval is reasonably approximated by the true growth rate (estimated as above) in the previous interval
 \begin{equation} \label{eq:hazard2}
@@
 \end{equation}
-Growth rates are estimated .... (see below).
 
-Second, we assume that the key parameters of eq.
-\ref{eq:lambda} are influenced by 3 factors: census year, species identify, and trait.
-Because traits are measured at species-level, the trait based effect is logically a subset of species level effect.
+Second, we assume that the three parameters of eq.
+\ref{eq:lambda} are influenced by 3 factors: wood density, species ID and census year.
+Because wood density is measured at species-level, its effect is logically a subset of the overall species level effect.
+
 To account for species strategy and partition variance into explained and unexplained at the observation, species and sampling period level we modelled $\alpha$, $\beta$ and $\gamma$ via an equation of the form:
 \begin{equation} \label{eq:alpha}
-\alpha_s = \alpha_{1,t} \, \alpha_{0,s} \, \rho_s^{\alpha_{2}}.
+\alpha_s = \alpha_{0,s} \, \alpha_{1,t} \, \rho_s^{\alpha_{2}}.
 \end{equation}
-Here $\alpha_{1,t}$ represents the sample-level random effect for which time $t$ an individual was sampled, $\alpha_{0,s}$ represents the species-level random effect of species $s$ within which individual $i$ belongs, and $\alpha_{2}$ represents the effect of standardised wood density $\rho_s$ on $\alpha_s$.
+Here $\alpha_{0,s}$ represents the species-level random effect of species $s$ within which individual $i$ belongs, $\alpha_{1,t}$ represents the sample-level random effect for which time $t$ an individual was sampled, and $\alpha_{2}$ represents the effect of standardised wood density $\rho_s$ on $\alpha_s$.
 The form of eq \ref{eq:alpha} ensures that parameters remain positive.
 On a log scale equation \ref{eq:alpha} becomes an additive model,$\log \alpha_s = \log \alpha_{0,s} + \log \alpha_{1,t} + \alpha_{2} \, \log \rho_s$, equivalent to a classic proportional hazards model widely used in survival analysis.
-We standardised wood density via the transformation: $\rho_s = \rho/600$, an operation which is equivalent to centring a log-transformed variable at a density of 600 kg m$^-3$.
-This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 600 (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant) [@Gelman:2007te].
+We standardised wood density via the transformation: $\rho_s = \rho/0.6$, an operation which is equivalent to centering a log-transformed variable at a density of 0.6 g cm${^-3}$.
+This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant) [@Gelman:2007te].
 
 The species and census-level random effects $\alpha_{0,s}, \alpha_{1,t}$ were assumed to follow a log-normal distribution:
-\begin{equation} \label{eq:alpha0}
-\log \alpha_i \sim \mathcal{N}\left(\mu_{\log \alpha_i}, \sigma_{\log \alpha_i}\right),\\
+\begin{equation} \label{eq:alpha0_alpha1}
+\alpha_i \sim \text{log-normal}\left(\mu_{\log \alpha_i}, \sigma_{\log \alpha_i}\right),\\
 \end{equation}
 where $\mu_{\log \alpha_i},  \sigma_{\log \alpha_i}$ are the mean and standard deviation of log-transformed variables.
+We estimated $\mu_{\log \alpha_{0,s}}$, the average species effect at zero growth using the following informative prior:
 
-Similar forms used for $\beta$ and $\gamma$.
+\begin{equation} \label{eq:alpha0}
+\mu_{\log \alpha_{0,s}} \sim \text{normal}\left(-0.87, 0.75\right).\\
+\end{equation}
 
-## Model fitting
+This distribution was derived from data presented by Kobe et al 1995, who estimated the probability of dying at zero growth/2.5 years for juveniles of 10 hardwood species across north-eastern America.
+Note that these data are also used to define the growth-mortality relationship in SORTIE-ND a spatially explicit forest dynamic model (REF).
+We converted these probabilities to instantaneous hazard rates by using $\lambda = -\log(1 - Pr(Death))/2.5$, and then log transforming these values and calculating the mean (i.e.
+-0.87) associated standard error (i.e.
+0.75).
+Based on this prior, we assumed that the average probability of death for an average species was 34%/yr with 95% of species effects occuring between 9% and 84%.
 
-We used Bayesian inference and fit models to data using Markov chain Monte Carlo (MCMC) sampling in R 3.1.3 using package rstan version 2.6 [@StanDevelopmentTeam:2015uz].
+EXAMPLE:
+```{r}
+# mean species instantaneous hazard rate
+M <- c(0.99,0.998,0.555,0.014,0.077,0.999,0.268,0.998,0.985)
+k = - log(1-M)/2.5
+mean(log(k))
+
+# standard error
+sd(log(k))/sqrt(9)
+```
+
+Similarly, we used prior information from Korning and Balslev et al (1994) to estimate the average species effect on $\gamma$, the growth-independent hazard rate, such that:
+
+\begin{equation} \label{eq:gamma0}
+\mu_{\log \gamma_{0,s}} \sim \text{normal}\left(-3.86, 0.2\right).\\
+\end{equation}
+
+This distribution was estimated the same way as in $\mu_{\log \alpha_{0,s}}$.
+That is we converted annual mortality probabilities of 17 tropical lowland tree species in Amazonian Ecuador that encompass the range of forest structure (i.e.
+understory species, small stemmed midcanopy species, large stemmed midcanopy species and canopy species) to instantaneous hazard rate, and then log transformed these and calculated the mean (i.e.
+-3.86) and standard error (i.e.
+0.2) These annual mortality estimates were averaged over size and growth rates, and as such are likely to be an overestimate of the true value of $\gamma$.
+However, based on this prior, we assume that the average growth-independent species mortality will be 2.1%/yr with 95% of species growth-independent probabilities of occuring between 1.4% and 3.1%.
+
+EXAMPLE:
+```{r}
+M <- c(8.6,1.8,0.6,3.8,0.9,0.7,4.3,4.7,1.8,6.4,0.8,1,2,1.8,3.7,2.2,2.2)/100 # From Korning Baslev 1994
+k = - log(1-M)/1
+mean(log(k))
+
+# standard error
+sd(log(k))/sqrt(length(M)) 
+```
+
+As we are fitting our model to two growth measures, we used weak priors to define the common log-normal distribution for the species effects on $\beta$:
+
+\begin{equation} \label{eq:beta_0,s}
+\mu_{\log \beta_{0,s}} \sim \text{lognormal}\left(0, 5\right).\\
+\end{equation}
+
+This prior, regardless of growth measure, spans a range such that the decay of the negative exponential can be rapid or is flat, suggesting no effect of growth rate on mortality rates.
+
+The census random effects, $\alpha_1$, $\beta_1$ and $\gamma_1$ were all sampled from lognormal distributions centered on zero with a standard deviations estimated from the data. For example:
+
+\begin{equation} \label{eq:a1b1c1,s}
+\alpha_{1,t} \sim \text{lognormal}\left(0, \sigma_{a1}\right).\\
+\end{equation}
+
+The effect of wood density on all three parameters (i.e. $\alpha_2$, $\beta_2$ and $\gamma_2$) were given normal priors centered on zero and a standard deviation of 5:
+
+\begin{equation} \label{eq:a2b2c2}
+\alpha_2, \beta_2, \gamma_2 \sim \text{normal}\left(0, 5\right).\\
+\end{equation}
+
+Standard deviations associated with both species random effects (i.e.
+$\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$) and those associated with census random effects (i.e.
+$\sigma_{a1}$, $\sigma_{b1}$, $\sigma_{c1}$), were all sampled from half cauchy priors with centered on zero with a scale of 2.5.
+
+\begin{equation} \label{eq:a1b1c1,s}
+\sigma_{a0},\sigma_{b0},\sigma_{c0},\sigma_{a1},\sigma_{b1},\sigma_{c1} \sim \text{cauchy+}\left(0, 2.5\right).\\
+\end{equation}
+
+We fitted these models using package rstan version 2.7 [@StanDevelopmentTeam:2015uz].
 Three independent chains were executed and in all cases modelled parameters converged within 1000 iterations.
@@
 
+\pagebreak
+
+## STAN MODEL
+```{r, eval=FALSE}
+data {
+  int<lower=1> n_obs;
+  int<lower=1> n_spp;
+  int<lower=1> spp[n_obs];
+  int<lower=1> n_census;
+  int<lower=1> census[n_obs];
+  int<lower=0, upper=1> y[n_obs];
+  vector[n_obs] census_length;
+  vector[n_obs] growth_dt;
+  vector[n_spp] rho_c;
+
+  // Held out data
+  int<lower=1> n_obs_heldout;
+  int<lower=1> n_spp_heldout;
+  int<lower=1> spp_heldout[n_obs_heldout];
+  int<lower=1> census_heldout[n_obs_heldout];
+  int<lower=0, upper=1> y_heldout[n_obs_heldout];
+  vector[n_obs_heldout] census_length_heldout;
+  vector[n_obs_heldout] growth_dt_heldout;
+  vector[n_spp_heldout] rho_c_heldout;
+}
+
+parameters {
+
+  // Mortality model parameters
+  real raw_log_a0[n_spp];
+  real mu_log_a0;
+  real<lower=0> sigma_log_a0;
+
+  real raw_log_a1[n_census];
+  real<lower=0> sigma_log_a1;
+
+  real raw_log_b0[n_spp];
+  real mu_log_b0;
+  real<lower=0> sigma_log_b0;
+
+  real raw_log_b1[n_census];
+  real<lower=0> sigma_log_b1;
+
+  real raw_log_c0[n_spp];
+  real mu_log_c0;
+  real<lower=0> sigma_log_c0;
+
+  real raw_log_c1[n_census];
+  real<lower=0> sigma_log_c1;
+
+  real a2;
+  real b2;
+  real c2;
+}
+
+model {
+
+  // Declaring mortality parameters
+  real alpha;
+  real a0[n_spp];
+  real a1[n_census];
+
+  real beta;
+  real b0[n_spp];
+  real b1[n_census];
+
+  real gamma;
+  real c0[n_spp];
+  real c1[n_census];
+
+  real cumulative_hazard;
+
+  // Calculating species random effects
+  for (s in 1:n_spp) {
+    a0[s] <- exp(raw_log_a0[s] * sigma_log_a0 + mu_log_a0); // e.g. implies lognormal(mu_log_a0, sigma_log_a0)
+    b0[s] <- exp(raw_log_b0[s] * sigma_log_b0 + mu_log_b0);
+    c0[s] <- exp(raw_log_c0[s] * sigma_log_c0 + mu_log_c0);
+  }
+
+  // Calculating census period random effects
+  for (t in 1:n_census) {
+    a1[t] <- exp(raw_log_a1[t] * sigma_log_a1); // e.g. implies lognormal(0, sigma_log_a1)
+    b1[t] <- exp(raw_log_b1[t] * sigma_log_b1);
+    c1[t] <- exp(raw_log_c1[t] * sigma_log_c1);
+  }
+
+  for (i in 1:n_obs) {
+    // Calculating mortality parameters
+    alpha <- a0[spp[i]] * a1[census[i]] * pow(rho_c[spp[i]], a2);
+    beta <- b0[spp[i]] * b1[census[i]] * pow(rho_c[spp[i]], b2);
+    gamma <- c0[spp[i]] * c1[census[i]] * pow(rho_c[spp[i]], c2);
+
+    // Likelihood for hazard model
+    cumulative_hazard <- -census_length[i] * (alpha * exp(-beta * growth_dt[i]) + gamma);
+
+    if (y[i] == 0) {
+      increment_log_prob(cumulative_hazard);
+    } else {
+      increment_log_prob(log1m_exp(cumulative_hazard));
+    }
+  }
+
+  // Priors
+
+  //Mortality model priors
+  raw_log_a0 ~ normal(0,1);
+  mu_log_a0 ~ normal(-0.87, 0.75);
+  sigma_log_a0 ~ cauchy(0, 2.5);
+
+  raw_log_b0 ~ normal(0, 1);
+  mu_log_b0 ~ normal(0, 5);
+  sigma_log_b0 ~ cauchy(0, 2.5);
+
+  raw_log_c0 ~ normal(0, 1);
+  mu_log_c0 ~ normal(-3.86, 0.2);
+  sigma_log_c0 ~ cauchy(0, 2.5);
+
+  raw_log_a1 ~ normal(0,1);
+  sigma_log_a1 ~ cauchy(0, 2.5);
+
+  raw_log_b1 ~ normal(0, 1);
+  sigma_log_b1 ~ cauchy(0, 2.5);
+
+  raw_log_c1 ~ normal(0, 1);
+  sigma_log_c1 ~ cauchy(0, 2.5);
+
+  a2 ~ normal(0,5);
+  b2 ~ normal(0,5);
+  c2 ~ normal(0,5);
+}
+```
+
+\pagebreak
+
+# OTHER STUFF
 To optimise calculations in stan, we tried to minimise the number of logarithmic and exponential transformations, resulting in the following algorithm:
@@
 
-### Measurement error and calculating true growth
-
-Observed dbh measurements are likely to include measurement error that can manifest into biologically unrealistic scenarios of negative growth and ultimately erroneous conclusions.
-We estimate this measurement error using another BCI dataset consisting of 1562 randomly selected trees remeasured within 30 days (see XXXX).
-Assuming that no growth occurred between observations, we estimated the standard deviation of measurement error (and its uncertainty) by sampling the discrepencies for each individual, $i$, as random realisations from a normal distribution centered on zero with a standard deviation, $\sigma_{error}$, which describes the variation in error:
-
-\begin{equation} \label{eq:error_model}
-discepency_i \sim \textrm{N}(0, \sigma_{error}). \end{equation}
-
-$\sigma_{error}$ was estimated to be 0.75 (sd = 0.01), meaning that 95% of the measurement errors were within +/- 1.47 cm of the true diameter.
-We then used this standard deviation of 0.75 in another Bayesian model to estimate the initial and final true dbh (i.e.
-observation - measurement error) for each individual.
-This was done by first modelling the observed initial (dbh1) and final dbh (dbh2) of each individual, $i$, as a random realisation from a normal distribution centered on their respective true estimates (i.e.
-$\widehat{dbh1_i}$ and $\widehat{dbh2_i}$), with a standard deviation of 0.75 (i.e.
-$\sigma_{error}$).
-
-\begin{equation} \label{eq:obs dbh1}
-dbh1_i \sim \textrm{N}(\widehat{dbh1_i}, 0.75). \end{equation}
-
-\begin{equation} \label{eq:obs dbh2}
-dbh2_i \sim \textrm{N}(\widehat{dbh2_i}, 0.75). \end{equation}
-
-$\widehat{dbh1_i}$ was estimated from a lognormal distribution:
-
-\begin{equation} \label{eq:true dbh1}
-\hat{dbh1_i} \sim \textrm{lognormal}(\mu_{dbh1}, \sigma_{dbh1}), \end{equation}
-
-where, $\mu_{dbh1}$ defines the log of the grand mean initial dbh, and $\sigma_{dbh1}$ defines that variation around that mean.
-$\mu_{dbh1}$ was estimated using a normal prior $\textrm{N}(0, 2.5)$, which represents our prior expectation that the mean initial dbh would likely occur between 0 and 148 cm.
-$\sigma_{dbh1}$ was estimated using a weakly informative half cauchy prior centered on zero and a scale parameter of 2.5.
-
-$\widehat{dbh2_i}$ was estimated as:
-
-\begin{equation} \label{eq:true dbh2}
-\widehat{dbh2_i} + R_i \, T_i, \end{equation}
-
-where, $R_i$ represents the true annual increase in dbh (cm) and $T_i$ represents the number of days between measurements divided by 365.25.
-We modelled $R_i$ as a random realisation from a lognormal distribution centered on zero and a standard deviation of 1, conveying our prior expectation that annual increases in dbh would be between 0.13 and 7.4 cm/yr.
-We used a lognormal as it ensured dbh growth cannot be negative.
-We then used the mean posterior estimate of $R_i$ as our estimate of true dbh growth for the subsequent mortality analysis.
-
-### Priors (NEEDS UPDATING)
-
-Because we used Bayesian inference we needed to specify prior distributions for all model parameters.
-The effects of wood density on all three parameters ($\alpha_{2}$, $\beta_{2}$ and $\gamma_{2}$), were sampled from weakly informative normal priors with a mean of 0 and a standard deviation of 5.
-
-census random effects: we assumed little variation so centred on zero with a standard deviation of 0.1
-alpha species mu - we extracted from Kobe et al 1995 and are the parameter estimates currently used in SORTIE
-gamma species mu - we extracted from chambers 2013
-beta species mu - we had no idea about, especially as this parameter is likely to change with different growth rates. As such we fitted a uniform prior bounded between -20 and 20.
-All sigma's were drawn from a half cauchy distribution with a mean of zero and a standard deviation of 2.5.
-
-and mean hyperparamters ($\mu_{\alpha_0}$, $\beta_{mu}$ and $\gamma_{\mu}$) we used stan's default uninformative prior, $\textrm{Uniform(-2,2)}.
-For hyperparamater standard deviations we used $\textrm{Uniform(0,2)}$.
-
-## To do
-
-Change so that baseline effect is census year, not species
-
 ## Analysis plan
 
-Step 1 - define a general structure of model we think will capture data, on basis of prior knowledge
-Step 2 - assuming species effects on all pars, ask which growth meausre best predicts data, compare 4 growth rates using cross validation
+Step 1 - define a general structure of model we think will capture data, on basis of prior knowledge [DONE]
+Step 2 - assuming species effects on all pars, ask which growth meausre best predicts data, compare 2 growth rates using cross validation
 Step 3 - fit full model with trait effects, look at effect sizes of WD on a, b and c
 Step 4 - based on 3, choose some subsets of models, compare these using both bayes factor (most parsimonious) and predictive capacity
-
-\pagebreak
-
-## Results
-**Take home message:**
-Wood density decreases intercept (log_a) & further increases the steepness of the growth-mortality slope (log_b). It also decreases the growth-independent hazard log_ho
-\pagebreak
-
-## Notes/Questions
-
-1 - How much data should we fit to this model?
-All of it would require the inclusion of a individual random effect to account for non-independence of repeatedly sampled individuals.
-We could subsample individuals in entire dataset...
-but again this would require non-independence to be account for.
-The advantage of this approach is we can also cross validate.
-Another way to cross validate is to run the model for each census on a subset of individuals spanning all species.
-Question is, how can we best stratify sampling to ensure low abundance species are also represented.
-
-2 - Growth is a predominant predictor in mortality models, however, it is not the only one.
-Size is also commonly used as a predictor of mortality.
-A typical finding from these size-mortality models is that mortality appears to be U-shaped.
-Could we account for this by fitting dbh and dbh \, wd interaction on the growth independent hazard (ho)?
-My concern is whether this makes sense as dbh and growth are highly correlated (~50%) consequently, these two predictors may not be identifiable.
-
-3 - Possibly not a major concern with size not in the model, but currently the data only utilises data from individuals that are at least 1.3 m high and have a diameter of > 1 cm.
-This means that we are missing alot of the mortality at the young age.
-Is it worth investing time in cleaning and preparing the BCI seedling data for inclusion?
-A difficulty to face is that the seedling data measures height while the main BCI dataset measures dbh.
-Maybe some type of allometric model using BCI?
-
-4 - Current model form at very low growth rate forces the hazard for high wood density species to become larger than low wood density species.
-Though this effect is not significant (95% credible intervals overlap)...
-it appears to be the result of the model form we are using.
-Discuss with Daniel.
-
-5 - Set model up to run on cluster.
-I've managed to dramatically speed up convergence time with these models.
-Approximately 5 hours.
-However, if we plan to run models containing multiple combinations of 4 growth rate types \, 3 parameter combinations.
-We will need to set this up on the cluster.
-
-6 - Need to integrate code into remaker
 
@@
 
+## Results
+**Prelim results:**
+
+**DBH_DT**
+Wood density effects on alpha, beta and gamma:
+*alpha* - no significant effect
+*beta* - positive effect (increases decay rate)
+*gamma* - negative effect (decreases hazard rate)
+
+**BASAL_DT**
+Wood density effects on alpha, beta and gamma:
+*alpha* - no significant effect
+*beta* - positive effect (increases decay rate)
+*gamma* - marginal negative effect (not significant @ p <0.05)
+
+**Best growth measure**
+Preliminary it appears to be dbh_dt as it has a higher log likelihood.
+
+\pagebreak
+
 # Figures
@@
        arrows(x0 = 11, y0 = 0.8, x1 = 11 ,y1 = 0.018)
-       text(11,1, 'ho')
+       text(11,1, 'gamma')
        axis(1, at=0, labels= 0)
@@
        curve((1-exp(-(exp(1 - exp(-0.3) * x) + exp(-4))))*100, xlim=c(0,12), ylim=c(0,100),
-             xlab=NA, ylab='Pr(Mortality)', xaxt='n', col='red')
+             xlab=NA, ylab='Pr(Mortality)/yr', xaxt='n', col='red')
        text(3.1,93, 'alpha')
@@
        arrows(x0 = 5, y0 = 30, x1 = 3 ,y1 = 30)
-       text(11,15, 'ho')
+       text(11,15, 'gamma')
        arrows(x0 = 11, y0 = 10, x1 = 11 ,y1 = exp(-4)*100)

### 2015-08-25 R/stan_functions.R
Removed data subsetting

@@ -34,7 +34,7 @@ run_single_stan_chain <- function(model, data, chain_id, iter=1000,
        iter = iter,
        chains=1, 
        chain_id=chain_id,
-       control =list(stepsize=0.01, adapt_delta=0.9, max_treedepth=15),
+       control =list(stepsize=0.01, max_treedepth=15),
        refresh=1,
        sample_file=sample_file,
        diagnostic_file=diagnostic_file)
@@ -42,8 +42,6 @@ run_single_stan_chain <- function(model, data, chain_id, iter=1000,
 
 prep_data_for_stan <- function(data, growth_measure) {
   full_data <- readRDS('export/bci_data_full.rds')
-  data$train <- data$train[1:50000,]
-  #data$heldout <- data$heldout[1:10000,]
   scale <- median(full_data$train[[growth_measure]])
   
   list(

### 2015-08-25 ms.Rmd
Updated priors

@@
 \begin{equation} \label{eq:alpha0}
-\mu_{\log \alpha_{0,s}} \sim \text{normal}\left(-0.87, 0.75\right).\\
+\mu_{\log \alpha_{0,s}} \sim \text{normal}\left(-0.61, 0.57\right).\\
 \end{equation}
@@
 We converted these probabilities to instantaneous hazard rates by using $\lambda = -\log(1 - Pr(Death))/2.5$, and then log transforming these values and calculating the mean (i.e.
--0.87) associated standard error (i.e.
-0.75).
-Based on this prior, we assumed that the average probability of death for an average species was 34%/yr with 95% of species effects occuring between 9% and 84%.
+-0.61) associated standard error (i.e.
+0.).
+Based on this prior, we assumed that the average probability of death for an average species was 42%/yr with 95% of species effects occuring between 16% and 81%.
 
@@
 ```{r}
-# mean species instantaneous hazard rate
-M <- c(0.99,0.998,0.555,0.014,0.077,0.999,0.268,0.998,0.985)
+# white ash
+median(c(0.999,0.55,0.998))
+# sugar maple
+median(c(0.998, 0.148, 0.999)
+# American beech
+median(c(0.014,0.12))
+# Yellow Birch
+median(c(0.555,0.995))
+# Red Maple
+0.99
+# Eastern hemlock
+0.077
+# Black cherry
+0.998
+# White oak
+0.82
+# Red oak
+0.985
+#White pine
+0.268
+
+# mean species instantaneous hazard rate (used medians for species with multiple measurements)
+M <- c(0.998,0.998,0.067,0.775,0.99,0.077,0.998,0.82,0.985, 0.268)
 k = - log(1-M)/2.5
@@
 # standard error
-sd(log(k))/sqrt(9)
+sd(log(k))/sqrt(10)
 ```
@@
   raw_log_c0 ~ normal(0, 1);
-  mu_log_c0 ~ normal(-3.86, 0.2);
+  mu_log_c0 ~ normal(-4.43, 1);
   sigma_log_c0 ~ cauchy(0, 2.5);
@@
   c2 ~ normal(0,5);
+}
+
+generated quantities {
+
+  // Declaring fitted parameters
+  real a0[n_spp];
+  real a1[n_census];
+
+  real b0[n_spp];
+  real b1[n_census];
+
+  real c0[n_spp];
+  real c1[n_census];
+
+  real alpha_fit;
+  real beta_fit;
+  real gamma_fit;
+  real cumulative_hazard_fit;
+  real log_lik_fit;
+  real sum_log_lik_fit;
+
+  // Declaring heldout parameters
+  real alpha_heldout;
+  real beta_heldout;
+  real gamma_heldout;
+  real cumulative_hazard_heldout;
+  real log_lik_heldout[n_obs_heldout];
+  real sum_log_lik_heldout;
+
+  // Initialization of summed log likelihoods
+  sum_log_lik_fit <- 0;
+  sum_log_lik_heldout <- 0;
+
+  // recalulate species random effects
+  for (s in 1:n_spp) {
+    a0[s] <- exp(raw_log_a0[s] * sigma_log_a0 + mu_log_a0);
+    b0[s] <- exp(raw_log_b0[s] * sigma_log_b0 + mu_log_b0);
+    c0[s] <- exp(raw_log_c0[s] * sigma_log_c0 + mu_log_c0);
+  }
+
+  // recalulate species random effects
+  for (t in 1:n_census) {
+    a1[t] <- exp(raw_log_a1[t] * sigma_log_a1);
+    b1[t] <- exp(raw_log_b1[t] * sigma_log_b1);
+    c1[t] <- exp(raw_log_c1[t] * sigma_log_c1);
+  }
+
+  // log likelihood for fitted model
+  for (i in 1:n_obs) {
+    alpha_fit <- a0[spp[i]] * a1[census[i]] * pow(rho_c[spp[i]], a2);
+    beta_fit <- b0[spp[i]] * b1[census[i]] * pow(rho_c[spp[i]], b2);
+    gamma_fit <- c0[spp[i]] * c1[census[i]] * pow(rho_c[spp[i]], c2);
+
+    cumulative_hazard_fit <- -census_length[i] * (alpha_fit * exp(-beta_fit * growth_dt[i]) + gamma_fit);
+
+    if (y[i] == 0) {
+      log_lik_fit <- cumulative_hazard_fit;
+    }
+    else {
+      log_lik_fit <- log1m_exp(cumulative_hazard_fit);
+    }
+    sum_log_lik_fit <- sum_log_lik_fit + log_lik_fit;
+  }
+
+  // log likelihood for held out data
+  for (j in 1:n_obs_heldout) {
+    alpha_heldout <- a0[spp_heldout[j]] * a1[census_heldout[j]] * pow(rho_c_heldout[spp_heldout[j]], a2);
+    beta_heldout <- b0[spp_heldout[j]] * b1[census_heldout[j]] * pow(rho_c_heldout[spp_heldout[j]], b2);
+    gamma_heldout <- c0[spp_heldout[j]] * c1[census_heldout[j]] * pow(rho_c_heldout[spp_heldout[j]], c2);
+
+    cumulative_hazard_heldout <- -census_length_heldout[j] * (alpha_heldout * exp(-beta_heldout * growth_dt_heldout[j]) + gamma_heldout);
+
+    if (y_heldout[j] == 0) {
+      log_lik_heldout[j] <- cumulative_hazard_heldout;
+    }
+    else {
+      log_lik_heldout[j] <- log1m_exp(cumulative_hazard_heldout);
+    }
+    sum_log_lik_heldout <- sum_log_lik_heldout + log_lik_heldout[j];
+  }
 }

### 2015-08-25 R/model_no_spp_err.R
Obsolete

@@ -1,143 +1,0 @@
-get_chunks_model1_no_spp_err <- function() {
-  list(
-    pars = c("c1_raw","c1_mu","c1_sigma","c2"),
-    parameters = "
-    real c1_raw[n_census];
-    real c1_mu;  
-    real<lower=0> c1_sigma;
-    real c2;",
-    
-    transformed_parameters = "
-    real c1[n_census];
-    real c_log[n_obs];
-    
-    for (t in 1:n_census) {
-    c1[t] <- c1_raw[t] * c1_sigma + c1_mu;
-    }
-    
-    for (i in 1:n_obs) {
-    c_log[i] <- c1[census[i]] + c2 * log_rho_cs[spp[i]];
-    p[i] <- inv_cloglog(log(census_length[i] * (exp(c_log[i]))));
-    }",
-    model = "
-  for (t in 1:n_census) {
-  c1_raw[t] ~ normal(0,1);
-  }
-  ",
-    r_model = function(stan_data, pars) {
-      for (i in 1:stan_data$n_obs) {
-        c_log[i] <- pars$c1_mu + pars$c2 * stan_data$log_rho_cs[stan_data$spp[i]]
-      }
-      inv_cloglog(log(stan_data$census_length * (exp(c_log))))
-    }
-  )
-}
-
-get_chunks_model2_no_spp_err <- function() {
-  list(
-    pars = c("a1_raw","a1_mu","a1_sigma","a2",
-             "b1_raw","b1_mu","b1_sigma","b2"),
-    parameters = "
-    real a1_raw[n_census];
-    real a1_mu;
-    real<lower=0> a1_sigma;
-  
-    real b1_raw[n_census];
-    real b1_mu;  
-    real<lower=0> b1_sigma;
-  
-    real a2;
-    real b2;",
-    transformed_parameters = "
-    real a1[n_census];
-    real a_log[n_obs];
-    
-    real b1[n_census];
-    real b_log[n_obs];
-    
-    for (t in 1:n_census) {
-    a1[t] <- a1_raw[t] * a1_sigma + a1_mu;
-    b1[t] <- b1_raw[t] * b1_sigma + b1_mu;
-    }
-    
-    for (i in 1:n_obs) {
-    a_log[i] <- a1[census[i]] + a2 * log_rho_cs[spp[i]];
-    b_log[i] <- b1[census[i]] + b2 * log_rho_cs[spp[i]];
[…]

### 2015-08-25 run_test.R
Obsolete

@@ -1,38 +1,0 @@
-# Download data for true dbh growth and save in relevant spot
-
-# remake::make("export/bci_data.rds")
-
-packages <- c("plyr","parallel", "rstan", "remake")
-sources <- c("R/dbh_error_model.r",
-             "R/true_dbh_model.r",
-             "R/model.R",
-             "R/task_compiler.R",
-             "R/stan_functions.R",
-             "R/utils.R")
-
-for (p in packages) {
-  library(p, character.only=TRUE, quietly=TRUE)
-}
-for (s in sources) {
-  source(s)
-}
-
-
-tasks <- tasks_growth(iter = 20, name = 'test')
-tasks <- df_to_list(tasks[c(1,2),])
-r1 <- model_compiler(tasks[[1]])
-r2 <- model_compiler(tasks[[2]])
-
-# First combine chains
-model <- sflist2stanfit(c(readRDS(r1), readRDS(r2)))
-
-# Will give you a trace plot of important parameters
-to.png(
-  traceplot(model, pars = c("mu_a0","sigma_log_a0","sigma_log_a1","a2",
-                       "mu_b0","sigma_log_b0","sigma_log_b1","b2",
-                       "mu_c0","sigma_log_c0","sigma_log_c1","c2"),ncol=3, nrow=4)
-  , paste0("results/", remake:::git_sha(), ".png"))
-
-# Check sampling diagnostics
-# Want max_depth to be less than 15 and n_divergent to = 0 for all (or most iterations)
-get_sampler_params(model)

### 2015-09-02 ms.Rmd
Updated manuscript. Still need to fix up cross validation section

@@
 where, $R_i$ represents the true annual increase in dbh (cm) and $T_i$ represents the number of days between measurements divided by 365.25.
-We modelled $R_i$ as a random realisation from a lognormal distribution centered on zero and a standard deviation of 1, conveying our prior expectation that annual increases in dbh would be between 0.13 and 7.4 cm/yr.
-We used a lognormal as it ensures dbh growth cannot be negative.
+We modelled $R_i$ as a random realisation from a half normal distribution centered on zero and a standard deviation of 5, conveying our prior expectation that annual increases in dbh would be between 0 and 10 cm/yr.
+We used a half normal as it ensures dbh growth cannot be negative.
 
-We calculated basal area growth rate using: $\frac{(\pi \, 0.25 \,\widehat{dbh_{i,t}}) - (\pi \, 0.25 \, \widehat{dbh_{i,t-1}})}{T_i}$.
+\begin{equation} \label{eq:obs dbh increment}
+R_i \sim \textrm{N+}(0, 5). \end{equation}.
 
-We then calculated the mean posterior growth rate estimate for each individual and used these as our estimates of true dbh growth for the subsequent mortality analysis.
+For computational efficiency, we used the posterior means of $\dbh_t-1$, $dbh_t$ to calculate:
+
+dbh growth rate:  $\frac{\widehat{dbh_{i,t}} - \widehat{dbh_{i,t-1}}}{T_i}$ 
+basal area growth rate: $\frac{(\pi \, 0.25 \,\widehat{dbh_{i,t}}) - (\pi \, 0.25 \, \widehat{dbh_{i,t-1}})}{T_i}$.
+
+These estimates of individual post growth were then used in the subsequent mortality analysis.
 
@@
 
-We extend a common functional form used to model the growth-mortality relationship [e.g. @Kobe:1997vy] to include a growth-independent baseline hazard, such that the functional form becomes
+We extend a common functional form used to model the growth-mortality relationship [e.g. @Kobe:1995] to include a growth-independent baseline hazard, such that the functional form becomes
 \begin{equation} \label{eq:lambda}
@@
 To incorporate varability in species mortality rates, we add allowed $\alpha$, $\beta$ and $\gamma$ to vary by wood density - a species trait thought to encapsulate a species growth strategy (REFs).
-In addition to this, we also allowed the three parameters to vary by species ID and census period, by doing this, we effectively allow the model to partition variation explained by wood density from the residual variation explained by species, census and individual levels [@Gelman:2007te; @Kery:2010tp].
+In addition to this, we also allowed the three parameters to vary by species ID, by doing this, we effectively allow the model to partition variation explained by wood density from the residual variation explained by species and individual levels [@Gelman:2007te; @Kery:2010tp].
 
@@
 
-To account for species strategy and partition variance into explained and unexplained at the observation, species and sampling period level we modelled $\alpha$, $\beta$ and $\gamma$ via an equation of the form:
-\begin{equation} \label{eq:alpha}
-\alpha_s = \alpha_{0,s} \, \alpha_{1,t} \, \rho_s^{\alpha_{2}}.
+To account for species strategy and partition variance into explained and unexplained at the observation and species level we modelled $\alpha$, $\beta$ and $\gamma$ via the following submodels:
+\begin{equation} \label{eq:submodels}
+\alpha_s = \alpha_{0,s} \, \rho_s^{\alpha_{1}},
+\beta_s = \beta_{0,s} \, \rho_s^{\alpha_{1}},
+\gamma_s = \gamma_{0,s} \, \rho_s^{\alpha_{1}}.
 \end{equation}
-Here $\alpha_{0,s}$ represents the species-level random effect of species $s$ within which individual $i$ belongs, $\alpha_{1,t}$ represents the sample-level random effect for which time $t$ an individual was sampled, and $\alpha_{2}$ represents the effect of standardised wood density $\rho_s$ on $\alpha_s$.
-The form of eq \ref{eq:alpha} ensures that parameters remain positive.
-On a log scale equation \ref{eq:alpha} becomes an additive model,$\log \alpha_s = \log \alpha_{0,s} + \log \alpha_{1,t} + \alpha_{2} \, \log \rho_s$, equivalent to a classic proportional hazards model widely used in survival analysis.
-We standardised wood density via the transformation: $\rho_s = \rho/0.6$, an operation which is equivalent to centering a log-transformed variable at a density of 0.6 g cm${^-3}$.
+Here $\alpha_{0,s}$,$\beta_{0,s}$,$\gamma_{0,s}$ represents the species-level random effect of species $s$ within which individual $i$ belongs and $\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of standardised wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively.
+The form of submodels in eq \ref{eq:submodels} ensures that parameters remain positive.
+On a log scale these submodels \ref{eq:submodels} become additive (e.g.
+$\log \alpha_s = \log \alpha_{0,s} + \alpha_{1} \, \log \rho_s$, equivalent to a classic proportional hazards model widely used in survival analysis.
+We standardised wood density via the transformation: $\rho_s = \rho/0.6$), an operation which is equivalent to centering a log-transformed variable at a density of 0.6 g cm${^-3}$.
 This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant) [@Gelman:2007te].
 
-The species and census-level random effects $\alpha_{0,s}, \alpha_{1,t}$ were assumed to follow a log-normal distribution:
-\begin{equation} \label{eq:alpha0_alpha1}
-\alpha_i \sim \text{log-normal}\left(\mu_{\log \alpha_i}, \sigma_{\log \alpha_i}\right),\\
+The species random effects on each parameter were assumed to be a random realistation from a log-normal distribution:
+\begin{equation} \label{eq:RE_species}
+\alpha_{0,s} \sim \text{log-normal}\left(\mu_{\log \alpha_0}, \sigma_{\log \alpha_0}\right),\\
+\beta_{0,s} \sim \text{log-normal}\left(\mu_{\log \beta_0}, \sigma_{\log \beta_0}\right),\\
+\alpha_{0,s} \sim \text{log-normal}\left(\mu_{\log \gamma_0}, \sigma_{\log \gamma_0}\right),\\
 \end{equation}
-where $\mu_{\log \alpha_i},  \sigma_{\log \alpha_i}$ are the mean and standard deviation of log-transformed variables.
-We estimated $\mu_{\log \alpha_{0,s}}$, the average species effect at zero growth using the following informative prior:
+
+We estimated $\mu_{\log \alpha_0}$, the average log species effect at zero growth using the following informative prior:
 
 \begin{equation} \label{eq:alpha0}
-\mu_{\log \alpha_{0,s}} \sim \text{normal}\left(-0.61, 0.57\right).\\
+\mu_{\log \alpha_{0}} \sim \text{normal}\left(-0.61, 0.57\right).\\
 \end{equation}
 
-This distribution was derived from data presented by Kobe et al 1995, who estimated the probability of dying at zero growth/2.5 years for juveniles of 10 hardwood species across north-eastern America.
+This distribution was derived from data published by Kobe et al 1995, who estimated the probability of dying at zero growth/2.5 years for juveniles of 10 hardwood species across north-eastern America.
 Note that these data are also used to define the growth-mortality relationship in SORTIE-ND a spatially explicit forest dynamic model (REF).
-We converted these probabilities to instantaneous hazard rates by using $\lambda = -\log(1 - Pr(Death))/2.5$, and then log transforming these values and calculating the mean (i.e.
--0.61) associated standard error (i.e.
-0.).
-Based on this prior, we assumed that the average probability of death for an average species was 42%/yr with 95% of species effects occuring between 16% and 81%.
+We converted these probabilities/2.5 years to instantaneous hazard rates by using $\lambda = -\log(1 - Pr(Death))/2.5$, and then log transforming these values and calculating the mean (i.e.
+-0.61) and standard error (i.e.
+0.57).
+Based on this prior, we assumed that the average probability of death was between 16 and 81% with the central tendency being 42%/yr.
 
@@
 
-Similarly, we used prior information from Korning and Balslev et al (1994) to estimate the average species effect on $\gamma$, the growth-independent hazard rate, such that:
+Similarly, we used prior information from Korning and Balslev et al (1994) to estimate, $\mu_{\log \gamma_0$, the average log species effect on $\gamma$, such that:
 
 \begin{equation} \label{eq:gamma0}
-\mu_{\log \gamma_{0,s}} \sim \text{normal}\left(-3.86, 0.2\right).\\
+\mu_{\log \gamma_0 \sim \text{normal}\left(-3.86, 0.2\right).\\
 \end{equation}
 
-This distribution was estimated the same way as in $\mu_{\log \alpha_{0,s}}$.
+This distribution was estimated the same way as in $\mu_{\log \alpha_0$.
 That is we converted annual mortality probabilities of 17 tropical lowland tree species in Amazonian Ecuador that encompass the range of forest structure (i.e.
@@
 
-As we are fitting our model to two growth measures, we used weak priors to define the common log-normal distribution for the species effects on $\beta$:
+As we fitted our model to two growth measures (i.e. dbh growth and basal area growth), we used weak normal priors on $\mu_{\log \beta_0$, the average log species decay rate where:
 
-\begin{equation} \label{eq:beta_0,s}
-\mu_{\log \beta_{0,s}} \sim \text{lognormal}\left(0, 5\right).\\
+\begin{equation} \label{eq:beta_0}
+\mu_{\log \beta_0 \sim \text{normal}\left(0, 5\right).\\
 \end{equation}
@@
 
-The census random effects, $\alpha_1$, $\beta_1$ and $\gamma_1$ were all sampled from lognormal distributions centered on zero with a standard deviations estimated from the data. For example:
-
-\begin{equation} \label{eq:a1b1c1,s}
-\alpha_{1,t} \sim \text{lognormal}\left(0, \sigma_{a1}\right).\\
-\end{equation}
-
-The effect of wood density on all three parameters (i.e. $\alpha_2$, $\beta_2$ and $\gamma_2$) were given normal priors centered on zero and a standard deviation of 5:
+The effect of wood density on all three parameters (i.e. $\alpha_1$, $\beta_1$ and $\gamma_1$) were given normal priors centered on zero and a standard deviation of 5:
 
@@
 
-Standard deviations associated with both species random effects (i.e.
-$\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$) and those associated with census random effects (i.e.
-$\sigma_{a1}$, $\sigma_{b1}$, $\sigma_{c1}$), were all sampled from half cauchy priors with centered on zero with a scale of 2.5.
+Standard deviations associated with species random effects (i.e.
+$\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$), were all sampled from positive half cauchy priors with centered on zero with a scale of 2.5.
 
 \begin{equation} \label{eq:a1b1c1,s}
-\sigma_{a0},\sigma_{b0},\sigma_{c0},\sigma_{a1},\sigma_{b1},\sigma_{c1} \sim \text{cauchy+}\left(0, 2.5\right).\\
+\sigma_{a0},\sigma_{b0},\sigma_{c0} \sim \text{cauchy+}\left(0, 2.5\right).\\
 \end{equation}
+
+## Cross validation
+
+To implement K-fold cross-validation in Stan we need to repeatedly fit the model to one subset of the data and use it to predict the held-out set.
+The way we recommend setting this up is to do the partitioning in R (or Python, or whatever data-processing environment is being used) and then pass the training data and held-out data to Stan in two pieces.
+With linear regression, for example, we start with the Stan model on page 4, passing in the training data as N, y, X and the held-out set as N holdout, y holdout, X holdout, with the data block augmented accordingly.
+We then keep the data block and model as is, and just alter the generated quantities block to operate on the held-out data.
+
+For a model $M$, training data $D$, and test data $D^*$, we want to calculate
+
+$$p(D^* | D,M) = \int p(\theta | D, M) \, p(D^* | \theta, M) \, d \theta$$
+
+The distribution $p(\theta | D, M)$ gives the likelihood of parameter values $\theta$ given the training data (obtained from fitting the model) and is calculated (via Bayes Rule) as
+
+$$p(\theta | D, M) = \frac{p(\theta) \, p(D|\theta)}{p(D)}.$$
+
+$p(D^* | \theta, M)$ is then the probability of observing the test data given a particular parameter set (i.e the likelihood of the data) and is calculated as
+$$p(D^* | \theta, M) = \prod_i p(D^*_i | \theta, M).$$
+
+Log-likelihoods are easier to deal with and so we have
+
+$$\log p(D^* | \theta, M) = \sum_i \log p(D^*_i | \theta, M).$$
+
+Aki \& Gelman 2014 provide methods for estimating "an importance-sampling approximated LOO directly using the log-likelihood evaluated at the posterior simulations of the parameter values"
 
@@
   int<lower=1> spp[n_obs];
-  int<lower=1> n_census;
-  int<lower=1> census[n_obs];
   int<lower=0, upper=1> y[n_obs];
@@
   int<lower=1> spp_heldout[n_obs_heldout];
-  int<lower=1> census_heldout[n_obs_heldout];
   int<lower=0, upper=1> y_heldout[n_obs_heldout];
@@
 
-  real raw_log_a1[n_census];
-  real<lower=0> sigma_log_a1;
-
   real raw_log_b0[n_spp];
@@
   real<lower=0> sigma_log_b0;
-
-  real raw_log_b1[n_census];
-  real<lower=0> sigma_log_b1;
 
@@
 
-  real raw_log_c1[n_census];
-  real<lower=0> sigma_log_c1;
-
-  real a2;
-  real b2;
-  real c2;
 }
@@
   real a0[n_spp];
-  real a1[n_census];
 
@@
   real b0[n_spp];
-  real b1[n_census];
 
@@
   real c0[n_spp];
-  real c1[n_census];
 
@@
 
-  // Calculating census period random effects
-  for (t in 1:n_census) {
-    a1[t] <- exp(raw_log_a1[t] * sigma_log_a1); // e.g. implies lognormal(0, sigma_log_a1)
-    b1[t] <- exp(raw_log_b1[t] * sigma_log_b1);
-    c1[t] <- exp(raw_log_c1[t] * sigma_log_c1);
-  }
-
   for (i in 1:n_obs) {
     // Calculating mortality parameters
-    alpha <- a0[spp[i]] * a1[census[i]] * pow(rho_c[spp[i]], a2);
-    beta <- b0[spp[i]] * b1[census[i]] * pow(rho_c[spp[i]], b2);
-    gamma <- c0[spp[i]] * c1[census[i]] * pow(rho_c[spp[i]], c2);
+    alpha <- a0[spp[i]]; 
+    beta <- b0[spp[i]]; 
+    gamma <- c0[spp[i]];
 
@@
   raw_log_a0 ~ normal(0,1);
-  mu_log_a0 ~ normal(-0.87, 0.75);
+  mu_log_a0 ~ normal(-0.61, 0.57);
   sigma_log_a0 ~ cauchy(0, 2.5);
@@
   raw_log_c0 ~ normal(0, 1);
-  mu_log_c0 ~ normal(-4.43, 1);
+  mu_log_c0 ~ normal(-3.86, 0.2);
   sigma_log_c0 ~ cauchy(0, 2.5);
 
-  raw_log_a1 ~ normal(0,1);
-  sigma_log_a1 ~ cauchy(0, 2.5);
-
-  raw_log_b1 ~ normal(0, 1);
-  sigma_log_b1 ~ cauchy(0, 2.5);
-
-  raw_log_c1 ~ normal(0, 1);
-  sigma_log_c1 ~ cauchy(0, 2.5);
-
-  a2 ~ normal(0,5);
-  b2 ~ normal(0,5);
-  c2 ~ normal(0,5);
 }
@@
   real a0[n_spp];
-  real a1[n_census];
-
   real b0[n_spp];
-  real b1[n_census];
-
   real c0[n_spp];
-  real c1[n_census];
 
@@
   real gamma_fit;
+
   real cumulative_hazard_fit;
@@
   real gamma_heldout;
+
   real cumulative_hazard_heldout;
@@
 
-  // recalulate species random effects
-  for (t in 1:n_census) {
-    a1[t] <- exp(raw_log_a1[t] * sigma_log_a1);
-    b1[t] <- exp(raw_log_b1[t] * sigma_log_b1);
-    c1[t] <- exp(raw_log_c1[t] * sigma_log_c1);
-  }
-
   // log likelihood for fitted model
   for (i in 1:n_obs) {
-    alpha_fit <- a0[spp[i]] * a1[census[i]] * pow(rho_c[spp[i]], a2);
-    beta_fit <- b0[spp[i]] * b1[census[i]] * pow(rho_c[spp[i]], b2);
-    gamma_fit <- c0[spp[i]] * c1[census[i]] * pow(rho_c[spp[i]], c2);
+    alpha_fit <- a0[spp[i]]; 
+    beta_fit <- b0[spp[i]]; 
+    gamma_fit <- c0[spp[i]];
 
@@
   for (j in 1:n_obs_heldout) {
-    alpha_heldout <- a0[spp_heldout[j]] * a1[census_heldout[j]] * pow(rho_c_heldout[spp_heldout[j]], a2);
-    beta_heldout <- b0[spp_heldout[j]] * b1[census_heldout[j]] * pow(rho_c_heldout[spp_heldout[j]], b2);
-    gamma_heldout <- c0[spp_heldout[j]] * c1[census_heldout[j]] * pow(rho_c_heldout[spp_heldout[j]], c2);
+    alpha_heldout <- a0[spp_heldout[j]]; 
+    beta_heldout <- b0[spp_heldout[j]]; 
+    gamma_heldout <- c0[spp_heldout[j]];
 
@@
 \pagebreak
-
-## Cross validation
-
-We will implement *K-fold cross validation*.
-
-According to [Vehtari & Gelman 2014](http://www.stat.columbia.edu/~gelman/research/unpublished/waic_stan.pdf):
-
-  To implement K-fold cross-validation in Stan we need to repeatedly fit the model to one subset of the data and use it to predict the held-out set.
-The way we recommend setting this up is to do the partitioning in R (or Python, or whatever data-processing environment is being used) and then pass the training data and held-out data to Stan in two pieces.
-With linear regression, for example, we start with the Stan model on page 4, passing in the training data as N, y, X and the held-out set as N holdout, y holdout, X holdout, with the data block augmented accordingly.
-We then keep the data block and model as is, and just alter the generated quantities block to operate on the held-out data.
-
-For a model $M$, training data $D$, and test data $D^*$, we want to calculate
-
-$$p(D^* | D,M) = \int p(\theta | D, M) \, p(D^* | \theta, M) \, d \theta$$
-
-The distribution $p(\theta | D, M)$ gives the likelihood of parameter values $\theta$ given the training data (obtained from fitting the model) and is calculated (via Bayes Rule) as
-
-$$p(\theta | D, M) = \frac{p(\theta) \, p(D|\theta)}{p(D)}.$$
-
-$p(D^* | \theta, M)$ is then the probability of observing the test data given a particular parameter set (i.e the likelihood of the data) and is calculated as
-$$p(D^* | \theta, M) = \prod_i p(D^*_i | \theta, M).$$
-
-Log-likelihoods are easier to deal with and so we have
-
-$$\log p(D^* | \theta, M) = \sum_i \log p(D^*_i | \theta, M).$$
-
-Aki \& Gelman 2014 provide methods for estimating "an importance-sampling approximated LOO directly using the log-likelihood evaluated at the posterior simulations of the parameter values"
 

### 2015-09-02 ms.Rmd
Fixed minor typos that caused pdf to fail

@@
 output: pdf_document
-csl: downloads/style.csl
-bibliography: data/refs.bib
+#csl: downloads/style.csl
+#bibliography: data/refs.bib
 ---
@@
 where, $\mu_{dbh_{t-1}}$ defines the log of the grand mean initial dbh, and $\sigma_{dbh_{t-1}}$ defines that variation around that mean.
-$\mu_{dbh_{t-1}}$ was estimated using a normal prior $\textrm{N}(0, 2.5)$, which represents our prior expectation that the mean initial dbh would likely occur between 0 and 148 cm.
+$\mu_{dbh_{t-1}}$ was estimated using a normal prior $\text{N}(0, 2.5)$, which represents our prior expectation that the mean initial dbh would likely occur between 0 and 148 cm.
 $\sigma_{dbh{t-1}}$ was estimated using a weakly informative half cauchy prior centered on zero and a scale parameter of 2.5.
@@
 \begin{equation} \label{eq:true dbh2}
-\widehat{dbh_{i,t-1}} + R_i \, T_i, \end{equation}
+\widehat{dbh_{i,t-1}} + R_i \, T_i, 
+\end{equation}
 
@@
 \begin{equation} \label{eq:obs dbh increment}
-R_i \sim \textrm{N+}(0, 5). \end{equation}.
+R_i \sim \textrm{N+}(0, 5). 
+\end{equation}
 
-For computational efficiency, we used the posterior means of $\dbh_t-1$, $dbh_t$ to calculate:
+For computational efficiency, we used the posterior means of $dbh_{t-1}$, $dbh_t$ to calculate:
 
@@
 \begin{equation} \label{eq:submodels}
-\alpha_s = \alpha_{0,s} \, \rho_s^{\alpha_{1}},
-\beta_s = \beta_{0,s} \, \rho_s^{\alpha_{1}},
+\alpha_s = \alpha_{0,s} \, \rho_s^{\alpha_{1}},\\
+\beta_s = \beta_{0,s} \, \rho_s^{\alpha_{1}}, \\
 \gamma_s = \gamma_{0,s} \, \rho_s^{\alpha_{1}}. 
 \end{equation}
+
 Here $\alpha_{0,s}$,$\beta_{0,s}$,$\gamma_{0,s}$ represents the species-level random effect of species $s$ within which individual $i$ belongs and $\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of standardised wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively.
@@
 $\log \alpha_s = \log \alpha_{0,s} + \alpha_{1} \, \log \rho_s$, equivalent to a classic proportional hazards model widely used in survival analysis.
-We standardised wood density via the transformation: $\rho_s = \rho/0.6$), an operation which is equivalent to centering a log-transformed variable at a density of 0.6 g cm${^-3}$.
+We standardised wood density via the transformation: $\rho_s = \frac{\rho}{0.6}$), an operation which is equivalent to centering a log-transformed variable at a density of 0.6 g cm${^-3}$.
 This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant) [@Gelman:2007te].
@@
 
-EXAMPLE:
 ```{r}
@@
 # sugar maple
-median(c(0.998, 0.148, 0.999)
+median(c(0.998, 0.148, 0.999))
 # American beech
@@
 
-Similarly, we used prior information from Korning and Balslev et al (1994) to estimate, $\mu_{\log \gamma_0$, the average log species effect on $\gamma$, such that:
+Similarly, we used prior information from Korning and Balslev et al (1994) to estimate, $\mu_{\log \gamma_0}$, the average log species effect on $\gamma$, such that:
 
 \begin{equation} \label{eq:gamma0}
-\mu_{\log \gamma_0 \sim \text{normal}\left(-3.86, 0.2\right).\\
+\mu_{\log \gamma_0} \sim \textrm{N}\left(-3.86, 0.2\right).\\
 \end{equation}
 
-This distribution was estimated the same way as in $\mu_{\log \alpha_0$.
+This distribution was estimated the same way as in $\mu_{\log \alpha_0}$.
 That is we converted annual mortality probabilities of 17 tropical lowland tree species in Amazonian Ecuador that encompass the range of forest structure (i.e.
@@
 sd(log(k))/sqrt(length(M))
+
 ```
-
-As we fitted our model to two growth measures (i.e. dbh growth and basal area growth), we used weak normal priors on $\mu_{\log \beta_0$, the average log species decay rate where:
+As we fitted our model to two growth measures (i.e. dbh growth and basal area growth), we used weak normal priors on $\mu_{\log \beta_0}$, the average log species decay rate where:
 
 \begin{equation} \label{eq:beta_0}
-\mu_{\log \beta_0 \sim \text{normal}\left(0, 5\right).\\
+\mu_{\log \beta_0} \sim \text{N}\left(0, 5\right).\\
 \end{equation}
@@
 
-\begin{equation} \label{eq:a2b2c2}
-\alpha_2, \beta_2, \gamma_2 \sim \text{normal}\left(0, 5\right).\\
+\begin{equation} \label{eq:a1b1c1}
+\alpha_1, \beta_1, \gamma_1 \sim \text{N}\left(0, 5\right).\\
 \end{equation}
@@
 
-\begin{equation} \label{eq:a1b1c1,s}
+\begin{equation} \label{eq:sigma}
 \sigma_{a0},\sigma_{b0},\sigma_{c0} \sim \text{cauchy+}\left(0, 2.5\right).\\

### 2016-01-11 ms.Rmd
edits to manuscript introduction

@@
 
-Understanding who lives or dies, and why, is critical to accurately predict population, community and ecosystem dynamics [@Hawkes:2000ib; @McDowell:2011dr].
-In plants, the carbon budget is considered a fundamental determinant of mortality, such that if rates of carbon assimilation are low or become insufficient for growth, an individual is more susceptible to dying [@Peet:1987va; @Hawkes:2000ib; @Keane:2001db; @McDowell:2011dr].
-Empirical studies examining intraspecific mortality have broadly supported this assumption, with many finding that the probability a plant dies decreases exponentially with increasing growth rates [@Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw].
-Consequently, most dynamic vegetation models, including tree gap models [@Botkin:1972ii; @Keane:2001db], SORTIE (REF) and Dynamic Global Vegetation Models (DGVM's; @Woodward:2004ft) use the growth-mortality relationship as the predominant, and in some cases, only mechanism describing plant mortality [@Hawkes:2000ib; @McDowell:2011dr].
-However, despite empirical evidence highlighting that mortality rates vary considerably between species, many of  dynamic model assume the shape of the growth-mortality relationship is constant across species.
+Understanding who lives or dies, and why, is critical to predicting population, community and ecosystem dynamics [@Hawkes:2000ib; @McDowell:2011dr].
+In plants, the carbon budget is considered a fundamental determinant of mortality, such that if rates of carbon assimilation are low or become insufficient for growth, an individual has a higher probability of dying [@Peet:1987va; @Hawkes:2000ib; @Keane:2001db; @McDowell:2011dr].
+Empirical studies examining intraspecific mortality have broadly supported this assumption, with many finding that probability of death declines exponentially with increasing growth rates [@Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw].
+Consequently, most dynamic vegetation models, including tree gap models [@Botkin:1972ii; @Keane:2001db], individual-based forest simulators (e.g.
+SORTIE; REF) and Dynamic Global Vegetation Models (DGVM's; @Woodward:2004ft) use this growth-mortality relationship as the predominant, and in some cases, only mechanism describing plant mortality [@Hawkes:2000ib; @McDowell:2011dr].
 
-Multispecies studies have shown that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species [@Poorter:2008iu; @Wright:2010fl], and thus, species have different mortality rates.
-Ecologists have attempted to explain this variation between species by using functional traits - attributes of species that are thought to influence vital rates (i.e.
-survival, growth and reproduction) and ultimately fittness [@Ackerly:2003eb; @Poorter:2008iu].
-Wood density, the biomass invested per unit wood volume, is one such functional trait that has been shown to be negatively correlated with tropical tree mortality [@Chave:2009iy; @Nascimento:2005wx; @King:2006he; @AlvarezClare:2007gv; @Chao:2008hg; @Poorter:2008iu; @Kraft:2010kq; @Wright:2010fl].
+However, many dynamic vegetation models assume the growth-mortality relationship is constant across species.
+This is despite empirical evidence highlighting that mortality rates vary considerably between species.
+In contrast to intraspecific studies, multispecies studies have shown that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species [@Poorter:2008iu; @Wright:2010fl].
+Ecologists have attempted to explain this interspecific variation by using functional traits - attributes of species that are thought to influence vital rates (i.e.
+survival, growth and reproduction) and ultimately fitness [@Ackerly:2003eb; @Poorter:2008iu].
+Wood density, the biomass invested per unit wood volume, is one such functional trait that has been consistently shown to be negatively correlated with tropical tree mortality [@Chave:2009iy; @Nascimento:2005wx; @King:2006he; @AlvarezClare:2007gv; @Chao:2008hg; @Poorter:2008iu; @Kraft:2010kq; @Wright:2010fl].
 The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to stem breakage [@Putz:1983wu; @vanGelder:2006ex; @Chave:2009iy], wood herbivory (NEED REF), embolism [@Hacke:2001kj, @Jacobsen:2005fx, @Preston:2006jn] and fungal and pathogen attack [@Augspurger:1984wx].
-However, research has also indicated that wood density may increase survival associated with growth-dependent mortality.
-For example, species with high wood density are often more shade tolerant, and thus exhibit a lower probability of dying at low growth rates [e.g.
+However, research has also shown that wood density may increase survival associated with growth-dependent mortality.
+For example, species with high wood density are often more shade tolerant, [ADD something about competitition from Kunstler et al Nature paper] and thus exhibit a lower probability of dying at low growth rates [e.g.
 @vanGelder:2006ex].
@@
 
-To date, most plant mortality research can be broadly seperated into two research focuses: 1) those that focus on estimating individual mortality within species [e.g.
-@Kobe:1997vy]; and 2) those that use suites of species to examine correlations between functional traits and population level estimates of plant mortality (e.g.
+To date, researchers have examined plant mortality by: 1) using individual based covariates (e.g.
+growth, size) within species [e.g.
+@Kobe:1997vy]; or 2) examining correlations between population level estimates of plant mortality (e.g.
 Wright et al 2010, Chave et al 2004).
-However, little crossover of ideas exists between these two research focuses.
-This is despite both individual and species level variation likely being required to build a general model of plant mortality.
-Furthermore, in both research focuses, the measure of growth rate used to examine the growth-mortality relationship appears to be based on researcher preference with little consideration given as to whether an area or diameter based growth rate is a better predictor of plant mortality.
+However, there has been little crossover of ideas between these two research focuses.
+For example, as far as we are aware, no study has examined how functional traits mediate the negative exponential relationship between growth and mortality.
+Furthermore, some fundamental questions have yet to be answered such as whether an area or diameter based growth rate better predicts plant mortality.
+Lastly, we are unaware of any attempt to partition plant mortality into growth
+
+Lastly, we are unaware of a model that attempts to simultaneously estimat
+
+Lastly, we are unaware of a model that attempts because most plant mortality data is binary (i.e. alive/dead),
+
+there has been there have been few attempts to partition mortality into growth related and stochastic
+
+ little consideration has been given to whether an area or diameter based growth rate better predicts plant mortality.
+This is despite both individual (e.g.
+growth rates) and species (e.g.
+function traits) characteristics likely to be required in order to develop an accurate and general model of how the growth effects plant mortality.
+Furthermore, both research focuses have used
+
+In this study, we reconcile both research focuses by using 15-years of growth and mortality data from 203 tropical rainforest species encompassing more than 180,000 individuals in a Bayesian hierarchical model that explicitly accounts for variability observed both within and between species.
+Specifically, our model uses the negative exponential growth-mortalty relationship commonly observed and fitted in intraspecific studies and extends this by both adding a growth-independent baseline hazard, and by allowing each parameter to vary as a function of species level traits (in this case wood density).
+
+## BASELINE HAZARDS
+## GROWTH MEASURE
 

### 2016-01-21 ms.Rmd
Restructured/rewrote introduction

--- 2015-09-02 ms.Rmd
@@
 ---
-title: "Wood density and the growth-dependent & independent mortality"
-author: "J S Camac"
-date: "17 August 2015"
+title: "Modelling plant mortality: integrating intra and interspecific approaches"
+author: James S Camac, Richard G FitzJohn, Mark Westoby, Daniel S Falster
+date: "21 January 2016"
+csl: data/style.csl
+bibliography: data/refs.bib
 output: pdf_document
-#csl: downloads/style.csl
-#bibliography: data/refs.bib
 ---
+# Abstract
 
 # Introduction
+Plant mortality is often modelled as a function of recent growth because it is thought to reflect the summed effects of many biotic and abiotic factors on a plants overall carbon budget [Kobe:1995tw; @Hawkes:2000ib; @Keane:2001db; @Wyckoff:2002ul].
+As such, growth rates are considered a highly integrated individual-based predictor of plant mortality [@Hawkes:2000ib].
+Within a species, the assumption of the growth-mortality relationship is that fast growing individuals (i.e.
+those that have larger carbon budgets) are more able to buffer stress and thus have higher survivorship relative to slow growing (i.e.
+low carbon budget) individuals [@Hawkes:2000ib; @Keane:2001db].
+Empirical studies examining this growth-mortality relationship have broadly supported this assumption, with many highlighting non-linear declines in mortality with increased growth [e.g.
+@Kobe:1997vy; @Wyckoff:2002ul; @Vieilledent:2010fv].
+Consequently, most dynamic vegetation models, including tree gap models [@Keane:2001db], individual-based forest simulators [e.g.
+SORTIE; @Pacala:1996vm] and Dynamic Global Vegetation Models [DGVMs; @Woodward:2004ft] use some measure of growth coupled with a stochastic constant (i.e.
+baseline hazard or minimum chance of death) as the fundamental determinants of plant mortality [@Hawkes:2000ib; @McDowell:2011dr].
 
-Understanding who lives or dies, and why, is critical to accurately predict population, community and ecosystem dynamics [@Hawkes:2000ib; @McDowell:2011dr].
-In plants, the carbon budget is considered a fundamental determinant of mortality, such that if rates of carbon assimilation are low or become insufficient for growth, an individual is more susceptible to dying [@Peet:1987va; @Hawkes:2000ib; @Keane:2001db; @McDowell:2011dr].
-Empirical studies examining intraspecific mortality have broadly supported this assumption, with many finding that the probability a plant dies decreases exponentially with increasing growth rates [@Buchman:1983vx; @Kobe:1997vy; @Wyckoff:2002ul; @Mantgem:2003dw].
-Consequently, most dynamic vegetation models, including tree gap models [@Botkin:1972ii; @Keane:2001db], SORTIE (REF) and Dynamic Global Vegetation Models (DGVM's; @Woodward:2004ft) use the growth-mortality relationship as the predominant, and in some cases, only mechanism describing plant mortality [@Hawkes:2000ib; @McDowell:2011dr].
-However, despite empirical evidence highlighting that mortality rates vary considerably between species, many of  dynamic model assume the shape of the growth-mortality relationship is constant across species.
+However, due to a paucity of data on plant mortality, most dynamic vegetation models have implemented mortality algorithms using overly simplistic assumptions [@Hawkes:2000ib; Wyckoff:2000un; @McDowell:2011dr].
+One of the strongest assumptions these models make is that both baseline hazard and the growth-mortality relationship are constant across species [@Pacala:1996vm, @Hawkes:2000ib; Wyckoff:2000un].
+That is, all species have the same random chance of death and exhibit the same tolerance to low growth.
+This is despite growing empirical evidence highlighting that mortality rates vary considerably between species, even within the same community (@Kobe:1997vy; @Wyckoff:2002ul; @Kunstler:2005bn; @Vieilledent:2010fv).
+More recently, however, attempts have been made to examine this inter-specific variation by correlating species intrinsic growth and mortality rates [e.g.
+@Poorter:2008iu; @Wright:2010fl].
+In contrast to intraspecific studies, these studies have shown that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species [Loehle:1996vw; @Poorter:2008iu; @Wright:2010fl].
+As such, species appear to exhibit a trade off between greater growth versus greater survival.
+Ecologists have attempted to explain this interspecific variation by using functional traits - attributes of species that are thought to influence vital rates (i.e.
+survival, growth and reproduction) and ultimately fitness [@Westoby:2002tb; @Ackerly:2003eb; @Falster:2015un].
 
-Multispecies studies have shown that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species [@Poorter:2008iu; @Wright:2010fl], and thus, species have different mortality rates.
-Ecologists have attempted to explain this variation between species by using functional traits - attributes of species that are thought to influence vital rates (i.e.
-survival, growth and reproduction) and ultimately fittness [@Ackerly:2003eb; @Poorter:2008iu].
-Wood density, the biomass invested per unit wood volume, is one such functional trait that has been shown to be negatively correlated with tropical tree mortality [@Chave:2009iy; @Nascimento:2005wx; @King:2006he; @AlvarezClare:2007gv; @Chao:2008hg; @Poorter:2008iu; @Kraft:2010kq; @Wright:2010fl].
-The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to stem breakage [@Putz:1983wu; @vanGelder:2006ex; @Chave:2009iy], wood herbivory (NEED REF), embolism [@Hacke:2001kj, @Jacobsen:2005fx, @Preston:2006jn] and fungal and pathogen attack [@Augspurger:1984wx].
-However, research has also indicated that wood density may increase survival associated with growth-dependent mortality.
-For example, species with high wood density are often more shade tolerant, and thus exhibit a lower probability of dying at low growth rates [e.g.
-@vanGelder:2006ex].
+Wood density, the biomass invested per unit wood volume, is one functional trait that has been consistently shown to be negatively correlated with tropical tree mortality [@Chave:2009iy; @Poorter:2008iu; @Kraft:2010kq; @Wright:2010fl].
+The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to random events such as stem breakage [@Putz:1983wu; @vanGelder:2006exa; @Chave:2009iy], embolism [@Hacke:2001kj, @Jacobsen:2005fx] and fungal and pathogen attack [@Augspurger:1984wx].
+As such, it is hypothesized that species with higher wood density are likely to exhibit lower baseline hazards.
+However, research has also shown that wood density may increase tolerance to low growth rates, and thus alter the growth-mortality relationship.
+For example, species with high wood density are more tolerant shade [@vanGelder:2006exa] and competition [@Kunstler:2016km].
 Studies have also shown that wood density may reduce drought induced carbon starvation [@McDowell:2011dr] and the susceptibility of ‘stressed’ individuals to opportunistic invertebrate, pathogen or fungal attack [@Augspurger:1984wx; @McDowell:2011dr].
 
-To date, most plant mortality research can be broadly seperated into two research focuses: 1) those that focus on estimating individual mortality within species [e.g.
-@Kobe:1997vy]; and 2) those that use suites of species to examine correlations between functional traits and population level estimates of plant mortality (e.g.
-Wright et al 2010, Chave et al 2004).
-However, little crossover of ideas exists between these two research focuses.
-This is despite both individual and species level variation likely being required to build a general model of plant mortality.
-Furthermore, in both research focuses, the measure of growth rate used to examine the growth-mortality relationship appears to be based on researcher preference with little consideration given as to whether an area or diameter based growth rate is a better predictor of plant mortality.
+What both intraspecific and interspecific studies have highlighted is that both individual (e.g.
+growth rates) and species (e.g.
+function traits) characteristics are required in order to develop generalizable models of plant mortality.
+However, to date, there has been little crossover of ideas between intraspecific and interspecific studies of plant mortality.
+For example, we are aware of only one study that uses individual based plant mortality empirical models to examine trait effects on baseline hazards [@AubryKientz:2013dg], and none that examine how traits mediate the growth-mortality relationship.
 
-In this study, we reconcile both research focuses by fitting 15-years of growth and mortality data from 203 tropical rainforest species encompassing more than 180,000 individuals in a Bayesian hierarchical model that explicitly accounts for variability observed both within and between species.
-Specifically, our model uses the negative exponential growth-mortalty relationship commonly observed and fitted in intraspecific studies and extends this by both adding a growth-independent baseline hazard, and by allowing each parameter to vary as a function of species level traits (in this case wood density).
-
-We then use this model, coupled with k-fold cross validation to address four questions:
-
-1. Which growth measure is most predictive of plant mortality?
-2. Does wood density influence both the growth-mortality trade-off and growth independent mortality?
-3. Does the effects of wood density vary with growth measure? and
-4. What is the most parsimonious mortality model in terms of wood density effects?
+In this study, we reconcile both intraspecific and interspecific theory by developing an individual based plant mortality model that incorporates individual based measurements of growth as well as the species trait wood density.
+Furthermore, we extend existing individual-based approaches [e.g.
+@Kobe:1995tw] by simultaneously estimating both baseline hazards and the growth-mortality relationship (hereon in referred to as growth-dependent hazard).
+We parameterise this model using 15-years of individual dbh growth and mortality data from 203 tropical rainforest species encompassing more than 180,000 individuals.
+Using this model,coupled with cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to the null model or growth-dependent only model (e.g.
+Kobe et al 1996) 2) Does allowing both hazards to vary by species improve predictive capacity?
+3) Does wood density predominately affect baseline hazards, or does it affect growth-dependent hazards or both?
+4) What better predicts plant mortality, absolute dbh growth or basal area growth?
 
@@
 
-Within the 50-ha BCI plot the diameter at breast height and survival status of all free-standing woody plants with a diameter greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter [@BarroColoradofores:2012up].
+Within the 50-ha BCI plot the diameter at breast height and survival status of all free-standing woody plants with a diameter greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter [@Condit:2012nz].
 For the purpose of this study, we discarded data collected in 1982 and 1985 censuses because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
@@
 
-We extend a common functional form used to model the growth-mortality relationship [e.g. @Kobe:1995] to include a growth-independent baseline hazard, such that the functional form becomes
+We extend a common functional form used to model the growth-mortality relationship [e.g. @Kobe:1995tw] to include a growth-independent baseline hazard, such that the functional form becomes
 \begin{equation} \label{eq:lambda}
@@
 
-```{r}
-# white ash
-median(c(0.999,0.55,0.998))
-# sugar maple
-median(c(0.998, 0.148, 0.999))
-# American beech
-median(c(0.014,0.12))
-# Yellow Birch
-median(c(0.555,0.995))
-# Red Maple
-0.99
-# Eastern hemlock
-0.077
-# Black cherry
-0.998
-# White oak
-0.82
-# Red oak
-0.985
-#White pine
-0.268
-
-# mean species instantaneous hazard rate (used medians for species with multiple measurements)
-M <- c(0.998,0.998,0.067,0.775,0.99,0.077,0.998,0.82,0.985, 0.268)
-k = - log(1-M)/2.5
-mean(log(k))
-
-# standard error
-sd(log(k))/sqrt(10)
-```
-
 Similarly, we used prior information from Korning and Balslev et al (1994) to estimate, $\mu_{\log \gamma_0}$, the average log species effect on $\gamma$, such that:
@@
 
-EXAMPLE:
-```{r}
-M <- c(8.6,1.8,0.6,3.8,0.9,0.7,4.3,4.7,1.8,6.4,0.8,1,2,1.8,3.7,2.2,2.2)/100 # From Korning Baslev 1994
-k = - log(1-M)/1
-mean(log(k))
-
-# standard error
-sd(log(k))/sqrt(length(M))
-
-```
 As we fitted our model to two growth measures (i.e. dbh growth and basal area growth), we used weak normal priors on $\mu_{\log \beta_0}$, the average log species decay rate where:
@@
 
-## STAN MODEL
-```{r, eval=FALSE}
-data {
-  int<lower=1> n_obs;
-  int<lower=1> n_spp;
-  int<lower=1> spp[n_obs];
-  int<lower=0, upper=1> y[n_obs];
-  vector[n_obs] census_length;
-  vector[n_obs] growth_dt;
-  vector[n_spp] rho_c;
-
-  // Held out data
-  int<lower=1> n_obs_heldout;
-  int<lower=1> n_spp_heldout;
-  int<lower=1> spp_heldout[n_obs_heldout];
-  int<lower=0, upper=1> y_heldout[n_obs_heldout];
-  vector[n_obs_heldout] census_length_heldout;
-  vector[n_obs_heldout] growth_dt_heldout;
-  vector[n_spp_heldout] rho_c_heldout;
-}
-
-parameters {
-
-  // Mortality model parameters
-  real raw_log_a0[n_spp];
-  real mu_log_a0;
-  real<lower=0> sigma_log_a0;
-
-  real raw_log_b0[n_spp];
-  real mu_log_b0;
-  real<lower=0> sigma_log_b0;
-
-  real raw_log_c0[n_spp];
-  real mu_log_c0;
-  real<lower=0> sigma_log_c0;
-
-}
-
-model {
-
-  // Declaring mortality parameters
-  real alpha;
-  real a0[n_spp];
-
-  real beta;
-  real b0[n_spp];
-
-  real gamma;
-  real c0[n_spp];
-
-  real cumulative_hazard;
-
-  // Calculating species random effects
-  for (s in 1:n_spp) {
-    a0[s] <- exp(raw_log_a0[s] * sigma_log_a0 + mu_log_a0); // e.g. implies lognormal(mu_log_a0, sigma_log_a0)
-    b0[s] <- exp(raw_log_b0[s] * sigma_log_b0 + mu_log_b0);
-    c0[s] <- exp(raw_log_c0[s] * sigma_log_c0 + mu_log_c0);
-  }
-
-  for (i in 1:n_obs) {
-    // Calculating mortality parameters
-    alpha <- a0[spp[i]]; 
-    beta <- b0[spp[i]]; 
-    gamma <- c0[spp[i]];
-
-    // Likelihood for hazard model
-    cumulative_hazard <- -census_length[i] * (alpha * exp(-beta * growth_dt[i]) + gamma);
-
-    if (y[i] == 0) {
-      increment_log_prob(cumulative_hazard);
-    } else {
-      increment_log_prob(log1m_exp(cumulative_hazard));
-    }
-  }
-
-  // Priors
-
-  //Mortality model priors
-  raw_log_a0 ~ normal(0,1);
-  mu_log_a0 ~ normal(-0.61, 0.57);
-  sigma_log_a0 ~ cauchy(0, 2.5);
-
-  raw_log_b0 ~ normal(0, 1);
-  mu_log_b0 ~ normal(0, 5);
-  sigma_log_b0 ~ cauchy(0, 2.5);
-
-  raw_log_c0 ~ normal(0, 1);
-  mu_log_c0 ~ normal(-3.86, 0.2);
-  sigma_log_c0 ~ cauchy(0, 2.5);
-
-}
-
-generated quantities {
-
-  // Declaring fitted parameters
-  real a0[n_spp];
-  real b0[n_spp];
-  real c0[n_spp];
-
-  real alpha_fit;
-  real beta_fit;
-  real gamma_fit;
-
-  real cumulative_hazard_fit;
-  real log_lik_fit;
-  real sum_log_lik_fit;
-
-  // Declaring heldout parameters
-  real alpha_heldout;
-  real beta_heldout;
-  real gamma_heldout;
-
-  real cumulative_hazard_heldout;
-  real log_lik_heldout[n_obs_heldout];
-  real sum_log_lik_heldout;
-
-  // Initialization of summed log likelihoods
-  sum_log_lik_fit <- 0;
-  sum_log_lik_heldout <- 0;
-
-  // recalulate species random effects
-  for (s in 1:n_spp) {
-    a0[s] <- exp(raw_log_a0[s] * sigma_log_a0 + mu_log_a0);
-    b0[s] <- exp(raw_log_b0[s] * sigma_log_b0 + mu_log_b0);
-    c0[s] <- exp(raw_log_c0[s] * sigma_log_c0 + mu_log_c0);
-  }
-
-  // log likelihood for fitted model
-  for (i in 1:n_obs) {
-    alpha_fit <- a0[spp[i]]; 
-    beta_fit <- b0[spp[i]]; 
-    gamma_fit <- c0[spp[i]];
-
-    cumulative_hazard_fit <- -census_length[i] * (alpha_fit * exp(-beta_fit * growth_dt[i]) + gamma_fit);
-
-    if (y[i] == 0) {
-      log_lik_fit <- cumulative_hazard_fit;
-    }
-    else {
-      log_lik_fit <- log1m_exp(cumulative_hazard_fit);
-    }
-    sum_log_lik_fit <- sum_log_lik_fit + log_lik_fit;
-  }
-
-  // log likelihood for held out data
-  for (j in 1:n_obs_heldout) {
-    alpha_heldout <- a0[spp_heldout[j]]; 
-    beta_heldout <- b0[spp_heldout[j]]; 
-    gamma_heldout <- c0[spp_heldout[j]];
-
-    cumulative_hazard_heldout <- -census_length_heldout[j] * (alpha_heldout * exp(-beta_heldout * growth_dt_heldout[j]) + gamma_heldout);
-
-    if (y_heldout[j] == 0) {
-      log_lik_heldout[j] <- cumulative_hazard_heldout;
-    }
-    else {
-      log_lik_heldout[j] <- log1m_exp(cumulative_hazard_heldout);
-    }
-    sum_log_lik_heldout <- sum_log_lik_heldout + log_lik_heldout[j];
-  }
-} 
-```
-
-\pagebreak
-
 # OTHER STUFF
@@
 
-## Analysis plan
-
-Step 1 - define a general structure of model we think will capture data, on basis of prior knowledge [DONE]
-Step 2 - assuming species effects on all pars, ask which growth meausre best predicts data, compare 2 growth rates using cross validation
-Step 3 - fit full model with trait effects, look at effect sizes of WD on a, b and c
-Step 4 - based on 3, choose some subsets of models, compare these using both bayes factor (most parsimonious) and predictive capacity
-
 \pagebreak
 
-\pagebreak
-
-## Results
-**Prelim results:**
-
-**DBH_DT**
-Wood density effects on alpha, beta and gamma:
-*alpha* - no significant effect
-*beta* - positive effect (increases decay rate)
-*gamma* - negative effect (decreases hazard rate)
-
-**BASAL_DT**
-Wood density effects on alpha, beta and gamma:
-*alpha* - no significant effect
-*beta* - positive effect (increases decay rate)
-*gamma* - marginal negative effect (not significant @ p <0.05)
-
-**Best growth measure**
-Preliminary it appears to be dbh_dt as it has a higher log likelihood.
-
-\pagebreak
-
-# Figures
-
-```{r model description, fig.height=6,fig.width=6, echo=FALSE}
-par(mfrow=c(2,1), mar=c(3,4,1,1), xaxs='i')
-       curve(exp(1 -exp(-0.3) * x) + exp(-4), xlim=c(0,12), ylim=c(0,5),
-             xlab=NA, ylab='Hazard rate',xaxt='n', col='red')
-       arrows(x0 = 2, y0 = 2.7, x1 = 0 ,y1 = 2.7)
-       text(3.1,2.7, 'alpha')
-       arrows(x0 = 4, y0 = 1.5, x1 = 1.5 ,y1 = 1)
-       text(4.5,1.7, 'beta')
-       arrows(x0 = 11, y0 = 0.8, x1 = 11 ,y1 = 0.018)
-       text(11,1, 'gamma')
-       axis(1, at=0, labels= 0)
-
-       curve((1-exp(-(exp(1 - exp(-0.3) * x) + exp(-4))))*100, xlim=c(0,12), ylim=c(0,100),
-             xlab=NA, ylab='Pr(Mortality)/yr', xaxt='n', col='red')
-       text(3.1,93, 'alpha')
-       arrows(x0 = 2, y0 = 93, x1 = 0 ,y1 = 93)
-       text(6,30, 'beta')
-       arrows(x0 = 5, y0 = 30, x1 = 3 ,y1 = 30)
-       text(11,15, 'gamma')
-       arrows(x0 = 11, y0 = 10, x1 = 11 ,y1 = exp(-4)*100)
-       axis(1, at=0, labels= 0)
-       mtext(side=1, 'Growth rate',line = 2)
-```
-
-**Fig 6:** Interpreting model parameters.
-Top = Hazard curves (i.e.
-instantaneous mortality rates); Bottom = Mortality probability/year based on estimated hazard rates.
-The intercept at zero growth is defined by $alpha$.
-The slope is defined by $beta$.
-The asymptote is defined by $\gamma$ and represents the growth-independent hazard.
-
 ## References

### 2016-01-25 R/process_output.R
Increased efficiency of processing functions and redid summarising functions

@@ -1,21 +1,9 @@
-# packages <- c("rstan")
-# sources <- c("R/model.R",
-#              "R/stan_functions.R",
-#              "R/utils.R")
-# 
-# for (p in packages) {
-#   library(p, character.only=TRUE, quietly=TRUE)
-# }
-# for (s in sources) {
-#   source(s)
-# }
-
 # Merge chains related to a given model/kfold combination
 combine_stan_chains <- function(files) {
-   sflist2stanfit(lapply(files, readRDS))
+  sflist2stanfit(lapply(files, readRDS))
 }
 
-# compile all models related to a given analyses
+# Compile all models related to a given analyses
 compile_models <- function(analysis) {
   if(!analysis %in% c("null_model","null_model_random_effects",
                       "no_gamma_model", "no_gamma_model_random_effects",
@@ -37,7 +25,7 @@ compile_models <- function(analysis) {
   
   fits <- lapply(sets, function(s) combine_stan_chains(s[['filename']]))
   pars <- lapply(sets,  function(s) s[1, c("analysis","growth_measure","rho_combo","kfold")])
-
+  
   list(model_info=pars, fits=fits)
 }
 
@@ -47,10 +35,10 @@ compile_multiple_analyses <- function(analysis) {
 }
 
 # Examine model diagnostics for single analysis
-kfold_model_diagnostics <- function(models) {
-  fits <- models$fits
-  info <- models$model_info
-  out1 <- do.call(rbind, lapply(fits, function(x) {
+kfold_diagnostics <- function(analysis) {
+  fits <- analysis$fits
+  info <- analysis$model_info
+  out1 <- bind_rows(lapply(fits, function(x) {
     summary_model <- summary(x)$summary
     sampler_params <- get_sampler_params(x, inc_warmup=FALSE)
     data.frame(
@@ -61,14 +49,14 @@ kfold_model_diagnostics <- function(models) {
       max_treedepth = max(sapply(sampler_params, function(y) y[,'treedepth__'])))
   }))
   
-  out2 <- do.call(rbind, lapply(info, function(x) {
+  out2 <- suppressWarnings(bind_rows(lapply(info, function(x) {
     data.frame(
       analysis = x$analysis,
       growth_measure = x$growth_measure,
       rho_combo = x$rho_combo,
-      kfold = as.integer(x$kfold)
-    )
-    }))
+      kfold = as.integer(x$kfold))
+  })))
+  
[…]

### 2016-01-25 R/stan_functions.R
Since models are running faster then expected Ive upped the iterations and acceptance rate to ensure more stability

@@ -71,7 +71,7 @@ model_compiler <- function(task) {
 }
 
 # Runs single chain
-run_single_stan_chain <- function(model, data, chain_id, iter=1000,
+run_single_stan_chain <- function(model, data, chain_id, iter=4000,
                                   sample_file=NA, diagnostic_file=NA) {
   data_for_stan <- prep_data_for_stan(data, model$growth_measure)
   stan(model_code = model$model_code,
@@ -81,7 +81,7 @@ run_single_stan_chain <- function(model, data, chain_id, iter=1000,
        iter = iter,
        chains=1,
        chain_id=chain_id,
-       control =list(stepsize=0.01, max_treedepth=15),
+       control =list(stepsize=0.01, adapt_delta=0.9, max_treedepth=15),
        refresh=1,
        sample_file=sample_file,
        diagnostic_file=diagnostic_file)
@@ -102,7 +102,7 @@ prep_data_for_stan <- function(data, growth_measure) {
     y_heldout = as.integer(data$heldout$dead_next_census),
     census_length_heldout = data$heldout$census_interval,
     growth_dt_heldout = data$heldout[[growth_measure]],
-    rho_c_heldout  = unique(data$heldout$rho/0.6)
+    rho_c_heldout = unique(data$heldout$rho/0.6)
   )
 }
 

### 2016-01-25 ms/ms.Rnw
Created manuscript folder

--- 2016-01-21 ms.Rmd
@@
----
-title: "Modelling plant mortality: integrating intra and interspecific approaches"
-author: James S Camac, Richard G FitzJohn, Mark Westoby, Daniel S Falster
-date: "21 January 2016"
-csl: data/style.csl
-bibliography: data/refs.bib
-output: pdf_document
----
-# Abstract
+\documentclass[a4paper,11pt]{article}
+\usepackage[osf]{mathpazo}
+\usepackage{ms}
+\usepackage[]{natbib}
+\usepackage{gensymb}
+%\raggedright
 
-# Introduction
-Plant mortality is often modelled as a function of recent growth because it is thought to reflect the summed effects of many biotic and abiotic factors on a plants overall carbon budget [Kobe:1995tw; @Hawkes:2000ib; @Keane:2001db; @Wyckoff:2002ul].
-As such, growth rates are considered a highly integrated individual-based predictor of plant mortality [@Hawkes:2000ib].
+\newcommand{\stan}{\texttt{stan}}
+
+\title{Modelling plant mortality: integrating intra and interspecific approaches}
+\author{James S. Camac\textasteriskcentered$^1$, Richard G. Fitzjohn$^1$, Mark Westoby$^1$, Daniel S. Falster$^1$}
+\affiliation{
+$^1$ Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia\\
+\textasteriskcentered Email for correspondence: \texttt{james.camac@gmail.com}\\
+Word count: }
+\date{}
+
+\bibliographystyle{mee.bst}
+
+\usepackage[title, titletoc, toc]{appendix}
+
+\mstype{Research Article}
+\runninghead{Incorporating traits into individual based mortality models}
+\keywords{traits, wood density, growth rates, bayesian, survival analysis}
+
+<<preload, echo=FALSE>>=
+n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
+n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
+proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique))
+n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
+obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
+@
+
+\begin{document}
+\mstitlepage
+\noindent
+\parindent=1.5em
+\addtolength{\parskip}{.3em}
+\doublespacing
+\linenumbers
+
+\section{Summary}\label{abstract}
+
+\section{Introduction}
+Plant mortality is often modelled as a function of recent growth because it is thought to reflect the summed effects of many biotic and abiotic factors on a plants overall carbon budget \citep{Kobe:1995tw, Hawkes:2000ib, Keane:2001db, Wyckoff:2002ul}.
+As such, growth rates are considered a highly integrated individual-based predictor of plant mortality \citep{Hawkes:2000ib}.
 Within a species, the assumption of the growth-mortality relationship is that fast growing individuals (i.e.
 those that have larger carbon budgets) are more able to buffer stress and thus have higher survivorship relative to slow growing (i.e.
-low carbon budget) individuals [@Hawkes:2000ib; @Keane:2001db].
-Empirical studies examining this growth-mortality relationship have broadly supported this assumption, with many highlighting non-linear declines in mortality with increased growth [e.g.
-@Kobe:1997vy; @Wyckoff:2002ul; @Vieilledent:2010fv].
-Consequently, most dynamic vegetation models, including tree gap models [@Keane:2001db], individual-based forest simulators [e.g.
-SORTIE; @Pacala:1996vm] and Dynamic Global Vegetation Models [DGVMs; @Woodward:2004ft] use some measure of growth coupled with a stochastic constant (i.e.
-baseline hazard or minimum chance of death) as the fundamental determinants of plant mortality [@Hawkes:2000ib; @McDowell:2011dr].
+low carbon budget) individuals \citep{Hawkes:2000ib, Keane:2001db}.
+Empirical studies examining this growth-mortality relationship have broadly supported this assumption, with many highlighting non-linear declines in mortality with increased growth \citep[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
+Consequently, most dynamic vegetation models, including tree gap models \citep{Keane:2001db}, individual-based forest simulators \citep[e.g.
+SORTIE][]{Pacala:1996vm} and Dynamic Global Vegetation Models \citep[DGVMs; ][]{Woodward:2004ft} use some measure of growth coupled with a stochastic constant (i.e.
+baseline hazard or minimum chance of death) as the fundamental determinants of plant mortality \citep{Hawkes:2000ib, McDowell:2011dr}.
 
-However, due to a paucity of data on plant mortality, most dynamic vegetation models have implemented mortality algorithms using overly simplistic assumptions [@Hawkes:2000ib; Wyckoff:2000un; @McDowell:2011dr].
-One of the strongest assumptions these models make is that both baseline hazard and the growth-mortality relationship are constant across species [@Pacala:1996vm, @Hawkes:2000ib; Wyckoff:2000un].
+However, due to a paucity of data on plant mortality, most dynamic vegetation models have implemented mortality algorithms using overly simplistic assumptions \citep{Hawkes:2000ib, Wyckoff:2000un, McDowell:2011dr}.
+One of the strongest assumptions these models make is that both baseline hazard and the growth-mortality relationship are constant across species \citep{Pacala:1996vm, Hawkes:2000ib, Wyckoff:2000un}.
 That is, all species have the same random chance of death and exhibit the same tolerance to low growth.
-This is despite growing empirical evidence highlighting that mortality rates vary considerably between species, even within the same community (@Kobe:1997vy; @Wyckoff:2002ul; @Kunstler:2005bn; @Vieilledent:2010fv).
-More recently, however, attempts have been made to examine this inter-specific variation by correlating species intrinsic growth and mortality rates [e.g.
-@Poorter:2008iu; @Wright:2010fl].
-In contrast to intraspecific studies, these studies have shown that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species [Loehle:1996vw; @Poorter:2008iu; @Wright:2010fl].
+This is despite growing empirical evidence highlighting that mortality rates vary considerably between species, even within the same community \citep{Kobe:1997vy, Wyckoff:2002ul, Kunstler:2005bn, Vieilledent:2010fv}.
+More recently, however, attempts have been made to examine this inter-specific variation by correlating species intrinsic growth and mortality rates \citep[e.g.][]{Poorter:2008iu, Wright:2010fl}.
+In contrast to intraspecific studies, these studies have shown that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species \citep{Loehle:1996vw, Poorter:2008iu, Wright:2010fl}.
 As such, species appear to exhibit a trade off between greater growth versus greater survival.
 Ecologists have attempted to explain this interspecific variation by using functional traits - attributes of species that are thought to influence vital rates (i.e.
-survival, growth and reproduction) and ultimately fitness [@Westoby:2002tb; @Ackerly:2003eb; @Falster:2015un].
+survival, growth and reproduction) and ultimately fitness \citep{Westoby:2002tb, Ackerly:2003eb, Falster:2015un}.
 
-Wood density, the biomass invested per unit wood volume, is one functional trait that has been consistently shown to be negatively correlated with tropical tree mortality [@Chave:2009iy; @Poorter:2008iu; @Kraft:2010kq; @Wright:2010fl].
-The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to random events such as stem breakage [@Putz:1983wu; @vanGelder:2006exa; @Chave:2009iy], embolism [@Hacke:2001kj, @Jacobsen:2005fx] and fungal and pathogen attack [@Augspurger:1984wx].
+Wood density, the biomass invested per unit wood volume, is one functional trait that has been consistently shown to be negatively correlated with tropical tree mortality \citep{Chave:2009iy, Poorter:2008iu, Kraft:2010kq, Wright:2010fl}.
+The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to random events such as stem breakage \citep{Putz:1983wu, vanGelder:2006exa, Chave:2009iy}, embolism \citep{Hacke:2001kj, Jacobsen:2005fx} and fungal and pathogen attack \citep{Augspurger:1984wx}.
 As such, it is hypothesized that species with higher wood density are likely to exhibit lower baseline hazards.
 However, research has also shown that wood density may increase tolerance to low growth rates, and thus alter the growth-mortality relationship.
-For example, species with high wood density are more tolerant shade [@vanGelder:2006exa] and competition [@Kunstler:2016km].
-Studies have also shown that wood density may reduce drought induced carbon starvation [@McDowell:2011dr] and the susceptibility of ‘stressed’ individuals to opportunistic invertebrate, pathogen or fungal attack [@Augspurger:1984wx; @McDowell:2011dr].
+For example, species with high wood density are more tolerant shade \citep{vanGelder:2006exa} and competition \citep{Kunstler:2016km}.
+Studies have also shown that wood density may reduce drought induced carbon starvation \citep{McDowell:2011dr} and the susceptibility of ‘stressed’ individuals to opportunistic invertebrate, pathogen or fungal attack \citep{Augspurger:1984wx, McDowell:2011dr}.
 
@@
 However, to date, there has been little crossover of ideas between intraspecific and interspecific studies of plant mortality.
-For example, we are aware of only one study that uses individual based plant mortality empirical models to examine trait effects on baseline hazards [@AubryKientz:2013dg], and none that examine how traits mediate the growth-mortality relationship.
+For example, we are aware of only one study that uses individual based plant mortality empirical models to examine trait effects on baseline hazards \citep{AubryKientz:2013dg}, and none that examine how traits mediate the growth-mortality relationship.
 
 In this study, we reconcile both intraspecific and interspecific theory by developing an individual based plant mortality model that incorporates individual based measurements of growth as well as the species trait wood density.
-Furthermore, we extend existing individual-based approaches [e.g.
-@Kobe:1995tw] by simultaneously estimating both baseline hazards and the growth-mortality relationship (hereon in referred to as growth-dependent hazard).
-We parameterise this model using 15-years of individual dbh growth and mortality data from 203 tropical rainforest species encompassing more than 180,000 individuals.
-Using this model,coupled with cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to the null model or growth-dependent only model (e.g.
-Kobe et al 1996) 2) Does allowing both hazards to vary by species improve predictive capacity?
+Furthermore, we extend existing individual-based approaches \citep[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship (hereon in referred to as growth-dependent hazard).
+We parameterise this model using 15-years of individual dbh growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals.
+Using this model,coupled with cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to the null model or growth-dependent only model 2) Does allowing both hazards to vary by species improve predictive capacity?
 3) Does wood density predominately affect baseline hazards, or does it affect growth-dependent hazards or both?
@@
 
-\pagebreak
+\clearpage
 
+\section{Methods}
+
+\subsection{Data}
+We parameterise plant mortality models using individual growth and survival data collected from a 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\degree;N, 79.85\degree;W).
+The plot is predominately undisturbed old-growth rainforest.
+The climate is warm throughout the year with a mean annual temperature of 26.1\degree;.
+Rainfall is seasonal with most of the 2500 mm falling between April and November.
+The island is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI) and the permanent plot was established in 1980.
+Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in (REFS).
+
+Within the 50-ha BCI plot the diameter at breast height and survival status of all free-standing woody plants at least 1.3 m tall with a diameter greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter \citep{Condit:2012nz}.
+For the purpose of modelling mortality as a function of growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
+We also excluded individuals based on several rules: 1) those that did not survive at least two censuses (two being required to estimate growth rates); 2) those that were not consistently measured at 1.3 m above ground; 3) those that were multi-stemmed; or 4) those that resprouted or '\textit{returned from the dead}'.
+We also discarded species that do not exhibit secondary growth (e.g.
+palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density.
+Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial dbh were also removed.
+Using this dataset, we then randomly selected one census observation per individual.
+This was done in order to reduce computational requirements.
+In total, \Sexpr{n_inds} individuals of \Sexpr{n_spp} species were included in our analysis.
+
+Wood density for each species was estimated by coring trees located within 15 km of the BCI plot because increment borers are prohibited within the plot \citep{Wright:2010fl}.
+Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100\degree C) and dividing by the fresh volume (as measured by water displacement).
+Wood density was then calculated as an area-weighted average (g cm$^-3$), where area refers to the annulus represented by each piece, assuming a circular trunk.
+
+\subsection{Estimating true growth}
+Field measurements almost always include observation error, and these errors can potentially manifest into biologically unrealistic scenarios that may ultimately bias interptations.
+In our dataset, \Sexpr{proportion_negative}\% of the observations were of negative dbh growth.
+While it is known that dbh can decline due to changes in water availability, growth We consider a biologically unrealistic scenario
+
+Here we quantify observation error by using another BCI dataset explicitly collected for this purpose.
+This dataset consistend of \Sexpr{n_obs_errors} randomly selected trees remasured within 30 days (see XXXX).
+By assuming that no growth occurred between the two observations, and that observation error did not vary with size, the standard deviation of measurement discrepencies
+
+explicitly collected to examine  consisting of \Sexpr{n_obs_errors} randomly selected trees remeasured within 30 days (see XXXX).
+By assuming that no growth occurred between the two observations, we estimated the standard deviation of measurement error
+$\sigma_{error}$ was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter.
+We then used this standard deviation of 0.75 in another Bayesian model to estimate the initial and final true dbh (i.e.
+observation - measurement error) for each individual.
+This was done by first modelling the observed initial ($dbh_{t-1}$) and final dbh ($dbh_{t}$) of each individual, $i$, as random realisations from a normal distribution centered on their respective true estimates (i.e.
+$\widehat{dbh1_i}$ and $\widehat{dbh2_i}$), with a standard deviation of 0.75 (i.e.
+$\sigma_{error}$).
+
+\bibliography{refs}
+
+### STILL CLEANING
 # Methods
@@
 
-We fit our model using growth and survival data collected from a 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15&deg;N, 79.85&deg;W).
-The plot is predominately undisturbed old-growth rainforest (48-ha), but 2 ha is of 100 year old secondary rainforest (REF).
+We parameterise our models using individual growth and survival data collected from a 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15&deg;N, 79.85&deg;W).
+The plot is predominately undisturbed old-growth rainforest.
 The climate at Barro Colorado Island is warm throughout the year with a mean annual temperature of 26.1&deg;.
@@
 The island is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI), which was granted long-term custodianship over the island by Panama's Environmental Authority.
-STRI have permission to establish the 50-ha plot as a permanent census in 1980.
+STRI establish the 50-ha plot as a permanent census in 1980.
 Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in (REFS).
@@
 For the purpose of this study, we discarded data collected in 1982 and 1985 censuses because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
-We excluded individual trees that did not survive at least two censuses (two being required to estimate growth rates), were not consistently measured at 1.3 m, were multi-stemmed, or had re-sprouted or 'returned from the dead'.
+We excluded individual trees that did not survive at least two censuses (two being required to estimate growth rates), were not consistently measured at 1.3 m above ground, were multi-stemmed, or had resprouted or 'returned from the dead'.
 We also discarded species that do not exhibit secondary growth (e.g.
-palms and ferns), contained fewer than 10 individuals or did not contain a wood density estimate were also discarded.
+palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density.
 Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial dbh were also removed.
-Lastly, in order to reduce computational requirements we randomly selected one observation per individual.
+Using this dataset, we then randomly selected one observation per individual.
+This was done in order to reduce computational requirements.
 In total, 180,500 individuals of 203 species were included in our analysis.
@@
 We estimate this measurement error using another BCI dataset consisting of 1562 randomly selected trees remeasured within 30 days (see XXXX).
-Assuming that no growth occurred between observations, we estimated the standard deviation of measurement error (and its uncertainty) by sampling the discrepencies for each individual, $i$, as random realisations from a normal distribution centered on zero with a standard deviation, $\sigma_{error}$, which describes the variation in error:
-
-\begin{equation} \label{eq:error_model}
-discepency_i \sim \textrm{N}(0, \sigma_{error}). \end{equation}
-
-$\sigma_{error}$ was estimated to be 0.75 (sd = 0.01), meaning that 95% of the measurement errors were within +/- 1.47 cm of the true diameter.
+By assuming that no growth occurred between the two observations, we estimated the standard deviation of measurement error
+$\sigma_{error}$ was estimated to be 0.75 (sd = 0.01), meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter.
 We then used this standard deviation of 0.75 in another Bayesian model to estimate the initial and final true dbh (i.e.
@@
 \pagebreak
-
-## References
+\end{document}

### 2016-01-25 ms/ms.tex
Created manuscript folder

--- 2016-01-25 ms/ms.Rnw
@@
-\documentclass[a4paper,11pt]{article}
+\documentclass[a4paper,11pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage[osf]{mathpazo}
@@
 
-<<preload, echo=FALSE>>=
-n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
-n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
-proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique))
-n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
-obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
-@
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 Furthermore, we extend existing individual-based approaches \citep[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship (hereon in referred to as growth-dependent hazard).
-We parameterise this model using 15-years of individual dbh growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals.
+We parameterise this model using 15-years of individual dbh growth and mortality data from 203 tropical rainforest species encompassing 180500 individuals.
 Using this model,coupled with cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to the null model or growth-dependent only model 2) Does allowing both hazards to vary by species improve predictive capacity?
@@
 This was done in order to reduce computational requirements.
-In total, \Sexpr{n_inds} individuals of \Sexpr{n_spp} species were included in our analysis.
+In total, 180500 individuals of 203 species were included in our analysis.
 
@@
 \subsection{Estimating true growth}
-Field measurements almost always include observation error, and these errors can potentially manifest into biologically unrealistic scenarios that may ultimately bias interptations.
-In our dataset, \Sexpr{proportion_negative}\% of the observations were of negative dbh growth.
-While it is known that dbh can decline due to changes in water availability, growth We consider a biologically unrealistic scenario
-
-Here we quantify observation error by using another BCI dataset explicitly collected for this purpose.
-This dataset consistend of \Sexpr{n_obs_errors} randomly selected trees remasured within 30 days (see XXXX).
-By assuming that no growth occurred between the two observations, and that observation error did not vary with size, the standard deviation of measurement discrepencies
-
-explicitly collected to examine  consisting of \Sexpr{n_obs_errors} randomly selected trees remeasured within 30 days (see XXXX).
+Field measurements almost always include observation error that can potentially manifest into biologically unrealistic scenarios of negative growth and ultimately biased conclusions \citep{Kery:2010tp}.
+In our dataset 0\% of observed growth rates were negative.
+We estimate this measurement error using another BCI dataset consisting of  randomly selected trees remeasured within 30 days (see XXXX).
 By assuming that no growth occurred between the two observations, we estimated the standard deviation of measurement error
-$\sigma_{error}$ was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter.
+$\sigma_{error}$ was estimated to be 0.75, meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter.
 We then used this standard deviation of 0.75 in another Bayesian model to estimate the initial and final true dbh (i.e.
@@
 
-### STILL CLEANING
-# Methods
-
-## Data
-
-We parameterise our models using individual growth and survival data collected from a 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15&deg;N, 79.85&deg;W).
-The plot is predominately undisturbed old-growth rainforest.
-The climate at Barro Colorado Island is warm throughout the year with a mean annual temperature of 26.1&deg;.
-Rainfall is seasonal with most of the 2500 mm falling between April and November (REF).
-The island is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI), which was granted long-term custodianship over the island by Panama's Environmental Authority.
-STRI establish the 50-ha plot as a permanent census in 1980.
-Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in (REFS).
-
-Within the 50-ha BCI plot the diameter at breast height and survival status of all free-standing woody plants with a diameter greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter [@Condit:2012nz].
-For the purpose of this study, we discarded data collected in 1982 and 1985 censuses because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
-We excluded individual trees that did not survive at least two censuses (two being required to estimate growth rates), were not consistently measured at 1.3 m above ground, were multi-stemmed, or had resprouted or 'returned from the dead'.
-We also discarded species that do not exhibit secondary growth (e.g.
-palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density.
-Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial dbh were also removed.
-Using this dataset, we then randomly selected one observation per individual.
-This was done in order to reduce computational requirements.
-In total, 180,500 individuals of 203 species were included in our analysis.
-
-Wood density for each species was estimated by coring trees located within 15 km of the BCI plot because increment borers are prohibited within the plot [@Wright:2010fl].
-Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100&deg;C) and dividing by the fresh volume (as measured by water displacement).
-Wood density was then calculated as an area-weighted average (g cm$^-3$), where area refers to the annulus represented by each piece, assuming a circular trunk.
-
-## Measurement error and calculating true growth
-
-Observed dbh measurements are likely to include measurement error that can manifest into biologically unrealistic scenarios of negative growth and ultimately erroneous conclusions.
-We estimate this measurement error using another BCI dataset consisting of 1562 randomly selected trees remeasured within 30 days (see XXXX).
-By assuming that no growth occurred between the two observations, we estimated the standard deviation of measurement error
-$\sigma_{error}$ was estimated to be 0.75 (sd = 0.01), meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter.
-We then used this standard deviation of 0.75 in another Bayesian model to estimate the initial and final true dbh (i.e.
-observation - measurement error) for each individual.
-This was done by first modelling the observed initial ($dbh_{t-1}$) and final dbh ($dbh_{t}$) of each individual, $i$, as random realisations from a normal distribution centered on their respective true estimates (i.e.
-$\widehat{dbh1_i}$ and $\widehat{dbh2_i}$), with a standard deviation of 0.75 (i.e.
-$\sigma_{error}$).
-
-\begin{equation} \label{eq:obs dbh1}
-dbh_{i,t-1} \sim \textrm{N}(\widehat{dbh_{i,t-1}}, 0.75). \end{equation}
-
-\begin{equation} \label{eq:obs dbh2}
-dbh_{i,t} \sim \textrm{N}(\widehat{dbh_{i,t}}, 0.75). \end{equation}
-
-$\widehat{dbh_{i,t-1}}$ was estimated from a lognormal distribution:
-
-\begin{equation} \label{eq:true dbh1}
-\widehat{dbh_{i,t-1}} \sim \textrm{lognormal}(\mu_{dbh_{i,t-1}}, \sigma_{dbh_{t-1}}), \end{equation}
-
-where, $\mu_{dbh_{t-1}}$ defines the log of the grand mean initial dbh, and $\sigma_{dbh_{t-1}}$ defines that variation around that mean.
-$\mu_{dbh_{t-1}}$ was estimated using a normal prior $\text{N}(0, 2.5)$, which represents our prior expectation that the mean initial dbh would likely occur between 0 and 148 cm.
-$\sigma_{dbh{t-1}}$ was estimated using a weakly informative half cauchy prior centered on zero and a scale parameter of 2.5.
-
-$\widehat{dbh_{i,t}}$ was estimated as:
-
-\begin{equation} \label{eq:true dbh2}
-\widehat{dbh_{i,t-1}} + R_i \, T_i, 
-\end{equation}
-
-where, $R_i$ represents the true annual increase in dbh (cm) and $T_i$ represents the number of days between measurements divided by 365.25.
-We modelled $R_i$ as a random realisation from a half normal distribution centered on zero and a standard deviation of 5, conveying our prior expectation that annual increases in dbh would be between 0 and 10 cm/yr.
-We used a half normal as it ensures dbh growth cannot be negative.
-
-\begin{equation} \label{eq:obs dbh increment}
-R_i \sim \textrm{N+}(0, 5). 
-\end{equation}
-
-For computational efficiency, we used the posterior means of $dbh_{t-1}$, $dbh_t$ to calculate:
-
-dbh growth rate:  $\frac{\widehat{dbh_{i,t}} - \widehat{dbh_{i,t-1}}}{T_i}$ 
-basal area growth rate: $\frac{(\pi \, 0.25 \,\widehat{dbh_{i,t}}) - (\pi \, 0.25 \, \widehat{dbh_{i,t-1}})}{T_i}$.
-
-These estimates of individual post growth were then used in the subsequent mortality analysis.
-
-## Mortality model description
-
-We extend a common functional form used to model the growth-mortality relationship [e.g. @Kobe:1995tw] to include a growth-independent baseline hazard, such that the functional form becomes
-\begin{equation} \label{eq:lambda}
-\lambda = \overbrace{\alpha \, \exp\left(-\beta \, G \right)}^{\text{Growth dependent}} + \overbrace{\gamma}^{\text{Growth independent}}
-\end{equation}
-
-By adding $\gamma$ we allow the functional form to asymptote at a mortality rate above zero, which allows for deaths to occur at high growth rates that would otherwise not be accounted for.
-The advantage of this functional form is that all parameters are biologically interpretable.
-$\alpha$ represents the intercept and biologically defines the sum of mortality rate at zero growth and baseline hazard.
-$\beta$ represents the function's decay, or in other words, the sensitivity of mortality rates to change with growth rate.
-$\gamma$, as stated before, represents asymptote and is referred to as the baseline hazard.
-$\alpha$ and $\beta$ are intrinsically linked to growth rate, and as such are growth-dependent parameters.
-By contrast, $\gamma$ accounts for mortality not related to growth caused by stachastic events (e.g.
-windfall, fire) and consequently is a growth-independent parameter.
-
-To incorporate varability in species mortality rates, we add allowed $\alpha$, $\beta$ and $\gamma$ to vary by wood density - a species trait thought to encapsulate a species growth strategy (REFs).
-In addition to this, we also allowed the three parameters to vary by species ID, by doing this, we effectively allow the model to partition variation explained by wood density from the residual variation explained by species and individual levels [@Gelman:2007te; @Kery:2010tp].
-
-## Bayesian Hierarchical Model implementation
-
-We have binary observations $y_i$ (0 = alive, 1 = dead) for each individual $i$.
-We also have a hypothesised underlying mortality model giving an instantaneous rate of mortality (or hazard) per unit time (eq.
-\ref{eq:lambda}).
-Based on standard demographic theory, the probability of death for individual $i$ across the census interval $t_1\rightarrow t_2$ is
-\begin{equation} \label{eq:pred_model}
-p_{i,t_1\rightarrow t_2} = 1 - \exp\left(\Lambda_{i,t_1\rightarrow t_2} \right),
-\end{equation}
-where $\Lambda_{i,t} = \int_{t1}^{t2} \lambda(t) \, dt$ is the cumulative hazard for that individual.
-Given estimates of $\lambda$, one can model the observations as a random realization from a Bernoulli distribution based on the probability that individual $i$ would die, $p_{i,t_1\rightarrow t_2}$:
-\begin{equation} \label{eq:obs_model}
-y_{i,t_1\rightarrow t_2} \sim \textrm{bernoulli}(p_{i,t_1\rightarrow t_2}).\end{equation}
-
-Our first hypothesis is that the cumulative hazard rate within a census interval is reasonably approximated by the true growth rate (estimated as above) in the previous interval
-\begin{equation} \label{eq:hazard2}
-\Lambda_{i,t} \approx (t_2-t_1) \lambda (t_1).
-\end{equation}
-
-Second, we assume that the three parameters of eq.
-\ref{eq:lambda} are influenced by 3 factors: wood density, species ID and census year.
-Because wood density is measured at species-level, its effect is logically a subset of the overall species level effect.
-
-To account for species strategy and partition variance into explained and unexplained at the observation and species level we modelled $\alpha$, $\beta$ and $\gamma$ via the following submodels:
-\begin{equation} \label{eq:submodels}
-\alpha_s = \alpha_{0,s} \, \rho_s^{\alpha_{1}},\\
-\beta_s = \beta_{0,s} \, \rho_s^{\alpha_{1}}, \\
-\gamma_s = \gamma_{0,s} \, \rho_s^{\alpha_{1}}. 
-\end{equation}
-
-Here $\alpha_{0,s}$,$\beta_{0,s}$,$\gamma_{0,s}$ represents the species-level random effect of species $s$ within which individual $i$ belongs and $\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of standardised wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively.
-The form of submodels in eq \ref{eq:submodels} ensures that parameters remain positive.
-On a log scale these submodels \ref{eq:submodels} become additive (e.g.
-$\log \alpha_s = \log \alpha_{0,s} + \alpha_{1} \, \log \rho_s$, equivalent to a classic proportional hazards model widely used in survival analysis.
-We standardised wood density via the transformation: $\rho_s = \frac{\rho}{0.6}$), an operation which is equivalent to centering a log-transformed variable at a density of 0.6 g cm${^-3}$.
-This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant) [@Gelman:2007te].
-
-The species random effects on each parameter were assumed to be a random realistation from a log-normal distribution:
-\begin{equation} \label{eq:RE_species}
-\alpha_{0,s} \sim \text{log-normal}\left(\mu_{\log \alpha_0}, \sigma_{\log \alpha_0}\right),\\
-\beta_{0,s} \sim \text{log-normal}\left(\mu_{\log \beta_0}, \sigma_{\log \beta_0}\right),\\
-\alpha_{0,s} \sim \text{log-normal}\left(\mu_{\log \gamma_0}, \sigma_{\log \gamma_0}\right),\\
-\end{equation}
-
-We estimated $\mu_{\log \alpha_0}$, the average log species effect at zero growth using the following informative prior:
-
-\begin{equation} \label{eq:alpha0}
-\mu_{\log \alpha_{0}} \sim \text{normal}\left(-0.61, 0.57\right).\\
-\end{equation}
-
-This distribution was derived from data published by Kobe et al 1995, who estimated the probability of dying at zero growth/2.5 years for juveniles of 10 hardwood species across north-eastern America.
-Note that these data are also used to define the growth-mortality relationship in SORTIE-ND a spatially explicit forest dynamic model (REF).
-We converted these probabilities/2.5 years to instantaneous hazard rates by using $\lambda = -\log(1 - Pr(Death))/2.5$, and then log transforming these values and calculating the mean (i.e.
--0.61) and standard error (i.e.
-0.57).
-Based on this prior, we assumed that the average probability of death was between 16 and 81% with the central tendency being 42%/yr.
-
-Similarly, we used prior information from Korning and Balslev et al (1994) to estimate, $\mu_{\log \gamma_0}$, the average log species effect on $\gamma$, such that:
-
-\begin{equation} \label{eq:gamma0}
-\mu_{\log \gamma_0} \sim \textrm{N}\left(-3.86, 0.2\right).\\
-\end{equation}
-
-This distribution was estimated the same way as in $\mu_{\log \alpha_0}$.
-That is we converted annual mortality probabilities of 17 tropical lowland tree species in Amazonian Ecuador that encompass the range of forest structure (i.e.
-understory species, small stemmed midcanopy species, large stemmed midcanopy species and canopy species) to instantaneous hazard rate, and then log transformed these and calculated the mean (i.e.
--3.86) and standard error (i.e.
-0.2) These annual mortality estimates were averaged over size and growth rates, and as such are likely to be an overestimate of the true value of $\gamma$.
-However, based on this prior, we assume that the average growth-independent species mortality will be 2.1%/yr with 95% of species growth-independent probabilities of occuring between 1.4% and 3.1%.
-
-As we fitted our model to two growth measures (i.e. dbh growth and basal area growth), we used weak normal priors on $\mu_{\log \beta_0}$, the average log species decay rate where:
-
-\begin{equation} \label{eq:beta_0}
-\mu_{\log \beta_0} \sim \text{N}\left(0, 5\right).\\
-\end{equation}
-
-This prior, regardless of growth measure, spans a range such that the decay of the negative exponential can be rapid or is flat, suggesting no effect of growth rate on mortality rates.
-
-The effect of wood density on all three parameters (i.e. $\alpha_1$, $\beta_1$ and $\gamma_1$) were given normal priors centered on zero and a standard deviation of 5:
-
-\begin{equation} \label{eq:a1b1c1}
-\alpha_1, \beta_1, \gamma_1 \sim \text{N}\left(0, 5\right).\\
-\end{equation}
-
-Standard deviations associated with species random effects (i.e.
-$\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$), were all sampled from positive half cauchy priors with centered on zero with a scale of 2.5.
-
-\begin{equation} \label{eq:sigma}
-\sigma_{a0},\sigma_{b0},\sigma_{c0} \sim \text{cauchy+}\left(0, 2.5\right).\\
-\end{equation}
-
-## Cross validation
-
-To implement K-fold cross-validation in Stan we need to repeatedly fit the model to one subset of the data and use it to predict the held-out set.
-The way we recommend setting this up is to do the partitioning in R (or Python, or whatever data-processing environment is being used) and then pass the training data and held-out data to Stan in two pieces.
-With linear regression, for example, we start with the Stan model on page 4, passing in the training data as N, y, X and the held-out set as N holdout, y holdout, X holdout, with the data block augmented accordingly.
-We then keep the data block and model as is, and just alter the generated quantities block to operate on the held-out data.
-
-For a model $M$, training data $D$, and test data $D^*$, we want to calculate
-
-$$p(D^* | D,M) = \int p(\theta | D, M) \, p(D^* | \theta, M) \, d \theta$$
-
-The distribution $p(\theta | D, M)$ gives the likelihood of parameter values $\theta$ given the training data (obtained from fitting the model) and is calculated (via Bayes Rule) as
-
-$$p(\theta | D, M) = \frac{p(\theta) \, p(D|\theta)}{p(D)}.$$
-
-$p(D^* | \theta, M)$ is then the probability of observing the test data given a particular parameter set (i.e the likelihood of the data) and is calculated as
-$$p(D^* | \theta, M) = \prod_i p(D^*_i | \theta, M).$$
-
-Log-likelihoods are easier to deal with and so we have
-
-$$\log p(D^* | \theta, M) = \sum_i \log p(D^*_i | \theta, M).$$
-
-Aki \& Gelman 2014 provide methods for estimating "an importance-sampling approximated LOO directly using the log-likelihood evaluated at the posterior simulations of the parameter values"
-
-We fitted these models using package rstan version 2.7 [@StanDevelopmentTeam:2015uz].
-Three independent chains were executed and in all cases modelled parameters converged within 1000 iterations.
-Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic [@Gelman:1992ts; @Brooks:1998ju].
-After discarding the first 1000 iterations as ‘burn in’, a further 1000 iterations were taken from the joint posterior.
-The residuals of the model were checked graphically against predictions with none showing any systematic pattern or severe heteroscedasticity.
-
-\pagebreak
-
-# OTHER STUFF
-To optimise calculations in stan, we tried to minimise the number of logarithmic and exponential transformations, resulting in the following algorithm:
-
-1. Estimate $\alpha, \beta, \gamma$
-  - Draw a number from standard lognormal distribution with location ($\mu$) and scale ($\sigma$) parameters being the mean and sd of logged variable.
-Note that this draw returns variable on non-logged axes, even though location and scale are specified on logged axes.
-It is equivalent to drawing a normal number on log-transformed axes and back transforming.
-$$z \sim \textrm{lognormal}\left(\mu_{\log \alpha_i}, \sigma_{\log \alpha_i}\right)$$
-  - model for $\mu_{\log \alpha_i}$ and $\sigma_{\log \alpha_i}$ stays the same, with priors being mean and variance of log-transformed values.
-2.  Estimate cumulative hazard $\Lambda$ as
-$$\Lambda = -T \, \left(\alpha_0\, \alpha_1 , \rho_s^{\alpha_2} \exp(\beta_0\, \beta_1 , \rho_s^{\beta_2}  * G) + \gamma_0\, \gamma_1 , \rho_s^{\gamma_2}\right)$$
-3. Increment log probability density by appropriate amount.
-  - If $y=0$ we have $$\Pr(y=0) = 1 - \Pr(y=1) = 1 - (1- \exp(-\Lambda) = \exp(-\Lambda),$$ so the log probability density is $$\log \Pr(y=0) = \log(\exp(-\Lambda)) = -\Lambda$$.
- - If $y=1$ we have $\Pr(y=1) = 1 - \exp(-\Lambda)$, so the log probability density is $$\log \Pr(y=1) = \log(1-\exp(-\Lambda)) = \textrm{log1m\_exp}(-\Lambda)$$.
-
-\pagebreak
 \end{document}

### 2016-01-25 R/process_output.R
Forgot to add comments to some functions

@@ -86,13 +86,13 @@ extract_loglik_samples <- function(analysis) {
     select(-modelid)
 }
 
-# Extract
+# Extract log likelihood samples for multiple analyses.
 extract_multi_analysis_loglik_samples <- function(list_of_analyses){
   samples <- lapply(list_of_analyses, extract_loglik_samples)
   plyr::ldply(samples, .id='model') %>%
     select(-model)
 }
-
+# Summarise log likelihood samples
 summarise_loglik_samples <- function(samples) {
   samples %>%
     group_by(analysis,growth_measure,rho_combo, likelihood) %>%

### 2016-01-29 ms/ms.Rnw
Rewrote methods and tweaked introduction

@@
-\documentclass[a4paper,11pt]{article}
+\documentclass[a4paper,12pt]{article}
 \usepackage[osf]{mathpazo}
@@
 \usepackage{gensymb}
+\usepackage{amsmath}
+\usepackage[title, titletoc, toc]{appendix}
 %\raggedright
@@
 
-\usepackage[title, titletoc, toc]{appendix}
-
 \mstype{Research Article}
@@
 n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
-proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique))
+proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
 n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
@@
 
-Wood density, the biomass invested per unit wood volume, is one functional trait that has been consistently shown to be negatively correlated with tropical tree mortality \citep{Chave:2009iy, Poorter:2008iu, Kraft:2010kq, Wright:2010fl}.
-The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to random events such as stem breakage \citep{Putz:1983wu, vanGelder:2006exa, Chave:2009iy}, embolism \citep{Hacke:2001kj, Jacobsen:2005fx} and fungal and pathogen attack \citep{Augspurger:1984wx}.
+Wood density, the biomass invested per unit wood volume, is one functional trait that has been consistently shown to be negatively correlated with tropical tree mortality \citep{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to random events such as stem breakage \citep{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \citep{Hacke:2001kj, Jacobsen:2005fx} and fungal and pathogen attack \citep{Augspurger:1984wx}.
 As such, it is hypothesized that species with higher wood density are likely to exhibit lower baseline hazards.
@@
 
-In this study, we reconcile both intraspecific and interspecific theory by developing an individual based plant mortality model that incorporates individual based measurements of growth as well as the species trait wood density.
+In this study, we reconcile both intraspecific and interspecific theory by developing an individual based models of plant mortality that incorporate individual based measurements of growth as well as the species trait wood density.
 Furthermore, we extend existing individual-based approaches \citep[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship (hereon in referred to as growth-dependent hazard).
@@
 \subsection{Data}
-We parameterise plant mortality models using individual growth and survival data collected from a 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\degree;N, 79.85\degree;W).
-The plot is predominately undisturbed old-growth rainforest.
-The climate is warm throughout the year with a mean annual temperature of 26.1\degree;.
-Rainfall is seasonal with most of the 2500 mm falling between April and November.
-The island is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI) and the permanent plot was established in 1980.
-Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in (REFS).
+We parameterise plant mortality models using individual growth and survival data \citep{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\degree;N, 79.85\degree;W).
+The climate on the island is warm throughout the year with a mean annual temperature of 26.1\degree and rainfall is seasonal with most of the 2500 mm falling between April and November.
+BCI is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI) and the permanent plot was established in 1980.
+Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in .
 
-Within the 50-ha BCI plot the diameter at breast height and survival status of all free-standing woody plants at least 1.3 m tall with a diameter greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter \citep{Condit:2012nz}.
+Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and contained a diameter at breast height greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter \citep{Condit:2012nz}.
 For the purpose of modelling mortality as a function of growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
@@
 Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial dbh were also removed.
-Using this dataset, we then randomly selected one census observation per individual.
-This was done in order to reduce computational requirements.
+In order to reduce computational requirements we then subsampled this dataset by randomly selecting one census observation per individual.
 In total, \Sexpr{n_inds} individuals of \Sexpr{n_spp} species were included in our analysis.
@@
 \subsection{Estimating true growth}
-Field measurements almost always include observation error, and these errors can potentially manifest into biologically unrealistic scenarios that may ultimately bias interptations.
-In our dataset, \Sexpr{proportion_negative}\% of the observations were of negative dbh growth.
-While it is known that dbh can decline due to changes in water availability, growth We consider a biologically unrealistic scenario
 
-Here we quantify observation error by using another BCI dataset explicitly collected for this purpose.
-This dataset consistend of \Sexpr{n_obs_errors} randomly selected trees remasured within 30 days (see XXXX).
-By assuming that no growth occurred between the two observations, and that observation error did not vary with size, the standard deviation of measurement discrepencies
+Field measurements almost always include observation error, errors that may ultimately bias conclusions.
+In our dataset, \Sexpr{proportion_negative}\% of observed growth rates were negative.
+While there are physiological processes (e.g.
+water depletation and bark drop) that can lead to trunk shrinkages overtime \citep{Pastur:2007kl}, it is biologically impossible for growth (i.e.
+the creation of new biomass) to be negative \citep{Clark:2007ve}.
+To ensure our model was not biased by these implausable values we  estimated '\textit{true growth}'(i.e.
+observed growth minus observation error) by first using another BCI dataset collected explicitly to examine obervation errors \citep{Condit:2012nz}.
+This dataset consisted of \Sexpr{n_obs_errors} randomly selected trees that had their dbh measured twice within a 30 day period \citep{Condit:2012nz}.
+Assuming no growth occurred over this period, any discrepency in measurements can be attributable to observation error.
+The standard deviation of these discrepencies was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter.
+Having an estimate of observation error, we then built an empirical Bayesian model to estimate '\textit{true}' initial and final dbh, and subsequently,'\textit{true growth}' for each individual.
+We did this by first modelling the observed initial ($dbh_{t-1}$) and final dbh ($dbh_{t}$) of each individual, $i$, as random realisations from a normal distribution centered on their respective true estimates (i.e.
+$\widehat{dbh1_i}$ and $\widehat{dbh2_i}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e.
+the observed measurement error):
 
-explicitly collected to examine  consisting of \Sexpr{n_obs_errors} randomly selected trees remeasured within 30 days (see XXXX).
-By assuming that no growth occurred between the two observations, we estimated the standard deviation of measurement error
-$\sigma_{error}$ was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter.
-We then used this standard deviation of 0.75 in another Bayesian model to estimate the initial and final true dbh (i.e.
-observation - measurement error) for each individual.
-This was done by first modelling the observed initial ($dbh_{t-1}$) and final dbh ($dbh_{t}$) of each individual, $i$, as random realisations from a normal distribution centered on their respective true estimates (i.e.
-$\widehat{dbh1_i}$ and $\widehat{dbh2_i}$), with a standard deviation of 0.75 (i.e.
-$\sigma_{error}$).
+\begin{equation}\label{eq:true_growth_obs} 
+\begin{aligned}
+dbh_{i,t-1} &\sim \mathcal{N}(\widehat{dbh}_{i,t-1}, 0.75) \\
+dbh_{i,t} &\sim \mathcal{N}(\widehat{dbh}_{i,t}, 0.75)
+\end{aligned} 
+\end{equation}
 
-\bibliography{refs}
-
-### STILL CLEANING
-# Methods
-
-## Data
-
-We parameterise our models using individual growth and survival data collected from a 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15&deg;N, 79.85&deg;W).
-The plot is predominately undisturbed old-growth rainforest.
-The climate at Barro Colorado Island is warm throughout the year with a mean annual temperature of 26.1&deg;.
-Rainfall is seasonal with most of the 2500 mm falling between April and November (REF).
-The island is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI), which was granted long-term custodianship over the island by Panama's Environmental Authority.
-STRI establish the 50-ha plot as a permanent census in 1980.
-Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in (REFS).
-
-Within the 50-ha BCI plot the diameter at breast height and survival status of all free-standing woody plants with a diameter greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter [@Condit:2012nz].
-For the purpose of this study, we discarded data collected in 1982 and 1985 censuses because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
-We excluded individual trees that did not survive at least two censuses (two being required to estimate growth rates), were not consistently measured at 1.3 m above ground, were multi-stemmed, or had resprouted or 'returned from the dead'.
-We also discarded species that do not exhibit secondary growth (e.g.
-palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density.
-Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial dbh were also removed.
-Using this dataset, we then randomly selected one observation per individual.
-This was done in order to reduce computational requirements.
-In total, 180,500 individuals of 203 species were included in our analysis.
-
-Wood density for each species was estimated by coring trees located within 15 km of the BCI plot because increment borers are prohibited within the plot [@Wright:2010fl].
-Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100&deg;C) and dividing by the fresh volume (as measured by water displacement).
-Wood density was then calculated as an area-weighted average (g cm$^-3$), where area refers to the annulus represented by each piece, assuming a circular trunk.
-
-## Measurement error and calculating true growth
-
-Observed dbh measurements are likely to include measurement error that can manifest into biologically unrealistic scenarios of negative growth and ultimately erroneous conclusions.
-We estimate this measurement error using another BCI dataset consisting of 1562 randomly selected trees remeasured within 30 days (see XXXX).
-By assuming that no growth occurred between the two observations, we estimated the standard deviation of measurement error
-$\sigma_{error}$ was estimated to be 0.75 (sd = 0.01), meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter.
-We then used this standard deviation of 0.75 in another Bayesian model to estimate the initial and final true dbh (i.e.
-observation - measurement error) for each individual.
-This was done by first modelling the observed initial ($dbh_{t-1}$) and final dbh ($dbh_{t}$) of each individual, $i$, as random realisations from a normal distribution centered on their respective true estimates (i.e.
-$\widehat{dbh1_i}$ and $\widehat{dbh2_i}$), with a standard deviation of 0.75 (i.e.
-$\sigma_{error}$).
-
-\begin{equation} \label{eq:obs dbh1}
-dbh_{i,t-1} \sim \textrm{N}(\widehat{dbh_{i,t-1}}, 0.75). \end{equation}
-
-\begin{equation} \label{eq:obs dbh2}
-dbh_{i,t} \sim \textrm{N}(\widehat{dbh_{i,t}}, 0.75). \end{equation}
-
-$\widehat{dbh_{i,t-1}}$ was estimated from a lognormal distribution:
-
-\begin{equation} \label{eq:true dbh1}
-\widehat{dbh_{i,t-1}} \sim \textrm{lognormal}(\mu_{dbh_{i,t-1}}, \sigma_{dbh_{t-1}}), \end{equation}
-
-where, $\mu_{dbh_{t-1}}$ defines the log of the grand mean initial dbh, and $\sigma_{dbh_{t-1}}$ defines that variation around that mean.
-$\mu_{dbh_{t-1}}$ was estimated using a normal prior $\text{N}(0, 2.5)$, which represents our prior expectation that the mean initial dbh would likely occur between 0 and 148 cm.
+The initial dbh, $\widehat{dbh}_{i,t-1}$, was sampled as a random realisations from a lognormal distribution:
+\begin{equation} \label{eq:true_dbh1}
+\widehat{dbh}_{i,t-1} \sim \mathcal{LN}(\mu_{dbh_{i,t-1}}, \sigma_{dbh_{t-1}}), 
+\end{equation}
+where, $\mu_{dbh_{t-1}}$ is the expected mean of log initial dbh, and $\sigma_{dbh_{t-1}}$ defines that variation around that mean.
+$\mu_{dbh_{t-1}}$ was estimated using a normal prior $\mathcal{N}(0, 2.5)$, which represents our prior expectation that the mean initial dbh would be greater than zero but less than 150 cm.
 $\sigma_{dbh{t-1}}$ was estimated using a weakly informative half cauchy prior centered on zero and a scale parameter of 2.5.
 
-$\widehat{dbh_{i,t}}$ was estimated as:
+The final dbh measurement, $\widehat{dbh}_{i,t}$ was estimated as:
+\begin{equation}\label{eq:true_dbh2}
+\widehat{dbh}_{i,t} = \widehat{dbh}_{i,t-1} + R_i \, T_i, 
+\end{equation}
+where, $R_i$ represents the annual true increase in dbh (cm) and $T_i$ represents the number of days between measurements divided by 365.25.
+We modelled $R_i$ as a random realisation from a lognormal distribution centered on zero and a standard deviation of 1.
+This prior conveyed our expectation that annual increases in dbh were unlikely to be greater than 7 cm/yr.
+Sampling from a lognormal distribution also ensured growth rates could not be negative.
 
-\begin{equation} \label{eq:true dbh2}
-\widehat{dbh_{i,t-1}} + R_i \, T_i, 
+Using this model we then used \stan's \textit{optimizing} function to maximise the joint posterior and extract point estimates of true initial dbh, $\widehat{dbh}_{i,t-1}$, true final dbh, $\widehat{dbh}_{i,t}$, and true annual dbh growth $R_i$.
+We used annual diameter growth, $R_i$, and annual basal area growth:$\frac{(\pi \, 0.25 \,\widehat{dbh}_{i,t}) - (\pi \, 0.25 \, \widehat{dbh}_{i,t-1})}{T_i}$ in the subsequent mortality analysis.
+
+\subsection{Mortality analysis}
+
+In survival analysis, the probability that an individual dies between $t_1$ and $t_2$, $p_{t_1\rightarrow t_2}$ is defined by:
+\begin{equation} \label{eq:mortality_model}
+p_{t_1\rightarrow t_2} = 1 - \exp\left((t_2-t_1)\,\lambda_i\right)
+\end{equation}
+where, $\lambda$, is the instantaneous hazard rate of death per unit time \citep{Kleinbaum:2012tn}.
+
+Here we examine the performance of three functional forms of $\lambda$ to predict observed patterns of tropical tree mortality.
+The simplest model (eqn: \ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plants growth rate.
+The second model (eqn: \ref{growth_haz} describes a growth-dependent hazard which assumes plants can only die due to a growth related cause.
+In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached.
+This negative exponential relationship has been used in other studies to model growth-dependent mortality \citep[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \citep{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}.
+The third model (eqn: \ref{add_haz}, and potentially the most biologically realistic, is the summation of the two previous models.
+Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at baseline hazard instead of zero.
+The advantage of this model is that it effectively accounts for two independent causes of mortality: that due to growth and that due to chance (i.e.
+stochastic events such as windthrow and fire).
+The parameters of eqn: \ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept (i.e.
+sum of both growth dependent and growth independent mortality at zero growth rate) and biologically defines the instantaneous mortality hazard $\lambda$ at zero growth.
+$\beta$ represents the exponential decay rate or the sensitivity of hazard rates to change with growth rate; and $\gamma$, is the asymptote and defines the baseline hazard.
+$\alpha$ and $\beta$ are intrinsically linked to growth rate, and as such are growth-dependent parameters.
+By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g.
+windfall, fire) that can kill a plant irrespective of its growth rate.
+\begin{align}
+\lambda &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
+\lambda &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
+\lambda &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} + 
+\overbrace{\gamma}^{\text{Baseline hazard}} \label{add_haz}
+\end{align}
+
+A central aim of this study is to identify what is the most predictive parameterisation of $\lambda$.
+To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison.
+Stage 1 involved comparing predictive performance of the three $\lambda$ models described above (eqn: \ref{base_haz,growth_haz,add_haz}).
+In this stage each model was fitted with fixed parameters (i.e.
+parameters were constant across species) and then again by allowing parameters to varying by species.
+This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) the extent to which species hazard rates differ.
+In stage 2 we used the most predictive model from stage 1 to examine whether dbh growth or basal area growth was a better predictor of plant mortality.
+In stage 3 we used the best model from stage 2, but without species random effects, to determine what combination of parameter-wood density effects best predicted tree mortality.
+This stage also allowed us to determine whether wood density primarily affected growth-dependent or growth-independent (i.e.
+baseline) mortality.
+Finally, in stage 4, we fit the best model based on the previous three steps to the entire dataset with species random effects on all parameters.
+We then examined wood density effect sizes and the proportion of species variability that it explained.
+
+\subsection{Model fitting}
+
+Our observations are of survival status $y_i$ (0 = alive, 1 = died) for each individual $i$ between two censuses.
+We model these observations as random realizations from a Bernoulli distribution based on the probability that individual $i$ dies between $t_1$ and $t_2$, $p_{i,t_1\rightarrow t_2}$:
+\begin{equation}
+y_{i,t_1\rightarrow t_2} \sim \textrm{Bernoulli}(p_{i,t_1\rightarrow t_2}).
 \end{equation}
 
-where, $R_i$ represents the true annual increase in dbh (cm) and $T_i$ represents the number of days between measurements divided by 365.25.
-We modelled $R_i$ as a random realisation from a half normal distribution centered on zero and a standard deviation of 5, conveying our prior expectation that annual increases in dbh would be between 0 and 10 cm/yr.
-We used a half normal as it ensures dbh growth cannot be negative.
+We modelled $p_{i,t_1\rightarrow t_2}$ using a complementary log-log link as:
+\begin{equation}
+\log(-\log(1 - p_{i,t_1\rightarrow t_2})) = \log((t_2-t_1)\,\lambda_i)
+\end{equation}.
+This parameterisation allowed for additive hazards such as those in \ref{add_haz} and is equivilant to a competiting risks analysis \citep{Kleinbaum:2012tn}.
 
-\begin{equation} \label{eq:obs dbh increment}
-R_i \sim \textrm{N+}(0, 5). 
+As we have multiple hypothesized functional forms of $\lambda$ (e.g.
+eqn: X ), each containing parameters that may or may not vary by species, we outline the fitting procedure of our most complicated model of which all others are nested within:
+\begin{equation}
+\lambda_i = \overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, Growth_i \right)}^{\text{Growth dependent}} + 
+\overbrace{\gamma_{s[i]}}^{\text{Growth independent}}.
 \end{equation}
 
-For computational efficiency, we used the posterior means of $dbh_{t-1}$, $dbh_t$ to calculate:
+In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ varying by species a function of both wood density (measured at species level) and a species random effect:
+\begin{equation}\label{eq:submodels}
+\begin{aligned}
+\alpha_s &= \alpha_{0,s} \, \rho_s^{\alpha_{1}},\\
+\beta_s &= \beta_{0,s} \, \rho_s^{\alpha_{1}}, \\
+\gamma_s &= \gamma_{0,s} \, \rho_s^{\alpha_{1}}. 
+\end{aligned}
+\end{equation}
+Here $\alpha_{0,s}$,$\beta_{0,s}$,$\gamma_{0,s}$ represents the species-level residual error not explained by wood density for species $s$ within which individual $i$ belongs.
+$\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of standardised wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively.
+The form of submodels in eqn: \ref{eq:submodels} ensures that parameters remain positive.
+On a log scale these submodels become additive (e.g.
+$\log \alpha_s = \log \alpha_{0,s} + \alpha_{1} \, \log \rho_s$.
+We standardised wood density via the transformation: $\rho_s = \frac{\rho}{0.6}$), an operation which is equivalent to centering a log-transformed $\rho$ at a density of 0.6 g cm${^-3}$.
+This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6 cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant).
 
-dbh growth rate:  $\frac{\widehat{dbh_{i,t}} - \widehat{dbh_{i,t-1}}}{T_i}$ 
-basal area growth rate: $\frac{(\pi \, 0.25 \,\widehat{dbh_{i,t}}) - (\pi \, 0.25 \, \widehat{dbh_{i,t-1}})}{T_i}$.
-
-These estimates of individual post growth were then used in the subsequent mortality analysis.
-
-## Mortality model description
-
-We extend a common functional form used to model the growth-mortality relationship [e.g. @Kobe:1995tw] to include a growth-independent baseline hazard, such that the functional form becomes
-\begin{equation} \label{eq:lambda}
-\lambda = \overbrace{\alpha \, \exp\left(-\beta \, G \right)}^{\text{Growth dependent}} + \overbrace{\gamma}^{\text{Growth independent}}
+Species random effects were modelled as random realisations from log-normal distributions:
+\begin{equation}
+\begin{aligned}
+\alpha_{0,s} &\sim \mathcal{LN}\left(\mu_{\log \alpha_0}, \sigma_{\log \alpha_0}\right),\\
+\beta_{0,s} &\sim \mathcal{LN}\left(\mu_{\log \beta_0}, \sigma_{\log \beta_0}\right),\\
+\gamma_{0,s} &\sim \mathcal{LN}\left(\mu_{\log \gamma_0}, \sigma_{\log \gamma_0}\right).
+\end{aligned}
 \end{equation}
 
-By adding $\gamma$ we allow the functional form to asymptote at a mortality rate above zero, which allows for deaths to occur at high growth rates that would otherwise not be accounted for.
-The advantage of this functional form is that all parameters are biologically interpretable.
-$\alpha$ represents the intercept and biologically defines the sum of mortality rate at zero growth and baseline hazard.
-$\beta$ represents the function's decay, or in other words, the sensitivity of mortality rates to change with growth rate.
-$\gamma$, as stated before, represents asymptote and is referred to as the baseline hazard.
-$\alpha$ and $\beta$ are intrinsically linked to growth rate, and as such are growth-dependent parameters.
-By contrast, $\gamma$ accounts for mortality not related to growth caused by stachastic events (e.g.
-windfall, fire) and consequently is a growth-independent parameter.
-
-To incorporate varability in species mortality rates, we add allowed $\alpha$, $\beta$ and $\gamma$ to vary by wood density - a species trait thought to encapsulate a species growth strategy (REFs).
-In addition to this, we also allowed the three parameters to vary by species ID, by doing this, we effectively allow the model to partition variation explained by wood density from the residual variation explained by species and individual levels [@Gelman:2007te; @Kery:2010tp].
-
-## Bayesian Hierarchical Model implementation
-
-We have binary observations $y_i$ (0 = alive, 1 = dead) for each individual $i$.
-We also have a hypothesised underlying mortality model giving an instantaneous rate of mortality (or hazard) per unit time (eq.
-\ref{eq:lambda}).
-Based on standard demographic theory, the probability of death for individual $i$ across the census interval $t_1\rightarrow t_2$ is
-\begin{equation} \label{eq:pred_model}
-p_{i,t_1\rightarrow t_2} = 1 - \exp\left(\Lambda_{i,t_1\rightarrow t_2} \right),
-\end{equation}
-where $\Lambda_{i,t} = \int_{t1}^{t2} \lambda(t) \, dt$ is the cumulative hazard for that individual.
-Given estimates of $\lambda$, one can model the observations as a random realization from a Bernoulli distribution based on the probability that individual $i$ would die, $p_{i,t_1\rightarrow t_2}$:
-\begin{equation} \label{eq:obs_model}
-y_{i,t_1\rightarrow t_2} \sim \textrm{bernoulli}(p_{i,t_1\rightarrow t_2}).\end{equation}
-
-Our first hypothesis is that the cumulative hazard rate within a census interval is reasonably approximated by the true growth rate (estimated as above) in the previous interval
-\begin{equation} \label{eq:hazard2}
-\Lambda_{i,t} \approx (t_2-t_1) \lambda (t_1).
+Both $\mu_{\log \alpha_0}$, the average log species residual error at zero growth, and $\mu_{\log \gamma_0}$, the average baseline log species residual error, from normal distributions centered on zero with a standard deviation of 2.5.
+\begin{equation}
+\mu_{\log \alpha_{0}},\mu_{\log \gamma_0} \sim \mathcal{N}\left(0, 2.5\right).\\
 \end{equation}
 
-Second, we assume that the three parameters of eq.
-\ref{eq:lambda} are influenced by 3 factors: wood density, species ID and census year.
-Because wood density is measured at species-level, its effect is logically a subset of the overall species level effect.
+We estimated $\mu_{\log \beta_0}$, the average species error on the exponential decay rate, using a weak normal prior centered on zero with a standard deviation of 5. 
+\begin{equation}
+\mu_{\log \beta_0} \sim \mathcal{N}\left(0, 5\right).\\
+\end{equation}
+As we were fitting these models to multiple growth measures (i.e.
+dbh increment, basal area growth), we intentially made this prior weak, such that the range of possible values spanned from a rapid exponential decay to no change in mortality with increasing growth.
 
-To account for species strategy and partition variance into explained and unexplained at the observation and species level we modelled $\alpha$, $\beta$ and $\gamma$ via the following submodels:
-\begin{equation} \label{eq:submodels}
-\alpha_s = \alpha_{0,s} \, \rho_s^{\alpha_{1}},\\
-\beta_s = \beta_{0,s} \, \rho_s^{\alpha_{1}}, \\
-\gamma_s = \gamma_{0,s} \, \rho_s^{\alpha_{1}}. 
+Standard deviations associated with species random effects (i.e. $\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$), were estimated using positive half cauchy priors centered on zero with a scale of 2.5.
+\begin{equation}
+\sigma_{a0},\sigma_{b0},\sigma_{c0} \sim \textrm{Cauchy+}\left(0, 2.5\right).\\
+\end{equation}
+For models that did not contain species random effects, the above sigma's were not calculated and only the hyperparameter means (i.e.
+$\mu_{\log \alpha_0}$,$\mu_{\log \beta}$,$\mu_{\log \gamma_0}$) were estimated.
+
+Wood density effects (i.e. $\alpha_1$, $\beta_1$ and $\gamma_1$) were estimated using normal priors centered on zero and a standard deviation of 5:
+\begin{equation} \label{eq:rho_effects}
+\alpha_1, \beta_1, \gamma_1 \sim \mathcal{N}\left(0, 5\right).\\
 \end{equation}
 
-Here $\alpha_{0,s}$,$\beta_{0,s}$,$\gamma_{0,s}$ represents the species-level random effect of species $s$ within which individual $i$ belongs and $\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of standardised wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively.
-The form of submodels in eq \ref{eq:submodels} ensures that parameters remain positive.
-On a log scale these submodels \ref{eq:submodels} become additive (e.g.
-$\log \alpha_s = \log \alpha_{0,s} + \alpha_{1} \, \log \rho_s$, equivalent to a classic proportional hazards model widely used in survival analysis.
-We standardised wood density via the transformation: $\rho_s = \frac{\rho}{0.6}$), an operation which is equivalent to centering a log-transformed variable at a density of 0.6 g cm${^-3}$.
-This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant) [@Gelman:2007te].
+All models were fit using package rstan version 2.8.1 \citep{StanDevelopmentTeam:2015uz}.
+Three independent chains were executed and in all cases modelled parameters converged within 2000 iterations.
+Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \citep{Gelman:1992ts, Brooks:1998ju}.
+After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.
 
-The species random effects on each parameter were assumed to be a random realistation from a log-normal distribution:
-\begin{equation} \label{eq:RE_species}
-\alpha_{0,s} \sim \text{log-normal}\left(\mu_{\log \alpha_0}, \sigma_{\log \alpha_0}\right),\\
-\beta_{0,s} \sim \text{log-normal}\left(\mu_{\log \beta_0}, \sigma_{\log \beta_0}\right),\\
-\alpha_{0,s} \sim \text{log-normal}\left(\mu_{\log \gamma_0}, \sigma_{\log \gamma_0}\right),\\
-\end{equation}
+\subsection{Model evaluation}
 
-We estimated $\mu_{\log \alpha_0}$, the average log species effect at zero growth using the following informative prior:
+We examined predictive accuracy of each model by using K-fold cross-validation.
+This involved randomly partitioning the data into ten folds of X observations, ensuring that each K-fold had at least one observation per species.
+We then fitted models to nine folds and predicted to the tenth fold.
+This was repeated until each fold had been used nine times for model fitting, and once as a heldout dataset for predicting.
+Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy (Stone:1997tg).
 
-\begin{equation} \label{eq:alpha0}
-\mu_{\log \alpha_{0}} \sim \text{normal}\left(-0.61, 0.57\right).\\
-\end{equation}
-
-This distribution was derived from data published by Kobe et al 1995, who estimated the probability of dying at zero growth/2.5 years for juveniles of 10 hardwood species across north-eastern America.
-Note that these data are also used to define the growth-mortality relationship in SORTIE-ND a spatially explicit forest dynamic model (REF).
-We converted these probabilities/2.5 years to instantaneous hazard rates by using $\lambda = -\log(1 - Pr(Death))/2.5$, and then log transforming these values and calculating the mean (i.e.
--0.61) and standard error (i.e.
-0.57).
-Based on this prior, we assumed that the average probability of death was between 16 and 81% with the central tendency being 42%/yr.
-
-Similarly, we used prior information from Korning and Balslev et al (1994) to estimate, $\mu_{\log \gamma_0}$, the average log species effect on $\gamma$, such that:
-
-\begin{equation} \label{eq:gamma0}
-\mu_{\log \gamma_0} \sim \textrm{N}\left(-3.86, 0.2\right).\\
-\end{equation}
-
-This distribution was estimated the same way as in $\mu_{\log \alpha_0}$.
-That is we converted annual mortality probabilities of 17 tropical lowland tree species in Amazonian Ecuador that encompass the range of forest structure (i.e.
-understory species, small stemmed midcanopy species, large stemmed midcanopy species and canopy species) to instantaneous hazard rate, and then log transformed these and calculated the mean (i.e.
--3.86) and standard error (i.e.
-0.2) These annual mortality estimates were averaged over size and growth rates, and as such are likely to be an overestimate of the true value of $\gamma$.
-However, based on this prior, we assume that the average growth-independent species mortality will be 2.1%/yr with 95% of species growth-independent probabilities of occuring between 1.4% and 3.1%.
-
-As we fitted our model to two growth measures (i.e. dbh growth and basal area growth), we used weak normal priors on $\mu_{\log \beta_0}$, the average log species decay rate where:
-
-\begin{equation} \label{eq:beta_0}
-\mu_{\log \beta_0} \sim \text{N}\left(0, 5\right).\\
-\end{equation}
-
-This prior, regardless of growth measure, spans a range such that the decay of the negative exponential can be rapid or is flat, suggesting no effect of growth rate on mortality rates.
-
-The effect of wood density on all three parameters (i.e. $\alpha_1$, $\beta_1$ and $\gamma_1$) were given normal priors centered on zero and a standard deviation of 5:
-
-\begin{equation} \label{eq:a1b1c1}
-\alpha_1, \beta_1, \gamma_1 \sim \text{N}\left(0, 5\right).\\
-\end{equation}
-
-Standard deviations associated with species random effects (i.e.
-$\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$), were all sampled from positive half cauchy priors with centered on zero with a scale of 2.5.
-
-\begin{equation} \label{eq:sigma}
-\sigma_{a0},\sigma_{b0},\sigma_{c0} \sim \text{cauchy+}\left(0, 2.5\right).\\
-\end{equation}
-
-## Cross validation
-
-To implement K-fold cross-validation in Stan we need to repeatedly fit the model to one subset of the data and use it to predict the held-out set.
-The way we recommend setting this up is to do the partitioning in R (or Python, or whatever data-processing environment is being used) and then pass the training data and held-out data to Stan in two pieces.
-With linear regression, for example, we start with the Stan model on page 4, passing in the training data as N, y, X and the held-out set as N holdout, y holdout, X holdout, with the data block augmented accordingly.
-We then keep the data block and model as is, and just alter the generated quantities block to operate on the held-out data.
-
-For a model $M$, training data $D$, and test data $D^*$, we want to calculate
-
-$$p(D^* | D,M) = \int p(\theta | D, M) \, p(D^* | \theta, M) \, d \theta$$
-
-The distribution $p(\theta | D, M)$ gives the likelihood of parameter values $\theta$ given the training data (obtained from fitting the model) and is calculated (via Bayes Rule) as
-
-$$p(\theta | D, M) = \frac{p(\theta) \, p(D|\theta)}{p(D)}.$$
-
-$p(D^* | \theta, M)$ is then the probability of observing the test data given a particular parameter set (i.e the likelihood of the data) and is calculated as
-$$p(D^* | \theta, M) = \prod_i p(D^*_i | \theta, M).$$
-
-Log-likelihoods are easier to deal with and so we have
-
-$$\log p(D^* | \theta, M) = \sum_i \log p(D^*_i | \theta, M).$$
-
-Aki \& Gelman 2014 provide methods for estimating "an importance-sampling approximated LOO directly using the log-likelihood evaluated at the posterior simulations of the parameter values"
-
-We fitted these models using package rstan version 2.7 [@StanDevelopmentTeam:2015uz].
-Three independent chains were executed and in all cases modelled parameters converged within 1000 iterations.
-Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic [@Gelman:1992ts; @Brooks:1998ju].
-After discarding the first 1000 iterations as ‘burn in’, a further 1000 iterations were taken from the joint posterior.
-The residuals of the model were checked graphically against predictions with none showing any systematic pattern or severe heteroscedasticity.
-
+Predictive accuracy was quantified as the mean (over 10 cross-validation replicates) of each of two threshold-independent performance metrics.
+First we compare log likelihoods of heldout data, which define the log probability of observing the data given the model.
+Secondly, in order to examine the model capacity to correctly discriminate between whether an individual survived or not, we used the area under the receiver operating characteristic (ROC) curve (AUC).
+AUC describes the model’s discrimination capacity as the frequency with which it predicts a higher probability of mortality for a randomly selected observed death than for a randomly selected observed survivor.
 \pagebreak
-
-# OTHER STUFF
-To optimise calculations in stan, we tried to minimise the number of logarithmic and exponential transformations, resulting in the following algorithm:
-
-1. Estimate $\alpha, \beta, \gamma$
-  - Draw a number from standard lognormal distribution with location ($\mu$) and scale ($\sigma$) parameters being the mean and sd of logged variable.
-Note that this draw returns variable on non-logged axes, even though location and scale are specified on logged axes.
-It is equivalent to drawing a normal number on log-transformed axes and back transforming.
-$$z \sim \textrm{lognormal}\left(\mu_{\log \alpha_i}, \sigma_{\log \alpha_i}\right)$$
-  - model for $\mu_{\log \alpha_i}$ and $\sigma_{\log \alpha_i}$ stays the same, with priors being mean and variance of log-transformed values.
-2.  Estimate cumulative hazard $\Lambda$ as
-$$\Lambda = -T \, \left(\alpha_0\, \alpha_1 , \rho_s^{\alpha_2} \exp(\beta_0\, \beta_1 , \rho_s^{\beta_2}  * G) + \gamma_0\, \gamma_1 , \rho_s^{\gamma_2}\right)$$
-3. Increment log probability density by appropriate amount.
-  - If $y=0$ we have $$\Pr(y=0) = 1 - \Pr(y=1) = 1 - (1- \exp(-\Lambda) = \exp(-\Lambda),$$ so the log probability density is $$\log \Pr(y=0) = \log(\exp(-\Lambda)) = -\Lambda$$.
- - If $y=1$ we have $\Pr(y=1) = 1 - \exp(-\Lambda)$, so the log probability density is $$\log \Pr(y=1) = \log(1-\exp(-\Lambda)) = \textrm{log1m\_exp}(-\Lambda)$$.
-
-\pagebreak
+\bibliography{refs}
 \end{document}

### 2016-01-29 ms/ms.Rnw
Fixed a few typos

@@
 obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
+n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))
 @
@@
 Here we examine the performance of three functional forms of $\lambda$ to predict observed patterns of tropical tree mortality.
-The simplest model (eqn: \ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plants growth rate.
+The simplest model (eqn:\ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plants growth rate.
 The second model (eqn: \ref{growth_haz} describes a growth-dependent hazard which assumes plants can only die due to a growth related cause.
@@
 This negative exponential relationship has been used in other studies to model growth-dependent mortality \citep[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \citep{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}.
-The third model (eqn: \ref{add_haz}, and potentially the most biologically realistic, is the summation of the two previous models.
+The third model (eqn:\ref{add_haz}, and potentially the most biologically realistic, is the summation of the two previous models.
 Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at baseline hazard instead of zero.
@@
 stochastic events such as windthrow and fire).
-The parameters of eqn: \ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept (i.e.
+The parameters of eqn:\ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept (i.e.
 sum of both growth dependent and growth independent mortality at zero growth rate) and biologically defines the instantaneous mortality hazard $\lambda$ at zero growth.
@@
 To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison.
-Stage 1 involved comparing predictive performance of the three $\lambda$ models described above (eqn: \ref{base_haz,growth_haz,add_haz}).
+Stage 1 involved comparing predictive performance of the three $\lambda$ models described in equations \ref{base_haz}, \ref{add_haz}.
 In this stage each model was fitted with fixed parameters (i.e.
@@
 \begin{equation}
-\log(-\log(1 - p_{i,t_1\rightarrow t_2})) = \log((t_2-t_1)\,\lambda_i)
-\end{equation}.
-This parameterisation allowed for additive hazards such as those in \ref{add_haz} and is equivilant to a competiting risks analysis \citep{Kleinbaum:2012tn}.
+\log(-\log(1 - p_{i,t_1\rightarrow t_2})) = \log((t_2-t_1)\,\lambda_i).
+\end{equation}
+This parameterisation allowed for additive hazards such as those in eqn:\ref{add_haz} and is equivilant to a competiting risks analysis \citep{Kleinbaum:2012tn}.
 
 As we have multiple hypothesized functional forms of $\lambda$ (e.g.
-eqn: X ), each containing parameters that may or may not vary by species, we outline the fitting procedure of our most complicated model of which all others are nested within:
+eqn:\ref{base_haz}, \ref{add_haz}), each containing parameters that may or may not vary by species, we outline the fitting procedure of our most complicated model of which all others are nested within:
 \begin{equation}
@@
 $\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of standardised wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively.
-The form of submodels in eqn: \ref{eq:submodels} ensures that parameters remain positive.
+The form of submodels in eqn:\ref{eq:submodels} ensures that parameters remain positive.
 On a log scale these submodels become additive (e.g.
@@
 We examined predictive accuracy of each model by using K-fold cross-validation.
-This involved randomly partitioning the data into ten folds of X observations, ensuring that each K-fold had at least one observation per species.
+This involved randomly partitioning the data into ten folds of \Sexpr{n_heldout} observations, ensuring that each K-fold had at least one observation per species.
 We then fitted models to nine folds and predicted to the tenth fold.
 This was repeated until each fold had been used nine times for model fitting, and once as a heldout dataset for predicting.
-Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy (Stone:1997tg).
+Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy \citep{Stone:1997tg}.
 

### 2016-02-02 R/process_output.R
Fixed log likelihood summarising function and added some plot functions for preliminary results

@@ -82,24 +82,93 @@ extract_loglik_samples <- function(analysis) {
   res <- plyr::ldply(lapply(samples, function(x) {
     gather(data.frame(x),'likelihood','estimate')}), .id='modelid')
   
-  left_join(info, res, 'modelid') %>%
-    select(-modelid)
+  left_join(info, res, 'modelid')
 }
 
 # Extract log likelihood samples for multiple analyses.
 extract_multi_analysis_loglik_samples <- function(list_of_analyses){
   samples <- lapply(list_of_analyses, extract_loglik_samples)
-  plyr::ldply(samples, .id='model') %>%
-    select(-model)
+  plyr::ldply(samples, .id='model')
 }
 # Summarise log likelihood samples
 summarise_loglik_samples <- function(samples) {
   samples %>%
-    group_by(analysis,growth_measure,rho_combo, likelihood) %>%
-    summarise(mean = mean(estimate),
-              `2.5%` = quantile(estimate, 0.025),
-              `10%` = quantile(estimate, 0.1),
-              `50%` = quantile(estimate, 0.5),
-              `90%` = quantile(estimate, 0.9),
-              `97.5%` = quantile(estimate, 0.975))
+    group_by(modelid, model, analysis,growth_measure,rho_combo,kfold, likelihood) %>%
+    summarise(klpd = mean(log_sum_exp(estimate))) %>%
+    ungroup() %>%
+    group_by(model, analysis, growth_measure, rho_combo, likelihood) %>%
+    summarise(elpd = mean(klpd),
+              st_err = sd(klpd)/sqrt(n())) %>%
+    mutate(ci = 1.96 * st_err,
+           upper_ci = elpd + ci,
+           lower_ci = elpd - ci) %>%
+    ungroup()
 }
+
+coefficent_plot_theme <- function() {
+  theme_classic() + theme(axis.title.x = element_text(face = "bold", size = 12),
+                          axis.title.y = element_blank(),
+                          axis.text.x = element_text(size = 10),
+                          axis.text.y = element_text(size = 12),
+                          plot.title = element_text(size = 18),
+                          title = element_text(face = "bold"),
+                          panel.margin = unit(4,"mm"))
+}
+
+test2 <-subset(test, model %in% c('rho_combinations','no_gamma_model', 'null_model') & likelihood =='sum_log_lik_heldout' & rho_combo=='' & growth_measure=='true_dbh_dt')
+
[…]

### 2016-02-02 R/model.R
Changed priors to work better across growth measures. Also added heldout loglik to monitoring list so we can calculate AUC #95. Also cleaned a few messy bits of code

@@ -1,11 +1,11 @@
 
-## NULL MODEL
+## baseline hazard model
 
-get_model_chunks_null <- function(tasks) {
+get_model_chunks_base_haz <- function(tasks) {
   
   list(
     growth_measure = tasks$growth_measure,
-    pars = c("log_hazard",
+    pars = c("gamma","log_lik_heldout",
              "sum_log_lik_fit","sum_log_lik_heldout"),
     data = "
         int<lower=1> n_obs;
@@ -15,22 +15,14 @@ get_model_chunks_null <- function(tasks) {
         // Held out data
         int<lower=1> n_obs_heldout;
         int<lower=0, upper=1> y_heldout[n_obs_heldout];
-        vector[n_obs_heldout] census_length_heldout;
-      ",
+        vector[n_obs_heldout] census_length_heldout;",
     parameters ="
-      // Mortality model parameters
-      real log_hazard;",
+      real<lower=0> gamma;",
     model ="
-      // Declaring mortality parameters
-      real hazard;
       real cumulative_hazard;
       
-      // Calculate fixed effects on normal scale
-      hazard <- exp(log_hazard);
-      
       for (i in 1:n_obs) {
-        // Likelihood for hazard model
-        cumulative_hazard <- -census_length[i] * hazard;
+        cumulative_hazard <- -census_length[i] * gamma;
         
         if (y[i] == 0) {
           increment_log_prob(cumulative_hazard);
@@ -38,14 +30,9 @@ get_model_chunks_null <- function(tasks) {
           increment_log_prob(log1m_exp(cumulative_hazard));
         }
       }
-      
       // Priors
-      
-      //Mortality model priors
-      log_hazard ~ normal(-3.86, 0.2);",
+      gamma ~ lognormal(0, 1);",
     generated_quantities ="
-      // Declaring fitted parameters
-      real hazard;
       real cumulative_hazard_fit;
       real log_lik_fit;
       real sum_log_lik_fit;
@@ -59,13 +46,10 @@ get_model_chunks_null <- function(tasks) {
       sum_log_lik_fit <- 0;
       sum_log_lik_heldout <- 0;
       
-      // recalulate fixed effects
-      hazard <- exp(log_hazard);
-      
       // log likelihood for fitted model
       for (i in 1:n_obs) {
[…]

### 2016-04-14 ms/ms.Rnw
Fixed typo
Changes to manuscript

@@
 \section{Introduction}
-Plant mortality is often modelled as a function of recent growth because it is thought to reflect the summed effects of many biotic and abiotic factors on a plants overall carbon budget \citep{Kobe:1995tw, Hawkes:2000ib, Keane:2001db, Wyckoff:2002ul}.
+Plant mortality is often modelled as a function of recent growth because it reflects the product of many biotic and abiotic factors on a plants overall carbon budget \citep{Kobe:1995tw, Hawkes:2000ib, Keane:2001db, Wyckoff:2002ul}.
 As such, growth rates are considered a highly integrated individual-based predictor of plant mortality \citep{Hawkes:2000ib}.
@@
 However, due to a paucity of data on plant mortality, most dynamic vegetation models have implemented mortality algorithms using overly simplistic assumptions \citep{Hawkes:2000ib, Wyckoff:2000un, McDowell:2011dr}.
-One of the strongest assumptions these models make is that both baseline hazard and the growth-mortality relationship are constant across species \citep{Pacala:1996vm, Hawkes:2000ib, Wyckoff:2000un}.
+One of the strongest assumptions these models make is that both stochastic hazards and the growth-mortality relationship are constant across species \citep{Pacala:1996vm, Hawkes:2000ib, Wyckoff:2000un}.
 That is, all species have the same random chance of death and exhibit the same tolerance to low growth.
 This is despite growing empirical evidence highlighting that mortality rates vary considerably between species, even within the same community \citep{Kobe:1997vy, Wyckoff:2002ul, Kunstler:2005bn, Vieilledent:2010fv}.
-More recently, however, attempts have been made to examine this inter-specific variation by correlating species intrinsic growth and mortality rates \citep[e.g.][]{Poorter:2008iu, Wright:2010fl}.
+More recently, however, attempts have been made to examine this inter-specific variation by correlating species intrinsic growth with population mortality rates \citep[e.g.][]{Poorter:2008iu, Wright:2010fl}.
 In contrast to intraspecific studies, these studies have shown that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species \citep{Loehle:1996vw, Poorter:2008iu, Wright:2010fl}.
@@
 
-Wood density, the biomass invested per unit wood volume, is one functional trait that has been consistently shown to be negatively correlated with tropical tree mortality \citep{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+Wood density, the biomass invested per unit wood volume, is one such functional trait that has consistently been shown to be negatively correlated with tropical tree mortality \citep{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
 The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to random events such as stem breakage \citep{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \citep{Hacke:2001kj, Jacobsen:2005fx} and fungal and pathogen attack \citep{Augspurger:1984wx}.
-As such, it is hypothesized that species with higher wood density are likely to exhibit lower baseline hazards.
+As such, it is hypothesized that species with higher wood density are likely to exhibit lower stochastic hazards.
 However, research has also shown that wood density may increase tolerance to low growth rates, and thus alter the growth-mortality relationship.
@@
 Furthermore, we extend existing individual-based approaches \citep[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship (hereon in referred to as growth-dependent hazard).
+We then compare the predictive performance of this model relative to simpler alternatives commonly used in predictive vegetation models.
 We parameterise this model using 15-years of individual dbh growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals.
-Using this model,coupled with cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to the null model or growth-dependent only model 2) Does allowing both hazards to vary by species improve predictive capacity?
+Using this model, coupled with cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to the null model or growth-dependent only model 2) Does allowing both hazards to vary by species improve predictive capacity?
 3) Does wood density predominately affect baseline hazards, or does it affect growth-dependent hazards or both?
@@
 BCI is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI) and the permanent plot was established in 1980.
-Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in .
+Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in (REF).
 
@@
 For the purpose of modelling mortality as a function of growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
-We also excluded individuals based on several rules: 1) those that did not survive at least two censuses (two being required to estimate growth rates); 2) those that were not consistently measured at 1.3 m above ground; 3) those that were multi-stemmed; or 4) those that resprouted or '\textit{returned from the dead}'.
+We also excluded individuals based on the following rules: 1) those that did not survive at least two censuses (two being required to estimate growth rates); 2) those that were not consistently measured at 1.3 m above ground; 3) those that were multi-stemmed; or 4) those that resprouted or '\textit{returned from the dead}'.
 We also discarded species that do not exhibit secondary growth (e.g.
@@
 Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial dbh were also removed.
-In order to reduce computational requirements we then subsampled this dataset by randomly selecting one census observation per individual.
+In order to reduce computational requirements we then subsampled this dataset by randomly selecting one observation per individual out of a possible three censuses.
 In total, \Sexpr{n_inds} individuals of \Sexpr{n_spp} species were included in our analysis.
@@
 The standard deviation of these discrepencies was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter.
-Having an estimate of observation error, we then built an empirical Bayesian model to estimate '\textit{true}' initial and final dbh, and subsequently,'\textit{true growth}' for each individual.
+Having an estimate of observation error, we then used Bayesian point estimation to maximise the joint posterior of the '\textit{true}' initial and final dbh, and subsequently,'\textit{true growth}' for each individual.
 We did this by first modelling the observed initial ($dbh_{t-1}$) and final dbh ($dbh_{t}$) of each individual, $i$, as random realisations from a normal distribution centered on their respective true estimates (i.e.
@@
 
-The initial dbh, $\widehat{dbh}_{i,t-1}$, was sampled as a random realisations from a lognormal distribution:
+The initial dbh for each individual $i$, $dbh_{i,t-1}$, was sampled as a random realisation from a lognormal distribution:
 \begin{equation} \label{eq:true_dbh1}
-\widehat{dbh}_{i,t-1} \sim \mathcal{LN}(\mu_{dbh_{i,t-1}}, \sigma_{dbh_{t-1}}), 
+dbh_{i,t-1} \sim \mathcal{LN}(\widehat{dbh}_{i,t-1}, \sigma_{dbh_{t-1}}), 
 \end{equation}
@@
 
-The final dbh measurement, $\widehat{dbh}_{i,t}$ was estimated as:
+The final dbh measurement for each individual $i$, $dbh_{i,t}$ was estimated as:
 \begin{equation}\label{eq:true_dbh2}
-\widehat{dbh}_{i,t} = \widehat{dbh}_{i,t-1} + R_i \, T_i, 
+dbh_{i,t} = \widehat{dbh}_{i,t-1} + R_i \, T_i, 
 \end{equation}
-where, $R_i$ represents the annual true increase in dbh (cm) and $T_i$ represents the number of days between measurements divided by 365.25.
+where, $R_i$ represents the annual true increase in dbh (cm) and $T_i$ represents the number of days between initial and final measurements divided by 365.25.
 We modelled $R_i$ as a random realisation from a lognormal distribution centered on zero and a standard deviation of 1.
-This prior conveyed our expectation that annual increases in dbh were unlikely to be greater than 7 cm/yr.
-Sampling from a lognormal distribution also ensured growth rates could not be negative.
+This prior conveyed our expectation that annual dbh growth could not be negative and is unlikely to be greater than 7 cm/yr.
 
 Using this model we then used \stan's \textit{optimizing} function to maximise the joint posterior and extract point estimates of true initial dbh, $\widehat{dbh}_{i,t-1}$, true final dbh, $\widehat{dbh}_{i,t}$, and true annual dbh growth $R_i$.
-We used annual diameter growth, $R_i$, and annual basal area growth:$\frac{(\pi \, 0.25 \,\widehat{dbh}_{i,t}) - (\pi \, 0.25 \, \widehat{dbh}_{i,t-1})}{T_i}$ in the subsequent mortality analysis.
+We then used annual diameter growth, $R_i$, and annual basal area growth:$\frac{(\pi \, 0.25 \,\widehat{dbh}_{i,t}) - (\pi \, 0.25 \, \widehat{dbh}_{i,t-1})}{T_i}$ in the subsequent mortality analysis.
 
@@
 \begin{equation} \label{eq:mortality_model}
-p_{t_1\rightarrow t_2} = 1 - \exp\left((t_2-t_1)\,\lambda_i\right)
+p_{t_1\rightarrow t_2} = 1 - \exp\left(-\lambda\,T\right)
 \end{equation}
-where, $\lambda$, is the instantaneous hazard rate of death per unit time \citep{Kleinbaum:2012tn}.
+where, $\lambda$, is the instantaneous hazard rate of death per unit time \citep{Kleinbaum:2012tn} and $T$ is the time lapsed between $t_1$ and $t_2$.
 
 Here we examine the performance of three functional forms of $\lambda$ to predict observed patterns of tropical tree mortality.
-The simplest model (eqn:\ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plants growth rate.
+The simplest model (eqn:\ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate.
 The second model (eqn: \ref{growth_haz} describes a growth-dependent hazard which assumes plants can only die due to a growth related cause.
@@
 The third model (eqn:\ref{add_haz}, and potentially the most biologically realistic, is the summation of the two previous models.
-Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at baseline hazard instead of zero.
-The advantage of this model is that it effectively accounts for two independent causes of mortality: that due to growth and that due to chance (i.e.
+Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero.
+The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e.
 stochastic events such as windthrow and fire).
-The parameters of eqn:\ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept (i.e.
-sum of both growth dependent and growth independent mortality at zero growth rate) and biologically defines the instantaneous mortality hazard $\lambda$ at zero growth.
+The parameters of eqn:\ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality hazard $\lambda$ at low annual growth rate.
 $\beta$ represents the exponential decay rate or the sensitivity of hazard rates to change with growth rate; and $\gamma$, is the asymptote and defines the baseline hazard.
-$\alpha$ and $\beta$ are intrinsically linked to growth rate, and as such are growth-dependent parameters.
+$\alpha$ and $\beta$ are intrinsically linked to a plants growth rate, and as such are growth-dependent parameters.
 By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g.
@@
 \begin{align}
-\lambda &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
-\lambda &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
-\lambda &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} + 
+\lambda_i &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
+\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
+\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} + 
 \overbrace{\gamma}^{\text{Baseline hazard}} \label{add_haz}
@@
 
-A central aim of this study is to identify what is the most predictive parameterisation of $\lambda$.
+The central aim of this study is to identify the most parsimonious combination of $\lambda$, growth measure (dbh increment growth or area growth) and wood density that best predicts tropical tree mortality.
 To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison.
-Stage 1 involved comparing predictive performance of the three $\lambda$ models described in equations \ref{base_haz}, \ref{add_haz}.
-In this stage each model was fitted with fixed parameters (i.e.
-parameters were constant across species) and then again by allowing parameters to varying by species.
-This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) the extent to which species hazard rates differ.
-In stage 2 we used the most predictive model from stage 1 to examine whether dbh growth or basal area growth was a better predictor of plant mortality.
-In stage 3 we used the best model from stage 2, but without species random effects, to determine what combination of parameter-wood density effects best predicted tree mortality.
-This stage also allowed us to determine whether wood density primarily affected growth-dependent or growth-independent (i.e.
-baseline) mortality.
-Finally, in stage 4, we fit the best model based on the previous three steps to the entire dataset with species random effects on all parameters.
-We then examined wood density effect sizes and the proportion of species variability that it explained.
+Stage 1 involved comparing predictive performance of the three $\lambda$ models described in equations \ref{base_haz} \textendash \ref{add_haz} using either annual dbh increment or area growth rates.
+In this stage each model was fitted without allowing parameters to vary by species.
+This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predicitive performance varied by growth measure.
+Following this, we used the best model of stage 1 but allowed its parameters to vary by species.
+This allowed us to determine how variable each parameter was across species and whether accounting for species differences significantly improved predictive performance.
+Because species responses varied considerably, we undertook a third stage of model fitting and selection that explicitly examined whether wood density (a species level trait) explained some of this interspecific variability.
+This involved using the best model observed in stage 1 and adding wood density effects to each parameter.
+Because we were interested in obtaining the most parimonous model of tropical tree mortality, we fitted the model under all combinations of wood density effects.
+This allowed us to determine the most parsimonous combination of wood density effects and whether it most strongly influenced growth-dependent or growth-independent mortality hazards.
+Following this we then fitted the best wood density combination model with species random effects to the entire dataset.
+This model was then used to examine mortality curves across wood density gradients.
+In order to determine the proportion of interspecific variation explained by wood density on each parameter it occurred on we reran the final model form with just random effects on parameters and calculated the percent reduction in unexplained species variation.
 
@@
 
-Our observations are of survival status $y_i$ (0 = alive, 1 = died) for each individual $i$ between two censuses.
-We model these observations as random realizations from a Bernoulli distribution based on the probability that individual $i$ dies between $t_1$ and $t_2$, $p_{i,t_1\rightarrow t_2}$:
+Our observations are of survival status $y_i$ (0 = alive, 1 = died) for each individual $i$ between census $t_1$ and $t_2$.
+We model these observations as random realizations from a Bernoulli distribution:
 \begin{equation}
@@
 
-We modelled $p_{i,t_1\rightarrow t_2}$ using a complementary log-log link as:
+Here, $p_{i,t_1\rightarrow t_2}$ is the probability individual $i$ dies over an observed time period $T_i$. We estimate this as:
 \begin{equation}
-\log(-\log(1 - p_{i,t_1\rightarrow t_2})) = \log((t_2-t_1)\,\lambda_i).
-\end{equation}
-This parameterisation allowed for additive hazards such as those in eqn:\ref{add_haz} and is equivilant to a competiting risks analysis \citep{Kleinbaum:2012tn}.
-
-As we have multiple hypothesized functional forms of $\lambda$ (e.g.
-eqn:\ref{base_haz}, \ref{add_haz}), each containing parameters that may or may not vary by species, we outline the fitting procedure of our most complicated model of which all others are nested within:
-\begin{equation}
-\lambda_i = \overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, Growth_i \right)}^{\text{Growth dependent}} + 
-\overbrace{\gamma_{s[i]}}^{\text{Growth independent}}.
+p_{i,t_1\rightarrow t_2} = 1 - exp(-\lambda_i\,T_i).
 \end{equation}
 
-In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ varying by species a function of both wood density (measured at species level) and a species random effect:
+Because we have multiple hypothesized functional forms of $\lambda_i$ (e.g.
+equations \ref{base_haz}\textendash\ref{add_haz}), we describe the fitting procedure of our most complicated model of which all others are nested within:
+\begin{equation}
+\lambda_i = \left(\overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, Growth_i \right)}^{\text{Growth dependent}} + 
+\overbrace{\gamma_{s[i]}}^{\text{Growth independent}}\right) \times \overbrace{\delta_{t[i]}}^{\text{Census effect}}
+\end{equation}
+
+In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ varying by species as a function of both wood density (measured at species level) and a species random effect:
 \begin{equation}\label{eq:submodels}
@@
 \end{equation}
+
 Here $\alpha_{0,s}$,$\beta_{0,s}$,$\gamma_{0,s}$ represents the species-level residual error not explained by wood density for species $s$ within which individual $i$ belongs.
-$\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of standardised wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively.
+$\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively.
 The form of submodels in eqn:\ref{eq:submodels} ensures that parameters remain positive.
@@
 This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6 cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant).
+In addition to this, we also centered $Growth_i$ on the lowest 5\% quantile for both dbh increment and area growth (0.172 and 0.338, respectively).
+This weas done to ensure there were sufficient data to accurately estimate $\alpha$ and that it would be interpreted as the hazard rate when growth rate was very low.
 
-Species random effects were modelled as random realisations from log-normal distributions:
+We modelled species random effects as random realisations from log-normal distributions:
 \begin{equation}
@@
 
-Both $\mu_{\log \alpha_0}$, the average log species residual error at zero growth, and $\mu_{\log \gamma_0}$, the average baseline log species residual error, from normal distributions centered on zero with a standard deviation of 2.5.
+All three mean hyperparameters, $\mu_{\log \alpha_0}$, the average log species residual error at zero growth, $\mu_{\log \beta_0}$, the average species error on the exponential decay rate, and $\mu_{\log \gamma_0}$, the average baseline log species residual error, were estimated using normal distributions centered on zero with a standard deviation of 2.5.
 \begin{equation}
-\mu_{\log \alpha_{0}},\mu_{\log \gamma_0} \sim \mathcal{N}\left(0, 2.5\right).\\
+\mu_{\log \alpha_{0}},\mu_{\log \beta_0},\mu_{\log \gamma_0} \sim \mathcal{N}\left(0, 2.5\right).\\
 \end{equation}
-
-We estimated $\mu_{\log \beta_0}$, the average species error on the exponential decay rate, using a weak normal prior centered on zero with a standard deviation of 5. 
-\begin{equation}
-\mu_{\log \beta_0} \sim \mathcal{N}\left(0, 5\right).\\
-\end{equation}
-As we were fitting these models to multiple growth measures (i.e.
-dbh increment, basal area growth), we intentially made this prior weak, such that the range of possible values spanned from a rapid exponential decay to no change in mortality with increasing growth.
 
@@
 \end{equation}
-For models that did not contain species random effects, the above sigma's were not calculated and only the hyperparameter means (i.e.
-$\mu_{\log \alpha_0}$,$\mu_{\log \beta}$,$\mu_{\log \gamma_0}$) were estimated.
 
-Wood density effects (i.e. $\alpha_1$, $\beta_1$ and $\gamma_1$) were estimated using normal priors centered on zero and a standard deviation of 5:
+For models that did not include species random effects, single intercepts were estimated for each parameter using log normal distributions centered on zero with standard deviations of 2.5.
+
+We estimated wood density effects (i.e. $\alpha_1$, $\beta_1$ and $\gamma_1$) using normal priors centered on zero and a standard deviation of 2.5:
 \begin{equation} \label{eq:rho_effects}
-\alpha_1, \beta_1, \gamma_1 \sim \mathcal{N}\left(0, 5\right).\\
+\alpha_1, \beta_1, \gamma_1 \sim \mathcal{N}\left(0, 2.5\right).\\
 \end{equation}
 
-All models were fit using package rstan version 2.8.1 \citep{StanDevelopmentTeam:2015uz}.
-Three independent chains were executed and in all cases modelled parameters converged within 2000 iterations.
+We accounted for possible variation in mortality rates across different census periods by allowing $lambda_i$ to be scaled by $\delta_t$ for each census $t$.
+We estimated $\delta_t$ as a random realization from a lognormal distribution centered on zero and a variance parameter $\sigma_{\delta}$ defines the variation in census effects:
+
+\begin{equation} \label{eq:census_effect}
+\delta_t \sim \mathcal{LN}\left(0, \sigma_{\delta}\right)\\
+\sigma_{\delta} \sim \textrm{Cauchy+}\left(0, 2.5\right).
+\end{equation}
+
+For each model we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\citep{StanDevelopmentTeam:2015uz}.
+We excuted three independent chains and in all cases modelled parameters converged within 2000 iterations.
 Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \citep{Gelman:1992ts, Brooks:1998ju}.
@@
 We examined predictive accuracy of each model by using K-fold cross-validation.
-This involved randomly partitioning the data into ten folds of \Sexpr{n_heldout} observations, ensuring that each K-fold had at least one observation per species.
+This involved randomly partitioning the data into ten folds of \Sexpr{n_heldout} observations, while ensuring that each K-fold had at least one observation per species.
 We then fitted models to nine folds and predicted to the tenth fold.
@@
 
-Predictive accuracy was quantified as the mean (over 10 cross-validation replicates) of each of two threshold-independent performance metrics.
-First we compare log likelihoods of heldout data, which define the log probability of observing the data given the model.
-Secondly, in order to examine the model capacity to correctly discriminate between whether an individual survived or not, we used the area under the receiver operating characteristic (ROC) curve (AUC).
-AUC describes the model’s discrimination capacity as the frequency with which it predicts a higher probability of mortality for a randomly selected observed death than for a randomly selected observed survivor.
+Predictive accuracy was quantified by estimating the average logarithmic loss $logloss$ of across K-folds for heldout data:
+\begin{equation} \label{eq:logloss}
+logloss = \frac{1}{K}\,\bigg(-\frac{1}{N}\,\sum_{i=1}^{N}\Big(y_i\,\log(p_{i,t_1\rightarrow t_2}) + ( 1 - y_i)\,\log(1 - p_{i,t_1\rightarrow t_2})\Big)\bigg),
+\end{equation}
+where $N$ refers to the number of heldout observations and $K$ is the number of K-folds.
+
+Logarithmic loss, also known as Log Loss, measures the accuracy of a model by incorporating probabilistic confidence.
+It does this by penalizing incorrect predictions based on how wrong the predicted probability is from the observation.
+For example, if an individual died between two time steps but the model predicted its probability of death at 0.1, then it would have a Log Loss of \Sexpr{round(- 1*log(0.1) + (1 - 1)*log(1 - 0.1),2)}, whereas if the the model predicted the probability of death was 0.9 then the Log Loss would be \Sexpr{round(- 1*log(0.9) + ( 1 - 1)*log(1 - 0.9),2)}.
+By minimising the Log Loss, we effectively maximise the model accuracy.
+In order to choose between models we averaged the estimated Log Loss across K-folds and reported the mean and standard error.
+
 \pagebreak

### 2016-05-31 ms/ms.Rnw
Edits to manuscript

@@
-\documentclass[a4paper,12pt]{article}
-\usepackage[osf]{mathpazo}
-\usepackage{ms}
-\usepackage[]{natbib}
-\usepackage{gensymb}
-\usepackage{amsmath}
-\usepackage[title, titletoc, toc]{appendix}
-%\raggedright
+\documentclass[9pt,twocolumn,twoside, lineno]{pnas-new}
+% Use the lineno option to display guide line numbers if required.
+% Note that the use of elements such as single-column equations
+% may affect the guide line number alignment.
 
-\newcommand{\stan}{\texttt{stan}}
+\templatetype{pnasresearcharticle} % Choose template 
+% {pnasresearcharticle} = Template for a two-column research article
+% {pnasmathematics} = Template for a one-column mathematics article
+% {pnasinvited} = Template for a PNAS invited submission
 
-\title{Modelling plant mortality: integrating intra and interspecific approaches}
-\author{James S. Camac\textasteriskcentered$^1$, Richard G. Fitzjohn$^1$, Mark Westoby$^1$, Daniel S. Falster$^1$}
-\affiliation{
-$^1$ Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia\\
-\textasteriskcentered Email for correspondence: \texttt{james.camac@gmail.com}\\
-Word count: }
-\date{}
+\title{Unifying intraspecific and interspecific variation in tropical tree mortality}
 
-\bibliographystyle{mee.bst}
+% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
+\author[a,1]{James S Camac}
+\author[a,2]{Mark Westoby} 
+\author[a,3]{Daniel S Falster}
 
-\mstype{Research Article}
-\runninghead{Incorporating traits into individual based mortality models}
-\keywords{traits, wood density, growth rates, bayesian, survival analysis}
+% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
+\affil[a]{Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia}
+
+% Please give the surname of the lead author for the running footer
+\leadauthor{Camac}
+
+% Please add here a significance statement to explain the relevance of your work
+\significancestatement{Add something significant here}
+
+% Please include corresponding author, author contribution and author declaration information
+\authorcontributions{J.C.
+designed and performed the analysis; D.F.
+conceived the idea.
+M.W.
+supervised the development of this work and provided biological and statistical advice.
+All authors contributed to the writing of this manuscript.}
+
+\authordeclaration{The authors declare that they have no competing financial interests.}
+\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: james.camac@gmail.com}
+
+% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
+\keywords{Tropical trees $|$ wood density $|$ growth $|$ mortality $|$ interspecific}
+
+\begin{abstract}
+Add abstract here
+\end{abstract}
+
+\dates{This manuscript was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
+
+\dates{This manuscript was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
 
@@
 \begin{document}
-\mstitlepage
-\noindent
-\parindent=1.5em
-\addtolength{\parskip}{.3em}
-\doublespacing
-\linenumbers
+% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
+% You should only change this length when you've finalised the article contents.
+\verticaladjustment{-2pt}
 
-\section{Summary}\label{abstract}
+\maketitle
+\thispagestyle{firststyle}
+\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
 
-\section{Introduction}
-Plant mortality is often modelled as a function of recent growth because it reflects the product of many biotic and abiotic factors on a plants overall carbon budget \citep{Kobe:1995tw, Hawkes:2000ib, Keane:2001db, Wyckoff:2002ul}.
-As such, growth rates are considered a highly integrated individual-based predictor of plant mortality \citep{Hawkes:2000ib}.
-Within a species, the assumption of the growth-mortality relationship is that fast growing individuals (i.e.
-those that have larger carbon budgets) are more able to buffer stress and thus have higher survivorship relative to slow growing (i.e.
-low carbon budget) individuals \citep{Hawkes:2000ib, Keane:2001db}.
-Empirical studies examining this growth-mortality relationship have broadly supported this assumption, with many highlighting non-linear declines in mortality with increased growth \citep[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
-Consequently, most dynamic vegetation models, including tree gap models \citep{Keane:2001db}, individual-based forest simulators \citep[e.g.
-SORTIE][]{Pacala:1996vm} and Dynamic Global Vegetation Models \citep[DGVMs; ][]{Woodward:2004ft} use some measure of growth coupled with a stochastic constant (i.e.
-baseline hazard or minimum chance of death) as the fundamental determinants of plant mortality \citep{Hawkes:2000ib, McDowell:2011dr}.
+% If your first paragraph (i.e.
+with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation.
+If this is the case, add \parshape=0 to the end of the list environment.
+\dropcap{P}lant mortality is often modelled as a function of recent growth because it reflects the product of many biotic and abiotic factors on a plants overall carbon budget and vigour \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db, Wyckoff:2002ul}.
+As such, growth rates are considered a highly integrated individual-based predictor of plant mortality \cite{Hawkes:2000ib}.
+Within a species, the assumption is that fast growing individuals (i.e.
+those that have larger carbon budgets) are more able to buffer stress and therefore have higher survivorship relative to slow growing (i.e.
+low carbon budget) individuals \cite{Hawkes:2000ib, Keane:2001db}.
+Empirical studies examining this growth\textendash mortality relationship have broadly supported this assumption, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
+Consequently, most dynamic vegetation models, including tree gap models \cite{Keane:2001db}, individual-based forest simulators \cite[e.g.
+SORTIE][]{Pacala:1996vm} and Dynamic Global Vegetation Models \cite[DGVMs; ][]{Woodward:2004ft} use some measure of growth coupled with a stochastic constant (i.e.
+baseline hazard or minimum chance of death) as the fundamental determinants of plant mortality \cite{Hawkes:2000ib, McDowell:2011dr}.
 
-However, due to a paucity of data on plant mortality, most dynamic vegetation models have implemented mortality algorithms using overly simplistic assumptions \citep{Hawkes:2000ib, Wyckoff:2000un, McDowell:2011dr}.
-One of the strongest assumptions these models make is that both stochastic hazards and the growth-mortality relationship are constant across species \citep{Pacala:1996vm, Hawkes:2000ib, Wyckoff:2000un}.
+However, due to a paucity of data on plant mortality, most dynamic vegetation models have implemented mortality algorithms with overly simplistic assumptions \cite{Hawkes:2000ib, Wyckoff:2000un, McDowell:2011dr}.
+One of the strongest assumptions made is that both stochastic hazards and the growth-mortality relationship are constant across species \cite{Pacala:1996vm, Hawkes:2000ib, Wyckoff:2000un}.
 That is, all species have the same random chance of death and exhibit the same tolerance to low growth.
-This is despite growing empirical evidence highlighting that mortality rates vary considerably between species, even within the same community \citep{Kobe:1997vy, Wyckoff:2002ul, Kunstler:2005bn, Vieilledent:2010fv}.
-More recently, however, attempts have been made to examine this inter-specific variation by correlating species intrinsic growth with population mortality rates \citep[e.g.][]{Poorter:2008iu, Wright:2010fl}.
-In contrast to intraspecific studies, these studies have shown that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species \citep{Loehle:1996vw, Poorter:2008iu, Wright:2010fl}.
+This is despite growing empirical evidence highlighting that mortality rates vary considerably between species, even within the same community \cite{Kobe:1997vy, Wyckoff:2002ul, Kunstler:2005bn, Vieilledent:2010fv}.
+Furthermore, because no systematic analysis has been conducted to determine what growth measure best predicts plant mortality, the growth-mortality relationship is often parameterised using a variety growth measures; the two most common being rates of dbh or basal area growth.
+
+Recently, research has focused on examining inter-specific variation by correlating intrinsic species' growth rates with population mortality rates \cite[e.g.][]{Poorter:2008iu, Wright:2010fl}.
+In contrast to intraspecific work, these studies have shown that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species \cite{Loehle:1996vw, Poorter:2008iu, Wright:2010fl}.
 As such, species appear to exhibit a trade off between greater growth versus greater survival.
-Ecologists have attempted to explain this interspecific variation by using functional traits - attributes of species that are thought to influence vital rates (i.e.
-survival, growth and reproduction) and ultimately fitness \citep{Westoby:2002tb, Ackerly:2003eb, Falster:2015un}.
+Ecologists have attempted to explain this interspecific variation by using a variety of functional traits \textemdash attributes of species that are thought to influence vital rates (i.e.
+survival, growth and reproduction) and ultimately fitness \cite{Westoby:2002tb, Ackerly:2003eb, Falster:2015un}.
 
-Wood density, the biomass invested per unit wood volume, is one such functional trait that has consistently been shown to be negatively correlated with tropical tree mortality \citep{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
-The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to random events such as stem breakage \citep{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \citep{Hacke:2001kj, Jacobsen:2005fx} and fungal and pathogen attack \citep{Augspurger:1984wx}.
+Wood density, the biomass invested per unit wood volume, is one such functional trait that has consistently been shown to be negatively correlated with tropical tree mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to random events such as stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx} and fungal and pathogen attack \cite{Augspurger:1984wx}.
 As such, it is hypothesized that species with higher wood density are likely to exhibit lower stochastic hazards.
 However, research has also shown that wood density may increase tolerance to low growth rates, and thus alter the growth-mortality relationship.
-For example, species with high wood density are more tolerant shade \citep{vanGelder:2006exa} and competition \citep{Kunstler:2016km}.
-Studies have also shown that wood density may reduce drought induced carbon starvation \citep{McDowell:2011dr} and the susceptibility of ‘stressed’ individuals to opportunistic invertebrate, pathogen or fungal attack \citep{Augspurger:1984wx, McDowell:2011dr}.
+For example, species with high wood density have been found to be more tolerant to both shade \cite{vanGelder:2006exa} and neighbor competition \cite{Kunstler:2016km}.
+Studies have also shown that wood density may reduce drought induced carbon starvation \cite{McDowell:2011dr} and the susceptibility of ‘stressed’ individuals to opportunistic invertebrate, pathogen or fungal attack \cite{Augspurger:1984wx, McDowell:2011dr}.
+Consequently, species with high wood density may be more tolerant to both stochastic events such as breakage as well as be more tolerant to periods of low growth.
 
@@
 growth rates) and species (e.g.
-function traits) characteristics are required in order to develop generalizable models of plant mortality.
+function traits) characteristics are likely to be required in order to develop general and accruate models of plant mortality.
 However, to date, there has been little crossover of ideas between intraspecific and interspecific studies of plant mortality.
-For example, we are aware of only one study that uses individual based plant mortality empirical models to examine trait effects on baseline hazards \citep{AubryKientz:2013dg}, and none that examine how traits mediate the growth-mortality relationship.
+For example, we are aware of only one study that uses individual based plant mortality empirical models to examine trait effects on baseline hazards \cite{AubryKientz:2013dg}, and none that examine how traits mediate the growth-mortality relationship.
 
-In this study, we reconcile both intraspecific and interspecific theory by developing an individual based models of plant mortality that incorporate individual based measurements of growth as well as the species trait wood density.
-Furthermore, we extend existing individual-based approaches \citep[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship (hereon in referred to as growth-dependent hazard).
-We then compare the predictive performance of this model relative to simpler alternatives commonly used in predictive vegetation models.
-We parameterise this model using 15-years of individual dbh growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals.
-Using this model, coupled with cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to the null model or growth-dependent only model 2) Does allowing both hazards to vary by species improve predictive capacity?
-3) Does wood density predominately affect baseline hazards, or does it affect growth-dependent hazards or both?
-4) What better predicts plant mortality, absolute dbh growth or basal area growth?
+In this study, we reconcile both intraspecific and interspecific theory by developing a unfied model that incorporates both individual growth and the species level trait, wood density.
+We extend existing individual-based approaches \cite[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship within a single model (hereon in referred to as growth-dependent hazard).
+We then compare the predictive performance of this model to other traditionally used mortality models.
+We parameterise these models using 15-years of individual growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals.
+Using these models, coupled with 10-fold cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to baseline only or growth-dependent only models?
+2) is dbh growth or basal area growth more predicitive of tropical tree mortality and does it's predictive capacity vary with model type?
+3) Does allowing both hazards to vary by species improve predictive capacity?
+4) Does wood density predominately affect baseline hazards, or does it affect growth-dependent hazards or both?
 
-\clearpage
+\section*{Results}
 
-\section{Methods}
+\begin{figure}
+\centering
+\includegraphics{figures/fig1.pdf}
+\caption{Average logarithmic loss $\pm$ 95\% BCIs from 10-fold cross validation.
+(A) Stage 1: Comparing three functional forms of tropical tree mortality using absolute diameter growth (red) and basal area growth (blue).
+(B) Comparing model with and without species random effects.
+(C) Comparing all combinations of wood density effects.}
+\label{fig1}
+\end{figure}
 
-\subsection{Data}
-We parameterise plant mortality models using individual growth and survival data \citep{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\degree;N, 79.85\degree;W).
-The climate on the island is warm throughout the year with a mean annual temperature of 26.1\degree and rainfall is seasonal with most of the 2500 mm falling between April and November.
+\subsection*{Most predictive functional form}
+
+Full model most predictive (also most biologically reasonable)
+DBH DT most predictive given functional form
+
+\subsection*{What is the most predictive combination of parameter-wood density effects?}
+
+wood density on gamma. Technically significant on alpha too... but effect size is small (can't show this though as focus is on predictive capacity with cross val.)
+
+\subsection*{How well does our model predict if we allow full random effects?}
+
+Seems to account for ~20-30\% of species variability (this will be confirmed once other models are finished).
+
+Figures:
+Single figure of logloss
+figure of species gamma values vs rho
+response curve for low and high wood density species from dataset.
+
+Species variability is highest on gamma (1.2) followed by alpha (0.8) and least on beta (0.3)
+Mention proportion of species variability on gamma explained by wood density.
+
+\section*{Discussion}
+
+- General overview of results:
+As expected, the most biologically reasonable model; the model with both a baseline hazard and growth-dependent hazard was the best predictor of tropical tree mortality, indicating that both growth dependent and independent hazards are needed to effectively model tree mortality.
+
+- Furthermore, we found that DBH was by far the better growth measure for predicting plant mortality.
+	- Why DBH?
+Could it be that DBH implicitly punishes smaller individuals relative to larger individuals.
+That is, small plants can only gain a small amount of dbh growth relative to a larger plant.
+As such, dbh growth implicitly captures the process whereby small individual plants are more prone to competitive effects as well as random breakage events caused by trampling, herbivory etc.
+
+- As expected, when allowing parameters to vary by species we found a significant improvement in model predictive capacity.
+		- This supports previous research that shows that species vary substantially in mortality rates.
+
+- When examining this interspecific variability we found that the effect of wood density improved model predictive capcity when it was included on the base-line hazard, with only marginal improvements when included on all three parameters.
+
+- Furthermore, we found that wood density explained XX\% of the total species variation for the baseline hazard.
+
+- Conclusion, species with higher wood density had lower baseline hazards supporting previous work that suggests that these species are less prone to stochastic events that involve breakage etc.
+
+\matmethods{
+\subsection*{Data}
+We parameterise plant mortality models using individual growth and survival data \cite{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\textdegree;N, 79.85\textdegree;W).
+The climate on the island is warm throughout the year with a mean annual temperature of 26.1\textdegree and rainfall is seasonal with most of the 2500 mm falling between April and November.
 BCI is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI) and the permanent plot was established in 1980.
@@
 
-Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and contained a diameter at breast height greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter \citep{Condit:2012nz}.
+Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and contained a diameter at breast height greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter \cite{Condit:2012nz}.
 For the purpose of modelling mortality as a function of growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
@@
 
-Wood density for each species was estimated by coring trees located within 15 km of the BCI plot because increment borers are prohibited within the plot \citep{Wright:2010fl}.
-Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100\degree C) and dividing by the fresh volume (as measured by water displacement).
+Wood density for each species was estimated by coring trees located within 15 km of the BCI plot because increment borers are prohibited within the plot \cite{Wright:2010fl}.
+Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100\textdegree C) and dividing by the fresh volume (as measured by water displacement).
 Wood density was then calculated as an area-weighted average (g cm$^-3$), where area refers to the annulus represented by each piece, assuming a circular trunk.
 
-\subsection{Estimating true growth}
+\subsection*{Estimating true growth}
 
@@
 While there are physiological processes (e.g.
-water depletation and bark drop) that can lead to trunk shrinkages overtime \citep{Pastur:2007kl}, it is biologically impossible for growth (i.e.
-the creation of new biomass) to be negative \citep{Clark:2007ve}.
+water depletation and bark drop) that can lead to trunk shrinkages overtime \cite{Pastur:2007kl}, it is biologically impossible for growth (i.e.
+the creation of new biomass) to be negative \cite{Clark:2007ve}.
 To ensure our model was not biased by these implausable values we estimated '\textit{true growth}'(i.e.
-observed growth minus observation error) by first using another BCI dataset collected explicitly to examine obervation errors \citep{Condit:2012nz}.
-This dataset consisted of \Sexpr{n_obs_errors} randomly selected trees that had their dbh measured twice within a 30 day period \citep{Condit:2012nz}.
+observed growth minus observation error) by first using another BCI dataset collected explicitly to examine obervation errors \cite{Condit:2012nz}.
+This dataset consisted of \Sexpr{n_obs_errors} randomly selected trees that had their dbh measured twice within a 30 day period \cite{Condit:2012nz}.
 Assuming no growth occurred over this period, any discrepency in measurements can be attributable to observation error.
@@
 
-Using this model we then used \stan's \textit{optimizing} function to maximise the joint posterior and extract point estimates of true initial dbh, $\widehat{dbh}_{i,t-1}$, true final dbh, $\widehat{dbh}_{i,t}$, and true annual dbh growth $R_i$.
-We then used annual diameter growth, $R_i$, and annual basal area growth:$\frac{(\pi \, 0.25 \,\widehat{dbh}_{i,t}) - (\pi \, 0.25 \, \widehat{dbh}_{i,t-1})}{T_i}$ in the subsequent mortality analysis.
+Using this model we then used \texttt{rstan} \textit{optimizing} function to maximise the joint posterior and extract point estimates of true initial dbh, $\widehat{dbh}_{i,t-1}$, true final dbh, $\widehat{dbh}_{i,t}$, and true annual dbh growth $R_i$.
+We then used annual diameter growth, $R_i$, and annual basal area growth: $\frac{(\pi \, 0.25 \,\widehat{dbh}_{i,t}) - (\pi \, 0.25 \, \widehat{dbh}_{i,t-1})}{T_i}$ in the subsequent mortality analysis.
 
-\subsection{Mortality analysis}
+\subsection*{Mortality analysis}
 
@@
 \end{equation}
-where, $\lambda$, is the instantaneous hazard rate of death per unit time \citep{Kleinbaum:2012tn} and $T$ is the time lapsed between $t_1$ and $t_2$.
+where, $\lambda$, is the instantaneous hazard rate of death per unit time \cite{Kleinbaum:2012tn} and $T$ is the time lapsed between $t_1$ and $t_2$.
 
@@
 In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached.
-This negative exponential relationship has been used in other studies to model growth-dependent mortality \citep[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \citep{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}.
+This negative exponential relationship has been used in other studies to model growth-dependent mortality \cite[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \cite{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}.
 The third model (eqn:\ref{add_haz}, and potentially the most biologically realistic, is the summation of the two previous models.
@@
 
-\subsection{Model fitting}
+\subsection*{Model fitting}
 
@@
 
-For each model we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\citep{StanDevelopmentTeam:2015uz}.
+For each model we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}.
 We excuted three independent chains and in all cases modelled parameters converged within 2000 iterations.
-Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \citep{Gelman:1992ts, Brooks:1998ju}.
+Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Gelman:1992ts, Brooks:1998ju}.
 After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.
 
-\subsection{Model evaluation}
+\subsection*{Model evaluation}
 
@@
 This was repeated until each fold had been used nine times for model fitting, and once as a heldout dataset for predicting.
-Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy \citep{Stone:1997tg}.
+Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy \cite{Stone:1997tg}.
 
@@
 In order to choose between models we averaged the estimated Log Loss across K-folds and reported the mean and standard error.
+}
 
-\pagebreak
-\bibliography{refs}
+\showmatmethods % Display the Materials and Methods section
+
+\acknow{
+J.C was supported by the Science and Industry Endowment Fund (SIEF; RP04-174).
+M.W was supported by a fellowship from the Australian Research Council (ARC).
+D.F.
+was supported by an ARC discovery grant.
+We thank Lachlan McCalman and Daniel Steinberg for statistical advice; Anias Gibert etc etc.}
+
+\showacknow % Display the acknowledgments section
+
+% \pnasbreak splits and balances the columns before the references.
+% If you see unexpected formatting errors, try commenting out this line
+% as it can run into problems with floats and footnotes on the final page.
+%\pnasbreak
+
+% Bibliography
+\bibliography{refs.bib}
+%\pnasbreak
 \end{document}

### 2016-05-31 R/figures.R
Changes to how to presentation of figures. More changes to come

@@ -45,52 +45,72 @@ plot_obs_v_pred_growth <- function(data) {
 }
 
 # Plot mean CV log likelihoods for functional comparison
-plot_function_growth_comparison <- function(logloss_summary) {
+plot_function_growth_comparison <- function(logloss_summary, title=NULL) {
   dat <- logloss_summary %>%
     filter(model!= "base_hazard" | growth_measure !='true_basal_area_dt') %>%
     mutate(growth_measure = replace(growth_measure, model=="base_hazard", "none")) %>%
     mutate(model = factor(model, levels=c('base_hazard','growth_hazard', 
                                           'base_growth_hazard'))) %>%
-    mutate(model = factor(model, labels = c( "Baseline hazard","Growth hazard","Baseline & growth hazard"))) %>%
     arrange(model)
   
   ggplot(dat, aes(x = model,y = mean, group = as.factor(growth_measure), colour=as.factor(growth_measure))) + 
     geom_pointrange(aes(ymin = `2.5%`, ymax=`97.5%`), position=position_dodge(.1), size=0.2) +
     scale_colour_manual('',values = c("true_basal_area_dt" ="blue","true_dbh_dt" ="red", "none" = "black")) +
-    ylab('Average Logarithmic Loss') + 
+    ylab('Logarithmic Loss') + 
     xlab('Model') +
+    scale_x_discrete(labels=c("base_hazard" = expression(gamma),
+                              "growth_hazard" = expression(alpha*"e"^{-beta~"growth"["i"]}),
+                              "base_growth_hazard" = expression(alpha*"e"^{-beta~"growth"["i"]} + gamma))) +
+    ylim(0.45,0.511) +
+    labs(title=title) +
     partial_plot_theme()
 }
 
 # Plot mean CV log likelihoods for growth comparison
-plot_random_effect_comparison <- function(log_loss_summary) {
+plot_random_effect_comparison <- function(log_loss_summary, title=NULL) {
   dat <- filter(log_loss_summary, growth_measure =='true_dbh_dt' &
                   (model =='base_growth_hazard' | model =='base_growth_hazard_re')) %>%
     mutate(model = factor(model, levels=c("base_growth_hazard","base_growth_hazard_re"))) %>%
-    mutate(model = factor(model, labels = c("Without random effects","With random effects"))) %>%
[…]

### 2016-06-01 R/figures.R
Updated figures based on new presentation style. Also added place holders for figures yet to be produced by Daniel and myself

@@ -2,7 +2,7 @@
 partial_plot_theme <- function(legend.position = "none", strips = FALSE,...) {
   sb <- if(strips==TRUE) element_rect(fill='lightgrey') else element_blank()
   st <- if(strips==TRUE) element_text(face='italic') else element_blank()
-  theme_classic(base_size = 7) + theme(strip.text = st,
+  theme_classic(base_size = 9) + theme(strip.text = st,
                           legend.title = element_blank(),
                           strip.background = sb,
                           legend.position = legend.position,
@@ -44,73 +44,94 @@ plot_obs_v_pred_growth <- function(data) {
   plot_grid(p1, p3, p2, p4, ncol=2, labels=LETTERS[1:4], label_size = 7)
 }
 
-# Plot mean CV log likelihoods for functional comparison
-plot_function_growth_comparison <- function(logloss_summary, title=NULL) {
-  dat <- logloss_summary %>%
-    filter(model!= "base_hazard" | growth_measure !='true_basal_area_dt') %>%
-    mutate(growth_measure = replace(growth_measure, model=="base_hazard", "none")) %>%
-    mutate(model = factor(model, levels=c('base_hazard','growth_hazard', 
-                                          'base_growth_hazard'))) %>%
-    arrange(model)
-  
-  ggplot(dat, aes(x = model,y = mean, group = as.factor(growth_measure), colour=as.factor(growth_measure))) + 
-    geom_pointrange(aes(ymin = `2.5%`, ymax=`97.5%`), position=position_dodge(.1), size=0.2) +
-    scale_colour_manual('',values = c("true_basal_area_dt" ="blue","true_dbh_dt" ="red", "none" = "black")) +
+plot_fig2 <- function(logloss_summaries) {
+  ggplot(logloss_summaries, aes(x = model_type,y = mean, group = growth_measure, fill=growth_measure, shape = model)) + 
+    geom_blank() +
+    geom_rect(aes(xmin = which(model_type == 'function_growth_comparison_base_hazard_none') -1,
+                  xmax= which(model_type == 'function_growth_comparison_base_growth_hazard_none')[-2]-0.5,
+                  ymin=-Inf, ymax=Inf), fill ='cornsilk2') +
[…]

### 2016-06-01 R/process_output.R
Added function that will combine logloss summaries in order to produce fig2

@@ -128,6 +128,41 @@ summarise_crossval_logloss <- function(comparison) {
     ungroup()
 }
 
+
+# Combines all cross val logloss outputs.
+combine_logloss_summaries <- function() {
+  logloss_func_growth <- summarise_crossval_logloss("function_growth_comparison")
+  logloss_rho_comparisons <- summarise_crossval_logloss("rho_combinations")
+  logloss_re_comparison <- summarise_crossval_logloss("species_random_effects")
+  
+  logloss_func_growth %>%
+    bind_rows(logloss_rho_comparisons, logloss_re_comparison) %>%
+    arrange(comparison, growth_measure) %>%
+    filter(model!= "base_hazard" | growth_measure !='true_basal_area_dt') %>% # removes unnecessary model base_hazard fit.
+    filter(comparison !='rho_combinations' |  rho_combo!='none') %>% # Already included in logloss_func_growth
+    mutate(growth_measure = replace(growth_measure, model=="base_hazard", "none"), # No growth for base model
+           model = replace(model, model=="base_growth_hazard_re", "base_growth_hazard")) %>% # rename for plotting purposes
+    mutate(model = factor(model, levels=c('base_hazard','growth_hazard','base_growth_hazard')),
+           growth_measure = factor(growth_measure, levels=c('none','true_basal_area_dt','true_dbh_dt')),
+           model_type = paste(comparison,model, rho_combo, sep='_')) %>%
+    mutate(model_type = factor(model_type, levels=c("function_growth_comparison_base_hazard_none",
+                                                    "function_growth_comparison_growth_hazard_none",
+                                                    "function_growth_comparison_base_growth_hazard_none",
+                                                    "rho_combinations_base_growth_hazard_a",
+                                                    "rho_combinations_base_growth_hazard_b",
+                                                    "rho_combinations_base_growth_hazard_ab",
+                                                    "rho_combinations_base_growth_hazard_c",
+                                                    "rho_combinations_base_growth_hazard_ac",
+                                                    "rho_combinations_base_growth_hazard_bc",
+                                                    "rho_combinations_base_growth_hazard_abc",
+                                                    "species_random_effects_base_growth_hazard_none"))) %>%
+    mutate(modelid = as.factor(as.numeric(model_type))) %>%
+    arrange(modelid) %>%
+    select(modelid, model_type, comparison, model, 
+           growth_measure, rho_combo,logloss, 
[…]

### 2016-06-08 ms/ms.Rnw
Added draft figures to manuscript
More figures and more model output processing

@@
 
+\begin{figure*}
+\centering
+\includegraphics{figures/fig1.pdf}
+\caption{three possible functional curves of plant mortality. Estimating growth. Cross validation proceedure}
+\label{fig1}
+\end{figure*}
+
 Wood density, the biomass invested per unit wood volume, is one such functional trait that has consistently been shown to be negatively correlated with tropical tree mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
@@
 \centering
-\includegraphics{figures/fig1.pdf}
+\includegraphics{figures/fig2.pdf}
 \caption{Average logarithmic loss $\pm$ 95\% BCIs from 10-fold cross validation.
@@
 (C) Comparing all combinations of wood density effects.}
-\label{fig1}
+\label{fig2}
 \end{figure}
 
-\subsection*{Most predictive functional form}
+\subsection*{Predictive performance as a function of model form and growth measure}
+We first investigated how model form and growth measure influenced predictive performance.
+To achieve this we first cross validated three model forms (Fig.
+\ref{fig1}) of tropical tree mortality using both dbh growth or basal area growth.
+We found that when the model accounted for both baseline and growth-dependent hazards we obtained a significant improvement in predictive capacity, relative to the other two simpler forms.
+However, this improvement was dependent on growth measure used, with predictive capacity increasing when dbh growth is used but not when basal area growth is used (Fig.
+\ref{fig2}).
 
-Full model most predictive (also most biologically reasonable)
-DBH DT most predictive given functional form
+\subsection*{Predictive performance and species effects}
+Using the most predictive model (i.e.
+that which used dbh growth and accounted for both baseline and growth-dependent hazards), we then examined how the inclusion of the effect of wood density on various combinations of parameters affected predictive performance.
+We found that the inclusion of wood density on any of the three parameters significantly improved predictive performance relative to a model without such effects.
+However, there was little differences in predictive performance across all wood density\textendash parameter combinations.
+Nonetheless, the most parsimonious model with lowest logarithmic loss was that which contained a wood density effect on only the baseline hazard ($\gamma$).
+Comparing this model to one that included species random effects on all parameters (i.e.
+the model that accounts for all species level variation) revealed that wood density accounted for only a small proportion of the maximum predictive gain achievable by accounting for species identification.
 
-\subsection*{What is the most predictive combination of parameter-wood density effects?}
+\subsection*{Effects of wood denisty}
 
-wood density on gamma. Technically significant on alpha too... but effect size is small (can't show this though as focus is on predictive capacity with cross val.)
-
-\subsection*{How well does our model predict if we allow full random effects?}
-
-Seems to account for ~20-30\% of species variability (this will be confirmed once other models are finished).
-
-Figures:
-Single figure of logloss
-figure of species gamma values vs rho
-response curve for low and high wood density species from dataset.
-
-Species variability is highest on gamma (1.2) followed by alpha (0.8) and least on beta (0.3)
-Mention proportion of species variability on gamma explained by wood density.
+\begin{figure}
+\centering
+\includegraphics{figures/fig3.pdf}
+\caption{Average logarithmic loss $\pm$ 95\% BCIs from 10-fold cross validation.
+(A) Stage 1: Comparing three functional forms of tropical tree mortality using absolute diameter growth (red) and basal area growth (blue).
+(B) Comparing model with and without species random effects.
+(C) Comparing all combinations of wood density effects.}
+\label{fig3}
+\end{figure}
 

### 2016-06-08 R/figures.R

@@ -2,8 +2,7 @@
 partial_plot_theme <- function(legend.position = "none", strips = FALSE,...) {
   sb <- if(strips==TRUE) element_rect(fill='lightgrey') else element_blank()
   st <- if(strips==TRUE) element_text(face='italic') else element_blank()
-  theme_classic(base_size = 9) + theme(strip.text = st,
-                          legend.title = element_blank(),
+  theme_classic(base_size = 7) + theme(strip.text = st,
                           strip.background = sb,
                           legend.position = legend.position,
                           axis.line.x = element_line(colour = 'black', size=0.5, linetype='solid'),
@@ -44,49 +43,6 @@ plot_obs_v_pred_growth <- function(data) {
   plot_grid(p1, p3, p2, p4, ncol=2, labels=LETTERS[1:4], label_size = 7)
 }
 
-plot_fig2 <- function(logloss_summaries) {
-  ggplot(logloss_summaries, aes(x = model_type,y = mean, group = growth_measure, fill=growth_measure, shape = model)) + 
-    geom_blank() +
-    geom_rect(aes(xmin = which(model_type == 'function_growth_comparison_base_hazard_none') -1,
-                  xmax= which(model_type == 'function_growth_comparison_base_growth_hazard_none')[-2]-0.5,
-                  ymin=-Inf, ymax=Inf), fill ='cornsilk2') +
-    geom_rect(aes(xmin = which(model_type == 'function_growth_comparison_base_growth_hazard_none')[-2]-0.5,
-                  xmax= which(model_type == 'rho_combinations_base_growth_hazard_bc')-0.5,
-                  ymin=-Inf, ymax=Inf), fill ='cornsilk') +
-    geom_pointrange(aes(ymin = `2.5%`, ymax=`97.5%`), position=position_dodge(.5)) +
-    ylab('Logarithmic Loss') + 
-    xlab('Model') +
-    scale_shape_manual(values = c(21, 22, 24)) +
-    scale_fill_manual(values =c('white','grey80','black')) +
-    scale_x_discrete(labels=c("function_growth_comparison_base_hazard_none" =  expression(gamma),
-                              "function_growth_comparison_growth_hazard_none" = expression(alpha*"e"^{-beta~"X"["i"]}),
[…]

### 2016-06-08 R/process_output.R

@@ -52,7 +52,7 @@ diagnose <- function(model) {
       n_divergent = sum(sapply(sampler_params, function(y) y[,'n_divergent__'])),
       max_treedepth = max(sapply(sampler_params, function(y) y[,'treedepth__'])))
   }))
-
+  
   out2 <- suppressWarnings(bind_rows(lapply(info, function(x) {
     data.frame(
       comparison = x$comparison,
@@ -100,9 +100,9 @@ logloss_samples <- function(model) {
 # Extract log loss samples for all models.
 extract_logloss_samples <- function(model) {
   if(is.null(model$fits)) { #Check to see if object is multi model 
-  samples <- lapply(model, logloss_samples)
-  plyr::ldply(samples, .id='modelid') %>%
-    select(-modelid)
+    samples <- lapply(model, logloss_samples)
+    plyr::ldply(samples, .id='modelid') %>%
+      select(-modelid)
   }
   else { 
     logloss_samples(model)
@@ -119,7 +119,7 @@ summarise_crossval_logloss <- function(comparison) {
     group_by(comparison, model, growth_measure, rho_combo, kfold, logloss) %>%
     summarise(kfold_logloss = mean(estimate)) %>%
     ungroup() %>%
-     group_by(comparison, model, growth_measure, rho_combo, logloss) %>%
+    group_by(comparison, model, growth_measure, rho_combo, logloss) %>%
     summarise(mean = mean(kfold_logloss),
               st_err = sd(kfold_logloss)/sqrt(n())) %>%
     mutate(ci = 1.96 * st_err,
@@ -127,7 +127,6 @@ summarise_crossval_logloss <- function(comparison) {
            `97.5%` = mean + ci) %>%
     ungroup()
 }
-
 
 # Combines all cross val logloss outputs.
 combine_logloss_summaries <- function() {
@@ -163,9 +162,11 @@ combine_logloss_summaries <- function() {
            mean, st_err,ci, `2.5%`,`97.5%`)
 }
 
+
+
 #
 get_times <- function(comparison) {
-   fits <- comparison$fits
+  fits <- comparison$fits
   info <- plyr::ldply(comparison$model_info, .id='modelid')
   times <- lapply(fits, function(x) 
     rstan::get_elapsed_time(x))
@@ -185,3 +186,80 @@ summarise_times <- function(times) {
     summarise(mn = median(total_hours))
   return(res)
 }
+
+
[…]

### 2016-06-08 R/stan_functions.R

@@ -184,14 +184,7 @@ prep_kfold_data_for_stan <- function(data, growth_measure) {
 }
 
 # Prepares full dataset for use with stan
-prep_full_data_for_stan <- function(data, growth_measure) {
-    switch(growth_measure,
-         'true_dbh_dt' = {
-           growth_dt <- data$true_dbh_dt - 0.172
-           },
-         'true_basal_area_dt' = {
-           growth_dt <- data$true_basal_area_dt - 0.338
-         })
+prep_full_data_for_stan <- function(data) {
   list(
     n_obs = nrow(data),
     n_census = max(data$censusid),
@@ -199,10 +192,11 @@ prep_full_data_for_stan <- function(data, growth_measure) {
     census = data$censusid,
     spp = data$sp_id,
     census_length = data$census_interval,
-    growth_dt = growth_dt,
+    growth_dt = data$true_dbh_dt - 0.172,
     rho_c  = unique(data$rho)/0.6,
-    y = as.integer(data$dead_next_census)
-  )
+    y = as.integer(data$dead_next_census),
+    sp = unique(data$sp),
+    raw_rho = unique(data$rho))
 }
 
 # Wrapper function to call either kfold or full data for stan
@@ -211,7 +205,7 @@ prep_data_for_stan <- function(data, growth_measure, crossval) {
     prep_kfold_data_for_stan(data, growth_measure)
   }
   else {
-    prep_full_data_for_stan(data, growth_measure)
+    prep_full_data_for_stan(data)
   }
 }
 

### 2016-06-10 R/figures.R
Draft schematic for fig 1

@@ -1,3 +1,17 @@
+# position label at a fractional x/y position on a plot
+label <- function(px, py, lab, ..., adj = c(0, 1)) {
+  usr <- par("usr")
+  x <- usr[1] + px * (usr[2] - usr[1])
+  y <- usr[3] + py * (usr[4] - usr[3])
+
+  if (par("ylog"))
+    y <- 10^y
+  if (par("xlog"))
+    x <- 10^x
+
+  text(x, y, lab, adj = adj, xpd=NA, ...)
+}
+
 # Plot theme
 partial_plot_theme <- function(legend.position = "none", strips = FALSE,...) {
   sb <- if(strips==TRUE) element_rect(fill='lightgrey') else element_blank()
@@ -42,6 +56,7 @@ plot_obs_v_pred_growth <- function(data) {
   
   plot_grid(p1, p3, p2, p4, ncol=2, labels=LETTERS[1:4], label_size = 7)
 }
+
 
 plot_fig2a <- function(logloss_summaries) {
   dat <- logloss_summaries %>%
@@ -135,40 +150,108 @@ plot_mu_curves <- function(model,wood_density=c(0.3,0.8), growth_range = c(0.03,
 
 # Manuscript figures
 
-plot_fig1 <- function() {
+plot_fig1 <- function(tree1, tree2, panelc) {
   growth_haz <- function(x) {
     1-exp(-0.15 * exp(-10 * x))
   }
-  
+
   base_growth_haz <- function(x) {
     1-exp(-(0.13 * exp(-10 * x) + 0.03))
   }
-  
-  layout(matrix(c(1,2,3,4,4,4,5,5,5), byrow=TRUE, ncol=3))
-  par(mar=c(1.3,1.3,1.3,0.1),ps = 7)
+
+  my_label <- function(text, x=-0.1) label(x, 1.3, text, cex=1.5)
+
+  layout(matrix(c(1,1,1,2,3,4,5,5,5), byrow=TRUE, ncol=3))
+  par(mar=c(2,2,3,1), oma=c(1,1,1,1))
+
+ # Tree growth diagram
+
+  plot(1:2, type='n', ann=FALSE, axes=FALSE, xlim=c(-1,1), ylim=c(-1,1))
+  my_label("a) Repeat census data")
+
+  # note we are only plotting to last viewport, but calls to other two needed to make figure work.
+  vps <- baseViewports()
[…]

### 2016-06-10 ms/ms.Rnw
Draft schematic for fig 1
Updates to manuscript and figures

@@
 
-\begin{figure*}
+\begin{figure}
 \centering
 \includegraphics{figures/fig1.pdf}
-\caption{three possible functional curves of plant mortality. Estimating growth. Cross validation proceedure}
+\caption{Outline of methodology. A) Three functional forms of plant mortality. B) Past growth used to predict future mortality. C) Cross validation proceedure}
 \label{fig1}
-\end{figure*}
+\end{figure}
 
@@
 \includegraphics{figures/fig2.pdf}
-\caption{Average logarithmic loss $\pm$ 95\% BCIs from 10-fold cross validation.
-(A) Stage 1: Comparing three functional forms of tropical tree mortality using absolute diameter growth (red) and basal area growth (blue).
-(B) Comparing model with and without species random effects.
-(C) Comparing all combinations of wood density effects.}
+\caption{Average logarithmic loss from 10-fold cross validation.
+(A) Example logarithmic loss curve for a single observed death.
+B) Comparing various functional forms of tropical tree mortality with different growth measures different (white = no growth, grey = basal area growth, black = dbh growth) and species effects.
+(C) Comparing model forms with various combinations of wood density effects.}
 \label{fig2}
@@
 
+\begin{figure}
+\centering
+\includegraphics{figures/fig3.pdf}
+\caption{Estimated log baseline mortality against wood density. Points are individual species and solid line is the mean trendline. All errors are $\pm$ 95\% BCI.}
+\label{fig3}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{figures/fig4.pdf}
+\caption{Estimated curves of instantaneous mortality rate and annual mortality probability.
+A,C) Highlight expected response curves for a hypothetical median species at low (0.3) and high (0.8) wood density.
+B,D) Predicted response curves for all species.
+Errors are $\pm$ 95\% BCI.}
+\label{fig4}
+\end{figure}
+
 \subsection*{Predictive performance and species effects}
@@
 \subsection*{Effects of wood denisty}
-
-\begin{figure}
-\centering
-\includegraphics{figures/fig3.pdf}
-\caption{Average logarithmic loss $\pm$ 95\% BCIs from 10-fold cross validation.
-(A) Stage 1: Comparing three functional forms of tropical tree mortality using absolute diameter growth (red) and basal area growth (blue).
-(B) Comparing model with and without species random effects.
-(C) Comparing all combinations of wood density effects.}
-\label{fig3}
-\end{figure}
 
@@
 was supported by an ARC discovery grant.
-We thank Lachlan McCalman and Daniel Steinberg for statistical advice; Anias Gibert etc etc.}
+We thank Lachlan McCalman and Daniel Steinberg for statistical advice; Anias Gibert etc etc.
+Fig.
+1a uses images by Tracey Saxby, IAN Image Library (http://ian.umces.edu/imagelibrary)}
 

### 2016-06-10 R/figures.R

@@ -253,9 +253,10 @@ fig.tree <- function(file_alive, file_dead) {
   grid.text(expression(paste(t[3])), x = x0, y = 0, just="top", gp=gp0)
 }
 plot_fig2 <- function(logloss_summaries) {
-  p1 <- plot_fig2a(logloss_summaries)
-  p2 <- plot_fig2b(logloss_summaries)
-  plot_grid(p1,p2, ncol=1, labels=LETTERS[1:2], label_size = 7)
+  p1 <- logloss_curve()
+  p2 <- plot_fig2a(logloss_summaries)
+  p3 <- plot_fig2b(logloss_summaries)
+  plot_grid(p1,p2,p3, ncol=1, labels=LETTERS[1:3], label_size = 7)
 }
 
 # Plot gamma vs wood density with mean trend line
@@ -265,7 +266,7 @@ med <- predict_mu_baseline_hazard(model, data)
 
 ggplot(spp, aes(x = wood_density,y = log(mean))) + 
       geom_pointrange(aes(ymin = log(`2.5%`), ymax=log(`97.5%`)), size=0.1) +
-      geom_ribbon(data = med, aes(ymin = log(`2.5%`),ymax = log(`97.5%`)), alpha=0.4, colour=NA) +
+      geom_ribbon(data = med, aes(ymin = log(`2.5%`),ymax = log(`97.5%`)), alpha=0.2, colour=NA) +
       geom_line(data = med, aes(x = wood_density, y= log(mean))) +
       partial_plot_theme() +
      ylab("log(instantaneous mortality rate)") +
@@ -294,4 +295,16 @@ plot_fig4 <- function(model, data) {
 #   partial_plot_theme()
 #}
 
+logloss_curve <- function() {
+  df <- data.frame(x = seq(0.001,1,length.out = 100))
+  logloss <- function(x, eps = 1e-15) {
+    predicted = pmin(pmax(x, eps), 1-eps)
+    - (1 * log(predicted) + (1 - 1) * log(1 - predicted))
+  }
+  ggplot(df, aes(x)) +
+    stat_function(fun = logloss) +
+    xlab("Predicted probability") +
+    ylab('Logarithmic loss') +
+  partial_plot_theme()
+}
 

### 2016-06-10 R/process_output.R

@@ -1,3 +1,11 @@
+# Calculate logloss outside of stan
+logloss = function(actual, predicted, eps = 1e-15) {
+   predicted = pmin(pmax(predicted, eps), 1-eps)
+   - (actual * log(predicted) + (1 - actual) * log(1 - predicted))
+}
+
+
+
 # Extracts optimization estimates
 extract_true_dbh_estimates <- function(optimization_results) {
   fit <- optimization_results$par
@@ -278,6 +286,29 @@ predict_mu_baseline_hazard <- function(model,data) {
               `97.5%` = quantile(inst_hazard, 0.975)) %>%
     ungroup()
 }
+
+# Predict observations outside stan
+predict_observations <- function(model, data) {
+spp_effects <- summarise_spp_params(model, data)
+wd_effects <- median(rstan::extract(model$fits[[1]], pars=c('c1'))$c1)
+census_error <- as.data.frame(rstan::extract(model$fits[[1]], pars=c('census_err'))$census_err) %>%
+  summarise_each(funs(median)) %>%
+  gather('censusid','census_err') %>%
+  mutate(censusid =c(1,2,3))
+
+plyr::ldply(spp, .id='parameter') %>%
+  select(parameter,sp, wood_density, median) %>%
+  spread(parameter, median) %>%
+  merge(data, by.all=sp) %>%
+  merge(census_error, by.all=censusid) %>%
+  mutate(wd_effects = wd_effects) %>%
+  ungroup() %>%
+  mutate(hazard_rate = (alpha * exp(-beta * (true_dbh_dt - 0.172)) + gamma*(rho/0.6)^wd_effects)*census_err,
+         prob_death = 1-exp(-census_interval * (alpha * exp(-beta * (true_dbh_dt - 0.172)) + gamma * (rho/0.6)^wd_effects)*census_err)) %>%
+  mutate(logloss = logloss(dead_next_census,prob_death))
+}
+
+
 
 
 

### 2016-06-14 ms/ms.Rnw
Added null model and made several changes to manuscript

--- 2016-05-31 ms/ms.Rnw
@@
 n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))
+
 @
@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics{figures/fig1.pdf}
+\caption{Outline of methodology. A) Past growth used to predict future mortality. B) Three functional forms of plant mortality. C) Cross validation proceedure}
+\label{fig1}
+\end{figure}
+
 In this study, we reconcile both intraspecific and interspecific theory by developing a unfied model that incorporates both individual growth and the species level trait, wood density.
@@
 
-\begin{figure}
-\centering
-\includegraphics{figures/fig1.pdf}
-\caption{Average logarithmic loss $\pm$ 95\% BCIs from 10-fold cross validation.
-(A) Stage 1: Comparing three functional forms of tropical tree mortality using absolute diameter growth (red) and basal area growth (blue).
-(B) Comparing model with and without species random effects.
-(C) Comparing all combinations of wood density effects.}
-\label{fig1}
+\subsection*{Predictive capacity of functional form and growth measure}
+Using 10-fold cross validation our first aim was to determine which model form (Fig.
+\ref{fig1}b) and growth measure (i.e.
+dbh growth or basal area growth) best predicted tree mortality.
+We found that the model which accounted for both baseline (i.e.
+growth-independent) and growth-dependent hazards had a significantly lower logarithmic loss, and thus, a higher predictive capacity, relative to either the baseline only or growth-dependent only model variants (Fig.
+\ref{fig2}a).
+However, this improvement was dependent on growth measure, with predictive capacity increasing when dbh growth was used but not when basal area growth was used (Fig.
+\ref{fig2}a).
+
+\begin{figure}[ht]
+\includegraphics{figures/fig2.pdf}
+\caption{Average logarithmic loss from 10-fold cross validation.
+A) Comparing various functional forms of tropical tree mortality with different growth measures different (white = no growth, grey = basal area growth, black = dbh growth) and species effects.
+B) Comparing model forms with various combinations of wood density effects.}
+\label{fig2}
 \end{figure}
 
-\subsection*{Most predictive functional form}
+\subsection*{Predictive capacity from including wood density}
+Using the model with both baseline and growth-dependent hazards coupled with dbh growth we then examined: 1) whether predictive capacity increased if parameters varied as a function of a species' wood density; 2) whether the effect of wood density was predominately on growth-dependent or baseline (growth-independent) hazard parameters; and 3) what proportional gain is achieved from including wood density relative to a full species random effect model that implicitly captures all species level variation.
+We found that the inclusion of wood density on any combination of the three parameters significantly improved predictive performance (i.e.
+lowered logarithmic loss) relative to a model without wood density effects (Fig.
+\ref{fig2}b).
+Furthermore, the greatest gain in predictive accuracy occurred when we allowed the baseline hazard $\gamma$ to vary by wood density, which accounted for \Sexpr{logloss_comparisons(logloss_summaries)[['rho_v_re']]}\% of the maximum predictive gain achievable in the fall species random effect model (Fig.
+\ref{fig2}a).
 
-Full model most predictive (also most biologically reasonable)
-DBH DT most predictive given functional form
+\begin{figure}[ht]
+\includegraphics{figures/fig3.pdf}
+\caption{Estimated log baseline mortality against wood density. Points are individual species and solid line is the mean trendline. All errors are $\pm$ 95\% BCI.}
+\label{fig3}
+\end{figure}
 
-\subsection*{What is the most predictive combination of parameter-wood density effects?}
+\subsection*{Effects of wood density and time}
+Once determining that the most predictive model was one that accounted for both growth-dependent and baseline hazards, used dbh growth, and allowed $\gamma$ to vary by wood density, we then fitted two final models to the entire dataset.
+The first contained both species random effects on all parameters and wood density on the baseline hazard, $\gamma$.
+This allowed us to examine the range of species growth-mortality curves, the effect of wood density and how mortality varied over time.
+We also fitted a second model that only contained species random effects.
+This allowed us to determine which parameter varied most by species and what proportion of the baseline hazard species variability was explained by wood density.
+We found that baseline hazard rates were significantly lower for species with high wood density relative to those with low wood density (Fig.
+\ref{fig3}).
+That is, high wood density species had a lower probability of dying irrespective of growth rate.
+For example, our model predicts that on average a species with a wood density of 0.3 has a instantaneous mortality rate of XX, whereas a species with a wood density of 0.8 has a instantenaous mortality rate of XX, translating to an annual mortality rate of XX\% and XX\%, respectively.
+Furthermore, our model also revealed that instantaneous mortality rates varied over time.
+Specifically, mortality rates at BCI decreased over time with hazard rates being scaled by \Sexpr{census_effects(final_model,1)},\Sexpr{census_effects(final_model,2)} and \Sexpr{census_effects(final_model,3)} for census periods, 1995\textendash 2000, 2000\textendash 2005 and 2005\textendash 2010, respectively.
 
-wood density on gamma. Technically significant on alpha too... but effect size is small (can't show this though as focus is on predictive capacity with cross val.)
+\begin{figure}[ht]
+\includegraphics{figures/fig4.pdf}[ht]
+\caption{Estimated curves of instantaneous mortality rate and annual mortality probability.
+A,C) Highlight expected response curves for a hypothetical median species at low (0.3) and high (0.8) wood density.
+B,D) Predicted response curves for all species.
+Errors are $\pm$ 95\% BCI.}
+\label{fig4}
+\end{figure}
 
-\subsection*{How well does our model predict if we allow full random effects?}
-
-Seems to account for ~20-30\% of species variability (this will be confirmed once other models are finished).
-
-Figures:
-Single figure of logloss
-figure of species gamma values vs rho
-response curve for low and high wood density species from dataset.
-
-Species variability is highest on gamma (1.2) followed by alpha (0.8) and least on beta (0.3)
-Mention proportion of species variability on gamma explained by wood density.
+\subsection*{Sources of variation}
+We found that the baseline hazard, $\gamma$, varied the most between species followed by $\alpha$ and least on $\beta$.
+We also found that the wood density effect on $\gamma$ explained approximately XX\% of the species variability on the baseline hazard.
 

### 2016-06-14 R/figures.R

@@ -167,7 +167,7 @@ plot_fig1 <- function(tree1, tree2, panelc) {
  # Tree growth diagram
 
   plot(1:2, type='n', ann=FALSE, axes=FALSE, xlim=c(-1,1), ylim=c(-1,1))
-  my_label("a) Repeat census data")
+  my_label("A) Repeat census data")
 
   # note we are only plotting to last viewport, but calls to other two needed to make figure work.
   vps <- baseViewports()
@@ -190,7 +190,7 @@ plot_fig1 <- function(tree1, tree2, panelc) {
   }
 
   hazard_plot(function(x) 0*x +0.03, expression(gamma))
-  my_label("b) Alternative mortality functions", x=-0.45)
+  my_label("B) Alternative mortality functions", x=-0.45)
   mtext("Mortality rate", 2, line=1, cex=0.75)
 
  # Example growth-dependent hazard
@@ -200,7 +200,7 @@ plot_fig1 <- function(tree1, tree2, panelc) {
   hazard_plot(base_growth_haz, expression(gamma + alpha*"e"^{-beta~"X"}))
 
   plot(NA, xlim = c(0, 1), ylim= c(0, 1), ann = FALSE, axes=FALSE)
-  my_label("c) 10-fold cross validation")
+  my_label("C) 10-fold cross validation")
 
   vps <- baseViewports()
   pushViewport(vps$inner, vps$figure, vps$plot)
@@ -253,10 +253,9 @@ fig.tree <- function(file_alive, file_dead) {
   grid.text(expression(paste(t[3])), x = x0, y = 0, just="top", gp=gp0)
 }
 plot_fig2 <- function(logloss_summaries) {
-  p1 <- logloss_curve()
-  p2 <- plot_fig2a(logloss_summaries)
-  p3 <- plot_fig2b(logloss_summaries)
-  plot_grid(p1,p2,p3, ncol=1, labels=LETTERS[1:3], label_size = 7)
+  p1 <- plot_fig2a(logloss_summaries)
+  p2 <- plot_fig2b(logloss_summaries)
+  plot_grid(p1,p2, ncol=1, labels=LETTERS[1:2], label_size = 7)
 }
 
 # Plot gamma vs wood density with mean trend line
@@ -275,7 +274,7 @@ ggplot(spp, aes(x = wood_density,y = log(mean))) +
 
 plot_fig4 <- function(model, data) {
[…]

### 2016-06-14 R/model.R

@@ -61,6 +61,77 @@ run_true_dbh_model <- function(data) {
   fit <- optimizing(model,
                     data = stan_data, iter=50000, refresh=1000)
 }
+
+# Null Model
+get_model_chunks_null <- function(tasks) {
+  
+  list(
+    growth_measure = tasks$growth_measure,
+    pars = c("log_gamma",
+             "logloss_heldout"),
+    data = "
+        // Fitted data
+        int<lower=1> n_obs;
+        vector[n_obs] census_length;
+        int<lower=0, upper=1> y[n_obs];
+        
+        // Heldout data
+        int<lower=1> n_obs_heldout;
+        vector[n_obs_heldout] census_length_heldout;
+        int<lower=0, upper=1> y_heldout[n_obs_heldout];",
+    parameters ="
+      real log_gamma;",
+    model ="
+      real cumulative_hazard;
+      real gamma;
+      
+      // Put gamma on normal scale
+      gamma <- exp(log_gamma);
+
+      for (i in 1:n_obs) {
+        cumulative_hazard <- -census_length[i] * gamma;
+        
+        if (y[i] == 0) {
+          increment_log_prob(cumulative_hazard);
+        } else {
+          increment_log_prob(log1m_exp(cumulative_hazard));
+        }
+      }
+      // Priors
+      log_gamma ~ normal(0, 2.5);",
+    generated_quantities ="
+      real gamma;
+      
+      // Declaring heldout
+      real cumulative_hazard_heldout;
+      real loglik_heldout;
+      real sum_loglik_heldout;
+      real logloss_heldout;
+
+      // Initialization of summed parameter
+      sum_loglik_heldout <- 0;
+
+      // Put normal on normal scale;
+      gamma <- exp(log_gamma);
+      
+      // Calculate log likelihood for heldout data
+      for (j in 1:n_obs_heldout) {
+        cumulative_hazard_heldout <- -census_length_heldout[j] * gamma;
+        
+        if (y_heldout[j] == 0) {
+          loglik_heldout <- cumulative_hazard_heldout;
+        }
+        else {
+          loglik_heldout <- log1m_exp(cumulative_hazard_heldout);
+        }
+        sum_loglik_heldout <- sum_loglik_heldout + loglik_heldout;
+      }
+        // Calculation of average negative log likelihood
+        logloss_heldout <- -sum_loglik_heldout/n_obs_heldout;"
+  )
+}
+
+
 ## baseline hazard model
 
 get_model_chunks_base_haz <- function(tasks) {

### 2016-06-14 R/process_output.R

@@ -309,6 +309,35 @@ plyr::ldply(spp, .id='parameter') %>%
 }
 
 
+predict_levels <- function(model, data) {
+spp_effects <- summarise_spp_params(model, data)
+wd_effects <- median(rstan::extract(model$fits[[1]], pars=c('c1'))$c1)
+census_error <- as.data.frame(rstan::extract(model$fits[[1]], pars=c('census_err'))$census_err) %>%
+  summarise_each(funs(median)) %>%
+  gather('censusid','census_err') %>%
+  mutate(censusid =c(1,2,3))
+
+test <-plyr::ldply(spp_effects, .id='parameter') %>%
+  select(parameter,sp, wood_density, median) %>%
+  spread(parameter, median) %>%
+  merge(data, by.all=sp) %>%
+  merge(census_error, by.all=censusid) %>%
+  mutate(wd_effects = wd_effects) %>%
+  ungroup() %>%
+  mutate(prob_death_all = 1-exp(-census_interval * (alpha * exp(-beta * (true_dbh_dt - 0.172)) + gamma * (rho/0.6)^wd_effects)*census_err),
+         prob_death_spp = 1-exp(-census_interval * (alpha * exp(-beta * (true_dbh_dt - 0.172)) + gamma)),
+         prob_death_rho = 1-exp(-census_interval * (mean(alpha) * exp(-mean(beta) * (true_dbh_dt - 0.172)) + mean(gamma) * (rho/0.6)^wd_effects)),
+         prob_death_cen = 1-exp(-census_interval * (mean(alpha) * exp(-mean(beta) * (true_dbh_dt - 0.172)) + mean(gamma)*census_err))) %>%
+  summarise(
+    spp_rho_census = sum((prob_death_all - mean(prob_death_all))^2),
+    spp = sum((prob_death_spp - mean(prob_death_spp))^2),
+    wood_density = sum((prob_death_rho - mean(prob_death_rho))^2),
+    census = sum((prob_death_cen - mean(prob_death_cen))^2)) %>%
+      mutate(total_sum_sq = sum(unexplained, spp, wood_density, census)) %>%
+  gather(level, sum_res_sq, -total_sum_sq) %>%
+  mutate(proportions = sum_res_sq/total_sum_sq)
+}
+
 
 
 

### 2016-06-14 R/stan_functions.R

@@ -3,6 +3,11 @@ kfold_tasks <- function(comparison,iter=4000,path='.') {
   n_kfolds = 10
   n_chains = 3
   switch(comparison,
+         "null_model" = {
+           growth_measure <- "true_dbh_dt"
+           rho_combo <- "none"
+           model  <- "null_model"
+         }, 
          "function_growth_comparison" = {
            growth_measure <- c("true_dbh_dt",'true_basal_area_dt')
            rho_combo <- "none"
@@ -89,6 +94,10 @@ model_compiler <- function(task) {
   model <- task$model
   ## Assemble the stan model:
   switch(model, 
+         "null_model"= {
+           chunks <- get_model_chunks_null(task)
+           chunks$crossval <- TRUE
+         },
          "base_hazard"= {
            chunks <- get_model_chunks_base_haz(task)
            chunks$crossval <- TRUE
@@ -240,6 +249,10 @@ precompile <- function(task) {
   model <- task$model
   ## Assemble the stan model:
   switch(model, 
+         "null_model"= {
+           chunks <- get_model_chunks_null(task)
+           chunks$crossval <- TRUE
+         },
          "base_hazard"= {
            chunks <- get_model_chunks_base_haz(task)
            chunks$crossval <- TRUE
@@ -291,17 +304,21 @@ precompile <- function(task) {
 
 #THIS ISN'T ELEGANT BUT IT WORKS
 precompile_crossval_models <- function() {
-  # Functional form/growth comparison
-  stage1 <- tasks_2_run(comparison = 'function_growth_comparison',iter = 10)
+  # Null model
+  stage1 <- tasks_2_run(comparison = 'null_model',iter = 10)
   vapply(df_to_list(stage1), precompile, character(1))
   
-  #Constant vs species random effects
-  stage2 <- tasks_2_run(comparison = 'species_random_effects',iter = 10)
+  # Functional form/growth comparison
+  stage2 <- tasks_2_run(comparison = 'function_growth_comparison',iter = 10)
   vapply(df_to_list(stage2), precompile, character(1))
   
-  #Rho combinations
-  stage3 <- tasks_2_run(comparison = 'rho_combinations',iter = 10)
+  #Constant vs species random effects
+  stage3 <- tasks_2_run(comparison = 'species_random_effects',iter = 10)
   vapply(df_to_list(stage3), precompile, character(1))
+  
+  #Rho combinations
[…]

### 2016-06-14 R/manuscript_functions.R
Functions to extract values for manuscript

@@ -1,0 +1,37 @@
+
+logloss_comparisons <- function(logloss_summary) {
+  logloss_summary %>%
+    filter(model_type %in% c("function_growth_comparison_base_growth_hazard_none", 
+                             "rho_combinations_base_growth_hazard_c",
+                             "species_random_effects_base_growth_hazard_none") & growth_measure=='true_dbh_dt') %>%
+    select(model_type, mean) %>%
+    mutate(model_type = factor(model_type, labels =c('base_growth','base_growth_rho','base_growth_re'))) %>%
+    spread(model_type, mean) %>%
+    mutate(rho_v_base = round((1-base_growth_rho/base_growth)*100,2),
+           re_v_base = round((1-base_growth_re/base_growth)*100,2),
+           rho_v_re = round((rho_v_base/re_v_base)*100,2))
+}
+
+census_effects <- function(model,census) {
+  model <- model$fits[[1]]
+  get_posterior_mean(model, 'census_err')[census,4]
+}
+
+# Predict median baseline hazard
+predict_point_mu_baseline_hazard <- function(model,data, wood_density) {
+  fit <- model$fits[[1]]
+  samples <- rstan::extract(fit, pars=c("mu_log_gamma","c1"))
+  samples <- as.data.frame(lapply(samples, as.vector)) 
+  covariates <- data.frame(wood_density = wood_density,
+                           wood_density_centered = wood_density/0.6)
+  samples %>%
+    merge(covariates) %>%
+    mutate(inst_hazard = exp(mu_log_gamma) * wood_density_centered^c1,
+           annual_mort = 1-exp(-exp(mu_log_gamma) * wood_density_centered^c1)) %>%
+    gather('prediction','estimate', -mu_log_gamma,-c1,-wood_density,-wood_density_centered) %>%
+    group_by(prediction) %>%
+    summarise(mean = mean(estimate),
+              `2.5%` = quantile(estimate,0.025),
+              `97.5%` = quantile(estimate, 0.975)) %>%
+    ungroup()
+}

### 2016-06-14 ms/ms.Rnw
Standardise variable names, terms and spelling throughout
- well, attempt to at least

@@
-\documentclass[9pt,twocolumn,twoside, lineno]{pnas-new}
+\documentclass[9pt,twocolumn,twoside]{pnas-new}
 % Use the lineno option to display guide line numbers if required.
@@
 % Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\affil[a]{Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia}
+\affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
 
@@
 designed and performed the analysis; D.F.
-conceived the idea.
-M.W.
-supervised the development of this work and provided biological and statistical advice.
+conceived the idea, supervized the development of this work; M.W.
+provided biological insight.
 All authors contributed to the writing of this manuscript.}
@@
 % Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
-% You should only change this length when you've finalised the article contents.
+% You should only change this length when you've finalized the article contents.
 \verticaladjustment{-2pt}
@@
 This is despite growing empirical evidence highlighting that mortality rates vary considerably between species, even within the same community \cite{Kobe:1997vy, Wyckoff:2002ul, Kunstler:2005bn, Vieilledent:2010fv}.
-Furthermore, because no systematic analysis has been conducted to determine what growth measure best predicts plant mortality, the growth-mortality relationship is often parameterised using a variety growth measures; the two most common being rates of dbh or basal area growth.
+Furthermore, because no systematic analysis has been conducted to determine what growth measure best predicts plant mortality, the growth-mortality relationship is often parameterized using a variety growth measures; the two most common being rates of DBH or basal area growth.
 
@@
 In this study, we reconcile both intraspecific and interspecific theory by developing a unfied model that incorporates both individual growth and the species level trait, wood density.
-We extend existing individual-based approaches \cite[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship within a single model (hereon in referred to as growth-dependent hazard).
+We extend existing individual-based approaches \cite[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship within a single model (hereafter referred to as growth-dependent hazard).
 We then compare the predictive performance of this model to other traditionally used mortality models.
-We parameterise these models using 15-years of individual growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals.
+We parameterize these models using 15-years of individual growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals.
 Using these models, coupled with 10-fold cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to baseline only or growth-dependent only models?
-2) is dbh growth or basal area growth more predicitive of tropical tree mortality and does it's predictive capacity vary with model type?
+2) is DBH growth or basal area growth more predictive of tropical tree mortality and does it's predictive capacity vary with model type?
 3) Does allowing both hazards to vary by species improve predictive capacity?
@@
 \ref{fig1}b) and growth measure (i.e.
-dbh growth or basal area growth) best predicted tree mortality.
+DBH growth or basal area growth) best predicted tree mortality.
 We found that the model which accounted for both baseline (i.e.
@@
 \ref{fig2}a).
-However, this improvement was dependent on growth measure, with predictive capacity increasing when dbh growth was used but not when basal area growth was used (Fig.
+However, this improvement was dependent on growth measure, with predictive capacity increasing when DBH growth was used but not when basal area growth was used (Fig.
 \ref{fig2}a).
@@
 \caption{Average logarithmic loss from 10-fold cross validation.
-A) Comparing various functional forms of tropical tree mortality with different growth measures different (white = no growth, grey = basal area growth, black = dbh growth) and species effects.
+A) Comparing various functional forms of tropical tree mortality with different growth measures different (white = no growth, grey = basal area growth, black = DBH growth) and species effects.
 B) Comparing model forms with various combinations of wood density effects.}
@@
 \subsection*{Predictive capacity from including wood density}
-Using the model with both baseline and growth-dependent hazards coupled with dbh growth we then examined: 1) whether predictive capacity increased if parameters varied as a function of a species' wood density; 2) whether the effect of wood density was predominately on growth-dependent or baseline (growth-independent) hazard parameters; and 3) what proportional gain is achieved from including wood density relative to a full species random effect model that implicitly captures all species level variation.
+Using the model with both baseline and growth-dependent hazards coupled with DBH growth we then examined: 1) whether predictive capacity increased if parameters varied as a function of a species' wood density; 2) whether the effect of wood density was predominately on growth-dependent or baseline (growth-independent) hazard parameters; and 3) what proportional gain is achieved from including wood density relative to a full species random effect model that implicitly captures all species level variation.
 We found that the inclusion of wood density on any combination of the three parameters significantly improved predictive performance (i.e.
@@
 \includegraphics{figures/fig3.pdf}
-\caption{Estimated log baseline mortality against wood density. Points are individual species and solid line is the mean trendline. All errors are $\pm$ 95\% BCI.}
+\caption{Estimated log baseline mortality against wood density. Points are individual species and solid line is the mean trend line. All errors are $\pm$ 95\% BCI.}
 \label{fig3}
@@
 \subsection*{Effects of wood density and time}
-Once determining that the most predictive model was one that accounted for both growth-dependent and baseline hazards, used dbh growth, and allowed $\gamma$ to vary by wood density, we then fitted two final models to the entire dataset.
+Once determining that the most predictive model was one that accounted for both growth-dependent and baseline hazards, used DBH growth, and allowed $\gamma$ to vary by wood density, we then fitted two final models to the entire dataset.
 The first contained both species random effects on all parameters and wood density on the baseline hazard, $\gamma$.
@@
 That is, high wood density species had a lower probability of dying irrespective of growth rate.
-For example, our model predicts that on average a species with a wood density of 0.3 has a instantaneous mortality rate of XX, whereas a species with a wood density of 0.8 has a instantenaous mortality rate of XX, translating to an annual mortality rate of XX\% and XX\%, respectively.
+For example, our model predicts that on average a species with a wood density of 0.3 has a instantaneous mortality rate of XX, whereas a species with a wood density of 0.8 has a instantaneous mortality rate of XX, translating to an annual mortality rate of XX\% and XX\%, respectively.
 Furthermore, our model also revealed that instantaneous mortality rates varied over time.
@@
 Could it be that DBH implicitly punishes smaller individuals relative to larger individuals.
-That is, small plants can only gain a small amount of dbh growth relative to a larger plant.
-As such, dbh growth implicitly captures the process whereby small individual plants are more prone to competitive effects as well as random breakage events caused by trampling, herbivory etc.
+That is, small plants can only gain a small amount of DBH growth relative to a larger plant.
+As such, DBH growth implicitly captures the process whereby small individual plants are more prone to competitive effects as well as random breakage events caused by trampling, herbivory etc.
 
@@
 
-- When examining this interspecific variability we found that the effect of wood density improved model predictive capcity when it was included on the base-line hazard, with only marginal improvements when included on all three parameters.
+- When examining this interspecific variability we found that the effect of wood density improved model predictive capacity when it was included on the base-line hazard, with only marginal improvements when included on all three parameters.
 
@@
 \subsection*{Data}
-We parameterise plant mortality models using individual growth and survival data \cite{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\textdegree;N, 79.85\textdegree;W).
+We parametrized plant mortality models using individual growth and survival data \cite{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\textdegree;N, 79.85\textdegree;W).
 The climate on the island is warm throughout the year with a mean annual temperature of 26.1\textdegree and rainfall is seasonal with most of the 2500 mm falling between April and November.
@@
 
-Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and contained a diameter at breast height greater or equal to 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter \cite{Condit:2012nz}.
+Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter \cite{Condit:2012nz}.
 For the purpose of modelling mortality as a function of growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
-We also excluded individuals based on the following rules: 1) those that did not survive at least two censuses (two being required to estimate growth rates); 2) those that were not consistently measured at 1.3 m above ground; 3) those that were multi-stemmed; or 4) those that resprouted or '\textit{returned from the dead}'.
+We also excluded individuals based on the following rules: 1) those that did not survive at least two censuses (two being required to estimate growth rate); 2) those that were not consistently measured at 1.3 m above ground; 3) those that were multi-stemmed; 4) those that resprouted or \textit{``returned from the dead''}; or 5) extreme outliers -- stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial diameter.
 We also discarded species that do not exhibit secondary growth (e.g.
 palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density.
-Extreme outliers: stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial dbh were also removed.
-In order to reduce computational requirements we then subsampled this dataset by randomly selecting one observation per individual out of a possible three censuses.
+In order to maintain statical independence of measurements, we then sub-sampled this dataset by randomly selecting one observation per individual out of a possible three censuses.
 In total, \Sexpr{n_inds} individuals of \Sexpr{n_spp} species were included in our analysis.
 
-Wood density for each species was estimated by coring trees located within 15 km of the BCI plot because increment borers are prohibited within the plot \cite{Wright:2010fl}.
+Wood density for each species was estimated by coring trees located within 15 km of the BCI plot, because increment borers are prohibited within the plot itself \cite{Wright:2010fl}.
 Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100\textdegree C) and dividing by the fresh volume (as measured by water displacement).
@@
 
-Field measurements almost always include observation error, errors that may ultimately bias conclusions.
+Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions.
 In our dataset, \Sexpr{proportion_negative}\% of observed growth rates were negative.
 While there are physiological processes (e.g.
-water depletation and bark drop) that can lead to trunk shrinkages overtime \cite{Pastur:2007kl}, it is biologically impossible for growth (i.e.
+water depletion and bark drop) that can lead to trunk shrinkages overtime \cite{Pastur:2007kl}, it is biologically impossible for growth (i.e.
 the creation of new biomass) to be negative \cite{Clark:2007ve}.
-To ensure our model was not biased by these implausable values we estimated '\textit{true growth}'(i.e.
-observed growth minus observation error) by first using another BCI dataset collected explicitly to examine obervation errors \cite{Condit:2012nz}.
-This dataset consisted of \Sexpr{n_obs_errors} randomly selected trees that had their dbh measured twice within a 30 day period \cite{Condit:2012nz}.
-Assuming no growth occurred over this period, any discrepency in measurements can be attributable to observation error.
-The standard deviation of these discrepencies was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within +/- 1.47 cm of the true diameter.
-Having an estimate of observation error, we then used Bayesian point estimation to maximise the joint posterior of the '\textit{true}' initial and final dbh, and subsequently,'\textit{true growth}' for each individual.
-We did this by first modelling the observed initial ($dbh_{t-1}$) and final dbh ($dbh_{t}$) of each individual, $i$, as random realisations from a normal distribution centered on their respective true estimates (i.e.
-$\widehat{dbh1_i}$ and $\widehat{dbh2_i}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e.
+To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e.
+observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors \cite{Condit:2012nz}.
+This dataset consisted of \Sexpr{n_obs_errors} randomly selected trees that had their diameter measured twice within a 30 day period \cite{Condit:2012nz}.
+Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error.
+The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter.
+Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual.
+We did this by first modelling the observed initial ($D_{i,0}$) and final diameter ($D_{i,1}$) of each individual, $i$, as random realisations from a normal distribution centerd on their respective true estimates (i.e.
+$\widehat{D_{i,0}}$ and $\widehat{D_{i,1}}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e.
 the observed measurement error):
@@
 \begin{aligned}
-dbh_{i,t-1} &\sim \mathcal{N}(\widehat{dbh}_{i,t-1}, 0.75) \\
-dbh_{i,t} &\sim \mathcal{N}(\widehat{dbh}_{i,t}, 0.75)
+D_{i,0} &\sim \mathcal{N}(\widehat{D}_{i,0}, 0.75) \\
+D_{i,1} &\sim \mathcal{N}(\widehat{D}_{i,1}, 0.75)
 \end{aligned} 
@@
 
-The initial dbh for each individual $i$, $dbh_{i,t-1}$, was sampled as a random realisation from a lognormal distribution:
+The initial diameter for each individual $i$, $\widehat{D}_{i,0}$, was sampled as a random realisation from a log-normal distribution:
 \begin{equation} \label{eq:true_dbh1}
-dbh_{i,t-1} \sim \mathcal{LN}(\widehat{dbh}_{i,t-1}, \sigma_{dbh_{t-1}}), 
+\widehat{D}_{i,0} \sim \log \mathcal{N}(\mu_{D_0}, \sigma_{D_0}),
 \end{equation}
-where, $\mu_{dbh_{t-1}}$ is the expected mean of log initial dbh, and $\sigma_{dbh_{t-1}}$ defines that variation around that mean.
-$\mu_{dbh_{t-1}}$ was estimated using a normal prior $\mathcal{N}(0, 2.5)$, which represents our prior expectation that the mean initial dbh would be greater than zero but less than 150 cm.
-$\sigma_{dbh{t-1}}$ was estimated using a weakly informative half cauchy prior centered on zero and a scale parameter of 2.5.
+where $\mu_{D_0}$ is the expected mean of log initial diameter and $\sigma_{D_0}$ defines that variation around that mean.
+$\mu_{D_0}$ was estimated using a normal prior $\mathcal{N}(0, 2.5)$, which represents our prior expectation that the mean initial diameter would be greater than zero but less than 150 cm.
+$\sigma_{D_0}$ was estimated using a weakly informative half Cauchy prior centerd on zero and a scale parameter of 2.5.
 
-The final dbh measurement for each individual $i$, $dbh_{i,t}$ was estimated as:
+The final diameter measurement for each individual $i$, $D_{i,1}$ was estimated as:
 \begin{equation}\label{eq:true_dbh2}
-dbh_{i,t} = \widehat{dbh}_{i,t-1} + R_i \, T_i, 
+\widehat{D}_{i,1} = \widehat{D}_{i,0} + \Delta D_{i,1} \, (t_1-t_0),
 \end{equation}
-where, $R_i$ represents the annual true increase in dbh (cm) and $T_i$ represents the number of days between initial and final measurements divided by 365.25.
-We modelled $R_i$ as a random realisation from a lognormal distribution centered on zero and a standard deviation of 1.
-This prior conveyed our expectation that annual dbh growth could not be negative and is unlikely to be greater than 7 cm/yr.
+where $\Delta D_{i,1}$ represents the annual true increase in diameter (cm) and $(t_1-t_0)$ represents the number of days between initial and final measurements divided by 365.25.
+We modelled $\Delta D_{i,1}$ as a random realisation from a log-normal distribution centerd on zero and a standard deviation of 1.
+This prior conveyed our expectation that annual diameter growth could not be negative and is unlikely to be greater than 7 cm/yr.
 
-Using this model we then used \texttt{rstan} \textit{optimizing} function to maximise the joint posterior and extract point estimates of true initial dbh, $\widehat{dbh}_{i,t-1}$, true final dbh, $\widehat{dbh}_{i,t}$, and true annual dbh growth $R_i$.
-We then used annual diameter growth, $R_i$, and annual basal area growth: $\frac{(\pi \, 0.25 \,\widehat{dbh}_{i,t}) - (\pi \, 0.25 \, \widehat{dbh}_{i,t-1})}{T_i}$ in the subsequent mortality analysis.
+Using this model we then used \texttt{rstan}'s \textit{optimizing} function to maximize the joint posterior and extract point estimates of true initial diameter, $\widehat{D}_{i,0}$, true final diameter, $\widehat{D}_{i,1}$, and true annual diameter growth $\Delta D_{i,1}$.
+We then used annual diameter growth, $\Delta D_{i,1}$, and annual basal area growth: $\frac{(\pi \, 0.25 \,\widehat{D}_{i,1}) - (\pi \, 0.25 \, \widehat{D}_{i,0})}{(t_1-t_0)}$ in the subsequent mortality analysis.
 
@@
 \begin{equation} \label{eq:mortality_model}
-p_{t_1\rightarrow t_2} = 1 - \exp\left(-\lambda\,T\right)
+p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
 \end{equation}
-where, $\lambda$, is the instantaneous hazard rate of death per unit time \cite{Kleinbaum:2012tn} and $T$ is the time lapsed between $t_1$ and $t_2$.
+where, $\lambda$ is the instantaneous hazard rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_1}^{t_2} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_1$ and $t_2$.
 
 Here we examine the performance of three functional forms of $\lambda$ to predict observed patterns of tropical tree mortality.
-The simplest model (eqn:\ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate.
-The second model (eqn: \ref{growth_haz} describes a growth-dependent hazard which assumes plants can only die due to a growth related cause.
+The simplest model (eq.
+\ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate.
+The second model (eq.
+\ref{growth_haz} describes a growth-dependent hazard which assumes plants can only die due to a growth related cause.
 In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached.
 This negative exponential relationship has been used in other studies to model growth-dependent mortality \cite[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \cite{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}.
-The third model (eqn:\ref{add_haz}, and potentially the most biologically realistic, is the summation of the two previous models.
+The third model (eq.
+\ref{add_haz}, and potentially the most biologically realistic, is the summation of the two previous models.
 Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero.
 The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e.
-stochastic events such as windthrow and fire).
-The parameters of eqn:\ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality hazard $\lambda$ at low annual growth rate.
+stochastic events such as wind throw and fire).
+The parameters of eq.
+\ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality hazard $\lambda$ at low annual growth rate.
 $\beta$ represents the exponential decay rate or the sensitivity of hazard rates to change with growth rate; and $\gamma$, is the asymptote and defines the baseline hazard.
@@
 \lambda_i &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
-\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
-\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, Growth_i \right)}^{\text{Growth-dependent hazard}} + 
+\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
+\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} +
 \overbrace{\gamma}^{\text{Baseline hazard}} \label{add_haz}
@@
 
-The central aim of this study is to identify the most parsimonious combination of $\lambda$, growth measure (dbh increment growth or area growth) and wood density that best predicts tropical tree mortality.
+The central aim of this study is to identify the most parsimonious combination of $\lambda$, growth measure (diameter increment growth or area growth) and wood density that best predicts tropical tree mortality.
 To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison.
-Stage 1 involved comparing predictive performance of the three $\lambda$ models described in equations \ref{base_haz} \textendash \ref{add_haz} using either annual dbh increment or area growth rates.
+Stage 1 involved comparing predictive performance of the three $\lambda$ models described in equations \ref{base_haz} \textendash \ref{add_haz} using either annual diameter increment or area growth rates.
 In this stage each model was fitted without allowing parameters to vary by species.
-This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predicitive performance varied by growth measure.
+This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure.
 Following this, we used the best model of stage 1 but allowed its parameters to vary by species.
@@
 This involved using the best model observed in stage 1 and adding wood density effects to each parameter.
-Because we were interested in obtaining the most parimonous model of tropical tree mortality, we fitted the model under all combinations of wood density effects.
-This allowed us to determine the most parsimonous combination of wood density effects and whether it most strongly influenced growth-dependent or growth-independent mortality hazards.
+Because we were interested in obtaining the most parsimonious model of tropical tree mortality, we fitted the model under all combinations of wood density effects.
+This allowed us to determine the most parsimonious combination of wood density effects and whether it most strongly influenced growth-dependent or growth-independent mortality hazards.
 Following this we then fitted the best wood density combination model with species random effects to the entire dataset.
@@
 
-Our observations are of survival status $y_i$ (0 = alive, 1 = died) for each individual $i$ between census $t_1$ and $t_2$.
-We model these observations as random realizations from a Bernoulli distribution:
+Our observations are for survival outcomes $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_1$ and $t_2$.
+We model these observations as random realizations from a Bernoulli distribution
 \begin{equation}
-y_{i,t_1\rightarrow t_2} \sim \textrm{Bernoulli}(p_{i,t_1\rightarrow t_2}).
+S_{i,t_1\rightarrow t_2} \sim \textrm{Bernoulli}(p_{i,t_1\rightarrow t_2}),
+\end{equation}
+where $p_{i,t_1\rightarrow t_2}$ is the probability individual $i$ dies in the interval $t_1$--$t_2$. Using standard survival analysis, this probability is given by
+\begin{equation}
+p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
+\end{equation}
+As we only have a point estimate of $\lambda_i(t)$ at $t=t_1$, the cumulative hazard from  $t_1$--$t_2$ is approximated as
+\begin{equation}
+  \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t \approx \lambda_i(t_1) (t_2-t_1).
 \end{equation}
 
-Here, $p_{i,t_1\rightarrow t_2}$ is the probability individual $i$ dies over an observed time period $T_i$. We estimate this as:
+As we have multiple hypothesized functional forms of $\lambda_i$ (e.g.
+equations \ref{base_haz}\textendash\ref{add_haz}), we describe here the fitting procedure of our most complicated model, of which all others may be considered special cases.
+The instantaneous rate of mortality is
 \begin{equation}
-p_{i,t_1\rightarrow t_2} = 1 - exp(-\lambda_i\,T_i).
-\end{equation}
-
-Because we have multiple hypothesized functional forms of $\lambda_i$ (e.g.
-equations \ref{base_haz}\textendash\ref{add_haz}), we describe the fitting procedure of our most complicated model of which all others are nested within:
-\begin{equation}
-\lambda_i = \left(\overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, Growth_i \right)}^{\text{Growth dependent}} + 
+\lambda_i = \left(\overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, X_i \right)}^{\text{Growth dependent}} +
 \overbrace{\gamma_{s[i]}}^{\text{Growth independent}}\right) \times \overbrace{\delta_{t[i]}}^{\text{Census effect}}
 \end{equation}  
+In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ are allowed to vary by species as a function of both wood density (measured at species level) and a species random effects, e.g.
+\begin{equation}\label{eq:submodels}
+\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{\rho_c}\right)^{\alpha_{1}},
+\end{equation}
+with similar formulations for $\beta_s$ and $\gamma_s$.
+Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$ within which individual $i$ belongs.
+The form of sub-model in eq.
+\ref{eq:submodels} ensures that parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ remain positive.
+On a log scale this model equates to an additive linear model , e.g.
+$\log \alpha_s = \log \alpha_{0,s} + (\alpha_{1} - \log \rho_c) \, \log \rho$.
 
-In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ varying by species as a function of both wood density (measured at species level) and a species random effect:
-\begin{equation}\label{eq:submodels}
-\begin{aligned}
-\alpha_s &= \alpha_{0,s} \, \rho_s^{\alpha_{1}},\\
-\beta_s &= \beta_{0,s} \, \rho_s^{\alpha_{1}}, \\
-\gamma_s &= \gamma_{0,s} \, \rho_s^{\alpha_{1}}. 
-\end{aligned}
-\end{equation}
-
-Here $\alpha_{0,s}$,$\beta_{0,s}$,$\gamma_{0,s}$ represents the species-level residual error not explained by wood density for species $s$ within which individual $i$ belongs.
-$\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively.
-The form of submodels in eqn:\ref{eq:submodels} ensures that parameters remain positive.
-On a log scale these submodels become additive (e.g.
-$\log \alpha_s = \log \alpha_{0,s} + \alpha_{1} \, \log \rho_s$.
-We standardised wood density via the transformation: $\rho_s = \frac{\rho}{0.6}$), an operation which is equivalent to centering a log-transformed $\rho$ at a density of 0.6 g cm${^-3}$.
-This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6 cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant).
-In addition to this, we also centered $Growth_i$ on the lowest 5\% quantile for both dbh increment and area growth (0.172 and 0.338, respectively).
-This weas done to ensure there were sufficient data to accurately estimate $\alpha$ and that it would be interpreted as the hazard rate when growth rate was very low.
+We standardized wood density via the transformation: $\frac{\rho}{\rho_c}$), where $\rho_c=0.6$ g cm${^-3}$.
+This operation is equivalent to centering a log-transformed $\rho$ around $\rho_c$ and enables mean hyper-parameters  to be interpreted as the value for a species with average wood density of 0.6 cm${^-3}$.
+Similarly, we also centered $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.
 
@@
 \begin{equation}
-\begin{aligned}
-\alpha_{0,s} &\sim \mathcal{LN}\left(\mu_{\log \alpha_0}, \sigma_{\log \alpha_0}\right),\\
-\beta_{0,s} &\sim \mathcal{LN}\left(\mu_{\log \beta_0}, \sigma_{\log \beta_0}\right),\\
-\gamma_{0,s} &\sim \mathcal{LN}\left(\mu_{\log \gamma_0}, \sigma_{\log \gamma_0}\right).
-\end{aligned}
+\alpha_{0,s} \sim \log \mathcal{N}\left(\mu_{\log \alpha_0}, \sigma_{\log \alpha_0}\right),
 \end{equation}
+with similar formulations for $\beta_{0,s}$ and $\gamma_{0,s}$.
 
@@
 
-Standard deviations associated with species random effects (i.e. $\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$), were estimated using positive half cauchy priors centered on zero with a scale of 2.5.
+Standard deviations associated with species random effects (i.e. $\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$), were estimated using positive half Cauchy priors centered on zero with a scale of 2.5.
 \begin{equation}
@@
 
-We accounted for possible variation in mortality rates across different census periods by allowing $lambda_i$ to be scaled by $\delta_t$ for each census $t$.
-We estimated $\delta_t$ as a random realization from a lognormal distribution centered on zero and a variance parameter $\sigma_{\delta}$ defines the variation in census effects:
+We accounted for possible variation in mortality rates across different census periods by allowing $\lambda_i$ to be scaled by $\delta_t$ for each census $t$.
+We estimated $\delta_t$ as a random realization from a log-normal distribution centered on zero and a variance parameter $\sigma_{\delta}$ defines the variation in census effects:
 
 \begin{equation} \label{eq:census_effect}
-\delta_t \sim \mathcal{LN}\left(0, \sigma_{\delta}\right)\\
+\delta_t \sim \log \mathcal{N}\left(0, \sigma_{\delta}\right)\\
 \sigma_{\delta} \sim \textrm{Cauchy+}\left(0, 2.5\right).
@@
 For each model we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}.
-We excuted three independent chains and in all cases modelled parameters converged within 2000 iterations.
+We executed three independent chains and in all cases modelled parameters converged within 2000 iterations.
 Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Gelman:1992ts, Brooks:1998ju}.
@@
 \subsection*{Model evaluation}
-
 We examined predictive accuracy of each model by using K-fold cross-validation.
@@
 We then fitted models to nine folds and predicted to the tenth fold.
-This was repeated until each fold had been used nine times for model fitting, and once as a heldout dataset for predicting.
+This was repeated until each fold had been used nine times for model fitting, and once as a held-out dataset for predicting.
 Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy \cite{Stone:1997tg}.
 
-Predictive accuracy was quantified by estimating the average logarithmic loss $logloss$ of across K-folds for heldout data:
+Predictive accuracy was quantified by estimating the average log loss across K-folds for held-out data, $\bar{\mathcal{L}}$ (Fig.
+\ref{fig1}).
+Logarithmic loss, also known as log loss and indicated by $\mathcal{L}$, measures the accuracy of a model by incorporating probabilistic confidence.
+It does this by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome (Fig ??).
+By minimizing the $\bar{\mathcal{L}}$, we maximize predictive skill.
+The average log loss across all individuals for the $k$th fold of held-out data, $\mathcal{L}_k$, is given by
 \begin{equation} \label{eq:logloss}
-logloss = \frac{1}{K}\,\bigg(-\frac{1}{N}\,\sum_{i=1}^{N}\Big(y_i\,\log(p_{i,t_1\rightarrow t_2}) + ( 1 - y_i)\,\log(1 - p_{i,t_1\rightarrow t_2})\Big)\bigg),
+\mathcal{L}_k = - \frac1{N_k}\,\sum_{i=1}^{N_k}\big(S_i\,\log(p_{i,t_1\rightarrow t_2}) + ( 1 - S_i)\,\log(1 - p_{i,t_1\rightarrow t_2})\big),
 \end{equation}
-where $N$ refers to the number of heldout observations and $K$ is the number of K-folds.
+where $N_k$ refers to the number of observations in fold.
 
-Logarithmic loss, also known as Log Loss, measures the accuracy of a model by incorporating probabilistic confidence.
-It does this by penalizing incorrect predictions based on how wrong the predicted probability is from the observation.
-For example, if an individual died between two time steps but the model predicted its probability of death at 0.1, then it would have a Log Loss of \Sexpr{round(- 1*log(0.1) + (1 - 1)*log(1 - 0.1),2)}, whereas if the the model predicted the probability of death was 0.9 then the Log Loss would be \Sexpr{round(- 1*log(0.9) + ( 1 - 1)*log(1 - 0.9),2)}.
-By minimising the Log Loss, we effectively maximise the model accuracy.
-In order to choose between models we averaged the estimated Log Loss across K-folds and reported the mean and standard error.
 }
@@
 
-\showacknow % Display the acknowledgments section
+\showacknow % Display the acknowledgements section
 

### 2016-06-14 ms/ms.Rnw
Standardise census indices throughout
- opting for t_1, t_2, t_3

@@
 Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual.
-We did this by first modelling the observed initial ($D_{i,0}$) and final diameter ($D_{i,1}$) of each individual, $i$, as random realisations from a normal distribution centerd on their respective true estimates (i.e.
-$\widehat{D_{i,0}}$ and $\widehat{D_{i,1}}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e.
+We did this by first modelling the observed initial ($D_{i,1}$) and final diameter ($D_{i,2}$) of each individual, $i$, as random realisations from a normal distribution centerd on their respective true estimates (i.e.
+$\widehat{D_{i,1}}$ and $\widehat{D_{i,2}}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e.
 the observed measurement error):
@@
 \begin{aligned}
-D_{i,0} &\sim \mathcal{N}(\widehat{D}_{i,0}, 0.75) \\
-D_{i,1} &\sim \mathcal{N}(\widehat{D}_{i,1}, 0.75)
+D_{i,1} &\sim \mathcal{N}(\widehat{D}_{i,1}, 0.75) \\
+D_{i,2} &\sim \mathcal{N}(\widehat{D}_{i,2}, 0.75)
 \end{aligned} 
@@
 
-The initial diameter for each individual $i$, $\widehat{D}_{i,0}$, was sampled as a random realisation from a log-normal distribution:
+The initial diameter for each individual $i$, $\widehat{D}_{i,1}$, was sampled as a random realisation from a log-normal distribution:
 \begin{equation} \label{eq:true_dbh1}
-\widehat{D}_{i,0} \sim \log \mathcal{N}(\mu_{D_0}, \sigma_{D_0}),
+\widehat{D}_{i,1} \sim \log \mathcal{N}(\mu_{D_1}, \sigma_{D_1}),
 \end{equation}
-where $\mu_{D_0}$ is the expected mean of log initial diameter and $\sigma_{D_0}$ defines that variation around that mean.
-$\mu_{D_0}$ was estimated using a normal prior $\mathcal{N}(0, 2.5)$, which represents our prior expectation that the mean initial diameter would be greater than zero but less than 150 cm.
-$\sigma_{D_0}$ was estimated using a weakly informative half Cauchy prior centerd on zero and a scale parameter of 2.5.
+where $\mu_{D_1}$ is the expected mean of log initial diameter and $\sigma_{D_1}$ defines that variation around that mean.
+$\mu_{D_1}$ was estimated using a normal prior $\mathcal{N}(0, 2.5)$, which represents our prior expectation that the mean initial diameter would be greater than zero but less than 150 cm.
+$\sigma_{D_1}$ was estimated using a weakly informative half Cauchy prior centerd on zero and a scale parameter of 2.5.
 
-The final diameter measurement for each individual $i$, $D_{i,1}$ was estimated as:
+The final diameter measurement for each individual $i$, $D_{i,2}$ was estimated as:
 \begin{equation}\label{eq:true_dbh2}
-\widehat{D}_{i,1} = \widehat{D}_{i,0} + \Delta D_{i,1} \, (t_1-t_0),
+\widehat{D}_{i,2} = \widehat{D}_{i,1} + \Delta D_{i,2} \, (t_2-t_1),
 \end{equation}
-where $\Delta D_{i,1}$ represents the annual true increase in diameter (cm) and $(t_1-t_0)$ represents the number of days between initial and final measurements divided by 365.25.
-We modelled $\Delta D_{i,1}$ as a random realisation from a log-normal distribution centerd on zero and a standard deviation of 1.
+where $\Delta D_{i,2}$ represents the annual true increase in diameter (cm) and $(t_2-t_1)$ represents the number of days between initial and final measurements divided by 365.25.
+We modelled $\Delta D_{i,2}$ as a random realisation from a log-normal distribution centerd on zero and a standard deviation of 1.
 This prior conveyed our expectation that annual diameter growth could not be negative and is unlikely to be greater than 7 cm/yr.
 
-Using this model we then used \texttt{rstan}'s \textit{optimizing} function to maximize the joint posterior and extract point estimates of true initial diameter, $\widehat{D}_{i,0}$, true final diameter, $\widehat{D}_{i,1}$, and true annual diameter growth $\Delta D_{i,1}$.
-We then used annual diameter growth, $\Delta D_{i,1}$, and annual basal area growth: $\frac{(\pi \, 0.25 \,\widehat{D}_{i,1}) - (\pi \, 0.25 \, \widehat{D}_{i,0})}{(t_1-t_0)}$ in the subsequent mortality analysis.
+Using this model we then used \texttt{rstan}'s \textit{optimizing} function to maximize the joint posterior and extract point estimates of true initial diameter, $\widehat{D}_{i,1}$, true final diameter, $\widehat{D}_{i,2}$, and true annual diameter growth $\Delta D_{i,2}$.
+We then used annual diameter growth, $\Delta D_{i,2}$, and annual basal area growth: $\frac{(\pi \, 0.25 \,\widehat{D}_{i,2}) - (\pi \, 0.25 \, \widehat{D}_{i,1})}{(t_2-t_1)}$ in the subsequent mortality analysis.
 
@@
 
-In survival analysis, the probability that an individual dies between $t_1$ and $t_2$, $p_{t_1\rightarrow t_2}$ is defined by:
+In survival analysis, the probability that an individual dies between $t_2$ and $t_3$, $p_{t_2\rightarrow t_3}$ is defined by:
 \begin{equation} \label{eq:mortality_model}
-p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
+p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
 \end{equation}
-where, $\lambda$ is the instantaneous hazard rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_1}^{t_2} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_1$ and $t_2$.
+where, $\lambda$ is the instantaneous hazard rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_2$ and $t_3$.
 
@@
 
-Our observations are for survival outcomes $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_1$ and $t_2$.
+Our observations are for survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$.
 We model these observations as random realizations from a Bernoulli distribution
 \begin{equation}
-S_{i,t_1\rightarrow t_2} \sim \textrm{Bernoulli}(p_{i,t_1\rightarrow t_2}),
+S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}),
 \end{equation}
-where $p_{i,t_1\rightarrow t_2}$ is the probability individual $i$ dies in the interval $t_1$--$t_2$. Using standard survival analysis, this probability is given by
+where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$. Using standard survival analysis, this probability is given by
 \begin{equation}
-p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
+p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
 \end{equation}
-As we only have a point estimate of $\lambda_i(t)$ at $t=t_1$, the cumulative hazard from  $t_1$--$t_2$ is approximated as
+As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
 \begin{equation}
-  \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t \approx \lambda_i(t_1) (t_2-t_1).
+  \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t \approx \lambda_i(t_2) (t_3-t_2).
 \end{equation}
@@
 \ref{fig1}).
-Logarithmic loss, also known as log loss and indicated by $\mathcal{L}$, measures the accuracy of a model by incorporating probabilistic confidence.
-It does this by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome (Fig ??).
+Logarithmic loss -- commonly known as log loss, and indicated by $\mathcal{L}$-- measures the accuracy of a model by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome (Fig ??).
 By minimizing the $\bar{\mathcal{L}}$, we maximize predictive skill.
@@
 \begin{equation} \label{eq:logloss}
-\mathcal{L}_k = - \frac1{N_k}\,\sum_{i=1}^{N_k}\big(S_i\,\log(p_{i,t_1\rightarrow t_2}) + ( 1 - S_i)\,\log(1 - p_{i,t_1\rightarrow t_2})\big),
+\mathcal{L}_k = - \frac1{N_k}\,\sum_{i=1}^{N_k}\big(S_i\,\log(p_{i,t_2\rightarrow t_3}) + ( 1 - S_i)\,\log(1 - p_{i,t_2\rightarrow t_3})\big),
 \end{equation}

### 2016-06-15 ms/ms.Rnw
Partly reorganise manuscript
Move sections of methods to end of introduction

Will need to re-evaluate the length, but this improves readability for now.

@@
 \begin{abstract}
-Add abstract here
+Individual death is a key process driving population dynamics, nutrient cycling, and evolution within plant communities.
+While past research has identified factors across a variety of scales influencing tree mortality -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these different factors are yet to be integrated within a predictive framework.
+Here we apply classic techniques from survival analysis within a hierarchical Bayesian and cross-validation framework to derive instantaneous mortality functions for tropical rain forest trees at Barro Colorado Island (BCI) Panama, integrating individual-level, species-level, and plot-level effects.
+In reaching this point, we show how mortality models including separate growth-dependent and growth-independent components, using stem diameter growth rather than basal-area growth, and attributing the effects of wood density to growth-independent death outperform alternative formulations.
+Moreover, we show that the effects of wood density -- a prominent trait known to influence mortality -- capture only a portion of total species-level differences; and that the year-to-year variation in mortality was large.
+Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI.
+Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
 \end{abstract}
@@
 If this is the case, add \parshape=0 to the end of the list environment.
+
 \dropcap{P}lant mortality is often modelled as a function of recent growth because it reflects the product of many biotic and abiotic factors on a plants overall carbon budget and vigour \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db, Wyckoff:2002ul}.
@@
 \includegraphics{figures/fig1.pdf}
-\caption{Outline of methodology. A) Past growth used to predict future mortality. B) Three functional forms of plant mortality. C) Cross validation proceedure}
+\caption{Outline of methodology.
+A) Our data consist of repeat measures on stem diameter ($D$) and status ($S$, alive or dead) for individual tress at specific census dates ($t_1$, $t_2$, $t_3$).
+B) We consider three alternative mortality functions, combining growth-independent and growth-dependent effects.
+For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
+C) The predictive skill of alternative models is evaluated via 10-fold cross validation.
+The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$).
+We then refit each alternative model 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing and training data.
+We then average the log loss $\mathcal{L_i}$ for each of these 10 model fits, to give an overall average log loss for each model $\bar{\mathcal{L}}$.}
 \label{fig1}
@@
 
-In this study, we reconcile both intraspecific and interspecific theory by developing a unfied model that incorporates both individual growth and the species level trait, wood density.
+In this study, we reconcile both intraspecific and interspecific theory by developing a unified model that incorporates both individual growth and the species level trait, wood density.
 We extend existing individual-based approaches \cite[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship within a single model (hereafter referred to as growth-dependent hazard).
@@
 4) Does wood density predominately affect baseline hazards, or does it affect growth-dependent hazards or both?
+
+\section*{Framework for modelling mortality}
+
+In classic survival analysis, the probability that an individual dies between $t_2$ and $t_3$, $p_{t_2\rightarrow t_3}$ is defined by
+\begin{equation} \label{eq:mortality_model}
+p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right),
+\end{equation}
+where $\lambda_i(t)$ is the instantaneous hazard (mortality) rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_2$ and $t_3$.
+The challenge then is to identify suitable forms of the instantaneous hazard (mortality) rate function.
+
+We examine the performance of three functional forms of $\lambda_i(t)$ to predict observed patterns of tropical tree mortality:
+\begin{align}
+\lambda_i &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
+\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
+\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} +
+\overbrace{\gamma.}^{\text{Baseline hazard}} \label{add_haz}
+\end{align}
+The simplest model (eq.
+\ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate.
+The second model (eq.
+\ref{growth_haz}) describes a growth-dependent hazard which assumes plants can only die due to a growth related cause.
+In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached.
+This negative exponential relationship has been used in other studies to model growth-dependent mortality \cite[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \cite{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}.
+The third model (eq.
+\ref{add_haz}) -- potentially the most biologically realistic -- is the summation of the two previous models.
+Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero.
+The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e.
+stochastic events such as wind throw and fire).
+The parameters of eq.
+\ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality hazard $\lambda$ at low annual growth rate.
+$\beta$ represents the exponential decay rate or the sensitivity of hazard rates to change with growth rate; and $\gamma$, is the asymptote and defines the baseline hazard.
+$\alpha$ and $\beta$ are intrinsically linked to a plants growth rate, and as such are growth-dependent parameters.
+By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g.
+windfall, fire) that can kill a plant irrespective of its growth rate.
+
+The central aim of this study is to identify the most parsimonious combination of $\lambda$, growth measure (diameter increment growth or area growth) and wood density that best predicts tropical tree mortality.
+To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison.
+Stage 1 involved comparing predictive performance of the three $\lambda$ models described in equations \ref{base_haz} \textendash \ref{add_haz} using either annual diameter increment or area growth rates.
+In this stage each model was fitted without allowing parameters to vary by species.
+This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure.
+Following this, we used the best model of stage 1 but allowed its parameters to vary by species.
+This allowed us to determine how variable each parameter was across species and whether accounting for species differences significantly improved predictive performance.
+Because species responses varied considerably, we undertook a third stage of model fitting and selection that explicitly examined whether wood density (a species level trait) explained some of this interspecific variability.
+This involved using the best model observed in stage 1 and adding wood density effects to each parameter.
+Finally we then fitted the best wood density combination model with species random effects to the entire dataset.
+This model was then used to examine mortality curves across wood density gradients.
+In order to determine the proportion of interspecific variation explained by wood density on each parameter it occurred on we reran the final model form with just random effects on parameters and calculated the percent reduction in unexplained species variation.
+
+At each stage, we evaluated the predictive accuracy of alternative models or predictor variables using K-fold cross-validation (Fig.
+\ref{fig1}).
+This involves randomly partitioning the data into ten folds of \Sexpr{n_heldout} observations, while ensuring that each K-fold had at least one observation per species.
+We then fitted models to nine folds and predicted to the tenth fold.
+This was repeated until each fold had been used nine times for model fitting, and once as a held-out dataset for predicting.
+Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy \cite{Hooten:2015, Stone:1997tg}.
 
@@
 
-\subsection*{Mortality analysis}
-
-In survival analysis, the probability that an individual dies between $t_2$ and $t_3$, $p_{t_2\rightarrow t_3}$ is defined by:
-\begin{equation} \label{eq:mortality_model}
-p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
-\end{equation}
-where, $\lambda$ is the instantaneous hazard rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_2$ and $t_3$.
-
-Here we examine the performance of three functional forms of $\lambda$ to predict observed patterns of tropical tree mortality.
-The simplest model (eq.
-\ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate.
-The second model (eq.
-\ref{growth_haz} describes a growth-dependent hazard which assumes plants can only die due to a growth related cause.
-In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached.
-This negative exponential relationship has been used in other studies to model growth-dependent mortality \cite[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \cite{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}.
-The third model (eq.
-\ref{add_haz}, and potentially the most biologically realistic, is the summation of the two previous models.
-Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero.
-The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e.
-stochastic events such as wind throw and fire).
-The parameters of eq.
-\ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality hazard $\lambda$ at low annual growth rate.
-$\beta$ represents the exponential decay rate or the sensitivity of hazard rates to change with growth rate; and $\gamma$, is the asymptote and defines the baseline hazard.
-$\alpha$ and $\beta$ are intrinsically linked to a plants growth rate, and as such are growth-dependent parameters.
-By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g.
-windfall, fire) that can kill a plant irrespective of its growth rate.
-\begin{align}
-\lambda_i &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
-\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
-\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} +
-\overbrace{\gamma}^{\text{Baseline hazard}} \label{add_haz}
-\end{align}
-
-The central aim of this study is to identify the most parsimonious combination of $\lambda$, growth measure (diameter increment growth or area growth) and wood density that best predicts tropical tree mortality.
-To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison.
-Stage 1 involved comparing predictive performance of the three $\lambda$ models described in equations \ref{base_haz} \textendash \ref{add_haz} using either annual diameter increment or area growth rates.
-In this stage each model was fitted without allowing parameters to vary by species.
-This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure.
-Following this, we used the best model of stage 1 but allowed its parameters to vary by species.
-This allowed us to determine how variable each parameter was across species and whether accounting for species differences significantly improved predictive performance.
-Because species responses varied considerably, we undertook a third stage of model fitting and selection that explicitly examined whether wood density (a species level trait) explained some of this interspecific variability.
-This involved using the best model observed in stage 1 and adding wood density effects to each parameter.
-Because we were interested in obtaining the most parsimonious model of tropical tree mortality, we fitted the model under all combinations of wood density effects.
-This allowed us to determine the most parsimonious combination of wood density effects and whether it most strongly influenced growth-dependent or growth-independent mortality hazards.
-Following this we then fitted the best wood density combination model with species random effects to the entire dataset.
-This model was then used to examine mortality curves across wood density gradients.
-In order to determine the proportion of interspecific variation explained by wood density on each parameter it occurred on we reran the final model form with just random effects on parameters and calculated the percent reduction in unexplained species variation.
-
 \subsection*{Model fitting}
-
 Our observations are for survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$.
@@
 \end{equation}
-where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$. Using standard survival analysis, this probability is given by
-\begin{equation}
-p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
-\end{equation}
+where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$ *(eq.
+\ref{eq:mortality_model}).
 As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
@@
 
-We standardized wood density via the transformation: $\frac{\rho}{\rho_c}$), where $\rho_c=0.6$ g cm${^-3}$.
+We standardized wood density via the transformation: $\frac{\rho}{\rho_c}$, where $\rho_c=0.6$ g cm${^-3}$.
 This operation is equivalent to centering a log-transformed $\rho$ around $\rho_c$ and enables mean hyper-parameters  to be interpreted as the value for a species with average wood density of 0.6 cm${^-3}$.
@@
 \subsection*{Model evaluation}
-We examined predictive accuracy of each model by using K-fold cross-validation.
-This involved randomly partitioning the data into ten folds of \Sexpr{n_heldout} observations, while ensuring that each K-fold had at least one observation per species.
-We then fitted models to nine folds and predicted to the tenth fold.
-This was repeated until each fold had been used nine times for model fitting, and once as a held-out dataset for predicting.
-Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy \cite{Stone:1997tg}.
-
 Predictive accuracy was quantified by estimating the average log loss across K-folds for held-out data, $\bar{\mathcal{L}}$ (Fig.

### 2016-06-15 R/figures.R

@@ -200,7 +200,7 @@ plot_fig1 <- function(tree1, tree2, panelc) {
   hazard_plot(growth_haz, expression(alpha*"e"^{-beta~"X"}))
   mtext("Growth rate", 1, line =1, cex=0.75)
  # Example full baseline + growth-dependent hazard
-  hazard_plot(base_growth_haz, expression(gamma + alpha*"e"^{-beta~"X"}))
+  hazard_plot(base_growth_haz, expression(alpha*"e"^{-beta~"X"}+gamma))
 
   plot(NA, xlim = c(0, 1), ylim= c(0, 1), ann = FALSE, axes=FALSE)
   my_label("C) 10-fold cross validation")
@@ -213,8 +213,8 @@ plot_fig1 <- function(tree1, tree2, panelc) {
               just=c("bottom"), height=unit(1, "npc"))
 
   popViewport(3)
-  text(1, 0.6,  expression(paste(bar("L"), "=")), cex=1.25, xpd=NA)
-  text(1, 0.4,  expression(sum("L"[i])/10), cex=1.25, xpd=NA)
+  text(1, 0.65,  expression(paste(bar("L"), "=")), cex=1.25, xpd=NA)
+  text(1, 0.35,  expression(sum("L"[i])/10), cex=1.25, xpd=NA)
 
 }
 

### 2016-06-17 R/figures.R
Added wood density units

@@ -67,7 +67,7 @@ plot_spp_curves <- function(model, data, growth_range= c(0.03,0.5), hazard_curve
       scale_y_continuous(expand=c(0,0), limits = c(0, NA)) +
       ylab("Annual mortality probability") +
       xlab("Annual dbh growth (cm)") +
-      scale_colour_gradient("wood density",low="blue", high="red") +
+      scale_colour_gradient(expression("wood density"~("g/cm"^3)),low="blue", high="red") +
       partial_plot_theme(legend.position="right")
   }
   else {
@@ -77,7 +77,7 @@ plot_spp_curves <- function(model, data, growth_range= c(0.03,0.5), hazard_curve
       scale_y_continuous(expand=c(0,0), limits = c(0, NA)) +
       ylab("Instantaneous mortality rate") +
       xlab("Annual dbh growth (cm)") +
-      scale_colour_gradient("wood density",low="blue", high="red") +
+      scale_colour_gradient(expression("wood density"~(g/cm^3)),low="blue", high="red") +
       partial_plot_theme(legend.position="right")
   }
 }
@@ -279,7 +279,7 @@ ggplot(spp, aes(x = wood_density,y = mean)) +
       partial_plot_theme() +
       scale_y_log10(breaks= breaks, labels = labels) +
      ylab("Instantaneous mortality rate") +
-     xlab("Wood density")
+     xlab(expression("wood density"~("g/cm"^3)))
 }
 
 plot_fig4 <- function(model, data) {
@@ -287,10 +287,10 @@ plot_fig4 <- function(model, data) {
     ggtitle('High/Low wood density') + 
     theme(title = element_text(size=6))
   p2 <- plot_spp_curves(model, data, hazard_curve = TRUE) +
-    partial_plot_theme(legend.position= c(0.8,0.65)) + 
+    partial_plot_theme(legend.position= c(0.8,0.5)) + 
     ggtitle('Species mortality curves') + 
     theme(title = element_text(size=6)) + 
-    theme(legend.key.size =unit(0.4, "cm"))
+    theme(legend.key.size =unit(0.4, "cm"), legend.title=element_text(size=5))
   p3 <- plot_mu_curves(model, hazard_curve= FALSE)
   p4 <- plot_spp_curves(model, data, hazard_curve = FALSE) + partial_plot_theme()
   plot_grid(p1,p2,p3,p4, ncol=2, labels=LETTERS[1:4], label_size = 7)

### 2016-06-17 R/stan_functions.R
Set seed

@@ -155,6 +155,7 @@ run_single_stan_chain <- function(model, data, chain_id, iter=4000,
        pars = model$pars,
        iter = iter,
        chains=1,
+       seed = 12345,
        chain_id=chain_id,
        control =list(stepsize=0.1, adapt_delta=0.99, max_treedepth=15))
 }

### 2016-06-17 ms/ms.Rnw
better figure captions

@@
 \includegraphics{figures/fig2.pdf}
-\caption{Average logarithmic loss from 10-fold cross validation.
-A) Comparing various functional forms of tropical tree mortality with different growth measures different (white = no growth, grey = basal area growth, black = DBH growth) and species effects.
-B) Comparing model forms with various combinations of wood density effects.}
+\caption{Average logarithmic loss (lower = higher accuracy).
+A) Four sequential stages of model selection.
+Null model: shows the expected logloss for the null model with no census effect; because mortality varied substantially between censuses it was included on all subsequent models; Functional form: comparison between baseline hazard model ($\gamma c$; circle), growth\textendash dependent hazard model ($(\alpha e^{\beta X_{i}}) c$; triangle) and model incorporating both baseline and growth\textendash dependent hazards ($(\alpha e^{\beta X_{i}} + \gamma) c$; square); Wood density: Most parsimonious effect of wood density; Species effects: Model with species random effects on all parameters \textendash indicating the total gain in predictive capacity if accounting for all species variability.
+B) Wood density effects on all parameter combinations in the model that incorporates both baseline and growth-dependent hazards.
+Colors represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth.
+Errors are $\pm$ 95\% BCI.}
 \label{fig2}
@@
 \includegraphics{figures/fig3.pdf}
-\caption{Estimated log baseline mortality against wood density. Points are individual species and solid line is the mean trend line. All errors are $\pm$ 95\% BCI.}
+\caption{Estimated relationship between wood density and baseline mortality rate. Points are species estimates and solid line is the mean species trendline. All errors are $\pm$ 95\% BCI.}
 \label{fig3}
@@
 \includegraphics{figures/fig4.pdf}[ht]
-\caption{Estimated curves of instantaneous mortality rate and annual mortality probability.
-A,C) Highlight expected response curves for a hypothetical median species at low (0.3) and high (0.8) wood density.
-B,D) Predicted response curves for all species.
+\caption{Predicted instantaneous mortality rates its derivative annual mortality probability.
+A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship.
+B,D) Predicted growth\textendash mortality response curves for each species.
 Errors are $\pm$ 95\% BCI.}

### 2016-07-04 R/figures.R
Shrink barplot width

@@ -103,7 +103,7 @@ plot_param_prop_explained <- function(param_prop_explained, ylab=NULL, xlab=NULL
            level = ifelse(param %in% c("wood_density","species"), "species","census"),
            level = factor(level, levels=c("species","census")))
   ggplot(dat, aes(x=level, y=proportion)) +
-    geom_bar(stat='identity', aes(fill=param)) +
+    geom_bar(stat='identity', aes(fill=param), width=0.5) +
     scale_x_discrete(labels=c('species','census')) +
     scale_y_continuous(expand=c(0,0), limits=c(0,1)) +
     scale_fill_manual(values = c("wood_density" ="darkgrey","species" ="black", "census"="black")) + 
@@ -116,8 +116,8 @@ plot_param_prop_explained <- function(param_prop_explained, ylab=NULL, xlab=NULL
 plot_base_v_growth_prop_explained <- function(param_prop_explained, ylab=NULL, xlab=NULL) {
   dat <- filter(param_prop_explained, param %in% c("baseline", "growth_dependent")) %>% droplevels()
   ggplot(dat, aes(x=param, y=proportion)) +
-    geom_bar(stat='identity', fill='black') +
-    scale_x_discrete(labels=c('growth dependent hazard','baseline hazard')) +
+    geom_bar(stat='identity', fill='black', width=0.5) +
+    scale_x_discrete(labels=c('growth dependent \n hazard','baseline \n hazard')) +
     scale_y_continuous(expand=c(0,0), limits=c(0,1)) +
     ylab(ylab) +
     xlab(xlab) +
@@ -368,9 +368,7 @@ plot_fig3 <- function(spp_params_covs, pred_mu_basehaz) {
 }
 
 plot_fig4 <- function(model, data) {
-  p1 <- plot_spp_curves(model, data, hazard_curve = FALSE) +
-    theme(title = element_text(size=6))
-  
+  p1 <- plot_spp_curves(model, data, hazard_curve = FALSE)
   p2 <- plot_mu_curves(model, hazard_curve= FALSE) +
     theme(legend.position= c(0.8,0.8),
           legend.key.size =unit(0.3, "cm"), 
@@ -386,8 +384,8 @@ plot_fig4 <- function(model, data) {
 # Proportion of variance explained
 plot_fig5 <- function(param_prop_explained) {
   p1 <- plot_param_prop_explained(param_prop_explained, ylab="Proportion of variation explained")
[…]

### 2016-07-04 ms/ms.Rnw
Move methods to supplementary

@@
 \author[a,1]{James S Camac}
-\author[a,2]{Mark Westoby} 
-\author[a,3]{Daniel S Falster}
+\author[a]{Mark Westoby} 
+\author[b]{S Joseph Wright}
+\author[a]{Daniel S Falster}
 
@@
 \affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
+\affil[b]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
 
@@
 designed and performed the analysis; D.F.
-conceived the idea, supervized the development of this work; M.W.
+conceived the idea, supervized the development of this work; S.J.W \& M.W.
 provided biological insight.
@@
 \begin{abstract}
-Individual death is a key process driving population dynamics, nutrient cycling, and evolution within plant communities.
-While past research has identified factors across a variety of scales influencing tree mortality -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these different factors are yet to be integrated within a predictive framework.
-Here we apply classic techniques from survival analysis within a hierarchical Bayesian and cross-validation framework to derive instantaneous mortality functions for tropical rain forest trees at Barro Colorado Island (BCI) Panama, integrating individual-level, species-level, and plot-level effects.
-In reaching this point, we show how mortality models including separate growth-dependent and growth-independent components, using stem diameter growth rather than basal-area growth, and attributing the effects of wood density to growth-independent death outperform alternative formulations.
+Death is an integral process driving population dynamics, nutrient cycling, and evolution within plant communities.
+While past research has identified factors influencing tree mortality -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these are yet to be integrated within a general predictive framework.
+Here we estimate instantenous mortality rates of 203 tropical rainforest tree species at Barro Colorado Island (BCI), Panama, by combining classic survival analysis techniques with a cross-validated hierarchical Bayesian framework that integrates individual-level, species-level, and plot-level effects.
+We show that greatest predictive capacity is achieved when when models simultaneously estimate growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth and account for wood density effects on growth-independent terms.
 Moreover, we show that the effects of wood density -- a prominent trait known to influence mortality -- capture only a portion of total species-level differences; and that the year-to-year variation in mortality was large.
-Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI.
+Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the largest driver of mortality at BCI.
 Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
@@
 This is despite growing empirical evidence highlighting that mortality rates vary considerably between species, even within the same community \cite{Kobe:1997vy, Wyckoff:2002ul, Kunstler:2005bn, Vieilledent:2010fv}.
-Furthermore, because no systematic analysis has been conducted to determine what growth measure best predicts plant mortality, the growth-mortality relationship is often parameterized using a variety growth measures; the two most common being rates of DBH or basal area growth.
+Furthermore, because no systematic analysis has been conducted to determine what growth measure best predicts plant mortality, the growth-mortality relationship is often parameterized using a variety growth measures; two of the most common being rates of DBH or basal area growth.
 
@@
 
-In this study, we reconcile both intraspecific and interspecific theory by developing a unified model that incorporates both individual growth and the species level trait, wood density.
-We extend existing individual-based approaches \cite[e.g.][]{Kobe:1995tw} by simultaneously estimating both baseline hazards and the growth-mortality relationship within a single model (hereafter referred to as growth-dependent hazard).
-We then compare the predictive performance of this model to other traditionally used mortality models.
-We parameterize these models using 15-years of individual growth and mortality data from 203 tropical rainforest species encompassing \Sexpr{n_inds} individuals.
-Using these models, coupled with 10-fold cross validation, we ask several questions: 1) Does simultaneously estimating baseline and growth-dependent hazards improve predictive capacity relative to baseline only or growth-dependent only models?
-2) is DBH growth or basal area growth more predictive of tropical tree mortality and does it's predictive capacity vary with model type?
-3) Does allowing both hazards to vary by species improve predictive capacity?
-4) Does wood density predominately affect baseline hazards, or does it affect growth-dependent hazards or both?
+In this study, we reconcile both intraspecific and interspecific mortality theory by combining classic survival analysis techniques with a cross-validated hierarchical Bayesian framework.
+Specifically, our framework  instantaneous mortality rates for 203 tropical rainforest trees at Barro Colorado Island (BCI) by integrating individual-level, species-level, and plot-level effects and simultaneously estimating growth-dependent and growth-independent mortality processes.
+Using this framework we
+identify the most parsimonious model of tropical tree instantaneous mortality rates.
+We also ask: 1) which  growth measure (i.e.
+diameter increment growth or area growth)best predicts growth-dependent mortality; 2) Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
+3) What is the proportion of species variability explained by wood density?
 
@@
 \end{equation}
-where $\lambda_i(t)$ is the instantaneous hazard (mortality) rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_2$ and $t_3$.
+where $\lambda_i(t)$ is the instantaneous mortality rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_2$ and $t_3$.
 The challenge then is to identify suitable forms of the instantaneous hazard (mortality) rate function.
@@
 The simplest model (eq.
-\ref{base_haz}), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate.
+\ref{base_haz}), assumes a single baseline mortality rate, which is constant irrespective of a plant's growth rate.
 The second model (eq.
-\ref{growth_haz}) describes a growth-dependent hazard which assumes plants can only die due to a growth related cause.
+\ref{growth_haz}) describes a growth-dependent mortality rate which assumes plants can only die due to a growth related cause.
 In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached.
@@
 \ref{add_haz}) -- potentially the most biologically realistic -- is the summation of the two previous models.
-Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero.
+Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike eq.
+\ref{growth_haz}, will asymptote at some baseline rate instead of zero.
 The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e.
@@
 The parameters of eq.
-\ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality hazard $\lambda$ at low annual growth rate.
-$\beta$ represents the exponential decay rate or the sensitivity of hazard rates to change with growth rate; and $\gamma$, is the asymptote and defines the baseline hazard.
+\ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality rate $\lambda$ at low annual growth rate.
+$\beta$ represents the exponential decay rate or the sensitivity of mortality rates to change with growth rate; and $\gamma$, is the asymptote that defines the baseline hazard.
 $\alpha$ and $\beta$ are intrinsically linked to a plants growth rate, and as such are growth-dependent parameters.
@@
 
+\begin{figure}[ht]
+\includegraphics{figures/fig5.pdf}[ht]
+\caption{Predicted instantaneous mortality rates its derivative annual mortality probability.
+A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship.
+B,D) Predicted growth\textendash mortality response curves for each species.
+Errors are $\pm$ 95\% BCI.}
+\label{fig5}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics{figures/fig6.pdf}[ht]
+\caption{Predicted instantaneous mortality rates its derivative annual mortality probability.
+A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship.
+B,D) Predicted growth\textendash mortality response curves for each species.
+Errors are $\pm$ 95\% BCI.}
+\label{fig6}
+\end{figure}
+
 \section*{Discussion}
@@
 The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter.
-Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual.
-We did this by first modelling the observed initial ($D_{i,1}$) and final diameter ($D_{i,2}$) of each individual, $i$, as random realisations from a normal distribution centerd on their respective true estimates (i.e.
-$\widehat{D_{i,1}}$ and $\widehat{D_{i,2}}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e.
-the observed measurement error):
+Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual (SUPP).
 
-\begin{equation}\label{eq:true_growth_obs} 
-\begin{aligned}
-D_{i,1} &\sim \mathcal{N}(\widehat{D}_{i,1}, 0.75) \\
-D_{i,2} &\sim \mathcal{N}(\widehat{D}_{i,2}, 0.75)
-\end{aligned} 
-\end{equation}
-
-The initial diameter for each individual $i$, $\widehat{D}_{i,1}$, was sampled as a random realisation from a log-normal distribution:
-\begin{equation} \label{eq:true_dbh1}
-\widehat{D}_{i,1} \sim \log \mathcal{N}(\mu_{D_1}, \sigma_{D_1}),
-\end{equation}
-where $\mu_{D_1}$ is the expected mean of log initial diameter and $\sigma_{D_1}$ defines that variation around that mean.
-$\mu_{D_1}$ was estimated using a normal prior $\mathcal{N}(0, 2.5)$, which represents our prior expectation that the mean initial diameter would be greater than zero but less than 150 cm.
-$\sigma_{D_1}$ was estimated using a weakly informative half Cauchy prior centerd on zero and a scale parameter of 2.5.
-
-The final diameter measurement for each individual $i$, $D_{i,2}$ was estimated as:
-\begin{equation}\label{eq:true_dbh2}
-\widehat{D}_{i,2} = \widehat{D}_{i,1} + \Delta D_{i,2} \, (t_2-t_1),
-\end{equation}
-where $\Delta D_{i,2}$ represents the annual true increase in diameter (cm) and $(t_2-t_1)$ represents the number of days between initial and final measurements divided by 365.25.
-We modelled $\Delta D_{i,2}$ as a random realisation from a log-normal distribution centerd on zero and a standard deviation of 1.
-This prior conveyed our expectation that annual diameter growth could not be negative and is unlikely to be greater than 7 cm/yr.
-
-Using this model we then used \texttt{rstan}'s \textit{optimizing} function to maximize the joint posterior and extract point estimates of true initial diameter, $\widehat{D}_{i,1}$, true final diameter, $\widehat{D}_{i,2}$, and true annual diameter growth $\Delta D_{i,2}$.
-We then used annual diameter growth, $\Delta D_{i,2}$, and annual basal area growth: $\frac{(\pi \, 0.25 \,\widehat{D}_{i,2}) - (\pi \, 0.25 \, \widehat{D}_{i,1})}{(t_2-t_1)}$ in the subsequent mortality analysis.
+\subsection*{Estimating mortality rates}
 
 \subsection*{Model fitting}
-Our observations are for survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$.
-We model these observations as random realizations from a Bernoulli distribution
-\begin{equation}
-S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}),
-\end{equation}
-where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$ *(eq.
-\ref{eq:mortality_model}).
-As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
-\begin{equation}
-  \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t \approx \lambda_i(t_2) (t_3-t_2).
-\end{equation}
 
-As we have multiple hypothesized functional forms of $\lambda_i$ (e.g.
-equations \ref{base_haz}\textendash\ref{add_haz}), we describe here the fitting procedure of our most complicated model, of which all others may be considered special cases.
-The instantaneous rate of mortality is
-\begin{equation}
-\lambda_i = \left(\overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, X_i \right)}^{\text{Growth dependent}} +
-\overbrace{\gamma_{s[i]}}^{\text{Growth independent}}\right) \times \overbrace{\delta_{t[i]}}^{\text{Census effect}}
-\end{equation}  
-In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ are allowed to vary by species as a function of both wood density (measured at species level) and a species random effects, e.g.
-\begin{equation}\label{eq:submodels}
-\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{\rho_c}\right)^{\alpha_{1}},
-\end{equation}
-with similar formulations for $\beta_s$ and $\gamma_s$.
-Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$ within which individual $i$ belongs.
-The form of sub-model in eq.
-\ref{eq:submodels} ensures that parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ remain positive.
-On a log scale this model equates to an additive linear model , e.g.
-$\log \alpha_s = \log \alpha_{0,s} + (\alpha_{1} - \log \rho_c) \, \log \rho$.
-
-We standardized wood density via the transformation: $\frac{\rho}{\rho_c}$, where $\rho_c=0.6$ g cm${^-3}$.
-This operation is equivalent to centering a log-transformed $\rho$ around $\rho_c$ and enables mean hyper-parameters  to be interpreted as the value for a species with average wood density of 0.6 cm${^-3}$.
-Similarly, we also centered $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.
-
-We modelled species random effects as random realisations from log-normal distributions:
-\begin{equation}
-\alpha_{0,s} \sim \log \mathcal{N}\left(\mu_{\log \alpha_0}, \sigma_{\log \alpha_0}\right),
-\end{equation}
-with similar formulations for $\beta_{0,s}$ and $\gamma_{0,s}$.
-
-All three mean hyperparameters, $\mu_{\log \alpha_0}$, the average log species residual error at zero growth, $\mu_{\log \beta_0}$, the average species error on the exponential decay rate, and $\mu_{\log \gamma_0}$, the average baseline log species residual error, were estimated using normal distributions centered on zero with a standard deviation of 2.5.
-\begin{equation}
-\mu_{\log \alpha_{0}},\mu_{\log \beta_0},\mu_{\log \gamma_0} \sim \mathcal{N}\left(0, 2.5\right).\\
-\end{equation}
-
-Standard deviations associated with species random effects (i.e. $\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$), were estimated using positive half Cauchy priors centered on zero with a scale of 2.5.
-\begin{equation}
-\sigma_{a0},\sigma_{b0},\sigma_{c0} \sim \textrm{Cauchy+}\left(0, 2.5\right).\\
-\end{equation}
-
-For models that did not include species random effects, single intercepts were estimated for each parameter using log normal distributions centered on zero with standard deviations of 2.5.
-
-We estimated wood density effects (i.e. $\alpha_1$, $\beta_1$ and $\gamma_1$) using normal priors centered on zero and a standard deviation of 2.5:
-\begin{equation} \label{eq:rho_effects}
-\alpha_1, \beta_1, \gamma_1 \sim \mathcal{N}\left(0, 2.5\right).\\
-\end{equation}
-
-We accounted for possible variation in mortality rates across different census periods by allowing $\lambda_i$ to be scaled by $\delta_t$ for each census $t$.
-We estimated $\delta_t$ as a random realization from a log-normal distribution centered on zero and a variance parameter $\sigma_{\delta}$ defines the variation in census effects:
-
-\begin{equation} \label{eq:census_effect}
-\delta_t \sim \log \mathcal{N}\left(0, \sigma_{\delta}\right)\\
-\sigma_{\delta} \sim \textrm{Cauchy+}\left(0, 2.5\right).
-\end{equation}
-
-For each model we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}.
+For all models we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}.
 We executed three independent chains and in all cases modelled parameters converged within 2000 iterations.

### 2016-07-04 ms/supp.Rnw
supp material

\documentclass[9pt,twocolumn,twoside]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} = Template for a one-column mathematics article
% {pnasinvited} = Template for a PNAS invited submission

\title{Unifying intraspecific and interspecific variation in tropical tree mortality}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{James S Camac}
\author[a]{Mark Westoby} 
\author[b]{S Joseph Wright}
\author[a]{Daniel S Falster}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
\affil[b]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}

% Please give the surname of the lead author for the running footer
\leadauthor{Camac}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Add something significant here}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{J.C. designed and performed the analysis; D.F. conceived the idea, supervized the development of this work; S.J.W \& M.W. provided biological insight. All authors contributed to the writing of this manuscript.}

\authordeclaration{The authors declare that they have no competing financial interests.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: james.camac@gmail.com}

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{Tropical trees $|$ wood density $|$ growth $|$ mortality $|$ interspecific}

\begin{document}
% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalized the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

\subsection*{Modelling true growth}

We did this by first modelling the observed initial ($D_{i,1}$) and final diameter ($D_{i,2}$) of each individual, $i$, as random realisations from a normal distribution centerd on their respective true estimates (i.e. $\widehat{D_{i,1}}$ and $\widehat{D_{i,2}}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e. the observed measurement error):

\begin{equation}\label{eq:true_growth_obs} 
\begin{aligned}
D_{i,1} &\sim \mathcal{N}(\widehat{D}_{i,1}, 0.75) \\
D_{i,2} &\sim \mathcal{N}(\widehat{D}_{i,2}, 0.75)
\end{aligned} 
\end{equation}

The initial diameter for each individual $i$, $\widehat{D}_{i,1}$, was sampled as a random realisation from a log-normal distribution:
\begin{equation} \label{eq:true_dbh1}
\widehat{D}_{i,1} \sim \log \mathcal{N}(\mu_{D_1}, \sigma_{D_1}),
\end{equation}
where $\mu_{D_1}$ is the expected mean of log initial diameter and $\sigma_{D_1}$ defines that variation around that mean. $\mu_{D_1}$ was estimated using a normal prior $\mathcal{N}(0, 2.5)$, which represents our prior expectation that the mean initial diameter would be greater than zero but less than 150 cm. $\sigma_{D_1}$ was estimated using a weakly informative half Cauchy prior centerd on zero and a scale parameter of 2.5.

The final diameter measurement for each individual $i$, $D_{i,2}$ was estimated as:
\begin{equation}\label{eq:true_dbh2}
\widehat{D}_{i,2} = \widehat{D}_{i,1} + \Delta D_{i,2} \, (t_2-t_1),
\end{equation}
where $\Delta D_{i,2}$ represents the annual true increase in diameter (cm) and $(t_2-t_1)$ represents the number of days between initial and final measurements divided by 365.25. We modelled $\Delta D_{i,2}$ as a random realisation from a log-normal distribution centerd on zero and a standard deviation of 1. This prior conveyed our expectation that annual diameter growth could not be negative and is unlikely to be greater than 7 cm/yr.

Using this model we then used \texttt{rstan}'s \textit{optimizing} function to maximize the joint posterior and extract point estimates of true initial diameter, $\widehat{D}_{i,1}$, true final diameter, $\widehat{D}_{i,2}$, and true annual diameter growth $\Delta D_{i,2}$. We then used annual diameter growth, $\Delta D_{i,2}$, and annual basal area growth: $\frac{(\pi \, 0.25 \,\widehat{D}_{i,2}) - (\pi \, 0.25 \, \widehat{D}_{i,1})}{(t_2-t_1)}$ in the subsequent mortality analysis.

\subsection*{Mortality model}
\subsection*{Model fitting}
Our observations are for survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$. We model these observations as random realizations from a Bernoulli distribution
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}),
\end{equation}
where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$ *(eq. \ref{eq:mortality_model}). As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
\begin{equation}
  \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t \approx \lambda_i(t_2) (t_3-t_2).
\end{equation}

As we have multiple hypothesized functional forms of $\lambda_i$ (e.g. equations \ref{base_haz}\textendash\ref{add_haz}), we describe here the fitting procedure of our most complicated model, of which all others may be considered special cases. The instantaneous rate of mortality is
\begin{equation}
\lambda_i = \left(\overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, X_i \right)}^{\text{Growth dependent}} +
\overbrace{\gamma_{s[i]}}^{\text{Growth independent}}\right) \times \overbrace{\delta_{t[i]}}^{\text{Census effect}}
\end{equation}  
In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ are allowed to vary by species as a function of both wood density (measured at species level) and a species random effects, e.g.
\begin{equation}\label{eq:submodels}
\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{\rho_c}\right)^{\alpha_{1}},
\end{equation}
with similar formulations for $\beta_s$ and $\gamma_s$. Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$ within which individual $i$ belongs. The form of sub-model in eq. \ref{eq:submodels} ensures that parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ remain positive. On a log scale this model equates to an additive linear model , e.g. $\log \alpha_s = \log \alpha_{0,s} + (\alpha_{1} - \log \rho_c) \, \log \rho$.

We standardized wood density via the transformation: $\frac{\rho}{\rho_c}$, where $\rho_c=0.6$ g cm${^-3}$. This operation is equivalent to centering a log-transformed $\rho$ around $\rho_c$ and enables mean hyper-parameters  to be interpreted as the value for a species with average wood density of 0.6 cm${^-3}$. Similarly, we also centered $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.

We modelled species random effects as random realisations from log-normal distributions:
\begin{equation}
\alpha_{0,s} \sim \log \mathcal{N}\left(\mu_{\log \alpha_0}, \sigma_{\log \alpha_0}\right),
\end{equation}
with similar formulations for $\beta_{0,s}$ and $\gamma_{0,s}$.

All three mean hyperparameters, $\mu_{\log \alpha_0}$, the average log species residual error at zero growth, $\mu_{\log \beta_0}$, the average species error on the exponential decay rate, and $\mu_{\log \gamma_0}$, the average baseline log species residual error, were estimated using normal distributions centered on zero with a standard deviation of 2.5.
\begin{equation}
\mu_{\log \alpha_{0}},\mu_{\log \beta_0},\mu_{\log \gamma_0} \sim \mathcal{N}\left(0, 2.5\right).\\
\end{equation}

Standard deviations associated with species random effects (i.e. $\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$), were estimated using positive half Cauchy priors centered on zero with a scale of 2.5.
\begin{equation}
\sigma_{a0},\sigma_{b0},\sigma_{c0} \sim \textrm{Cauchy+}\left(0, 2.5\right).\\
\end{equation}

For models that did not include species random effects, single intercepts were estimated for each parameter using log normal distributions centered on zero with standard deviations of 2.5.

We estimated wood density effects (i.e. $\alpha_1$, $\beta_1$ and $\gamma_1$) using normal priors centered on zero and a standard deviation of 2.5:
\begin{equation} \label{eq:rho_effects}
\alpha_1, \beta_1, \gamma_1 \sim \mathcal{N}\left(0, 2.5\right).\\
\end{equation}

We accounted for possible variation in mortality rates across different census periods by allowing $\lambda_i$ to be scaled by $\delta_t$ for each census $t$. We estimated $\delta_t$ as a random realization from a log-normal distribution centered on zero and a variance parameter $\sigma_{\delta}$ defines the variation in census effects:

\begin{equation} \label{eq:census_effect}
\delta_t \sim \log \mathcal{N}\left(0, \sigma_{\delta}\right)\\
\sigma_{\delta} \sim \textrm{Cauchy+}\left(0, 2.5\right).
\end{equation}

### 2016-07-04 ms/supp.Rnw
Fixed typos so that supplement will at least compile. Very draft form at this stage

@@
 
+\dates{This manuscript was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
+
+<<preload, echo=FALSE>>=
+n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
+n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
+proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
+n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
+obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
+n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))
+
+@
+
 \begin{document}
-% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
-% You should only change this length when you've finalized the article contents.
-\verticaladjustment{-2pt}
-
-\maketitle
-\thispagestyle{firststyle}
-\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
-
+\section*{Supplementary}
 \subsection*{Modelling true growth}
@@
 \subsection*{Mortality model}
-\subsection*{Model fitting}
 Our observations are for survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$.
@@
 \end{equation}
+
+\end{document}

### 2016-07-05 ms/ms.Rnw
removed pesky typo where ht occured after figure
Attempt at new introduction

@@
 
-\title{Unifying intraspecific and interspecific variation in tropical tree mortality}
+\title{Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework}
 
@@
 \author[a,1]{James S Camac}
-\author[a]{Mark Westoby} 
-\author[b]{S Joseph Wright}
-\author[a]{Daniel S Falster}
+\author[a,2]{Mark Westoby}
+\author[a,3]{Daniel S Falster}
 
@@
 \affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
-\affil[b]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
 
@@
 % Please add here a significance statement to explain the relevance of your work
-\significancestatement{Add something significant here}
+% 120 words
+\significancestatement{Add something significant here.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
+Sed a magna eget lorem posuere euismod ac in enim.
+Nam porttitor placerat est, quis finibus quam efficitur nec.
+Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
+Maecenas eros sapien, semper quis velit nec, porta convallis est.
+Vestibulum porttitor diam ac massa porta pharetra.
+In a viverra justo.
+Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
+Aenean dignissim sollicitudin pharetra.
+Integer aliquet odio odio, vel bibendum mi lobortis nec.
+Aliquam non fermentum elit, id pellentesque ex.}
 
@@
 designed and performed the analysis; D.F.
-conceived the idea, supervized the development of this work; S.J.W \& M.W.
+conceived the idea, supervized the development of this work; M.W.
 provided biological insight.
@@
 
+% 250 words
 \begin{abstract}
-Death is an integral process driving population dynamics, nutrient cycling, and evolution within plant communities.
-While past research has identified factors influencing tree mortality -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these are yet to be integrated within a general predictive framework.
-Here we estimate instantenous mortality rates of 203 tropical rainforest tree species at Barro Colorado Island (BCI), Panama, by combining classic survival analysis techniques with a cross-validated hierarchical Bayesian framework that integrates individual-level, species-level, and plot-level effects.
-We show that greatest predictive capacity is achieved when when models simultaneously estimate growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth and account for wood density effects on growth-independent terms.
+Tree death is a key process driving population dynamics, nutrient cycling, and evolution within plant communities.
+While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these distinct drivers are yet to be integrated within a cohesive and predictive framework.
+Here we apply classic techniques from survival analysis within a hierarchical Bayesian and cross-validation framework to derive instantaneous mortality functions for tropical rain forest trees at Barro Colorado Island (BCI) Panama, integrating individual-level, species-level, and plot-level effects.
+In reaching this point, we show how mortality models including separate growth-dependent and growth-independent components, using stem diameter growth rather than basal-area growth, and attributing the effects of wood density to growth-independent death, outperform alternative formulations.
 Moreover, we show that the effects of wood density -- a prominent trait known to influence mortality -- capture only a portion of total species-level differences; and that the year-to-year variation in mortality was large.
-Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the largest driver of mortality at BCI.
+Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI.
 Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
@@
 n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))
-
 @
@@
 
-\dropcap{P}lant mortality is often modelled as a function of recent growth because it reflects the product of many biotic and abiotic factors on a plants overall carbon budget and vigour \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db, Wyckoff:2002ul}.
-As such, growth rates are considered a highly integrated individual-based predictor of plant mortality \cite{Hawkes:2000ib}.
-Within a species, the assumption is that fast growing individuals (i.e.
-those that have larger carbon budgets) are more able to buffer stress and therefore have higher survivorship relative to slow growing (i.e.
-low carbon budget) individuals \cite{Hawkes:2000ib, Keane:2001db}.
-Empirical studies examining this growth\textendash mortality relationship have broadly supported this assumption, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
-Consequently, most dynamic vegetation models, including tree gap models \cite{Keane:2001db}, individual-based forest simulators \cite[e.g.
-SORTIE][]{Pacala:1996vm} and Dynamic Global Vegetation Models \cite[DGVMs; ][]{Woodward:2004ft} use some measure of growth coupled with a stochastic constant (i.e.
-baseline hazard or minimum chance of death) as the fundamental determinants of plant mortality \cite{Hawkes:2000ib, McDowell:2011dr}.
+\dropcap{P}lant mortality rates are known to vary among individuals within a species, among coexisting species within the same forest, from year-to-year, and between forests [refs].
+Current estimates suggest the range of variation is large, e.g.
+from xxx xxx among plants within the same forest, and from XX-XXX among forests within any given year, or between years.
+This variation has considerable consequences for forest structure and dynamics.
+Death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools, while at the same time opening up a gap that restarts a successional race, during which 10's-100's more plants may die while competing for a spot in the sun.
+In models of forest dynamics, variation in mortality rates has been found to have a larger impact on forest structure than absolute growth rates \cite{Johnson-2016}.
+Improving our ability to understand the mortality process is therefore a priority for making skilful predictions about forest dynamics and carbon, water or nutrient cycles; especially in an era of rapid environmental change.
 
-However, due to a paucity of data on plant mortality, most dynamic vegetation models have implemented mortality algorithms with overly simplistic assumptions \cite{Hawkes:2000ib, Wyckoff:2000un, McDowell:2011dr}.
-One of the strongest assumptions made is that both stochastic hazards and the growth-mortality relationship are constant across species \cite{Pacala:1996vm, Hawkes:2000ib, Wyckoff:2000un}.
-That is, all species have the same random chance of death and exhibit the same tolerance to low growth.
-This is despite growing empirical evidence highlighting that mortality rates vary considerably between species, even within the same community \cite{Kobe:1997vy, Wyckoff:2002ul, Kunstler:2005bn, Vieilledent:2010fv}.
-Furthermore, because no systematic analysis has been conducted to determine what growth measure best predicts plant mortality, the growth-mortality relationship is often parameterized using a variety growth measures; two of the most common being rates of DBH or basal area growth.
-
-Recently, research has focused on examining inter-specific variation by correlating intrinsic species' growth rates with population mortality rates \cite[e.g.][]{Poorter:2008iu, Wright:2010fl}.
-In contrast to intraspecific work, these studies have shown that species with intrinsically faster growth rates exhibit higher mortality rates relative to slower growing species \cite{Loehle:1996vw, Poorter:2008iu, Wright:2010fl}.
-As such, species appear to exhibit a trade off between greater growth versus greater survival.
-Ecologists have attempted to explain this interspecific variation by using a variety of functional traits \textemdash attributes of species that are thought to influence vital rates (i.e.
-survival, growth and reproduction) and ultimately fitness \cite{Westoby:2002tb, Ackerly:2003eb, Falster:2015un}.
-
-Wood density, the biomass invested per unit wood volume, is one such functional trait that has consistently been shown to be negatively correlated with tropical tree mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
-The underlying mechanism of this relationship is thought to be due to denser wood conveying enhanced physical resistance to random events such as stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx} and fungal and pathogen attack \cite{Augspurger:1984wx}.
-As such, it is hypothesized that species with higher wood density are likely to exhibit lower stochastic hazards.
-However, research has also shown that wood density may increase tolerance to low growth rates, and thus alter the growth-mortality relationship.
-For example, species with high wood density have been found to be more tolerant to both shade \cite{vanGelder:2006exa} and neighbor competition \cite{Kunstler:2016km}.
-Studies have also shown that wood density may reduce drought induced carbon starvation \cite{McDowell:2011dr} and the susceptibility of ‘stressed’ individuals to opportunistic invertebrate, pathogen or fungal attack \cite{Augspurger:1984wx, McDowell:2011dr}.
-Consequently, species with high wood density may be more tolerant to both stochastic events such as breakage as well as be more tolerant to periods of low growth.
-
-What both intraspecific and interspecific studies have highlighted is that both individual (e.g.
-growth rates) and species (e.g.
-function traits) characteristics are likely to be required in order to develop general and accruate models of plant mortality.
-However, to date, there has been little crossover of ideas between intraspecific and interspecific studies of plant mortality.
-For example, we are aware of only one study that uses individual based plant mortality empirical models to examine trait effects on baseline hazards \cite{AubryKientz:2013dg}, and none that examine how traits mediate the growth-mortality relationship.
+Two substantial challenges exist when studying the mortality process in trees.
+The first is that very large population sizes and/or very long periods of observation are needed for effective inference.
+This requirement arises is in part because the observable outcome -- alive vs.
+dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate).
+In classic survival analysis \cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
+\begin{equation} \label{eq:mortality_model}
+p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
+\end{equation}
+Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$ \cite{Kleinbaum:2012tn}.
+The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
+\begin{equation}
+S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
+\end{equation}
+Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are needed, e.g.
+100s-1000s records.
+As such, detailed studying of mortality for forest trees has only recently become possible, with the accumulation of data from repeat surveys in long-term monitoring plots containing 1000's trees and spanning several decades \cite{Anderson-Teixeira-2015a}.
 
@@
 B) We consider three alternative mortality functions, combining growth-independent and growth-dependent effects.
+The parameters of the model are biologically interpretable: $\alpha$ defines the instantaneous mortality rate at low growth rate; $\beta$ reflects the sensitivity of mortality rate to changes ingrowth rate; and $\gamma$ is the asymptote, or baseline hazard.
+Combined $\alpha$ and $\beta$ capture growth-dependent mortality, while $\gamma$ captures growth-independent mortality (e.g.
+windfall, fire) that kill a plant irrespective of its growth rate.
 For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
-C) The predictive skill of alternative models is evaluated via 10-fold cross validation.
+C) Log-loss function
+D) The predictive skill of alternative models is evaluated via 10-fold cross validation.
 The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$).
-We then refit each alternative model 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing and training data.
+Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing and training data.
 We then average the log loss $\mathcal{L_i}$ for each of these 10 model fits, to give an overall average log loss for each model $\bar{\mathcal{L}}$.}
@@
 
-In this study, we reconcile both intraspecific and interspecific mortality theory by combining classic survival analysis techniques with a cross-validated hierarchical Bayesian framework.
-Specifically, our framework  instantaneous mortality rates for 203 tropical rainforest trees at Barro Colorado Island (BCI) by integrating individual-level, species-level, and plot-level effects and simultaneously estimating growth-dependent and growth-independent mortality processes.
-Using this framework we
-identify the most parsimonious model of tropical tree instantaneous mortality rates.
-We also ask: 1) which  growth measure (i.e.
-diameter increment growth or area growth)best predicts growth-dependent mortality; 2) Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
-3) What is the proportion of species variability explained by wood density?
+The other major challenge when studying mortality -- similar to that faced in other fields like medicine and engineering -- is to derive suitable shapes for the underlying instantaneous mortality (hazard) function.
+Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate, traits, and size (XX).
+Generally, however, these influences have not been formalised as hazard functions.
+Yet such knowledge is needed for predicting mortality into the future and in models of ecosystem function (XX).
+For projecting the outcomes of a population, hazard functions are invaluable because they ....
 
-\section*{Framework for modelling mortality}
+A particular challenge is to integrate the apparently conflicting relationships of mortality to growth observed within and across species.
+Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
+Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
+Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species also having higher mortality than individuals from slower growing species \cite{Poorter:2008iu, Wright:2010fl}.
+This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume).
+Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack \cite{Augspurger:1984wx}, and ultimately reduced mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
 
-In classic survival analysis, the probability that an individual dies between $t_2$ and $t_3$, $p_{t_2\rightarrow t_3}$ is defined by
-\begin{equation} \label{eq:mortality_model}
-p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right),
-\end{equation}
-where $\lambda_i(t)$ is the instantaneous mortality rate of death per unit time \cite{Kleinbaum:2012tn} and $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the cumulative hazard between $t_2$ and $t_3$.
-The challenge then is to identify suitable forms of the instantaneous hazard (mortality) rate function.
-
-We examine the performance of three functional forms of $\lambda_i(t)$ to predict observed patterns of tropical tree mortality:
-\begin{align}
-\lambda_i &= \overbrace{\gamma}^{\text{baseline hazard}} \label{base_haz}\\
-\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} \label{growth_haz}\\
-\lambda_i &= \overbrace{\alpha \, \exp\left(-\beta \, X_i \right)}^{\text{Growth-dependent hazard}} +
-\overbrace{\gamma.}^{\text{Baseline hazard}} \label{add_haz}
-\end{align}
-The simplest model (eq.
-\ref{base_haz}), assumes a single baseline mortality rate, which is constant irrespective of a plant's growth rate.
-The second model (eq.
-\ref{growth_haz}) describes a growth-dependent mortality rate which assumes plants can only die due to a growth related cause.
-In this model, we make an explicit assumption that the growth-mortality relationship is captured by a negative exponential curve whereby the probability of death is high at low growth rates and as growth increases the probability of dying decays at an exponential rate until an asymptote at zero is reached.
-This negative exponential relationship has been used in other studies to model growth-dependent mortality \cite[e.g.][]{Kobe:1995tw} and broadly reflects patterns observed in non-parametric growth-mortality models \cite{Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}.
+In this study, we reconcile both intraspecific and interspecific drivers of instantaneous mortality into a unified model, including individual-level, species-level, and plot-level effects, within a cross-validated hierarchical Bayesian framework for 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama.
+To derive this unified model, we compare the skill of three alternative functional forms for $\lambda_i(t)$ in predicting patterns of tropical tree mortality in a diverse tropical forest in Panama (Fig.
+\ref{fig1}B).
+The simplest model (Fig.
+\ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate.
+The second model (Fig.
+\ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases.
+Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached.
 The third model (eq.
-\ref{add_haz}) -- potentially the most biologically realistic -- is the summation of the two previous models.
-Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth, but, unlike eq.
-\ref{growth_haz}, will asymptote at some baseline rate instead of zero.
+Fig.
+\ref{fig1}B, right) is the summation of the two previous models.
+Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero.
 The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e.
 stochastic events such as wind throw and fire).
-The parameters of eq.
-\ref{add_haz} are also biologically interpretable: $\alpha$ represents the intercept and defines the instantaneous mortality rate $\lambda$ at low annual growth rate.
-$\beta$ represents the exponential decay rate or the sensitivity of mortality rates to change with growth rate; and $\gamma$, is the asymptote that defines the baseline hazard.
-$\alpha$ and $\beta$ are intrinsically linked to a plants growth rate, and as such are growth-dependent parameters.
-By contrast, $\gamma$ is a growth-independent parameter that accounts for stochastic events (e.g.
-windfall, fire) that can kill a plant irrespective of its growth rate.
+Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (refs).
+However, these alternative models have not previously been systematically compared.
 
-The central aim of this study is to identify the most parsimonious combination of $\lambda$, growth measure (diameter increment growth or area growth) and wood density that best predicts tropical tree mortality.
-To do this, we used K-fold cross-validation and four iterative stages of model fitting and comparison.
-Stage 1 involved comparing predictive performance of the three $\lambda$ models described in equations \ref{base_haz} \textendash \ref{add_haz} using either annual diameter increment or area growth rates.
+In deriving the unified model, we address the following specific hypotheses:
+\begin{enumerate}\itemsep 0pt
+  \item Does mortality vary substantially between censuses?
+  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
+  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
+  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
+  \item What is the proportion of total inter-specific variability explained by wood density?
+\end{enumerate}
+Answering these questions involves comparing predictive skill of alternative models.
+To achieve this, we used K-fold cross-validation (Fig.
+\ref{fig1}D).
+Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}.
+Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations.
+We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the five hypotheses evaluated above, and gradually increasing model complexity.
+At each stage, the best model from the previous stage was taken as input for the next stage.
+for example, stage 2 involved comparing predictive performance of the three $\lambda$ models described in (Fig.
+\ref{fig1}B) using either annual diameter increment or area growth rates.
 In this stage each model was fitted without allowing parameters to vary by species.
 This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure.
-Following this, we used the best model of stage 1 but allowed its parameters to vary by species.
+Following this, we used the best model of stage 2 but allowed its parameters to vary by species.
 This allowed us to determine how variable each parameter was across species and whether accounting for species differences significantly improved predictive performance.
-Because species responses varied considerably, we undertook a third stage of model fitting and selection that explicitly examined whether wood density (a species level trait) explained some of this interspecific variability.
-This involved using the best model observed in stage 1 and adding wood density effects to each parameter.
-Finally we then fitted the best wood density combination model with species random effects to the entire dataset.
-This model was then used to examine mortality curves across wood density gradients.
-In order to determine the proportion of interspecific variation explained by wood density on each parameter it occurred on we reran the final model form with just random effects on parameters and calculated the percent reduction in unexplained species variation.
-
-At each stage, we evaluated the predictive accuracy of alternative models or predictor variables using K-fold cross-validation (Fig.
-\ref{fig1}).
-This involves randomly partitioning the data into ten folds of \Sexpr{n_heldout} observations, while ensuring that each K-fold had at least one observation per species.
-We then fitted models to nine folds and predicted to the tenth fold.
-This was repeated until each fold had been used nine times for model fitting, and once as a held-out dataset for predicting.
-Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s predictive accuracy \cite{Hooten:2015, Stone:1997tg}.
 
@@
 \begin{figure}[ht]
-\includegraphics{figures/fig4.pdf}[ht]
+\includegraphics{figures/fig4.pdf}
 \caption{Predicted instantaneous mortality rates its derivative annual mortality probability.
@@
 \begin{figure}[ht]
-\includegraphics{figures/fig5.pdf}[ht]
+\includegraphics{figures/fig5.pdf}
 \caption{Predicted instantaneous mortality rates its derivative annual mortality probability.
@@
 \begin{figure}[ht]
-\includegraphics{figures/fig6.pdf}[ht]
+\includegraphics{figures/fig6.pdf}
 \caption{Predicted instantaneous mortality rates its derivative annual mortality probability.
@@
 
+The most complicated function is
+\begin{equation}
+\lambda_i = \big(\underbrace{\alpha \, \exp\left(-\beta \, X_i \right)}_{\text{Growth-dependent}} +  \underbrace{\gamma}_{\text{Baseline}}\big) \underbrace{\delta_t}_{\text{Census-effect}}.
+ \label{add_haz}
+\end{equation}
+
 \subsection*{Model fitting}

### 2016-07-05 R/figures.R

@@ -143,6 +143,8 @@ logloss_curve <- function() {
 # Map of plot gap index with recruits overlayed for 1985 to 1990 and 1990 to 1995
 plot_recruits_gapindex <- function(gap_index_raster, recruit_gap_conditions) {
   #pdf(f <- tempfile(fileext = '.pdf'), 4.5, 4.5)
+  greys <- grey.colors(20, start =0.1, end=1.0)
+
   print(rasterVis::levelplot(raster::stack(gap_index_raster), 
                        names.attr=paste("Recruits from", names(gap_index_raster)),
                        layout=c(1, 2), 
@@ -150,13 +152,13 @@ plot_recruits_gapindex <- function(gap_index_raster, recruit_gap_conditions) {
                        scales=list(alternating=FALSE, tck=c(0.5, 0.5), 
                                    x=list(at=seq(0, 1000, 200)),
                                    y=list(at=seq(0, 400, 100))),
-                       col.regions=viridisLite::viridis(100),
-                       at=seq(0,1,0.1), 
+                       col.regions= greys[c(1,1, seq_along(greys), 20, 20)],
+                       at=seq(0,1,0.05),
                        par.settings = list(fontsize = list(text = 7), 
                                            strip.background=list(col='white'),
                                            strip.border=list(col='transparent'))) +
           latticeExtra::layer(sp::sp.points(SpatialPoints(recruit_gap_conditions[[panel.number()]]), 
-                                        pch='.', col='black', alpha=0.4), 
+                                        pch='+', col='red', alpha=0.4, cex=0.1),
                               data=list(recruit_gap_conditions=recruit_gap_conditions)))
   
 }
@@ -218,7 +220,7 @@ plot_fig1 <- function(tree1, tree2, panelc) {
   
   my_label <- function(text, x=-0.1) label(x, 1.3, text, cex=1.5)
   
-  layout(matrix(c(1,1,1,2,3,4,5,5,5), byrow=TRUE, ncol=3))
+  layout(matrix(c(1,1,1,2,3,4,5,5,5,6,6,6), byrow=TRUE, ncol=3))
   par(mar=c(2,2,3,1), oma=c(1,1,1,1))
   
   # Tree growth diagram
[…]

### 2016-07-07 R/figures.R
Resized figure 1

@@ -210,23 +210,25 @@ plot_spp_param_by_covariate <- function(data, focal_param, covariate ="mean_gap_
 # Manuscript figures
 
 plot_fig1 <- function(tree1, tree2, panelc) {
+  # Example growth hazard function
   growth_haz <- function(x) {
     1-exp(-0.15 * exp(-10 * x))
   }
   
+  # Example base-growth hazard function
   base_growth_haz <- function(x) {
     1-exp(-(0.13 * exp(-10 * x) + 0.03))
   }
+  # Function to add labels to panels
+  my_label <- function(text, x=-0.1) label(x, 1.3, text)
   
-  my_label <- function(text, x=-0.1) label(x, 1.3, text, cex=1.5)
-  
+  # Plot layout
   layout(matrix(c(1,1,1,2,3,4,5,5,5,6,6,6), byrow=TRUE, ncol=3))
-  par(mar=c(2,2,3,1), oma=c(1,1,1,1))
+  par(mar=c(1,1,2.5,1), oma=c(1,1,1,1), cex=0.5)
   
   # Tree growth diagram
-  
   plot(1:2, type='n', ann=FALSE, axes=FALSE, xlim=c(-1,1), ylim=c(-1,1))
-  my_label("A) Repeat census data")
+  my_label("A) Repeat census data", x=-0.06)
   
   # note we are only plotting to last viewport, but calls to other two needed to make figure work.
   vps <- baseViewports()
@@ -234,41 +236,47 @@ plot_fig1 <- function(tree1, tree2, panelc) {
   fig.tree(tree1, tree2)
   popViewport(3)
   
-  # Example baseline hazard
+  # Produce empty plot
   empty_plot <- function(){
+    par(mar=c(2,2,3,1))
     plot(NA, ylim = c(0,0.14),  xlim=c(0,1), xaxs='i', axes=FALSE, ann=FALSE)
-    axis(1, at = c(-1, 2), labels=NA, lwd=1.5)
-    axis(2, at = c(-1, 2),labels=NA, lwd=1.5)
+    axis(1, at = c(-1, 2), labels=NA)
+    axis(2, at = c(-1, 2),labels=NA)
   }
   
+  # Plot hazard curves
   hazard_plot <- function(f, text){
     x <- seq(0,1,length.out = 50)
     empty_plot()
-    lines(x, f(x), type='l', lwd=2, col="red")
-    text(0.1, 0.12, text, cex=1.5, pos=4, xpd=NA)
+    lines(x, f(x), type='l', col="red")
[…]

### 2016-07-07 R/manuscript_functions.R
Retweaked these functions

@@ -1,37 +1,59 @@
+# Manuscript functions
 
-logloss_comparisons <- function(logloss_summary) {
-  logloss_summary %>%
-    filter(model_type %in% c("function_growth_comparison_base_growth_hazard_none", 
-                             "rho_combinations_base_growth_hazard_c",
-                             "species_random_effects_base_growth_hazard_none") & growth_measure=='true_dbh_dt') %>%
-    select(model_type, mean) %>%
-    mutate(model_type = factor(model_type, labels =c('base_growth','base_growth_rho','base_growth_re'))) %>%
-    spread(model_type, mean) %>%
-    mutate(rho_v_base = round((1-base_growth_rho/base_growth)*100,2),
-           re_v_base = round((1-base_growth_re/base_growth)*100,2),
-           rho_v_re = round((rho_v_base/re_v_base)*100,2))
+# Number of individuals used
+n_inds <- function(data) {
+  as.character(round(nrow(data),-2))
 }
 
+# Number of species used in study
+n_spp <- function(data) {
+  length(unique(data$sp))
+}
+
+# Proportion of negative observations
+prop_neg_growth <- function(data) {
+  round(nrow(subset(data, obs_dbh_dt<0))/nrow(data)*100)
+}
+
+# Number of observations in remeasurement error data
+nobs_meas_err_data <- function(data) {
+  nrow(data$discrep)
+}
+
+# Observed standard deviation of measurement error
+obs_error_sigma <- function(data) {
+  round(sd(data$discrep),2)
+}
+
+# Number of heldout observations
+n_heldout <- function(data) {
+  as.character(nrow(data[[1]]$heldout))
+}
+  
+# Proportion of deaths by census
+prop_deaths <- function(data, census) {
+res <-data %>%
+  group_by(censusid) %>%
+  summarise(deaths = sum(dead_next_census),
+            survivors = sum(dead_next_census==0)) %>%
+    ungroup() %>%
+  mutate(
+    total = survivors + deaths,
+    prop_died = deaths/total,
+    census = factor(censusid, labels=c('1995to2000','2000to2005','2005to2010'))) %>%
+  select(census,censusid, total, deaths, survivors, prop_died)
+return(signif(res[res$censusid==census,"prop_died"],2)*100)
+}
+
+# Extract census scaling factor
 census_effects <- function(model,census) {
-  model <- model$fits[[1]]
[…]

### 2016-07-07 ms/ms.Rnw
Edited intro. rewrote results. Need to add a couple of numbers and references in places

@@
 \author[a,1]{James S Camac}
-\author[a,2]{Mark Westoby}
-\author[a,3]{Daniel S Falster}
+\author[a]{Mark Westoby}
+\author[b]{S Joseph Wright}
+\author[a]{Daniel S Falster}
 
@@
 % Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
-\keywords{Tropical trees $|$ wood density $|$ growth $|$ mortality $|$ interspecific}
+\keywords{Bayesian $|$ Growth rates $|$ Hazard rates $|$ Tropical trees $|$ Wood density}
 
@@
 \begin{abstract}
-Tree death is a key process driving population dynamics, nutrient cycling, and evolution within plant communities.
+Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities.
 While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these distinct drivers are yet to be integrated within a cohesive and predictive framework.
-Here we apply classic techniques from survival analysis within a hierarchical Bayesian and cross-validation framework to derive instantaneous mortality functions for tropical rain forest trees at Barro Colorado Island (BCI) Panama, integrating individual-level, species-level, and plot-level effects.
-In reaching this point, we show how mortality models including separate growth-dependent and growth-independent components, using stem diameter growth rather than basal-area growth, and attributing the effects of wood density to growth-independent death, outperform alternative formulations.
-Moreover, we show that the effects of wood density -- a prominent trait known to influence mortality -- capture only a portion of total species-level differences; and that the year-to-year variation in mortality was large.
+Here we apply classic survival analysis techniques within a cross-validated hierarchical Bayesian framework to derive instantaneous mortality functions that integrate individual, species, and plot-level effects for 203 tropical rain forest tree species at Barro Colorado Island (BCI) Panama.
+We show that mortality models that separate growth-dependent and growth-independent components, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations.
+Moreover, we show that the effects of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability in species mortality rates, and that temporal variation in mortality was large.
 Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI.
@@
 \doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
-
-<<preload, echo=FALSE>>=
-n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
-n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
-proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
-n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
-obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
-n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))
-@
 
@@
 
-\dropcap{P}lant mortality rates are known to vary among individuals within a species, among coexisting species within the same forest, from year-to-year, and between forests [refs].
-Current estimates suggest the range of variation is large, e.g.
-from xxx xxx among plants within the same forest, and from XX-XXX among forests within any given year, or between years.
+\dropcap{R}ates of plant mortality are known to be highly variable among individuals, among coexisting species, between forests and from year-to-year [refs].
 This variation has considerable consequences for forest structure and dynamics.
-Death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools, while at the same time opening up a gap that restarts a successional race, during which 10's-100's more plants may die while competing for a spot in the sun.
+Death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools (Chave et al 200X), while at the same time opening up a gap that restarts a successional race, during which 10's-100's more plants may die while competing for a spot in the sun.
 In models of forest dynamics, variation in mortality rates has been found to have a larger impact on forest structure than absolute growth rates \cite{Johnson-2016}.
-Improving our ability to understand the mortality process is therefore a priority for making skilful predictions about forest dynamics and carbon, water or nutrient cycles; especially in an era of rapid environmental change.
+As such, improving our ability to understand the mortality process is therefore a priority for making skilful predictions about forest dynamics and carbon, water or nutrient cycles; especially in an era of rapid environmental change.
 
-Two substantial challenges exist when studying the mortality process in trees.
-The first is that very large population sizes and/or very long periods of observation are needed for effective inference.
-This requirement arises is in part because the observable outcome -- alive vs.
+Two substantial challenges exist when studying tree mortality in tropical rainforests.
+The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates.
+This requirement arises in part because the observable outcome -- alive vs.
 dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate).
@@
 \end{equation}
-Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are needed, e.g.
-100s-1000s records.
-As such, detailed studying of mortality for forest trees has only recently become possible, with the accumulation of data from repeat surveys in long-term monitoring plots containing 1000's trees and spanning several decades \cite{Anderson-Teixeira-2015a}.
+Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required.
+As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys in long-term monitoring plots containining thousands of individuals and spanning several decades \cite{Anderson-Teixeira-2015a}.
 
-\begin{figure}[ht]
+The other major challenge when studying mortality -- similar to that faced in other fields like medicine and engineering -- is to derive suitable shapes for the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality.
+Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate, traits, and size (XX).
+Generally, however, these influences have not been formalised as hazard functions (XX).
+Yet such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function (XX).
+This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.
+
+A particular challenge is to integrate the observed conflicting relationships between mortality and growth observed within and across species.
+Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
+Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
+Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality than individuals from slower growing species \cite{Poorter:2008iu, Wright:2010fl}.
+This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume).
+Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack \cite{Augspurger:1984wx}, and ultimately reduced mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+
+\begin{figure}[!hb]
 \centering
@@
 \caption{Outline of methodology.
-A) Our data consist of repeat measures on stem diameter ($D$) and status ($S$, alive or dead) for individual tress at specific census dates ($t_1$, $t_2$, $t_3$).
-B) We consider three alternative mortality functions, combining growth-independent and growth-dependent effects.
-The parameters of the model are biologically interpretable: $\alpha$ defines the instantaneous mortality rate at low growth rate; $\beta$ reflects the sensitivity of mortality rate to changes ingrowth rate; and $\gamma$ is the asymptote, or baseline hazard.
+A) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, alive or dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
+B) We consider three alternative hazard functions: 1) a baseline hazard, 2) a growth-dependent hazard; and 3) a function that combines both baseline and growth-dependent hazards.
+The parameters of the models are biologically interpretable: $\alpha$ defines the instantaneous mortality rate at low growth rate; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote, or baseline hazard.
 Combined $\alpha$ and $\beta$ capture growth-dependent mortality, while $\gamma$ captures growth-independent mortality (e.g.
@@
 For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
-C) Log-loss function
-D) The predictive skill of alternative models is evaluated via 10-fold cross validation.
+C) Log-loss function highlighting that the further the predicted probability is from truth the higher the logloss.
+D) The predictive accuracy of alternative models is evaluated via 10-fold cross validation.
 The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$).
 Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing and training data.
-We then average the log loss $\mathcal{L_i}$ for each of these 10 model fits, to give an overall average log loss for each model $\bar{\mathcal{L}}$.}
+We then average the estimated log loss for each of these 10 model fits.}
 \label{fig1}
@@
 
-The other major challenge when studying mortality -- similar to that faced in other fields like medicine and engineering -- is to derive suitable shapes for the underlying instantaneous mortality (hazard) function.
-Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate, traits, and size (XX).
-Generally, however, these influences have not been formalised as hazard functions.
-Yet such knowledge is needed for predicting mortality into the future and in models of ecosystem function (XX).
-For projecting the outcomes of a population, hazard functions are invaluable because they ....
-
-A particular challenge is to integrate the apparently conflicting relationships of mortality to growth observed within and across species.
-Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
-Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
-Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species also having higher mortality than individuals from slower growing species \cite{Poorter:2008iu, Wright:2010fl}.
-This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume).
-Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack \cite{Augspurger:1984wx}, and ultimately reduced mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
-
-In this study, we reconcile both intraspecific and interspecific drivers of instantaneous mortality into a unified model, including individual-level, species-level, and plot-level effects, within a cross-validated hierarchical Bayesian framework for 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama.
-To derive this unified model, we compare the skill of three alternative functional forms for $\lambda_i(t)$ in predicting patterns of tropical tree mortality in a diverse tropical forest in Panama (Fig.
-\ref{fig1}B).
+In this study, we reconcile both intraspecific and interspecific drivers of instantaneous mortality by developing a unified model that incorporates individual-level, species-level, and plot-level effects, within a cross-validated hierarchical Bayesian framework.
+To derive this unified model, we compare the predictive accuracy of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig.
+\ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama.
 The simplest model (Fig.
@@
 \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases.
+
 Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached.
@@
 \ref{fig1}B, right) is the summation of the two previous models.
-Specifically, this model allows $\lambda$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero.
+Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero.
 The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e.
 stochastic events such as wind throw and fire).
-Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (refs).
+Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (Moorcroft 2001).
 However, these alternative models have not previously been systematically compared.
 
-In deriving the unified model, we address the following specific hypotheses:
+In deriving the unified model, we address the following specific questions:
 \begin{enumerate}\itemsep 0pt
@@
   \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
-  \item What is the proportion of total inter-specific variability explained by wood density?
+  \item What is the proportion of inter-specific variability explained by wood density?
+  \item What is the proportion of total variability explained by species, census, baseline hazard and growth-dependent hazard?
 \end{enumerate}
-Answering these questions involves comparing predictive skill of alternative models.
+Answering these questions involves comparing predictive accuracy of alternative models.
 To achieve this, we used K-fold cross-validation (Fig.
@@
 Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations.
-We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the five hypotheses evaluated above, and gradually increasing model complexity.
-At each stage, the best model from the previous stage was taken as input for the next stage.
-for example, stage 2 involved comparing predictive performance of the three $\lambda$ models described in (Fig.
-\ref{fig1}B) using either annual diameter increment or area growth rates.
-In this stage each model was fitted without allowing parameters to vary by species.
-This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure.
-Following this, we used the best model of stage 2 but allowed its parameters to vary by species.
-This allowed us to determine how variable each parameter was across species and whether accounting for species differences significantly improved predictive performance.
+For example, some of our models took over a two weeks to complete.
+We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increasing model complexity (see methods).
 
@@
 
-\subsection*{Predictive capacity of functional form and growth measure}
-Using 10-fold cross validation our first aim was to determine which model form (Fig.
-\ref{fig1}b) and growth measure (i.e.
-DBH growth or basal area growth) best predicted tree mortality.
-We found that the model which accounted for both baseline (i.e.
-growth-independent) and growth-dependent hazards had a significantly lower logarithmic loss, and thus, a higher predictive capacity, relative to either the baseline only or growth-dependent only model variants (Fig.
-\ref{fig2}a).
-However, this improvement was dependent on growth measure, with predictive capacity increasing when DBH growth was used but not when basal area growth was used (Fig.
-\ref{fig2}a).
+\subsection*{Mortality strongly varies over time}
+Tree mortality progressively decreased over time.
+Between 1995 and 2010, the highest proportion of trees dying between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and lastly 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%).
+Consequently, when we allowed hazard rates to be scaled by census (i.e.
+adding term $\delta$) we observed a significant increase in predictive accuracy (i.e.
+lower logarithmic loss; Fig.
+\ref{fig2}A).
+Assuming the long-term average census effect ($\delta$) is 1, then our model predicted deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.
 
-\begin{figure}[ht]
+\begin{figure}[!h]
 \includegraphics{figures/fig2.pdf}
-\caption{Average logarithmic loss (lower = higher accuracy).
-A) Four sequential stages of model selection.
-Null model: shows the expected logloss for the null model with no census effect; because mortality varied substantially between censuses it was included on all subsequent models; Functional form: comparison between baseline hazard model ($\gamma c$; circle), growth\textendash dependent hazard model ($(\alpha e^{\beta X_{i}}) c$; triangle) and model incorporating both baseline and growth\textendash dependent hazards ($(\alpha e^{\beta X_{i}} + \gamma) c$; square); Wood density: Most parsimonious effect of wood density; Species effects: Model with species random effects on all parameters \textendash indicating the total gain in predictive capacity if accounting for all species variability.
-B) Wood density effects on all parameter combinations in the model that incorporates both baseline and growth-dependent hazards.
+\caption{Mean ($\pm$ 95\% credible intervals) logarithmic loss whereby lower values equal higher predictive accuracy. 
+A) Five sequential stages of model selection with each stage increasing in complexity.
+Null: null model ($\gamma$); Census: inclusion of census effects ($\delta$); Growth rate: the inclusion of growth rate.
+This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{\beta X_{i}})\delta$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{\beta X_{i}} + \gamma)\delta$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
+B) Change in logloss under all wood density\textendash parameter combinations.
 Colors represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth.
-Errors are $\pm$ 95\% BCI.}
+Symbols represent general functional form: null/baseline (circle), growth\textendash dependent hazard model (triangle) and model incorporating both baseline and growth\textendash dependent hazards (square).
+Note: errors are too small to see in panel A.}
 \label{fig2}
@@
 
-\subsection*{Predictive capacity from including wood density}
-Using the model with both baseline and growth-dependent hazards coupled with DBH growth we then examined: 1) whether predictive capacity increased if parameters varied as a function of a species' wood density; 2) whether the effect of wood density was predominately on growth-dependent or baseline (growth-independent) hazard parameters; and 3) what proportional gain is achieved from including wood density relative to a full species random effect model that implicitly captures all species level variation.
-We found that the inclusion of wood density on any combination of the three parameters significantly improved predictive performance (i.e.
-lowered logarithmic loss) relative to a model without wood density effects (Fig.
-\ref{fig2}b).
-Furthermore, the greatest gain in predictive accuracy occurred when we allowed the baseline hazard $\gamma$ to vary by wood density, which accounted for \Sexpr{logloss_comparisons(logloss_summaries)[['rho_v_re']]}\% of the maximum predictive gain achievable in the fall species random effect model (Fig.
-\ref{fig2}a).
+\subsection*{Growth hazard functions}
+We found that the hazard function that incorporates both growth-dependent and baseline hazard terms (i.e.
+$(\alpha e^{\beta X_{i}} + \gamma)\delta$) and used stem diameter growth rather than basal-area growth, significantly outperformed both the baseline only (i.e.$\gamma \delta$) or growth-dependent only (i.e.
+$(\alpha e^{\beta X_{i}})\delta$) hazard functions (Fig.
+\ref{fig2}A).
+Surprisingly, we found that the growth-dependent only function did not improve predictive capacity over the baseline only function.
+Furthermore, we found that while the hazard function that incorporates both baseline and growth-dependent terms significantly improved model accuracy, this was only apparent if dbh growth was used.
 
-\begin{figure}[ht]
+\begin{figure}[h!]
 \includegraphics{figures/fig3.pdf}
-\caption{Estimated relationship between wood density and baseline mortality rate. Points are species estimates and solid line is the mean species trendline. All errors are $\pm$ 95\% BCI.}
+\caption{Relationship between species baseline mortality rate and wood density.
+Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study.
+Blue trendline with grey shading shows average ($\pm$ 95\% credible intervals) expected relationship.}
 \label{fig3}
@@
 
-\subsection*{Effects of wood density and time}
-Once determining that the most predictive model was one that accounted for both growth-dependent and baseline hazards, used DBH growth, and allowed $\gamma$ to vary by wood density, we then fitted two final models to the entire dataset.
-The first contained both species random effects on all parameters and wood density on the baseline hazard, $\gamma$.
-This allowed us to examine the range of species growth-mortality curves, the effect of wood density and how mortality varied over time.
-We also fitted a second model that only contained species random effects.
-This allowed us to determine which parameter varied most by species and what proportion of the baseline hazard species variability was explained by wood density.
-We found that baseline hazard rates were significantly lower for species with high wood density relative to those with low wood density (Fig.
+\subsection*{Effect of wood density on mortality rates}
+The inclusion of wood density on either baseline or growth-dependent parameters significantly improved model accuracy relative to a hazard function without such effects (Fig.
+\ref{fig2} A,B).
+The most parsimonious combination that provided the largest gain in predictive accuracy (i.e.
+lowest logarithmic loss) occurred when only the baseline hazard term ($\gamma$) was varied as a function of species wood density ($\rho$; Fig.
+\ref{fig2} B).
+We found that wood density was negatively correlated with species baseline mortality rate (Fig.
 \ref{fig3}).
-That is, high wood density species had a lower probability of dying irrespective of growth rate.
-For example, our model predicts that on average a species with a wood density of 0.3 has a instantaneous mortality rate of XX, whereas a species with a wood density of 0.8 has a instantaneous mortality rate of XX, translating to an annual mortality rate of XX\% and XX\%, respectively.
-Furthermore, our model also revealed that instantaneous mortality rates varied over time.
-Specifically, mortality rates at BCI decreased over time with hazard rates being scaled by \Sexpr{census_effects(final_model,1)},\Sexpr{census_effects(final_model,2)} and \Sexpr{census_effects(final_model,3)} for census periods, 1995\textendash 2000, 2000\textendash 2005 and 2005\textendash 2010, respectively.
+This meant that species with low wood density had on average higher growth-independent mortality rates ((Fig.
+\ref{fig4} A,B), and thus higher annual probability of death (Fig.
+\ref{fig4} C,D), relative to high wood density species.
+While, wood density did improve model accuracy it only accounted for XX\% of the total gain in predictive capacity that could be obtained by accounting for species differences.
 
-\begin{figure}[ht]
+\subsection*{Proportion of variation explained}
+Predicted mortality rates were most variable between species accounting for approximately XX\% followed by between census periods (accounting for XX\%).
+Wood density, a species level trait, accounted for approximately XX\% of the total species variability, suggesting that other species differences contribute to the ramaining XX\% unexplained species variation.
+
+As expected, the growth-dependent hazard terms (i.e.
+the part of the model allowed to vary by individual growth) accounted for XX\% of the overall variation in mortality rates while the baseline (growth-independent) hazard accounted for XX\%.
+
+\begin{figure}[h!]
 \includegraphics{figures/fig4.pdf}
-\caption{Predicted instantaneous mortality rates its derivative annual mortality probability.
-A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship.
-B,D) Predicted growth\textendash mortality response curves for each species.
-Errors are $\pm$ 95\% BCI.}
+\caption{Predicted changes in instantaneous mortality rates (A,B), and its derivative, annual mortality probability (C,D), with increasing growth rates.
+A,C) Estimated species curves for all 203 BCI species used in this study.
+B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species.}
 \label{fig4}
@@
 
-\subsection*{Sources of variation}
-We found that the baseline hazard, $\gamma$, varied the most between species followed by $\alpha$ and least on $\beta$.
-We also found that the wood density effect on $\gamma$ explained approximately XX\% of the species variability on the baseline hazard.
+\subsection*{Mortality rates, species light requirements and species maximum dbh}
+We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were positively correlated with gap index (our measure of species light requirement; Fig.
+\ref{fig6} A,C, E).
+By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index.
+The combination of these correlations suggest that species that predominately recruit in gaps (i.e.
+higher gap index) were more prone to death due to both low growth stochastic chance and required faster growth rates for a given mortality rate relative to a species that readily recruits in non-gaps.
+We detected no strong correlations between estimated species parameters and their associated maximum DBH (Fig.
+\ref{fig6} B,D,F,H).
 
-\begin{figure}[ht]
+\begin{figure}[h!]
 \includegraphics{figures/fig5.pdf}
-\caption{Predicted instantaneous mortality rates its derivative annual mortality probability.
-A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship.
-B,D) Predicted growth\textendash mortality response curves for each species.
-Errors are $\pm$ 95\% BCI.}
+\caption{Proportion of predicted variation explained.
+A) Variation explained by wood density (grey) as well as species and census random effects.
+B) Variation explained by growth\textendash dependent and baseline (stochastic) hazards.}
 \label{fig5}
@@
 
-\begin{figure}[ht]
+\section*{Discussion}
+
+Add something significant here.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
+Sed a magna eget lorem posuere euismod ac in enim.
+Nam porttitor placerat est, quis finibus quam efficitur nec.
+Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
+Maecenas eros sapien, semper quis velit nec, porta convallis est.
+Vestibulum porttitor diam ac massa porta pharetra.
+In a viverra justo.
+Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
+Aenean dignissim sollicitudin pharetra.
+Integer aliquet odio odio, vel bibendum mi lobortis nec.
+Aliquam non fermentum elit, id pellentesque ex.
+
+Add something significant here.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
+Sed a magna eget lorem posuere euismod ac in enim.
+Nam porttitor placerat est, quis finibus quam efficitur nec.
+Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
+Maecenas eros sapien, semper quis velit nec, porta convallis est.
+Vestibulum porttitor diam ac massa porta pharetra.
+In a viverra justo.
+Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
+Aenean dignissim sollicitudin pharetra.
+Integer aliquet odio odio, vel bibendum mi lobortis nec.
+Aliquam non fermentum elit, id pellentesque ex.
+
+Add something significant here.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
+Sed a magna eget lorem posuere euismod ac in enim.
+Nam porttitor placerat est, quis finibus quam efficitur nec.
+Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
+Maecenas eros sapien, semper quis velit nec, porta convallis est.
+Vestibulum porttitor diam ac massa porta pharetra.
+In a viverra justo.
+Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
+Aenean dignissim sollicitudin pharetra.
+Integer aliquet odio odio, vel bibendum mi lobortis nec.
+Aliquam non fermentum elit, id pellentesque ex.
+
+Add something significant here.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
+Sed a magna eget lorem posuere euismod ac in enim.
+Nam porttitor placerat est, quis finibus quam efficitur nec.
+Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
+Maecenas eros sapien, semper quis velit nec, porta convallis est.
+Vestibulum porttitor diam ac massa porta pharetra.
+In a viverra justo.
+Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
+Aenean dignissim sollicitudin pharetra.
+Integer aliquet odio odio, vel bibendum mi lobortis nec.
+Aliquam non fermentum elit, id pellentesque ex.
+
+Add something significant here.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
+Sed a magna eget lorem posuere euismod ac in enim.
+Nam porttitor placerat est, quis finibus quam efficitur nec.
+Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
+Maecenas eros sapien, semper quis velit nec, porta convallis est.
+Vestibulum porttitor diam ac massa porta pharetra.
+In a viverra justo.
+Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
+Aenean dignissim sollicitudin pharetra.
+Integer aliquet odio odio, vel bibendum mi lobortis nec.
+Aliquam non fermentum elit, id pellentesque ex.
+
+\begin{figure}[h!]
 \includegraphics{figures/fig6.pdf}
-\caption{Predicted instantaneous mortality rates its derivative annual mortality probability.
-A,C) Effect of high (0.8) and low (0.3) wood density on growth-mortality relationship.
-B,D) Predicted growth\textendash mortality response curves for each species.
-Errors are $\pm$ 95\% BCI.}
+\caption{Posthoc correlations between estimated species parameters and measures of species light requirement (A,C,E,G) and species maximum DBH (B,D,F,H).
+Points are mean ($\pm$ 95\% credible intervals) for each of the 203 species used in this study.
+Blue line shows the average trendline as determined from standard linear regression}
 \label{fig6}
@@
 
-\section*{Discussion}
+Add something significant here.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
+Sed a magna eget lorem posuere euismod ac in enim.
+Nam porttitor placerat est, quis finibus quam efficitur nec.
+Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
+Maecenas eros sapien, semper quis velit nec, porta convallis est.
+Vestibulum porttitor diam ac massa porta pharetra.
+In a viverra justo.
+Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
+Aenean dignissim sollicitudin pharetra.
+Integer aliquet odio odio, vel bibendum mi lobortis nec.
+Aliquam non fermentum elit, id pellentesque ex.
 
@@
 In order to maintain statical independence of measurements, we then sub-sampled this dataset by randomly selecting one observation per individual out of a possible three censuses.
-In total, \Sexpr{n_inds} individuals of \Sexpr{n_spp} species were included in our analysis.
+In total, \Sexpr{n_inds(BCI_model_dataset_true_unique)} individuals of \Sexpr{n_spp(BCI_model_dataset_true_unique)} species were included in our analysis.
 
@@
 Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions.
-In our dataset, \Sexpr{proportion_negative}\% of observed growth rates were negative.
+In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of observed growth rates were negative.
 While there are physiological processes (e.g.
@@
 observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors \cite{Condit:2012nz}.
-This dataset consisted of \Sexpr{n_obs_errors} randomly selected trees that had their diameter measured twice within a 30 day period \cite{Condit:2012nz}.
+This dataset consisted of \Sexpr{nobs_meas_err_data(BCI_dbh_error_data)} randomly selected trees that had their diameter measured twice within a 30 day period \cite{Condit:2012nz}.
 Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error.
-The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter.
+The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma(BCI_dbh_error_data)}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter.
 Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual (SUPP).
@@
 \subsection*{Estimating mortality rates}
+At each stage, the best model from the previous stage was taken as input for the next stage.
+Specifically, stage 1 fitted the simplest model (Fig.
+\ref{fig1}B) and compared it to the same model but one which varied by census.
+This allowed us to determine whether $\lambda$ significantly varied over-time and thus whether accounting for it improved predictive capacity.
+Stage 2 involved comparing predictive performance of the three $\lambda$ models described in (Fig.
+\ref{fig1}B) using either annual diameter increment or area growth rates.
+This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure.
+In stage 3, we begin to examined inter-specific variability by using the best model of stage 2 but allowing various combinations of parameters to vary as a function of species wood density.
+This allowed us to determine if wood density improved predictive capacity, and if so, whether it predominately improved predictive capacity via growth-dependent or growth-independent processes.
+Stage 4 also examined inter-specific variability, however, here the focus was on using the best model of stage 2 and estimating the total predictive gain obtained if all parameters were allowed to vary as a function of species ID.
+The last stage was fitting the best model, with species random effects and wood density to the entire dataset (i.e.
+not cross validated).
+This allowed us to examine the range of species hazard curves, how they are affected by wood density, census and posthoc covariates, and the proportion of predicted variation explained by species, wood density, census and growth-dependent and baseline hazard terms.
 

### 2016-07-18 R/manuscript_functions.R
extract proportion of variation wood density attributes

@@ -53,7 +53,27 @@ census_effects <- function(model,census) {
 
 # Extract proportion explained.
 extract_prop_explained <- function(data, param) {
+  if(param == "species") {
+    signif(data[data$param=='species','proportion'] + data[data$param=='wood_density','proportion'],2)*100
+  }
+  else {
   signif(data[data$param==param,'proportion'],2)*100
+  }
 }
 
-
+# Prop change in logloss.
+pred_gain_wd <- function(data) {
+  
+  df <- data %>%
+    filter(growth_measure=='true_dbh_dt' &
+             model_type %in% c('function_growth_comparison_base_growth_hazard_none',
+                               'rho_combinations_base_growth_hazard_c',
+                               'species_random_effects_base_growth_hazard_none')) %>%
+    select(comparison, mean) %>%
+    spread(comparison, mean) %>%
+    mutate(rho_diff = function_growth_comparison - rho_combinations,
+           spp_diff = function_growth_comparison - species_random_effects,
+           prop_spp_gain = rho_diff/spp_diff)
+  
+  return(signif(df$prop_spp_gain,2)*100)
+}

### 2016-07-18 ms/ms.Rnw
Draft dicussion

@@
 
+% Enable cross referencing to Evidence document
+\usepackage{xr}
+\externaldocument{ms-suppinfo}
 \title{Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework}
@@
 While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these distinct drivers are yet to be integrated within a cohesive and predictive framework.
-Here we apply classic survival analysis techniques within a cross-validated hierarchical Bayesian framework to derive instantaneous mortality functions that integrate individual, species, and plot-level effects for 203 tropical rain forest tree species at Barro Colorado Island (BCI) Panama.
+Here we apply classic survival analysis techniques within a cross-validated hierarchical Bayesian framework to derive instantaneous mortality functions that integrate individual, species, and temporal effects for 203 tropical rain forest tree species at Barro Colorado Island (BCI) Panama.
 We show that mortality models that separate growth-dependent and growth-independent components, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations.
@@
 Two substantial challenges exist when studying tree mortality in tropical rainforests.
-The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates.
+The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates (Poorter et al 2008).
 This requirement arises in part because the observable outcome -- alive vs.
@@
 This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume).
-Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack \cite{Augspurger:1984wx}, and ultimately reduced mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack \cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
 
@@
 
-In this study, we reconcile both intraspecific and interspecific drivers of instantaneous mortality by developing a unified model that incorporates individual-level, species-level, and plot-level effects, within a cross-validated hierarchical Bayesian framework.
+In this study, we reconcile both intraspecific and interspecific drivers of instantaneous mortality by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework.
 To derive this unified model, we compare the predictive accuracy of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig.
@@
 stochastic events such as wind throw and fire).
-Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (Moorcroft 2001).
+Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (Moorcroft 2001; McDowell et al 2011).
 However, these alternative models have not previously been systematically compared.
@@
 \begin{enumerate}\itemsep 0pt
-  \item Does mortality vary substantially between censuses?
   \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
@@
   \item What is the proportion of total variability explained by species, census, baseline hazard and growth-dependent hazard?
+  \item Does mortality vary substantially between censuses?
 \end{enumerate}
@@
 Tree mortality progressively decreased over time.
-Between 1995 and 2010, the highest proportion of trees dying between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and lastly 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%).
+Between 1995 and 2010, the highest proportion of trees death occured between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and lastly 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%).
 Consequently, when we allowed hazard rates to be scaled by census (i.e.
@@
 We found that the hazard function that incorporates both growth-dependent and baseline hazard terms (i.e.
-$(\alpha e^{\beta X_{i}} + \gamma)\delta$) and used stem diameter growth rather than basal-area growth, significantly outperformed both the baseline only (i.e.$\gamma \delta$) or growth-dependent only (i.e.
-$(\alpha e^{\beta X_{i}})\delta$) hazard functions (Fig.
+$(\alpha e^{\beta X_{i}} + \gamma)\delta$), significantly outperformed both the baseline only (i.e.$\gamma \delta$) or growth-dependent only (i.e.
+$(\alpha e^{\beta X_{i}})\delta$) hazard functions, irrespective of growth rate (Fig.
 \ref{fig2}A).
-Surprisingly, we found that the growth-dependent only function did not improve predictive capacity over the baseline only function.
-Furthermore, we found that while the hazard function that incorporates both baseline and growth-dependent terms significantly improved model accuracy, this was only apparent if dbh growth was used.
+However, predictive accuracy was higher for stem diameter relative to basal-area growth.
+Perhaps surprisingly, we detected no significant improvement in predictive accuracy between the baseline only and growth-dependent only hazard functions.
 
@@
 \ref{fig3}).
-This meant that species with low wood density had on average higher growth-independent mortality rates ((Fig.
+This meant that species with low wood density and fast individual growths had on average higher mortality rates ((Fig.
 \ref{fig4} A,B), and thus higher annual probability of death (Fig.
 \ref{fig4} C,D), relative to high wood density species.
-While, wood density did improve model accuracy it only accounted for XX\% of the total gain in predictive capacity that could be obtained by accounting for species differences.
+While, wood density did improve model accuracy it only accounted for \Sexpr{pred_gain_wd(logloss_summaries)}\% of the total gain in predictive capacity that could be obtained by accounting for species differences.
 
 \subsection*{Proportion of variation explained}
-Predicted mortality rates were most variable between species accounting for approximately XX\% followed by between census periods (accounting for XX\%).
-Wood density, a species level trait, accounted for approximately XX\% of the total species variability, suggesting that other species differences contribute to the ramaining XX\% unexplained species variation.
-
+Approximately \Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of variation in predicted mortality rates captured by our model was due to species differences, \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of which was captured by wood density.
+Census effects accounted for approximately \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation in mortality rates.
 As expected, the growth-dependent hazard terms (i.e.
-the part of the model allowed to vary by individual growth) accounted for XX\% of the overall variation in mortality rates while the baseline (growth-independent) hazard accounted for XX\%.
+the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the overall variation in mortality rates.
 
@@
 
-\subsection*{Mortality rates, species light requirements and species maximum dbh}
+\subsection*{Mortality rates, maximum species size and species light requirements}
 We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were positively correlated with gap index (our measure of species light requirement; Fig.
@@
 By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index.
-The combination of these correlations suggest that species that predominately recruit in gaps (i.e.
-higher gap index) were more prone to death due to both low growth stochastic chance and required faster growth rates for a given mortality rate relative to a species that readily recruits in non-gaps.
+The combination of these correlations suggest that species which predominately recruit in gaps (i.e.
+higher gap index) were more prone to death due to both low growth and stochastic chance.
+They also required faster growth rates to achieve a given mortality rate relative to a species that recruits in non-gaps.
 We detected no strong correlations between estimated species parameters and their associated maximum DBH (Fig.
@@
 \caption{Proportion of predicted variation explained.
-A) Variation explained by wood density (grey) as well as species and census random effects.
-B) Variation explained by growth\textendash dependent and baseline (stochastic) hazards.}
+A) Predicted variation explained by wood density (grey) as well as species and census random effects.
+B) Predicted variation explained by growth\textendash dependent and baseline (stochastic) hazards.}
 \label{fig5}
@@
 \section*{Discussion}
+Our Bayesian framework coupled with cross validation revealed that the most accurate and parsimonious model of tropical tree death was that which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) attributed the effect of wood density to growth-independent mortality; and 4) incorporated temporal varability.
+Moreover, we found that rates of tropical tree mortality varied substantially between species and that wood density, a species level functional trait, only explained a small proportion of the overall inter-specific variation.
 
-Add something significant here.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
-Sed a magna eget lorem posuere euismod ac in enim.
-Nam porttitor placerat est, quis finibus quam efficitur nec.
-Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
-Maecenas eros sapien, semper quis velit nec, porta convallis est.
-Vestibulum porttitor diam ac massa porta pharetra.
-In a viverra justo.
-Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
-Aenean dignissim sollicitudin pharetra.
-Integer aliquet odio odio, vel bibendum mi lobortis nec.
-Aliquam non fermentum elit, id pellentesque ex.
+The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards (Moorcroft, Sortie,McDowell 2011).
+We show that regardless of growth measure, incorporating both hazards significantly improves model predictive accuracy.
+This is because the additive effects of both hazards explicitly allow models to account for two independent causes of plant mortality.
+The growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stressors (e.g.
+competition, parasitization, localized herbivory).
+By contrast, the growth-independent hazard accounts for deaths caused by stochastic events that arise irrespective of an individuals growth rate (e.g.
+windthrow).
 
-Add something significant here.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
-Sed a magna eget lorem posuere euismod ac in enim.
-Nam porttitor placerat est, quis finibus quam efficitur nec.
-Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
-Maecenas eros sapien, semper quis velit nec, porta convallis est.
-Vestibulum porttitor diam ac massa porta pharetra.
-In a viverra justo.
-Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
-Aenean dignissim sollicitudin pharetra.
-Integer aliquet odio odio, vel bibendum mi lobortis nec.
-Aliquam non fermentum elit, id pellentesque ex.
+Another advantage of partitioning mortality into growth-depedent and growth-independent hazards was that we could estimate the proportion of variation each attributed.
+Like many other studies (Denslow JS (1987) King DA, Davies SJ, Supardi MNN, Tan S (2005), Ruger 2011), our analyses highlight the importance of light competition as a dominant factor influence tropical tree demographic rates.
+Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted varability in mortality rates (Fig 5), and thus, suggests that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.
 
-Add something significant here.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
-Sed a magna eget lorem posuere euismod ac in enim.
-Nam porttitor placerat est, quis finibus quam efficitur nec.
-Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
-Maecenas eros sapien, semper quis velit nec, porta convallis est.
-Vestibulum porttitor diam ac massa porta pharetra.
-In a viverra justo.
-Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
-Aenean dignissim sollicitudin pharetra.
-Integer aliquet odio odio, vel bibendum mi lobortis nec.
-Aliquam non fermentum elit, id pellentesque ex.
+Incorporating the effect of wood density on mortality rates also improved predictive performance.
+Our analyses revealed that the most parsimonious combination of wood density effects was when it was attributed to only growth-independent mortality.
+Specifically, we found that high wood density species had lower growth-independent mortality rates relative to their low wood density counterparts.
+This finding corroborating the observed negative correlation between mortality and wood density reported elsewhere \cite{Kraft:2010kq,Chave:2009iy}.
+More importantly, however, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}.
 
-Add something significant here.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
-Sed a magna eget lorem posuere euismod ac in enim.
-Nam porttitor placerat est, quis finibus quam efficitur nec.
-Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
-Maecenas eros sapien, semper quis velit nec, porta convallis est.
-Vestibulum porttitor diam ac massa porta pharetra.
-In a viverra justo.
-Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
-Aenean dignissim sollicitudin pharetra.
-Integer aliquet odio odio, vel bibendum mi lobortis nec.
-Aliquam non fermentum elit, id pellentesque ex.
-
-Add something significant here.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
-Sed a magna eget lorem posuere euismod ac in enim.
-Nam porttitor placerat est, quis finibus quam efficitur nec.
-Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
-Maecenas eros sapien, semper quis velit nec, porta convallis est.
-Vestibulum porttitor diam ac massa porta pharetra.
-In a viverra justo.
-Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
-Aenean dignissim sollicitudin pharetra.
-Integer aliquet odio odio, vel bibendum mi lobortis nec.
-Aliquam non fermentum elit, id pellentesque ex.
+While wood density effects are now being incorporated in mortality algorithms of many vegetation models (REFS), our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation.
+This means that such models are likely to severely underestimate the true varability in mortality rates.
+Consequently, because variation in mortality rates are known to have significant impacts on forest dynamics \cite{Johnson-2016}, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics, with flow on effects for ecosystem proporties such as GPP and NPP.
 
@@
 Points are mean ($\pm$ 95\% credible intervals) for each of the 203 species used in this study.
-Blue line shows the average trendline as determined from standard linear regression}
+Blue line shows the average trendline as determined from standard linear regression.
+Gap index is a binary species level index of the light required for recruitment.
+Here low numbers indicate the species readily recruits under dense canopy and 1 indicates that a species only ever recruits in gaps.}
 \label{fig6}
@@
 
-Add something significant here.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
-Sed a magna eget lorem posuere euismod ac in enim.
-Nam porttitor placerat est, quis finibus quam efficitur nec.
-Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
-Maecenas eros sapien, semper quis velit nec, porta convallis est.
-Vestibulum porttitor diam ac massa porta pharetra.
-In a viverra justo.
-Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
-Aenean dignissim sollicitudin pharetra.
-Integer aliquet odio odio, vel bibendum mi lobortis nec.
-Aliquam non fermentum elit, id pellentesque ex.
+The underestimation of inter-specific variation suggests that there are other species attributes besides wood density that affect tree mortality rates.
+Poorter et al.
+(2006) suggested that a species light demand and maximum size (a correlate to species longevity; Lieberman et al 1985) are two major axes of inter-specific variation in tropical rainforests.
+While we did not explicitly model mortality as a function of these traits, our posthoc correlations (FIG) revealed that some inter-specific variation was indeed correlated with light demand, but not maximum size.
+We found that light demanding species (i.e.
+those with high gap index) were more susceptible to death due to both low growth and stochastic chance and required faster growth rates to achieve a given mortality rate relative to a species that readily recruits in shaded conditions.
 
-- General overview of results:
-As expected, the most biologically reasonable model; the model with both a baseline hazard and growth-dependent hazard was the best predictor of tropical tree mortality, indicating that both growth dependent and independent hazards are needed to effectively model tree mortality.
+We recommend research should focus on examining how other traits influence growth-dependent and growth-independent hazards and what combination of traits minimises unexplained species variability.
+Researchers, however, should be mindful that the inclusion of additional traits comes at the cost of added complexity, increased uncertainty (i.e.
+more parameters to estimate) and increased minimum knowledge; all of which diminish the interpretability of models, but also the applicability of predicting mortality rates in many data poor ecosystems.
+Because of this, we also recommend future research focus on developing more integrative correlates of plant mortality.
+For example, we may have captured more inter-specific variability in baseline hazards had we used a integrative measure of plant durability such as a trait that incorporates both wood density and stem architecture (REF).
 
-- Furthermore, we found that DBH was by far the better growth measure for predicting plant mortality.
-	- Why DBH?
-Could it be that DBH implicitly punishes smaller individuals relative to larger individuals.
-That is, small plants can only gain a small amount of DBH growth relative to a larger plant.
-As such, DBH growth implicitly captures the process whereby small individual plants are more prone to competitive effects as well as random breakage events caused by trampling, herbivory etc.
+Future research should also focus on examining how climate variables influence both growth-depedent and growth-independent hazards, particularly as climate-driven mortality is increasing globally (McDowell 2011, Get a Zepple paper).
+Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010.
+Unfortuantely, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3).
+However, based on anedoctal evidence, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by a 1998 El-Nino (REF).
+Furthermore, future research should also examine how different growth measures influence predictive accuracy of mortality models.
+We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive accuracy, and we are unsure how other common measures such as biomass growth (Chave's biomass paper) will compare.
 
-- As expected, when allowing parameters to vary by species we found a significant improvement in model predictive capacity.
-		- This supports previous research that shows that species vary substantially in mortality rates.
-
-- When examining this interspecific variability we found that the effect of wood density improved model predictive capacity when it was included on the base-line hazard, with only marginal improvements when included on all three parameters.
-
-- Furthermore, we found that wood density explained XX\% of the total species variation for the baseline hazard.
-
-- Conclusion, species with higher wood density had lower baseline hazards supporting previous work that suggests that these species are less prone to stochastic events that involve breakage etc.
+Here we showcase a new framework for modelling tropical tree mortality that attempts to unifies empirical evidence from within and between species studies.
+This framework also provides an approach for partitioning mortality rates into growth-dependent and growth-independent hazards.
+Our findings reveal that while wood density is an important trait affecting mortality rates, we are still only capturing a fraction of the overall species variability in mortality rates.
 
@@
 
-The most complicated function is
-\begin{equation}
-\lambda_i = \big(\underbrace{\alpha \, \exp\left(-\beta \, X_i \right)}_{\text{Growth-dependent}} +  \underbrace{\gamma}_{\text{Baseline}}\big) \underbrace{\delta_t}_{\text{Census-effect}}.
- \label{add_haz}
-\end{equation}
-
 \subsection*{Model fitting}

### 2016-07-19 ms/ms.Rnw
Added references

@@
 This variation has considerable consequences for forest structure and dynamics.
-Death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools (Chave et al 200X), while at the same time opening up a gap that restarts a successional race, during which 10's-100's more plants may die while competing for a spot in the sun.
-In models of forest dynamics, variation in mortality rates has been found to have a larger impact on forest structure than absolute growth rates \cite{Johnson-2016}.
+Death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools\cite{Chave:2005hz}, while at the same time opening up a gap that restarts a successional race, during which 10's-100's more plants may die while competing for a spot in the sun.
+In models of forest dynamics, variation in mortality rates has been found to have a larger impact on forest structure than absolute growth rates\cite{Johnson-2016}.
 As such, improving our ability to understand the mortality process is therefore a priority for making skilful predictions about forest dynamics and carbon, water or nutrient cycles; especially in an era of rapid environmental change.
@@
 \end{equation}
-Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$ \cite{Kleinbaum:2012tn}.
+Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$\cite{Kleinbaum:2012tn}.
 The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
@@
 Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required.
-As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys in long-term monitoring plots containining thousands of individuals and spanning several decades \cite{Anderson-Teixeira-2015a}.
+As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys in long-term monitoring plots containining thousands of individuals and spanning several decades\cite{Anderson-Teixeira-2015a}.
 
 The other major challenge when studying mortality -- similar to that faced in other fields like medicine and engineering -- is to derive suitable shapes for the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality.
-Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate, traits, and size (XX).
-Generally, however, these influences have not been formalised as hazard functions (XX).
-Yet such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function (XX).
+Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate\cite{Kobe:1995tw,Wyckoff:2002ul}, traits\cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size\cite{Vieilledent:2010fv, Hurst:2011ky}.
+Generally, however, these influences have not been formalised as hazard functions\cite{Zens:2003jv}.
+Yet such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function\cite{Hawkes:2000ib;Zens:2003jv,McDowell:2011dr}.
 This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.
@@
 A particular challenge is to integrate the observed conflicting relationships between mortality and growth observed within and across species.
-Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
-Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth \cite[e.g.][]{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
-Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality than individuals from slower growing species \cite{Poorter:2008iu, Wright:2010fl}.
+Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
+Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
+Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality than individuals from slower growing species\cite{Poorter:2008iu, Wright:2010fl}.
 This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume).
-Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack \cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack\cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events\cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
 
@@
 
-The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards (Moorcroft, Sortie,McDowell 2011).
+The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards\cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}.
 We show that regardless of growth measure, incorporating both hazards significantly improves model predictive accuracy.
@@
 Another advantage of partitioning mortality into growth-depedent and growth-independent hazards was that we could estimate the proportion of variation each attributed.
-Like many other studies (Denslow JS (1987) King DA, Davies SJ, Supardi MNN, Tan S (2005), Ruger 2011), our analyses highlight the importance of light competition as a dominant factor influence tropical tree demographic rates.
-Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted varability in mortality rates (Fig 5), and thus, suggests that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.
+Like many other studies\cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlight the importance of light competition an important factor influencing tropical tree demographic rates.
+Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted varability in mortality rates (Fig.
+\ref{fig5}), and thus, suggests that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.
 
@@
 Specifically, we found that high wood density species had lower growth-independent mortality rates relative to their low wood density counterparts.
-This finding corroborating the observed negative correlation between mortality and wood density reported elsewhere \cite{Kraft:2010kq,Chave:2009iy}.
-More importantly, however, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall \cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}.
+This finding corroborating the observed negative correlation between mortality and wood density reported elsewhere\cite{Kraft:2010kq,Chave:2009iy}.
+More importantly, however, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}.
 
-While wood density effects are now being incorporated in mortality algorithms of many vegetation models (REFS), our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation.
+While wood density effects are now being incorporated in mortality algorithms of many vegetation models\cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation.
 This means that such models are likely to severely underestimate the true varability in mortality rates.
-Consequently, because variation in mortality rates are known to have significant impacts on forest dynamics \cite{Johnson-2016}, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics, with flow on effects for ecosystem proporties such as GPP and NPP.
+Consequently, because variation in mortality rates are known to have significant impacts on forest dynamics\cite{Johnson-2016}, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics, with flow on effects for ecosystem proporties such as GPP and NPP.
 
@@
 The underestimation of inter-specific variation suggests that there are other species attributes besides wood density that affect tree mortality rates.
-Poorter et al.
-(2006) suggested that a species light demand and maximum size (a correlate to species longevity; Lieberman et al 1985) are two major axes of inter-specific variation in tropical rainforests.
-While we did not explicitly model mortality as a function of these traits, our posthoc correlations (FIG) revealed that some inter-specific variation was indeed correlated with light demand, but not maximum size.
-We found that light demanding species (i.e.
-those with high gap index) were more susceptible to death due to both low growth and stochastic chance and required faster growth rates to achieve a given mortality rate relative to a species that readily recruits in shaded conditions.
+It has been suggested that a species light demand and maximum size are two major axes of inter-specific variation in tropical rainforests\cite{Poorter:2006tq}.
+While we did not explicitly model mortality as a function of these traits, our posthoc correlations  \ref{fig6} revealed that some inter-specific variation was indeed correlated with light demand.
+Specifically, we found that light demanding species (i.e.
+those with high gap index) were more susceptible to death due to both low growth and stochastic chance and required faster growth rates to achieve a given mortality rate relative to a species that readily recruits in shaded conditions, supporting findings by others\cite{Ruger:2011cv}.
+By contrast, we detected no correlations between a species maximum stem diameter and mortality rates, supporting findings by others\cite{Wright:2010fl}.
 
@@
 Because of this, we also recommend future research focus on developing more integrative correlates of plant mortality.
-For example, we may have captured more inter-specific variability in baseline hazards had we used a integrative measure of plant durability such as a trait that incorporates both wood density and stem architecture (REF).
+For example, we may have captured more inter-specific variability in baseline hazards had we used a integrative measure of plant durability such as a trait that incorporates both wood density and stem architecture.
 
-Future research should also focus on examining how climate variables influence both growth-depedent and growth-independent hazards, particularly as climate-driven mortality is increasing globally (McDowell 2011, Get a Zepple paper).
+Future research should also focus on examining how climate variables influence both growth-depedent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}.
 Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010.
 Unfortuantely, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3).
-However, based on anedoctal evidence, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by a 1998 El-Nino (REF).
-Furthermore, future research should also examine how different growth measures influence predictive accuracy of mortality models.
-We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive accuracy, and we are unsure how other common measures such as biomass growth (Chave's biomass paper) will compare.
+However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by a El-Nino in 1997\textendash 1998.
+Droughts are known to have significant impacts on tree mortality rates in these ecosystems\cite{Condit:1995wl, Mcdowell:2013wf} Furthermore, future research should also examine how different growth measures influence predictive accuracy of mortality models.
+We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive accuracy, and we are unsure how other common measures such as biomass growth\cite{Chave:2005hz} will compare.
 
@@
 \subsection*{Data}
-We parametrized plant mortality models using individual growth and survival data \cite{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\textdegree;N, 79.85\textdegree;W).
+We parametrized plant mortality models using individual growth and survival data\cite{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\textdegree;N, 79.85\textdegree;W).
 The climate on the island is warm throughout the year with a mean annual temperature of 26.1\textdegree and rainfall is seasonal with most of the 2500 mm falling between April and November.
 BCI is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI) and the permanent plot was established in 1980.
-Detailed descriptions on the flora, fauna, geology and climate of BCI can be found in (REF).
 
-Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter \cite{Condit:2012nz}.
-For the purpose of modelling mortality as a function of growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
+Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter\cite{Condit:2012nz}.
+For the purpose of modelling mortality as a function of past growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
+Consequently, we modelled tree mortality as a function of past growth for censuses 1995\textendash 2000, 2000\textendash 2005 and 2005\textendash 2010.
 We also excluded individuals based on the following rules: 1) those that did not survive at least two censuses (two being required to estimate growth rate); 2) those that were not consistently measured at 1.3 m above ground; 3) those that were multi-stemmed; 4) those that resprouted or \textit{``returned from the dead''}; or 5) extreme outliers -- stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial diameter.
@@
 
-Wood density for each species was estimated by coring trees located within 15 km of the BCI plot, because increment borers are prohibited within the plot itself \cite{Wright:2010fl}.
+Wood density for each species was estimated by coring trees located within 15 km of the BCI plot, because increment borers are prohibited within the plot itself\cite{Wright:2010fl}.
 Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100\textdegree C) and dividing by the fresh volume (as measured by water displacement).
 Wood density was then calculated as an area-weighted average (g cm$^-3$), where area refers to the annulus represented by each piece, assuming a circular trunk.
+
+In order to calculate our gap index (our measure of a species light dependence) we also used annual canopy census data collected during 1985\textendash 1990 and 1990\textendash 1995.
+The canopy census recorded the presence of vegetation in six height intervals, 0–2, 2–5, 5–10, 10–20, 20–30 and 30 m every 5-m across the 50 ha.
+We calculated the minimum observed number of vegetation strata above 2 m all 5x5 cells in each census period.
+We then transformed this to a binary gap index by where 1 refers to no vegetation above 2 m and 0 referring to densely vegetated by subtracting the above values from 5 (i.e., 0 = dense, 5 = gap).
+To acknowledge the influence of gaps on surrounding cells' light availability, we then calculated a weighted focal sum using a 3 x 3 cell moving window, for which the central cell was assigned a weight of 8, and the eight neighbouring cells were each assigned weights of 1.
+The combined impact of vegetation density of neighbouring cells on light availability at the central cell was therefore assumed to be equivalent to the impact of vegetation density at the central cell.
 
@@
 While there are physiological processes (e.g.
-water depletion and bark drop) that can lead to trunk shrinkages overtime \cite{Pastur:2007kl}, it is biologically impossible for growth (i.e.
-the creation of new biomass) to be negative \cite{Clark:2007ve}.
+water depletion and bark drop) that can lead to trunk shrinkages overtime\cite{Pastur:2007kl}, it is biologically impossible for growth (i.e.
+the creation of new biomass) to be negative\cite{Clark:2007ve}.
 To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e.
-observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors \cite{Condit:2012nz}.
-This dataset consisted of \Sexpr{nobs_meas_err_data(BCI_dbh_error_data)} randomly selected trees that had their diameter measured twice within a 30 day period \cite{Condit:2012nz}.
+observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors\cite{Condit:2012nz}.
+This dataset consisted of \Sexpr{nobs_meas_err_data(BCI_dbh_error_data)} randomly selected trees that had their diameter measured twice within a 30 day period\cite{Condit:2012nz}.
 Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error.
@@
 We executed three independent chains and in all cases modelled parameters converged within 2000 iterations.
-Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Gelman:1992ts, Brooks:1998ju}.
+Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic\cite{Gelman:1992ts, Brooks:1998ju}.
 After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.

### 2016-07-19 ms/supp.Rnw
Format for supplementary materials
- layout for title etc
- add figures
- add references to figure in main text

--- 2016-07-04 ms/supp.Rnw
@@
-\documentclass[9pt,twocolumn,twoside]{pnas-new}
+\documentclass[9pt,onecolumn,twoside]{pnas-new}
 % Use the lineno option to display guide line numbers if required.
@@
 
-\templatetype{pnasresearcharticle} % Choose template 
-% {pnasresearcharticle} = Template for a two-column research article
-% {pnasmathematics} = Template for a one-column mathematics article
-% {pnasinvited} = Template for a PNAS invited submission
+\templatetype{pnasresearcharticle}
 
-\title{Unifying intraspecific and interspecific variation in tropical tree mortality}
+\title{
+{\normalfont \fontsize{12pt}{16pt}
+{\color{black50} Supporting information for:}\\} \vskip6pt
+\fontsize{16pt}{20pt} Unifying intraspecific and interspecific variation in tropical tree mortality
+}
 
@@
 
-% Please give the surname of the lead author for the running footer
-\leadauthor{Camac}
+\dates{This supplement was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
 
-% Please add here a significance statement to explain the relevance of your work
-\significancestatement{Add something significant here}
+\hypersetup{
+     linkcolor=pnasbluetext
+}
 
-% Please include corresponding author, author contribution and author declaration information
-\authorcontributions{J.C.
-designed and performed the analysis; D.F.
-conceived the idea, supervized the development of this work; S.J.W \& M.W.
-provided biological insight.
-All authors contributed to the writing of this manuscript.}
-
-\authordeclaration{The authors declare that they have no competing financial interests.}
-\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: james.camac@gmail.com}
-
-% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
-\keywords{Tropical trees $|$ wood density $|$ growth $|$ mortality $|$ interspecific}
+\RequirePackage[colorlinks=true, allcolors=blue]{hyperref}
 
 \begin{document}
-% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
-% You should only change this length when you've finalized the article contents.
-\verticaladjustment{-2pt}
 
 \maketitle
-\thispagestyle{firststyle}
-\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
 
-\subsection*{Modelling true growth}
+\tableofcontents
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{secnumdepth}{0}
+
+\section{Appendix A - Supplementary text}
+
+Example reference \cite{Johnson-2016}.
+
+<<preload, echo=FALSE>>=
+n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
+n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
+proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
+n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
+obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
+n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))
+@
+
+\subsection{Modelling true growth}
 
@@
 
-\subsection*{Mortality model}
-\subsection*{Model fitting}
+\subsection{Mortality model}
 Our observations are for survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$.
@@
 \end{equation}
+
+\clearpage
+\section{Appendix 2 - Example stan code}
+
+Include example model here
+
+Describe how to get other models?
+
+\clearpage
+\section{Supplementary Figures}
+
+\begin{figure}[!hb]
+\centering
+\includegraphics[width=15cm,height=20cm,keepaspectratio]{figures/supp_fig1}
+\caption{ About this figure}
+\label{figS1}
+\end{figure}
+
+\clearpage
+\begin{figure}[!hb]
+\centering
+\includegraphics[width=15cm,height=20cm,keepaspectratio]{figures/supp_fig2}
+\caption{ About this figure}
+\label{figS2}
+\end{figure}
+
+\clearpage
+
+\section{References}
+
+\bibliography{refs.bib}
+\end{document}

### 2016-10-19 ms/ms.Rnw
Format for supplementary materials
- layout for title etc
- add figures
- add references to figure in main text
Edits to manuscript and figure code

@@
 \usepackage{xr}
-\externaldocument{ms-suppinfo}
+\externaldocument{supp}
+
 \title{Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework}
@@
 % 120 words
-\significancestatement{Add something significant here.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex.
-Sed a magna eget lorem posuere euismod ac in enim.
-Nam porttitor placerat est, quis finibus quam efficitur nec.
-Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem.
-Maecenas eros sapien, semper quis velit nec, porta convallis est.
-Vestibulum porttitor diam ac massa porta pharetra.
-In a viverra justo.
-Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus.
-Aenean dignissim sollicitudin pharetra.
-Integer aliquet odio odio, vel bibendum mi lobortis nec.
-Aliquam non fermentum elit, id pellentesque ex.}
+\significancestatement{Tree mortality is a fundamental demographic process which affects forest dynamics and consequently, a terrestrial ecosystems ability to store carbon.
+Here, for the first time, we use observed survival records over a 15 year period from more than 180,000 individuals, to simultaneously estimate growth-dependent and growth-independent (i.e.
+death by stocastic chance) mortality across 203 tropical forest tree species.
+We found that growth-dependent mortality was the predominate factor influencing tree mortality rates at Barro Colorado Island (BCI).
+Furthermore, we found that wood density most strongly influenced mortality rates by decreasing growth-independent mortality.
+However, wood density only accounted for a small fraction of the overall species variability observed in growth-independent mortality rates, suggesting that there must be other species traits that strongly affect mortality.
+Which traits and how many of them, however, remains unknown.}
 
@@
 Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities.
-While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these distinct drivers are yet to be integrated within a cohesive and predictive framework.
+While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits and temporal plot-scale effects -- these distinct drivers are yet to be integrated within a cohesive and predictive framework.
 Here we apply classic survival analysis techniques within a cross-validated hierarchical Bayesian framework to derive instantaneous mortality functions that integrate individual, species, and temporal effects for 203 tropical rain forest tree species at Barro Colorado Island (BCI) Panama.
-We show that mortality models that separate growth-dependent and growth-independent components, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations.
+We show that mortality models that partition mortality into growth-dependent and growth-independent components, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations.
 Moreover, we show that the effects of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability in species mortality rates, and that temporal variation in mortality was large.
 Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI.
-Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
+Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
 \end{abstract}
@@
 This variation has considerable consequences for forest structure and dynamics.
-Death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools\cite{Chave:2005hz}, while at the same time opening up a gap that restarts a successional race, during which 10's-100's more plants may die while competing for a spot in the sun.
-In models of forest dynamics, variation in mortality rates has been found to have a larger impact on forest structure than absolute growth rates\cite{Johnson-2016}.
-As such, improving our ability to understand the mortality process is therefore a priority for making skilful predictions about forest dynamics and carbon, water or nutrient cycles; especially in an era of rapid environmental change.
+For example, death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools\cite{Chave:2005hz}.
+Furthermore it creates a gap in the canopy that can restart a successional race, during which 10's-100's more plants may die while competing for a spot in the sun.
+In models of forest dynamics, variation in mortality rates have been found to have a larger impact on forest structure than variation in absolute growth rates\cite{Johnson-2016}.
+As such, improving our ability to understand the mortality process is a priority for making accruate predictions about forest and carbon dynamics as well as water or nutrient cycles; especially in an era of rapid environmental change.
 
-Two substantial challenges exist when studying tree mortality in tropical rainforests.
-The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates (Poorter et al 2008).
+Two difficulties arise when studying tree mortality in tropical rainforests.
+The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates\cite{Poorter:2008iu}.
 This requirement arises in part because the observable outcome -- alive vs.
 dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate).
-In classic survival analysis \cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
+In classic survival analysis\cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
 \begin{equation} \label{eq:mortality_model}
@@
 
-The other major challenge when studying mortality -- similar to that faced in other fields like medicine and engineering -- is to derive suitable shapes for the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality.
+The other difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality.
 Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate\cite{Kobe:1995tw,Wyckoff:2002ul}, traits\cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size\cite{Vieilledent:2010fv, Hurst:2011ky}.
 Generally, however, these influences have not been formalised as hazard functions\cite{Zens:2003jv}.
-Yet such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function\cite{Hawkes:2000ib;Zens:2003jv,McDowell:2011dr}.
+Yet, such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function\cite{Hawkes:2000ib;Zens:2003jv,McDowell:2011dr}.
 This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.
 
-A particular challenge is to integrate the observed conflicting relationships between mortality and growth observed within and across species.
+A particular challenge in developing hazard functions for plants is to integrate the observed conflicting relationships between mortality and growth within and across species.
 Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
-Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
-Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality than individuals from slower growing species\cite{Poorter:2008iu, Wright:2010fl}.
+Empirical studies provide broad support of this theory, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
+By contrast, across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality rates than individuals from slower growing species\cite{Poorter:2008iu, Wright:2010fl}.
 This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume).
 Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack\cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events\cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+
+Another major challenge is using binary observations (e.g.
+alive = 0; dead = 1) to infer the contribution of growth-dependent and growth-independent (i.e.
+stochastic events) hazards on tree mortality rates.
+By partitioning mortality rates into these two processes model predictive capacity may be improved and it allows us to narrow in on whether growth-dependent or growth-independent processes are the predominate cause of death in an ecosystem.
+For example, are tree deaths in rainforest predominately a result of competitive influences (i.e.
+growth-dependent processes) or are they more likely to result from growth-independent events such as treefall?
+More importantly, however, such partitions will allow us to examine how plant traits affect each hazard.
+For example, higher wood density is thought to convey higher survival by decreasing mortality aociated with stoachastic events such as those listed above.
+However, research has also shown that wood density may increase tolerance to low growth rates, and thus alter the growth-dependent mortality.
+For example, studies have found that species with high wood density are more tolerant to both shade \cite{vanGelder:2006exa}, neighbor competition \cite{Kunstler:2016km} and drought induced carbon starvation \cite{McDowell:2011dr}.
+Consequently, wood density may affect a trees response to both growth-dependent and growth-independent hazards.
+The question is, which hazard does it affect most?
 
@@
 
-In this study, we reconcile both intraspecific and interspecific drivers of instantaneous mortality by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework.
+In this study, we reconcile both intraspecific and interspecific drivers and attempt to partition instantaneous mortality rates into growth-dependent and growth-independent hazards by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework.
 To derive this unified model, we compare the predictive accuracy of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig.
@@
 \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases.
-
 Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached.
@@
   \item What is the proportion of total variability explained by species, census, baseline hazard and growth-dependent hazard?
-  \item Does mortality vary substantially between censuses?
+  \item Does mortality rates vary substantially between censuses?
 \end{enumerate}
@@
 
-\subsection*{Mortality strongly varies over time}
+\subsection*{Mortality over time}
 Tree mortality progressively decreased over time.
@@
 
+Example referencing suppmat figure (Fig. \ref{figS1}).
+
 \matmethods{
@@
 
-In order to calculate our gap index (our measure of a species light dependence) we also used annual canopy census data collected during 1985\textendash 1990 and 1990\textendash 1995.
-The canopy census recorded the presence of vegetation in six height intervals, 0–2, 2–5, 5–10, 10–20, 20–30 and 30 m every 5-m across the 50 ha.
-We calculated the minimum observed number of vegetation strata above 2 m all 5x5 cells in each census period.
-We then transformed this to a binary gap index by where 1 refers to no vegetation above 2 m and 0 referring to densely vegetated by subtracting the above values from 5 (i.e., 0 = dense, 5 = gap).
-To acknowledge the influence of gaps on surrounding cells' light availability, we then calculated a weighted focal sum using a 3 x 3 cell moving window, for which the central cell was assigned a weight of 8, and the eight neighbouring cells were each assigned weights of 1.
-The combined impact of vegetation density of neighbouring cells on light availability at the central cell was therefore assumed to be equivalent to the impact of vegetation density at the central cell.
-
 \subsection*{Estimating true growth}
@@
 the creation of new biomass) to be negative\cite{Clark:2007ve}.
-To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e.
-observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors\cite{Condit:2012nz}.
-This dataset consisted of \Sexpr{nobs_meas_err_data(BCI_dbh_error_data)} randomly selected trees that had their diameter measured twice within a 30 day period\cite{Condit:2012nz}.
-Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error.
-The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma(BCI_dbh_error_data)}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter.
-Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual (SUPP).
+To ensure our mortality model was not biased by these implausible values we first developed a model to estimate \textit{``true growth''}(i.e.
+observed growth minus observation error).
+Details of this model can be found in supplementary material (REF, FigS1).
 
@@
 After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.
+These models were then cross validated using 10-fold cross validation (REF).
 
 \subsection*{Model evaluation}
-Predictive accuracy was quantified by estimating the average log loss across K-folds for held-out data, $\bar{\mathcal{L}}$ (Fig.
+Predictive accuracy was quantified by estimating the average log loss across 10-folds for held-out data, $\bar{\mathcal{L}}$ (Fig.
 \ref{fig1}).
-Logarithmic loss -- commonly known as log loss, and indicated by $\mathcal{L}$-- measures the accuracy of a model by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome (Fig ??).
+Logarithmic loss -- commonly known as log loss, and indicated by $\mathcal{L}$-- measures the accuracy of a model by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome (Fig.
+\ref{fig1}C).
 By minimizing the $\bar{\mathcal{L}}$, we maximize predictive skill.
@@
 where $N_k$ refers to the number of observations in fold.
+
+\subsection*{Posthoc correlations}
+
+In order to calculate our gap index (our measure of a species light dependence) we also used annual canopy census data collected during 1985\textendash 1990 and 1990\textendash 1995.
+The canopy census recorded the presence of vegetation in six height intervals, 0–2, 2–5, 5–10, 10–20, 20–30 and 30 m every 5-m across the 50 ha.
+We calculated the minimum observed number of vegetation strata above 2 m all 5x5 cells in each census period.
+We then transformed this to a binary gap index by where 1 refers to no vegetation above 2 m and 0 referring to densely vegetated by subtracting the above values from 5 (i.e., 0 = dense, 5 = gap).
+To acknowledge the influence of gaps on surrounding cells' light availability, we then calculated a weighted focal sum using a 3 x 3 cell moving window, for which the central cell was assigned a weight of 8, and the eight neighbouring cells were each assigned weights of 1.
+The combined impact of vegetation density of neighbouring cells on light availability at the central cell was therefore assumed to be equivalent to the impact of vegetation density at the central cell.
 
@@
 was supported by an ARC discovery grant.
-We thank Lachlan McCalman and Daniel Steinberg for statistical advice; Anias Gibert etc etc.
+We thank Rich FitzJohn, Lachlan McCalman and Daniel Steinberg for computational advice; Anias Gibert etc etc.
 Fig.

### 2016-10-19 ms/supp.Rnw
Edits to manuscript and figure code

@@
 
-<<preload, echo=FALSE>>=
-n_inds <- as.character(round(nrow(BCI_model_dataset_true_unique),-2))
-n_spp <- length(unique(BCI_model_dataset_true_unique$sp))
-proportion_negative <- round(nrow(subset(BCI_model_dataset_true_unique, obs_dbh_dt<0))/nrow(BCI_model_dataset_true_unique)*100)
-n_obs_errors <- nrow(BCI_dbh_error_data$discrep)
-obs_error_sigma <- round(sd(BCI_dbh_error_data$discrep),2)
-n_heldout <- as.character(nrow(BCI_training_sets[[1]]$heldout))
-@
-
 \subsection{Modelling true growth}
 
-We did this by first modelling the observed initial ($D_{i,1}$) and final diameter ($D_{i,2}$) of each individual, $i$, as random realisations from a normal distribution centerd on their respective true estimates (i.e.
-$\widehat{D_{i,1}}$ and $\widehat{D_{i,2}}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e.
+Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions.
+In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of observed growth rates were negative.
+While there are physiological processes (e.g.
+water depletion and bark drop) that can lead to trunk shrinkages overtime\cite{Pastur:2007kl}, it is biologically impossible for growth (i.e.
+the creation of new biomass) to be negative\cite{Clark:2007ve}.
+To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e.
+observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors\cite{Condit:2012nz}.
+This dataset consisted of \Sexpr{nobs_meas_err_data(BCI_dbh_error_data)} randomly selected trees that had their diameter measured twice within a 30 day period\cite{Condit:2012nz}.
+Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error.
+The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma(BCI_dbh_error_data)}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter.
+Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual.
+
+We modelled true growth by first estimating the observed initial ($D_{i,1}$) and final diameter ($D_{i,2}$) of each individual, $i$, as random realisations from a normal distribution centerd on their respective true estimates (i.e.
+$\widehat{D_{i,1}}$ and $\widehat{D_{i,2}}$), with the standard deviation fixed at \Sexpr{obs_error_sigma(BCI_dbh_error_data)} (i.e.
 the observed measurement error):
@@
 \subsection{Mortality model}
-Our observations are for survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$.
+We modelled tree mortality using survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$.
 We model these observations as random realizations from a Bernoulli distribution
@@
 \clearpage
-\section{Appendix 2 - Example stan code}
+\section{Appendix 2 - Stan code for final model fit with species random effects}
 
-Include example model here
-
-Describe how to get other models?
+<<echo=FALSE>>=
+fullmodelcode()
+@
 
@@
 \centering
-\includegraphics[width=15cm,height=20cm,keepaspectratio]{figures/supp_fig1}
-\caption{ About this figure}
+\includegraphics{figures/supp_fig1.png}
+\caption{Observed vs predicted individual growth rates and stem diameters}
 \label{figS1}
@@
 \centering
-\includegraphics[width=15cm,height=20cm,keepaspectratio]{figures/supp_fig2}
-\caption{ About this figure}
+\includegraphics{figures/supp_fig2}
+\caption{Map of BCI 50 ha plot showing tree recruits (red crosses) against gap index for censuses 1985\textendash 1990 and 1990\textendash 1995.
+Colour gradient indicates the minimum amount of canopy shading above 2 m in a 5 year census.
+Dark colours indicate dense canopy shading, light colours indicate little or no canopy shading above 2 m.}
 \label{figS2}
+\end{figure}
+
+\clearpage
+\begin{figure}[!hb]
+\centering
+\includegraphics{figures/supp_fig3}
+\caption{alpha}
+\label{figS3}
+\end{figure}
+
+\clearpage
+\begin{figure}[!hb]
+\centering
+\includegraphics{figures/supp_fig4}
+\caption{alpha}
+\label{figS4}
+\end{figure}
+
+\clearpage
+\begin{figure}[!hb]
+\centering
+\includegraphics{figures/supp_fig5}
+\caption{alpha}
+\label{figS5}
 \end{figure}

### 2016-10-19 R/bci_data_processing.R

@@ -117,6 +117,8 @@ BCI_clean <- function(BCI_data, spp_table) {
         # Remove observations without a species code
         !is.na(sp)
     ) %>%
+    # fix missing species name.
+    mutate(species = replace(species, sp == "swars2", "Swartzia simplex2")) %>%
     # For each individual..
     group_by(treeid) %>%
     # Filter plants with multiple stems
@@ -161,7 +163,7 @@ BCI_clean <- function(BCI_data, spp_table) {
     # ensures at least 1 individual is in the heldout dataset
     filter(n_ind >=10) %>%
     ungroup() %>%
-    select(gx,gy,sp,n_ind,treeid,census,exactdate,julian,census_interval,pom,nostems,
+    select(gx,gy,species,sp,n_ind,treeid,census,exactdate,julian,census_interval,pom,nostems,
            dbh_prev,dbh,dead_next_census)
   
 }
@@ -250,7 +252,7 @@ reduce_to_single_ind_obs <- function(data) {
     filter(keep) %>%
     select(-keep) %>%
     ungroup() %>%
-    select(gx, gy,sp, sp_id, censusid, dead_next_census,
+    select(gx, gy,species,sp, sp_id, censusid, dead_next_census,
            census_interval, rho, dbh_prev, dbh)
 }
 

### 2016-10-19 R/figures.R

@@ -42,7 +42,7 @@ plot_obs_v_pred_growth <- function(data) {
 
 # Plot species predicted mortality v growth curves
 plot_spp_curves <- function(model, data, growth_range= c(0.03,0.5), hazard_curve = FALSE, 
-                            ylab="Annual mortality probability", xlab="Annual mortality probability") {
+                            ylab="Annual mortality probability", xlab="Annual dbh growth (cm)") {
   preds <- predict_spp_hazard(model, data, growth_range)
   
   breaks <- c(0.0001,0.001,0.01,0.1, 1, 10, 100)
@@ -161,6 +161,47 @@ plot_recruits_gapindex <- function(gap_index_raster, recruit_gap_conditions) {
                                         pch='+', col='red', alpha=0.4, cex=0.1),
                               data=list(recruit_gap_conditions=recruit_gap_conditions)))
   
+}
+
+# Plot species parameters
+
+plot_spp_params <- function(model, data, parameter,xlab ='Log effect size', title=NULL) {
+  # Extract hyper distribution summary
+  hyper <- summarise_hyper_params(model,data) %>%
+    subset(., grepl(parameter, names(.))) %>%
+    {
+      hyper_param <- as.character(TRUE)
+      species <- as.character('Hyper distribution')
+      mean <- .[[paste0('mu_log_',parameter)]]$mean
+      sd <- .[[paste0('sigma_log_',parameter)]]$mean
+      `2.5%` <- qnorm(0.025,mean,sd)
+      `97.5%` <- qnorm(0.975,mean,sd)
+      cbind.data.frame(hyper_param,species,mean,sd,`2.5%`,`97.5%`, stringsAsFactors=FALSE)
+    }
+  # Extract species param summaries
+  spp <- summarise_spp_params(model,data, TRUE) %>% 
+    .[[paste0('log_',parameter)]] %>%
+    mutate(hyper_param = as.character(FALSE),
+           sp = as.character(sp),
+           species = as.character(species))
+  
+  # Merge dataframes and prep for plotting
+  dat <-bind_rows(hyper, spp) %>%
+    mutate(
+      species = as.factor(species)) %>%
+    mutate(species = factor(species, levels=species[order(hyper_param,mean)], ordered=TRUE))
+  
+  #Plot 
+  ggplot(dat, aes(x = mean,y = species, col=hyper_param)) + 
[…]

### 2016-10-19 R/manuscript_functions.R

@@ -77,3 +77,11 @@ pred_gain_wd <- function(data) {
   
   return(signif(df$prop_spp_gain,2)*100)
 }
+
+# Extract code for full model
+fullmodelcode <- function() {
+  task <- tasks_2_run("final_model")[1,]
+  chunks <- get_final_model_chunks(task)
+  model <- make_stan_model(chunks)
+  writeLines(model$model_code)
+}

### 2016-10-19 R/process_output.R

@@ -189,23 +189,55 @@ summarise_times <- function(times) {
   return(res)
 }
 
-# Summarise posteriors of species level parameters
-summarise_spp_params <- function(model, data) {
+# Summarise hyper parameters
+summarise_hyper_params <- function(model, data) {
   fit <- model$fits[[1]]
   dat <- prep_full_data_for_stan(data)
-  samples <- rstan::extract(fit, pars=c("alpha","beta","gamma"))
-  
-  samples[["alpha_gamma"]] <- samples$alpha + samples$gamma
+  samples <- rstan::extract(fit, pars=c("mu_log_alpha","mu_log_beta","mu_log_gamma",
+                                        "sigma_log_alpha", "sigma_log_beta", "sigma_log_gamma"))
   
   lapply(samples, function(x) {
     cbind.data.frame(
-      sp = dat$sp,
-      wood_density = dat$raw_rho,
-      mean = apply(x,2, mean),
-      median = apply(x,2, median),
-      sd = apply(x,2,sd),
-      aperm(apply(x,2, quantile, c(0.025,0.975)), c(2,1)))
+      mean = mean(x),
+      median = median(x),
+      sd = sd(x),
+      t(quantile(x, c(0.025,0.975))))
   })
+}
+
+# Summarise posteriors of species level parameters
+summarise_spp_params <- function(model, data, logscale=FALSE) {
+  fit <- model$fits[[1]]
+  dat <- prep_full_data_for_stan(data)
+  samples <- rstan::extract(fit, pars=c("alpha","beta","gamma"))
+  samples[["alpha_gamma"]] <- samples$alpha + samples$gamma
+  
+  if(logscale==TRUE) {
+    res <-lapply(samples, function(x) {
+      cbind.data.frame(
+        species = dat$species,
+        sp = dat$sp,
+        wood_density = dat$raw_rho,
+        mean = apply(log(x),2, mean),
+        median = apply(log(x),2, median),
+        sd = apply(log(x),2,sd),
+        aperm(apply(log(x),2, quantile, c(0.025,0.975)), c(2,1)))
+    })
+    names(res) <- paste0('log_', names(res))
+    return(res)
+  }
+  else {
+    lapply(samples, function(x) {
+      cbind.data.frame(
+        species = dat$species,
+        sp = dat$sp,
+        wood_density = dat$raw_rho,
+        mean = apply(x,2, mean),
[…]

### 2016-10-19 R/stan_functions.R

@@ -205,6 +205,7 @@ prep_full_data_for_stan <- function(data) {
     growth_dt = data$true_dbh_dt - 0.172,
     rho_c  = unique(data$rho)/0.6,
     y = as.integer(data$dead_next_census),
+    species = unique(data$species),
     sp = unique(data$sp),
     raw_rho = unique(data$rho))
 }

### 2016-10-27 ms/ms.Rnw
More changes to intro and discussion

@@
 Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities.
-While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits and temporal plot-scale effects -- these distinct drivers are yet to be integrated within a cohesive and predictive framework.
-Here we apply classic survival analysis techniques within a cross-validated hierarchical Bayesian framework to derive instantaneous mortality functions that integrate individual, species, and temporal effects for 203 tropical rain forest tree species at Barro Colorado Island (BCI) Panama.
-We show that mortality models that partition mortality into growth-dependent and growth-independent components, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations.
-Moreover, we show that the effects of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability in species mortality rates, and that temporal variation in mortality was large.
-Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI.
-Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
+While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits and temporal effects -- these distinct drivers are yet to be integrated within a cohesive and predictive framework.
+In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive instantaneous mortality functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
+Specifically, we develop mortality functions that not only integrate individual, species and temporal effects, but also simultaneously estimate the relative contribution of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate.
+We show that functions that partition mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations.
+Moreover, we show that the effect of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability observed between species, and that temporal variation in mortality was large.
+Lastly, our analysis indicate that growth-dependent processes are the predominant contributor to rates of mortality at BCI.
+Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests.
+More importantly, it highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
 \end{abstract}
@@
 
-\dropcap{R}ates of plant mortality are known to be highly variable among individuals, among coexisting species, between forests and from year-to-year [refs].
+\dropcap{R}ates of plant mortality are known to be highly variable among individuals, coexisting species, between forests and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}.
 This variation has considerable consequences for forest structure and dynamics.
@@
 Generally, however, these influences have not been formalised as hazard functions\cite{Zens:2003jv}.
-Yet, such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function\cite{Hawkes:2000ib;Zens:2003jv,McDowell:2011dr}.
+Yet, such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function\cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.
 This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.
 
-A particular challenge in developing hazard functions for plants is to integrate the observed conflicting relationships between mortality and growth within and across species.
+\section*{Unifying intra- and inter-specific variation}
+
+A particular challenge in developing hazard functions for plants is to integrate the conflicting relationships observed between mortality and growth within and across species.
 Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
-Empirical studies provide broad support of this theory, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
-By contrast, across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality rates than individuals from slower growing species\cite{Poorter:2008iu, Wright:2010fl}.
+Empirical studies provide broadly support this theory, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
+By contrast, across species, there tends to be a strong trade-off between growth and survival, with faster growing species having higher mortality rates than slower growing species\cite{Poorter:2008iu, Wright:2010fl}.
 This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume).
@@
 
-Another major challenge is using binary observations (e.g.
-alive = 0; dead = 1) to infer the contribution of growth-dependent and growth-independent (i.e.
-stochastic events) hazards on tree mortality rates.
-By partitioning mortality rates into these two processes model predictive capacity may be improved and it allows us to narrow in on whether growth-dependent or growth-independent processes are the predominate cause of death in an ecosystem.
-For example, are tree deaths in rainforest predominately a result of competitive influences (i.e.
-growth-dependent processes) or are they more likely to result from growth-independent events such as treefall?
-More importantly, however, such partitions will allow us to examine how plant traits affect each hazard.
-For example, higher wood density is thought to convey higher survival by decreasing mortality aociated with stoachastic events such as those listed above.
-However, research has also shown that wood density may increase tolerance to low growth rates, and thus alter the growth-dependent mortality.
-For example, studies have found that species with high wood density are more tolerant to both shade \cite{vanGelder:2006exa}, neighbor competition \cite{Kunstler:2016km} and drought induced carbon starvation \cite{McDowell:2011dr}.
-Consequently, wood density may affect a trees response to both growth-dependent and growth-independent hazards.
-The question is, which hazard does it affect most?
+Another major challenge is estimating the relative contribution of growth-dependent and growth-independent hazards on an individual's overall instantaneous mortality rate.
+While plants can be killed by many factors, they all broadly fall into two classes.
+Plants can die due to insufficient carbon assimulation required for growth and repair (i.e.
+growth-dependent hazards), or they can die because of stochastic events that are independent of an individual's growth rate (e.g.
+windfall, fire).
+By developing models that can simultaneously estimate growth-dependent and growth-independent hazards we can begin to quantify their relative contribution and examine how they vary by ecosystem.
+More importantly, such models will allow us to determine how plant traits influence each hazard.
+For example, wood density is thought to primarily affect growth-independent mortality by reducing stem breakage \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+However, recent research has also indicated that higher wood density conveys increased tolerance to shade \cite{vanGelder:2006exa}, neighbor competition \cite{Kunstler:2016km} and drought induced carbon starvation \cite{McDowell:2011dr}.
+The question is, does wood density affect both hazards?
+And if so, which hazard does it affect most?
+Answering such questions will ultimately allow us to incorporate more biological information into existing mortality functions, which in turn, will increase model predictive capacity.
 
@@
 stochastic events such as wind throw and fire).
-Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (Moorcroft 2001; McDowell et al 2011).
+Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}.
 However, these alternative models have not previously been systematically compared.
@@
   \item What is the proportion of inter-specific variability explained by wood density?
-  \item What is the proportion of total variability explained by species, census, baseline hazard and growth-dependent hazard?
+  \item What is the proportion of total variability explained by species, census and growth-dependent and growth-independent hazards?
   \item Does mortality rates vary substantially between censuses?
@@
 competition, parasitization, localized herbivory).
-By contrast, the growth-independent hazard accounts for deaths caused by stochastic events that arise irrespective of an individuals growth rate (e.g.
+By contrast, the growth-independent hazard (i.e.
+baseline hazard) accounts for deaths caused by stochastic events that arise irrespective of an individuals growth rate (e.g.
 windthrow).
 
-Another advantage of partitioning mortality into growth-depedent and growth-independent hazards was that we could estimate the proportion of variation each attributed.
-Like many other studies\cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlight the importance of light competition an important factor influencing tropical tree demographic rates.
-Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted varability in mortality rates (Fig.
-\ref{fig5}), and thus, suggests that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.
+Another advantage of partitioning mortality into growth-dependent and growth-independent hazards was that we could estimate the proportion of variation attributed to each.
+Like many other studies\cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlight the importance of light competition in influencing tropical tree demographic rates.
+Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted variability in mortality rates (Fig.
+\ref{fig5}), indicating that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.
 
@@
 Specifically, we found that high wood density species had lower growth-independent mortality rates relative to their low wood density counterparts.
-This finding corroborating the observed negative correlation between mortality and wood density reported elsewhere\cite{Kraft:2010kq,Chave:2009iy}.
+This finding corroborates the observed negative correlation observed between mortality and wood density reported elsewhere\cite{Kraft:2010kq,Chave:2009iy}.
 More importantly, however, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}.
@@
 
-The underestimation of inter-specific variation suggests that there are other species attributes besides wood density that affect tree mortality rates.
+The high unexplained inter-specific variation suggests that there are other species traits besides wood density that affect tree mortality rates.
 It has been suggested that a species light demand and maximum size are two major axes of inter-specific variation in tropical rainforests\cite{Poorter:2006tq}.
@@
 
-Future research should also focus on examining how climate variables influence both growth-depedent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}.
+Furthermore, future research should also focus on examining how climate variables influence both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}.
 Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010.
@@
 However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by a El-Nino in 1997\textendash 1998.
-Droughts are known to have significant impacts on tree mortality rates in these ecosystems\cite{Condit:1995wl, Mcdowell:2013wf} Furthermore, future research should also examine how different growth measures influence predictive accuracy of mortality models.
+Droughts are known to have significant impacts on tree mortality rates in these ecosystems\cite{Condit:1995wl, Mcdowell:2013wf}.
+
+The length between censuses should also be examined to determine its affect on the estimation of growth-dependent and growth-independent hazards.
+Large census intervals are likely to lead to underestimation of growth-dependent mortality, and consequently, an overestimation of growth-independent mortality.
+This is because if a census interval is too wide, it will fail to account for deaths attributed to rapid declines in growth caused by senescence (especially for short-lived plants) or events such as drought.
+Because our analysis estimated an individual's past growth from 5 year census intervals, and because it included short-lived species (i.e.
+those with high annual mortality rates; see Fig.
+\ref{fig4}) and a drought period between 1995 and 2000, our estimates of growth-dependent and growth-independent hazards are likely to be biased.
+
+Lastly, research should also examine how different growth measures influence predictive accuracy of mortality models.
 We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive accuracy, and we are unsure how other common measures such as biomass growth\cite{Chave:2005hz} will compare.
@@
 Our findings reveal that while wood density is an important trait affecting mortality rates, we are still only capturing a fraction of the overall species variability in mortality rates.
-
-Example referencing suppmat figure (Fig. \ref{figS1}).
 
@@
 observed growth minus observation error).
-Details of this model can be found in supplementary material (REF, FigS1).
+Details of this model can be found in supplementary material (FigS1).
 
@@
 After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.
-These models were then cross validated using 10-fold cross validation (REF).
+These models were then cross validated using 10-fold cross validation.
 

### 2016-10-29 ms/ms.Rnw
Second proof read

@@
 In models of forest dynamics, variation in mortality rates have been found to have a larger impact on forest structure than variation in absolute growth rates\cite{Johnson-2016}.
-As such, improving our ability to understand the mortality process is a priority for making accruate predictions about forest and carbon dynamics as well as water or nutrient cycles; especially in an era of rapid environmental change.
+As such, improving our ability to understand the mortality process is a priority for making accurate predictions about forest and carbon dynamics as well as water or nutrient cycles; especially in an era of rapid environmental change.
 
@@
 Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required.
-As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys in long-term monitoring plots containining thousands of individuals and spanning several decades\cite{Anderson-Teixeira-2015a}.
+As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in monitoring plots containining thousands of individuals\cite{Anderson-Teixeira-2015a}.
 
@@
 
-\section*{Unifying intra- and inter-specific variation}
+\section*{A unified model of tree mortality}
 
@@
 Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
-Empirical studies provide broadly support this theory, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
-By contrast, across species, there tends to be a strong trade-off between growth and survival, with faster growing species having higher mortality rates than slower growing species\cite{Poorter:2008iu, Wright:2010fl}.
+Empirical studies broadly support this theory, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
+By contrast, across species, there tends to be a strong trade-off between growth and survival, with faster growing species typically exhibiting higher mortality rates than slower growing species\cite{Poorter:2008iu, Wright:2010fl}.
 This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume).
@@
 Another major challenge is estimating the relative contribution of growth-dependent and growth-independent hazards on an individual's overall instantaneous mortality rate.
-While plants can be killed by many factors, they all broadly fall into two classes.
-Plants can die due to insufficient carbon assimulation required for growth and repair (i.e.
+While plants can be killed by many factors, they all broadly fall into two independent hazards.
+Plants can die as a result of insufficient carbon assimulation required for growth and repair (i.e.
 growth-dependent hazards), or they can die because of stochastic events that are independent of an individual's growth rate (e.g.
 windfall, fire).
-By developing models that can simultaneously estimate growth-dependent and growth-independent hazards we can begin to quantify their relative contribution and examine how they vary by ecosystem.
-More importantly, such models will allow us to determine how plant traits influence each hazard.
+By developing models that can simultaneously estimate growth-dependent and growth-independent hazards we can begin to examine their relative contribution and how these vary by ecosystem.
+More importantly, such models will allow us to determine how plant traits and other factors influence each hazard.
 For example, wood density is thought to primarily affect growth-independent mortality by reducing stem breakage \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
-However, recent research has also indicated that higher wood density conveys increased tolerance to shade \cite{vanGelder:2006exa}, neighbor competition \cite{Kunstler:2016km} and drought induced carbon starvation \cite{McDowell:2011dr}.
+However, recent research has also indicated that higher wood density may convey increased tolerance to shade \cite{vanGelder:2006exa}, neighbor competition \cite{Kunstler:2016km} and drought induced carbon starvation \cite{McDowell:2011dr}.
 The question is, does wood density affect both hazards?
@@
 
-In this study, we reconcile both intraspecific and interspecific drivers and attempt to partition instantaneous mortality rates into growth-dependent and growth-independent hazards by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework.
-To derive this unified model, we compare the predictive accuracy of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig.
+In this study, we reconcile both intraspecific and interspecific factors and attempt to partition instantaneous mortality rates into growth-dependent and growth-independent hazards.
+We achieve this by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework.
+To derive this unified model, we compare the capacity of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig.
 \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama.
@@
 \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases.
-Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached.
+Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality rate decaying exponentially until an asymptote at zero is reached.
 The third model (eq.
@@
 Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero.
-The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e.
-stochastic events such as wind throw and fire).
+The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth (i.e.
+growth-dependent hazard) and death due to stochastic events such as windthrow or fire (i.e.
+growth-independent or baseline hazard).
 Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}.
@@
 Answering these questions involves comparing predictive accuracy of alternative models.
-To achieve this, we used K-fold cross-validation (Fig.
+To achieve this, we used 10-fold cross-validation (Fig.
 \ref{fig1}D).
@@
 Tree mortality progressively decreased over time.
-Between 1995 and 2010, the highest proportion of trees death occured between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and lastly 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%).
+Between 1995 and 2010, the highest proportion of trees death occured between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%).
 Consequently, when we allowed hazard rates to be scaled by census (i.e.
@@
 $(\alpha e^{\beta X_{i}} + \gamma)\delta$), significantly outperformed both the baseline only (i.e.$\gamma \delta$) or growth-dependent only (i.e.
-$(\alpha e^{\beta X_{i}})\delta$) hazard functions, irrespective of growth rate (Fig.
+$(\alpha e^{\beta X_{i}})\delta$) hazard functions (Fig.
 \ref{fig2}A).
-However, predictive accuracy was higher for stem diameter relative to basal-area growth.
+Moreover, we found that predictive accuracy was higher for stem diameter growth relative to basal-area growth.
 Perhaps surprisingly, we detected no significant improvement in predictive accuracy between the baseline only and growth-dependent only hazard functions.
@@
 \ref{fig2} A,B).
-The most parsimonious combination that provided the largest gain in predictive accuracy (i.e.
-lowest logarithmic loss) occurred when only the baseline hazard term ($\gamma$) was varied as a function of species wood density ($\rho$; Fig.
+However, the most parsimonious model with highest predictive accuracy (i.e.
+lowest logarithmic loss) was that which attributed the wood density effect, $\rho$, to the baseline hazard term ($\gamma$; Fig.
 \ref{fig2} B).
-We found that wood density was negatively correlated with species baseline mortality rate (Fig.
+Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate (Fig.
 \ref{fig3}).
-This meant that species with low wood density and fast individual growths had on average higher mortality rates ((Fig.
+This meant that fast growing individuals from a low wood density species had on higher average mortality rates ((Fig.
 \ref{fig4} A,B), and thus higher annual probability of death (Fig.
-\ref{fig4} C,D), relative to high wood density species.
-While, wood density did improve model accuracy it only accounted for \Sexpr{pred_gain_wd(logloss_summaries)}\% of the total gain in predictive capacity that could be obtained by accounting for species differences.
+\ref{fig4} C,D), relative to fast growing individuals of high wood density species.
 
 \subsection*{Proportion of variation explained}
-Approximately \Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of variation in predicted mortality rates captured by our model was due to species differences, \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of which was captured by wood density.
-Census effects accounted for approximately \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation in mortality rates.
-As expected, the growth-dependent hazard terms (i.e.
-the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the overall variation in mortality rates.
+By partitioning the modelled variation between our two random effects (i.e.
+species and census; Fig.
+\ref{fig5}A), we found that approximately \Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of variation in predicted mortality rates was due to species differences, \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of which was captured by wood density.
+By contrast, census effects accounted for approximately \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation in mortality rates.
+When we partitioned modelled variation into the two independent hazards we found that the growth-dependent hazard (i.e.
+the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the overall variation in mortality rates, and the remainder was attributed to the baseline hazard.
 
@@
 Our Bayesian framework coupled with cross validation revealed that the most accurate and parsimonious model of tropical tree death was that which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) attributed the effect of wood density to growth-independent mortality; and 4) incorporated temporal varability.
-Moreover, we found that rates of tropical tree mortality varied substantially between species and that wood density, a species level functional trait, only explained a small proportion of the overall inter-specific variation.
+Moreover, we found that rates of tropical tree mortality varied substantially between species and that wood density, a species level functional trait, explained only a small proportion of the overall inter-specific variation.
 
@@
 Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted variability in mortality rates (Fig.
-\ref{fig5}), indicating that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.
+\ref{fig5}).
+This indicates that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.
 
 Incorporating the effect of wood density on mortality rates also improved predictive performance.
-Our analyses revealed that the most parsimonious combination of wood density effects was when it was attributed to only growth-independent mortality.
-Specifically, we found that high wood density species had lower growth-independent mortality rates relative to their low wood density counterparts.
+Our analyses revealed that the most parsimonious combination of wood density effects was when it was attributed to only the growth-independent hazard term.
+Specifically, high wood density species had lower baseline rates relative to low wood density counterparts.
 This finding corroborates the observed negative correlation observed between mortality and wood density reported elsewhere\cite{Kraft:2010kq,Chave:2009iy}.
-More importantly, however, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}.
+More importantly, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}.
 
@@
 This means that such models are likely to severely underestimate the true varability in mortality rates.
-Consequently, because variation in mortality rates are known to have significant impacts on forest dynamics\cite{Johnson-2016}, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics, with flow on effects for ecosystem proporties such as GPP and NPP.
+Consequently, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics, with flow on effects for ecosystem proporties such as GPP and NPP\cite{Johnson-2016}.
 
@@
 
-The high unexplained inter-specific variation suggests that there are other species traits besides wood density that affect tree mortality rates.
-It has been suggested that a species light demand and maximum size are two major axes of inter-specific variation in tropical rainforests\cite{Poorter:2006tq}.
-While we did not explicitly model mortality as a function of these traits, our posthoc correlations \ref{fig6} revealed that some inter-specific variation was indeed correlated with light demand.
+The high unexplained inter-specific variation indicates that there are other species traits, besides wood density, that affect tree mortality rates.
+This unexplained variation may possibly be explained by a species' light demand and maximum height, which has recently been proposed as two major axes of inter-specific variation in tropical rainforests\cite{Poorter:2006tq}.
+While we did not model mortality as a function of these traits, our posthoc correlations (Fig.
+\ref{fig6}) revealed that some inter-specific variation was indeed correlated with light demand.
 Specifically, we found that light demanding species (i.e.
-those with high gap index) were more susceptible to death due to both low growth and stochastic chance and required faster growth rates to achieve a given mortality rate relative to a species that readily recruits in shaded conditions, supporting findings by others\cite{Ruger:2011cv}.
-By contrast, we detected no correlations between a species maximum stem diameter and mortality rates, supporting findings by others\cite{Wright:2010fl}.
+those that typically recruit in gaps; high gap index) had higher baseline hazard rates and were more susceptible to dying as a result of low growth, relative to those that readily recruit in shade (i.e.
+low gap index), supporting findings by others\cite{Ruger:2011cv}.
+By contrast, we detected no correlations between a species maximum stem diameter and mortality rates, both contradicting\cite{Poorter:2006tq} and supporting\cite{Wright:2010fl} findings by others.
 
-We recommend research should focus on examining how other traits influence growth-dependent and growth-independent hazards and what combination of traits minimises unexplained species variability.
-Researchers, however, should be mindful that the inclusion of additional traits comes at the cost of added complexity, increased uncertainty (i.e.
+We recommend future research should focus on examining how other traits influence growth-dependent and growth-independent hazards and what combination of traits minimises unexplained species variability.
+Having said this, researchers should be mindful that the inclusion of additional traits comes at the cost of complexity, increased uncertainty (i.e.
 more parameters to estimate) and increased minimum knowledge; all of which diminish the interpretability of models, but also the applicability of predicting mortality rates in many data poor ecosystems.
@@
 
-Furthermore, future research should also focus on examining how climate variables influence both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}.
+Furthermore, future research should focus on examining how climate variables influence both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}.
 Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010.
@@
 
-The length between censuses should also be examined to determine its affect on the estimation of growth-dependent and growth-independent hazards.
-Large census intervals are likely to lead to underestimation of growth-dependent mortality, and consequently, an overestimation of growth-independent mortality.
+How the length between censuses affect the estimation of growth-dependent and growth-independent hazards should also be examined.
+Large census intervals are likely to lead to underestimation of growth-dependent mortality, and consequently, overestimate the relative contribution of growth-independent hazards.
 This is because if a census interval is too wide, it will fail to account for deaths attributed to rapid declines in growth caused by senescence (especially for short-lived plants) or events such as drought.
@@
 \ref{fig1}).
-Logarithmic loss -- commonly known as log loss, and indicated by $\mathcal{L}$-- measures the accuracy of a model by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome (Fig.
+Logarithmic loss -- commonly known as log loss, and indicated by $\mathcal{L}$, measures the accuracy of a model by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome, $S_i$ (Fig.
 \ref{fig1}C).
-By minimizing the $\bar{\mathcal{L}}$, we maximize predictive skill.
+By minimizing the $\bar{\mathcal{L}}$, we maximize predictive capacity.
 The average log loss across all individuals for the $k$th fold of held-out data, $\mathcal{L}_k$, is given by

### 2016-10-29 ms/supp.Rnw
Second proof read

@@
 \clearpage
-\section{Appendix 2 - Stan code for final model fit with species random effects}
+\section{Appendix 2 - Stan code for final model fit with species random effects and wood density effect}
 
@@
 \centering
-\includegraphics{figures/supp_fig1.png}
+\includegraphics[width=\textwidth]{figures/supp_fig1.png}
 \caption{Observed vs predicted individual growth rates and stem diameters}
@@
 \includegraphics{figures/supp_fig3}
-\caption{alpha}
+\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\alpha$) for each species. Red = Model estimated hyper-distribution species effects were sampled from}
 \label{figS3}
@@
 \includegraphics{figures/supp_fig4}
-\caption{alpha}
+\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\beta$) for each species. Red = Model estimated hyper-distribution species effects were sampled from}
 \label{figS4}
@@
 \includegraphics{figures/supp_fig5}
-\caption{alpha}
+\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\gamma$) for each species. Red = Model estimated hyper-distribution species effects were sampled from}
 \label{figS5}

### 2016-10-29 ms/supp.Rnw
forgot to save these small changes

@@
 
-Example reference \cite{Johnson-2016}.
-
 \subsection{Modelling true growth}
 
-Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions.
-In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of observed growth rates were negative.
-While there are physiological processes (e.g.
-water depletion and bark drop) that can lead to trunk shrinkages overtime\cite{Pastur:2007kl}, it is biologically impossible for growth (i.e.
-the creation of new biomass) to be negative\cite{Clark:2007ve}.
 To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e.
@@
 \end{equation}
-where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$ *(eq.
-\ref{eq:mortality_model}).
+where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$.
 As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
@@
 
-As we have multiple hypothesized functional forms of $\lambda_i$ (e.g.
-equations \ref{base_haz}\textendash\ref{add_haz}), we describe here the fitting procedure of our most complicated model, of which all others may be considered special cases.
+As we have multiple hypothesized functional forms of $\lambda_i$, we describe here the fitting procedure of our most complicated model, of which all others may be considered special cases.
 The instantaneous rate of mortality is

### 2017-05-31 ms/ms.Rnw
forgot to save these small changes
Edits throughout
- refine and shorten text
- update PNAS style
- better formatting of Supp Matt
- update units in figures

@@
-\documentclass[9pt,twocolumn,twoside]{pnas-new}
+\documentclass[9pt,twocolumn,twoside]{pnas-new}\usepackage[]{graphicx}\usepackage[]{color}
+
 % Use the lineno option to display guide line numbers if required.
@@
 
-\title{Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework}
+\title{Unifying intra- and inter-specific variation in tropical tree mortality}
 
@@
 \author[b]{S Joseph Wright}
-\author[a]{Daniel S Falster}
+\author[a,c]{Daniel S Falster}
 
@@
 \affil[b]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
+\affil[c]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales}
 
@@
 % 120 words
-\significancestatement{Tree mortality is a fundamental demographic process which affects forest dynamics and consequently, a terrestrial ecosystems ability to store carbon.
+\significancestatement{Tree mortality is a fundamental demographic process affecting forest dynamics and carbon cycling.
 Here, for the first time, we use observed survival records over a 15 year period from more than 180,000 individuals, to simultaneously estimate growth-dependent and growth-independent (i.e.
-death by stocastic chance) mortality across 203 tropical forest tree species.
-We found that growth-dependent mortality was the predominate factor influencing tree mortality rates at Barro Colorado Island (BCI).
+stochastic) mortality across 203 tropical forest tree species.
+We found that growth-dependent mortality was the predominate factor influencing tree mortality rates at Barro Colorado Island.
 Furthermore, we found that wood density most strongly influenced mortality rates by decreasing growth-independent mortality.
@@
 % Please include corresponding author, author contribution and author declaration information
-\authorcontributions{J.C.
-designed and performed the analysis; D.F.
-conceived the idea, supervized the development of this work; M.W.
-provided biological insight.
-All authors contributed to the writing of this manuscript.}
+\authorcontributions{J.C. performed research; D.S.F. conceived the idea; S.J.W. collected data. All authors contributed to the writing of this manuscript.}
 
-\authordeclaration{The authors declare that they have no competing financial interests.}
+\authordeclaration{The authors declare that they have no competing financial interests.\\
+
+Data deposition: Code and data reproducing the results in this paper are available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.
+}
+
 \correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: james.camac@gmail.com}
@@
 Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities.
-While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits and temporal effects -- these distinct drivers are yet to be integrated within a cohesive and predictive framework.
+While past research has identified factors influencing tree mortality across a variety of scales, these distinct drivers are yet to be integrated within a unified, predictive framework.
 In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive instantaneous mortality functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
@@
 \doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 \ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
-
-% If your first paragraph (i.e.
-with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation.
-If this is the case, add \parshape=0 to the end of the list environment.
 
@@
 This requirement arises in part because the observable outcome -- alive vs.
-dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate).
-In classic survival analysis\cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
+dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$, also called the ``hazard function''.
+In classic survival analysis\cite{Kleinbaum:2012tn}, the probability an individual dies between times $t_2$ and $t_3$ is a function of $\lambda_i(t)$:
 \begin{equation} \label{eq:mortality_model}
@@
 \end{equation}
-Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required.
-As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in monitoring plots containining thousands of individuals\cite{Anderson-Teixeira-2015a}.
+Since trees are long-lived and we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required.
+As such, detailed studies of tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals\cite{Anderson-Teixeira-2015a}.
 
-The other difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality.
+Another difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of hazard functions that skilfully predict patterns in mortality.
 Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate\cite{Kobe:1995tw,Wyckoff:2002ul}, traits\cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size\cite{Vieilledent:2010fv, Hurst:2011ky}.
-Generally, however, these influences have not been formalised as hazard functions\cite{Zens:2003jv}.
-Yet, such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function\cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.
-This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.
+Generally, however, these influences have not been integrated into a common hazard function\cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.
 
-\section*{A unified model of tree mortality}
+\section*{Towards a unified model of tree mortality}
 
-A particular challenge in developing hazard functions for plants is to integrate the conflicting relationships observed between mortality and growth within and across species.
-Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
-Empirical studies broadly support this theory, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}.
-By contrast, across species, there tends to be a strong trade-off between growth and survival, with faster growing species typically exhibiting higher mortality rates than slower growing species\cite{Poorter:2008iu, Wright:2010fl}.
-This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume).
-Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack\cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events\cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+A specific challenge in developing hazard functions for plants is to estimate the relative contribution of growth-dependent and growth-independent hazards on an individual's overall hazard.
+While plants die via many causes, these broadly fall into two categories.
+The first are growth-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair.
+The second are growth-independent hazards, where plants die because of stochastic events, irrespective of their growth rate, such as  windfall or fire.
+To capture this distinction, we assume an individual's hazard is the sum of growth-independent ($\lambda_{\textrm{b},i}(t)$) and growth-dependent ($\lambda_{\textrm{g},i}(t)$) rates:
+\begin{equation} \label{eq:additive}
+\lambda_i(t)  = \lambda_{\textrm{b},i}(t) + \lambda_{\textrm{g},i}(t).
+\end{equation}
 
-Another major challenge is estimating the relative contribution of growth-dependent and growth-independent hazards on an individual's overall instantaneous mortality rate.
-While plants can be killed by many factors, they all broadly fall into two independent hazards.
-Plants can die as a result of insufficient carbon assimulation required for growth and repair (i.e.
-growth-dependent hazards), or they can die because of stochastic events that are independent of an individual's growth rate (e.g.
-windfall, fire).
-By developing models that can simultaneously estimate growth-dependent and growth-independent hazards we can begin to examine their relative contribution and how these vary by ecosystem.
-More importantly, such models will allow us to determine how plant traits and other factors influence each hazard.
-For example, wood density is thought to primarily affect growth-independent mortality by reducing stem breakage \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
-However, recent research has also indicated that higher wood density may convey increased tolerance to shade \cite{vanGelder:2006exa}, neighbor competition \cite{Kunstler:2016km} and drought induced carbon starvation \cite{McDowell:2011dr}.
-The question is, does wood density affect both hazards?
-And if so, which hazard does it affect most?
-Answering such questions will ultimately allow us to incorporate more biological information into existing mortality functions, which in turn, will increase model predictive capacity.
+A further challenge in developing hazard functions for plants is to integrate the conflicting relationships observed within and across species between mortality and growth.
+Within species, mortality rates are lower for fast-growing individuals, presumably because those individuals have superior carbon budgets and are thus able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
+Empirical studies broadly support this theory, with many indicating exponential declines in mortality with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, such that with parameters $\alpha, \beta$:
+\begin{equation} \label{eq:gd}
+\lambda_{\textrm{g},i}(t) = \alpha \textrm{e}^{-\beta X_i(t)}.
+\end{equation}
+
+By contrast, across species, there is a strong trade-off between growth and survival, with individuals from faster-growing species  exhibiting higher mortality rates than individuals from slower-growing species\cite{Poorter:2008iu, Wright:2010fl}.
+This trade-off may arise via traits with antagonistic effects, such as wood density.
+Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage\cite{vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack\cite{Augspurger:1984wx}, and ultimately reduced mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
 
@@
 C) Log-loss function highlighting that the further the predicted probability is from truth the higher the logloss.
-D) The predictive accuracy of alternative models is evaluated via 10-fold cross validation.
+D) The predictive skill of alternative models is evaluated via 10-fold cross validation.
 The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$).
@@
 
-In this study, we reconcile both intraspecific and interspecific factors and attempt to partition instantaneous mortality rates into growth-dependent and growth-independent hazards.
+In this study, we attempt to reconcile intra-specific and inter-specific factors as well as partition instantaneous mortality rates into growth-dependent and growth-independent rates.
 We achieve this by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework.
-To derive this unified model, we compare the capacity of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig.
+To derive this unified model, we compare the skill of three alternative hazard functions ($\lambda_i(t)$; Fig.
 \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama.
-The simplest model (Fig.
-\ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate.
+Beginning from eq.\ref{eq:additive}, the simplest model (Fig.
+\ref{fig1}B, left), assumes a single baseline, growth-independent hazard rate.
 The second model (Fig.
-\ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases.
-Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality rate decaying exponentially until an asymptote at zero is reached.
+\ref{fig1}B, middle) assumes the risk of dying eventually declines to zero as growth increases (eq.
+\ref{eq:gd}).
 The third model (eq.
 Fig.
-\ref{fig1}B, right) is the summation of the two previous models.
-Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero.
-The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth (i.e.
-growth-dependent hazard) and death due to stochastic events such as windthrow or fire (i.e.
-growth-independent or baseline hazard).
-Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}.
-However, these alternative models have not previously been systematically compared.
+\ref{fig1}B, right) is the summation of the two previous models, which allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, asymptoting at some baseline hazard $\geq 0$.
+Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared.
 
-In deriving the unified model, we address the following specific questions:
-\begin{enumerate}\itemsep 0pt
-  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
-  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
-  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
-  \item What is the proportion of inter-specific variability explained by wood density?
-  \item What is the proportion of total variability explained by species, census and growth-dependent and growth-independent hazards?
-  \item Does mortality rates vary substantially between censuses?
-\end{enumerate}
-Answering these questions involves comparing predictive accuracy of alternative models.
+To derive the unified model, we asked the following questions:
+1) Does mortality rates vary substantially between censuses, such that a specific census effect is needed?
+2) Does a hazard function including both growth-dependent and independent terms outperform a function including only one of these?
+3) Which growth measure (stem diameter or stem area increment) more skilfully predicts growth-dependent mortality?
+4) Does including wood density (a species trait) improve model skill, and if so, does it affect parameters of growth-dependent or independent hazard or both?
+5) What proportion of inter-specific variability in hazard is explained by wood density?
+6) What proportion of total variability in hazard is explained by species, census, growth-dependent and growth-independent hazards?
+
+Answering these questions involves comparing the  predictive skill of alternative models.
 To achieve this, we used 10-fold cross-validation (Fig.
-\ref{fig1}D).
-Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}.
-Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations.
-For example, some of our models took over a two weeks to complete.
-We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increasing model complexity (see methods).
+\ref{fig1}D), to systematically assesses each model's skill in predicting outcomes in novel data (i.e.
+not used in model calibration) \cite{Hooten:2015, Stone:1997tg}.
+Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations.
+We therefore fit models across five iterative stages of model development, corresponding to the six questions evaluated above, of gradually increased complexity (see Methods).
 
@@
 \subsection*{Mortality over time}
-Tree mortality progressively decreased over time.
-Between 1995 and 2010, the highest proportion of trees death occured between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%).
+Comparing the 3 intervals between censuses from 1995 to 2010, we found tree mortality progressively decreasing over time.
+The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%).
 Consequently, when we allowed hazard rates to be scaled by census (i.e.
-adding term $\delta$) we observed a significant increase in predictive accuracy (i.e.
-lower logarithmic loss; Fig.
+adding term $\delta$) we observed a significant increase in predictive skill (i.e.
+Fig.
 \ref{fig2}A).
-Assuming the long-term average census effect ($\delta$) is 1, then our model predicted deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.
+Assuming the long-term average census effect ($\delta$) is 1, deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.
 
@@
 \includegraphics{figures/fig2.pdf}
-\caption{Mean ($\pm$ 95\% credible intervals) logarithmic loss whereby lower values equal higher predictive accuracy. 
+\caption{Mean ($\pm$ 95\% credible intervals) logarithmic loss whereby lower values equal higher predictive skill.
 A) Five sequential stages of model selection with each stage increasing in complexity.
-Null: null model ($\gamma$); Census: inclusion of census effects ($\delta$); Growth rate: the inclusion of growth rate.
-This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{\beta X_{i}})\delta$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{\beta X_{i}} + \gamma)\delta$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
+Null: null model ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate.
+This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
 B) Change in logloss under all wood density\textendash parameter combinations.
@@
 \subsection*{Growth hazard functions}
-We found that the hazard function that incorporates both growth-dependent and baseline hazard terms (i.e.
-$(\alpha e^{\beta X_{i}} + \gamma)\delta$), significantly outperformed both the baseline only (i.e.$\gamma \delta$) or growth-dependent only (i.e.
-$(\alpha e^{\beta X_{i}})\delta$) hazard functions (Fig.
+We found that a hazard function with both growth-dependent and growth-independent terms (i.e.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$), significantly outperformed both the baseline only (i.e.$\gamma \delta_t$) or growth-dependent only (i.e.
+$(\alpha e^{-\beta X_{i}})\delta_t$) functions (Fig.
 \ref{fig2}A).
-Moreover, we found that predictive accuracy was higher for stem diameter growth relative to basal-area growth.
-Perhaps surprisingly, we detected no significant improvement in predictive accuracy between the baseline only and growth-dependent only hazard functions.
+Moreover, we found that predictive skill was higher when using stem diameter growth over basal-area growth.
 
@@
 \subsection*{Effect of wood density on mortality rates}
-The inclusion of wood density on either baseline or growth-dependent parameters significantly improved model accuracy relative to a hazard function without such effects (Fig.
+Including wood density ($\rho$) as an effect on either baseline or growth-dependent parameters significantly improved model skill relative to a hazard function without such effects (Fig.
 \ref{fig2} A,B).
-However, the most parsimonious model with highest predictive accuracy (i.e.
-lowest logarithmic loss) was that which attributed the wood density effect, $\rho$, to the baseline hazard term ($\gamma$; Fig.
+The most parsimonious model, with highest predictive skill, was that which attributed the wood density effect to the baseline hazard term ($\gamma$; Fig.
 \ref{fig2} B).
-Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate (Fig.
+Wood density was therefore negatively correlated with a species' baseline mortality rate (Fig.
 \ref{fig3}).
-This meant that fast growing individuals from a low wood density species had on higher average mortality rates ((Fig.
+
+% This meant that fast growing individuals from a low wood density species had on higher average mortality rates (Fig.
 \ref{fig4} A,B), and thus higher annual probability of death (Fig.
@@
 
-\begin{figure}[h!]
-\includegraphics{figures/fig4.pdf}
-\caption{Predicted changes in instantaneous mortality rates (A,B), and its derivative, annual mortality probability (C,D), with increasing growth rates.
-A,C) Estimated species curves for all 203 BCI species used in this study.
-B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species.}
-\label{fig4}
-\end{figure}
-
 \subsection*{Mortality rates, maximum species size and species light requirements}
-We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were positively correlated with gap index (our measure of species light requirement; Fig.
+We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were weakly and positively correlated with gap index (our measure of species light requirement; Fig.
 \ref{fig6} A,C, E).
@@
 The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards\cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}.
-We show that regardless of growth measure, incorporating both hazards significantly improves model predictive accuracy.
+We show that regardless of growth measure, incorporating both hazards significantly improves model predictive skill.
 This is because the additive effects of both hazards explicitly allow models to account for two independent causes of plant mortality.
@@
 
+\begin{figure}[h]
+\includegraphics{figures/fig4.pdf}
+\caption{Predicted changes in instantaneous mortality rates (A,B), and annual mortality probability (C,D), with increasing growth rates.
+A,C) Estimated species curves for all 203 BCI species used in this study.
+B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species.}
+\label{fig4}
+\end{figure}
+
 Incorporating the effect of wood density on mortality rates also improved predictive performance.
@@
 This finding corroborates the observed negative correlation observed between mortality and wood density reported elsewhere\cite{Kraft:2010kq,Chave:2009iy}.
-More importantly, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}.
+More importantly, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall\cite{vanGelder:2006exa,Chave:2009iy}.
 
-While wood density effects are now being incorporated in mortality algorithms of many vegetation models\cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation.
-This means that such models are likely to severely underestimate the true varability in mortality rates.
-Consequently, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics, with flow on effects for ecosystem proporties such as GPP and NPP\cite{Johnson-2016}.
+While wood density effects are now being incorporated in mortality algorithms of many vegetation models\cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation (Fig.
+\ref{fig4}).
+This means that such models are likely to severely underestimate the true variability in mortality rates.
+Consequently, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics, with flow on effects for ecosystem properties such as GPP and NPP\cite{Johnson-2016}.
 
@@
 Unfortuantely, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3).
-However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by a El-Nino in 1997\textendash 1998.
+However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by an El-Nino in 1997\textendash 1998.
 Droughts are known to have significant impacts on tree mortality rates in these ecosystems\cite{Condit:1995wl, Mcdowell:2013wf}.
@@
 
-Lastly, research should also examine how different growth measures influence predictive accuracy of mortality models.
-We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive accuracy, and we are unsure how other common measures such as biomass growth\cite{Chave:2005hz} will compare.
+Lastly, research should also examine how different growth measures influence predictive skill of mortality models.
+We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive skill, and we are unsure how other common measures such as biomass growth\cite{Chave:2005hz} will compare.
 
@@
 \matmethods{
+\vspace{-2em}
 \subsection*{Data}
-We parametrized plant mortality models using individual growth and survival data\cite{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\textdegree;N, 79.85\textdegree;W).
-The climate on the island is warm throughout the year with a mean annual temperature of 26.1\textdegree and rainfall is seasonal with most of the 2500 mm falling between April and November.
-BCI is managed exclusively for field research by the Smithsonian Tropical Research Institute (STRI) and the permanent plot was established in 1980.
+We derived plant mortality models using individual growth and survival data\cite{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\textdegree;N, 79.85\textdegree;W).
+The climate on the island is warm and rainfall is seasonal with falling between April and November.
+BCI is managed exclusively for field research by the Smithsonian Tropical Research Institute.
 
 Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter\cite{Condit:2012nz}.
-For the purpose of modelling mortality as a function of past growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm whereas later censuses were rounded to the nearest millimetre.
+For the purpose of modelling mortality as a function of past growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm, whereas later censuses were rounded to the nearest millimetre.
 Consequently, we modelled tree mortality as a function of past growth for censuses 1995\textendash 2000, 2000\textendash 2005 and 2005\textendash 2010.
-We also excluded individuals based on the following rules: 1) those that did not survive at least two censuses (two being required to estimate growth rate); 2) those that were not consistently measured at 1.3 m above ground; 3) those that were multi-stemmed; 4) those that resprouted or \textit{``returned from the dead''}; or 5) extreme outliers -- stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial diameter.
-We also discarded species that do not exhibit secondary growth (e.g.
+We discarded species that do not exhibit secondary growth (e.g.
 palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density.
-In order to maintain statical independence of measurements, we then sub-sampled this dataset by randomly selecting one observation per individual out of a possible three censuses.
+We also excluded individuals that: 1) did not survive at least two censuses (two being required to estimate growth rate); 2) were not consistently measured at 1.3 m above ground; 3) were multi-stemmed; 4) resprouted or \textit{``returned from the dead''}; or 5) were extreme outliers -- stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial diameter.
+In order to maintain statical independence of measurements, we sub-sampled the dataset by randomly selecting one observation per individual out of a possible three censuses.
 In total, \Sexpr{n_inds(BCI_model_dataset_true_unique)} individuals of \Sexpr{n_spp(BCI_model_dataset_true_unique)} species were included in our analysis.
@@
 Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100\textdegree C) and dividing by the fresh volume (as measured by water displacement).
-Wood density was then calculated as an area-weighted average (g cm$^-3$), where area refers to the annulus represented by each piece, assuming a circular trunk.
+Wood density was then calculated as an area-weighted average (g cm$^{-3}$), where area refers to the annulus represented by each piece, assuming a circular trunk.
 
@@
 
-Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions.
-In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of observed growth rates were negative.
-While there are physiological processes (e.g.
-water depletion and bark drop) that can lead to trunk shrinkages overtime\cite{Pastur:2007kl}, it is biologically impossible for growth (i.e.
-the creation of new biomass) to be negative\cite{Clark:2007ve}.
-To ensure our mortality model was not biased by these implausible values we first developed a model to estimate \textit{``true growth''}(i.e.
-observed growth minus observation error).
-Details of this model can be found in supplementary material (FigS1).
+Field measurements of diameter inevitably include observation error, which effect estimates of growth rates.
+In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of estimated growth rates were negative.
+To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community.
+Details of this model can be found in Supplementary Material S1.
 
-\subsection*{Estimating mortality rates}
+\subsection*{Hazard functions}
+We fit a series of models with increasing complexity.
 At each stage, the best model from the previous stage was taken as input for the next stage.
-Specifically, stage 1 fitted the simplest model (Fig.
-\ref{fig1}B) and compared it to the same model but one which varied by census.
-This allowed us to determine whether $\lambda$ significantly varied over-time and thus whether accounting for it improved predictive capacity.
-Stage 2 involved comparing predictive performance of the three $\lambda$ models described in (Fig.
-\ref{fig1}B) using either annual diameter increment or area growth rates.
+In stage 1 we fit the simplest model (a constant) and compared to a same where hazard rate varied by census.
+Stage 2 involved comparing predictive performance of the three functions described in (Fig.
+\ref{fig1}B) using either stem diameter or stem area growth rate.
 This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure.
-In stage 3, we begin to examined inter-specific variability by using the best model of stage 2 but allowing various combinations of parameters to vary as a function of species wood density.
+In stage 3, we examined inter-specific variability by using the best model of stage 2 but allowing various combinations of parameters to vary as a function of species wood density.
 This allowed us to determine if wood density improved predictive capacity, and if so, whether it predominately improved predictive capacity via growth-dependent or growth-independent processes.
 Stage 4 also examined inter-specific variability, however, here the focus was on using the best model of stage 2 and estimating the total predictive gain obtained if all parameters were allowed to vary as a function of species ID.
-The last stage was fitting the best model, with species random effects and wood density to the entire dataset (i.e.
-not cross validated).
+The last stage was fitting the best model, with species random effects and wood density to the entire dataset.
 This allowed us to examine the range of species hazard curves, how they are affected by wood density, census and posthoc covariates, and the proportion of predicted variation explained by species, wood density, census and growth-dependent and baseline hazard terms.
@@
 After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.
-These models were then cross validated using 10-fold cross validation.
 
 \subsection*{Model evaluation}
-Predictive accuracy was quantified by estimating the average log loss across 10-folds for held-out data, $\bar{\mathcal{L}}$ (Fig.
+Predictive skill was quantified by estimating the average log loss across 10-folds for held-out data, $\bar{\mathcal{L}}$ (Fig.
 \ref{fig1}).
-Logarithmic loss -- commonly known as log loss, and indicated by $\mathcal{L}$, measures the accuracy of a model by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome, $S_i$ (Fig.
+Logarithmic loss -- commonly known as log loss, $\mathcal{L}$, measures the skill of a model by penalizing incorrect predictions, based on how wrong the predicted probability is from the observed outcome, $S_i$ (Fig.
 \ref{fig1}C).
-By minimizing the $\bar{\mathcal{L}}$, we maximize predictive capacity.
-The average log loss across all individuals for the $k$th fold of held-out data, $\mathcal{L}_k$, is given by
+Lower $\mathcal{L}$ implies greater skill.
+The average log loss across all individuals for the $k$th fold of held-out data, $\mathcal{L}_k$, is then
 \begin{equation} \label{eq:logloss}
@@
 
-In order to calculate our gap index (our measure of a species light dependence) we also used annual canopy census data collected during 1985\textendash 1990 and 1990\textendash 1995.
-The canopy census recorded the presence of vegetation in six height intervals, 0–2, 2–5, 5–10, 10–20, 20–30 and 30 m every 5-m across the 50 ha.
+In order to calculate our gap index (our measure of a species light dependence) we also used annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995.
+The canopy census recorded the presence of vegetation in six height intervals, 0–2, 2–5, 5–10, 10–20, 20–30 and 30 m every 5 m across the 50 ha.
 We calculated the minimum observed number of vegetation strata above 2 m all 5x5 cells in each census period.

### 2017-05-31 ms/supp.Rnw
Edits throughout
- refine and shorten text
- update PNAS style
- better formatting of Supp Matt
- update units in figures

@@
-\documentclass[9pt,onecolumn,twoside]{pnas-new}
+\documentclass[9pt,twocolumn,twoside]{pnas-new}
 % Use the lineno option to display guide line numbers if required.
@@
 \affil[b]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
+\affil[c]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales}
+
+% Please give the surname of the lead author for the running footer
+\leadauthor{Camac}
 
@@
 
-\tableofcontents
-
 \renewcommand{\thefigure}{S\arabic{figure}}
 \renewcommand{\thetable}{S\arabic{table}}
-\setcounter{secnumdepth}{0}
+\renewcommand{\theequation}{S\arabic{equation}}
+\renewcommand{\thesection}{S\arabic{section}}
 
-\section{Appendix A - Supplementary text}
+% \setcounter{secnumdepth}{0}
 
-\subsection{Modelling true growth}
+Our model is presented in detail in ...... Some intro text.
+
+\section{Estimating True Growth}
 
@@
 where $\Delta D_{i,2}$ represents the annual true increase in diameter (cm) and $(t_2-t_1)$ represents the number of days between initial and final measurements divided by 365.25.
-We modelled $\Delta D_{i,2}$ as a random realisation from a log-normal distribution centerd on zero and a standard deviation of 1.
-This prior conveyed our expectation that annual diameter growth could not be negative and is unlikely to be greater than 7 cm/yr.
+We modelled $\Delta D_{i,2}$ as a random realisation from a log-normal distribution centered on zero and a standard deviation of 1.
+This prior conveyed our expectation that annual diameter growth could not be negative and is unlikely to be greater than 7 cm yr$^{-1}$.
 
@@
 
-\subsection{Mortality model}
+\section{Mortality model}
+
 We modelled tree mortality using survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$.
@@
 \ref{eq:submodels} ensures that parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ remain positive.
-On a log scale this model equates to an additive linear model , e.g.
-$\log \alpha_s = \log \alpha_{0,s} + (\alpha_{1} - \log \rho_c) \, \log \rho$.
+On a log scale this model equates to an additive linear model , i.e.
+\begin{equation}\label{eq:submodels_log}
+\log \alpha_s = \log \alpha_{0,s} + (\alpha_{1} - \log \rho_c) \, \log \rho.
+\end{equation}
 
-We standardized wood density via the transformation: $\frac{\rho}{\rho_c}$, where $\rho_c=0.6$ g cm${^-3}$.
-This operation is equivalent to centering a log-transformed $\rho$ around $\rho_c$ and enables mean hyper-parameters  to be interpreted as the value for a species with average wood density of 0.6 cm${^-3}$.
+We standardized wood density via the transformation: $\frac{\rho}{\rho_c}$, where $\rho_c=0.6$ g cm$^{-3}$.
+This operation is equivalent to centering a log-transformed $\rho$ around $\rho_c$ and enables mean hyper-parameters  to be interpreted as the value for a species with average wood density of 0.6 cm$^{-3}$.
 Similarly, we also centered $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.
@@
 
-\clearpage
-\section{Appendix 2 - Stan code for final model fit with species random effects and wood density effect}
+\section{Numerical optimisation}
 
-<<echo=FALSE>>=
-fullmodelcode()
+\section{Example Stan code}
+
+Models were implemented in the `Stan` framework, which uses a specific language to encode model structure \cite{StanDevelopmentTeam:2015uz}.
+Code for all models implemented are available in the supplied source code.
+As an example, we include here the code for the final model fit, including with species random effects, wood density effect and census effect.
+
+<<echo=FALSE, highlight=TRUE, comment = ''>>=
+task <- tasks_2_run("final_model")[1, ]
+chunks <- get_final_model_chunks(task)
+model <- make_stan_model(chunks)
+code <- gsub("\n      ", "\n", model$model_code)
+code <- strsplit(code, "\n")[[1]]
+writeLines(strwrap(code, width = 50, exdent = 10))
 @
 
+\section{References}
+\bibliography{refs.bib}
+
 \clearpage
-\section{Supplementary Figures}
-
-\begin{figure}[!hb]
+\begin{figure*}[!hb]
 \centering
-\includegraphics[width=\textwidth]{figures/supp_fig1.png}
+\includegraphics[width=0.6\textheight]{figures/supp_fig1.png}
 \caption{Observed vs predicted individual growth rates and stem diameters}
 \label{figS1}
-\end{figure}
+\end{figure*}
 
 \clearpage
-\begin{figure}[!hb]
+\begin{figure*}[!hb]
 \centering
-\includegraphics{figures/supp_fig2}
+\includegraphics[width=\textwidth]{figures/supp_fig2}
 \caption{Map of BCI 50 ha plot showing tree recruits (red crosses) against gap index for censuses 1985\textendash 1990 and 1990\textendash 1995.
@@
 \label{figS2}
-\end{figure}
+\end{figure*}
 
 \clearpage
-\begin{figure}[!hb]
+\begin{figure*}[!hb]
 \centering
-\includegraphics{figures/supp_fig3}
+\includegraphics[height=0.8\textheight]{figures/supp_fig3}
 \caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\alpha$) for each species. Red = Model estimated hyper-distribution species effects were sampled from}
 \label{figS3}
-\end{figure}
+\end{figure*}
 
 \clearpage
-\begin{figure}[!hb]
+\begin{figure*}[!hb]
 \centering
@@
 \label{figS4}
-\end{figure}
+\end{figure*}
 
 \clearpage
-\begin{figure}[!hb]
+\begin{figure*}[!hb]
 \centering
@@
 \label{figS5}
-\end{figure}
+\end{figure*}
 
-\clearpage
-
-\section{References}
-
-\bibliography{refs.bib}
 \end{document}

### 2017-05-31 R/figures.R

@@ -24,8 +24,8 @@ plot_obs_v_pred_growth <- function(data) {
     geom_abline(intercept = 0, slope = 1, col='red', size= 0.2) +
     scale_x_continuous(expand=c(0,0), limit=c(NA, 5)) +
     scale_y_continuous(expand=c(0,0), limit=c(0, 5)) +
-    ylab(expression(widehat("DBH growth")~(cm/yr))) + 
-    xlab('DBH growth (cm/yr)') +
+    ylab(expression(widehat("DBH growth")~("cm yr"^-1))) +
+    xlab(expression("DBH growth"~("cm yr"^-1))) +
     partial_plot_theme()
   
   p4 <- ggplot(data = data,  aes(x = obs_basal_area_dt, y = true_basal_area_dt)) + 
@@ -33,8 +33,8 @@ plot_obs_v_pred_growth <- function(data) {
     geom_abline(intercept = 0, slope = 1, col='red', size= 0.2) +
     scale_x_continuous(expand=c(0,0), limit=c(NA, 300)) +
     scale_y_continuous(expand=c(0,0), limit=c(0, 300)) +
-    ylab(expression(widehat("Basal area growth")~(cm^2/yr))) + 
-    xlab(expression("Basal area growth"~(cm^2/yr))) +
+    ylab(expression(widehat("Basal area growth")~("cm"^2~"yr"^-1))) +
+    xlab(expression("Basal area growth"~("cm"^2~"yr"^-1))) +
     partial_plot_theme()
   
   plot_grid(p1, p2, p3, p4, ncol=2, labels=LETTERS[1:4], label_size = 7)
@@ -42,7 +42,9 @@ plot_obs_v_pred_growth <- function(data) {
 
 # Plot species predicted mortality v growth curves
 plot_spp_curves <- function(model, data, growth_range= c(0.03,0.5), hazard_curve = FALSE, 
-                            ylab="Annual mortality probability", xlab="Annual dbh growth (cm)") {
+                            ylab="Annual mortality probability", xlab=expression("Annual dbh growth"~("cm yr"^-1))) {
+
+
   preds <- predict_spp_hazard(model, data, growth_range)
   
   breaks <- c(0.0001,0.001,0.01,0.1, 1, 10, 100)
@@ -55,7 +57,7 @@ plot_spp_curves <- function(model, data, growth_range= c(0.03,0.5), hazard_curve
       scale_y_continuous(expand=c(0,0), limits = c(0, NA)) +
       ylab(ylab) +
       xlab(xlab) +
[…]

### 2017-05-31 ms/ms.Rnw
Minor edits from James

--- 2016-10-29 ms/ms.Rnw
@@
 \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases.
-Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality rate decaying exponentially until an asymptote at zero is reached.
+Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality rate decaying exponentially towards an asymptote at zero.
 The third model (eq.
@@
 Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations.
-For example, some of our models took over a two weeks to complete.
-We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increasing model complexity (see methods).
+For example, some of our models took over two weeks to complete.
+We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increased model complexity (see methods).
 
@@
 A) Five sequential stages of model selection with each stage increasing in complexity.
-Null: null model ($\gamma$); Census: inclusion of census effects ($\delta$); Growth rate: the inclusion of growth rate.
-This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{\beta X_{i}})\delta$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{\beta X_{i}} + \gamma)\delta$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
+Null: null model ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate.
+This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
 B) Change in logloss under all wood density\textendash parameter combinations.
@@
 We found that the hazard function that incorporates both growth-dependent and baseline hazard terms (i.e.
-$(\alpha e^{\beta X_{i}} + \gamma)\delta$), significantly outperformed both the baseline only (i.e.$\gamma \delta$) or growth-dependent only (i.e.
-$(\alpha e^{\beta X_{i}})\delta$) hazard functions (Fig.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$), significantly outperformed both the baseline only (i.e.$\gamma \delta_t$) or growth-dependent only (i.e.
+$(\alpha e^{-\beta X_{i}})\delta_t$) hazard functions (Fig.
 \ref{fig2}A).
 Moreover, we found that predictive accuracy was higher for stem diameter growth relative to basal-area growth.
-Perhaps surprisingly, we detected no significant improvement in predictive accuracy between the baseline only and growth-dependent only hazard functions.
+Perhaps surprisingly, we detected no significant improvement in predictive accuracy when examining models that only accounted for baseline or growth-dependent only hazards.
 
@@
 \ref{fig3}).
-This meant that fast growing individuals from a low wood density species had on higher average mortality rates ((Fig.
+This meant that fast growing individuals from a low wood density species had on higher average mortality rates (Fig.
 \ref{fig4} A,B), and thus higher annual probability of death (Fig.
@@
 \subsection*{Mortality rates, maximum species size and species light requirements}
-We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were positively correlated with gap index (our measure of species light requirement; Fig.
+We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were weakly and positively correlated with gap index (our measure of species light requirement; Fig.
 \ref{fig6} A,C, E).
@@
 This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure.
-In stage 3, we begin to examined inter-specific variability by using the best model of stage 2 but allowing various combinations of parameters to vary as a function of species wood density.
+In stage 3, we examined inter-specific variability by using the best model of stage 2 but allowing various combinations of parameters to vary as a function of species wood density.
 This allowed us to determine if wood density improved predictive capacity, and if so, whether it predominately improved predictive capacity via growth-dependent or growth-independent processes.

### 2017-06-01 ms/ms.Rnw
Finished draft
- length of main text now within allowed 6 pages
- reworked final part of introduction, merging sections from the list of questions and methods into single para.
- shortened methods and added detail
- added intro to Supp Matt
- added details on numerical optimisations to SuppMatt

--- 2017-05-31 ms/ms.Rnw
@@
 \end{equation}
-Since trees are long-lived and we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required.
+Since trees are long-lived and we are trying to estimate the shape of a continuous hazard function from binary data, large sample-sizes are required.
 As such, detailed studies of tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals\cite{Anderson-Teixeira-2015a}.
@@
 Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate\cite{Kobe:1995tw,Wyckoff:2002ul}, traits\cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size\cite{Vieilledent:2010fv, Hurst:2011ky}.
-Generally, however, these influences have not been integrated into a common hazard function\cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.
+However, these influences have not yet been integrated into a common hazard function\cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.
 
@@
 The second are growth-independent hazards, where plants die because of stochastic events, irrespective of their growth rate, such as  windfall or fire.
-To capture this distinction, we assume an individual's hazard is the sum of growth-independent ($\lambda_{\textrm{b},i}(t)$) and growth-dependent ($\lambda_{\textrm{g},i}(t)$) rates:
-\begin{equation} \label{eq:additive}
-\lambda_i(t)  = \lambda_{\textrm{b},i}(t) + \lambda_{\textrm{g},i}(t).
-\end{equation}
+We therefore assume an individual's hazard is the sum of growth-independent and growth-dependent rates.
 
@@
 Within species, mortality rates are lower for fast-growing individuals, presumably because those individuals have superior carbon budgets and are thus able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
-Empirical studies broadly support this theory, with many indicating exponential declines in mortality with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, such that with parameters $\alpha, \beta$:
-\begin{equation} \label{eq:gd}
-\lambda_{\textrm{g},i}(t) = \alpha \textrm{e}^{-\beta X_i(t)}.
-\end{equation}
-
+Empirical studies broadly support this theory, with many indicating exponential declines in mortality with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}.
 By contrast, across species, there is a strong trade-off between growth and survival, with individuals from faster-growing species  exhibiting higher mortality rates than individuals from slower-growing species\cite{Poorter:2008iu, Wright:2010fl}.
 This trade-off may arise via traits with antagonistic effects, such as wood density.
-Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage\cite{vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack\cite{Augspurger:1984wx}, and ultimately reduced mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+Denser wood -- which is expensive to build and thus slows growth --  reduces stem breakage\cite{vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, pathogen attack\cite{Augspurger:1984wx}, and ultimately mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
 
@@
 In this study, we attempt to reconcile intra-specific and inter-specific factors as well as partition instantaneous mortality rates into growth-dependent and growth-independent rates.
-We achieve this by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework.
-To derive this unified model, we compare the skill of three alternative hazard functions ($\lambda_i(t)$; Fig.
-\ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama.
-Beginning from eq.\ref{eq:additive}, the simplest model (Fig.
-\ref{fig1}B, left), assumes a single baseline, growth-independent hazard rate.
-The second model (Fig.
-\ref{fig1}B, middle) assumes the risk of dying eventually declines to zero as growth increases (eq.
-\ref{eq:gd}).
-The third model (eq.
+We achieve this by evaluating the following, unified hazard function, incorporating individual-, species-, and census-level effects:
+\begin{equation} \label{eq:model}
+\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s[i]}}^{\text{Growth independent}} + \overbrace{\alpha_{s[i]} \, \textrm{e}^{-\beta_{s[i]} \, X_i(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
+\end{equation}
+Eq.
+\ref{eq:model} allows for additive growth-independent and growth-dependent effects, includes a negative exponential effect of growth rate, and allows for mortality to vary temporally among censuses, via the random effect $\delta_{t}$.
+Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ are allowed to vary by species.
+Here we include an effect of a species trait (wood density, $\rho$), as well as a species random effect that captures any additional species-level differences not represented $\rho$ (for details see Methods).
+
+To validate this model, we fit a series of models with increasing complexity, with eq.
+\ref{eq:model} being the most complex, and compare their skill in predicting patterns of tropical tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_unique)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_unique)} tropical rainforest species at Barro Colorado Island (BCI), Panama (Fig.
+\ref{fig1}).
+The data are repeat censuses of stem diameter and tree status (alive vs dead) taken over a 15 yr period (Fig.
+\ref{fig1}A).
+We compare the skill of different hazard functions in predicting outcomes in novel data (i.e.
+not used in model calibration) via 10-fold cross-validation (Fig.
+\ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}.
+Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations.
+We therefore fit models across five iterative stages of model development.
+At each stage, the best model from the previous stage was taken as input for the next stage.
+
+In stage 1 we ask whether mortality rates vary substantially between censuses, so that a census effect is needed?
+The null model is simply a constant, invariant with respect to species, growth rate, or year.
+In stage 2 we assess whether a hazard function including both growth-dependent and independent terms outperform a function including only one of these?
+Species effects were excluded, so we are simply asking which of three functional forms for $\lambda_i(t)$ (Fig.
+\ref{fig1}B) best represents the data.
+The simplest form (Fig.
+\ref{fig1}B, left) assumes a constant growth-independent hazard rate.
+The second form (Fig.
+\ref{fig1}B, middle) assumes the risk of dying eventually declines to zero as growth increases.
+The third form (eq.
 Fig.
@@
 Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared.
-
-To derive the unified model, we asked the following questions:
-1) Does mortality rates vary substantially between censuses, such that a specific census effect is needed?
-2) Does a hazard function including both growth-dependent and independent terms outperform a function including only one of these?
-3) Which growth measure (stem diameter or stem area increment) more skilfully predicts growth-dependent mortality?
-4) Does including wood density (a species trait) improve model skill, and if so, does it affect parameters of growth-dependent or independent hazard or both?
-5) What proportion of inter-specific variability in hazard is explained by wood density?
-6) What proportion of total variability in hazard is explained by species, census, growth-dependent and growth-independent hazards?
-
-Answering these questions involves comparing the  predictive skill of alternative models.
-To achieve this, we used 10-fold cross-validation (Fig.
-\ref{fig1}D), to systematically assesses each model's skill in predicting outcomes in novel data (i.e.
-not used in model calibration) \cite{Hooten:2015, Stone:1997tg}.
-Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations.
-We therefore fit models across five iterative stages of model development, corresponding to the six questions evaluated above, of gradually increased complexity (see Methods).
+We also investigate which growth measure (stem diameter or stem area increment) more skilfully predicts growth-dependent mortality.
+In stage 3, we examined whether including wood density (a species trait) improved model skill, and if so, whether the effect of wood density was primarily on growth-dependent or independent hazards, or both?
+Finally, we fit a model allowing other species-level variation in parameters, via a random effect.
+This allowed us to ask what proportion of inter-specific variability in hazard is explained by wood density?
+Using this final model we also conducted some posthoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.
 
@@
 \subsection*{Mortality over time}
-Comparing the 3 intervals between censuses from 1995 to 2010, we found tree mortality progressively decreasing over time.
+Comparing the three intervals between censuses from 1995 to 2010, we found tree mortality progressively decreasing over time.
 The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%).
 Consequently, when we allowed hazard rates to be scaled by census (i.e.
-adding term $\delta$) we observed a significant increase in predictive skill (i.e.
-Fig.
+adding term $\delta$) we observed a significant increase in predictive skill (Fig.
 \ref{fig2}A).
@@
 \subsection*{Growth hazard functions}
-We found that a hazard function with both growth-dependent and growth-independent terms (i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$), significantly outperformed both the baseline only (i.e.$\gamma \delta_t$) or growth-dependent only (i.e.
-$(\alpha e^{-\beta X_{i}})\delta_t$) functions (Fig.
+Comparing the three hazard functions in Fig.
+\ref{fig1}B, we found that the third function -- with both growth-dependent and growth-independent terms, i.e.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ -- significantly outperformed both the baseline only or growth-dependent only functions (Fig.
 \ref{fig2}A).
-Moreover, we found that predictive skill was higher when using stem diameter growth over basal-area growth.
+Moreover, we found that predictive skill was higher when using stem diameter growth over stem-area growth (Fig.
+\ref{fig2}A).
 
@@
 species and census; Fig.
-\ref{fig5}A), we found that approximately \Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of variation in predicted mortality rates was due to species differences, \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of which was captured by wood density.
+\ref{fig5}A), we found that approximately \Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of variation in predicted mortality rates was due to species differences, only \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of which was captured by wood density.
 By contrast, census effects accounted for approximately \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation in mortality rates.
@@
 
-\begin{figure}[h!]
+\begin{figure}[t]
 \includegraphics{figures/fig5.pdf}
@@
 competition, parasitization, localized herbivory).
-By contrast, the growth-independent hazard (i.e.
-baseline hazard) accounts for deaths caused by stochastic events that arise irrespective of an individuals growth rate (e.g.
+By contrast, the growth-independent hazard accounts for deaths caused by events that arise irrespective of an individuals growth rate (e.g.
 windthrow).
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
 \includegraphics{figures/fig4.pdf}
@@
 This means that such models are likely to severely underestimate the true variability in mortality rates.
-Consequently, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics, with flow on effects for ecosystem properties such as GPP and NPP\cite{Johnson-2016}.
+Consequently, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.
 
-\begin{figure}[h!]
+\begin{figure}[t]
 \includegraphics{figures/fig6.pdf}
@@
 
-The high unexplained inter-specific variation indicates that there are other species traits, besides wood density, that affect tree mortality rates.
+We found a high level of inter-specific variation was unexplained by wood density, suggesting here are other traits affecting tree mortality rates.
 This unexplained variation may possibly be explained by a species' light demand and maximum height, which has recently been proposed as two major axes of inter-specific variation in tropical rainforests\cite{Poorter:2006tq}.
@@
 By contrast, we detected no correlations between a species maximum stem diameter and mortality rates, both contradicting\cite{Poorter:2006tq} and supporting\cite{Wright:2010fl} findings by others.
-
 We recommend future research should focus on examining how other traits influence growth-dependent and growth-independent hazards and what combination of traits minimises unexplained species variability.
-Having said this, researchers should be mindful that the inclusion of additional traits comes at the cost of complexity, increased uncertainty (i.e.
-more parameters to estimate) and increased minimum knowledge; all of which diminish the interpretability of models, but also the applicability of predicting mortality rates in many data poor ecosystems.
-Because of this, we also recommend future research focus on developing more integrative correlates of plant mortality.
-For example, we may have captured more inter-specific variability in baseline hazards had we used a integrative measure of plant durability such as a trait that incorporates both wood density and stem architecture.
 
@@
 Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010.
-Unfortuantely, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3).
-However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by an El-Nino in 1997\textendash 1998.
-Droughts are known to have significant impacts on tree mortality rates in these ecosystems\cite{Condit:1995wl, Mcdowell:2013wf}.
+Unfortunately, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3).
+However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by an El-Nino in 1997\textendash 1998 \cite{Condit:1995wl, Mcdowell:2013wf}.
 
@@
 
-Here we showcase a new framework for modelling tropical tree mortality that attempts to unifies empirical evidence from within and between species studies.
+Here we showcase a new framework for modelling tropical tree mortality that unifies empirical evidence from within and between species studies.
 This framework also provides an approach for partitioning mortality rates into growth-dependent and growth-independent hazards.
@@
 \subsection*{Data}
-We derived plant mortality models using individual growth and survival data\cite{Condit:2012nz} collected from a relatively undisturbed 50-ha tropical rainforest plot on Barro Colorado Island (BCI), Panama (9.15\textdegree;N, 79.85\textdegree;W).
-The climate on the island is warm and rainfall is seasonal with falling between April and November.
-BCI is managed exclusively for field research by the Smithsonian Tropical Research Institute.
+We derived plant mortality models using individual growth and survival data collected from a relatively undisturbed 50-ha tropical rainforest plot on BCI, Panama (9.15\textdegree;N, 79.85\textdegree;W).
+The climate on the island is warm and rainfall is seasonal with most falling between April and November \cite{Condit:2012nz}.
 
-Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981-1983, 1985, and every 5 years thereafter\cite{Condit:2012nz}.
-For the purpose of modelling mortality as a function of past growth, we discarded data collected in 1982 and 1985 because diameter measurements were rounded to the nearest 5 mm, whereas later censuses were rounded to the nearest millimetre.
-Consequently, we modelled tree mortality as a function of past growth for censuses 1995\textendash 2000, 2000\textendash 2005 and 2005\textendash 2010.
+Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981\textendash1983, 1985, and every 5 years thereafter\cite{Condit:2012nz}.
+For the purpose of modelling mortality as a function of past growth, we discarded data collected prior to 1990 because diameter measurements were rounded to the nearest 5 mm, whereas later censuses were rounded to the nearest millimetre.
+Consequently, we modelled tree mortality as a function of past growth for censuses 1995\textendash2000, 2000\textendash2005 and 2005\textendash2010.
 We discarded species that do not exhibit secondary growth (e.g.
 palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density.
-We also excluded individuals that: 1) did not survive at least two censuses (two being required to estimate growth rate); 2) were not consistently measured at 1.3 m above ground; 3) were multi-stemmed; 4) resprouted or \textit{``returned from the dead''}; or 5) were extreme outliers -- stems which grew more than 5 cm/yr or shrunk more than 25\% of their initial diameter.
+We also excluded individuals that: 1) did not survive at least two censuses (two being required to estimate growth rate); 2) were not consistently measured at 1.3 m above ground; 3) were multi-stemmed; 4) resprouted or \textit{``returned from the dead''}; or 5) were extreme outliers -- stems which grew more than 5 cm yr$^{-1}$ or shrunk more than 25\% of their initial diameter.
 In order to maintain statical independence of measurements, we sub-sampled the dataset by randomly selecting one observation per individual out of a possible three censuses.
-In total, \Sexpr{n_inds(BCI_model_dataset_true_unique)} individuals of \Sexpr{n_spp(BCI_model_dataset_true_unique)} species were included in our analysis.
 
-Wood density for each species was estimated by coring trees located within 15 km of the BCI plot, because increment borers are prohibited within the plot itself\cite{Wright:2010fl}.
+Wood density for each species was estimated by coring trees located within 15 km of the BCI plot \cite{Wright:2010fl}.
 Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100\textdegree C) and dividing by the fresh volume (as measured by water displacement).
-Wood density was then calculated as an area-weighted average (g cm$^{-3}$), where area refers to the annulus represented by each piece, assuming a circular trunk.
-
-\subsection*{Estimating true growth}
-
-Field measurements of diameter inevitably include observation error, which effect estimates of growth rates.
-In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of estimated growth rates were negative.
-To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community.
-Details of this model can be found in Supplementary Material S1.
-
-\subsection*{Hazard functions}
-We fit a series of models with increasing complexity.
-At each stage, the best model from the previous stage was taken as input for the next stage.
-In stage 1 we fit the simplest model (a constant) and compared to a same where hazard rate varied by census.
-Stage 2 involved comparing predictive performance of the three functions described in (Fig.
-\ref{fig1}B) using either stem diameter or stem area growth rate.
-This allowed us to determine: 1) whether predictive performance was substantially improved with increasing functional complexity; and 2) whether predictive performance varied by growth measure.
-In stage 3, we examined inter-specific variability by using the best model of stage 2 but allowing various combinations of parameters to vary as a function of species wood density.
-This allowed us to determine if wood density improved predictive capacity, and if so, whether it predominately improved predictive capacity via growth-dependent or growth-independent processes.
-Stage 4 also examined inter-specific variability, however, here the focus was on using the best model of stage 2 and estimating the total predictive gain obtained if all parameters were allowed to vary as a function of species ID.
-The last stage was fitting the best model, with species random effects and wood density to the entire dataset.
-This allowed us to examine the range of species hazard curves, how they are affected by wood density, census and posthoc covariates, and the proportion of predicted variation explained by species, wood density, census and growth-dependent and baseline hazard terms.
 
@@
 
-For all models we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}.
+Eqs.
+\ref{eq:mortality_model}-\ref{eq:model} were fit to the data using Bayesian inference and with covariates for growth rate in previous census and wood density, as well as random effects.
+Growth rates were estimated from field measurements of diameter, which inevitably include observation error.
+In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of estimated growth rates were negative.
+To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see Supplementary Material S1 for details).
+The parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ were modelled as a function of both wood density (measured at species level) and a species-level random effect:
+\begin{equation}\label{eq:submodels}
+\alpha_s = \alpha_{0,s} \, (\frac{\rho}{\rho_c})^{\alpha_{1}},
+\end{equation}
+with similar formulations for $\beta_s$ and $\gamma_s$.
+Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$, modelled as random realisations from log-normal distributions.
+The form of eq.
+\ref{eq:submodels} ensures that parameters remain positive; and on a log scale this equates to an additive linear model centered around $\rho_c$.
+We also centered growth rate $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.
+Weak priors on all hyper-parameters were set (see Supplementary Material S2 for details).
+Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}.
 We executed three independent chains and in all cases modelled parameters converged within 2000 iterations.
@@
 
-\subsection*{Model evaluation}
+\subsection*{Evaluation model skill}
 Predictive skill was quantified by estimating the average log loss across 10-folds for held-out data, $\bar{\mathcal{L}}$ (Fig.
@@
 \subsection*{Posthoc correlations}
-
-In order to calculate our gap index (our measure of a species light dependence) we also used annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995.
-The canopy census recorded the presence of vegetation in six height intervals, 0–2, 2–5, 5–10, 10–20, 20–30 and 30 m every 5 m across the 50 ha.
+We calculated a gap index as a measure of a species' light dependence, using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995.
+The canopy census recorded the presence of vegetation in six height intervals, 0\textendash2, 2\textendash5, 5\textendash10, 10\textendash20, 20\textendash30 and 30 m every 5 m across the 50 ha.
 We calculated the minimum observed number of vegetation strata above 2 m all 5x5 cells in each census period.
-We then transformed this to a binary gap index by where 1 refers to no vegetation above 2 m and 0 referring to densely vegetated by subtracting the above values from 5 (i.e., 0 = dense, 5 = gap).
-To acknowledge the influence of gaps on surrounding cells' light availability, we then calculated a weighted focal sum using a 3 x 3 cell moving window, for which the central cell was assigned a weight of 8, and the eight neighbouring cells were each assigned weights of 1.
+We then transformed this to a binary gap index by where 1 refers to no vegetation above 2 m and 0 referring to densely vegetated, by subtracting the above values from 5 (i.e., 0 = dense, 5 = gap).
+To account for the influence of gaps on surrounding cells' light availability, we then calculated a weighted focal sum using a 3 x 3 cell moving window, for which the central cell was assigned a weight of 8, and the eight neighbouring cells were each assigned weights of 1.
 The combined impact of vegetation density of neighbouring cells on light availability at the central cell was therefore assumed to be equivalent to the impact of vegetation density at the central cell.
@@
 J.C was supported by the Science and Industry Endowment Fund (SIEF; RP04-174).
-M.W was supported by a fellowship from the Australian Research Council (ARC).
 D.F.
-was supported by an ARC discovery grant.
-We thank Rich FitzJohn, Lachlan McCalman and Daniel Steinberg for computational advice; Anias Gibert etc etc.
+and M.W were supported by fellowships from the Australian Research Council (ARC).
+We thank Rich FitzJohn, Lachlan McCalman and Daniel Steinberg for computational advice.
 Fig.
-1a uses images by Tracey Saxby, IAN Image Library (http://ian.umces.edu/imagelibrary)}
+1a uses images by Tracey Saxby, IAN Image Library.}
 
@@
 
-% \pnasbreak splits and balances the columns before the references.
-% If you see unexpected formatting errors, try commenting out this line
-% as it can run into problems with floats and footnotes on the final page.
-%\pnasbreak
+\vspace{-5em}
+\pnasbreak
 

### 2017-06-01 ms/supp.Rnw
Finished draft
- length of main text now within allowed 6 pages
- reworked final part of introduction, merging sections from the list of questions and methods into single para.
- shortened methods and added detail
- added intro to Supp Matt
- added details on numerical optimisations to SuppMatt

@@
 
-Our model is presented in detail in ...... Some intro text.
+Here we provide some further details on methods and additional results, supplementing those presented in the main text.
+Section S1-Estimating True Growth describes a probabilistic methods used to estimate \textit{``true growth''} from field measurements,, taking into account measurement error and the distribution of growth rate across the community.
+Section S2-Parameter distributions describes the distributions applied to species-level random effects and priors put over parameters in the fitted model.
+Section S3-Numerical optimisation outlines some numerical concerns in estimating logloss, while Section S4-Example Stan code gives an example of the code fitted to the data.
+Code and data reproducing the results in this paper are available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.
 
@@
 
-To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e.
-observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors\cite{Condit:2012nz}.
+ Growth rates were estimated from field measurements of diameter, which inevitably include observation error.
+In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of estimated growth rates were negative.
+To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community.
+To achieve this we used another dataset collected at BCI to explicitly to examine observation errors\cite{Condit:2012nz}.
 This dataset consisted of \Sexpr{nobs_meas_err_data(BCI_dbh_error_data)} randomly selected trees that had their diameter measured twice within a 30 day period\cite{Condit:2012nz}.
@@
 
-\section{Mortality model}
-
-We modelled tree mortality using survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$.
-We model these observations as random realizations from a Bernoulli distribution
-\begin{equation}
-S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}),
-\end{equation}
-where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$.
-As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
-\begin{equation}
-  \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t \approx \lambda_i(t_2) (t_3-t_2).
-\end{equation}
-
-As we have multiple hypothesized functional forms of $\lambda_i$, we describe here the fitting procedure of our most complicated model, of which all others may be considered special cases.
-The instantaneous rate of mortality is
-\begin{equation}
-\lambda_i = \left(\overbrace{\alpha_{s[i]} \, \exp\left(-\beta_{s[i]} \, X_i \right)}^{\text{Growth dependent}} +
-\overbrace{\gamma_{s[i]}}^{\text{Growth independent}}\right) \times \overbrace{\delta_{t[i]}}^{\text{Census effect}}
-\end{equation}  
-In this model, $\alpha_s$, $\beta_s$ and $\gamma_s$ are allowed to vary by species as a function of both wood density (measured at species level) and a species random effects, e.g.
-\begin{equation}\label{eq:submodels}
-\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{\rho_c}\right)^{\alpha_{1}},
-\end{equation}
-with similar formulations for $\beta_s$ and $\gamma_s$.
-Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$ within which individual $i$ belongs.
-The form of sub-model in eq.
-\ref{eq:submodels} ensures that parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ remain positive.
-On a log scale this model equates to an additive linear model , i.e.
-\begin{equation}\label{eq:submodels_log}
-\log \alpha_s = \log \alpha_{0,s} + (\alpha_{1} - \log \rho_c) \, \log \rho.
-\end{equation}
-
-We standardized wood density via the transformation: $\frac{\rho}{\rho_c}$, where $\rho_c=0.6$ g cm$^{-3}$.
-This operation is equivalent to centering a log-transformed $\rho$ around $\rho_c$ and enables mean hyper-parameters  to be interpreted as the value for a species with average wood density of 0.6 cm$^{-3}$.
-Similarly, we also centered $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.
+\section{Parameter distributions}
 
@@
 \section{Numerical optimisation}
+
+To evaluate eq.
+5 from the main text (log loss), we apply the following substitutions, to reduce the number of $\log$ and $\exp$ operations, and thereby maintain numerical accuracy.
+From the main text, equation 5 states that the log loss for a single data point is
+\begin{equation} \label{eq:logloss_i}
+\mathcal{L}_i = S_i\,\log(p_{i,t_2\rightarrow t_3}) + ( 1 - S_i)\,\log(1 - p_{i,t_2\rightarrow t_3}),
+\end{equation}
+where $p_{i,t_2\rightarrow t_3} = 1 - \exp\left(-\Lambda\right)$ is the probability an individual dies between times $t_2$ and $t_3$ (from eq.
+1), $\Lambda$ is the cumulative hazard from $t_2\rightarrow t_3$, and $S_i$ is the observed outcome (0 = alive, 1 = died).
+Note that $\textrm{Pr}(S_i =0) = 1- p_{i,t_2\rightarrow t_3} = \exp\left(-\Lambda\right)$.
+Numerical calculations of log loss can then be simplified as follows:
+
+If $S_i =0 \rightarrow \mathcal{L}_i = \log(1-p) = -\Lambda$.
+
+If $S_i = 1  \rightarrow \mathcal{L}_i = \log(p) = \textrm{log1m\_exp}(-\Lambda)$.
+
+\noindent In Stan, the function $ \textrm{log1m\_exp}$ is is defined so that $ \textrm{log1m\_exp}(x)$ has the same value as $\log(1.0-\exp(x))$, but the computation is faster, more accurate, and more stable \cite{StanDevelopmentTeam:2015uz}.
 

### 2017-06-01 ms/ms.Rnw
Minor edits

@@
 Combined $\alpha$ and $\beta$ capture growth-dependent mortality, while $\gamma$ captures growth-independent mortality (e.g.
-windfall, fire) that kill a plant irrespective of its growth rate.
+windfall, fire) that kill a plant, irrespective of its growth rate.
 For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
@@
 
-In stage 1 we ask whether mortality rates vary substantially between censuses, so that a census effect is needed?
+In stage 1 we ask whether mortality rates vary substantially between censuses to establish whether a census effect is required.
 The null model is simply a constant, invariant with respect to species, growth rate, or year.
@@
 The second form (Fig.
-\ref{fig1}B, middle) assumes the risk of dying eventually declines to zero as growth increases.
+\ref{fig1}B, middle) assumes the risk of dying eventually declines towards an asymptote of zero as growth increases.
 The third form (eq.
 Fig.
-\ref{fig1}B, right) is the summation of the two previous models, which allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, asymptoting at some baseline hazard $\geq 0$.
+\ref{fig1}B, right) is the summation of the two previous models, which allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, asymptoting at some baseline hazard $\textgreater 0$.
 Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared.
 We also investigate which growth measure (stem diameter or stem area increment) more skilfully predicts growth-dependent mortality.
-In stage 3, we examined whether including wood density (a species trait) improved model skill, and if so, whether the effect of wood density was primarily on growth-dependent or independent hazards, or both?
-Finally, we fit a model allowing other species-level variation in parameters, via a random effect.
-This allowed us to ask what proportion of inter-specific variability in hazard is explained by wood density?
-Using this final model we also conducted some posthoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.
+In stage 3, we examined whether including wood density (a species-level trait) improved model skill, and if so, whether the effect of wood density was primarily on growth-dependent or growth-independent hazards, or both?
+Finally, we fit a model which allowed parameters to vary by species-level variation that was otherwise not captured by wood density.
+This allowed us to ask what proportion of inter-specific hazard variability is explained by wood density.
+Using the final 'best' model we also conducted some posthoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.
 
@@
 \subsection*{Mortality over time}
-Comparing the three intervals between censuses from 1995 to 2010, we found tree mortality progressively decreasing over time.
+Comparing the three 5-year intervals between censuses from 1995 to 2010, we found tree mortality progressively decreasing over time.
 The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%).
@@
 
-We found a high level of inter-specific variation was unexplained by wood density, suggesting here are other traits affecting tree mortality rates.
+We found a high level of inter-specific variation was unexplained by wood density, suggesting that other traits are affecting tree mortality rates.
 This unexplained variation may possibly be explained by a species' light demand and maximum height, which has recently been proposed as two major axes of inter-specific variation in tropical rainforests\cite{Poorter:2006tq}.
@@
 We also excluded individuals that: 1) did not survive at least two censuses (two being required to estimate growth rate); 2) were not consistently measured at 1.3 m above ground; 3) were multi-stemmed; 4) resprouted or \textit{``returned from the dead''}; or 5) were extreme outliers -- stems which grew more than 5 cm yr$^{-1}$ or shrunk more than 25\% of their initial diameter.
-In order to maintain statical independence of measurements, we sub-sampled the dataset by randomly selecting one observation per individual out of a possible three censuses.
+In order to maintain statistical independence of measurements and avoid estimating 180,500 individual parameters (computationaly inviable), we sub-sampled the dataset by randomly selecting one observation per individual out of a possible three censuses.
 

### 2017-06-01 ms/ms.Rnw
Minor edits

@@
 \templatetype{pnasresearcharticle} % Choose template
-% {pnasresearcharticle} = Template for a two-column research article
-% {pnasmathematics} = Template for a one-column mathematics article
-% {pnasinvited} = Template for a PNAS invited submission
-
-% Enable cross referencing to Evidence document
-\usepackage{xr}
-\externaldocument{supp}
 
@@
 
-% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
 \author[a,1]{James S Camac}
@@
 
-% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
 \affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
@@
 
-% Please give the surname of the lead author for the running footer
 \leadauthor{Camac}
@@
 
-% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
+% Please include two to five keywords, separated by the pipe symbol
 \keywords{Bayesian $|$ Growth rates $|$ Hazard rates $|$ Tropical trees $|$ Wood density}
@@
 \ref{fig1}).
-The data are repeat censuses of stem diameter and tree status (alive vs dead) taken over a 15 yr period (Fig.
+The data are repeat censuses of stem diameter and tree status (alive vs.
+dead) taken over a 15 yr period (Fig.
 \ref{fig1}A).
@@
 \section*{Discussion}
-Our Bayesian framework coupled with cross validation revealed that the most accurate and parsimonious model of tropical tree death was that which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) attributed the effect of wood density to growth-independent mortality; and 4) incorporated temporal varability.
+Our Bayesian framework coupled with cross validation revealed that the most accurate and parsimonious model of tropical tree death was that which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) attributed the effect of wood density to growth-independent mortality; and 4) incorporated temporal variability.
 Moreover, we found that rates of tropical tree mortality varied substantially between species and that wood density, a species level functional trait, explained only a small proportion of the overall inter-specific variation.
@@
 This is because the additive effects of both hazards explicitly allow models to account for two independent causes of plant mortality.
-The growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stressors (e.g.
-competition, parasitization, localized herbivory).
+The growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses (e.g.
+competition, parasites, herbivory).
 By contrast, the growth-independent hazard accounts for deaths caused by events that arise irrespective of an individuals growth rate (e.g.
@@
 We also excluded individuals that: 1) did not survive at least two censuses (two being required to estimate growth rate); 2) were not consistently measured at 1.3 m above ground; 3) were multi-stemmed; 4) resprouted or \textit{``returned from the dead''}; or 5) were extreme outliers -- stems which grew more than 5 cm yr$^{-1}$ or shrunk more than 25\% of their initial diameter.
-In order to maintain statistical independence of measurements and avoid estimating 180,500 individual parameters (computationaly inviable), we sub-sampled the dataset by randomly selecting one observation per individual out of a possible three censuses.
+In order to maintain statistical independence of measurements and avoid estimating 180,500 individual parameters, we sub-sampled the dataset by randomly selecting one observation per individual out of a possible three censuses.
 
@@
 In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of estimated growth rates were negative.
-To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see Supplementary Material S1 for details).
+To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see Supplementary Material S1 for details; Fig.
+S1).
 The parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ were modelled as a function of both wood density (measured at species level) and a species-level random effect:
@@
 Weak priors on all hyper-parameters were set (see Supplementary Material S2 for details).
-Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}.
+Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz} and employing some numerical optimisations (see Supplementary Material S3-S4 for details and example code).
 We executed three independent chains and in all cases modelled parameters converged within 2000 iterations.
@@
 After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.
+Full distributions for the final model are shown in Figs.
+S3\textendash S5.
 
-\subsection*{Evaluation model skill}
+\subsection*{Evaluating model skill}
 Predictive skill was quantified by estimating the average log loss across 10-folds for held-out data, $\bar{\mathcal{L}}$ (Fig.
@@
 We calculated a gap index as a measure of a species' light dependence, using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995.
-The canopy census recorded the presence of vegetation in six height intervals, 0\textendash2, 2\textendash5, 5\textendash10, 10\textendash20, 20\textendash30 and 30 m every 5 m across the 50 ha.
-We calculated the minimum observed number of vegetation strata above 2 m all 5x5 cells in each census period.
-We then transformed this to a binary gap index by where 1 refers to no vegetation above 2 m and 0 referring to densely vegetated, by subtracting the above values from 5 (i.e., 0 = dense, 5 = gap).
-To account for the influence of gaps on surrounding cells' light availability, we then calculated a weighted focal sum using a 3 x 3 cell moving window, for which the central cell was assigned a weight of 8, and the eight neighbouring cells were each assigned weights of 1.
-The combined impact of vegetation density of neighbouring cells on light availability at the central cell was therefore assumed to be equivalent to the impact of vegetation density at the central cell.
+The canopy census recorded the presence of vegetation in six height intervals: 0\textendash2, 2\textendash5, 5\textendash10, 10\textendash20, 20\textendash30 and > 30 m, in all 5x5 m subplots across the 50 ha plot.
+For each subplot, we calculated the minimum number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap).
+We then calculated a weighted sum for each subplot, using a 3 x 3 cell moving window, where the central cell was assigned a weight of 8, and the neighbouring cells were assigned weights of 1.
+The gap index for each species was then calculated by averaging the light level encountered by new saplings as they appeared in the census (Fig.
+S2).
 

### 2017-06-01 ms/supp.Rnw
Minor edits

@@
 {\color{black50} Supporting information for:}\\} \vskip6pt
-\fontsize{16pt}{20pt} Unifying intraspecific and interspecific variation in tropical tree mortality
+\fontsize{16pt}{20pt} Unifying intra- and inter-specific variation in tropical tree mortality
 }
@@
 \author[b]{S Joseph Wright}
-\author[a,c]{Daniel S Falster}
+\author[c]{Daniel S Falster}
 
-% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
 \affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
@@
 We then used annual diameter growth, $\Delta D_{i,2}$, and annual basal area growth: $\frac{(\pi \, 0.25 \,\widehat{D}_{i,2}) - (\pi \, 0.25 \, \widehat{D}_{i,1})}{(t_2-t_1)}$ in the subsequent mortality analysis.
+Fig.
+\ref{figS1} illustrates observed and estimated diameters, and growth rate estimated from raw data including measurement error and from the probabilistic model.
 
@@
 \caption{Map of BCI 50 ha plot showing tree recruits (red crosses) against gap index for censuses 1985\textendash 1990 and 1990\textendash 1995.
-Colour gradient indicates the minimum amount of canopy shading above 2 m in a 5 year census.
+Colour gradient indicate the minimum amount of canopy shading above 2 m in a 5 year census.
 Dark colours indicate dense canopy shading, light colours indicate little or no canopy shading above 2 m.}
@@
 \includegraphics[height=0.8\textheight]{figures/supp_fig3}
-\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\alpha$) for each species. Red = Model estimated hyper-distribution species effects were sampled from}
+\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\alpha$) for each species. Red = Model estimated hyper-distribution species effects were sampled from.}
 \label{figS3}
@@
 \includegraphics{figures/supp_fig4}
-\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\beta$) for each species. Red = Model estimated hyper-distribution species effects were sampled from}
+\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\beta$) for each species. Red = Model estimated hyper-distribution species effects were sampled from.}
 \label{figS4}
@@
 \includegraphics{figures/supp_fig5}
-\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\gamma$) for each species. Red = Model estimated hyper-distribution species effects were sampled from}
+\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\gamma$) for each species. Red = Model estimated hyper-distribution species effects were sampled from.}
 \label{figS5}

### 2017-06-08 R/build.R
Add additional extensions to the latex clean functin

@@ -48,7 +48,7 @@ latex_build <- function(filename, bibliography=NULL,
 latex_clean <- function(filename) {
   filebase <- sub(".tex$", "", filename)
   exts <- c(".log", ".aux", ".bbl", ".blg", ".fls", ".out", ".snm",
-            ".nav", ".tdo", ".toc", ".fdb_latexmk")
+            ".nav", ".tdo", ".toc", ".fdb_latexmk",".xwm")
   aux <- paste0(filebase, exts)
   file.remove(aux[file.exists(aux)])
 }

### 2017-06-08 ms/ms.Rnw
Fixed author addresses and minor typos

@@
 
-\author[a,1]{James S Camac}
+\author[a,b,1]{James S Camac}
+\author[c]{Richard Condit}
+\author[d]{Richard G FitzJohn}
+\author[e]{Lachlan McCalman}
+\author[e]{Daniel Steinberg}
 \author[a]{Mark Westoby}
-\author[b]{S Joseph Wright}
-\author[a,c]{Daniel S Falster}
+\author[f]{S Joseph Wright}
+\author[a,g]{Daniel S Falster}
 
-\affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
-\affil[b]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
-\affil[c]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales}
+\affil[a]{Department of Biological Sciences, Macquarie University, New South Wales, Australia}
+\affil[b]{Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Victoria, Australia}
+\affil[c]{Field Museum of Natural History and Morton Arboretum, Chicago, Illinois, USA}
+\affil[d]{Imperial College, London, United Kingdom}
+\affil[e]{Data61, CSIRO, Australia}
+\affil[f]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
+\affil[g]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, NSW 2052, Australia}
 
@@
 % Please include corresponding author, author contribution and author declaration information
-\authorcontributions{J.C. performed research; D.S.F. conceived the idea; S.J.W. collected data. All authors contributed to the writing of this manuscript.}
+\authorcontributions{D.F. conceived idea; all authors designed research; J.C. \& R.F. performed analysis; R.C. \& S.J.W. collected data. J.C \& D.F. wrote the manuscript}
 
@@
 For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
-C) Log-loss function highlighting that the further the predicted probability is from truth the higher the logloss.
+C) Predictions for plant survival and death are compared to observed outcomes. Incorrect predictions are penalised via the log-loss function.
 D) The predictive skill of alternative models is evaluated via 10-fold cross validation.
@@
 This allowed us to ask what proportion of inter-specific hazard variability is explained by wood density.
-Using the final 'best' model we also conducted some posthoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.
+Using the final `best' model we also conducted some posthoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.
 

### 2017-06-08 ms/ms.Rnw
Diverse minor edits throughout
- adding units and details to axis labels
- tightening paragraphs to save a line, where possible
- more details where needed

@@
 \author[f]{S Joseph Wright}
-\author[a,g]{Daniel S Falster}
+\author[g]{Daniel S Falster}
 
@@
 However, wood density only accounted for a small fraction of the overall species variability observed in growth-independent mortality rates, suggesting that there must be other species traits that strongly affect mortality.
-Which traits and how many of them, however, remains unknown.}
+Which traits remains unknown.}
 
 % Please include corresponding author, author contribution and author declaration information
-\authorcontributions{D.F. conceived idea; all authors designed research; J.C. \& R.F. performed analysis; R.C. \& S.J.W. collected data. J.C \& D.F. wrote the manuscript}
+\authorcontributions{D.F. conceived idea; all authors designed research; J.C. \& R.F. performed analysis; R.C. \& S.W. collected data. J.C \& D.F. wrote the manuscript}
 
@@
 
-Data deposition: Code and data reproducing the results in this paper are available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.
+Data deposition: Code reproducing the results is available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.
+Data from BCI is available at DOIs \href{http://doi.org/10.5479/data.bci.20130603.}{10.5479/data.bci.20130603.}, \href{http://doi.org/10.5479/data.bci20140711}{10.5479/data.bci20140711}.
 }
@@
 \end{equation}
-Since trees are long-lived and we are trying to estimate the shape of a continuous hazard function from binary data, large sample-sizes are required.
-As such, detailed studies of tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals\cite{Anderson-Teixeira-2015a}.
+As trees are long-lived and we are trying to estimate the shape of a continuous hazard function from binary data, large sample-sizes are required.
+Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals\cite{Anderson-Teixeira-2015a}.
 
@@
 
-In this study, we attempt to reconcile intra-specific and inter-specific factors as well as partition instantaneous mortality rates into growth-dependent and growth-independent rates.
+Here, we attempt to reconcile intra- and inter-specific factors as well as partition instantaneous mortality rates into growth-dependent and growth-independent rates.
 We achieve this by evaluating the following, unified hazard function, incorporating individual-, species-, and census-level effects:
@@
 
-To validate this model, we fit a series of models with increasing complexity, with eq.
-\ref{eq:model} being the most complex, and compare their skill in predicting patterns of tropical tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_unique)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_unique)} tropical rainforest species at Barro Colorado Island (BCI), Panama (Fig.
+To validate this model, we fit a series of models with increasing complexity (eq.
+\ref{eq:model} being the most complex) and compare their skill in predicting patterns of tropical tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_unique)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_unique)} tree species at Barro Colorado Island (BCI), Panama (Fig.
 \ref{fig1}).
@@
 Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations.
-We therefore fit models across five iterative stages of model development.
+We therefore fit models across five stages of model development.
 At each stage, the best model from the previous stage was taken as input for the next stage.
 
-In stage 1 we ask whether mortality rates vary substantially between censuses to establish whether a census effect is required.
+In stage 1 we ask whether mortality rates vary substantially between censuses, to establish whether a census effect is required.
 The null model is simply a constant, invariant with respect to species, growth rate, or year.
-In stage 2 we assess whether a hazard function including both growth-dependent and independent terms outperform a function including only one of these?
+In stage 2 we ask whether a hazard function including both growth-dependent and independent terms outperforms a function including only one of these?
 Species effects were excluded, so we are simply asking which of three functional forms for $\lambda_i(t)$ (Fig.
@@
 The simplest form (Fig.
-\ref{fig1}B, left) assumes a constant growth-independent hazard rate.
+\ref{fig1}B, left) assumes a constant (growth-independent) hazard rate.
 The second form (Fig.
-\ref{fig1}B, middle) assumes the risk of dying eventually declines towards an asymptote of zero as growth increases.
+\ref{fig1}B, middle) assumes the risk of dying declines towards an asymptote of zero as growth increases.
 The third form (eq.
 Fig.
-\ref{fig1}B, right) is the summation of the two previous models, which allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, asymptoting at some baseline hazard $\textgreater 0$.
+\ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike B, asymptoting at some baseline rate $\textgreater 0$.
 Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared.
 We also investigate which growth measure (stem diameter or stem area increment) more skilfully predicts growth-dependent mortality.
-In stage 3, we examined whether including wood density (a species-level trait) improved model skill, and if so, whether the effect of wood density was primarily on growth-dependent or growth-independent hazards, or both?
-Finally, we fit a model which allowed parameters to vary by species-level variation that was otherwise not captured by wood density.
+In stage 3, we examined whether including wood density (a species-level trait) improved model skill, and if so, whether the effect of wood density was on growth-dependent or growth-independent hazard, or both?
+Finally, we fit a model that allowed parameters to vary by species-level effects not captured by wood density.
 This allowed us to ask what proportion of inter-specific hazard variability is explained by wood density.
@@
 \includegraphics{figures/fig2.pdf}
-\caption{Mean ($\pm$ 95\% credible intervals) logarithmic loss whereby lower values equal higher predictive skill.
+\caption{Comparing predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% credible intervals) logarithmic loss, with lower values implying greater skill.
 A) Five sequential stages of model selection with each stage increasing in complexity.
-Null: null model ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate.
-This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
-B) Change in logloss under all wood density\textendash parameter combinations.
+Null: constant ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate.
+This stage compares two hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and growth-independent + growth-dependent hazard ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
+B) Comparing predictive skill for wood density parameter combinations.
 Colors represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth.
-Symbols represent general functional form: null/baseline (circle), growth\textendash dependent hazard model (triangle) and model incorporating both baseline and growth\textendash dependent hazards (square).
-Note: errors are too small to see in panel A.}
+Symbols represent functional form: null and growth-independent (circle), growth-dependent hazard (triangle), both growth-independent and growth-dependent hazards (square).
+Errors in A are too small to see.}
 \label{fig2}
@@
 \ref{fig1}B, we found that the third function -- with both growth-dependent and growth-independent terms, i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ -- significantly outperformed both the baseline only or growth-dependent only functions (Fig.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ -- significantly outperformed both the growth-independent only or growth-dependent only functions (Fig.
 \ref{fig2}A).
@@
 \includegraphics{figures/fig3.pdf}
-\caption{Relationship between species baseline mortality rate and wood density.
+\caption{Relationship between species growth-independent mortality rate and wood density.
 Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study.
@@
 \subsection*{Effect of wood density on mortality rates}
-Including wood density ($\rho$) as an effect on either baseline or growth-dependent parameters significantly improved model skill relative to a hazard function without such effects (Fig.
+Including wood density as an effect on either growth-independent or growth-dependent parameters significantly improved model skill, relative to a model without such effects (Fig.
 \ref{fig2} A,B).
-The most parsimonious model, with highest predictive skill, was that which attributed the wood density effect to the baseline hazard term ($\gamma$; Fig.
+The most parsimonious model, with highest predictive skill, was that attributing the wood density effect to the growth-independent hazard term ($\gamma$; Fig.
 \ref{fig2} B).
-Wood density was therefore negatively correlated with a species' baseline mortality rate (Fig.
+Thus wood density was correlated with a species' growth-independent mortality rate (Fig.
 \ref{fig3}).
@@
 When we partitioned modelled variation into the two independent hazards we found that the growth-dependent hazard (i.e.
-the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the overall variation in mortality rates, and the remainder was attributed to the baseline hazard.
+the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the overall variation in mortality rates, and the remainder was attributed to the growth-independent hazard.
 
 \subsection*{Mortality rates, maximum species size and species light requirements}
-We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the baseline hazard term, were weakly and positively correlated with gap index (our measure of species light requirement; Fig.
+We found that $\alpha$, the effect of low growth rate, and $\gamma$, the growth-independent hazard term, were weakly and positively correlated with gap index, a measure of species light requirement (Fig.
 \ref{fig6} A,C, E).
-By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index.
+By contrast, $\beta$ (the exponential decay with increasing growth rate) was negatively correlated gap index.
 The combination of these correlations suggest that species which predominately recruit in gaps (i.e.
@@
 \includegraphics{figures/fig5.pdf}
-\caption{Proportion of predicted variation explained.
-A) Predicted variation explained by wood density (grey) as well as species and census random effects.
-B) Predicted variation explained by growth\textendash dependent and baseline (stochastic) hazards.}
+\caption{Proportion of predicted variation in 5-yr survival explained by different covariates.
+A) Wood density (grey) as well as total species and census random effects.
+B) Growth-dependent and growth-independent (baseline) hazards.}
 \label{fig5}
@@
 competition, parasites, herbivory).
-By contrast, the growth-independent hazard accounts for deaths caused by events that arise irrespective of an individuals growth rate (e.g.
-windthrow).
+By contrast, the growth-independent hazard accounts for deaths caused by events that arise irrespective of an individual's growth rate (e.g.
+windthrow, lightning strike).
 
@@
 
-How the length between censuses affect the estimation of growth-dependent and growth-independent hazards should also be examined.
-Large census intervals are likely to lead to underestimation of growth-dependent mortality, and consequently, overestimate the relative contribution of growth-independent hazards.
-This is because if a census interval is too wide, it will fail to account for deaths attributed to rapid declines in growth caused by senescence (especially for short-lived plants) or events such as drought.
+Whether the length between censuses biases estimation of growth-dependent and growth-independent hazards should also be examined.
+Long census intervals may lead to underestimation of growth-dependent mortality, and consequently, overestimate the relative contribution of growth-independent hazards.
+This is because if a census interval is too wide, it will fail to account for deaths attributed to rapid declines in growth caused (especially for short-lived plants) or events such as drought.
 Because our analysis estimated an individual's past growth from 5 year census intervals, and because it included short-lived species (i.e.
@@
 
-Lastly, research should also examine how different growth measures influence predictive skill of mortality models.
-We have no clear explanation as to why DBH growth outperformed basal area growth in terms of model predictive skill, and we are unsure how other common measures such as biomass growth\cite{Chave:2005hz} will compare.
+Lastly, research should also examine how different growth measures alter the predictive skill of mortality models.
+We found that DBH growth outperformed basal area growth, but have as yet  no clear explanation as to why, or how other common measures such as biomass growth\cite{Chave:2005hz} will compare.
 
@@
 \subsection*{Posthoc correlations}
-We calculated a gap index as a measure of a species' light dependence, using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995.
-The canopy census recorded the presence of vegetation in six height intervals: 0\textendash2, 2\textendash5, 5\textendash10, 10\textendash20, 20\textendash30 and > 30 m, in all 5x5 m subplots across the 50 ha plot.
-For each subplot, we calculated the minimum number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap).
-We then calculated a weighted sum for each subplot, using a 3 x 3 cell moving window, where the central cell was assigned a weight of 8, and the neighbouring cells were assigned weights of 1.
-The gap index for each species was then calculated by averaging the light level encountered by new saplings as they appeared in the census (Fig.
+We calculated a gap index as a measure of a species' light dependence, using canopy census data collected from 1985\textendash1995\cite{Hubbell:2014}.
+The canopy census recorded, in all 5x5 m subplots across the 50 ha plot, the presence of leaf in six height intervals between 0, 2, 5, 10, 20, 30 and > 30 m.
+For each subplot, we calculated the minimum number of strata > 2 m containing leaf; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap).
+We then calculated a weighted mean for each subplot, averaging each cell's value with mean of the 8 neighbouring cells.
+The gap index for each species is then the average light level encountered by new saplings appearing in the census (Fig.
 S2).
@@
 
-\acknow{
-J.C was supported by the Science and Industry Endowment Fund (SIEF; RP04-174).
+\acknow{We thank H.
+Muller-Landau and XX anonymous reviewers for feedback and B.
+Carpenter for technical advice.
+We acknowledge S.
+Hubbell, R.
+Foster, R.
+Pérez, S.
+Aguilar, S.
+Lao, S.
+Dolins, \& hundreds of field workers for their contribution; and the National Science Foundation, Smithsonian Tropical Research Institute, \& MacArthur Foundation for funding the design, collection, quality control and management of long-term growth data at BCI.
+Fig.
+1a uses images by Tracey Saxby, IAN Image Library.
+J.C., R.F., L.M.
+and D.S.
+were supported by the Science and Industry Endowment Fund (SIEF; RP04-174).
 D.F.
-and M.W were supported by fellowships from the Australian Research Council (ARC).
-We thank Rich FitzJohn, Lachlan McCalman and Daniel Steinberg for computational advice.
-Fig.
-1a uses images by Tracey Saxby, IAN Image Library.}
+and M.W.
+were supported by fellowships from the Australian Research Council.}
 

### 2017-06-08 ms/supp.Rnw
Diverse minor edits throughout
- adding units and details to axis labels
- tightening paragraphs to save a line, where possible
- more details where needed

@@
 
-% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
 \author[a,1]{James S Camac}
+\author[b]{Richard Condit}
+\author[c]{Richard G FitzJohn}
+\author[d]{Lachlan McCalman}
+\author[d]{Daniel Steinberg}
 \author[a]{Mark Westoby}
-\author[b]{S Joseph Wright}
-\author[c]{Daniel S Falster}
+\author[e]{S Joseph Wright}
+\author[f]{Daniel S Falster}
 
 \affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
-\affil[b]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
-\affil[c]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales}
+\affil[b]{Field Museum of Natural History and Morton Arboretum, Chicago, Illinois, USA}
+\affil[c]{Imperial College, London, United Kingdom}
+\affil[d]{Data61, CSIRO, Australia}
+\affil[e]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
+\affil[f]{Evolution and Ecology Research Centre, University of New South Wales, Australia}
 

### 2017-06-08 R/figures.R

@@ -108,7 +108,7 @@ plot_param_prop_explained <- function(param_prop_explained, ylab=NULL, xlab=NULL
            level = factor(level, levels=c("species","census")))
   ggplot(dat, aes(x=level, y=proportion)) +
     geom_bar(stat='identity', aes(fill=param), width=0.5) +
-    scale_x_discrete(labels=c('species','census')) +
+    scale_x_discrete(labels=c('species\n','census\n')) +
     scale_y_continuous(expand=c(0,0), limits=c(0,1)) +
     scale_fill_manual(values = c("wood_density" ="darkgrey","species" ="black", "census"="black")) + 
     ylab(ylab) +
@@ -121,7 +121,7 @@ plot_base_v_growth_prop_explained <- function(param_prop_explained, ylab=NULL, x
   dat <- filter(param_prop_explained, param %in% c("baseline", "growth_dependent")) %>% droplevels()
   ggplot(dat, aes(x=param, y=proportion)) +
     geom_bar(stat='identity', fill='black', width=0.5) +
-    scale_x_discrete(labels=c('growth dependent','baseline')) +
+    scale_x_discrete(labels=c('growth-\ndependent','growth-\nindependent')) +
     scale_y_continuous(expand=c(0,0), limits=c(0,1)) +
     ylab(ylab) +
     xlab(xlab) +
@@ -313,7 +313,7 @@ plot_fig1 <- function(tree1, tree2, panelc) {
   p <- seq(0,1, length.out=200)
   lines(p, -log(1-p), type='l', col="red")
   my_label("C) Penalty for incorrect prediction", x=-0.137)
-  mtext("Probability of outcome, p", 1, line=2.1, cex=0.5)
+  mtext("Probability of outcome", 1, line=2.1, cex=0.5)
   mtext("log loss", 2, line=2.1, cex=0.5)
 
   # Plot cross validation proceedure
@@ -378,12 +378,12 @@ plot_fig2a <- function(logloss_summaries) {
              comparison %in% c("null_model","function_growth_comparison","species_random_effects")) %>%
     mutate(comparison = replace(comparison, comparison =="function_growth_comparison" & model=="base_hazard", "census"),
       comparison = factor(comparison, levels=c('null_model','census','function_growth_comparison','rho_combinations','species_random_effects'),
[…]

### 2017-06-16 ms/ms.Rnw
Rearranged code so it was presented in a more logical order. Also renamed some functions so that they were more easily interpreted. Lastly made additional edits to manuscript

@@
 \author[f]{S Joseph Wright}
-\author[g]{Daniel S Falster}
+\author[a,g]{Daniel S Falster}
 
 \affil[a]{Department of Biological Sciences, Macquarie University, New South Wales, Australia}
-\affil[b]{Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Victoria, Australia}
+\affil[b]{Current Address: Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Victoria, Australia}
 \affil[c]{Field Museum of Natural History and Morton Arboretum, Chicago, Illinois, USA}
@@
 \affil[f]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
-\affil[g]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, NSW 2052, Australia}
+\affil[g]{Current Address: Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, NSW 2052, Australia}
 
@@
 \significancestatement{Tree mortality is a fundamental demographic process affecting forest dynamics and carbon cycling.
-Here, for the first time, we use observed survival records over a 15 year period from more than 180,000 individuals, to simultaneously estimate growth-dependent and growth-independent (i.e.
-stochastic) mortality across 203 tropical forest tree species.
+Here, for the first time, we use observed survival records over a 15 year period from more than 180,000 individuals, to simultaneously estimate growth-dependent and growth-independent mortality across 203 tropical forest tree species.
 We found that growth-dependent mortality was the predominate factor influencing tree mortality rates at Barro Colorado Island.
@@
 However, wood density only accounted for a small fraction of the overall species variability observed in growth-independent mortality rates, suggesting that there must be other species traits that strongly affect mortality.
-Which traits remains unknown.}
+Which traits, however, remains unknown.}
 
 % Please include corresponding author, author contribution and author declaration information
-\authorcontributions{D.F. conceived idea; all authors designed research; J.C. \& R.F. performed analysis; R.C. \& S.W. collected data. J.C \& D.F. wrote the manuscript}
+\authorcontributions{D.F.
+conceived idea; all authors designed research; J.C.
+\& R.F.
+performed analysis in discussion with L.M, D.S.
+\& M.W; R.C.
+\& S.W.
+collected data.
+J.C \& D.F.
+wrote the manuscript}
 
@@
 
-Data deposition: Code reproducing the results is available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.
+Data deposition: Code reproducing the results in this paper is available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.
 Data from BCI is available at DOIs \href{http://doi.org/10.5479/data.bci.20130603.}{10.5479/data.bci.20130603.}, \href{http://doi.org/10.5479/data.bci20140711}{10.5479/data.bci20140711}.
@@
 Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities.
-While past research has identified factors influencing tree mortality across a variety of scales, these distinct drivers are yet to be integrated within a unified, predictive framework.
+While past research has identified factors influencing tree mortality across a variety of scales, these distinct drivers are yet to be integrated within a unified predictive framework.
 In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive instantaneous mortality functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
-Specifically, we develop mortality functions that not only integrate individual, species and temporal effects, but also simultaneously estimate the relative contribution of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate.
+Specifically, we develop mortality functions that not only integrate individual, species and temporal effects, but also simultaneously estimate the independent contribution of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate.
 We show that functions that partition mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations.
-Moreover, we show that the effect of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability observed between species, and that temporal variation in mortality was large.
-Lastly, our analysis indicate that growth-dependent processes are the predominant contributor to rates of mortality at BCI.
+Moreover, we show that temporal variation in mortality is large and that the effect of wood density -- a prominent trait known to influence tree mortality -- explains only \Sexpr{extract_prop_explained(param_variance_explained,I('wood_density'))}\% of the total variability observed among species.
+Lastly, our analysis show that growth-dependent processes are the predominant contributor to rates of tree mortality at BCI.
 Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests.
-More importantly, it highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
+It also highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
 \end{abstract}
@@
 
-\dropcap{R}ates of plant mortality are known to be highly variable among individuals, coexisting species, between forests and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}.
+\dropcap{R}ates of plant mortality are known to be highly variable among coexisting species, between forests and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}.
 This variation has considerable consequences for forest structure and dynamics.
-For example, death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools\cite{Chave:2005hz}.
-Furthermore it creates a gap in the canopy that can restart a successional race, during which 10's-100's more plants may die while competing for a spot in the sun.
-In models of forest dynamics, variation in mortality rates have been found to have a larger impact on forest structure than variation in absolute growth rates\cite{Johnson-2016}.
+For example, death of a single large tree can transfer up to 20000 kg of carbon from living to decaying carbon pools\cite{Chave:2005hz}.
+Furthermore it creates a gap in the canopy that can restart a successional race, during which 100's-1000's more plants may die while competing for a spot in the sun.
+In models of forest dynamics, variation in mortality rates has been shown to have a larger impact on forest structure than variation in absolute growth rates\cite{Johnson-2016}.
 As such, improving our ability to understand the mortality process is a priority for making accurate predictions about forest and carbon dynamics as well as water or nutrient cycles; especially in an era of rapid environmental change.
@@
 This requirement arises in part because the observable outcome -- alive vs.
-dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$, also called the ``hazard function''.
+dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$, also called the ``hazard function''.
 In classic survival analysis\cite{Kleinbaum:2012tn}, the probability an individual dies between times $t_2$ and $t_3$ is a function of $\lambda_i(t)$:
@@
 \end{equation}
-As trees are long-lived and we are trying to estimate the shape of a continuous hazard function from binary data, large sample-sizes are required.
+As trees are long-lived and we are trying to estimate the shape of a continuous hazard function from binary data, large sample sizes are required.
 Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals\cite{Anderson-Teixeira-2015a}.
@@
 Eq.
-\ref{eq:model} allows for additive growth-independent and growth-dependent effects, includes a negative exponential effect of growth rate, and allows for mortality to vary temporally among censuses, via the random effect $\delta_{t}$.
+\ref{eq:model} allows for additive growth-independent and growth-dependent effects, includes a negative exponential effect of growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
 Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ are allowed to vary by species.
-Here we include an effect of a species trait (wood density, $\rho$), as well as a species random effect that captures any additional species-level differences not represented $\rho$ (for details see Methods).
+Here we include an effect of a species trait (wood density, $\rho$), as well as a species random effect that captures any remaining species-level differences not accounted for by wood density (for details see Methods).
 
@@
 We compare the skill of different hazard functions in predicting outcomes in novel data (i.e.
-not used in model calibration) via 10-fold cross-validation (Fig.
+not used in model fitting) via 10-fold cross-validation (Fig.
 \ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}.
 Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations.
-We therefore fit models across five stages of model development.
+We therefore fit models across five iterative stages of model development.
 At each stage, the best model from the previous stage was taken as input for the next stage.
@@
 The null model is simply a constant, invariant with respect to species, growth rate, or year.
-In stage 2 we ask whether a hazard function including both growth-dependent and independent terms outperforms a function including only one of these?
+In stage 2 we assess whether a hazard function including both growth-dependent and growth-independent terms outperform a function including only one of these.
 Species effects were excluded, so we are simply asking which of three functional forms for $\lambda_i(t)$ (Fig.
-\ref{fig1}B) best represents the data.
+\ref{fig1}B) best predicts the data.
 The simplest form (Fig.
-\ref{fig1}B, left) assumes a constant (growth-independent) hazard rate.
+\ref{fig1}B, left) assumes a constant growth-independent hazard rate.
 The second form (Fig.
@@
 Fig.
-\ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike B, asymptoting at some baseline rate $\textgreater 0$.
+\ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, asymptoting at some baseline hazard $\textgreater 0$.
 Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared.
 We also investigate which growth measure (stem diameter or stem area increment) more skilfully predicts growth-dependent mortality.
-In stage 3, we examined whether including wood density (a species-level trait) improved model skill, and if so, whether the effect of wood density was on growth-dependent or growth-independent hazard, or both?
-Finally, we fit a model that allowed parameters to vary by species-level effects not captured by wood density.
+
+In stage 3, we examined whether including wood density (a species-level trait) improved model skill, and if so, whether the effect of wood density was on growth-dependent, growth-independent, or both hazards.
+Finally, we fit a model that allowed parameters to vary by species-level variation that was otherwise not captured by wood density.
 This allowed us to ask what proportion of inter-specific hazard variability is explained by wood density.
-Using the final `best' model we also conducted some posthoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.
+Using the final `best' model we also conducted posthoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.
 
@@
 \includegraphics{figures/fig2.pdf}
-\caption{Comparing predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% credible intervals) logarithmic loss, with lower values implying greater skill.
+\caption{Comparing predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% credible intervals) logarithmic loss, with lower values implying greater predictive skill.
 A) Five sequential stages of model selection with each stage increasing in complexity.
-Null: constant ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate.
-This stage compares two hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and growth-independent + growth-dependent hazard ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
+Null: null model ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate.
+This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
 B) Comparing predictive skill for wood density parameter combinations.
-Colors represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth.
-Symbols represent functional form: null and growth-independent (circle), growth-dependent hazard (triangle), both growth-independent and growth-dependent hazards (square).
-Errors in A are too small to see.}
+Colors for both panels represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth.
+For both panels symbols represent functional form: null and growth-independent (circle), growth-dependent hazard (triangle), both growth-independent and growth-dependent hazards (square).
+Errors in panel A are too small to see.}
 \label{fig2}
@@
 \ref{fig1}B, we found that the third function -- with both growth-dependent and growth-independent terms, i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ -- significantly outperformed both the growth-independent only or growth-dependent only functions (Fig.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ -- significantly outperformed both the baseline only or growth-dependent only functions (Fig.
 \ref{fig2}A).
@@
 \includegraphics{figures/fig3.pdf}
-\caption{Relationship between species growth-independent mortality rate and wood density.
+\caption{Relationship between species growth-dependent mortality rate and wood density.
 Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study.
@@
 \subsection*{Effect of wood density on mortality rates}
-Including wood density as an effect on either growth-independent or growth-dependent parameters significantly improved model skill, relative to a model without such effects (Fig.
+Including wood density as an effect on either growth-independent or growth-dependent parameters significantly improved model skill relative to a model without such effects (Fig.
 \ref{fig2} A,B).
@@
 \ref{fig2} B).
-Thus wood density was correlated with a species' growth-independent mortality rate (Fig.
+Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate (Fig.
 \ref{fig3}).
-
-% This meant that fast growing individuals from a low wood density species had on higher average mortality rates (Fig.
+This meant that fast growing individuals from a low wood density species had, on average, higher mortality rates (Fig.
 \ref{fig4} A,B), and thus higher annual probability of death (Fig.
 \ref{fig4} C,D), relative to fast growing individuals of high wood density species.
+For example, the model predicted that a species with a wood density of 0.3 g $\textrm{cm}^{-3}$ had an estimated mean probability of dying of \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.3),2)['mean']} yr$^{-1}$ compared to \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.8),2)['mean']} yr$^{-1}$ for a species with a wood density of 0.8 g $\textrm{cm}^{-3}$.
+
+\begin{figure}[t]
+\includegraphics{figures/fig4.pdf}
+\caption{Predicted changes in instantaneous mortality rates (A,B), and annual mortality probability (C,D), with increasing growth rates.
+A,C) Estimated species curves for all 203 BCI species used in this study.
+B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species.}
+\label{fig4}
+\end{figure}
 
@@
 \subsection*{Mortality rates, maximum species size and species light requirements}
-We found that $\alpha$, the effect of low growth rate, and $\gamma$, the growth-independent hazard term, were weakly and positively correlated with gap index, a measure of species light requirement (Fig.
+We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the growth-independent hazard term, were weakly and positively correlated with gap index (our measure of species light requirement; Fig.
 \ref{fig6} A,C, E).
-By contrast, $\beta$ (the exponential decay with increasing growth rate) was negatively correlated gap index.
+By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index.
 The combination of these correlations suggest that species which predominately recruit in gaps (i.e.
@@
 \includegraphics{figures/fig5.pdf}
-\caption{Proportion of predicted variation in 5-yr survival explained by different covariates.
-A) Wood density (grey) as well as total species and census random effects.
+\caption{Proportion of predicted variation in 5-year survival explained by different covariates.
+A) Wood density (grey) as well as species and census random effects.
 B) Growth-dependent and growth-independent (baseline) hazards.}
@@
 \section*{Discussion}
-Our Bayesian framework coupled with cross validation revealed that the most accurate and parsimonious model of tropical tree death was that which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) attributed the effect of wood density to growth-independent mortality; and 4) incorporated temporal variability.
+Our Bayesian framework coupled with cross validation revealed that the most explanatory and parsimonious model of tropical tree death was that which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) attributed the effect of wood density to growth-independent mortality; and 4) incorporated temporal variability.
 Moreover, we found that rates of tropical tree mortality varied substantially between species and that wood density, a species level functional trait, explained only a small proportion of the overall inter-specific variation.
@@
 We show that regardless of growth measure, incorporating both hazards significantly improves model predictive skill.
-This is because the additive effects of both hazards explicitly allow models to account for two independent causes of plant mortality.
-The growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses (e.g.
+This is because the growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses (e.g.
 competition, parasites, herbivory).
@@
 This indicates that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.
-
-\begin{figure}[t]
-\includegraphics{figures/fig4.pdf}
-\caption{Predicted changes in instantaneous mortality rates (A,B), and annual mortality probability (C,D), with increasing growth rates.
-A,C) Estimated species curves for all 203 BCI species used in this study.
-B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species.}
-\label{fig4}
-\end{figure}
 
@@
 
-We found a high level of inter-specific variation was unexplained by wood density, suggesting that other traits are affecting tree mortality rates.
-This unexplained variation may possibly be explained by a species' light demand and maximum height, which has recently been proposed as two major axes of inter-specific variation in tropical rainforests\cite{Poorter:2006tq}.
+We found that wood density only explained \Sexpr{extract_prop_explained(param_variance_explained,I('wood_density'))}\% of the total variation among species, suggesting that other traits are affecting tree mortality rates.
+This unexplained variation may possibly be explained by a species' light demand and maximum height, which have recently been proposed as two major axes of inter-specific variation in tropical rainforests\cite{Poorter:2006tq}.
 While we did not model mortality as a function of these traits, our posthoc correlations (Fig.
@@
 
-Whether the length between censuses biases estimation of growth-dependent and growth-independent hazards should also be examined.
-Long census intervals may lead to underestimation of growth-dependent mortality, and consequently, overestimate the relative contribution of growth-independent hazards.
-This is because if a census interval is too wide, it will fail to account for deaths attributed to rapid declines in growth caused (especially for short-lived plants) or events such as drought.
+Whether the length between censuses affect the estimation of growth-dependent and growth-independent hazards should also be examined.
+Large census intervals may lead to underestimation of growth-dependent mortality, and consequently, overestimate the relative contribution of growth-independent hazards.
+This is because if a census interval is too wide, it will fail to account for deaths attributed to rapid declines in growth caused by senescence (especially for short-lived plants) or events such as drought.
 Because our analysis estimated an individual's past growth from 5 year census intervals, and because it included short-lived species (i.e.
@@
 Lastly, research should also examine how different growth measures alter the predictive skill of mortality models.
-We found that DBH growth outperformed basal area growth, but have as yet  no clear explanation as to why, or how other common measures such as biomass growth\cite{Chave:2005hz} will compare.
+We found that DBH growth outperformed basal area growth in terms of model predictive skill, but have as yet no clear explanation as to why, or how other common measures such as biomass growth\cite{Chave:2005hz} will compare.
 
@@
 \subsection*{Posthoc correlations}
-We calculated a gap index as a measure of a species' light dependence, using canopy census data collected from 1985\textendash1995\cite{Hubbell:2014}.
-The canopy census recorded, in all 5x5 m subplots across the 50 ha plot, the presence of leaf in six height intervals between 0, 2, 5, 10, 20, 30 and > 30 m.
-For each subplot, we calculated the minimum number of strata > 2 m containing leaf; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap).
-We then calculated a weighted mean for each subplot, averaging each cell's value with mean of the 8 neighbouring cells.
+We calculated a gap index as a measure of a species' light dependence using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995.
+The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m).
+For each subplot, we calculated the minimum number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap).
+We then calculated a weighted sum for each subplot, using the values from the 8 neighbouring cells.
 The gap index for each species is then the average light level encountered by new saplings appearing in the census (Fig.

### 2017-06-16 R/bci_data_processing.R

@@ -3,7 +3,7 @@
 
 # DOWNLOADING FUNCTIONS
 
-#Download BCI data
+# Download BCI data
 BCI_download_50ha_plot_full<- function(dest) {
   url <-"https://repository.si.edu/bitstream/handle/10088/20925/bci.full.Rdata31Aug2012.zip"
   download(url, dest, mode="wb")
@@ -21,10 +21,27 @@ BCI_download_species_table <- function(dest) {
   download(url, dest, mode="wb")
 }
 
-# LOADING FUNCTIONS
+# REMAKE EXPORT FUNCTION
+## Required for remake (1 function -> n file outputs)
+export_data <- function(data, filename) {
+  filename_fmt <- sub("\\.rds$", "_%s.rds", filename)
+  filename_sub <- sprintf(filename_fmt, seq_along(data))
+  for (i in seq_along(data)) {
+    saveRDS(data[[i]], filename_sub[[i]])
+  }
+}
 
-#Load 50ha census data
+# GENERIC LOADING FUNCTION
+# loads an RData file, and returns it
+load_rdata <- function(file) {
+  v <- load(file)
+  get(v)
+}
+
+# SPECIFIC LOADING FUNCTIONS
+# Load 50ha census data
 BCI_load_50ha_plot <- function(path_to_zip) {
+  
   
   tmp <- tempfile()
   unzip(path_to_zip, exdir=tmp)
@@ -54,7 +71,7 @@ BCI_load_dbh_error_data <- function(file) {
   data
 }
 
-#Load BCI canopy data
+# Load BCI canopy data
 BCI_load_canopy <- function(path_to_zip) {
   
   tmp <- tempfile()
@@ -88,9 +105,56 @@ BCI_load_nomenclature <- function(file){
 }
 
 # DATA PROCESSING FUNCTION
-# Note all sub functions are used here are at the bottom
 
 BCI_clean <- function(BCI_data, spp_table) {
+  # NESTED FUNCTIONS
+  # Drop last observation
+  drop_last <- function(x) {
+    if(length(x) > 0)
+      x[seq_len(length(x)-1)]
+    else
+      NULL
+  }
+
+  # SUB FUNCTIONS USED TO PROCESS DATA
+  #Look up family
+  lookup_family <- function(tag, spp_table){
+    i <- match(tag, tolower(spp_table[['sp']]))
+    spp_table$family[i]
+  }
+  
+  #Look up species code
[…]

### 2017-06-16 R/figures.R

@@ -1,258 +1,28 @@
-# Plot observed vs predicted true dbh and growth
-plot_obs_v_pred_growth <- function(data) {
-  
-  p1 <- ggplot(data = data,  aes(x = dbh_prev, y = true_dbh1)) + 
-    geom_point(alpha = 0.4, size= 0.2) +
-    geom_abline(intercept = 0, slope = 1,col='red', size= 0.2) +
-    scale_x_continuous(expand=c(0,0), limit=c(NA, 160)) +
-    scale_y_continuous(expand=c(0,0), limit=c(0, 160)) +
-    ylab(expression(widehat("DBH"["t1"])~(cm))) + 
-    xlab("DBH"["t1"]~("cm")) +
-    partial_plot_theme()
-  
-  p2 <- ggplot(data = data,  aes(x = dbh, y = true_dbh2)) + 
-    geom_point(alpha = 0.4,size= 0.2) +
-    geom_abline(intercept = 0, slope = 1,col='red', size= 0.2) +
-    scale_x_continuous(expand=c(0,0), limit=c(NA, 160)) +
-    scale_y_continuous(expand=c(0,0), limit=c(0, 160)) +
-    ylab(expression(widehat("DBH"["t2"])~(cm))) + 
-    xlab("DBH"["t2"]~("cm")) +
-    partial_plot_theme()
-  
-  p3 <- ggplot(data = data,  aes(x = obs_dbh_dt, y = true_dbh_dt)) + 
-    geom_point(alpha = 0.4, size= 0.2) +
-    geom_abline(intercept = 0, slope = 1, col='red', size= 0.2) +
-    scale_x_continuous(expand=c(0,0), limit=c(NA, 5)) +
-    scale_y_continuous(expand=c(0,0), limit=c(0, 5)) +
-    ylab(expression(widehat("DBH growth")~("cm yr"^-1))) +
-    xlab(expression("DBH growth"~("cm yr"^-1))) +
-    partial_plot_theme()
-  
-  p4 <- ggplot(data = data,  aes(x = obs_basal_area_dt, y = true_basal_area_dt)) + 
-    geom_point(alpha = 0.4, size= 0.2) +
-    geom_abline(intercept = 0, slope = 1, col='red', size= 0.2) +
-    scale_x_continuous(expand=c(0,0), limit=c(NA, 300)) +
-    scale_y_continuous(expand=c(0,0), limit=c(0, 300)) +
[…]

### 2017-06-16 R/manuscript_functions.R

@@ -42,7 +42,7 @@ res <-data %>%
     prop_died = deaths/total,
     census = factor(censusid, labels=c('1995to2000','2000to2005','2005to2010'))) %>%
   select(census,censusid, total, deaths, survivors, prop_died)
-return(signif(res[res$censusid==census,"prop_died"],2)*100)
+signif(res[res$censusid==census,"prop_died"],2)*100
 }
 
 # Extract census scaling factor
@@ -84,4 +84,29 @@ fullmodelcode <- function() {
   chunks <- get_final_model_chunks(task)
   model <- make_stan_model(chunks)
   writeLines(model$model_code)
+}
+
+# Extract baseline mortality estimate (prob death or hazard values) for different wood density
+extract_baseline_estimate <- function(model, hazard=FALSE, wood_density) {
+  fit <- model$fits[[1]]
+  samples <- rstan::extract(fit, pars=c("mu_log_gamma","c1"))
+  samples <- as.data.frame(lapply(samples, as.vector)) 
+  
+  
+  output <-samples %>%
+    mutate(wood_density_centered = wood_density/0.6,
+           inst_hazard = exp(mu_log_gamma) * wood_density_centered^c1,
+           prob_death = 1 - exp(-exp(mu_log_gamma) * wood_density_centered^c1))
+  
+  if(hazard==FALSE) {
+    output %>%
+      summarise(mean = mean(prob_death),
+                `2.5%` = quantile(prob_death,0.025),
+                `97.5%` = quantile(prob_death, 0.975))
+  } else {
+    output %>%
+      summarise(mean = mean(inst_hazard),
+                `2.5%` = quantile(inst_hazard,0.025),
+                `97.5%` = quantile(inst_hazard, 0.975))
+  }
 }

### 2017-06-16 R/process_output.R

@@ -1,4 +1,7 @@
-# Extracts optimization estimates
+## Process output functions
+# Note most of these functions require stan_functions.R to be sourced
+
+# Extracts optimization estimates for dbh at each time
 extract_true_dbh_estimates <- function(optimization_results) {
   fit <- optimization_results$par
   data.frame(fit, nm=names(fit)) %>%
@@ -8,24 +11,6 @@ extract_true_dbh_estimates <- function(optimization_results) {
     spread(variable, fit) %>%
     mutate(ind = as.integer(ind)) %>%
     arrange(ind)
-}
-
-# rstan's chain merger
-combine_stan_chains <- function(files) {
-  rstan::sflist2stanfit(lapply(files, readRDS))
-}
-
-# Sub-function to compile chains for our workflow
-compile_chains <- function(comparison) {
-  if(!comparison %in% c("null_model","function_growth_comparison","species_random_effects","rho_combinations","final_model","final_base_growth_hazard_re")) {
-    stop('comparison can only be one of the following: 
-                      "null_model","function_growth_comparison","species_random_effects","rho_combinations","final_model", "final_base_growth_hazard_re"')
-  }
-  tasks <- tasks_2_run(comparison)
-  sets <- split(tasks,  list(tasks$comparison,tasks$model,tasks$growth_measure,tasks$rho_combo,tasks$kfold), sep='_', drop=TRUE)
-  pars <- lapply(sets,  function(s) s[1, c("comparison","model","growth_measure","rho_combo","kfold")])
-  fits <- lapply(sets, function(s) combine_stan_chains(s[['filename']]))
-  list(model_info=pars, fits=fits)
 }
 
 # Compile chains for multiple model comparisons
@@ -38,7 +23,39 @@ compile_models <- function(comparison) {
   }
 }
 
-# Diagnostic summary function
+# Sub-function to compile_models for our workflow
+compile_chains <- function(comparison) {
+  if(!comparison %in% c("null_model","function_growth_comparison","species_random_effects","rho_combinations","final_model","final_base_growth_hazard_re")) {
+    stop('comparison can only be one of the following: 
+         "null_model","function_growth_comparison","species_random_effects","rho_combinations","final_model", "final_base_growth_hazard_re"')
+  }
+  tasks <- tasks_2_run(comparison)
[…]

### 2017-06-16 R/stan_functions.R

@@ -1,4 +1,18 @@
-# Kfold tasks
+# FUNCTIONS FOR COMPILING AND DISTRIBUTING STAN MODELS
+
+#### TASK FUNCTIONS ####
+
+# Wrapper function to allow for both full data and kfold tasks
+tasks_2_run <- function(comparison,iter=4000,path='.') {
+  if(comparison %in% c("final_model","final_base_growth_hazard_re")) {
+    full_data_tasks(comparison, iter, path)
+  } 
+  else {
+    kfold_tasks(comparison, iter, path)
+  }
+}
+
+# CROSS VALIDATION
 kfold_tasks <- function(comparison,iter=4000,path='.') {
   n_kfolds = 10
   n_chains = 3
@@ -27,7 +41,7 @@ kfold_tasks <- function(comparison,iter=4000,path='.') {
          },
          stop('comparison can only be one of the following: 
                       "function_growth_comparison","species_random_effects","rho_combinations"')
-         )
+  )
   ret <- expand.grid(comparison=comparison,
                      model = model,
                      iter=iter,
@@ -46,7 +60,7 @@ kfold_tasks <- function(comparison,iter=4000,path='.') {
   return(ret)
 }
 
-# Full data tasks
+# FINAL MODEL
 full_data_tasks <- function(model, iter=4000,path='.') {
   n_chains = 3
   switch(model,
@@ -59,35 +73,27 @@ full_data_tasks <- function(model, iter=4000,path='.') {
            comparison = "final_base_growth_hazard_re"
          }, 
          stop("Model must be either 'final_model' or 'final_base_growth_hazard_re'"))
-         
-         ret <- expand.grid(comparison=comparison,
-                            model=model,
-                            iter=iter,
-                            chain=seq_len(n_chains),
-                            growth_measure= "true_dbh_dt",
-                            rho_combo= rho_combo,
-                            kfold=0,
-                            stringsAsFactors=FALSE)
-         
-         ret$modelid <- rep(1:nrow(unique(ret[,c('comparison','model','growth_measure','rho_combo','kfold')])),each = n_chains)
-         ret <- ret %>%
-           mutate(jobid = seq_len(n()),
-                  filename = sprintf("%s/results/chain_fits/%s/%d.rds", path, model, jobid),
-                  fold_data = sprintf("%s/precompile/bci_data_full.rds", path))
-         return(ret)
+  
[…]

### 2017-06-16 R/utils.R

@@ -1,67 +1,0 @@
-
-
-#loads an RData file, and returns it
-load_rdata <- function(file) {
-  v <- load(file)
-  get(v)
-}
-
-## Really ugly working around something I've not worked out how to do
-## in remake (1 function -> n file outputs)
-export_data <- function(data, filename) {
-  filename_fmt <- sub("\\.rds$", "_%s.rds", filename)
-  filename_sub <- sprintf(filename_fmt, seq_along(data))
-  for (i in seq_along(data)) {
-    saveRDS(data[[i]], filename_sub[[i]])
-  }
-}
-
-# Convert df to list
-df_to_list <- function(x) {
-  attr(x, "out.attrs") <- NULL # expand.grid leaves this behind.
-  unname(lapply(split(x, seq_len(nrow(x))), as.list))
-}
-
-# Drop last observation
-drop_last <- function(x) {
-  if(length(x) > 0)
-    x[seq_len(length(x)-1)]
-  else
-    NULL
-}
-
-# Calculate logloss
-logloss = function(actual, predicted, eps = 1e-15) {
-  predicted = pmin(pmax(predicted, eps), 1-eps)
-  - (actual * log(predicted) + (1 - actual) * log(1 - predicted))
-}
-# Sum of squares
-sum_squares <- function(x) {
-  sum((x-mean(x))^2)
-}
-
-# position label at a fractional x/y position on a plot
-label <- function(px, py, lab, ..., adj = c(0, 1)) {
-  usr <- par("usr")
-  x <- usr[1] + px * (usr[2] - usr[1])
-  y <- usr[3] + py * (usr[4] - usr[3])
-  
-  if (par("ylog"))
-    y <- 10^y
-  if (par("xlog"))
-    x <- 10^x
-  
-  text(x, y, lab, adj = adj, xpd=NA, ...)
-}
-
-# Plot theme
-partial_plot_theme <- function(legend.position = "none", strips = FALSE,...) {
-  sb <- if(strips==TRUE) element_rect(fill='lightgrey') else element_blank()
-  st <- if(strips==TRUE) element_text() else element_blank()
-  theme_classic(base_size = 7) + theme(strip.text = st,
-                                       strip.background = sb,
-                                       legend.position = legend.position,
-                                       axis.line.x = element_line(colour = 'black', size=0.5, linetype='solid'),
-                                       axis.line.y = element_line(colour = 'black', size=0.5, linetype='solid'),
-                                       plot.margin = unit(c(3,3,3,3), "mm"))
-}

### 2017-06-16 ms/supp.Rnw
Add authors

--- 2017-06-01 ms/supp.Rnw
@@
 
-% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\author[a,1]{James S Camac}
+\author[a,b,1]{James S Camac}
+\author[c]{Richard Condit}
+\author[d]{Richard G FitzJohn}
+\author[e]{Lachlan McCalman}
+\author[e]{Daniel Steinberg}
 \author[a]{Mark Westoby}
-\author[b]{S Joseph Wright}
-\author[c]{Daniel S Falster}
+\author[f]{S Joseph Wright}
+\author[a,g]{Daniel S Falster}
 
-\affil[a]{Department of Biological Sciences, Macquarie University NSW 2109, Australia}
-\affil[b]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
-\affil[c]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales}
+\affil[a]{Department of Biological Sciences, Macquarie University, New South Wales, Australia}
+\affil[b]{Current Address: Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Victoria, Australia}
+\affil[c]{Field Museum of Natural History and Morton Arboretum, Chicago, Illinois, USA}
+\affil[d]{Imperial College, London, United Kingdom}
+\affil[e]{Data61, CSIRO, Australia}
+\affil[f]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
+\affil[g]{Current Address: Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, NSW 2052, Australia}
 

### 2017-06-16 ms/ms.Rnw
Changes to sub sampling

@@
 To validate this model, we fit a series of models with increasing complexity (eq.
-\ref{eq:model} being the most complex) and compare their skill in predicting patterns of tropical tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_unique)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_unique)} tree species at Barro Colorado Island (BCI), Panama (Fig.
+\ref{eq:model} being the most complex) and compare their skill in predicting patterns of tropical tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_dbh)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_dbh)} tree species at Barro Colorado Island (BCI), Panama (Fig.
 \ref{fig1}).
@@
 Comparing the three 5-year intervals between censuses from 1995 to 2010, we found tree mortality progressively decreasing over time.
-The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%).
+The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 3)}\%).
 Consequently, when we allowed hazard rates to be scaled by census (i.e.
@@
 Growth rates were estimated from field measurements of diameter, which inevitably include observation error.
-In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of estimated growth rates were negative.
+In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_dbh)}\% of estimated growth rates were negative.
 To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see Supplementary Material S1 for details; Fig.

### 2017-06-16 R/bci_data_processing.R

@@ -258,29 +258,6 @@ add_true_growth <- function(data, true_dbh_mod) {
 
 # CROSS VALIDATION FUNCTIONS
 
-# Randomly sample one observation for each individual
-reduce_to_single_ind_obs <- function(data) {
-  # set seed so that same subsetting is implemented on all machines
-  set.seed(523)
-  
-  # returns vector of same length which is all FALSE except
-  # for a single, randomly placed TRUE
-  sample_one <- function(x) {
-    seq_along(x) == sample(length(x), 1)
-  }
-  
-  ret <-
-    data %>%
-    group_by(treeid) %>%
-    mutate(keep = sample_one(treeid)) %>%
-    filter(keep) %>%
-    select(-keep) %>%
-    ungroup() %>%
-    select(gx, gy,species,sp, sp_id, censusid, dead_next_census,
-           census_interval, rho, dbh_prev, dbh)
-}
-
-
 # Split into k equally sized datasets
 split_into_kfolds <- function(data, k=10) {
   # make dataset an even multiple of 10

### 2017-10-23 ms/ms.Rnw

@@
 \affil[f]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
-\affil[g]{Current Address: Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, NSW 2052, Australia}
+\affil[g]{Current Address: Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, Sydney NSW 2052, Australia}
 

### 2017-10-24 ms/ms.Rnw
Updated some of the intext functions and text around subsampling

@@
 \significancestatement{Tree mortality is a fundamental demographic process affecting forest dynamics and carbon cycling.
-Here, for the first time, we use observed survival records over a 15 year period from more than 180,000 individuals, to simultaneously estimate growth-dependent and growth-independent mortality across 203 tropical forest tree species.
+Here, for the first time, we use over 400,000 observed survival records collected over a 15 year period from more than 180,000 individuals, to simultaneously estimate growth-dependent and growth-independent mortality across 203 tropical forest tree species.
 We found that growth-dependent mortality was the predominate factor influencing tree mortality rates at Barro Colorado Island.
@@
 We show that functions that partition mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations.
-Moreover, we show that temporal variation in mortality is large and that the effect of wood density -- a prominent trait known to influence tree mortality -- explains only \Sexpr{extract_prop_explained(param_variance_explained,I('wood_density'))}\% of the total variability observed among species.
+Moreover, we show that the effect of wood density -- a prominent trait known to influence tree mortality -- explains only \Sexpr{extract_prop_explained(param_variance_explained,I('wood_density'))}\% of the total variability observed among species.
 Lastly, our analysis show that growth-dependent processes are the predominant contributor to rates of tree mortality at BCI.
@@
 \ref{fig1}A).
+In total \Sexpr{n_obs(BCI_model_dataset_true_dbh)} observations were used to fit these models.
 We compare the skill of different hazard functions in predicting outcomes in novel data (i.e.
@@
 Consequently, when we allowed hazard rates to be scaled by census (i.e.
-adding term $\delta$) we observed a significant increase in predictive skill (Fig.
+adding term $\delta$) we observed a small, but significant, increase in predictive skill (Fig.
 \ref{fig2}A).
-Assuming the long-term average census effect ($\delta$) is 1, deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.
 
@@
 Furthermore, future research should focus on examining how climate variables influence both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}.
-Our analyses revealed a significant decline in mortality at BCI between 1995 and 2010.
+Our analyses revealed a a marginal decline in mortality at BCI between 1995 and 2010.
 Unfortunately, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3).
@@
 We also excluded individuals that: 1) did not survive at least two censuses (two being required to estimate growth rate); 2) were not consistently measured at 1.3 m above ground; 3) were multi-stemmed; 4) resprouted or \textit{``returned from the dead''}; or 5) were extreme outliers -- stems which grew more than 5 cm yr$^{-1}$ or shrunk more than 25\% of their initial diameter.
-In order to maintain statistical independence of measurements and avoid estimating 180,500 individual parameters, we sub-sampled the dataset by randomly selecting one observation per individual out of a possible three censuses.
+In total \Sexpr{n_obs(BCI_model_dataset_true_dbh)} observations were used in this study comprising \Sexpr{n_inds(BCI_model_dataset_true_dbh)} individual trees and \Sexpr{n_spp(BCI_model_dataset_true_dbh)} species.
+Because of computational difficulties, the models fit in this study do not include individual random effects, as this would require estimation of an additional \Sexpr{n_inds(BCI_model_dataset_true_dbh)} parameters.
+Instead, our models assume that repeat measurements of an individual are independent of one another.
+We believe this is a reasonable assumption given that there is approximately 5-years between censuses.
 

### 2017-10-24 R/manuscript_functions.R

@@ -1,13 +1,16 @@
 # Manuscript functions
 
+n_obs <- function(data) {
+  prettyNum(nrow(data), big.mark = ",")
+}
 # Number of individuals used
 n_inds <- function(data) {
-  as.character(round(nrow(data),-2))
+  prettyNum(length(unique(data$treeid)), big.mark = ",")
 }
-
+prettyNum(sum(c(199630,211350,722040,15274700,13019600)),big.mark=",")
 # Number of species used in study
 n_spp <- function(data) {
-  length(unique(data$sp))
+  prettyNum(length(unique(data$sp)),big.mark = ",")
 }
 
 # Proportion of negative observations
@@ -17,7 +20,7 @@ prop_neg_growth <- function(data) {
 
 # Number of observations in remeasurement error data
 nobs_meas_err_data <- function(data) {
-  nrow(data$discrep)
+  prettyNum(nrow(data),big.mark = ",")
 }
 
 # Observed standard deviation of measurement error
@@ -27,7 +30,7 @@ obs_error_sigma <- function(data) {
 
 # Number of heldout observations
 n_heldout <- function(data) {
-  as.character(nrow(data[[1]]$heldout))
+ prettyNum(nrow(data[[1]]$heldout), big.mark = ",")
 }
   
 # Proportion of deaths by census

### 2017-11-10 ms/ms.Rnw
Edits throughout, refine Fig 5
- change Fig5 from double to single panel
- spaces before all \cite calls
- remove some text near end of discussion
- rework sections on variance explained

@@
 \affil[f]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
-\affil[g]{Current Address: Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, Sydney NSW 2052, Australia}
+\affil[g]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, Sydney NSW 2052, Australia}
 
@@
 We found that growth-dependent mortality was the predominate factor influencing tree mortality rates at Barro Colorado Island.
-Furthermore, we found that wood density most strongly influenced mortality rates by decreasing growth-independent mortality.
-However, wood density only accounted for a small fraction of the overall species variability observed in growth-independent mortality rates, suggesting that there must be other species traits that strongly affect mortality.
-Which traits, however, remains unknown.}
+Furthermore, we found that while wood density influenced mortality rates by decreasing growth-independent mortality, wood density only accounted for a small fraction of the overall species variability in mortality rates, suggesting that there must be other species traits that strongly affect mortality.}
 
@@
 In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive instantaneous mortality functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
-Specifically, we develop mortality functions that not only integrate individual, species and temporal effects, but also simultaneously estimate the independent contribution of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate.
-We show that functions that partition mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations.
+Specifically, we develop mortality functions that not only integrate individual, species, and temporal effects, while also partitioning the contributions of growth-dependent and growth-independent effects on the overall instantaneous mortality rate.
+We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent mortality outperform alternative formulations.
 Moreover, we show that the effect of wood density -- a prominent trait known to influence tree mortality -- explains only \Sexpr{extract_prop_explained(param_variance_explained,I('wood_density'))}\% of the total variability observed among species.
@@
 
-\dropcap{R}ates of plant mortality are known to be highly variable among coexisting species, between forests and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}.
+\dropcap{R}ates of plant mortality are known to vary widely among individuals within species, among coexisting species, between forests, and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}.
 This variation has considerable consequences for forest structure and dynamics.
-For example, death of a single large tree can transfer up to 20000 kg of carbon from living to decaying carbon pools\cite{Chave:2005hz}.
-Furthermore it creates a gap in the canopy that can restart a successional race, during which 100's-1000's more plants may die while competing for a spot in the sun.
-In models of forest dynamics, variation in mortality rates has been shown to have a larger impact on forest structure than variation in absolute growth rates\cite{Johnson-2016}.
-As such, improving our ability to understand the mortality process is a priority for making accurate predictions about forest and carbon dynamics as well as water or nutrient cycles; especially in an era of rapid environmental change.
+For example, death of a single large tree can transfer up to 20000 kg of carbon from living to decaying carbon pools \cite{Chave:2005hz}.
+Furthermore it creates a gap in the canopy that can restart a successional race, during which 100's of plants may die while competing for a spot in the sun.
+In models of forest dynamics, variation in mortality rates has been shown to have a larger impact on forest structure than variation in absolute growth rates \cite{Johnson-2016}.
+Improving our understanding of the mortality process is therefore a priority for making accurate predictions about population, carbon and nutrient dynamics of forests; especially in an era of rapid environmental change.
 
 Two difficulties arise when studying tree mortality in tropical rainforests.
-The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates\cite{Poorter:2008iu}.
+The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates \cite{Poorter:2008iu}.
 This requirement arises in part because the observable outcome -- alive vs.
 dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$, also called the ``hazard function''.
-In classic survival analysis\cite{Kleinbaum:2012tn}, the probability an individual dies between times $t_2$ and $t_3$ is a function of $\lambda_i(t)$:
+In classic survival analysis \cite{Kleinbaum:2012tn}, the probability an individual dies between times $t_2$ and $t_3$ is a function of $\lambda_i(t)$:
 \begin{equation} \label{eq:mortality_model}
@@
 \end{equation}
-As trees are long-lived and we are trying to estimate the shape of a continuous hazard function from binary data, large sample sizes are required.
-Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals\cite{Anderson-Teixeira-2015a}.
+As trees are long-lived and we are trying to estimate the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes are required.
+Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.
 
 Another difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of hazard functions that skilfully predict patterns in mortality.
-Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate\cite{Kobe:1995tw,Wyckoff:2002ul}, traits\cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size\cite{Vieilledent:2010fv, Hurst:2011ky}.
-However, these influences have not yet been integrated into a common hazard function\cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.
+Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate \cite{Kobe:1995tw,Wyckoff:2002ul}, traits \cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size \cite{Vieilledent:2010fv, Hurst:2011ky}.
+However, these influences have not yet been integrated into a common hazard function \cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.
 
@@
 The second are growth-independent hazards, where plants die because of stochastic events, irrespective of their growth rate, such as windfall or fire.
-We therefore assume an individual's hazard is the sum of growth-independent and growth-dependent rates.
+An individual's total hazard is the sum of growth-independent and growth-dependent components.
 
 A further challenge in developing hazard functions for plants is to integrate the conflicting relationships observed within and across species between mortality and growth.
-Within species, mortality rates are lower for fast-growing individuals, presumably because those individuals have superior carbon budgets and are thus able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
+Within species, mortality rates are lower for fast-growing individuals, presumably because those individuals have superior carbon budgets and are thus able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
 Empirical studies broadly support this theory, with many indicating exponential declines in mortality with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}.
-By contrast, across species, there is a strong trade-off between growth and survival, with individuals from faster-growing species  exhibiting higher mortality rates than individuals from slower-growing species\cite{Poorter:2008iu, Wright:2010fl}.
+By contrast, across species, there is a strong trade-off between growth and survival, with individuals from faster-growing species  exhibiting higher mortality rates than individuals from slower-growing species \cite{Poorter:2008iu, Wright:2010fl}.
 This trade-off may arise via traits with antagonistic effects, such as wood density.
-Denser wood -- which is expensive to build and thus slows growth --  reduces stem breakage\cite{vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, pathogen attack\cite{Augspurger:1984wx}, and ultimately mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+Denser wood -- which is expensive to build and thus slows growth --  reduces stem breakage \cite{vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, pathogen attack \cite{Augspurger:1984wx}, and thereby mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
 
@@
 \ref{eq:model} allows for additive growth-independent and growth-dependent effects, includes a negative exponential effect of growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
-Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ are allowed to vary by species.
+Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$.
 Here we include an effect of a species trait (wood density, $\rho$), as well as a species random effect that captures any remaining species-level differences not accounted for by wood density (for details see Methods).
@@
 
-In stage 1 we ask whether mortality rates vary substantially between censuses, to establish whether a census effect is required.
-The null model is simply a constant, invariant with respect to species, growth rate, or year.
-In stage 2 we assess whether a hazard function including both growth-dependent and growth-independent terms outperform a function including only one of these.
+In stage 1, we ask whether mortality rates vary substantially between censuses, to establish whether a census effect is required.
+The null model is a constant, invariant with respect to species, growth rate, or year.
+In stage 2, we assess whether a hazard function including both growth-dependent and growth-independent terms outperform a function including only one of these.
 Species effects were excluded, so we are simply asking which of three functional forms for $\lambda_i(t)$ (Fig.
@@
 Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared.
-We also investigate which growth measure (stem diameter or stem area increment) more skilfully predicts growth-dependent mortality.
-
-In stage 3, we examined whether including wood density (a species-level trait) improved model skill, and if so, whether the effect of wood density was on growth-dependent, growth-independent, or both hazards.
+We also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.
+In stage 3, we examine whether including wood density (a species-level trait) improved model skill, and if so, whether the effect of wood density was on growth-dependent, growth-independent, or both hazards.
 Finally, we fit a model that allowed parameters to vary by species-level variation that was otherwise not captured by wood density.
 This allowed us to ask what proportion of inter-specific hazard variability is explained by wood density.
-Using the final `best' model we also conducted posthoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.
+Using the final ``best'' model we also conducted post-hoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.
 
@@
 \subsection*{Mortality over time}
-Comparing the three 5-year intervals between censuses from 1995 to 2010, we found tree mortality progressively decreasing over time.
+Comparing the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates progressively decreasing over time.
 The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 3)}\%).
@@
 A) Five sequential stages of model selection with each stage increasing in complexity.
-Null: null model ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate.
-This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: Inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
+Null: constant ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate.
+This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
 B) Comparing predictive skill for wood density parameter combinations.
-Colors for both panels represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth.
+Colors represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth.
 For both panels symbols represent functional form: null and growth-independent (circle), growth-dependent hazard (triangle), both growth-independent and growth-dependent hazards (square).
@@
 
-\subsection*{Growth hazard functions}
+\subsection*{Hazard functions}
 Comparing the three hazard functions in Fig.
-\ref{fig1}B, we found that the third function -- with both growth-dependent and growth-independent terms, i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ -- significantly outperformed both the baseline only or growth-dependent only functions (Fig.
+\ref{fig1}B, we found that the third function --- with both growth-dependent and growth-independent terms, i.e.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the growth-independent only or growth-dependent only functions (Fig.
 \ref{fig2}A).
@@
 
-\subsection*{Effect of wood density on mortality rates}
+\subsection*{Wood density and other species effects}
 Including wood density as an effect on either growth-independent or growth-dependent parameters significantly improved model skill relative to a model without such effects (Fig.
@@
 \ref{fig2} B).
-Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate (Fig.
+Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate, $\gamma$ (Fig.
 \ref{fig3}).
@@
 \ref{fig4} C,D), relative to fast growing individuals of high wood density species.
-For example, the model predicted that a species with a wood density of 0.3 g $\textrm{cm}^{-3}$ had an estimated mean probability of dying of \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.3),2)['mean']} yr$^{-1}$ compared to \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.8),2)['mean']} yr$^{-1}$ for a species with a wood density of 0.8 g $\textrm{cm}^{-3}$.
+For example, a species with a wood density of 0.3 g $\textrm{cm}^{-3}$ had an estimated mean probability of dying of \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.3),2)['mean']} yr$^{-1}$ compared to \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.8),2)['mean']} yr$^{-1}$ for a species with a wood density of 0.8 g $\textrm{cm}^{-3}$.
+Incorporating an additional species-level random effect to capture any additional inter specific differences substantially improved model skill relative to a model with only wood density (Fig.
+\ref{fig2} A).
 
@@
 \subsection*{Proportion of variation explained}
-By partitioning the modelled variation between our two random effects (i.e.
-species and census; Fig.
-\ref{fig5}A), we found that approximately \Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of variation in predicted mortality rates was due to species differences, only \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of which was captured by wood density.
-By contrast, census effects accounted for approximately \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation in mortality rates.
-When we partitioned modelled variation into the two independent hazards we found that the growth-dependent hazard (i.e.
-the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the overall variation in mortality rates, and the remainder was attributed to the growth-independent hazard.
+To compliment the above analysis based on skill in predicting held-out data, we used the final fitted model to estimate the amount of variation being contributed by different factors in 1-yr survival for all individuals (Fig.
+\ref{fig5}).
+All species-level effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr survival.
+However, wood density alone captured only \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of total variation.
+Census effects accounted for an even smaller fraction: only \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation.
+Comparing the two independent hazards, growth-dependent hazards (i.e.
+the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total variation, while growth-independent hazards accounted for a smaller proportion (Fig.
+\ref{fig5}).
 
 \subsection*{Mortality rates, maximum species size and species light requirements}
-We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the growth-independent hazard term, were weakly and positively correlated with gap index (our measure of species light requirement; Fig.
-\ref{fig6} A,C, E).
+To determine whether species-level parameters were associated with their maximum size or light requirement, we conducted a post-hoc analysis comparing the fitted parameters to these traits.
+We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the growth-independent hazard term, were weakly and positively correlated with gap index (a measure of species light requirement; Fig.
+\ref{fig6} A,C,E).
 By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index.
-The combination of these correlations suggest that species which predominately recruit in gaps (i.e.
-higher gap index) were more prone to death due to both low growth and stochastic chance.
-They also required faster growth rates to achieve a given mortality rate relative to a species that recruits in non-gaps.
-We detected no strong correlations between estimated species parameters and their associated maximum DBH (Fig.
-\ref{fig6} B,D,F,H).
+These correlations suggest that species which predominately recruit in gaps are more prone to death, due to both low growth and stochastic chance.
+They also required faster growth rates to achieve a given mortality rate relative to a species that recruit in shade.
+Correlations between estimated species parameters and their associated maximum DBH were weak (Fig.
+\ref{fig6}).
 
@@
 \includegraphics{figures/fig5.pdf}
-\caption{Proportion of predicted variation in 5-year survival explained by different covariates.
-A) Wood density (grey) as well as species and census random effects.
-B) Growth-dependent and growth-independent (baseline) hazards.}
+\caption{Proportional reduction in predicted 1-year survival from removing different effects.
+Effects are not mutually exclusive so sum to more than 1.0.
+Note also that the effect of "wood density" is nested within the effect of "species".}
 \label{fig5}
@@
 Our Bayesian framework coupled with cross validation revealed that the most explanatory and parsimonious model of tropical tree death was that which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) attributed the effect of wood density to growth-independent mortality; and 4) incorporated temporal variability.
-Moreover, we found that rates of tropical tree mortality varied substantially between species and that wood density, a species level functional trait, explained only a small proportion of the overall inter-specific variation.
+Moreover, we found that rates of tropical tree mortality varied substantially between species and that wood density, a species level functional trait, explained only a limited proportion of the overall inter-specific variation.
 
-The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards\cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}.
+The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}.
 We show that regardless of growth measure, incorporating both hazards significantly improves model predictive skill.
@@
 
-Another advantage of partitioning mortality into growth-dependent and growth-independent hazards was that we could estimate the proportion of variation attributed to each.
-Like many other studies\cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlight the importance of light competition in influencing tropical tree demographic rates.
+Additionally, the partitioning of mortality into growth-dependent and growth-independent effects allowed us to estimate the proportion of variation attributed to each.
+Like many other studies \cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlight the importance of light competition in influencing tropical tree demographic rates.
 Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted variability in mortality rates (Fig.
 \ref{fig5}).
-This indicates that carbon budget deficiencies were a major contributor to mortality rates at BCI between 1995 and 2010.
+This suggests deficiencies in carbon budget are a major contributor to tree death on BCI.
 
@@
 Specifically, high wood density species had lower baseline rates relative to low wood density counterparts.
-This finding corroborates the observed negative correlation observed between mortality and wood density reported elsewhere\cite{Kraft:2010kq,Chave:2009iy}.
-More importantly, our analyses support the theory that wood density predominately reduce mortality rates by decreasing a species vulnerability to growth-independent stochastic events such as windthrow, trampling and treefall\cite{vanGelder:2006exa,Chave:2009iy}.
+This finding corroborates the observed negative correlation observed between mortality and wood density reported elsewhere \cite{Kraft:2010kq,Chave:2009iy}.
+More importantly, our analyses support the theory that wood density reduces mortality rates by decreasing a species' vulnerability to growth-independent threats, such as windthrow, trampling and treefall \cite{vanGelder:2006exa,Chave:2009iy}.
 
-While wood density effects are now being incorporated in mortality algorithms of many vegetation models\cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation (Fig.
-\ref{fig4}).
+While wood density effects are now being incorporated in mortality algorithms of many vegetation models \cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation (Fig.
+\ref{fig4}-\ref{fig5}).
 This means that such models are likely to severely underestimate the true variability in mortality rates.
@@
 \includegraphics{figures/fig6.pdf}
-\caption{Posthoc correlations between estimated species parameters and measures of species light requirement (A,C,E,G) and species maximum DBH (B,D,F,H).
+\caption{Posthoc correlations between estimated species parameters and measures of species light requirement (left) and species maximum DBH (right).
 Points are mean ($\pm$ 95\% credible intervals) for each of the 203 species used in this study.
@@
 
-We found that wood density only explained \Sexpr{extract_prop_explained(param_variance_explained,I('wood_density'))}\% of the total variation among species, suggesting that other traits are affecting tree mortality rates.
-This unexplained variation may possibly be explained by a species' light demand and maximum height, which have recently been proposed as two major axes of inter-specific variation in tropical rainforests\cite{Poorter:2006tq}.
-While we did not model mortality as a function of these traits, our posthoc correlations (Fig.
-\ref{fig6}) revealed that some inter-specific variation was indeed correlated with light demand.
-Specifically, we found that light demanding species (i.e.
-those that typically recruit in gaps; high gap index) had higher baseline hazard rates and were more susceptible to dying as a result of low growth, relative to those that readily recruit in shade (i.e.
-low gap index), supporting findings by others\cite{Ruger:2011cv}.
-By contrast, we detected no correlations between a species maximum stem diameter and mortality rates, both contradicting\cite{Poorter:2006tq} and supporting\cite{Wright:2010fl} findings by others.
-We recommend future research should focus on examining how other traits influence growth-dependent and growth-independent hazards and what combination of traits minimises unexplained species variability.
+We found that wood density only explained \Sexpr{extract_prop_explained(param_variance_explained,I('wood_density'))}\% of the total \Sexpr{extract_prop_explained(param_variance_explained,I('species'))}\% variation explained by species effects, suggesting that other traits are also affecting tree mortality rates.
+Post-hoc analyses suggest this unexplained variation was in part related to a species' light demand, but not its maximum height.
+Both traits have been proposed major axes of inter-specific variation in tropical rainforests \cite{Poorter:2006tq} (Fig.
+\ref{fig6}).
+Specifically, light demanding species (i.e.
+high gap index) had higher growth-independent hazard rates and were more susceptible to dying as a result of low growth, relative to those that readily recruit in shade (i.e.
+low gap index), supporting past findings \cite{Ruger:2011cv}.
+By contrast, we detected no correlations between a species maximum stem diameter and mortality rates, both contradicting \cite{Poorter:2006tq} and supporting \cite{Wright:2010fl} previous results.
+Future research should therefore resolve how other traits influence growth-dependent and growth-independent hazards, and identify the combination of traits needed to improve model predictive skill.
 
-Furthermore, future research should focus on examining how climate variables influence both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing globally\cite{vanMantgem:2009bi, Mcdowell:2013wf}.
-Our analyses revealed a a marginal decline in mortality at BCI between 1995 and 2010.
+Future research should also focus on resolving how climate variables influence both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing globally \cite{vanMantgem:2009bi, Mcdowell:2013wf}.
+Our analyses revealed a marginal decline in mortality at BCI between 1995 and 2010.
 Unfortunately, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3).
-However, the high numbers of deaths that occurred 1995 and 2000, were likely the product of a severe drought caused by an El-Nino in 1997\textendash 1998 \cite{Condit:1995wl, Mcdowell:2013wf}.
+However, the high numbers of deaths that occurred 1995 and 2000, were potentially the product of a severe drought caused by an El-Nino in 1997\textendash 1998 \cite{Condit:1995wl, Mcdowell:2013wf}.
 
@@
 This is because if a census interval is too wide, it will fail to account for deaths attributed to rapid declines in growth caused by senescence (especially for short-lived plants) or events such as drought.
-Because our analysis estimated an individual's past growth from 5 year census intervals, and because it included short-lived species (i.e.
-those with high annual mortality rates; see Fig.
-\ref{fig4}) and a drought period between 1995 and 2000, our estimates of growth-dependent and growth-independent hazards are likely to be biased.
-
-Lastly, research should also examine how different growth measures alter the predictive skill of mortality models.
-We found that DBH growth outperformed basal area growth in terms of model predictive skill, but have as yet no clear explanation as to why, or how other common measures such as biomass growth\cite{Chave:2005hz} will compare.
 
@@
 
-Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981\textendash1983, 1985, and every 5 years thereafter\cite{Condit:2012nz}.
+Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981\textendash1983, 1985, and every 5 years thereafter \cite{Condit:2012nz}.
 For the purpose of modelling mortality as a function of past growth, we discarded data collected prior to 1990 because diameter measurements were rounded to the nearest 5 mm, whereas later censuses were rounded to the nearest millimetre.
@@
 palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density.
-We also excluded individuals that: 1) did not survive at least two censuses (two being required to estimate growth rate); 2) were not consistently measured at 1.3 m above ground; 3) were multi-stemmed; 4) resprouted or \textit{``returned from the dead''}; or 5) were extreme outliers -- stems which grew more than 5 cm yr$^{-1}$ or shrunk more than 25\% of their initial diameter.
+We also excluded individuals that: 1) did not survive at least two censuses (two being required to estimate growth rate); 2) were not consistently measured at 1.3 m above ground; 3) were multi-stemmed; 4) resprouted or seemingly ``returned from the dead''; or 5) were extreme outliers -- stems which grew more than 5 cm yr$^{-1}$ or shrunk more than 25\% of their initial diameter.
 In total \Sexpr{n_obs(BCI_model_dataset_true_dbh)} observations were used in this study comprising \Sexpr{n_inds(BCI_model_dataset_true_dbh)} individual trees and \Sexpr{n_spp(BCI_model_dataset_true_dbh)} species.
-Because of computational difficulties, the models fit in this study do not include individual random effects, as this would require estimation of an additional \Sexpr{n_inds(BCI_model_dataset_true_dbh)} parameters.
+Because of computational costs, the models fit in this study do not include individual random effects, as this would require estimation of an additional \Sexpr{n_inds(BCI_model_dataset_true_dbh)} parameters.
 Instead, our models assume that repeat measurements of an individual are independent of one another.
@@
 \begin{equation}\label{eq:submodels}
-\alpha_s = \alpha_{0,s} \, (\frac{\rho}{\rho_c})^{\alpha_{1}},
+\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{\rho_c}\right)^{\alpha_{1}},
 \end{equation}
@@
 We executed three independent chains and in all cases modelled parameters converged within 2000 iterations.
-Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic\cite{Gelman:1992ts, Brooks:1998ju}.
+Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Gelman:1992ts, Brooks:1998ju}.
 After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.
@@
 \begin{equation} \label{eq:logloss}
-\mathcal{L}_k = - \frac1{N_k}\,\sum_{i=1}^{N_k}\big(S_i\,\log(p_{i,t_2\rightarrow t_3}) + ( 1 - S_i)\,\log(1 - p_{i,t_2\rightarrow t_3})\big),
+\mathcal{L}_k = - \frac1{N_k}\,\sum_{i=1}^{N_k}\left(S_i\,\log(p_{i,t_2\rightarrow t_3}) + ( 1 - S_i)\,\log(1 - p_{i,t_2\rightarrow t_3})\right),
 \end{equation}

### 2017-11-10 R/figures.R

@@ -313,39 +313,22 @@ plot_fig4b <- function(model,wood_density=c(0.3,0.8), growth_range = c(0.03,0.5)
 
 #### FIGURE 5 ######
 # Proportion of variance explained
-plot_fig5 <- function(param_prop_explained) {
-  p1 <- plot_fig5a(param_prop_explained, ylab="Proportion of variation")
-  p2 <- plot_fig5b(param_prop_explained)
-  plot_grid(p1,p2, ncol=2, labels=LETTERS[1:2], label_size = 7)
-}
+plot_fig5 <- function(param_variance_explained) {
 
-# FIG 5 PANEL FUNCTIONS #
-# Plot proportions of predicted variance explained by wood density, species and census (fig5a)
-plot_fig5a <- function(param_prop_explained, ylab=NULL, xlab=NULL) {
-  dat <- filter(param_prop_explained, !param %in% c("baseline", "growth_dependent")) %>% droplevels() %>%
-    mutate(param = factor(param, levels=c("wood_density","species","census")),
-           level = ifelse(param %in% c("wood_density","species"), "species","census"),
-           level = factor(level, levels=c("species","census")))
-  ggplot(dat, aes(x=level, y=proportion)) +
+  dat <- param_variance_explained %>%
+  # Force order of levels in factor param, otherwise makes alphabetical
+    transform(param = factor(param, levels = param))
+
+  p1 <- ggplot(dat, aes(x=param, y=proportion)) +
     geom_bar(stat='identity', aes(fill=param), width=0.5) +
-    scale_x_discrete(labels=c('species\n','census\n')) +
-    scale_y_continuous(expand=c(0,0), limits=c(0,1)) +
-    scale_fill_manual(values = c("wood_density" ="darkgrey","species" ="black", "census"="black")) + 
-    ylab(ylab) +
-    xlab(xlab) +
-    plot_theme(legend.position="none") 
-}
-
-# Plot proportions of predicted variance explained by baseline and growth dependent hazards
-plot_fig5b <- function(param_prop_explained, ylab=NULL, xlab=NULL) {
-  dat <- filter(param_prop_explained, param %in% c("baseline", "growth_dependent")) %>% droplevels()
-  ggplot(dat, aes(x=param, y=proportion)) +
     geom_bar(stat='identity', fill='black', width=0.5) +
[…]

### 2017-11-10 R/manuscript_functions.R

@@ -56,12 +56,7 @@ census_effects <- function(model,census) {
 
 # Extract proportion explained.
 extract_prop_explained <- function(data, param) {
-  if(param == "species") {
-    signif(data[data$param=='species','proportion'] + data[data$param=='wood_density','proportion'],2)*100
-  }
-  else {
   signif(data[data$param==param,'proportion'],2)*100
-  }
 }
 
 # Prop change in logloss.

### 2017-11-10 R/process_output.R

@@ -127,7 +127,7 @@ combine_logloss_summaries <- function() {
 # Summarise log loss samples
 summarise_crossval_logloss <- function(comparison) {
   models <- compile_models(comparison)
-  # We don't make an explict target of compiled models because of
+  # We don't make an explicit target of compiled models because of
   # a lack of support for long vectors in digest (remake issue #76)
   samples <- extract_logloss_samples(models)
   samples %>%
@@ -149,6 +149,7 @@ extract_logloss_samples <- function(model) {
     samples <- lapply(model, logloss_samples)
     plyr::ldply(samples, .id='modelid') %>%
       select(-modelid)
+
   }
   else { 
     logloss_samples(model)
@@ -341,14 +342,16 @@ predict_observations <- function(model, data) {
 
 # Calculate proportion of model variance explained by each parameter
 get_param_variance_explained <- function(model, data) {
+
   # Sum of squares
   sum_squares <- function(x) {
     sum((x-mean(x))^2)
   }
   # Extract species random effects
   spp_effects <- summarise_spp_params(model, data)
+
   # Extract wood density effect
-  wd_effect <- mean(rstan::extract(model$fits[[1]], pars=c('c1'))$c1)
+  c1 <- mean(rstan::extract(model$fits[[1]], pars=c('c1'))$c1)
   # Extract hyper parameters
   mu_effects <- rstan::extract(model$fits[[1]], pars=c('mu_log_alpha', 'mu_log_beta','mu_log_gamma'))
   mu_effects <- as.data.frame(lapply(mu_effects,as.vector)) %>% summarise_all(funs(mean))
@@ -359,30 +362,45 @@ get_param_variance_explained <- function(model, data) {
     tidyr::gather('censusid','census_err') %>%
     mutate(censusid =c(1,2,3)) # To match with data
   
-  plyr::ldply(spp_effects, .id='parameter') %>%
+
+  # function to predict survival for different levels of effects and time interval dt
+  surv <- function(growth_indepenent, growth_depenent, census, dt = 1) {
+    1-exp(-dt * ((growth_indepenent + growth_depenent)*census))
+  }
+
+  effects <- plyr::ldply(spp_effects, .id='parameter') %>%
     select(parameter,sp, wood_density, mean) %>%
     tidyr::spread(parameter, mean) %>%
     merge(data, by.all=sp) %>%
[…]

### 2017-11-10 ms/ms.Rnw
Minor fixes

@@
 \includegraphics{figures/fig5.pdf}
-\caption{Proportional reduction in predicted 1-year survival from removing different effects.
+\caption{Proportional reduction in predicted 1-year survival across all individuals from removing different effects.
 Effects are not mutually exclusive so sum to more than 1.0.
-Note also that the effect of "wood density" is nested within the effect of "species".}
+Note also that the effect of "wood density" is a subset of the overall "species" effect.}
 \label{fig5}

### 2017-11-10 ms/supp.Rnw
Minor fixes

@@
 Here we provide some further details on methods and additional results, supplementing those presented in the main text.
-Section S1-Estimating True Growth describes a probabilistic methods used to estimate \textit{``true growth''} from field measurements,, taking into account measurement error and the distribution of growth rate across the community.
-Section S2-Parameter distributions describes the distributions applied to species-level random effects and priors put over parameters in the fitted model.
-Section S3-Numerical optimisation outlines some numerical concerns in estimating logloss, while Section S4-Example Stan code gives an example of the code fitted to the data.
+Section ``S1-Estimating True Growth'' describes a probabilistic methods used to estimate individual growth from field measurements,, taking into account measurement error and the distribution of growth rate across the community.
+Section ``S2-Parameter distributions'' describes the distributions applied to species-level random effects and priors put over parameters in the fitted model.
+Section ``S3-Numerical optimisation'' outlines some numerical concerns in estimating logloss, while Section ``S4-Example Stan code'' gives an example of the code fitted to the data.
 Code and data reproducing the results in this paper are available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.
@@
  Growth rates were estimated from field measurements of diameter, which inevitably include observation error.
-In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of estimated growth rates were negative.
+In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_dbh)}\% of estimated growth rates were negative.
 To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community.
@@
 
-\noindent In Stan, the function $ \textrm{log1m\_exp}$ is is defined so that $ \textrm{log1m\_exp}(x)$ has the same value as $\log(1.0-\exp(x))$, but the computation is faster, more accurate, and more stable \cite{StanDevelopmentTeam:2015uz}.
+\noindent In Stan, the function $ \textrm{log1m\_exp}$ is defined so that $ \textrm{log1m\_exp}(x)$ has the same value as $\log(1.0-\exp(x))$, but the computation is faster, more accurate, and more stable \cite{StanDevelopmentTeam:2015uz}.
 
@@
 
-Models were implemented in the `Stan` framework, which uses a specific language to encode model structure \cite{StanDevelopmentTeam:2015uz}.
+Models were implemented in the `Stan` framework, which uses its own specific syntax to encode models \cite{StanDevelopmentTeam:2015uz}.
 Code for all models implemented are available in the supplied source code.

### 2017-11-24 R/process_output.R
Added a table of hyper parameters

@@ -411,3 +411,40 @@ merge_spp_params_covs <- function(spp_params,recruit_gap_conditions, raw_plot_da
       select(species,sp,wood_density, dbh_95, mean_gap_index, mean, median, sd, `2.5%`,`97.5%`)
   }))
 }
+
+
+# Create table of hyper parameter estimates
+param_table <- function(model) {
+  
+  param1 <- as.data.frame(
+    summary(model$fits[[1]], 
+            c('mu_log_alpha',
+              "mu_log_beta",
+              "mu_log_gamma"))$summary) %>%
+    mutate(Parameter = row.names(.)) %>%
+    select(Parameter, mean, `2.5%`, `97.5%`) %>%
+    mutate_at(vars(-Parameter), funs(exp))
+  
+  param2 <- as.data.frame(summary(model$fits[[1]], 'c1')$summary) %>%
+    mutate(Parameter = row.names(.)) %>%
+    select(Parameter, mean, `2.5%`, `97.5%`)
+  
+  x <- bind_rows(param1,param2) %>%
+    select(Parameter, "Geometric mean" = mean, `2.5%`, `97.5%`) %>%
+    mutate_at(vars(-Parameter), funs(round(.,4))) %>%
+    mutate(Parameter = factor(Parameter, 
+                              levels = c("c1",
+                                         "mu_log_gamma",
+                                         "mu_log_beta",
+                                         "mu_log_alpha"),
+                              labels = c("Wood density ($\\rho$)",
+                                         "$\\gamma$",
+                                         "$\\beta$",
+                                         "$\\alpha$")))
+  
+  Hmisc::latex(x, file = "", booktabs = TRUE, rowname = NULL,  
+               colnamesTexCmd="bfseries",
+               col.just = c(rep("c", 4)), label = "table:1",
+               where = "!h",
+               caption ="Hyper parameters estimated from full model")
+}

### 2017-11-24 ms/ms.Rnw
Removed some text about drought and added text about measurement methodology changes based on Richard's comments

@@
 \ref{fig2}A).
-
+A summary of hyper parameter estimates can be found in Table \ref{table:1}.
+<<param-table,results="asis",center=TRUE, echo= FALSE>>= 
+param_table(final_model)
+@
 \begin{figure}[h!]
@@
 A,C) Estimated species curves for all 203 BCI species used in this study.
-B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species.}
+B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species.
+Note: Species parameter estimates can be found in the supplementary Figures S3-S5}
 \label{fig4}
@@
 Unfortunately, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3).
-However, the high numbers of deaths that occurred 1995 and 2000, were potentially the product of a severe drought caused by an El-Nino in 1997\textendash 1998 \cite{Condit:1995wl, Mcdowell:2013wf}.
 
@@
 This is because if a census interval is too wide, it will fail to account for deaths attributed to rapid declines in growth caused by senescence (especially for short-lived plants) or events such as drought.
+Although the opposite is also possible as drought has also been associated with increased tree growth \cite{Condit2017}.
 
@@
 Within the 50-ha plot the diameter at breast height and survival status of all free-standing woody plants that were at least 1.3 m tall and had diameter $\geq$ 1 cm were recorded in 1981\textendash1983, 1985, and every 5 years thereafter \cite{Condit:2012nz}.
-For the purpose of modelling mortality as a function of past growth, we discarded data collected prior to 1990 because diameter measurements were rounded to the nearest 5 mm, whereas later censuses were rounded to the nearest millimetre.
+For the purpose of modelling mortality as a function of past growth, we discarded data collected prior to 1990.
+This was because diameter measurements were rounded to the nearest 5 mm for individuals with dbh < 55 mm, whereas in later censuses all individuals were measured to the nearest millimetre \cite{Condit2017}.
 Consequently, we modelled tree mortality as a function of past growth for censuses 1995\textendash2000, 2000\textendash2005 and 2005\textendash2010.
@@
 After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.
-Full distributions for the final model are shown in Figs.
+Species parameter estimates from the final model are shown in Figs.
 S3\textendash S5.

### 2017-11-27 ms/manuscript.Rnw
updated file names

--- 2017-11-24 ms/ms.Rnw
@@
 \ref{fig2}A).
+Individual census effects can be found in Table S1.
 
@@
 \ref{fig2}A).
-A summary of hyper parameter estimates can be found in Table \ref{table:1}.
-<<param-table,results="asis",center=TRUE, echo= FALSE>>= 
-param_table(final_model)
-@
+A summary of hyper parameter estimates can be found in Table S1.
+
 \begin{figure}[h!]
@@
 Incorporating an additional species-level random effect to capture any additional inter specific differences substantially improved model skill relative to a model with only wood density (Fig.
-\ref{fig2} A).
+\ref{fig2}A).
 
@@
 B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species.
-Note: Species parameter estimates can be found in the supplementary Figures S3-S5}
+Note: Species parameter estimates can be found in the supplementary Figures S3-S5.}
 \label{fig4}
@@
 \subsection*{Proportion of variation explained}
-To compliment the above analysis based on skill in predicting held-out data, we used the final fitted model to estimate the amount of variation being contributed by different factors in 1-yr survival for all individuals (Fig.
+To compliment the above analysis based on skill in predicting held-out data, we used the final fitted model to estimate the amount of variation captured by different factors in 1-yr survival for all individuals (Fig.
 \ref{fig5}).
-All species-level effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr survival.
-However, wood density alone captured only \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of total variation.
-Census effects accounted for an even smaller fraction: only \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation.
-Comparing the two independent hazards, growth-dependent hazards (i.e.
-the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total variation, while growth-independent hazards accounted for a smaller proportion (Fig.
+Species-level effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr survival.
+Wood density explained \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of this species variation.
+Census accounted for only \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation.
+The growth-dependent hazard (i.e.
+the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total variation, while growth-independent hazards accounted for the remainder (Fig.
 \ref{fig5}).
@@
 \includegraphics{figures/fig5.pdf}
-\caption{Proportional reduction in predicted 1-year survival across all individuals from removing different effects.
-Effects are not mutually exclusive so sum to more than 1.0.
-Note also that the effect of "wood density" is a subset of the overall "species" effect.}
+\caption{Variation captured in predicted 1-year survival across all individuals as determined by removing different effects.
+Effects are not mutually exclusive thus sum to more than 1.0.
+Note "wood density" is a subset of the overall "species" effect.}
 \label{fig5}

### 2017-11-27 ms/manuscript.Rnw
minor changes

@@
 In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive instantaneous mortality functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
-Specifically, we develop mortality functions that not only integrate individual, species, and temporal effects, while also partitioning the contributions of growth-dependent and growth-independent effects on the overall instantaneous mortality rate.
+Specifically, we develop mortality functions that not only integrate individual, species, and temporal effects, but also partition the contributions of growth-dependent and growth-independent effects on the overall instantaneous mortality rate.
 We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent mortality outperform alternative formulations.

### 2017-12-01 ms/manuscript.Rnw
Clarifying Fig 1C

@@
 \caption{Outline of methodology.
-A) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, alive or dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
+A) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
 B) We consider three alternative hazard functions: 1) a baseline hazard, 2) a growth-dependent hazard; and 3) a function that combines both baseline and growth-dependent hazards.
@@
 For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
-C) Predictions for plant survival and death are compared to observed outcomes. Incorrect predictions are penalised via the log-loss function.
+C) Each model's skill in predicting observed outcomes ($S$) is quantified via the log-loss function (eq. \ref{eq:logloss}).
 D) The predictive skill of alternative models is evaluated via 10-fold cross validation.

### 2017-12-01 R/figures.R

@@ -62,7 +62,7 @@ plot_fig1 <- function(tree1, tree2, panelc) {
   hazard_plot <- function(f, text){
     x <- seq(0,1,length.out = 50)
     empty_plot()
-    lines(x, f(x), type='l', col="red")
+    lines(x, f(x), type='l', col="orange")
     text(0.1, 0.12, text, pos=4, xpd=NA)
   }
   
@@ -82,9 +82,13 @@ plot_fig1 <- function(tree1, tree2, panelc) {
   par(mar=c(2.5,3,2,1))
   plot(NA, xlim = c(0, 1), ylim= c(0, 5), ann = FALSE, las=1)
   p <- seq(0,1, length.out=200)
-  lines(p, -log(1-p), type='l', col="red")
+  lines(p, -log(1-p), type='l', col="blue")
+  text(0.9, 3, expression(paste(S[i],"=0")), col="blue", pos=3, xpd=NA)
+  lines(p, -log(p), type='l', col="red")
+  text(0.1, 3, expression(paste(S[i],"=1")), col="red", pos=3, xpd=NA)
+
   my_label("C) Penalty for incorrect prediction", x=-0.137)
-  mtext("Probability of outcome", 1, line=2.1, cex=0.5)
+  mtext(expression(paste("Probability of death, ", p[i])), 1, line=2.1, cex=0.5)
   mtext("log loss", 2, line=2.1, cex=0.5)
   
   # Plot cross validation procedure

### 2017-12-01 ms/manuscript.Rnw
Clarify figs 4-5: both now using 1-yr mortality

@@
 This meant that fast growing individuals from a low wood density species had, on average, higher mortality rates (Fig.
-\ref{fig4} A,B), and thus higher annual probability of death (Fig.
+\ref{fig4} A,B), and thus higher probability of death across a 1-yr period (Fig.
 \ref{fig4} C,D), relative to fast growing individuals of high wood density species.
@@
 \includegraphics{figures/fig4.pdf}
-\caption{Predicted changes in instantaneous mortality rates (A,B), and annual mortality probability (C,D), with increasing growth rates.
-A,C) Estimated species curves for all 203 BCI species used in this study.
-B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species.
-Note: Species parameter estimates can be found in the supplementary Figures S3-S5.}
+\caption{Predicted changes in instantaneous mortality rate (A,B), and 1-yr mortality probability (C,D), with increasing  growth rate.
+A,C) Estimated curves for all 203 BCI species used in this study.
+B,D) The average ($\pm$ 95\% credible intervals) curve for a high (0.8) and low (0.3) wood density species.
+See Figs.
+S3-S5 for species-level parameter estimates.}
 \label{fig4}
@@
 \subsection*{Proportion of variation explained}
-To compliment the above analysis based on skill in predicting held-out data, we used the final fitted model to estimate the amount of variation captured by different factors in 1-yr survival for all individuals (Fig.
+To compliment the above analysis based on skill in predicting held-out data, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1yr for all individuals (Fig.
 \ref{fig5}).
-Species-level effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr survival.
+Species-level effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr mortality.
 Wood density explained \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of this species variation.
@@
 \includegraphics{figures/fig5.pdf}
-\caption{Variation captured in predicted 1-year survival across all individuals as determined by removing different effects.
+\caption{Proportion of variation in predicted 1-year mortality (Fig.
+\ref{fig4}C) for all individuals captured by different effects.
 Effects are not mutually exclusive thus sum to more than 1.0.

### 2017-12-01 R/figures.R

@@ -233,7 +233,7 @@ plot_fig3 <- function(spp_params_covs, pred_mu_basehaz) {
     geom_line(data = pred_mu_basehaz, aes(x = wood_density, y= mean), col='blue') +
     plot_theme() +
     scale_y_log10(breaks= breaks, labels = labels) +
-    ylab(expression("Growth-independent mortality rate,"~(gamma))) +
+    ylab(expression("Growth-independent mortality rate,"~gamma~("yr"^-1))) +
     xlab(expression("Wood density"~("g cm"^-3)))
 }
 #### FIGURE 4 ######
@@ -247,7 +247,7 @@ plot_fig4 <- function(model, data) {
           legend.text = element_text(size=4),
           legend.title.align =0.75)
   
-  p3 <- plot_fig4a(model, data, hazard_curve = FALSE)
+  p3 <- plot_fig4a(model, data, hazard_curve = FALSE, ylab = "Probability 1-yr mortality")
   p4 <- plot_fig4b(model, hazard_curve= FALSE, ylab=NULL)
   
   plot_grid(p1,p2,p3,p4, ncol=2, labels=LETTERS[1:4], label_size = 7)
@@ -257,7 +257,7 @@ plot_fig4 <- function(model, data) {
 
 # Plot species predicted mortality v growth curves (Fig4a)
 plot_fig4a <- function(model, data, growth_range= c(0.03,0.5), hazard_curve = FALSE, 
-                       ylab="Annual mortality probability", xlab=expression("Annual dbh growth"~("cm yr"^-1))) {
+                       ylab="", xlab=expression("Annual dbh growth"~("cm yr"^-1))) {
   
   
   preds <- predict_spp_hazard(model, data, growth_range)
@@ -289,7 +289,7 @@ plot_fig4a <- function(model, data, growth_range= c(0.03,0.5), hazard_curve = FA
 
 # Plot average species curve (Fig4b)
 plot_fig4b <- function(model,wood_density=c(0.3,0.8), growth_range = c(0.03,0.5), hazard_curve = FALSE, 
-                       ylab="Annual mortality probability", xlab=expression("Annual dbh growth"~("cm yr"^-1))) {
+                       ylab="1-yr mortality probability", xlab=expression("Annual dbh growth"~("cm yr"^-1))) {
   preds <- predict_mu_hazards(model,wood_density, growth_range, hazard_curve)
   
[…]

### 2017-12-01 R/process_output.R

@@ -363,8 +363,8 @@ get_param_variance_explained <- function(model, data) {
     mutate(censusid =c(1,2,3)) # To match with data
   
 
-  # function to predict survival for different levels of effects and time interval dt
-  surv <- function(growth_indepenent, growth_depenent, census, dt = 1) {
+  # function to predict probability of death for different levels of effects and time interval dt
+  mort <- function(growth_indepenent, growth_depenent, census, dt = 1) {
     1-exp(-dt * ((growth_indepenent + growth_depenent)*census))
   }
 
@@ -379,14 +379,14 @@ get_param_variance_explained <- function(model, data) {
       rho_c = rho/0.6, # centers mean to what models used
       true_dbh_dt_c = true_dbh_dt - 0.172 # centers growth to what models used
       ) %>%
-    # Now calculate predicted 1-yr removing different effects including 
+    # Now calculate predicted 1-yr mortality removing different effects included
     mutate(
-      full_model = surv( alpha * exp(-beta * true_dbh_dt_c), gamma, census_err ),
-      full_minus_census = surv( alpha * exp(-beta * true_dbh_dt_c), gamma, mean(census_err) ),
-      full_minus_rho = surv( alpha * exp(-beta * true_dbh_dt_c), gamma/(rho_c^c1), census_err ),
-      full_minus_spp = surv( median( alpha) * exp(-median(beta) * true_dbh_dt_c), median(gamma), census_err ),
-      full_minus_growthdep = surv(median( alpha * exp(-beta * true_dbh_dt_c)), gamma, census_err),
-      full_minus_growthindep =  surv( alpha * exp(-beta * true_dbh_dt_c), median(gamma), census_err )
+      full_model = mort( alpha * exp(-beta * true_dbh_dt_c), gamma, census_err ),
+      full_minus_census = mort( alpha * exp(-beta * true_dbh_dt_c), gamma, mean(census_err) ),
+      full_minus_rho = mort( alpha * exp(-beta * true_dbh_dt_c), gamma/(rho_c^c1), census_err ),
+      full_minus_spp = mort( median( alpha) * exp(-median(beta) * true_dbh_dt_c), median(gamma), census_err ),
+      full_minus_growthdep = mort(median( alpha * exp(-beta * true_dbh_dt_c)), gamma, census_err),
+      full_minus_growthindep =  mort( alpha * exp(-beta * true_dbh_dt_c), median(gamma), census_err )
       ) %>%
     select(
       full_model, 

### 2017-12-02 ms/manuscript.Rnw
Minor changes to some text to improve clarity

@@
-\documentclass[9pt,twocolumn,twoside]{pnas-new}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[9pt,twocolumn,twoside, lineno]{pnas-new}\usepackage[]{graphicx}\usepackage[]{color}
 
@@
 Another difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of hazard functions that skilfully predict patterns in mortality.
-Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate \cite{Kobe:1995tw,Wyckoff:2002ul}, traits \cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size \cite{Vieilledent:2010fv, Hurst:2011ky}.
+Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate \cite{Kobe:1995tw,Wyckoff:2002ul}, traits \cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size \cite{Vieilledent:2010fv}.
 However, these influences have not yet been integrated into a common hazard function \cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.
@@
 Within species, mortality rates are lower for fast-growing individuals, presumably because those individuals have superior carbon budgets and are thus able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
-Empirical studies broadly support this theory, with many indicating exponential declines in mortality with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}.
+Empirical studies broadly support this theory, with many indicating exponential declines in mortality with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv}.
 By contrast, across species, there is a strong trade-off between growth and survival, with individuals from faster-growing species  exhibiting higher mortality rates than individuals from slower-growing species \cite{Poorter:2008iu, Wright:2010fl}.
@@
 C) Each model's skill in predicting observed outcomes ($S$) is quantified via the log-loss function (eq. \ref{eq:logloss}).
-D) The predictive skill of alternative models is evaluated via 10-fold cross validation.
+D) The predictive skill of alternative models was evaluated via 10-fold cross validation.
 The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$).
-Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing and training data.
-We then average the estimated log loss for each of these 10 model fits.}
+Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing (1 fold; orange) and training (9 folds; green) data.
+Predictive capacity was assessed by averaging the log loss's obtained from the 10 test data predictions.}
 \label{fig1}
@@
 \includegraphics{figures/fig2.pdf}
-\caption{Comparing predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% credible intervals) logarithmic loss, with lower values implying greater predictive skill.
+\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% credible intervals) logarithmic loss, with lower values implying greater predictive skill.
 A) Five sequential stages of model selection with each stage increasing in complexity.
@@
 This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
-B) Comparing predictive skill for wood density parameter combinations.
-Colors represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth.
-For both panels symbols represent functional form: null and growth-independent (circle), growth-dependent hazard (triangle), both growth-independent and growth-dependent hazards (square).
-Errors in panel A are too small to see.}
+B) Predictive skill for wood density parameter combinations.
+Shading represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth.
+For both panels symbols represent functional form: null (circle), growth-dependent hazard (triangle), both growth-independent and growth-dependent hazards (square).}
 \label{fig2}
@@
 \includegraphics{figures/fig3.pdf}
-\caption{Relationship between species growth-dependent mortality rate and wood density.
+\caption{Relationship between species growth-independent mortality rate and wood density.
 Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study.
@@
 \ref{fig2}A).
+See Figs.
+S3-S5 for species-level parameter estimates.
 
@@
 A,C) Estimated curves for all 203 BCI species used in this study.
-B,D) The average ($\pm$ 95\% credible intervals) curve for a high (0.8) and low (0.3) wood density species.
-See Figs.
-S3-S5 for species-level parameter estimates.}
+B,D) The average ($\pm$ 95\% credible intervals) curve for a high (0.8) and low (0.3) wood density species.}
 \label{fig4}
@@
 \subsection*{Proportion of variation explained}
-To compliment the above analysis based on skill in predicting held-out data, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1yr for all individuals (Fig.
+To compliment the above analysis based on skill in predicting held-out (test) data, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1yr for all individuals (Fig.
 \ref{fig5}).
+We achieved this by removing a parameter sequentially and calculating the proportion of the sum of squares attributed to a submodel relative to the full model.
 Species-level effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr mortality.
@@
 \includegraphics{figures/fig5.pdf}
-\caption{Proportion of variation in predicted 1-year mortality (Fig.
-\ref{fig4}C) for all individuals captured by different effects.
+\caption{Proportion of variation in predicted 1-year mortality for all individuals captured by different effects.
 Effects are not mutually exclusive thus sum to more than 1.0.
@@
 \caption{Posthoc correlations between estimated species parameters and measures of species light requirement (left) and species maximum DBH (right).
-Points are mean ($\pm$ 95\% credible intervals) for each of the 203 species used in this study.
+Points are means ($\pm$ 95\% credible intervals) for each of the 203 species used in this study.
 Blue line shows the average trendline as determined from standard linear regression.
-Gap index is a binary species level index of the light required for recruitment.
-Here low numbers indicate the species readily recruits under dense canopy and 1 indicates that a species only ever recruits in gaps.}
+Gap index is a species level index of the light required for recruitment.
+Low gap index indicates the species readily recruits under dense canopy and 1 indicates that a species only ever recruits in gaps.}
 \label{fig6}
@@
 We executed three independent chains and in all cases modelled parameters converged within 2000 iterations.
-Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Gelman:1992ts, Brooks:1998ju}.
+Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Brooks:1998ju}.
 After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.
@@
 The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m).
-For each subplot, we calculated the minimum number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap).
-We then calculated a weighted sum for each subplot, using the values from the 8 neighbouring cells.
-The gap index for each species is then the average light level encountered by new saplings appearing in the census (Fig.
+For each subplot, we calculated the number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap).
+As light may penetrate into a subplot from the edge of a subplot, we rescaled this index to account for values in the eight immediate neighbouring subplots.
+Specifically, we used a weighted sum approach whereby the central subplot is assigned a weight of 8 and the eight neighbouring subplots are assigned a weight of 1.
+This meant that the contribution of the central plot was equivalent to the combined effect of all eight neighbouring plots.
+These weighted values were then summed for each central cell and then these values were normalised to be between 0 and 1.
+The gap index for each species was then estimated as the average light level encountered by new saplings appearing in the census (Fig.
 S2).
@@
 \acknow{We thank H.
-Muller-Landau and XX anonymous reviewers for feedback and B.
+Muller-Landau and anonymous reviewers for feedback and B.
 Carpenter for technical advice.

### 2017-12-02 R/figures.R
Needed to increase ylim for fig 6 and adjusted some colours and text in fig 1

@@ -62,7 +62,7 @@ plot_fig1 <- function(tree1, tree2, panelc) {
   hazard_plot <- function(f, text){
     x <- seq(0,1,length.out = 50)
     empty_plot()
-    lines(x, f(x), type='l', col="orange")
+    lines(x, f(x), type='l', col="#d7301f")
     text(0.1, 0.12, text, pos=4, xpd=NA)
   }
   
@@ -82,10 +82,10 @@ plot_fig1 <- function(tree1, tree2, panelc) {
   par(mar=c(2.5,3,2,1))
   plot(NA, xlim = c(0, 1), ylim= c(0, 5), ann = FALSE, las=1)
   p <- seq(0,1, length.out=200)
-  lines(p, -log(1-p), type='l', col="blue")
-  text(0.9, 3, expression(paste(S[i],"=0")), col="blue", pos=3, xpd=NA)
-  lines(p, -log(p), type='l', col="red")
-  text(0.1, 3, expression(paste(S[i],"=1")), col="red", pos=3, xpd=NA)
+  lines(p, -log(1-p), type='l', col="#5e3c99")
+  text(0.85, 3, expression(paste(S[i],"=0 (survived)")), col="#5e3c99", pos=3, xpd=NA)
+  lines(p, -log(p), type='l', col="#e66101")
+  text(0.15, 3, expression(paste(S[i],"=1 (died)")), col="#e66101", pos=3, xpd=NA)
 
   my_label("C) Penalty for incorrect prediction", x=-0.137)
   mtext(expression(paste("Probability of death, ", p[i])), 1, line=2.1, cex=0.5)
@@ -104,7 +104,7 @@ plot_fig1 <- function(tree1, tree2, panelc) {
               just=c("bottom"), height=unit(1, "npc"))
   
   popViewport(3)
-  text(0.85, 0.45,  expression(paste(bar("logloss"), " =", sum("logloss"[i])/10)), xpd=NA)
+  text(0.85, 0.45,  expression(paste(bar("logloss"), " =", sum(italic("L")[i])/10)), xpd=NA)
 }
 
 # FIGURE 1 sub functions #
@@ -183,7 +183,7 @@ plot_fig2a <- function(logloss_summaries) {
     geom_pointrange(aes(ymin = `2.5%`, ymax=`97.5%`), position=position_dodge(0.5), stroke = 0.5, size=0.4) +
     ylab('Logarithmic loss') + 
     xlab('Hazard function') +
-    scale_shape_manual(values = c(23,21, 24, 22)) +
[…]

### 2017-12-02 ms/manuscript.Rnw
Changed subsection title to better reflect what was done

@@
 
-\subsection*{Proportion of variation explained}
+\subsection*{Parameter contribution to predicted variation}
 To compliment the above analysis based on skill in predicting held-out (test) data, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1yr for all individuals (Fig.

### 2017-12-03 R/figures.R
Need to increase the xlim for figS1

@@ -437,8 +437,8 @@ plot_figS1 <- function(data) {
   p4 <- ggplot(data = data,  aes(x = obs_basal_area_dt, y = true_basal_area_dt)) + 
     geom_point(alpha = 0.4, size= 0.2) +
     geom_abline(intercept = 0, slope = 1, col='red', size= 0.2) +
-    scale_x_continuous(expand=c(0,0), limit=c(NA, 300)) +
-    scale_y_continuous(expand=c(0,0), limit=c(0, 300)) +
+    scale_x_continuous(expand=c(0,0), limit=c(NA, 600)) +
+    scale_y_continuous(expand=c(0,0), limit=c(0, 600)) +
     ylab(expression("Predicted basal area growth"~("cm"^2~"yr"^-1))) +
     xlab(expression("Observed basal area growth"~("cm"^2~"yr"^-1))) +
     plot_theme() +

### 2017-12-03 ms/manuscript.Rnw
Updated rstan citations

@@
 Weak priors on all hyper-parameters were set (see Supplementary Material S2 for details).
-Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz} and employing some numerical optimisations (see Supplementary Material S3-S4 for details and example code).
+Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{rstan2017} and employing some numerical optimisations (see Supplementary Material S3-S4 for details and example code).
 We executed three independent chains and in all cases modelled parameters converged within 2000 iterations.

### 2017-12-03 ms/supplementary.Rnw
updated file names
Updated rstan citations

--- 2017-11-10 ms/supp.Rnw
@@
 
-\noindent In Stan, the function $ \textrm{log1m\_exp}$ is defined so that $ \textrm{log1m\_exp}(x)$ has the same value as $\log(1.0-\exp(x))$, but the computation is faster, more accurate, and more stable \cite{StanDevelopmentTeam:2015uz}.
+\noindent In Stan, the function $ \textrm{log1m\_exp}$ is defined so that $ \textrm{log1m\_exp}(x)$ has the same value as $\log(1.0-\exp(x))$, but the computation is faster, more accurate, and more stable \cite{StanDevelopmentTeam:2017uz}.
 
@@
 \clearpage
+
+<<param-table,results="asis",center=TRUE, echo= FALSE>>= 
+hyperparam_table(final_model)
+@
+
+\clearpage
+
 \begin{figure*}[!hb]

### 2017-12-03 ms/manuscript.Rnw
Minor rewording to shorten paragraphs

@@
 \ref{fig5}).
-We achieved this by removing a parameter sequentially and calculating the proportion of the sum of squares attributed to a submodel relative to the full model.
+We achieved this by removing each parameter and calculating the reduction in the sum of squares for this model relative to the full model.
 Species-level effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr mortality.
@@
 
-Future research should also focus on resolving how climate variables influence both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing globally \cite{vanMantgem:2009bi, Mcdowell:2013wf}.
-Our analyses revealed a marginal decline in mortality at BCI between 1995 and 2010.
-Unfortunately, we were unable to statistically attribute this temporal change to particular climatic variables due to low temporal replication (N=3).
+Future research should also resolve how climate influences both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing \cite{vanMantgem:2009bi, Mcdowell:2013wf}.
+By contrast, our analyses revealed a marginal decline in mortality at BCI from 1995 to 2010.
+Attributing temporal changes to particular climatic variables is challenging, however, due to low number temporal replication (n=3).
 
-Whether the length between censuses affect the estimation of growth-dependent and growth-independent hazards should also be examined.
-Large census intervals may lead to underestimation of growth-dependent mortality, and consequently, overestimate the relative contribution of growth-independent hazards.
-This is because if a census interval is too wide, it will fail to account for deaths attributed to rapid declines in growth caused by senescence (especially for short-lived plants) or events such as drought.
-Although the opposite is also possible as drought has also been associated with increased tree growth \cite{Condit2017}.
+We should also consider how the interval between censuses affects the estimation of growth-dependent and growth-independent hazards.
+Large census intervals may underestimate growth-dependent mortality, as wide census intervals will not capture deaths due to rapid declines in growth, or events such as drought (although drought might also increase tree growth \cite{Condit2017}).
+Consequently, we may overestimate the relative contribution of growth-independent hazards.
 
@@
 Weak priors on all hyper-parameters were set (see Supplementary Material S2 for details).
-Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{rstan2017} and employing some numerical optimisations (see Supplementary Material S3-S4 for details and example code).
+Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{rstan2017} and employing some numerical optimisations (see Supplementary Material S3-S4 for details).
 We executed three independent chains and in all cases modelled parameters converged within 2000 iterations.

### 2017-12-04 ms/manuscript.Rnw
final set up fixes

@@
 \dates{This manuscript was compiled on \today}
-\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
+%\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
 
 \dates{This manuscript was compiled on \today}
-\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
+%\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 \ref{fig1}B, we found that the third function --- with both growth-dependent and growth-independent terms, i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the growth-independent only or growth-dependent only functions (Fig.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the growth-independent only (i.e.
+null) or growth-dependent only functions (Fig.
 \ref{fig2}A).
@@
 \subsection*{Parameter contribution to predicted variation}
-To compliment the above analysis based on skill in predicting held-out (test) data, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1yr for all individuals (Fig.
+To compliment the analysis, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1 year for all individuals (Fig.
 \ref{fig5}).
 We achieved this by removing each parameter and calculating the reduction in the sum of squares for this model relative to the full model.
-Species-level effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr mortality.
+Species effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr mortality.
 Wood density explained \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of this species variation.
-Census accounted for only \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation.
-The growth-dependent hazard (i.e.
-the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total variation, while growth-independent hazards accounted for the remainder (Fig.
+Census accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation.
+The growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total variation, while growth-independent hazards accounted for the remainder (Fig.
 \ref{fig5}).
@@
 with similar formulations for $\beta_s$ and $\gamma_s$.
-Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$, modelled as random realisations from log-normal distributions.
+Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$.
+These random effects were modelled as random realisations from log-normal distributions.
 The form of eq.
@@
 This meant that the contribution of the central plot was equivalent to the combined effect of all eight neighbouring plots.
-These weighted values were then summed for each central cell and then these values were normalised to be between 0 and 1.
-The gap index for each species was then estimated as the average light level encountered by new saplings appearing in the census (Fig.
+These weighted values were then summed and rescaled between 0 and 1 by dividing by the maximum value estimated across all subplots.
+The gap index for each species was estimated as the mean light index encountered by new saplings appearing in the census (Fig.
 S2).

### 2017-12-04 R/figures.R

@@ -269,7 +269,7 @@ plot_fig4a <- function(model, data, growth_range= c(0.03,0.5), hazard_curve = FA
     ggplot(preds, aes(x = dbh_growth,y = annual_prob_mort, group = sp, colour = wood_density)) + 
       geom_line(alpha=0.4, size=0.3) +
       scale_x_continuous(expand=c(0,0)) +
-      scale_y_continuous(expand=c(0,0), limits = c(0, NA)) +
+      scale_y_continuous(expand=c(0,0), limits = c(0, 1)) +
       ylab(ylab) +
       xlab(xlab) +
       scale_colour_gradient(expression("Wood density"~("g cm"^-3)),low="blue", high="red") +

### 2017-12-04 ms/manuscript.Rnw
Minor changes to addresses and include zip unzip in dockerfile for downloading and uncompressing data

@@
 \author[f]{S Joseph Wright}
-\author[g]{Daniel S Falster}
+\author[a,g]{Daniel S Falster}
 
-\affil[a]{Department of Biological Sciences, Macquarie University, New South Wales, Australia}
-\affil[b]{Current Address: Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Victoria, Australia}
+\affil[a]{Department of Biological Sciences, Macquarie University, Sydney, New South Wales, Australia}
+\affil[b]{Current Address: Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Melbourne, Victoria, Australia}
 \affil[c]{Field Museum of Natural History and Morton Arboretum, Chicago, Illinois, USA}
@@
 \affil[f]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
-\affil[g]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, Sydney NSW 2052, Australia}
+\affil[g]{Current Address: Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, Sydney, New South Wales, Australia}
 

### 2017-12-04 ms/supplementary.Rnw
typo

@@
 Here we provide some further details on methods and additional results, supplementing those presented in the main text.
-Section ``S1-Estimating True Growth'' describes a probabilistic methods used to estimate individual growth from field measurements,, taking into account measurement error and the distribution of growth rate across the community.
+Section ``S1-Estimating True Growth'' describes a probabilistic methods used to estimate individual growth from field measurements, taking into account measurement error and the distribution of growth rate across the community.
 Section ``S2-Parameter distributions'' describes the distributions applied to species-level random effects and priors put over parameters in the fitted model.

### 2017-12-05 ms/supplementary.Rnw
Fixed author affiliations to match submitted manuscript

@@
 
-\affil[a]{Department of Biological Sciences, Macquarie University, New South Wales, Australia}
-\affil[b]{Current Address: Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Victoria, Australia}
+\affil[a]{Department of Biological Sciences, Macquarie University, Sydney, New South Wales, Australia}
+\affil[b]{Current Address: Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Melbourne, Victoria, Australia}
 \affil[c]{Field Museum of Natural History and Morton Arboretum, Chicago, Illinois, USA}
@@
 \affil[f]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
-\affil[g]{Current Address: Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, NSW 2052, Australia}
+\affil[g]{Current Address: Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, Sydney, New South Wales, Australia}
 

### 2018-07-16 ms/manuscript.Rnw
Beginning of manuscript edits

@@
 \significancestatement{Tree mortality is a fundamental demographic process affecting forest dynamics and carbon cycling.
-Here, for the first time, we use over 400,000 observed survival records collected over a 15 year period from more than 180,000 individuals, to simultaneously estimate growth-dependent and growth-independent mortality across 203 tropical forest tree species.
-We found that growth-dependent mortality was the predominate factor influencing tree mortality rates at Barro Colorado Island.
-Furthermore, we found that while wood density influenced mortality rates by decreasing growth-independent mortality, wood density only accounted for a small fraction of the overall species variability in mortality rates, suggesting that there must be other species traits that strongly affect mortality.}
+Here, for the first time, we build a statistical mortality model that not only integrates intra-specific and inter-specific variation but also simulataneously estimates the independent contributions of two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth.
+We parameterise this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals encompassed within 203 species.
+We show that both growth-independent and growth-dependent hazards are significant contribtors to tree mortality at Barro Colorado Island.
+Moreover, we show that species traits -- wood density, maximum dbh and shade intolerance -- all variously affect rates of tree mortality but ultimately only account for a small fraction of the total species variability.}
 
@@
 While past research has identified factors influencing tree mortality across a variety of scales, these distinct drivers are yet to be integrated within a unified predictive framework.
-In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive instantaneous mortality functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
-Specifically, we develop mortality functions that not only integrate individual, species, and temporal effects, but also partition the contributions of growth-dependent and growth-independent effects on the overall instantaneous mortality rate.
-We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent mortality outperform alternative formulations.
-Moreover, we show that the effect of wood density -- a prominent trait known to influence tree mortality -- explains only \Sexpr{extract_prop_explained(param_variance_explained,I('wood_density'))}\% of the total variability observed among species.
-Lastly, our analysis show that growth-dependent processes are the predominant contributor to rates of tree mortality at BCI.
-Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests.
+In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive a instantaneous mortality function for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
+Specifically, we develop a mortality function that not only integrates individual, species, and temporal effects, but also partitions the contributions of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate.
+We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and tolerance to shade outperform alternative formulations.
+Moreover, we show that while these three traits can affect both growth-dependent and growth-independent hazards, they explain only only a small fraction of the total variability observed among species.
+Combined, this study provides a statistical framework for predicting individual-level mortality in highly diverse tropical forests.
 It also highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
@@
 
-\dropcap{R}ates of plant mortality are known to vary widely among individuals within species, among coexisting species, between forests, and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}.
+\dropcap{R}ates of plant mortality are known to vary widely among individuals within a species, among coexisting species, between forests, and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}.
 This variation has considerable consequences for forest structure and dynamics.
-For example, death of a single large tree can transfer up to 20000 kg of carbon from living to decaying carbon pools \cite{Chave:2005hz}.
-Furthermore it creates a gap in the canopy that can restart a successional race, during which 100's of plants may die while competing for a spot in the sun.
-In models of forest dynamics, variation in mortality rates has been shown to have a larger impact on forest structure than variation in absolute growth rates \cite{Johnson-2016}.
+For example, the death of a single large tree can transfer up to 20000 kg of carbon from living to decaying carbon pools \cite{Chave:2005hz} and creates a gap in the canopy that can restart a successional race, during which 100's of plants may die while competing for a spot in the sun.
+In models of forest dynamics, variation in tree mortality rates has been shown to have significantly larger impacts on forest structure relative to variation in absolute growth rates \cite{Johnson-2016}.
 Improving our understanding of the mortality process is therefore a priority for making accurate predictions about population, carbon and nutrient dynamics of forests; especially in an era of rapid environmental change.
@@
 Two difficulties arise when studying tree mortality in tropical rainforests.
-The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates \cite{Poorter:2008iu}.
-This requirement arises in part because the observable outcome -- alive vs.
-dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$, also called the ``hazard function''.
-In classic survival analysis \cite{Kleinbaum:2012tn}, the probability an individual dies between times $t_2$ and $t_3$ is a function of $\lambda_i(t)$:
+The first is that large population sizes and long periods of observation are required to make accurate inferences into how various factors affect tree mortality rates \cite{Poorter:2008iu}.
+This requirement arises in part because the observed outcome -- alive vs.
+dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$, also known as the ``hazard function''.
+In classic survival analysis \cite{Kleinbaum:2012tn}, the probability an individual dies between times $t_1$ and $t_2$ is a function of both $\lambda_i(t)$, and the amount of time lapsed between $t_1$ and $t_2$:
 \begin{equation} \label{eq:mortality_model}
-p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
+p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
 \end{equation}
-Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$\cite{Kleinbaum:2012tn}.
-The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
+Here $\int_{t_1}^{t_2} \lambda_i(t) {\rm d}t$ can be thought of as the cumulative hazard an individual obtains between $t_1$ and $t_2$\cite{Kleinbaum:2012tn}.
+The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then a realisation of this cumulative hazard on a probability scale.
 \begin{equation}
-S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
+S_{i,t_1\rightarrow t_2} \sim \textrm{Bernoulli}(p_{i,t_1\rightarrow t_2}).
 \end{equation}
-As trees are long-lived and we are trying to estimate the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes are required.
+As trees are long-lived and we are trying to estimate the shape of a continuous hazard function ($\lambda_i$) from binary data, large sample sizes collected over long durations are required.
 Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.
@@
 Another difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of hazard functions that skilfully predict patterns in mortality.
-Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate \cite{Kobe:1995tw,Wyckoff:2002ul}, traits \cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size \cite{Vieilledent:2010fv}.
+Past research has identified a range of factors correlated with tree survival, including an individual's growth rate \cite{Kobe:1995tw,Wyckoff:2002ul}, and size \cite{Vieilledent:2010fv} as well as species-level traits \cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}.
 However, these influences have not yet been integrated into a common hazard function \cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.
@@
 
-A specific challenge in developing hazard functions for plants is to estimate the relative contribution of growth-dependent and growth-independent hazards on an individual's overall hazard.
-While plants die via many causes, these broadly fall into two categories.
+A specific challenge in developing hazard functions for plants is to estimate the independent contributions of growth-dependent and growth-independent hazards on an individual's overall hazard.
+While plants die via many causes, these broadly fall into two main types of hazards.
 The first are growth-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair.
 The second are growth-independent hazards, where plants die because of stochastic events, irrespective of their growth rate, such as windfall or fire.
-An individual's total hazard is the sum of growth-independent and growth-dependent components.
+An individual's total hazard is therefore the sum of these independent hazards.
 
-A further challenge in developing hazard functions for plants is to integrate the conflicting relationships observed within and across species between mortality and growth.
+A further challenge in developing hazard functions for plants is the need to integrate the conflicting relationships observed within and across species between mortality and growth.
 Within species, mortality rates are lower for fast-growing individuals, presumably because those individuals have superior carbon budgets and are thus able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
-Empirical studies broadly support this theory, with many indicating exponential declines in mortality with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv}.
-By contrast, across species, there is a strong trade-off between growth and survival, with individuals from faster-growing species  exhibiting higher mortality rates than individuals from slower-growing species \cite{Poorter:2008iu, Wright:2010fl}.
-This trade-off may arise via traits with antagonistic effects, such as wood density.
-Denser wood -- which is expensive to build and thus slows growth --  reduces stem breakage \cite{vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, pathogen attack \cite{Augspurger:1984wx}, and thereby mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+Empirical studies broadly support this theory, with many indicating exponential-like declines in mortality with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv}.
+By contrast, across species, there is a strong trade-off between growth and survival, with faster-growing species exhibiting higher mortality rates relative to slower-growing species \cite{Poorter:2008iu, Wright:2010fl}.
+This trade-off may arise via species traits with antagonistic effects, such as wood density.
+Denser wood -- which is expensive to build and thus slows individual growth --  reduces stem breakage \cite{vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, pathogen attack \cite{Augspurger:1984wx}, and thereby mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
 
@@
 
-Here, we attempt to reconcile intra- and inter-specific factors as well as partition instantaneous mortality rates into growth-dependent and growth-independent rates.
-We achieve this by evaluating the following, unified hazard function, incorporating individual-, species-, and census-level effects:
+Here, we attempt to reconcile intra- and inter-specific variation as well as account for growth-dependent and growth-independent hazards.
+We achieve this by evaluating the following, unified hazard function, incorporating individual-, species-, and temporal effects:
 \begin{equation} \label{eq:model}
@@
 Eq.
-\ref{eq:model} allows for additive growth-independent and growth-dependent effects, includes a negative exponential effect of growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
+\ref{eq:model} allows for additive growth-independent and growth-dependent effects, includes a negative exponential effect of individual's growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
 Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$.
-Here we include an effect of a species trait (wood density, $\rho$), as well as a species random effect that captures any remaining species-level differences not accounted for by wood density (for details see Methods).
+Here we include the effects of three species traits (wood density, $\rho$, shade-intolerance, $\upsilon$, and maximum dbh, $\psi$), as well as a species random effect that captures any remaining species-level differences that is not otherwise accounted for by these traits (for details see Methods).
 
-To validate this model, we fit a series of models with increasing complexity (eq.
-\ref{eq:model} being the most complex) and compare their skill in predicting patterns of tropical tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_dbh)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_dbh)} tree species at Barro Colorado Island (BCI), Panama (Fig.
+To validate this model, we fit a series of other commonly used parametric hazard functions and compare their skill in predicting tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_dbh)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_dbh)} tree species at Barro Colorado Island (BCI), Panama (Fig.
 \ref{fig1}).
 The data are repeat censuses of stem diameter and tree status (alive vs.
-dead) taken over a 15 yr period (Fig.
+dead) taken approximately at 5-year intervals over a 15 yr period (Fig.
 \ref{fig1}A).
@@
 Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations.
-We therefore fit models across five iterative stages of model development.
-At each stage, the best model from the previous stage was taken as input for the next stage.
+We therefore fit models across seven iterative stages of model development.
 
-In stage 1, we ask whether mortality rates vary substantially between censuses, to establish whether a census effect is required.
-The null model is a constant, invariant with respect to species, growth rate, or year.
-In stage 2, we assess whether a hazard function including both growth-dependent and growth-independent terms outperform a function including only one of these.
+In stage 1, we ask whether mortality rates vary substantially among censuses, and thus, whether a census effect is required.
+We achieved this by fitting a constant that is invariant with respect to species, growth rate, or census period (i.e.
+a null model) and compared it to the same model with a census random effect.
+In stage 2, we assess whether a function incorporating both growth-dependent and growth-independent hazards outperform functions that only account for one or the other of these.
 Species effects were excluded, so we are simply asking which of three functional forms for $\lambda_i(t)$ (Fig.
-\ref{fig1}B) best predicts the data.
+\ref{fig1}B) best predicts the survival data.
 The simplest form (Fig.
@@
 \ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, asymptoting at some baseline hazard $\textgreater 0$.
-Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared.
+Combined, these three models capture a variety of functional responses previously proposed (e.g.
+\cite{Kobe:1995tw, Wyckoff:2000un}), including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared.
 We also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.

### 2018-07-16 R/table_1.R
Table to replace the explained variance figure of the old manuscript

@@ -1,0 +1,135 @@
+# Prepare table 1 for manuscript
+#' 
+# Prepare table 1 for manuscript
+#' @param model List containing stan fit object
+#' @param data Dataframe used to fit model
+#' @return Dataframe
+#' @details This function examines how both R2 and AUC change when certain parameters are removed from the model.
+#' R2 is calculated as a function of a the modelled mean probability of death across individuals for a given species
+#' against the observed proportion of individuals that died per species. AUC simply compares the raw observations against the model
+#' predicted probabilities for each individual.
+#' @author James Camac (\email{james.camac@gmail.com}) & Daniel Falster (\email{daniel.falster@unsw.edu.au})
+#' @export
+table_1 <- function(model, data) {
+  
+  `%>%` <- magrittr::`%>%`
+  
+  # Extract mean parameter estimates
+  mu_params <- summarise_hyper_params(model, data) %>%
+    dplyr::select(param, mean) %>%
+    filter(param %in% c("mu_log_alpha","mu_log_beta","mu_log_gamma")) %>%
+    mutate_at(vars(mean),funs(exp)) %>%
+    mutate(param = paste0("mu_",sub('.*\\_', '', param))) %>%
+    tidyr::spread(param, mean)
+  
+  # Extract mean species parameter estimates
+  # These parameters willbe alpha beta and gamma. They encompass any trait effects
+  spp_params <- summarise_spp_params(model, data) %>%
+    dplyr::select(param,species,sp,mean) %>%
+    dplyr::mutate(sp = as.character(sp),
+                  species = as.character(species)) %>%
+    dplyr::distinct() %>%
+    tidyr::spread(param,mean)
+  
+  # Extract trait effect
+  traits <- summarise_trait_params(model) %>%
+    dplyr::select(param, mean) %>%
+    tidyr::spread(param, mean)
+  
+  # Extract census effects
+  census_error <- as.data.frame(rstan::extract(model$fits[[1]], pars=c('census_err'))$census_err) %>%
+    dplyr::rename(`1` = V1, `2`= V2, `3` = V3) %>%
+    tidyr::gather(censusid, estimate) %>%
+    dplyr::group_by(censusid) %>%
[…]

### 2018-07-16 ms/manuscript.Rnw
Few extra edits

@@
 In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive a instantaneous mortality function for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
-Specifically, we develop a mortality function that not only integrates individual, species, and temporal effects, but also partitions the contributions of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate.
+Specifically, we develop an empricial mortality function that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the contributions of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate.
 We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and tolerance to shade outperform alternative formulations.

### 2018-07-17 ms/manuscript.Rnw
Changes to figures and manuscript

--- 2018-07-16 ms/manuscript.Rnw
@@
 \significancestatement{Tree mortality is a fundamental demographic process affecting forest dynamics and carbon cycling.
-Here, for the first time, we build a statistical mortality model that not only integrates intra-specific and inter-specific variation but also simulataneously estimates the independent contributions of two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth.
+Here, for the first time, we build a statistical mortality model that not only integrates intra-specific and inter-specific variation but also, for the first time, simulataneously estimates the independent contributions of two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth.
 We parameterise this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals encompassed within 203 species.
@@
 Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities.
-While past research has identified factors influencing tree mortality across a variety of scales, these distinct drivers are yet to be integrated within a unified predictive framework.
-In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive a instantaneous mortality function for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
-Specifically, we develop a mortality function that not only integrates individual, species, and temporal effects, but also partitions the contributions of growth-dependent and growth-independent hazards on the overall instantaneous mortality rate.
+While past research has identified variety of factors influencing tree mortality at both individual and species scales, these distinct drivers are yet to be integrated within a unified predictive framework.
+In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive hazard functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
+Specifically, we develop an empricial mortality model that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the independent contributions of growth-related and growth-independent hazards on the overall mortality rates.
 We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and tolerance to shade outperform alternative formulations.
-Moreover, we show that while these three traits can affect both growth-dependent and growth-independent hazards, they explain only only a small fraction of the total variability observed among species.
+Moreover, we show that while these three traits do affect both growth-dependent and growth-independent hazards, they account for only a small fraction of the total variability observed among species.
 Combined, this study provides a statistical framework for predicting individual-level mortality in highly diverse tropical forests.
@@
 In models of forest dynamics, variation in tree mortality rates has been shown to have significantly larger impacts on forest structure relative to variation in absolute growth rates \cite{Johnson-2016}.
-Improving our understanding of the mortality process is therefore a priority for making accurate predictions about population, carbon and nutrient dynamics of forests; especially in an era of rapid environmental change.
+Improving our understanding of the mortality process is therefore a priority for making accurate predictions about population, carbon and nutrient dynamics of forests; especially in an era of increased climate-driven forest die-off \cite{Anderegg:2012ep}.
 
@@
 This requirement arises in part because the observed outcome -- alive vs.
-dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$, also known as the ``hazard function''.
+dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the mortality rate for individual $i$ at time $t$, also known as the ``hazard function or hazard rate''.
 In classic survival analysis \cite{Kleinbaum:2012tn}, the probability an individual dies between times $t_1$ and $t_2$ is a function of both $\lambda_i(t)$, and the amount of time lapsed between $t_1$ and $t_2$:
@@
 \end{equation}
-As trees are long-lived and we are trying to estimate the shape of a continuous hazard function ($\lambda_i$) from binary data, large sample sizes collected over long durations are required.
+
+As trees are long-lived and we are trying to estimate the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes collected over long durations are required.
 Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.
@@
 Another difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of hazard functions that skilfully predict patterns in mortality.
-Past research has identified a range of factors correlated with tree survival, including an individual's growth rate \cite{Kobe:1995tw,Wyckoff:2002ul}, and size \cite{Vieilledent:2010fv} as well as species-level traits \cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}.
-However, these influences have not yet been integrated into a common hazard function \cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.
+Past research has identified a range of factors correlated with tree survival, including an individual's growth rate \cite{Kobe:1995tw,Wyckoff:2002ul}, and size \cite{Vieilledent:2010fv} as well as species-level traits such as wood density (REFS), maximum size (REFS) and measures of shade tolerance (REFS) \cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}.
+However, these intra- and inter-specific effects have not yet been integrated into a common hazard function that accounts for the two fundamental hazards faced by plants -- death due to low carbon budget and death due to stochastic events that are unrelated to carbon budget (e.g.
+fire, windthrow) \cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.
 
@@
 
-A specific challenge in developing hazard functions for plants is to estimate the independent contributions of growth-dependent and growth-independent hazards on an individual's overall hazard.
-While plants die via many causes, these broadly fall into two main types of hazards.
-The first are growth-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair.
-The second are growth-independent hazards, where plants die because of stochastic events, irrespective of their growth rate, such as windfall or fire.
-An individual's total hazard is therefore the sum of these independent hazards.
+A specific challenge in developing biologically plausible hazard functions for plants is the need to estimate the independent contributions of carbon-dependent and carbon-independent hazards on an individual's overall hazard.
+While plants die via many causes, these broadly fall into two main hazards.
+The first are carbon-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair.
+The second are carbon-independent hazards, where plants die because of stochastic events that are unrelated to their growth rate, such as windfall or fire.
+An individual's total hazard at a given time point is therefore the sum of these two independent hazards, and is contrary to many traditional survival analyses done for plants which assume these hazards are proportional to eachother (REF).
 
@@
 By contrast, across species, there is a strong trade-off between growth and survival, with faster-growing species exhibiting higher mortality rates relative to slower-growing species \cite{Poorter:2008iu, Wright:2010fl}.
-This trade-off may arise via species traits with antagonistic effects, such as wood density.
+This trade-off between growth and survival may arise via species traits with antagonistic effects, such as wood density.
 Denser wood -- which is expensive to build and thus slows individual growth --  reduces stem breakage \cite{vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, pathogen attack \cite{Augspurger:1984wx}, and thereby mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
@@
 
-Here, we attempt to reconcile intra- and inter-specific variation as well as account for growth-dependent and growth-independent hazards.
-We achieve this by evaluating the following, unified hazard function, incorporating individual-, species-, and temporal effects:
+Here, we attempt to develop a empirical model capable of reconciling these conflicting growth-mortality relationships observed among individuals and between species while also explicitly estimating the independent contributions of carbon-dependent and independent hazards.
+We achieve this by evaluating the following, unified hazard function that incorporates individual-, species-, and temporal effects:
 \begin{equation} \label{eq:model}
@@
 Eq.
-\ref{eq:model} allows for additive growth-independent and growth-dependent effects, includes a negative exponential effect of individual's growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
+\ref{eq:model} allows for the independent effects of carbon-dependent and carbon-independent hazards, includes a negative exponential effect of individual's growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
 Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$.
-Here we include the effects of three species traits (wood density, $\rho$, shade-intolerance, $\upsilon$, and maximum dbh, $\psi$), as well as a species random effect that captures any remaining species-level differences that is not otherwise accounted for by these traits (for details see Methods).
+Here we include the effects of three species traits thought to be major drivers of inter-specific variation in tropical tree mortality -- wood density ($\rho$), shade-intolerance ($\upsilon$) and maximum dbh ($\psi$).
+We also include species random effects that captures any remaining species-level differences that are not otherwise accounted for by species traits (for details see Methods).
 
-To validate this model, we fit a series of other commonly used parametric hazard functions and compare their skill in predicting tree mortality for \Sexpr{n_inds(BCI_model_dataset_true_dbh)} individuals from \Sexpr{n_spp(BCI_model_dataset_true_dbh)} tree species at Barro Colorado Island (BCI), Panama (Fig.
+To validate this model, we fit a series of other commonly used parametric hazard functions and compare their skill in predicting tree mortality for \Sexpr{n_inds(BCI_training_full)} individuals from \Sexpr{n_spp(BCI_training_full)} tree species at Barro Colorado Island (BCI), Panama (Fig.
 \ref{fig1}).
@@
 \ref{fig1}A).
-In total \Sexpr{n_obs(BCI_model_dataset_true_dbh)} observations were used to fit these models.
+In total \Sexpr{n_obs(BCI_training_full)} observations were used to fit these models.
 We compare the skill of different hazard functions in predicting outcomes in novel data (i.e.
@@
 We achieved this by fitting a constant that is invariant with respect to species, growth rate, or census period (i.e.
-a null model) and compared it to the same model with a census random effect.
+the null model) and compared it to the same model but with a census random effect.
 In stage 2, we assess whether a function incorporating both growth-dependent and growth-independent hazards outperform functions that only account for one or the other of these.
@@
 Comparing the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates progressively decreasing over time.
-The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 3)}\%).
+The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%).
 Consequently, when we allowed hazard rates to be scaled by census (i.e.
@@
 
-\begin{figure}[!h]
+\begin{figure*}[!ht]
 \includegraphics{figures/fig2.pdf}
@@
 \label{fig2}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[h!]
+\begin{figure*}[h!]
 \includegraphics{figures/fig3.pdf}
@@
 \label{fig3}
-\end{figure}
+\end{figure*}
 
-\subsection*{Wood density and other species effects}
+\subsection*{Species traits}
 Including wood density as an effect on either growth-independent or growth-dependent parameters significantly improved model skill relative to a model without such effects (Fig.
@@
 
-\begin{figure}[t]
+To determine whether species-level parameters were associated with their maximum size or light requirement, we conducted a post-hoc analysis comparing the fitted parameters to these traits.
+We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the growth-independent hazard term, were weakly and positively correlated with gap index (a measure of species light requirement).
+By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index.
+These correlations suggest that species which predominately recruit in gaps are more prone to death, due to both low growth and stochastic chance.
+They also required faster growth rates to achieve a given mortality rate relative to a species that recruit in shade.
+Correlations between estimated species parameters and their associated maximum DBH were weak).
+
+\begin{figure}[h!]
 \includegraphics{figures/fig4.pdf}
-\caption{Predicted changes in instantaneous mortality rate (A,B), and 1-yr mortality probability (C,D), with increasing growth rate.
+\caption{Predicted changes in mortality rate (A,B), and 1-yr mortality probability (C,D), with increasing growth rate.
 A,C) Estimated curves for all 203 BCI species used in this study.
@@
 \subsection*{Parameter contribution to predicted variation}
-To compliment the analysis, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1 year for all individuals (Fig.
-\ref{fig5}).
+To compliment the analysis, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1 year for all individuals (Table XX).
 We achieved this by removing each parameter and calculating the reduction in the sum of squares for this model relative to the full model.
-Species effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr mortality.
-Wood density explained \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of this species variation.
-Census accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation.
-The growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total variation, while growth-independent hazards accounted for the remainder (Fig.
-\ref{fig5}).
+Species effects contributed...
 
-\subsection*{Mortality rates, maximum species size and species light requirements}
-To determine whether species-level parameters were associated with their maximum size or light requirement, we conducted a post-hoc analysis comparing the fitted parameters to these traits.
-We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the growth-independent hazard term, were weakly and positively correlated with gap index (a measure of species light requirement; Fig.
-\ref{fig6} A,C,E).
-By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index.
-These correlations suggest that species which predominately recruit in gaps are more prone to death, due to both low growth and stochastic chance.
-They also required faster growth rates to achieve a given mortality rate relative to a species that recruit in shade.
-Correlations between estimated species parameters and their associated maximum DBH were weak (Fig.
-\ref{fig6}).
-
-\begin{figure}[t]
-\includegraphics{figures/fig5.pdf}
-\caption{Proportion of variation in predicted 1-year mortality for all individuals captured by different effects.
+% \begin{figure}[t]
+% \includegraphics{figures/fig5.pdf}
+% \caption{Proportion of variation in predicted 1-year mortality for all individuals captured by different effects.
 Effects are not mutually exclusive thus sum to more than 1.0.
 Note "wood density" is a subset of the overall "species" effect.}
-\label{fig5}
-\end{figure}
+% \label{fig5}
+% \end{figure}
 
@@
 Like many other studies \cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlight the importance of light competition in influencing tropical tree demographic rates.
-Specifically, we found that the growth-dependent hazard accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total predicted variability in mortality rates (Fig.
-\ref{fig5}).
+Specifically, we found that the growth-dependent hazard accounted for XX of the total predicted variability in mortality rates (Table X).
 This suggests deficiencies in carbon budget are a major contributor to tree death on BCI.
@@
 While wood density effects are now being incorporated in mortality algorithms of many vegetation models \cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation (Fig.
-\ref{fig4}-\ref{fig5}).
+\ref{fig4}, Table X).
 This means that such models are likely to severely underestimate the true variability in mortality rates.
@@
 
-\begin{figure}[t]
-\includegraphics{figures/fig6.pdf}
-\caption{Posthoc correlations between estimated species parameters and measures of species light requirement (left) and species maximum DBH (right).
-Points are means ($\pm$ 95\% credible intervals) for each of the 203 species used in this study.
-Blue line shows the average trendline as determined from standard linear regression.
-Gap index is a species level index of the light required for recruitment.
-Low gap index indicates the species readily recruits under dense canopy and 1 indicates that a species only ever recruits in gaps.}
-\label{fig6}
-\end{figure}
-
-We found that wood density only explained \Sexpr{extract_prop_explained(param_variance_explained,I('wood_density'))}\% of the total \Sexpr{extract_prop_explained(param_variance_explained,I('species'))}\% variation explained by species effects, suggesting that other traits are also affecting tree mortality rates.
+We found that wood density only explained XX of the total XX variation explained by species effects, suggesting that other traits are also affecting tree mortality rates.
 Post-hoc analyses suggest this unexplained variation was in part related to a species' light demand, but not its maximum height.
 Both traits have been proposed major axes of inter-specific variation in tropical rainforests \cite{Poorter:2006tq} (Fig.
-\ref{fig6}).
+XX).
 Specifically, light demanding species (i.e.
@@
 We also excluded individuals that: 1) did not survive at least two censuses (two being required to estimate growth rate); 2) were not consistently measured at 1.3 m above ground; 3) were multi-stemmed; 4) resprouted or seemingly ``returned from the dead''; or 5) were extreme outliers -- stems which grew more than 5 cm yr$^{-1}$ or shrunk more than 25\% of their initial diameter.
-In total \Sexpr{n_obs(BCI_model_dataset_true_dbh)} observations were used in this study comprising \Sexpr{n_inds(BCI_model_dataset_true_dbh)} individual trees and \Sexpr{n_spp(BCI_model_dataset_true_dbh)} species.
-Because of computational costs, the models fit in this study do not include individual random effects, as this would require estimation of an additional \Sexpr{n_inds(BCI_model_dataset_true_dbh)} parameters.
+In total \Sexpr{n_obs(BCI_training_full)} observations were used in this study comprising \Sexpr{n_inds(BCI_training_full)} individual trees and \Sexpr{n_spp(BCI_training_full)} species.
+Because of computational costs, the models fit in this study do not include individual random effects, as this would require estimation of an additional \Sexpr{n_inds(BCI_training_full)} parameters.
 Instead, our models assume that repeat measurements of an individual are independent of one another.
@@
 Growth rates were estimated from field measurements of diameter, which inevitably include observation error.
-In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_dbh)}\% of estimated growth rates were negative.
+In our dataset, \Sexpr{prop_neg_growth(BCI_training_full)}\% of estimated growth rates were negative.
 To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see Supplementary Material S1 for details; Fig.

### 2018-07-17 R/plot_fig4.R

@@ -14,12 +14,65 @@ plot_fig4 <- function(model,
                       growth_range = c(0.03,0.5),
                       hazard_curve = FALSE) {
   
-  p1 <- plot_spp_curves(model, data, growth_range, trait_name = "wood_density", hazard_curve, xlab= NULL, ylab=expression("1-yr mortality probability"), legend_label = expression(atop("Wood density",(g~cm^-3))))
-  p2 <- plot_mu_curves(model, growth_range, trait_name = "wood_density", trait_values = c(0.2,0.8), hazard_curve, xlab=NULL, ylab = NULL, legend_label = expression(atop("Wood density",(g~cm^-3))))
-  p3 <- plot_spp_curves(model, data, growth_range, trait_name = "dbh_95", hazard_curve, xlab= NULL, ylab=expression("1-yr mortality probability"), legend_label = expression(atop("Max dbh","(cm)")))
-  p4 <- plot_mu_curves(model, growth_range, trait_name = "dbh_95", trait_values = c(3,180),hazard_curve, xlab=NULL, ylab = NULL, legend_label = expression(atop("Max dbh","(cm)")))
-  p5 <- plot_spp_curves(model, data, growth_range, trait_name = "gap_index", hazard_curve, xlab= expression("Annual dbh growth"~("cm yr"^-1)), ylab=expression("1-yr mortality probability"), legend_label = expression(atop("Shade","intolerance")))
-  p6 <- plot_mu_curves(model, growth_range, trait_name = "gap_index", trait_values = c(0.5, 0.8), hazard_curve, xlab=expression("Annual dbh growth"~("cm yr"^-1)), ylab = NULL, legend_label = expression(atop("Shade","intolerance"))) 
+  p1 <- plot_spp_curves(model, 
+                        data, 
+                        growth_range, 
+                        trait_name = "wood_density", 
+                        hazard_curve, 
+                        colour_by_trait = FALSE, 
+                        xlab= NULL, 
+                        ylab=("1-yr mortality probability"))
   
-  cowplot::plot_grid(p1,p2,p3,p4,p5,p6, ncol=2, labels=LETTERS[1:6], label_size = 7)
+  p2 <- plot_mu_curves(model, 
+                       growth_range, 
+                       trait_name = "wood_density", 
+                       trait_values = c(0.2,0.8), 
+                       hazard_curve, 
+                       xlab=NULL, 
+                       ylab = ("1-yr mortality probability"), 
[…]

### 2018-07-17 R/plot_mu_curves.R

@@ -30,9 +30,9 @@ plot_mu_curves <- function(model,
   
   p1 <- ggplot2::ggplot(preds, ggplot2::aes(x = dbh_growth,y = mean, group = type, colour = factor(trait_value), fill=factor(trait_value))) + 
     ggplot2::geom_line() +
-    viridis::scale_color_viridis(name=legend_label, discrete = TRUE) +
+    ggplot2::scale_colour_manual(name=legend_label, values = c("blue","red")) +
     ggplot2::geom_ribbon(ggplot2::aes(ymin = `2.5%`,ymax = `97.5%`), alpha=0.5, colour=NA) +
-    viridis::scale_fill_viridis(legend_label, discrete=TRUE) +
+    ggplot2::scale_fill_manual(legend_label,values = c("blue","red")) +
     ggplot2::scale_x_continuous(expand=c(0,0)) +
     ggplot2::xlab(xlab)
   

### 2018-07-17 R/plot_spp_curves.R

@@ -6,6 +6,7 @@
 #' @param growth_range Numeric vector. The growth range to predict curve over
 #' @param trait_name Character. Name of trait to use as colour gradient
 #' @param hazard_curve Logical. Whether to plot hazard (TRUE) or probability (FALSE) curve
+#' @param colour_by_trait Logical. If TRUE species curves will be coloured based on species trait value. Otherwise not.
 #' @param ylab Character. Y axis title
 #' @param xlab Character. X axis title
 #' @param legend_position The position of legends ("none", "left", "right", "bottom", "top", or two-element numeric vector)
@@ -19,6 +20,7 @@ plot_spp_curves <- function(model,
                             growth_range = c(0.03,0.5), 
                             trait_name, 
                             hazard_curve = FALSE, 
+                            colour_by_trait = TRUE,
                             ylab= "Trait value", 
                             xlab= expression("Annual dbh growth"~("cm yr"^-1)), 
                             legend_position = "right",
@@ -27,44 +29,53 @@ plot_spp_curves <- function(model,
   `%>%` <- magrittr::`%>%`
   
   preds <- predict_spp_hazard_curves(model, data, growth_range) %>%
-    filter(trait == trait_name)
+    dplyr::filter(trait == trait_name)
   
-  base::switch(trait_name,
-               "wood_density" = {
-                 legend_limit = c(floor(min(preds$trait_value)/0.1)*0.1,ceiling(max(preds$trait_value)/0.1)*0.1)
-                 legend_breaks <- pretty(legend_limit, n = 5,min.n = 5)
-               },
-               "dbh_95" = {
-                 legend_limit = c(1,ceiling(max(preds$trait_value)/100)*100)
-                 legend_breaks <- pretty(legend_limit, n = 5,min.n = 5)
-               },
-               "gap_index" = {
-                 legend_limit = c(floor(min(preds$trait_value)/0.1)*0.1,ceiling(max(preds$trait_value)/0.1)*0.1)
-                 legend_breaks <- pretty(legend_limit, n = 5,min.n = 5)
-               })
+  if(isTRUE(colour_by_trait)) {
+    base::switch(trait_name,
+                 "wood_density" = {
[…]

### 2018-07-17 R/predict_spp_hazard_curves.R

@@ -16,7 +16,7 @@ predict_spp_hazard_curves <- function(model, data, growth_range = c(0.03,0.5)) {
                                    dbh_growth_centered = seq(min(growth_range),max(growth_range),length.out = 100) - 0.172)
   
   res <- spp_parameters %>%
-    select(-c(median,sd,`2.5%`,`97.5%`)) %>%
+    dplyr::select(-c(median,sd,`2.5%`,`97.5%`)) %>%
     tidyr::spread(param, mean) %>%
     base::merge(growth_rates) %>%
     dplyr::mutate(

### 2018-07-17 ms/manuscript.Rnw
Some more work on the front end

@@
 dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the mortality rate for individual $i$ at time $t$, also known as the ``hazard function or hazard rate''.
-In classic survival analysis \cite{Kleinbaum:2012tn}, the probability an individual dies between times $t_1$ and $t_2$ is a function of both $\lambda_i(t)$, and the amount of time lapsed between $t_1$ and $t_2$:
+In classic survival analysis \cite{Kleinbaum:2012tn, Zens:2003jv}, the probability an individual dies between times $t_1$ and $t_2$ is a function of both $\lambda_i(t)$, and the amount of time lapsed between $t_1$ and $t_2$:
 \begin{equation} \label{eq:mortality_model}
@@
 
-Another difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of hazard functions that skilfully predict patterns in mortality.
-Past research has identified a range of factors correlated with tree survival, including an individual's growth rate \cite{Kobe:1995tw,Wyckoff:2002ul}, and size \cite{Vieilledent:2010fv} as well as species-level traits such as wood density (REFS), maximum size (REFS) and measures of shade tolerance (REFS) \cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}.
-However, these intra- and inter-specific effects have not yet been integrated into a common hazard function that accounts for the two fundamental hazards faced by plants -- death due to low carbon budget and death due to stochastic events that are unrelated to carbon budget (e.g.
-fire, windthrow) \cite{Hawkes:2000ib,Zens:2003jv,McDowell:2011dr}.
-
 \section*{Towards a unified model of tree mortality}
 
-A specific challenge in developing biologically plausible hazard functions for plants is the need to estimate the independent contributions of carbon-dependent and carbon-independent hazards on an individual's overall hazard.
-While plants die via many causes, these broadly fall into two main hazards.
-The first are carbon-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair.
-The second are carbon-independent hazards, where plants die because of stochastic events that are unrelated to their growth rate, such as windfall or fire.
-An individual's total hazard at a given time point is therefore the sum of these two independent hazards, and is contrary to many traditional survival analyses done for plants which assume these hazards are proportional to eachother (REF).
-
-A further challenge in developing hazard functions for plants is the need to integrate the conflicting relationships observed within and across species between mortality and growth.
+A major challenge in developing hazard functions for plants is the need to integrate the conflicting relationships observed within and across species between mortality and growth.
 Within species, mortality rates are lower for fast-growing individuals, presumably because those individuals have superior carbon budgets and are thus able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
@@
 Denser wood -- which is expensive to build and thus slows individual growth --  reduces stem breakage \cite{vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, pathogen attack \cite{Augspurger:1984wx}, and thereby mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
+However, to date, these intra- and inter-specific effects have not yet been integrated into a single hazard function in such a way that it allows one to examine their effects on the two fundamental hazards faced by plants.
+
+Plants die via many causes, however, these broadly fall into two main categories \cite{Hawkes:2000ib,McDowell:2011dr}.
+The first are carbon-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair.
+The second are carbon-independent hazards, where plants die because of chance events that are unrelated to their growth rate, such as windfall or fire.
+An individual's total hazard at a given time point is therefore the sum of these two independent hazards.
+However, to date, statistical models of plant mortality have not attempted to simulataneously estimate these hazards from standard survival data, let alone examine how intra- and inter-specific factors affect them.
 
@@
 
-To validate this model, we fit a series of other commonly used parametric hazard functions and compare their skill in predicting tree mortality for \Sexpr{n_inds(BCI_training_full)} individuals from \Sexpr{n_spp(BCI_training_full)} tree species at Barro Colorado Island (BCI), Panama (Fig.
+To validate this model, we also fit two alternative hazard functions: 1) a baseline hazard, 2) a growth-dependent hazard, and then compare their skill in predicting the observed mortality for \Sexpr{n_inds(BCI_training_full)} individual from \Sexpr{n_spp(BCI_training_full)} tree species at Barro Colorado Island (BCI), Panama (Fig.
 \ref{fig1}).

### 2018-07-20 ms/manuscript.Rnw
More edits to improve readability

@@
 Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities.
-While past research has identified variety of factors influencing tree mortality at both individual and species scales, these distinct drivers are yet to be integrated within a unified predictive framework.
+While past research has identified a variety of factors influencing tree mortality at both individual and species scales, these distinct drivers are yet to be integrated within a unified predictive framework.
 In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive hazard functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
@@
 This variation has considerable consequences for forest structure and dynamics.
-For example, the death of a single large tree can transfer up to 20000 kg of carbon from living to decaying carbon pools \cite{Chave:2005hz} and creates a gap in the canopy that can restart a successional race, during which 100's of plants may die while competing for a spot in the sun.
+For example, the death of a single large tree can transfer up to 20,000 kg of carbon from living to decaying carbon pools \cite{Chave:2005hz} and creates a gap in the canopy that can restart a successional race, during which 100's of plants may die while competing for a spot in the sun.
 In models of forest dynamics, variation in tree mortality rates has been shown to have significantly larger impacts on forest structure relative to variation in absolute growth rates \cite{Johnson-2016}.
@@
 
+Plants die because of many causes, however, these broadly fall into two main categories \cite{Hawkes:2000ib,McDowell:2011dr}.
+The first are carbon-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair.
+The second are carbon-independent hazards, where plants die because of chance events that are unrelated to their carbon budget, such as windfall or fire.
+An individual's total hazard at a given time point is therefore the sum of these two independent hazards.
+However, to date, most statistical models of tree mortality do not explicitly estimate the independent effects of these two fundamanetal hazards faced by plants.
+Instead, plant mortality is typically modelled using: 1) logistic regression (i.e.
+with logit link), which does not estimate a rate, but rather the log odds of death at a given time point; or 2) via a proportional hazards approach such as a Cox proportional-hazard model or complementary-log-log link binomial regression, both of which assume hazards are multiplicative, and therefore, not independent of one another \cite{Kleinbaum:2012tn}.
+
+There are several distinct advantages of modelling tree mortality as a function of two additive, and thus independent hazards: 1) The model explicitly accounts for the independent effects of the two fundamental types of hazards faced by plants; 2) it allows for the examination of the relative contributions each hazard provides to the overall hazard rate; 3) it allows for examination of how various factors affect each hazard term, and because of this; 4) allows ecological theories of how factors such as species traits affect plant mortality to be empirically tested.
+
 A major challenge in developing hazard functions for plants is the need to integrate the conflicting relationships observed within and across species between mortality and growth.
@@
 
-Plants die via many causes, however, these broadly fall into two main categories \cite{Hawkes:2000ib,McDowell:2011dr}.
-The first are carbon-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair.
-The second are carbon-independent hazards, where plants die because of chance events that are unrelated to their growth rate, such as windfall or fire.
-An individual's total hazard at a given time point is therefore the sum of these two independent hazards.
-However, to date, statistical models of plant mortality have not attempted to simulataneously estimate these hazards from standard survival data, let alone examine how intra- and inter-specific factors affect them.
-
 \begin{figure}[!hb]
@@
 A) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
-B) We consider three alternative hazard functions: 1) a baseline hazard, 2) a growth-dependent hazard; and 3) a function that combines both baseline and growth-dependent hazards.
-The parameters of the models are biologically interpretable: $\alpha$ defines the instantaneous mortality rate at low growth rate; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote, or baseline hazard.
-Combined $\alpha$ and $\beta$ capture growth-dependent mortality, while $\gamma$ captures growth-independent mortality (e.g.
+B) We consider three alternative hazard functions: 1) a growth-independent baseline hazard (i.e.
+average hazard), 2) a growth-dependent hazard; and 3) a function that combines both growth dependent and independent hazards.
+The parameters of the models are biologically interpretable: $\alpha$ defines the the low growth rate effect on the overall hazard; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote (Panel B; right), or baseline hazard (Panel B; left).
+In the most complex model (Panel B; right), $\alpha$ and $\beta$ capture growth-dependent hazards, while $\gamma$ captures growth-independent hazards (e.g.
 windfall, fire) that kill a plant, irrespective of its growth rate.
@@
 
-Here, we attempt to develop a empirical model capable of reconciling these conflicting growth-mortality relationships observed among individuals and between species while also explicitly estimating the independent contributions of carbon-dependent and independent hazards.
-We achieve this by evaluating the following, unified hazard function that incorporates individual-, species-, and temporal effects:
+Here, we develop a tree mortality model that not only reconciles the conflicting growth-mortality relationships observed among individuals and between species, but also explicitly accounts for carbon-dependent and carbon-independent hazards.
+We achieve this by evaluating the following unified hazard function that incorporates individual-, species-, and temporal effects:
 \begin{equation} \label{eq:model}
-\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s[i]}}^{\text{Growth independent}} + \overbrace{\alpha_{s[i]} \, \textrm{e}^{-\beta_{s[i]} \, X_i(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
+\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s[i]}}^{\text{Carbon independent}} + \overbrace{\alpha_{s[i]} \, \textrm{e}^{-\beta_{s[i]} \, X_i(t)}}^{\text{Carbon dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
 \end{equation}
 Eq.
-\ref{eq:model} allows for the independent effects of carbon-dependent and carbon-independent hazards, includes a negative exponential effect of individual's growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
+\ref{eq:model} allows for the additive, thus independent, effects of carbon-dependent and carbon-independent hazards, includes a negative exponential effect of individual's growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
 Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$.
-Here we include the effects of three species traits thought to be major drivers of inter-specific variation in tropical tree mortality -- wood density ($\rho$), shade-intolerance ($\upsilon$) and maximum dbh ($\psi$).
-We also include species random effects that captures any remaining species-level differences that are not otherwise accounted for by species traits (for details see Methods).
+Here we include the effects of three species traits thought to be major drivers of inter-specific variation in tropical tree mortality -- wood density ($\rho$), light requirement ($\upsilon$) and maximum dbh ($\psi$).
+We also include species random effects in order to capture species-level differences that are not otherwise accounted for by species traits (for details see Methods).
 
-To validate this model, we also fit two alternative hazard functions: 1) a baseline hazard, 2) a growth-dependent hazard, and then compare their skill in predicting the observed mortality for \Sexpr{n_inds(BCI_training_full)} individual from \Sexpr{n_spp(BCI_training_full)} tree species at Barro Colorado Island (BCI), Panama (Fig.
+To validate this model, we also fit two alternative hazard functions: 1) a baseline hazard (i.e.
+a null model) and 2) a growth-dependent only hazard.
+We then compare each functions ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from \Sexpr{n_spp(BCI_training_full)} species recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama (Fig.
 \ref{fig1}).
@@
 Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations.
-We therefore fit models across seven iterative stages of model development.
+We therefore fit models across six iterative stages of model development.
 
-In stage 1, we ask whether mortality rates vary substantially among censuses, and thus, whether a census effect is required.
-We achieved this by fitting a constant that is invariant with respect to species, growth rate, or census period (i.e.
-the null model) and compared it to the same model but with a census random effect.
-In stage 2, we assess whether a function incorporating both growth-dependent and growth-independent hazards outperform functions that only account for one or the other of these.
-Species effects were excluded, so we are simply asking which of three functional forms for $\lambda_i(t)$ (Fig.
-\ref{fig1}B) best predicts the survival data.
+In stage 1, we first ask whether average mortality rates vary substantially among censuses, and thus, whether a census effect is required.
+We achieved this by fitting a constant, $\gamma$, that is invariant with respect to species, growth rate, or census period (i.e.
+the null model) and compared it to a constant that is scaled by a census random effect, $\gamma\delta_{t}$.
+This stage revealed that there was significant variation in mortality rates observed among censuses, and as such a random effect for census was included as a scaling factor for all subsequent models.
+
+In stage 2, we assessed the predictive performance of three alternative hazard functions (Fig.
+\ref{fig1}B).
 The simplest form (Fig.
-\ref{fig1}B, left) assumes a constant growth-independent hazard rate.
+\ref{fig1}B, left) assumes a constant hazard.
 The second form (Fig.
 \ref{fig1}B, middle) assumes the risk of dying declines towards an asymptote of zero as growth increases.
-The third form (eq.
-Fig.
-\ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, asymptoting at some baseline hazard $\textgreater 0$.
-Combined, these three models capture a variety of functional responses previously proposed (e.g.
-\cite{Kobe:1995tw, Wyckoff:2000un}), including effects represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, which have not previously been systematically compared.
-We also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.
-In stage 3, we examine whether including wood density (a species-level trait) improved model skill, and if so, whether the effect of wood density was on growth-dependent, growth-independent, or both hazards.
-Finally, we fit a model that allowed parameters to vary by species-level variation that was otherwise not captured by wood density.
-This allowed us to ask what proportion of inter-specific hazard variability is explained by wood density.
-Using the final ``best'' model we also conducted post-hoc tests, to determine how species-level parameters were associated with their maximum size and light requirement.
+The third form (Fig.
+\ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike the second functional form examined here, asymptotes at some baseline hazard $\textgreater 0$.
+Combined, these three models capture a variety of functional responses that are both used in empirical studies of plant mortality \cite{Kobe:1995tw,Wyckoff:2000un} and are represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, but which have not previously been systematically compared.
+In doing this, we also also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.
+
+In the remaining stages, we then take the most predictive hazard function and growth rate measure from stage 2 and examine how species traits, wood density, light demand and maximum dbh, affected model predictive performance.
+First, we individually fit an effect of each trait on all parameters (stage 3), we then examine 2-trait combinations on all parameters (stage 4) as well as the effect of three traits on all parameters (stage 5).
+Lastly, we fit a model that had included a species random effect on each parameter (stage 6).
+This allowed us to 1) examine the maximum predictive capacity when accounting for species differences; and 2) provides a relative indication on how much of the total species variation is captured by these three traits.
+Because these models were computationally expensive to run (some requiring over 400 computing hours), we were unable to examine interactions among traits, or determine the optimal combination of traits with parameters.
 
@@
 \subsection*{Mortality over time}
-Comparing the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates progressively decreasing over time.
-The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%).
+Comparing the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates declining over time.
+The highest proportion of tree death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%).
 Consequently, when we allowed hazard rates to be scaled by census (i.e.
-adding term $\delta$) we observed a small, but significant, increase in predictive skill (Fig.
+adding the term $\delta$) we observed a small increase in predictive skill (Fig.
 \ref{fig2}A).
@@
 \includegraphics{figures/fig2.pdf}
-\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% credible intervals) logarithmic loss, with lower values implying greater predictive skill.
-A) Five sequential stages of model selection with each stage increasing in complexity.
-Null: constant ($\gamma$); Census: inclusion of census effects ($\delta_t$); Growth rate: the inclusion of growth rate.
-This includes two possible hazard functions: growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); WD: inclusion of wood density effect ($\rho$) on $\gamma$; Species: the inclusion of species random effects $s$ on all parameters.
-B) Predictive skill for wood density parameter combinations.
-Shading represent growth measure used: white = no growth measure, grey = basal area growth, black = dbh growth.
-For both panels symbols represent functional form: null (circle), growth-dependent hazard (triangle), both growth-independent and growth-dependent hazards (square).}
+\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% confidence intervals) logarithmic loss, with lower values implying greater predictive skill.
+A) Six sequential stages of model selection with each stage increasing in complexity.
+\textbf{Stage 1:} Baseline: constant ($\gamma$) vs Baseline with census effect ($\gamma\delta_t$); \textbf{Stage 2:} Best model of stage 1 compared to hazard functions that also account for individual growth rate.
+This includes two possible hazard functions: a growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters.
+Symbols represent the three hazard functional form: baseline (circle), growth-dependent (triangle), and hazard containing both growth-independent and growth-dependent hazards (square).}
 \label{fig2}
@@
 \subsection*{Hazard functions}
-Comparing the three hazard functions in Fig.
-\ref{fig1}B, we found that the third function --- with both growth-dependent and growth-independent terms, i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the growth-independent only (i.e.
-null) or growth-dependent only functions (Fig.
-\ref{fig2}A).
+Comparing the three hazard functions summarised in Fig.
+\ref{fig1}B, we found that the third function --- the function containing both growth-dependent and growth-independent hazard terms, i.e.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline hazard or growth-dependent only functions (Fig.
+\ref{fig2}).
 Moreover, we found that predictive skill was higher when using stem diameter growth over stem-area growth (Fig.
@@
 \includegraphics{figures/fig3.pdf}
-\caption{Relationship between species growth-independent mortality rate and wood density.
+\caption{Estimated relationships between Relationship between species growth-independent mortality rate and wood density.
 Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study.

### 2018-07-20 R/plot_fig2.R

@@ -28,11 +28,16 @@ plot_fig2 <- function(logloss_summaries) {
                                                                      "rho_size_all",
                                                                      "gap_size_all"), "2_traits"),
                   comparison = replace(comparison, comparison == "multi_trait_all", "3_traits"),
-                  comparison = replace(comparison, comparison =="function_growth_comparison" & model=="base_hazard", "census"),
+                  comparison = replace(comparison, comparison =="function_growth_comparison" & model=="base_hazard", "null_model"),
+                  stage = factor(comparison, 
+                                 levels=c("null_model","function_growth_comparison","1_trait","2_traits", 
+                                          "3_traits","species_random_effects"),
+                                 labels = c("bold(Stage~1)","bold(Stage~2)","bold(Stage~3)", 
+                                            "bold(Stage~4)", "bold(Stage~5)","bold(Stage~6)")),
                   comparison = factor(comparison, 
-                                      levels=c("null_model","census","function_growth_comparison","1_trait","2_traits", 
+                                      levels=c("null_model","function_growth_comparison","1_trait","2_traits", 
                                                "3_traits","species_random_effects"),
-                                      labels = c("Null","Census","Growth~rate","1~trait~(alpha~beta~gamma)", 
+                                      labels = c("Baseline","Growth~rate","1~trait~(alpha~beta~gamma)", 
                                                  "2~traits~(alpha~beta~gamma)", "3~traits~(alpha~beta~gamma)","Species")),
                   model_type = factor(model_type, levels = c("null_model_null_model_none",
                                                              "function_growth_comparison_base_hazard_none",
@@ -66,9 +71,10 @@ plot_fig2 <- function(logloss_summaries) {
                                        "gap_size_all_base_growth_hazard_gap_size_abc" = expression(psi + upsilon),
                                        "multi_trait_all_base_growth_hazard_rho_gap_size_abc" = expression(psi + rho + upsilon),
                                        "species_random_effects_base_growth_hazard_none" = expression((alpha[s]*"e"^{-beta[s]~"X"["i"]} + gamma[s])~delta["t"]))) +
-    ggplot2::facet_grid(.~comparison, scales='free_x', drop=TRUE, space = "free_x", labeller = ggplot2::label_parsed) +
[…]

### 2018-07-20 R/plot_fig3.R

@@ -9,19 +9,25 @@
 
 plot_fig3 <- function(spp_params, mu_trend) {
   
+  spp_params <- spp_params %>%
+    mutate(trait_value2 = ifelse(trait=="dbh_95", log10(trait_value),trait_value))
+  
+  mu_trend <- mu_trend %>%
+    mutate(trait_value2 = ifelse(trait=="dbh_95", log10(trait_value),trait_value))
+  
   breaks <- c(0.001,0.01,0.1, 1, 10)
   labels <- sapply(log10(breaks),function(i) as.expression(bquote(10^ .(i))))
-  trait_label_lookup <- c(wood_density ="Wood~density~(g~cm^-3)", gap_index = "Shade~intolerance", dbh_95 = "Max~dbh~(cm)")
+  trait_label_lookup <- c(wood_density ="Wood~density~(g~cm^-3)", gap_index = "Shade~intolerance", dbh_95 = "log10(Max~dbh~(cm))")
   param_label_lookup <- c(alpha = "Low~growth~effect~(alpha)", beta = "Growth~decay~rate~(beta)", gamma ="Baseline~hazard~(gamma)")
   
-  ggplot2::ggplot(spp_params, ggplot2::aes(x = trait_value,y = mean)) + 
-    ggplot2::geom_ribbon(data = mu_trend, ggplot2::aes(x = trait_value,ymin = `2.5%`,ymax = `97.5%`), alpha=0.5, colour=NA) +
+  ggplot2::ggplot(spp_params, ggplot2::aes(x = trait_value2,y = mean)) + 
+    ggplot2::geom_ribbon(data = mu_trend, ggplot2::aes(x = trait_value2,ymin = `2.5%`,ymax = `97.5%`), alpha=0.5, colour=NA) +
     ggplot2::facet_grid(param~trait, scales = "free",
                         labeller = labeller(param = ggplot2::as_labeller(param_label_lookup, default = ggplot2::label_parsed), 
                                             trait = ggplot2::as_labeller(trait_label_lookup, default = ggplot2::label_parsed))) +
     ggplot2::geom_pointrange(ggplot2::aes(ymin = `2.5%`, ymax=`97.5%`), size=0.25, shape= 16) +
     ggplot2::geom_point(shape= 21, fill='red') +
-    ggplot2::geom_line(data = mu_trend, ggplot2::aes(x = trait_value, y= mean), col='blue') +
[…]

### 2018-07-20 R/summarise_trait_params.R

@@ -12,7 +12,7 @@ summarise_trait_params <- function(model) {
   fit <- model$fits[[1]] # Indexing to avoid list name
   samples <- rstan::extract(fit, pars=c("a1","a2","a3","b1","b2","b3","c1","c2","c3"))
   
-  res <-lapply(samples, function(x) {
+  lapply(samples, function(x) {
     cbind.data.frame(
       mean = mean(x),
       median = median(x),

### 2018-07-20 ms/manuscript.Rnw
More changes to improve manuscript. Still got to fix up some text

@@
 \includegraphics{figures/fig3.pdf}
-\caption{Estimated relationships between Relationship between species growth-independent mortality rate and wood density.
-Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study.
-Blue trendline with grey shading shows average ($\pm$ 95\% credible intervals) expected relationship.}
+\caption{Estimated relationships between between species traits (maximum dbh, light demand and wood density) and model parameters $\alpha$, $\beta$ and $\gamma$.
+Points are the estimated mean ($\pm$ 95\% credible intervals) parameters for each of the 203 Barro Colorado Island species used in this study.
+Blue trendline with grey shading shows average ($\pm$ 95\% credible intervals) expected relationship across observed trait ranges.}
 \label{fig3}
@@
 \subsection*{Species traits}
+Using the hazard function with highest predictive skill (i.e.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$), we examined how both the individual and combined effects of species' maximum dbh, light demand and wood density affected model accuracy.
+We found that when we examined each trait individually on all three parameters ($\alpha$, $\beta$ and $\gamma$), both wood density and light demand significantly improved predictive performance relative to the model with no traits.
+By contrast, we found only marginal performance gains when a species' maximum dbh was included on all three parameters (Fig.~\ref{fig2} Stage 3).
+
+Additional predictive gains were also obtained when multiple traits were considered.
+Specifically, we found that when the effects of light demand and maximum dbh where included on all parameters, predictive accuracy was significantly higher than any of the 1-trait models (Fig.~\ref{fig2} Stage 4).
+Additional, albeit marginal, gains in predictive performance were also obtained when the effects of all three traits where included on $\alpha$, $\beta$ and $\gamma$ (Fig.~\ref{fig2} Stage 5).
+However, relative to fitting species random effects on each parameter, the performance gain from these traits was small, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig.
+\ref{fig2} Stage 6).
+See Figs.
+S3-S5 for species-level parameter estimates.
+
+Close examination of the how wood three traits on $\alpha$, $\beta$ and $\gamma$ revealed varied and contrasting effects.
+
 Including wood density as an effect on either growth-independent or growth-dependent parameters significantly improved model skill relative to a model without such effects (Fig.
@@
 \includegraphics{figures/fig4.pdf}
-\caption{Predicted changes in mortality rate (A,B), and 1-yr mortality probability (C,D), with increasing growth rate.
-A,C) Estimated curves for all 203 BCI species used in this study.
-B,D) The average ($\pm$ 95\% credible intervals) curve for a high (0.8) and low (0.3) wood density species.}
+\caption{Predicted changes in annual mortality probability with increasing growth rate.
+A) Estimated average curves for all 203 Barro Colorado Island species used in this study.
+B,C,D) represent average ($\pm$ 95\% credible intervals) curve for a high (red) and low (blue) species trait values for wood density (B), maximum dbh (C), and light demand (D).}
 \label{fig4}
@@
 
-% \begin{figure}[t]
-% \includegraphics{figures/fig5.pdf}
-% \caption{Proportion of variation in predicted 1-year mortality for all individuals captured by different effects.
-Effects are not mutually exclusive thus sum to more than 1.0.
-Note "wood density" is a subset of the overall "species" effect.}
-% \label{fig5}
-% \end{figure}
+<<results="asis", echo=FALSE>>=
+Hmisc::latex(table_1(final_model, BCI_training_full),file = "", booktabs = TRUE, rowname = NULL,  
+               colnamesTexCmd="bfseries",
+               col.just = c(rep("c", 4)), label = "table:1",
+               where = "!h",
+               caption ="Relative effects of parameter removal on both r2. r2 estimates are derived 
+               from modelled predictions vs the observed proportion of individuals that died between 1995 and 2010 for each species.")
+@
 
@@
 Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic \cite{Brooks:1998ju}.
-After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.
+After discarding the first 1000 iterations as ‘burn in’, a further 1000 iterations were taken from the joint posterior.
 Species parameter estimates from the final model are shown in Figs.

### 2018-07-20 R/plot_fig2.R

@@ -75,6 +75,6 @@ plot_fig2 <- function(logloss_summaries) {
     plot_theme(strips = TRUE) +
     ggplot2::theme(axis.text.x = ggplot2::element_text(angle=15, hjust = 1),
                    panel.spacing = unit(.5, 'pt'),
-                   strip.text.x = ggplot2::element_text(size=4),
+                   strip.text.x = ggplot2::element_text(size=5),
                    strip.background=element_rect(colour ="lightgrey"))
 }

### 2018-07-20 R/plot_fig4.R

@@ -65,7 +65,7 @@ plot_fig4 <- function(model,
                        xlab=expression("Annual dbh growth"~("cm yr"^-1)), 
                        ylab = ("1-yr mortality probability"), 
                        legend_position = c(0.75,0.8),
-                       legend_label = expression("Shade intolerance"))  +
+                       legend_label = expression("Light demand"))  +
           ggplot2::theme(legend.background = ggplot2::element_rect(fill = "transparent"),
                      legend.key = ggplot2::element_rect(fill = "transparent", 
                                                         color = "transparent"),

### 2018-07-20 R/table_1.R

@@ -117,7 +117,7 @@ table_1 <- function(model, data) {
     dplyr::ungroup() %>%
     tidyr::gather(Model, mean_prob_death, -c(sp,prop_died)) %>%
     dplyr::group_by(Model) %>%
-    dplyr::mutate(r2 = cor(mean_prob_death,prop_died)^2) %>%
+    dplyr::mutate(r2 = round(cor(mean_prob_death,prop_died)^2,3)) %>%
     dplyr::select(-c(sp, prop_died,mean_prob_death)) %>%
     dplyr::distinct()
   
@@ -131,5 +131,7 @@ table_1 <- function(model, data) {
     dplyr::summarise_all(.funs = AUC, response =predictions$dead_next_census) %>%
     tidyr::gather(Model, AUC)
   
-  dplyr::left_join(R2, auc, by="Model")
+  #x <- dplyr::left_join(R2, auc, by="Model")
+  
+ R2
 }

### 2018-07-20 ms/manuscript.Rnw
Minor changes to authors, address,  spacing

@@
 
-\author[a,b,1]{James S Camac}
-\author[c]{Richard Condit}
-\author[d]{Richard G FitzJohn}
-\author[e]{Lachlan McCalman}
-\author[e]{Daniel Steinberg}
+\author[a,1]{James S Camac}
+\author[b]{Richard Condit}
+\author[a,2]{Richard G FitzJohn}
+\author[c]{Lachlan McCalman}
+\author[c]{Daniel Steinberg}
 \author[a]{Mark Westoby}
-\author[f]{S Joseph Wright}
-\author[a,g]{Daniel S Falster}
+\author[d]{S Joseph Wright}
+\author[e]{Daniel S Falster}
 
 \affil[a]{Department of Biological Sciences, Macquarie University, Sydney, New South Wales, Australia}
-\affil[b]{Current Address: Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Melbourne, Victoria, Australia}
-\affil[c]{Field Museum of Natural History and Morton Arboretum, Chicago, Illinois, USA}
-\affil[d]{Imperial College, London, United Kingdom}
-\affil[e]{Data61, CSIRO, Australia}
-\affil[f]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
-\affil[g]{Current Address: Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, Sydney, New South Wales, Australia}
+\affil[b]{Field Museum of Natural History and Morton Arboretum, Chicago, Illinois, USA}
+\affil[c]{Data61, CSIRO, Australia}
+\affil[d]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
+\affil[e]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, Sydney, New South Wales, Australia}
 
@@
 % Please include corresponding author, author contribution and author declaration information
-\authorcontributions{D.F.
-conceived idea; all authors designed research; J.C.
-\& R.F.
+\authorcontributions{D.S.F.
+conceived idea; all authors designed research; J.S.C., D.S.F.
+\& R.G.F.
 performed analysis in discussion with L.M, D.S.
@@
 collected data.
-J.C \& D.F.
+J.S.C \& D.S.F.
 wrote the manuscript}
@@
 
-\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: james.camac@gmail.com}
+\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed.
+E-mail: james.camac@gmail.com.
+Present address: Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Melbourne, Victoria, Australia
+
+\textsuperscript{2} Present address: Department of Infectious Disease Epidemiology, Imperial College London, Faculty of Medicine, Norfolk Place, London W2 1PG, United Kingdom
+}
 
@@
 
-\begin{figure*}[h!]
+\begin{figure*}[ht!]
 \includegraphics{figures/fig3.pdf}
@@
 1a uses images by Tracey Saxby, IAN Image Library.
-J.C., R.F., L.M.
+J.S.C., R.G.F., L.M.
 and D.S.
 were supported by the Science and Industry Endowment Fund (SIEF; RP04-174).
-D.F.
+D.S.F.
 and M.W.

### 2018-07-20 R/plot_fig4.R

@@ -21,7 +21,7 @@ plot_fig4 <- function(model,
                         hazard_curve, 
                         colour_by_trait = FALSE, 
                         xlab= NULL, 
-                        ylab=("1-yr mortality probability"))
+                        ylab=NULL)
   
   p2 <- plot_mu_curves(model, 
                        growth_range, 
@@ -29,7 +29,7 @@ plot_fig4 <- function(model,
                        trait_values = c(0.2,0.8), 
                        hazard_curve, 
                        xlab=NULL, 
-                       ylab = ("1-yr mortality probability"), 
+                       ylab =NULL, 
                        legend_position = c(0.75,0.7),
                        legend_label = expression("Wood density"~(g~cm^-3))) +
           ggplot2::theme(legend.background = ggplot2::element_rect(fill = "transparent"),
@@ -46,7 +46,7 @@ plot_fig4 <- function(model,
                        trait_values = c(3,180),
                        hazard_curve, 
                        xlab=NULL, 
-                       ylab = ("1-yr mortality probability"), 
+                       ylab = NULL, 
                        legend_position = c(0.75,0.7),                       
                        legend_label = expression("Maximum dbh (cm)")) +
           ggplot2::theme(legend.background = ggplot2::element_rect(fill = "transparent"),
@@ -63,7 +63,7 @@ plot_fig4 <- function(model,
                        trait_values = c(0.5, 0.8), 
                        hazard_curve, 
                        xlab=expression("Annual dbh growth"~("cm yr"^-1)), 
-                       ylab = ("1-yr mortality probability"), 
+                       ylab = NULL, 
                        legend_position = c(0.75,0.8),
                        legend_label = expression("Light demand"))  +
           ggplot2::theme(legend.background = ggplot2::element_rect(fill = "transparent"),
@@ -74,5 +74,6 @@ plot_fig4 <- function(model,
                      legend.key.size = unit(0.5, "cm"),
                      legend.title.align =2)
   
+  # ("1-yr mortality probability")
   cowplot::plot_grid(p1,p2,p3,p4, ncol=1, labels=LETTERS[1:4], label_size = 7)
 }

### 2018-07-21 ms/manuscript.Rnw
Various tightenings

@@
 
-\dropcap{R}ates of plant mortality are known to vary widely among individuals within a species, among coexisting species, between forests, and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}.
-This variation has considerable consequences for forest structure and dynamics.
-For example, the death of a single large tree can transfer up to 20,000 kg of carbon from living to decaying carbon pools \cite{Chave:2005hz} and creates a gap in the canopy that can restart a successional race, during which 100's of plants may die while competing for a spot in the sun.
-In models of forest dynamics, variation in tree mortality rates has been shown to have significantly larger impacts on forest structure relative to variation in absolute growth rates \cite{Johnson-2016}.
-Improving our understanding of the mortality process is therefore a priority for making accurate predictions about population, carbon and nutrient dynamics of forests; especially in an era of increased climate-driven forest die-off \cite{Anderegg:2012ep}.
+\dropcap{R}ates of plant mortality vary widely among individuals within a species, among coexisting species, between forests, and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}.
+This variation has large consequences for forest structure and dynamics.
+The death of a large tree can transfer up to 20,000 kg of carbon from living to decaying carbon pools \cite{Chave:2005hz}, while also creating a gap in the canopy that restarts a successional race, during which 100's of plants will die while competing for a spot in the sun.
+Models of forest dynamics suggest variation in tree mortality rates can have larger impacts on forest structure relative than variation in growth rates \cite{Johnson-2016}.
+Improving our understanding of the mortality process is therefore a priority for accurately predicting changes in population, carbon and nutrient dynamics of forests; especially in an era of rapid environmental change \cite{Anderegg:2012ep}.
 
@@
 This requirement arises in part because the observed outcome -- alive vs.
-dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the mortality rate for individual $i$ at time $t$, also known as the ``hazard function or hazard rate''.
-In classic survival analysis \cite{Kleinbaum:2012tn, Zens:2003jv}, the probability an individual dies between times $t_1$ and $t_2$ is a function of both $\lambda_i(t)$, and the amount of time lapsed between $t_1$ and $t_2$:
+dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the mortality (or `hazard`) rate for individual $i$ at time $t$.
+In classic survival analysis \cite{Kleinbaum:2012tn, Zens:2003jv}, the probability an individual dies between times $t_1$ and $t_2$ is a function of both $\lambda_i(t)$, and the amount of time lapsed:
 \begin{equation} \label{eq:mortality_model}
@@
 \end{equation}
-Here $\int_{t_1}^{t_2} \lambda_i(t) {\rm d}t$ can be thought of as the cumulative hazard an individual obtains between $t_1$ and $t_2$\cite{Kleinbaum:2012tn}.
-The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then a realisation of this cumulative hazard on a probability scale.
+Here $\int_{t_1}^{t_2} \lambda_i(t) {\rm d}t$ is the cumulative hazard an individual experiences from  $t_1$ and $t_2$\cite{Kleinbaum:2012tn}.
+The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then a realisation of this cumulative hazard on a probability scale:
 \begin{equation}
@@
 \end{equation}
-
-As trees are long-lived and we are trying to estimate the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes collected over long durations are required.
+Second, as trees are long-lived and we are estimating the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes collected over long durations are required.
 Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.
@@
 \begin{equation} \label{eq:model}
-\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s[i]}}^{\text{Carbon independent}} + \overbrace{\alpha_{s[i]} \, \textrm{e}^{-\beta_{s[i]} \, X_i(t)}}^{\text{Carbon dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
+\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Carbon independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{s,i}(t)}}^{\text{Carbon dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
 \end{equation}
@@
 \ref{fig2}A).
-Individual census effects can be found in Table S1.
+Table S1 gives individual census effects.
 
@@
 \ref{fig2}A).
-A summary of hyper parameter estimates can be found in Table S1.
+Table S1 gives a summary of hyper parameter estimates.
 

### 2018-07-21 ms/manuscript.Rnw
Added more text to the results and fixed some figures

@@
 
-In the remaining stages, we then take the most predictive hazard function and growth rate measure from stage 2 and examine how species traits, wood density, light demand and maximum dbh, affected model predictive performance.
-First, we individually fit an effect of each trait on all parameters (stage 3), we then examine 2-trait combinations on all parameters (stage 4) as well as the effect of three traits on all parameters (stage 5).
-Lastly, we fit a model that had included a species random effect on each parameter (stage 6).
+In the remaining stages, we then take the most predictive hazard function and growth rate measure from stage 2 and examine how species traits: wood density, light demand and maximum dbh, affected model predictive performance.
+Specifically, we first examined each trait individually on all three $\alpha$, $\beta$ and $\gamma$ (stage 3), we also examined 2-trait (stage 4) and 3-trait (stage 5) combinations on these three parameters.
+Lastly, we fit a model that included a species random effect on each parameter (stage 6).
 This allowed us to 1) examine the maximum predictive capacity when accounting for species differences; and 2) provides a relative indication on how much of the total species variation is captured by these three traits.
@@
 \subsection*{Mortality over time}
-Comparing the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates declining over time.
+Examining the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates declining over time.
 The highest proportion of tree death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%).
@@
 \ref{fig1}B, we found that the third function --- the function containing both growth-dependent and growth-independent hazard terms, i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline hazard or growth-dependent only functions (Fig.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline and growth-dependent only hazard functions (Fig.
 \ref{fig2}).
-Moreover, we found that predictive skill was higher when using stem diameter growth over stem-area growth (Fig.
+Moreover, we found that predictive skill was higher when using stem diameter growth relative to stem-area growth (Fig.
 \ref{fig2}A).
@@
 Using the hazard function with highest predictive skill (i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$), we examined how both the individual and combined effects of species' maximum dbh, light demand and wood density affected model accuracy.
-We found that when we examined each trait individually on all three parameters ($\alpha$, $\beta$ and $\gamma$), both wood density and light demand significantly improved predictive performance relative to the model with no traits.
-By contrast, we found only marginal performance gains when a species' maximum dbh was included on all three parameters (Fig.~\ref{fig2} Stage 3).
-
-Additional predictive gains were also obtained when multiple traits were considered.
-Specifically, we found that when the effects of light demand and maximum dbh where included on all parameters, predictive accuracy was significantly higher than any of the 1-trait models (Fig.~\ref{fig2} Stage 4).
-Additional, albeit marginal, gains in predictive performance were also obtained when the effects of all three traits where included on $\alpha$, $\beta$ and $\gamma$ (Fig.~\ref{fig2} Stage 5).
-However, relative to fitting species random effects on each parameter, the performance gain from these traits was small, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$), we examined how model predictive accuracy was influenced by adding the effects of species' maximum dbh, light demand and wood density to $\alpha$, $\beta$ and $\gamma$.
+In general we found that including traits effects on these parameters decreased logarithmic loss, and thus, increased model predictive accuracy (Fig.~\ref{fig2}).
+Moreover, we found that incorporating the effect of all three traits on these parameters resulted in better predictive model relative to models that only incorporated one or two of the traits.
+However, relative to a model including species random effects on each parameter, which account for unexplained species differences, the performance gain from these traits was small, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig.
 \ref{fig2} Stage 6).
@@
 
-Close examination of the how wood three traits on $\alpha$, $\beta$ and $\gamma$ revealed varied and contrasting effects.
-
-Including wood density as an effect on either growth-independent or growth-dependent parameters significantly improved model skill relative to a model without such effects (Fig.
-\ref{fig2} A,B).
-The most parsimonious model, with highest predictive skill, was that attributing the wood density effect to the growth-independent hazard term ($\gamma$; Fig.
-\ref{fig2} B).
-Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate, $\gamma$ (Fig.
-\ref{fig3}).
-This meant that fast growing individuals from a low wood density species had, on average, higher mortality rates (Fig.
-\ref{fig4} A,B), and thus higher probability of death across a 1-yr period (Fig.
-\ref{fig4} C,D), relative to fast growing individuals of high wood density species.
-For example, a species with a wood density of 0.3 g $\textrm{cm}^{-3}$ had an estimated mean probability of dying of \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.3),2)['mean']} yr$^{-1}$ compared to \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.8),2)['mean']} yr$^{-1}$ for a species with a wood density of 0.8 g $\textrm{cm}^{-3}$.
-Incorporating an additional species-level random effect to capture any additional inter specific differences substantially improved model skill relative to a model with only wood density (Fig.
-\ref{fig2}A).
-See Figs.
-S3-S5 for species-level parameter estimates.
-
-To determine whether species-level parameters were associated with their maximum size or light requirement, we conducted a post-hoc analysis comparing the fitted parameters to these traits.
-We found that both $\alpha$, the effect of low growth rate, and $\gamma$, the growth-independent hazard term, were weakly and positively correlated with gap index (a measure of species light requirement).
-By contrast, $\beta$ (the parameter that defines the exponential decay with increasing growth rate) was negatively correlated gap index.
-These correlations suggest that species which predominately recruit in gaps are more prone to death, due to both low growth and stochastic chance.
-They also required faster growth rates to achieve a given mortality rate relative to a species that recruit in shade.
-Correlations between estimated species parameters and their associated maximum DBH were weak).
+Examining the effects of maximum dbh, light demand and wood density on $\alpha$, $\beta$ and $\gamma$ revealed that each trait influenced tree mortality in different ways (Fig.
+\ref{fig3}-\ref{fig4}).
+Specifically, we found that light demanding species were more susceptible to low growth rates (i.e.
+higher $\alpha$), had higher growth-independent hazards (i.e.
+higher $\gamma$) and exhibited shallower declines in growth-dependent hazards with increasing growth rates (i.e.
+lower $\beta$).
+By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards.
+Species' maximum dbh was found to only have a significant effect on $\gamma$, whereby species that can achieve larger dbh exhibited lower growth-independent hazards.
 
@@
 A) Estimated average curves for all 203 Barro Colorado Island species used in this study.
-B,C,D) represent average ($\pm$ 95\% credible intervals) curve for a high (red) and low (blue) species trait values for wood density (B), maximum dbh (C), and light demand (D).}
+B,C,D) represent average ($\pm$ 95\% credible intervals) curve for a high (red) and low (blue) trait values.
+B) wood density values 0.2 vs 0.8 (g cm$^{-3}$); C) maximum dbh values 3 vs 180 cm; D) light demand values 0.5 vs 0.8.}
 \label{fig4}
@@
                colnamesTexCmd="bfseries",
-               col.just = c(rep("c", 4)), label = "table:1",
+               col.just = c(rep("c", 3)), label = "table:1",
                where = "!h",

### 2018-07-21 R/plot_fig3.R

@@ -29,7 +29,7 @@ plot_fig3 <- function(spp_params, mu_trend) {
     ggplot2::geom_point(shape= 21, fill='red') +
     ggplot2::geom_line(data = mu_trend, ggplot2::aes(x = trait_value2, y= mean), col='blue') +
     ggplot2::scale_y_log10(breaks= breaks, labels = labels) +
-    ggplot2::ylab(expression("Instantaneous hazard rate"~("yr"^-1))) +
+    ggplot2::ylab(expression("Hazard rate"~("yr"^-1))) +
     ggplot2::xlab("Species value") +
     plot_theme(strips = TRUE) +
     theme(panel.border = ggplot2::element_rect(color="grey", size=0.5, linetype="solid", fill=NA),

### 2018-07-21 R/plot_fig4.R

@@ -20,42 +20,29 @@ plot_fig4 <- function(model,
                         trait_name = "wood_density", 
                         hazard_curve, 
                         colour_by_trait = FALSE, 
-                        xlab= NULL, 
-                        ylab=NULL)
+                        xlab= "", 
+                        ylab=NULL) +
+    theme(plot.margin=unit(c(0.35,0.2,0.1,0.5),"cm"))
   
   p2 <- plot_mu_curves(model, 
                        growth_range, 
                        trait_name = "wood_density", 
                        trait_values = c(0.2,0.8), 
                        hazard_curve, 
-                       xlab=NULL, 
-                       ylab =NULL, 
-                       legend_position = c(0.75,0.7),
-                       legend_label = expression("Wood density"~(g~cm^-3))) +
-          ggplot2::theme(legend.background = ggplot2::element_rect(fill = "transparent"),
-                     legend.key = ggplot2::element_rect(fill = "transparent", 
-                                                        color = "transparent"),
-                     legend.title = ggplot2::element_text(size=5),
-                     legend.text = ggplot2::element_text(size=5),
-                     legend.key.size = unit(0.5, "cm"),
-                     legend.title.align=0.5)
+                       xlab="", 
+                       ylab =NULL) +
+    plot_theme() + 
+    theme(plot.margin=unit(c(0.35,0.2,0.1,0.5),"cm"))
 
   p3 <- plot_mu_curves(model,
                        growth_range, 
                        trait_name = "dbh_95", 
                        trait_values = c(3,180),
                        hazard_curve, 
-                       xlab=NULL, 
-                       ylab = NULL, 
-                       legend_position = c(0.75,0.7),                       
-                       legend_label = expression("Maximum dbh (cm)")) +
-          ggplot2::theme(legend.background = ggplot2::element_rect(fill = "transparent"),
-                     legend.key = ggplot2::element_rect(fill = "transparent", 
-                                                        color = "transparent"),
-                     legend.title = ggplot2::element_text(size=5),
-                     legend.text = ggplot2::element_text(size=5),
-                     legend.key.size = unit(0.5, "cm"),
-                     legend.title.align=0.5)
+                       xlab= "", 
+                       ylab = NULL) +
+        plot_theme() +
+        theme(plot.margin=unit(c(0.35,0.2,0.1,0.5),"cm"))
   
   p4 <- plot_mu_curves(model, 
[…]

### 2018-07-21 R/table_1.R

@@ -99,14 +99,14 @@ table_1 <- function(model, data) {
       prop_died,
       dead_next_census,
       `Full model` = full_model, 
-      `Full - Census error` = full_minus_census, 
-      `Full - Wood density` = full_minus_rho, 
-      `Full - Gap index` = full_minus_gap,
-      `Full - Max dbh` = full_minus_dbh,
-      `Full - Species error` = full_minus_sppre,
-      `Full - Species traits & species error` = full_minus_spp,
-      `Full - Growth dependent hazard` = full_minus_growthdep, 
-      `Full - Growth independent hazard` = full_minus_growthindep)
+      `Full minus census error` = full_minus_census, 
+      `Full minus wood density` = full_minus_rho, 
+      `Full minus light demand` = full_minus_gap,
+      `Full minus max dbh` = full_minus_dbh,
+      `Full minus species error` = full_minus_sppre,
+      `Full minus species traits & species error` = full_minus_spp,
+      `Full minus growth dependent hazard` = full_minus_growthdep, 
+      `Full minus growth independent hazard` = full_minus_growthindep)
   
   
   # Calculate R2 relative to observed proportion of death per species
@@ -117,7 +117,7 @@ table_1 <- function(model, data) {
     dplyr::ungroup() %>%
     tidyr::gather(Model, mean_prob_death, -c(sp,prop_died)) %>%
     dplyr::group_by(Model) %>%
-    dplyr::mutate(r2 = round(cor(mean_prob_death,prop_died)^2,3)) %>%
+    dplyr::mutate(`$r^2$` = round(cor(mean_prob_death,prop_died)^2,3)) %>%
     dplyr::select(-c(sp, prop_died,mean_prob_death)) %>%
     dplyr::distinct()
   
@@ -129,9 +129,9 @@ table_1 <- function(model, data) {
   auc <- predictions %>%
     dplyr::select(-c(dead_next_census,sp, prop_died)) %>%
     dplyr::summarise_all(.funs = AUC, response =predictions$dead_next_census) %>%
-    tidyr::gather(Model, AUC)
+    tidyr::gather(Model, AUC) %>%
+    dplyr::mutate(AUC = round(AUC,3))
   
-  #x <- dplyr::left_join(R2, auc, by="Model")
+  dplyr::left_join(R2, auc, by="Model")
   
- R2
 }

### 2018-07-23 ms/manuscript.Rnw
Final touches to the main text. Need to make minor edits to the methods still

@@
 \significancestatement{Tree mortality is a fundamental demographic process affecting forest dynamics and carbon cycling.
-Here, for the first time, we build a statistical mortality model that not only integrates intra-specific and inter-specific variation but also, for the first time, simulataneously estimates the independent contributions of two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth.
-We parameterise this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals encompassed within 203 species.
-We show that both growth-independent and growth-dependent hazards are significant contribtors to tree mortality at Barro Colorado Island.
-Moreover, we show that species traits -- wood density, maximum dbh and shade intolerance -- all variously affect rates of tree mortality but ultimately only account for a small fraction of the total species variability.}
+Here, we build a mortality model that not only integrates intra and inter-specific variation but also, for the first time, simulataneously estimates the independent contributions of two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth.
+We parameterise this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals across 203 species.
+We show that both growth-dependent and -independent hazards are significant contribtors to tree mortality at Barro Colorado Island.
+Moreover, we show that species traits -- wood density, maximum dbh and light demand -- all variously affect tree mortality but ultimately only account for a small fraction of the total variability among species.}
 
@@
 Models of forest dynamics suggest variation in tree mortality rates can have larger impacts on forest structure relative than variation in growth rates \cite{Johnson-2016}.
-Improving our understanding of the mortality process is therefore a priority for accurately predicting changes in population, carbon and nutrient dynamics of forests; especially in an era of rapid environmental change \cite{Anderegg:2012ep}.
+Improving our understanding of the mortality process is therefore a priority for accurately predicting changes in population, carbon and nutrient dynamics of forests; especially in an era of rapid environmental change \cite{Anderegg:2012ep,Mcdowell:2013wf}.
 
@@
 
+\begin{figure*}[!ht]
+\includegraphics{figures/fig2.pdf}
+\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% confidence intervals) logarithmic loss, with lower values implying greater predictive skill.
+A) Six sequential stages of model selection with each stage increasing in complexity.
+\textbf{Stage 1:} Baseline: constant ($\gamma$) vs Baseline with census effect ($\gamma\delta_t$); \textbf{Stage 2:} Best model of stage 1 compared to hazard functions that also account for individual growth rate.
+This includes two possible hazard functions: a growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters.
+Symbols represent the three hazard functional form: baseline (circle), growth-dependent (triangle), and hazard containing both growth-independent and growth-dependent hazards (square).}
+\label{fig2}
+\end{figure*}
+
 To validate this model, we also fit two alternative hazard functions: 1) a baseline hazard (i.e.
@@
 Consequently, when we allowed hazard rates to be scaled by census (i.e.
-adding the term $\delta$) we observed a small increase in predictive skill (Fig.
+adding the term $\delta_t$) we observed a small increase in predictive skill (Fig.
 \ref{fig2}A).
 Table S1 gives individual census effects.
-
-\begin{figure*}[!ht]
-\includegraphics{figures/fig2.pdf}
-\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% confidence intervals) logarithmic loss, with lower values implying greater predictive skill.
-A) Six sequential stages of model selection with each stage increasing in complexity.
-\textbf{Stage 1:} Baseline: constant ($\gamma$) vs Baseline with census effect ($\gamma\delta_t$); \textbf{Stage 2:} Best model of stage 1 compared to hazard functions that also account for individual growth rate.
-This includes two possible hazard functions: a growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters.
-Symbols represent the three hazard functional form: baseline (circle), growth-dependent (triangle), and hazard containing both growth-independent and growth-dependent hazards (square).}
-\label{fig2}
-\end{figure*}
 
@@
 \subsection*{Species traits}
-Using the hazard function with highest predictive skill (i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$), we examined how model predictive accuracy was influenced by adding the effects of species' maximum dbh, light demand and wood density to $\alpha$, $\beta$ and $\gamma$.
-In general we found that including traits effects on these parameters decreased logarithmic loss, and thus, increased model predictive accuracy (Fig.~\ref{fig2}).
-Moreover, we found that incorporating the effect of all three traits on these parameters resulted in better predictive model relative to models that only incorporated one or two of the traits.
-However, relative to a model including species random effects on each parameter, which account for unexplained species differences, the performance gain from these traits was small, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig.
+Using the hazard function form with highest predictive skill (i.e.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$; Fig.~\ref{fig1}b right), we examined how effects of a species' maximum dbh ($\psi$), light demand ($\upsilon$) and wood density ($\rho$) placed on $\alpha$, $\beta$ and $\gamma$ affected model predictive accuracy.
+In general we found that including traits effects on these parameters decreased logarithmic loss, and thus, increased predictive accuracy (Fig.~\ref{fig2}).
+Moreover, we found that incorporating the effect of all three traits on these parameters resulted in higher predictive capacity relative to models that only incorporated one or two of the traits.
+However, this increase in predictive accuracy was small relative to including species random effects on each parameter, which account for unexplained species differences, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig.
 \ref{fig2} Stage 6).
@@
 
-Examining the effects of maximum dbh, light demand and wood density on $\alpha$, $\beta$ and $\gamma$ revealed that each trait influenced tree mortality in different ways (Fig.
+\begin{figure}[h!]
+\includegraphics{figures/fig4.pdf}
+\caption{Predicted changes in annual mortality probability with increasing growth rate.
+A) Estimated average curves for all 203 Barro Colorado Island species used in this study.
+B,C,D) represent average ($\pm$ 95\% credible intervals) curve for a high (red) and low (blue) trait values.
+B) wood density values 0.2 vs 0.8 (g cm$^{-3}$); C) maximum dbh values 3 vs 180 cm; D) light demand values 0.5 vs 0.8.
+Note that the intercept is the sum of $\alpha$ and $\gamma$ (i.e.
+the stochastic chance of death plus effect of low growth.}
+\label{fig4}
+\end{figure}
+
+In order to examine how species traits -- maximum dbh, light demand and wood density -- affect tree mortality rates, we fitted a final model containing all three species traits and a species random effect (which captures inter-specific variation that unexplained by traits) on each of $\alpha$, $\beta$ and $\gamma$.
+We found that each trait influenced tree mortality in different ways (Fig.
 \ref{fig3}-\ref{fig4}).
@@
 By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards.
-Species' maximum dbh was found to only have a significant effect on $\gamma$, whereby species that can achieve larger dbh exhibited lower growth-independent hazards.
+Species' maximum dbh only had significant effects on $\gamma$, whereby species that can achieve larger dbh exhibited lower growth-independent hazards.
 
-\begin{figure}[h!]
-\includegraphics{figures/fig4.pdf}
-\caption{Predicted changes in annual mortality probability with increasing growth rate.
-A) Estimated average curves for all 203 Barro Colorado Island species used in this study.
-B,C,D) represent average ($\pm$ 95\% credible intervals) curve for a high (red) and low (blue) trait values.
-B) wood density values 0.2 vs 0.8 (g cm$^{-3}$); C) maximum dbh values 3 vs 180 cm; D) light demand values 0.5 vs 0.8.}
-\label{fig4}
-\end{figure}
+\subsection*{Parameter relative importance}
 
-\subsection*{Parameter contribution to predicted variation}
-To compliment the analysis, we used the final fitted model to estimate the amount of variation captured by different factors in the predicted probability of dying across 1 year for all individuals (Table XX).
-We achieved this by removing each parameter and calculating the reduction in the sum of squares for this model relative to the full model.
-Species effects contributed...
+Using the final model, we then examine: 1) its ability distinguish between individuals that died andindividuals that survived (i.e.
+via AUROC); and 2) the amount of variation (i.e.
+$r^2$) the model captures relative to the proportions of individual deaths observed per species between 1995 and 2010 (Table.~\ref{table:1}).
+We found that the full model had an AUCROC of \Sexpr{table1[[1,3]]}, indicating that the model had a fair ability to correctly discriminate observed deaths from survivals.
+Moreover, we found that the full model also explained \Sexpr{table1[[1,2]]*100}\% of the total variation observed among species.
 
 <<results="asis", echo=FALSE>>=
-Hmisc::latex(table_1(final_model, BCI_training_full),file = "", booktabs = TRUE, rowname = NULL,  
+Hmisc::latex(table1,file = "", booktabs = TRUE, rowname = NULL,  
                colnamesTexCmd="bfseries",
@@
                where = "!h",
-               caption ="Relative effects of parameter removal on both r2. r2 estimates are derived 
-               from modelled predictions vs the observed proportion of individuals that died between 1995 and 2010 for each species.")
+               caption ="Relative effects of parameter removal on both $r^2$ and AUROC. $r^2$ estimates are derived 
+               from modelled predictions vs proportion of individual deaths observed between 1995 and 2010 for each species.
+As such, $r^2$ measures the amount of species-level variation explained by the model or sub-models.
+By contrast, AUROC examines the model's ability to distinguish individuals that died from individuals that survived.")
 @
 
+We then examined the relative contribution species traits, random effects and growth-dependent and independent hazards (Table.~\ref{table:1}) had on overall hazard rates by individually removing their effects and examining changes in both $r^2$ and AUROC.
+First, we found that the removal of census random effect $\delta_t$ had little impact on either $r^2$ or AUROC.
+The removal of either wood density or light demand resulted in similar magnitude effects on both $r^2$ and AUROC.
+By contrast, the removal of maximum dbh had a slightly smaller effect on these metrics.
+We detected a significant drop in both $r^2$ and AUROC when species random effects (i.e.
+the parameters that capture variation that is otherwise unexplained by traits) were removed, further highlighting that maximum dbh, light demand and wood density explain only a small fraction of the overall inter-specific variation observed among species.
+As expected, the removal of both species traits and species random effects resulted in a model that was unable to capture any inter-specific variation observed among species, and thus had low discriminatory ability.
+Lastly, we found the removal of the growth-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the growth-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation.
+However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the growth-dependent hazard resulted in substantially lower predictivie ability.
+
 \section*{Discussion}
-Our Bayesian framework coupled with cross validation revealed that the most explanatory and parsimonious model of tropical tree death was that which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) attributed the effect of wood density to growth-independent mortality; and 4) incorporated temporal variability.
-Moreover, we found that rates of tropical tree mortality varied substantially between species and that wood density, a species level functional trait, explained only a limited proportion of the overall inter-specific variation.
+Our Bayesian framework coupled with cross validation revealed that the most predictive hazard model of tropical tree death was one which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) included the effects of species' maximum dbh, wood density and light demand; and 4) incorporated temporal variability.
+Moreover, we found that rates of tropical tree mortality varied substantially between species and that the a species' maximum dbh, wood density and light demand explained only a fraction of the overall inter-specific variation.
 
-The findings of this study provide empirical support for dynamic vegetation models that estimate mortality as the sum of growth-dependent and growth-independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}.
+The findings of this study provide empirical support for dynamic vegetation models that predict tree mortality as a function of both growth-dependent and independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}.
 We show that regardless of growth measure, incorporating both hazards significantly improves model predictive skill.
@@
 
-Additionally, the partitioning of mortality into growth-dependent and growth-independent effects allowed us to estimate the proportion of variation attributed to each.
-Like many other studies \cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlight the importance of light competition in influencing tropical tree demographic rates.
-Specifically, we found that the growth-dependent hazard accounted for XX of the total predicted variability in mortality rates (Table X).
-This suggests deficiencies in carbon budget are a major contributor to tree death on BCI.
+Additionally, the partitioning of mortality rates into both carbon-dependent and carbon-independent effects allowed us to not only estimate their relative contribution, but also how species traits affect each.
+Like many other studies \cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlighted the importance of light competition in influencing tropical tree demographic rates.
+Specifically, we found that that the removal of the growth-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table1}).
+This suggests deficiencies in carbon budgets are a major contributor to tree death on Barro Colorado Island.
 
-Incorporating the effect of wood density on mortality rates also improved predictive performance.
-Our analyses revealed that the most parsimonious combination of wood density effects was when it was attributed to only the growth-independent hazard term.
-Specifically, high wood density species had lower baseline rates relative to low wood density counterparts.
-This finding corroborates the observed negative correlation observed between mortality and wood density reported elsewhere \cite{Kraft:2010kq,Chave:2009iy}.
-More importantly, our analyses support the theory that wood density reduces mortality rates by decreasing a species' vulnerability to growth-independent threats, such as windthrow, trampling and treefall \cite{vanGelder:2006exa,Chave:2009iy}.
+Our analyses support previous research that advocated that wood density, species maximum size and light demand important explanatory variables of inter-specific variation in tropical rainforests \cite{Poorter:2006tq, Chave:2009iy}.
+Specificially, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates.
+By contrast, both high dbh or wood density species were found to be more tolerant of growth-independent hazards, corrobrating evidence reported elsewhere\cite{Chave:2009iy, Kraft:2010kq}.
+However, our analyses also revealed that high wood density species were also more tolerant of low growth rates, and thus, better able to tolerate shade and competiton \cite{vanGelder:2006exa, Kunstler:2016km}.
 
-While wood density effects are now being incorporated in mortality algorithms of many vegetation models \cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation (Fig.
-\ref{fig4}, Table X).
-This means that such models are likely to severely underestimate the true variability in mortality rates.
-Consequently, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.
+While the effects of species' wood density, maximum size and light demand are now varying being incorporated in mortality algorithms of many vegetation models \cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation (Fig.
+\ref{fig4}, Table~\ref{table:1}).
+This means that such models are likely to severely underestimate the true variability in tropical rainforest mortality rates.
+As a consequence, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.
 
-We found that wood density only explained XX of the total XX variation explained by species effects, suggesting that other traits are also affecting tree mortality rates.
-Post-hoc analyses suggest this unexplained variation was in part related to a species' light demand, but not its maximum height.
-Both traits have been proposed major axes of inter-specific variation in tropical rainforests \cite{Poorter:2006tq} (Fig.
-XX).
-Specifically, light demanding species (i.e.
-high gap index) had higher growth-independent hazard rates and were more susceptible to dying as a result of low growth, relative to those that readily recruit in shade (i.e.
-low gap index), supporting past findings \cite{Ruger:2011cv}.
-By contrast, we detected no correlations between a species maximum stem diameter and mortality rates, both contradicting \cite{Poorter:2006tq} and supporting \cite{Wright:2010fl} previous results.
-Future research should therefore resolve how other traits influence growth-dependent and growth-independent hazards, and identify the combination of traits needed to improve model predictive skill.
+% Daniel I find these next three paragraphs as ho humm. I'm advocating for their deletion. Instead, we could just end with the above? Thoughts?
 
-Future research should also resolve how climate influences both growth-dependent and growth-independent hazards, particularly as climate-driven mortality is increasing \cite{vanMantgem:2009bi, Mcdowell:2013wf}.
-By contrast, our analyses revealed a marginal decline in mortality at BCI from 1995 to 2010.
+%Future research should focus on resolving how climate influences both carbon-dependent and carbon-independent hazards, particularly as climate-driven mortality is increasing \cite{vanMantgem:2009bi,Anderegg:2012ep,Mcdowell:2013wf}.
+Our analyses revealed a marginal decline in mortality at BCI from 1995 to 2010.
 Attributing temporal changes to particular climatic variables is challenging, however, due to low number temporal replication (n=3).
 
-We should also consider how the interval between censuses affects the estimation of growth-dependent and growth-independent hazards.
-Large census intervals may underestimate growth-dependent mortality, as wide census intervals will not capture deaths due to rapid declines in growth, or events such as drought (although drought might also increase tree growth \cite{Condit2017}).
-Consequently, we may overestimate the relative contribution of growth-independent hazards.
+%We should also consider how the interval between censuses affects the estimation of carbon-dependent and carbon-independent hazards.
+Large census intervals may underestimate carbon-dependent mortality, as wide census intervals will not capture deaths due to rapid declines in growth, or events such as drought (although drought might also increase tree growth \cite{Condit2017}).
+Consequently, we may overestimate the relative contribution of carbon-independent hazards.
 
-Here we showcase a new framework for modelling tropical tree mortality that unifies empirical evidence from within and between species studies.
-This framework also provides an approach for partitioning mortality rates into growth-dependent and growth-independent hazards.
+%Here we showcase a new framework for modelling tropical tree mortality that unifies empirical evidence from within and between species studies.
+This framework also provides an approach for partitioning mortality rates into carbon-dependent and carbon-independent hazards.
 Our findings reveal that while wood density is an important trait affecting mortality rates, we are still only capturing a fraction of the overall species variability in mortality rates.
@@
 Cores were broken into pieces, each 5 cm long and specific gravity of each piece was determined by oven drying (100\textdegree C) and dividing by the fresh volume (as measured by water displacement).
+
+We calculated a species' light demand using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995.
+The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m).
+For each subplot, we calculated the number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap).
+As light may penetrate into a subplot from the edge of a subplot, we rescaled this index to account for values in the eight immediate neighbouring subplots.
+Specifically, we used a weighted sum approach whereby the central subplot is assigned a weight of 8 and the eight neighbouring subplots are assigned a weight of 1.
+This meant that the contribution of the central plot was equivalent to the combined effect of all eight neighbouring plots.
+These weighted values were then summed and rescaled between 0 and 1 by dividing by the maximum value estimated across all subplots.
+The light  demand for each species was estimated as the mean light index encountered by new saplings appearing in the census (Fig.
+S2).
+
+A species' maximum dbh was determined from the data, by taking the 95th percentile.
 
@@
 
-\subsection*{Posthoc correlations}
-We calculated a gap index as a measure of a species' light dependence using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995.
-The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m).
-For each subplot, we calculated the number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap).
-As light may penetrate into a subplot from the edge of a subplot, we rescaled this index to account for values in the eight immediate neighbouring subplots.
-Specifically, we used a weighted sum approach whereby the central subplot is assigned a weight of 8 and the eight neighbouring subplots are assigned a weight of 1.
-This meant that the contribution of the central plot was equivalent to the combined effect of all eight neighbouring plots.
-These weighted values were then summed and rescaled between 0 and 1 by dividing by the maximum value estimated across all subplots.
-The gap index for each species was estimated as the mean light index encountered by new saplings appearing in the census (Fig.
-S2).
-
-}
-
 \showmatmethods % Display the Materials and Methods section

### 2018-07-23 R/table_1.R

@@ -129,8 +129,8 @@ table_1 <- function(model, data) {
   auc <- predictions %>%
     dplyr::select(-c(dead_next_census,sp, prop_died)) %>%
     dplyr::summarise_all(.funs = AUC, response =predictions$dead_next_census) %>%
-    tidyr::gather(Model, AUC) %>%
-    dplyr::mutate(AUC = round(AUC,3))
+    tidyr::gather(Model, AUROC) %>%
+    dplyr::mutate(AUROC = round(AUROC,3))
   
   dplyr::left_join(R2, auc, by="Model")
   

### 2018-07-23 ms/manuscript.Rnw
fixed typo

@@
 Like many other studies \cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlighted the importance of light competition in influencing tropical tree demographic rates.
-Specifically, we found that that the removal of the growth-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table1}).
+Specifically, we found that that the removal of the growth-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table:1}).
 This suggests deficiencies in carbon budgets are a major contributor to tree death on Barro Colorado Island.
@@
 Specificially, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates.
-By contrast, both high dbh or wood density species were found to be more tolerant of growth-independent hazards, corrobrating evidence reported elsewhere\cite{Chave:2009iy, Kraft:2010kq}.
+By contrast, both high dbh or wood density species were found to be more tolerant of growth-independent hazards, corrobrating evidence reported elsewhere \cite{Chave:2009iy, Kraft:2010kq}.
 However, our analyses also revealed that high wood density species were also more tolerant of low growth rates, and thus, better able to tolerate shade and competiton \cite{vanGelder:2006exa, Kunstler:2016km}.
@@
 
-A species' maximum dbh was determined from the data, by taking the 95th percentile.
+A species' maximum dbh was determined from the data, by taking the 95th percentile observed for each species.
 
@@
 where $N_k$ refers to the number of observations in fold.
+}
 

### 2018-07-23 ms/manuscript.Rnw
some final changes to the document

@@
 % 120 words
-\significancestatement{Tree mortality is a fundamental demographic process affecting forest dynamics and carbon cycling.
-Here, we build a mortality model that not only integrates intra and inter-specific variation but also, for the first time, simulataneously estimates the independent contributions of two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth.
+\significancestatement{We build a mortality model that not only integrates intra and inter-specific variation but also, simulataneously estimates the two fundamental hazards faced by plants -- Death due to low carbon budget and death independent of carbon budget -- and how a species' light demand, maximum dbh and wood density affect each.
 We parameterise this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals across 203 species.
-We show that both growth-dependent and -independent hazards are significant contribtors to tree mortality at Barro Colorado Island.
-Moreover, we show that species traits -- wood density, maximum dbh and light demand -- all variously affect tree mortality but ultimately only account for a small fraction of the total variability among species.}
+We show that carbon-dependent hazards are a significant contribtor to tree mortality at Barro Colorado Island.
+Moreover, we show that species traits -- wood density, maximum dbh and light demand -- all affect tree mortality rates but ultimately only account for a small fraction of the total variability among species.}
 
@@
 In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive hazard functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
-Specifically, we develop an empricial mortality model that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the independent contributions of growth-related and growth-independent hazards on the overall mortality rates.
-We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and tolerance to shade outperform alternative formulations.
-Moreover, we show that while these three traits do affect both growth-dependent and growth-independent hazards, they account for only a small fraction of the total variability observed among species.
+Specifically, we develop an empricial mortality model that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the independent contributions of carbon-related hazards and carbon-independent hazards on the overall mortality rates.
+We show that functions that separate mortality rates into carbon-dependent and carbon-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and light demand outperform alternative formulations.
+Moreover, we show that while these three traits do affect both carbon-dependent and carbon-independent hazards, they account for only a small fraction of the total variability observed among species.
 Combined, this study provides a statistical framework for predicting individual-level mortality in highly diverse tropical forests.
@@
 
-There are several distinct advantages of modelling tree mortality as a function of two additive, and thus independent hazards: 1) The model explicitly accounts for the independent effects of the two fundamental types of hazards faced by plants; 2) it allows for the examination of the relative contributions each hazard provides to the overall hazard rate; 3) it allows for examination of how various factors affect each hazard term, and because of this; 4) allows ecological theories of how factors such as species traits affect plant mortality to be empirically tested.
+There are several distinct advantages of modelling tree mortality as a function of two additive, and thus independent hazards: 1) The model explicitly accounts for the independent effects of the two fundamental types of hazards faced by plants; 2) it allows for the examination of the relative contributions each hazard provides to the overall hazard rate; 3) it allows for examination of how various factors affect each hazard term, and because of this; 4) allows ecological theories of how factors such as species traits affect plant mortality to be empirically tested using commonly collected data.
 
@@
 A) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
-B) We consider three alternative hazard functions: 1) a growth-independent baseline hazard (i.e.
-average hazard), 2) a growth-dependent hazard; and 3) a function that combines both growth dependent and independent hazards.
-The parameters of the models are biologically interpretable: $\alpha$ defines the the low growth rate effect on the overall hazard; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote (Panel B; right), or baseline hazard (Panel B; left).
-In the most complex model (Panel B; right), $\alpha$ and $\beta$ capture growth-dependent hazards, while $\gamma$ captures growth-independent hazards (e.g.
-windfall, fire) that kill a plant, irrespective of its growth rate.
+B) We consider three alternative hazard functions: 1) a carbon-independent baseline hazard (i.e.
+average hazard), 2) a carbon-dependent hazard, where growth rate is a proxy for carbon budget; and 3) a function that combines both carbon-dependent and independent hazards.
+The parameters of the models are biologically interpretable: $\alpha$ defines the low growth rate (i.e.
+low carbon budget) effect on the overall hazard; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote (Panel B; right).
+In the most complex model (Panel B; right), $\alpha$ and $\beta$ capture carbon-dependent hazards, while $\gamma$ captures carbon-independent hazards that kill a plant, irrespective of its growth rate/carbon budget (e.g.
+windfall, fire).
 For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
-C) Each model's skill in predicting observed outcomes ($S$) is quantified via the log-loss function (eq. \ref{eq:logloss}).
+C) Each model's skill in predicting observed outcomes ($S$) is quantified via the log-loss function (eq. \ref{eq:logloss}). Lower logloss means higher predictive accuracy.
 D) The predictive skill of alternative models was evaluated via 10-fold cross validation.
@@
 Eq.
-\ref{eq:model} allows for the additive, thus independent, effects of carbon-dependent and carbon-independent hazards, includes a negative exponential effect of individual's growth rate, and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
+\ref{eq:model} allows for the additive, thus independent, effects of carbon-dependent and carbon-independent hazards, includes a negative exponential effect of individual's growth rate (our measure of carbon budget), and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
 Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$.
-Here we include the effects of three species traits thought to be major drivers of inter-specific variation in tropical tree mortality -- wood density ($\rho$), light requirement ($\upsilon$) and maximum dbh ($\psi$).
+Here we include the effects of three species traits thought to be major drivers of inter-specific variation in tropical tree mortality -- wood density ($\rho$), light rdemand ($\upsilon$) and maximum dbh ($\psi$).
 We also include species random effects in order to capture species-level differences that are not otherwise accounted for by species traits (for details see Methods).
@@
 \includegraphics{figures/fig2.pdf}
-\caption{Predictive skill of alternative hazard functions. Values are mean ($\pm$ 95\% confidence intervals) logarithmic loss, with lower values implying greater predictive skill.
+\caption{Predictive skill of alternative hazard functions.
+Values are mean ($\pm$ 95\% confidence intervals) logarithmic loss, with lower values implying greater predictive skill and a logloss $\approx$ 0.693 indicating a model unable to distinguish an observed death from a survival.
 A) Six sequential stages of model selection with each stage increasing in complexity.
-\textbf{Stage 1:} Baseline: constant ($\gamma$) vs Baseline with census effect ($\gamma\delta_t$); \textbf{Stage 2:} Best model of stage 1 compared to hazard functions that also account for individual growth rate.
-This includes two possible hazard functions: a growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters.
-Symbols represent the three hazard functional form: baseline (circle), growth-dependent (triangle), and hazard containing both growth-independent and growth-dependent hazards (square).}
+\textbf{Stage 1:} Baseline: constant ($\gamma$) vs Baseline with census effect ($\gamma\delta_t$); \textbf{Stage 2:} Best model of stage 1 compared to hazard functions that also account for individual growth rate (i.e.
+carbon budget).
+This includes two possible hazard functions: a carbon-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + carbon-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters.
+Symbols represent the three hazard functional form: baseline (circle), carbon-dependent (triangle), and hazard containing both carbon-independent and carbon-dependent hazards (square).}
 \label{fig2}
@@
 To validate this model, we also fit two alternative hazard functions: 1) a baseline hazard (i.e.
-a null model) and 2) a growth-dependent only hazard.
+a null model) and 2) a carbon-dependent only hazard.
 We then compare each functions ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from \Sexpr{n_spp(BCI_training_full)} species recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama (Fig.
@@
 We achieved this by fitting a constant, $\gamma$, that is invariant with respect to species, growth rate, or census period (i.e.
-the null model) and compared it to a constant that is scaled by a census random effect, $\gamma\delta_{t}$.
+a null model) and compared it to a constant that is scaled by a census random effect, $\gamma\delta_{t}$.
 This stage revealed that there was significant variation in mortality rates observed among censuses, and as such a random effect for census was included as a scaling factor for all subsequent models.
@@
 The second form (Fig.
-\ref{fig1}B, middle) assumes the risk of dying declines towards an asymptote of zero as growth increases.
+\ref{fig1}B, middle) assumes the risk of dying declines towards an asymptote of zero as growth, and thus carbon budget, increases.
 The third form (Fig.
-\ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike the second functional form examined here, asymptotes at some baseline hazard $\textgreater 0$.
+\ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike the second functional form examined here, asymptotes at some carbon-independent baseline hazard $\textgreater 0$.
 Combined, these three models capture a variety of functional responses that are both used in empirical studies of plant mortality \cite{Kobe:1995tw,Wyckoff:2000un} and are represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, but which have not previously been systematically compared.
-In doing this, we also also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.
+In doing this, we also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts carbon-dependent mortality.
 
-In the remaining stages, we then take the most predictive hazard function and growth rate measure from stage 2 and examine how species traits: wood density, light demand and maximum dbh, affected model predictive performance.
-Specifically, we first examined each trait individually on all three $\alpha$, $\beta$ and $\gamma$ (stage 3), we also examined 2-trait (stage 4) and 3-trait (stage 5) combinations on these three parameters.
+In the remaining stages, we take the most predictive hazard function and growth rate measure from stage 2 and examine how species traits: wood density, light demand and maximum dbh, affected model predictive performance.
+Specifically, we examined each trait individually on all three $\alpha$, $\beta$ and $\gamma$ (stage 3), as well as 2-trait (stage 4) and 3-trait (stage 5) combinations on these three parameters.
 Lastly, we fit a model that included a species random effect on each parameter (stage 6).
 This allowed us to 1) examine the maximum predictive capacity when accounting for species differences; and 2) provides a relative indication on how much of the total species variation is captured by these three traits.
-Because these models were computationally expensive to run (some requiring over 400 computing hours), we were unable to examine interactions among traits, or determine the optimal combination of traits with parameters.
+Because these models were computationally expensive to run (with some models requiring over 400 computing hours), we were unable to examine interactions among traits or determine the most predictive combination of traits effects and parameters in a cross validated framework.
 
@@
 Comparing the three hazard functions summarised in Fig.
-\ref{fig1}B, we found that the third function --- the function containing both growth-dependent and growth-independent hazard terms, i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline and growth-dependent only hazard functions (Fig.
+\ref{fig1}B, we found that the third function --- the function containing both carbon-dependent and carbon-independent hazard terms, i.e.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline and carbon-dependent only hazard functions (Fig.
 \ref{fig2}).
@@
 \ref{fig2}A).
-Table S1 gives a summary of hyper parameter estimates.
+Table S1 gives a summary of mean hyper parameter estimates.
 
@@
 \includegraphics{figures/fig3.pdf}
-\caption{Estimated relationships between between species traits (maximum dbh, light demand and wood density) and model parameters $\alpha$, $\beta$ and $\gamma$.
+\caption{Estimated relationships between between species traits (maximum dbh, light demand and wood density) and model parameters $\alpha$ (low growth effect), $\beta$ (growth-mortality decay rate) and $\gamma$ (baseline/carbon-independent hazard).
 Points are the estimated mean ($\pm$ 95\% credible intervals) parameters for each of the 203 Barro Colorado Island species used in this study.
@@
 $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$; Fig.~\ref{fig1}b right), we examined how effects of a species' maximum dbh ($\psi$), light demand ($\upsilon$) and wood density ($\rho$) placed on $\alpha$, $\beta$ and $\gamma$ affected model predictive accuracy.
-In general we found that including traits effects on these parameters decreased logarithmic loss, and thus, increased predictive accuracy (Fig.~\ref{fig2}).
+In general we found that including trait effects on these parameters decreased logarithmic loss, and thus, increased predictive accuracy (Fig.~\ref{fig2}).
 Moreover, we found that incorporating the effect of all three traits on these parameters resulted in higher predictive capacity relative to models that only incorporated one or two of the traits.
@@
 lower $\beta$).
-By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards.
-Species' maximum dbh only had significant effects on $\gamma$, whereby species that can achieve larger dbh exhibited lower growth-independent hazards.
-
+By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower carbon-independent hazards.
+Species' maximum dbh only had significant effects on $\gamma$, whereby species that can achieve larger dbh exhibited lower carbon-independent hazards.
+Table S1 for trait effects on each parameter.
 \subsection*{Parameter relative importance}
 
-Using the final model, we then examine: 1) its ability distinguish between individuals that died andindividuals that survived (i.e.
-via AUROC); and 2) the amount of variation (i.e.
-$r^2$) the model captures relative to the proportions of individual deaths observed per species between 1995 and 2010 (Table.~\ref{table:1}).
+Using the final model, we then examine: 1) its ability distinguish between individuals that died and individuals that survived via AUROC; and 2) the amount of inter-specific variation the model explains (i.e.
+$r^2$) relative to the proportions of individual deaths observed per species between 1995 and 2010 (Table.~\ref{table:1}).
 We found that the full model had an AUCROC of \Sexpr{table1[[1,3]]}, indicating that the model had a fair ability to correctly discriminate observed deaths from survivals.
@@
 
-We then examined the relative contribution species traits, random effects and growth-dependent and independent hazards (Table.~\ref{table:1}) had on overall hazard rates by individually removing their effects and examining changes in both $r^2$ and AUROC.
+We then examined the relative contribution species traits, random effects and carbon-dependent and independent hazards (Table.~\ref{table:1}) had on overall hazard rates by individually removing their effects and examining changes in both $r^2$ and AUROC.
 First, we found that the removal of census random effect $\delta_t$ had little impact on either $r^2$ or AUROC.
 The removal of either wood density or light demand resulted in similar magnitude effects on both $r^2$ and AUROC.
-By contrast, the removal of maximum dbh had a slightly smaller effect on these metrics.
+By contrast, the removal of maximum dbh had a slightly smaller effect on both metrics.
 We detected a significant drop in both $r^2$ and AUROC when species random effects (i.e.
@@
 As expected, the removal of both species traits and species random effects resulted in a model that was unable to capture any inter-specific variation observed among species, and thus had low discriminatory ability.
-Lastly, we found the removal of the growth-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the growth-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation.
-However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the growth-dependent hazard resulted in substantially lower predictivie ability.
+Lastly, we found the removal of the carbon-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the carbon-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation.
+However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the carbon-dependent hazard resulted in substantially lower predictive ability.
 
 \section*{Discussion}
-Our Bayesian framework coupled with cross validation revealed that the most predictive hazard model of tropical tree death was one which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) included the effects of species' maximum dbh, wood density and light demand; and 4) incorporated temporal variability.
+Our Bayesian framework coupled with cross validation revealed that the most predictive hazard model of tropical tree death was one which: 1) partitioned mortality into carbon-dependent and carbon-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) included the effects of species' maximum dbh, wood density and light demand; and 4) incorporated temporal variability.
 Moreover, we found that rates of tropical tree mortality varied substantially between species and that the a species' maximum dbh, wood density and light demand explained only a fraction of the overall inter-specific variation.
 
-The findings of this study provide empirical support for dynamic vegetation models that predict tree mortality as a function of both growth-dependent and independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}.
-We show that regardless of growth measure, incorporating both hazards significantly improves model predictive skill.
-This is because the growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses (e.g.
+The findings of this study provide empirical support for dynamic vegetation models that predict tree mortality as a function of both carbon-dependent and independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}.
+We show that regardless of growth measure (our proxy for carbon budget), incorporating both hazards significantly improves model predictive skill.
+This is because the carbon-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses (e.g.
 competition, parasites, herbivory).
-By contrast, the growth-independent hazard accounts for deaths caused by events that arise irrespective of an individual's growth rate (e.g.
+By contrast, the carbon-independent hazard accounts for deaths caused by events that arise irrespective of an individual's carbon budget (e.g.
 windthrow, lightning strike).
@@
 Like many other studies \cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlighted the importance of light competition in influencing tropical tree demographic rates.
-Specifically, we found that that the removal of the growth-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table:1}).
+Specifically, we found that that the removal of the carbon-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table:1}).
 This suggests deficiencies in carbon budgets are a major contributor to tree death on Barro Colorado Island.
@@
 By contrast, both high dbh or wood density species were found to be more tolerant of growth-independent hazards, corrobrating evidence reported elsewhere \cite{Chave:2009iy, Kraft:2010kq}.
-However, our analyses also revealed that high wood density species were also more tolerant of low growth rates, and thus, better able to tolerate shade and competiton \cite{vanGelder:2006exa, Kunstler:2016km}.
+However, our analyses has also supports other work \cite{vanGelder:2006exa, Kunstler:2016km} which highlights that higher wood density species are also more tolerant of low growth rates, and thus, better able to tolerate shade and competiton.
 
-While the effects of species' wood density, maximum size and light demand are now varying being incorporated in mortality algorithms of many vegetation models \cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation (Fig.
+While the effects of species' wood density, maximum size and light demand are now varyingly being incorporated into mortality algorithms of many vegetation models \cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation (Fig.
 \ref{fig4}, Table~\ref{table:1}).
@@
 This was because diameter measurements were rounded to the nearest 5 mm for individuals with dbh < 55 mm, whereas in later censuses all individuals were measured to the nearest millimetre \cite{Condit2017}.
-Consequently, we modelled tree mortality as a function of past growth for censuses 1995\textendash2000, 2000\textendash2005 and 2005\textendash2010.
+Consequently, we modelled tree mortality as a function of past growth (our measure of available carbon budget) for censuses 1995\textendash2000, 2000\textendash2005 and 2005\textendash2010.
 We discarded species that do not exhibit secondary growth (e.g.
@@
 
-We calculated a species' light demand using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995.
+A species' light demand was calculated using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995.
 The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m).
@@
 
-A species' maximum dbh was determined from the data, by taking the 95th percentile observed for each species.
+A species' maximum dbh was determined from directly from the data, by taking the maximum dbh within the 95th percentile observed for each species.
 
@@
 Eqs.
-\ref{eq:mortality_model}-\ref{eq:model} were fit to the data using Bayesian inference and with covariates for growth rate in previous census and wood density, as well as random effects.
+\ref{eq:mortality_model}-\ref{eq:model} were fit to the data using Bayesian inference and with covariates for growth rate in previous census, species traits -- maximum dbh, light demand and wood density, as well as random effects.
 Growth rates were estimated from field measurements of diameter, which inevitably include observation error.
@@
 S1).
-The parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ were modelled as a function of both wood density (measured at species level) and a species-level random effect:
+The parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ were modelled as a function of species traits -- maximum dbh $\psi$, wood density, $\rho$, and light demand, $\upsilon$ -- and a species-level random effect:
 \begin{equation}\label{eq:submodels}
-\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{\rho_c}\right)^{\alpha_{1}},
+\alpha_s = \alpha_{0,s} \, \left(\frac{\rho}{0.6}\right)^{\alpha_{1}}\left(\frac{\upsilon}{0.7}\right)^{\alpha_{2}}\left(\frac{\psi}{15}\right)^{\alpha_{3}},
 \end{equation}
 with similar formulations for $\beta_s$ and $\gamma_s$.
-Here $\alpha_{1}$ captures the effect of wood density $\rho$ on $\alpha_s$, while $\alpha_{0,s}$ captures any other species-level residual error not explained by wood density for species $s$.
+Here $\alpha_{1}$ is the effect of wood density $\rho$, $\alpha_{2}$ is the effect of light demand, $\upsilon$, and $\alpha_{3}$ is the effect of maximum  dbh, $\psi$.
+$\alpha_{0,s}$ captures any other species-level residual error that is otherwise not explained by these three species traits.
 These random effects were modelled as random realisations from log-normal distributions.
 The form of eq.
-\ref{eq:submodels} ensures that parameters remain positive; and on a log scale this equates to an additive linear model centered around $\rho_c$.
+\ref{eq:submodels} ensures that parameters remain positive; and on a log scale this equates to an additive linear model centered around the denominator.
+In this case we centered wood density on 0.6 g cm$^{-3}$, light demand on 0.7 and maximum dbh on 15 cm.
 We also centered growth rate $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.
@@
 Dolins, \& hundreds of field workers for their contribution; and the National Science Foundation, Smithsonian Tropical Research Institute, \& MacArthur Foundation for funding the design, collection, quality control and management of long-term growth data at BCI.
-Fig.
+We also Fig.
 1a uses images by Tracey Saxby, IAN Image Library.

### 2018-07-23 ms/supplementary.Rnw
some final changes to the document

@@
  Growth rates were estimated from field measurements of diameter, which inevitably include observation error.
-In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_dbh)}\% of estimated growth rates were negative.
+In our dataset, \Sexpr{prop_neg_growth(BCI_training_full)}\% of estimated growth rates were negative.
 To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community.
@@
 
-We estimated wood density effects (i.e. $\alpha_1$, $\beta_1$ and $\gamma_1$) using normal priors centered on zero and a standard deviation of 2.5:
+Effects of species traits -- wood density, $\rho$, light demand, $\upsilon$ and maximum dbh, $\psi$ (i.e.
+$\alpha_{1-3}$, $\beta_{1-3}$ and $\gamma_{1-3}$) using normal priors centered on zero and a standard deviation of 2.5:
 \begin{equation} \label{eq:rho_effects}
-\alpha_1, \beta_1, \gamma_1 \sim \mathcal{N}\left(0, 2.5\right).\\
+\alpha_{1-3}, \beta_{1-3}, \gamma_{1-3} \sim \mathcal{N}\left(0, 2.5\right).\\
 \end{equation}
@@
 <<echo=FALSE, highlight=TRUE, comment = ''>>=
-task <- tasks_2_run("final_model")[1, ]
-chunks <- get_final_model_chunks(task)
-model <- make_stan_model(chunks)
-code <- gsub("\n      ", "\n", model$model_code)
+model <- final_model$fits[[1]]@stanmodel@model_code
+code <- gsub("\n      ", "\n", model)
 code <- strsplit(code, "\n")[[1]]
@@
 <<param-table,results="asis",center=TRUE, echo= FALSE>>= 
-hyperparam_table(final_model)
+tableS1(final_model)
 @

### 2018-07-23 R/tableS1.R

@@ -1,0 +1,69 @@
+# Prepare table S1 for manuscript
+#' 
+# Prepare table S1 for manuscript
+#' @param model List containing stan fit object
+#' @return Dataframe
+#' @author James Camac (\email{james.camac@gmail.com}) & Daniel Falster (\email{daniel.falster@unsw.edu.au})
+#' @export
+tableS1 <- function(model) {
+  
+  param1 <- base::as.data.frame(
+    rstan::summary(model$fits[[1]],
+                   c('mu_log_alpha',
+                     "mu_log_beta",
+                     "mu_log_gamma"))$summary) %>%
+    dplyr::mutate(Parameter = row.names(.)) %>%
+    dplyr::select(Parameter, mean, `2.5%`, `97.5%`) %>%
+    dplyr::mutate_at(vars(-Parameter), dplyr::funs(exp))
+  
+  param2 <- base::as.data.frame(rstan::summary(model$fits[[1]], c('a1','a2','a3',
+                                                                  'b1','b2','b3',
+                                                                  'c1','c2','c3'))$summary) %>%
+                                  dplyr::mutate(Parameter = row.names(.)) %>%
+                                  dplyr::select(Parameter, mean, `2.5%`, `97.5%`)
+  
+  param3 <- base::as.data.frame(rstan::summary(model$fits[[1]], 'census_err')$summary) %>%
+    dplyr::mutate(Parameter = row.names(.)) %>%
+    dplyr::select(Parameter, mean, `2.5%`, `97.5%`)
+  
+  x <- dplyr::bind_rows(param1,param2, param3) %>%
+    dplyr::select(Parameter, "Geometric mean" = mean, `2.5%`, `97.5%`) %>%
+    dplyr::mutate_at(dplyr::vars(-Parameter), dplyr::funs(round(.,4))) %>%
+    dplyr::mutate(Parameter = factor(Parameter,
+                                     levels = c("census_err[3]",
+                                                "census_err[2]",
+                                                "census_err[1]",
+                                                "a1",
+                                                "a2",
+                                                "a3",
+                                                "b1",
+                                                "b2",
+                                                "b3",
+                                                "c1",
+                                                "c2",
+                                                "c3",
+                                                "mu_log_gamma",
+                                                "mu_log_beta",
+                                                "mu_log_alpha"),
+                                     labels = c("Census 3 effect ($\\delta_3$)",
+                                                "Census 2 effect ($\\delta_2$)",
[…]

### 2018-07-23 R/table_1.R

@@ -105,8 +105,8 @@ table_1 <- function(model, data) {
       `Full minus max dbh` = full_minus_dbh,
       `Full minus species error` = full_minus_sppre,
       `Full minus species traits & species error` = full_minus_spp,
-      `Full minus growth dependent hazard` = full_minus_growthdep, 
-      `Full minus growth independent hazard` = full_minus_growthindep)
+      `Full minus carbon-dependent hazard` = full_minus_growthdep, 
+      `Full minus carbon-independent hazard` = full_minus_growthindep)
   
   
   # Calculate R2 relative to observed proportion of death per species

### 2018-07-30 ms/manuscript.Rnw
Some of the changes suggested by @dfalster

@@
 % 120 words
-\significancestatement{We build a mortality model that not only integrates intra and inter-specific variation but also, simulataneously estimates the two fundamental hazards faced by plants -- Death due to low carbon budget and death independent of carbon budget -- and how a species' light demand, maximum dbh and wood density affect each.
+\significancestatement{We build a mortality model that not only integrates intra and inter-specific variation but also, simulataneously estimates the two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth -- and how a species' light demand, maximum dbh and wood density affect each.
 We parameterise this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals across 203 species.
-We show that carbon-dependent hazards are a significant contribtor to tree mortality at Barro Colorado Island.
+We show that growth-dependent hazards are a significant contribtor to tree mortality at Barro Colorado Island.
 Moreover, we show that species traits -- wood density, maximum dbh and light demand -- all affect tree mortality rates but ultimately only account for a small fraction of the total variability among species.}
@@
 In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive hazard functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
-Specifically, we develop an empricial mortality model that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the independent contributions of carbon-related hazards and carbon-independent hazards on the overall mortality rates.
-We show that functions that separate mortality rates into carbon-dependent and carbon-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and light demand outperform alternative formulations.
-Moreover, we show that while these three traits do affect both carbon-dependent and carbon-independent hazards, they account for only a small fraction of the total variability observed among species.
+Specifically, we develop an empricial mortality model that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the independent contributions of growth-related hazards and growth-independent hazards on the overall mortality rates.
+We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and light demand outperform alternative formulations.
+Moreover, we show that while these three traits do affect both growth-dependent and growth-independent hazards, they account for only a small fraction of the total variability observed among species.
 Combined, this study provides a statistical framework for predicting individual-level mortality in highly diverse tropical forests.
@@
 Plants die because of many causes, however, these broadly fall into two main categories \cite{Hawkes:2000ib,McDowell:2011dr}.
-The first are carbon-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair.
-The second are carbon-independent hazards, where plants die because of chance events that are unrelated to their carbon budget, such as windfall or fire.
+The first are growth-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair.
+The second are growth-independent hazards, where plants die because of chance events that are unrelated to their carbon budget, such as windfall or fire.
 An individual's total hazard at a given time point is therefore the sum of these two independent hazards.
@@
 \caption{Outline of methodology.
-A) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
-B) We consider three alternative hazard functions: 1) a carbon-independent baseline hazard (i.e.
-average hazard), 2) a carbon-dependent hazard, where growth rate is a proxy for carbon budget; and 3) a function that combines both carbon-dependent and independent hazards.
+A) We consider three alternative hazard functions: 1) a growth-independent baseline hazard (i.e.
+average hazard; Panel A; left), 2) a growth-dependent hazard, where growth rate is a proxy for carbon budget (Panel A; middle); and 3) a function that combines both growth-dependent and independent hazards (Panel A; right).
 The parameters of the models are biologically interpretable: $\alpha$ defines the low growth rate (i.e.
-low carbon budget) effect on the overall hazard; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote (Panel B; right).
-In the most complex model (Panel B; right), $\alpha$ and $\beta$ capture carbon-dependent hazards, while $\gamma$ captures carbon-independent hazards that kill a plant, irrespective of its growth rate/carbon budget (e.g.
+low carbon budget) effect on the overall hazard; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote.
+In the most complex model (Panel A; right), $\alpha$ and $\beta$ capture growth-dependent hazards, while $\gamma$ captures growth-independent hazards that kill a plant, irrespective of its growth rate/carbon budget (e.g.
 windfall, fire).
 For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
+B) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
 C) Each model's skill in predicting observed outcomes ($S$) is quantified via the log-loss function (eq. \ref{eq:logloss}). Lower logloss means higher predictive accuracy.
@@
 
-Here, we develop a tree mortality model that not only reconciles the conflicting growth-mortality relationships observed among individuals and between species, but also explicitly accounts for carbon-dependent and carbon-independent hazards.
+Here, we develop a tree mortality model that not only reconciles the conflicting growth-mortality relationships observed among individuals and between species, but also explicitly accounts for growth-dependent and growth-independent hazards.
 We achieve this by evaluating the following unified hazard function that incorporates individual-, species-, and temporal effects:
 \begin{equation} \label{eq:model}
-\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Carbon independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{s,i}(t)}}^{\text{Carbon dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
+\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Growth independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{s,i}(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
 \end{equation}
 Eq.
-\ref{eq:model} allows for the additive, thus independent, effects of carbon-dependent and carbon-independent hazards, includes a negative exponential effect of individual's growth rate (our measure of carbon budget), and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
+\ref{eq:model} allows for the additive, thus independent, effects of growth-dependent and growth-independent hazards, includes a negative exponential effect of individual's growth rate (our measure of carbon budget), and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
 Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$.
@@
 carbon budget).
-This includes two possible hazard functions: a carbon-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + carbon-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters.
-Symbols represent the three hazard functional form: baseline (circle), carbon-dependent (triangle), and hazard containing both carbon-independent and carbon-dependent hazards (square).}
+This includes two possible hazard functions: a growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters.
+Symbols represent the three hazard functional form: baseline (circle), growth-dependent (triangle), and hazard containing both growth-independent and growth-dependent hazards (square).}
 \label{fig2}
@@
 To validate this model, we also fit two alternative hazard functions: 1) a baseline hazard (i.e.
-a null model) and 2) a carbon-dependent only hazard.
+a null model) and 2) a growth-dependent only hazard.
 We then compare each functions ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from \Sexpr{n_spp(BCI_training_full)} species recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama (Fig.
@@
 The third form (Fig.
-\ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike the second functional form examined here, asymptotes at some carbon-independent baseline hazard $\textgreater 0$.
+\ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike the second functional form examined here, asymptotes at some growth-independent baseline hazard $\textgreater 0$.
 Combined, these three models capture a variety of functional responses that are both used in empirical studies of plant mortality \cite{Kobe:1995tw,Wyckoff:2000un} and are represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, but which have not previously been systematically compared.
-In doing this, we also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts carbon-dependent mortality.
+In doing this, we also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.
 
-In the remaining stages, we take the most predictive hazard function and growth rate measure from stage 2 and examine how species traits: wood density, light demand and maximum dbh, affected model predictive performance.
+In the remaining stages, we take the most predictive hazard function and growth rate measure from stage 2 and examine how three species traits considered major explanatory variables of inter-specific variation in tropical rainforests \cite{king:2006va,Poorter:2006tq, Chave:2009iy}: wood density, light demand and maximum dbh, affected model predictive performance.
 Specifically, we examined each trait individually on all three $\alpha$, $\beta$ and $\gamma$ (stage 3), as well as 2-trait (stage 4) and 3-trait (stage 5) combinations on these three parameters.
@@
 \ref{fig2}A).
-Table S1 gives individual census effects.
+Table S1 provides individual census effects.
 
@@
 Comparing the three hazard functions summarised in Fig.
-\ref{fig1}B, we found that the third function --- the function containing both carbon-dependent and carbon-independent hazard terms, i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline and carbon-dependent only hazard functions (Fig.
+\ref{fig1}B, we found that the third function --- the function containing both growth-dependent and growth-independent hazard terms, i.e.
+$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline and growth-dependent only hazard functions (Fig.
 \ref{fig2}).
@@
 \ref{fig2}A).
-Table S1 gives a summary of mean hyper parameter estimates.
+Table S1 gives a summary of mean parameter estimates.
 
@@
 \includegraphics{figures/fig3.pdf}
-\caption{Estimated relationships between between species traits (maximum dbh, light demand and wood density) and model parameters $\alpha$ (low growth effect), $\beta$ (growth-mortality decay rate) and $\gamma$ (baseline/carbon-independent hazard).
+\caption{Estimated relationships between between species traits (maximum dbh, light demand and wood density) and model parameters $\alpha$ (low growth effect), $\beta$ (growth-mortality decay rate) and $\gamma$ (baseline/growth-independent hazard).
 Points are the estimated mean ($\pm$ 95\% credible intervals) parameters for each of the 203 Barro Colorado Island species used in this study.
@@
 \ref{fig2} Stage 6).
-See Figs.
+See Tables S1 for trait coefficients and Table S2 or Figs.
 S3-S5 for species-level parameter estimates.
@@
 lower $\beta$).
-By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower carbon-independent hazards.
-Species' maximum dbh only had significant effects on $\gamma$, whereby species that can achieve larger dbh exhibited lower carbon-independent hazards.
+By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards.
+Species' maximum dbh only had significant effects on $\gamma$, whereby species that can achieve larger dbh exhibited lower growth-independent hazards.
 Table S1 for trait effects on each parameter.
@@
 
-We then examined the relative contribution species traits, random effects and carbon-dependent and independent hazards (Table.~\ref{table:1}) had on overall hazard rates by individually removing their effects and examining changes in both $r^2$ and AUROC.
+We then examined the relative contribution species traits, random effects and growth-dependent and independent hazards (Table.~\ref{table:1}) had on overall hazard rates by individually removing their effects and examining changes in both $r^2$ and AUROC.
 First, we found that the removal of census random effect $\delta_t$ had little impact on either $r^2$ or AUROC.
@@
 As expected, the removal of both species traits and species random effects resulted in a model that was unable to capture any inter-specific variation observed among species, and thus had low discriminatory ability.
-Lastly, we found the removal of the carbon-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the carbon-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation.
-However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the carbon-dependent hazard resulted in substantially lower predictive ability.
+Lastly, we found the removal of the growth-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the growth-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation.
+However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the growth-dependent hazard resulted in substantially lower predictive ability.
 
 \section*{Discussion}
-Our Bayesian framework coupled with cross validation revealed that the most predictive hazard model of tropical tree death was one which: 1) partitioned mortality into carbon-dependent and carbon-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) included the effects of species' maximum dbh, wood density and light demand; and 4) incorporated temporal variability.
+Our Bayesian framework coupled with cross validation revealed that the most predictive hazard model of tropical tree death was one which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) included the effects of species' maximum dbh, wood density and light demand; and 4) incorporated temporal variability.
 Moreover, we found that rates of tropical tree mortality varied substantially between species and that the a species' maximum dbh, wood density and light demand explained only a fraction of the overall inter-specific variation.
 
-The findings of this study provide empirical support for dynamic vegetation models that predict tree mortality as a function of both carbon-dependent and independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}.
+The findings of this study provide empirical support for dynamic vegetation models that predict tree mortality as a function of both growth-dependent and independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}.
 We show that regardless of growth measure (our proxy for carbon budget), incorporating both hazards significantly improves model predictive skill.
-This is because the carbon-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses (e.g.
+This is because the growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses (e.g.
 competition, parasites, herbivory).
-By contrast, the carbon-independent hazard accounts for deaths caused by events that arise irrespective of an individual's carbon budget (e.g.
+By contrast, the growth-independent hazard accounts for deaths caused by events that arise irrespective of an individual's carbon budget (e.g.
 windthrow, lightning strike).
 
-Additionally, the partitioning of mortality rates into both carbon-dependent and carbon-independent effects allowed us to not only estimate their relative contribution, but also how species traits affect each.
-Like many other studies \cite{King:1994vg, Ruger:2011cv, Kunstler:2016km}, our analyses highlighted the importance of light competition in influencing tropical tree demographic rates.
-Specifically, we found that that the removal of the carbon-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table:1}).
+Additionally, the partitioning of mortality rates into both growth-dependent and growth-independent effects allowed us to not only estimate their relative contribution, but also how species traits affect each.
+Like many other studies \cite{King:1994vg,king:2006va, Ruger:2011cv, Kunstler:2016km}, our analyses highlighted the importance of light competition in influencing tropical tree demographic rates.
+Specifically, we found that that the removal of the growth-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table:1}).
 This suggests deficiencies in carbon budgets are a major contributor to tree death on Barro Colorado Island.
 
-Our analyses support previous research that advocated that wood density, species maximum size and light demand important explanatory variables of inter-specific variation in tropical rainforests \cite{Poorter:2006tq, Chave:2009iy}.
+Our analyses support previous research that advocated that wood density, species maximum size and light demand important explanatory variables of inter-specific variation in tropical rainforests \cite{king:2006va,Poorter:2006tq, Chave:2009iy}.
 Specificially, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates.
@@
 A species' maximum dbh was determined from directly from the data, by taking the maximum dbh within the 95th percentile observed for each species.
+
+Weak correlation (\textless 0.36) was found between species' wood density, light demand and maximum dbh (See Fig. S6)
 
@@
 \showacknow % Display the acknowledgements section
+\vspace{-5em}
 
-\vspace{-5em}
-\pnasbreak
-
+\newpage
 % Bibliography
 \bibliography{refs.bib}
-%\pnasbreak
+
 \end{document}

### 2018-07-30 R/plot_fig1.R

@@ -25,22 +25,20 @@ plot_fig1 <- function(file_alive, file_dead, file_crossval) {
     1-exp(-(0.13 * exp(-10 * x) + 0.03))
   }
   
+  # Plot hazard curves
+  hazard_plot <- function(f, text){
+    x <- seq(0,1,length.out = 50)
+    empty_plot()
+    lines(x, f(x), type='l', col="#d7301f")
+    text(0.1, 0.12, text, pos=4, xpd=NA)
+  }
+  
   # Function to add labels to panels
   my_label <- function(text, x=-0.1) label(x, 1.3, text)
   
   # Plot layout
-  layout(matrix(c(1,1,1,2,3,4,5,5,5,6,6,6), byrow=TRUE, ncol=3))
+  layout(matrix(c(1,2,3,4,4,4,5,5,5,6,6,6), byrow=TRUE, ncol=3))
   par(mar=c(1,1,2.5,1), oma=c(1,1,1,1), cex=0.5)
-  
-  # Tree growth diagram
-  plot(1:2, type='n', ann=FALSE, axes=FALSE, xlim=c(-1,1), ylim=c(-1,1))
-  my_label("A) Repeat census data", x=-0.06)
-  
-  # note we are only plotting to last viewport, but calls to other two needed to make figure work.
-  vps <- baseViewports()
-  pushViewport(vps$inner, vps$figure, vps$plot)
-  plot_tree_life(file_alive, file_dead)
-  popViewport(3)
   
   # Produce empty plot
   empty_plot <- function(){
@@ -50,17 +48,9 @@ plot_fig1 <- function(file_alive, file_dead, file_crossval) {
     axis(2, at = c(-1, 2),labels=NA)
   }
   
-  # Plot hazard curves
-  hazard_plot <- function(f, text){
-    x <- seq(0,1,length.out = 50)
-    empty_plot()
-    lines(x, f(x), type='l', col="#d7301f")
-    text(0.1, 0.12, text, pos=4, xpd=NA)
-  }
-  
   # Plot base hazard
   hazard_plot(base_only_haz, expression(gamma))
-  my_label("B) Alternative mortality functions", x=-0.36)
+  my_label("A) Alternative mortality functions", x=-0.36)
   mtext("Mortality rate", 2, line=1, cex=0.5)
   
   # Plot growth-dependent hazard
@@ -68,7 +58,17 @@ plot_fig1 <- function(file_alive, file_dead, file_crossval) {
[…]

### 2018-07-30 R/plot_fig2.R

@@ -62,7 +62,7 @@ plot_fig2 <- function(logloss_summaries) {
     ggplot2::scale_x_discrete(labels=c("null_model_null_model_none" = expression(gamma),
                                        "function_growth_comparison_base_hazard_none" = expression(gamma~delta["t"]),
                                        "function_growth_comparison_growth_hazard_none" = expression((alpha*"e"^{-beta~"X"["i"]})~delta["t"]),
-                                       "function_growth_comparison_base_growth_hazard_none" = expression((alpha*"e"^{-beta~"X"["i"]} + gamma)~delta["t"]),
+                                       "function_growth_comparison_base_growth_hazard_none" = expression((gamma + alpha*"e"^{-beta~"X"["i"]})~delta["t"]),
                                        "size_combinations_base_growth_hazard_abc" = expression("Max dbh"~(psi)),
                                        "rho_combinations_base_growth_hazard_abc" = expression("Wood density"~(rho)),
                                        "gap_combinations_base_growth_hazard_abc" = expression("Light index"~(upsilon)),
@@ -70,7 +70,7 @@ plot_fig2 <- function(logloss_summaries) {
                                        "rho_gap_all_base_growth_hazard_rho_gap_abc" = expression(rho + upsilon),
                                        "gap_size_all_base_growth_hazard_gap_size_abc" = expression(psi + upsilon),
                                        "multi_trait_all_base_growth_hazard_rho_gap_size_abc" = expression(psi + rho + upsilon),
-                                       "species_random_effects_base_growth_hazard_none" = expression((alpha[s]*"e"^{-beta[s]~"X"["i"]} + gamma[s])~delta["t"]))) +
+                                       "species_random_effects_base_growth_hazard_none" = expression((gamma + alpha[s]*"e"^{-beta[s]~"X"["i"]})~delta["t"]))) +
     ggplot2::facet_grid(.~stage + comparison, scales='free_x', drop=TRUE, space = "free_x", labeller = labeller(comparison = ggplot2::label_parsed, stage = ggplot2::label_parsed)) +
     plot_theme(strips = TRUE) +
     ggplot2::theme(axis.text.x = ggplot2::element_text(angle=15, hjust = 1),

### 2018-07-30 R/plot_fig3.R

@@ -10,15 +10,19 @@
 plot_fig3 <- function(spp_params, mu_trend) {
   
   spp_params <- spp_params %>%
-    mutate(trait_value2 = ifelse(trait=="dbh_95", log10(trait_value),trait_value))
+    mutate(trait_value2 = ifelse(trait=="dbh_95", log10(trait_value),trait_value),
+           param = factor(param, levels=c("gamma","alpha","beta")),
+           trait = factor(trait, levels=c("gap_index","wood_density","dbh_95")))
   
   mu_trend <- mu_trend %>%
-    mutate(trait_value2 = ifelse(trait=="dbh_95", log10(trait_value),trait_value))
+    mutate(trait_value2 = ifelse(trait=="dbh_95", log10(trait_value),trait_value),
+           param = factor(param, levels=c("gamma","alpha","beta")),
+           trait = factor(trait, levels=c("gap_index","wood_density","dbh_95")))
   
-  breaks <- c(0.001,0.01,0.1, 1, 10)
-  labels <- sapply(log10(breaks),function(i) as.expression(bquote(10^ .(i))))
-  trait_label_lookup <- c(wood_density ="Wood~density~(g~cm^-3)", gap_index = "Light~demand", dbh_95 = "log10(Max~dbh~(cm))")
-  param_label_lookup <- c(alpha = "Low~growth~effect~(alpha)", beta = "Growth~decay~rate~(beta)", gamma ="Baseline~hazard~(gamma)")
+  breaks <- c(0.001,0.01,0.1, 1, 5, 10, 20, 40, 80)
+  labels <- sapply(round(log10(breaks),2),function(i) as.expression(bquote(10^ .(i))))
+  trait_label_lookup <- c(gap_index = "Light~demand", wood_density ="Wood~density~(g~cm^-3)", dbh_95 = "log10(Max~dbh~(cm))")
+  param_label_lookup <- c(gamma ="Baseline~hazard~(gamma)",alpha = "Low~growth~effect~(alpha)", beta = "Growth~decay~rate~(beta)")
   
   ggplot2::ggplot(spp_params, ggplot2::aes(x = trait_value2,y = mean)) + 
     ggplot2::geom_ribbon(data = mu_trend, ggplot2::aes(x = trait_value2,ymin = `2.5%`,ymax = `97.5%`), alpha=0.5, colour=NA) +
@@ -29,7 +33,7 @@ plot_fig3 <- function(spp_params, mu_trend) {
[…]

### 2018-07-30 R/plot_fig4.R

@@ -26,6 +26,16 @@ plot_fig4 <- function(model,
   
   p2 <- plot_mu_curves(model, 
                        growth_range, 
+                       trait_name = "gap_index", 
+                       trait_values = c(0.5, 0.8), 
+                       hazard_curve, 
+                       xlab= NULL, 
+                       ylab = NULL) +
+    plot_theme() +
+    theme(plot.margin=unit(c(0.35,0.2,0.5,0.5),"cm"))
+  
+  p3 <- plot_mu_curves(model, 
+                       growth_range, 
                        trait_name = "wood_density", 
                        trait_values = c(0.2,0.8), 
                        hazard_curve, 
@@ -34,7 +44,7 @@ plot_fig4 <- function(model,
     plot_theme() + 
     theme(plot.margin=unit(c(0.35,0.2,0.5,0.5),"cm"))
 
-  p3 <- plot_mu_curves(model,
+  p4 <- plot_mu_curves(model,
                        growth_range, 
                        trait_name = "dbh_95", 
                        trait_values = c(3,180),
@@ -44,20 +54,10 @@ plot_fig4 <- function(model,
         plot_theme() +
         theme(plot.margin=unit(c(0.35,0.2,0.5,0.5),"cm"))
   
-  p4 <- plot_mu_curves(model, 
-                       growth_range, 
-                       trait_name = "gap_index", 
-                       trait_values = c(0.5, 0.8), 
-                       hazard_curve, 
-                       xlab= NULL, 
-                       ylab = NULL) +
-    plot_theme() +
-    theme(plot.margin=unit(c(0.35,0.2,0.5,0.5),"cm"))
-  
   # ("1-yr mortality probability")
   cowplot::plot_grid(p1,p2,p3,p4, ncol=1, labels=c("A) Species curves",
-                                                   "B) Wood density",
-                                                   "C) Maximum dbh",
+                                                   "B) Maximum dbh",
+                                                   "C) Wood density",
                                                    "D) Light demand"), 
                      label_size = 7, 
                      label_fontface ="plain") +

### 2018-07-30 R/tableS1.R

@@ -63,7 +63,7 @@ tableS1 <- function(model) {
   
   Hmisc::latex(x, file = "", booktabs = TRUE, rowname = NULL,
                colnamesTexCmd="bfseries",
-               col.just = c(rep("c", 4)), label = "table:1",
+               col.just = c("c",rep("l", 3)), label = "table:S1",
                where = "!h",
                caption ="Mean hyper-parameters and coefficients for both species trait and census effects as estimated from full model")
 }

### 2018-07-30 R/tableS2.R

@@ -1,0 +1,26 @@
+# Prepare table S2 for manuscript
+#' 
+# Prepare table S2 for manuscript
+#' @param spp_params Estimated species level parameters
+#' @return Dataframe
+#' @author James Camac (\email{james.camac@gmail.com}) & Daniel Falster (\email{daniel.falster@unsw.edu.au})
+#' @export
+tableS2 <- function(spp_params) {
+  
+  x <- spp_params %>%
+    dplyr::select(param,species, mean, `2.5%`,`97.5%`) %>%
+    dplyr::distinct() %>%
+    mutate_if(is.numeric, funs(round),4) %>%
+    mutate(stat = paste0(mean," (",`2.5%`,",",`97.5%`,")")) %>%
+    dplyr::select(species, param, stat) %>%
+    tidyr::spread(param, stat) %>%
+    dplyr::select("Species" = species, "$\\gamma$" = gamma, "$\\alpha$" = alpha, "$\\beta$" = beta)
+  
+  Hmisc::latex(x, file = "", booktabs = TRUE, rowname = NULL,
+               colnamesTexCmd="bfseries",
+               col.just = c("c",rep("l", 3)), label = "table:S2",
+               where = "!h",
+               longtable = TRUE,
+               lines.page = 51,
+               caption ="Species geometric mean ($\\pm$ 95$\\%$ credible interval) for $\\alpha$, $\\beta$ and $\\gamma$. Estimates derived from full model")
+}

### 2018-08-08 ms/manuscript.Rnw
Moderate restructue

@@
 
-\title{Unifying intra- and inter-specific variation in tropical tree mortality}
+\title{Unify intra- and inter-specific drivers of tropical tree mortality via additive hazards approach}
 
@@
 % 120 words
-\significancestatement{We build a mortality model that not only integrates intra and inter-specific variation but also, simulataneously estimates the two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth -- and how a species' light demand, maximum dbh and wood density affect each.
-We parameterise this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals across 203 species.
-We show that growth-dependent hazards are a significant contribtor to tree mortality at Barro Colorado Island.
+\significancestatement{We build a mortality model that not only integrates intra and inter-specific variation but also, simultaneously estimates the two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth -- and how a species' light demand, maximum dbh and wood density affect each.
+We parameterize this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals across 203 species.
+We show that growth-dependent hazards are a significant contributor to tree mortality at Barro Colorado Island.
 Moreover, we show that species traits -- wood density, maximum dbh and light demand -- all affect tree mortality rates but ultimately only account for a small fraction of the total variability among species.}
@@
 
-\dropcap{R}ates of plant mortality vary widely among individuals within a species, among coexisting species, between forests, and from year-to-year \cite{Kobe:1995tw,Wyckoff:2002ul,Poorter:2008iu}.
-This variation has large consequences for forest structure and dynamics.
-The death of a large tree can transfer up to 20,000 kg of carbon from living to decaying carbon pools \cite{Chave:2005hz}, while also creating a gap in the canopy that restarts a successional race, during which 100's of plants will die while competing for a spot in the sun.
-Models of forest dynamics suggest variation in tree mortality rates can have larger impacts on forest structure relative than variation in growth rates \cite{Johnson-2016}.
-Improving our understanding of the mortality process is therefore a priority for accurately predicting changes in population, carbon and nutrient dynamics of forests; especially in an era of rapid environmental change \cite{Anderegg:2012ep,Mcdowell:2013wf}.
+Tree death is a key process driving population dynamics, nutrient cycling, and evolution within plant communities.
+While the risk of death varies among forests and from year-to-year \cite{vanMantgem:2009bi}, trees growing within the same forest also vary remarkably in realised risk of death\cite{Condit:2006}.
+In tropical forests, annual mortality can differ 20-60 fold among populations of coexisting species \cite{Condit:2006}.
+These differences reflect differences in the way trees from different species are constructed and function, and arise via a coordinated axis of life-history variation, in which individuals from species with higher potential-growth rates also experience higher mortality \cite{Wright:2010fl}.
+At the same time, the risk of death varies widely within species, with those individuals growing slower or sitting lower in the light spectrum at greater risk \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv}.
+While these inter- and intra-specific patterns of mortality are widely recognised, they show conflicting relationships with tree growth rate (increasing with growth rate across species, decreasing with growth rate within species), making it difficult to integrate both sources of variation into a predictive model.
 
-Two difficulties arise when studying tree mortality in tropical rainforests.
-The first is that large population sizes and long periods of observation are required to make accurate inferences into how various factors affect tree mortality rates \cite{Poorter:2008iu}.
-This requirement arises in part because the observed outcome -- alive vs.
-dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the mortality (or `hazard`) rate for individual $i$ at time $t$.
-In classic survival analysis \cite{Kleinbaum:2012tn, Zens:2003jv}, the probability an individual dies between times $t_1$ and $t_2$ is a function of both $\lambda_i(t)$, and the amount of time lapsed:
+In this article we outline a new approach for predicting mortality among individuals within a forest, incorporating both inter and intra-specific variation in risk.
+Our approach builds on the notion that the total risk of death at any time is the additive effect of at least two, largely independent risks \cite{Keane:2001db}.
+Although plants die because of many causes, these broadly fall into two main categories \cite{Keane:2001db, McDowell:2011dr}.
+The first are growth-independent hazards, which cause death irrespective of the tree's current growth rate, and may include windfall, fire, and certain types of pathogen or herbivore attack.
+These hazards are assumed to occur at some average rate (Fig 1a, left).
+The second are growth-dependent hazards, which cause slower growing plants within each species to die at a faster rate, presumably because inferior carbon budgets make them unable to tolerate diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
+Many empirical studies suggest an exponential decline in risk with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv} (Fig 1a, centre).
+An individual's total risk of dying is then the sum of these two (Fig 1a, right).
+Biologically, the additive model implies that even fast-growing individuals experience some risk of death.
+
+While already applied in some vegetation models \cite{Moorcroft:2001ws}, the additive hazards model has not previously been applied in empirical studies.
+Moreover, we note that the underlying mathematical structure of statistical methods used in most previous empirical studies imply responses similar to either the left or centre panels of Fig 1a (see SI Appendix S1).
+In particular, commonly used methods -- including exponential models \cite{Kobe:1995tw, Wyckoff:2000un}, logistic regression \cite{Ruger:2011cv}, generalized linear models \cite{Visser_functional_2016}, and Cox proportional-hazard models \cite{Neumann_climate_2017} -- handle covariates in a manner that allows predicted risk to decline to zero for fast growing individuals (as in Fig 1a centre; SI Appendix S1).
+These methods may therefore fail to capture the full range of variation observed within and among species.
+
+\section*{An additive hazards models with species effects}
+
+In forests, mortality rates are usually estimated from census data, where the size and fate of individual trees is recorded at successive time points \cite{Anderson-Teixeira-2015a} (Fig 1b).
+The size and status of individuals at each census -- alive vs.
+dead -- can then be used to estimate $\lambda_i(t)$, a continuous, latent, mortality (or `hazard`) rate for all individuals in group $i$ at time $t$.
+Using this approach, the probability an individual dies between times $t_1$ and $t_2$ results from the cumulative hazard they are exposed to between censuses\cite{Kleinbaum:2012tn}:
 \begin{equation} \label{eq:mortality_model}
-p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
+p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right),
 \end{equation}
-Here $\int_{t_1}^{t_2} \lambda_i(t) {\rm d}t$ is the cumulative hazard an individual experiences from  $t_1$ and $t_2$\cite{Kleinbaum:2012tn}.
-The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then a realisation of this cumulative hazard on a probability scale:
-\begin{equation}
-S_{i,t_1\rightarrow t_2} \sim \textrm{Bernoulli}(p_{i,t_1\rightarrow t_2}).
+The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then modelled as a binary draw.
+The challenge is then to estimate an appropriate form for the hazard rate, $\lambda_i(t)$.
+
+Here we propose the following hazard function, incorporating individual-, species-, and temporal effects:
+\begin{equation} \label{eq:model}
+\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Growth independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{s,i}(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
 \end{equation}
-Second, as trees are long-lived and we are estimating the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes collected over long durations are required.
-Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.
-
-\section*{Towards a unified model of tree mortality}
-
-Plants die because of many causes, however, these broadly fall into two main categories \cite{Hawkes:2000ib,McDowell:2011dr}.
-The first are growth-dependent hazards, where plants die because of insufficient carbon assimilation for growth and repair.
-The second are growth-independent hazards, where plants die because of chance events that are unrelated to their carbon budget, such as windfall or fire.
-An individual's total hazard at a given time point is therefore the sum of these two independent hazards.
-However, to date, most statistical models of tree mortality do not explicitly estimate the independent effects of these two fundamanetal hazards faced by plants.
-Instead, plant mortality is typically modelled using: 1) logistic regression (i.e.
-with logit link), which does not estimate a rate, but rather the log odds of death at a given time point; or 2) via a proportional hazards approach such as a Cox proportional-hazard model or complementary-log-log link binomial regression, both of which assume hazards are multiplicative, and therefore, not independent of one another \cite{Kleinbaum:2012tn}.
-
-There are several distinct advantages of modelling tree mortality as a function of two additive, and thus independent hazards: 1) The model explicitly accounts for the independent effects of the two fundamental types of hazards faced by plants; 2) it allows for the examination of the relative contributions each hazard provides to the overall hazard rate; 3) it allows for examination of how various factors affect each hazard term, and because of this; 4) allows ecological theories of how factors such as species traits affect plant mortality to be empirically tested using commonly collected data.
-
-A major challenge in developing hazard functions for plants is the need to integrate the conflicting relationships observed within and across species between mortality and growth.
-Within species, mortality rates are lower for fast-growing individuals, presumably because those individuals have superior carbon budgets and are thus able to tolerate or repair diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
-Empirical studies broadly support this theory, with many indicating exponential-like declines in mortality with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv}.
-By contrast, across species, there is a strong trade-off between growth and survival, with faster-growing species exhibiting higher mortality rates relative to slower-growing species \cite{Poorter:2008iu, Wright:2010fl}.
-This trade-off between growth and survival may arise via species traits with antagonistic effects, such as wood density.
-Denser wood -- which is expensive to build and thus slows individual growth --  reduces stem breakage \cite{vanGelder:2006exa,Chave:2009iy}, embolism \cite{Hacke:2001kj, Jacobsen:2005fx}, pathogen attack \cite{Augspurger:1984wx}, and thereby mortality \cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.
-However, to date, these intra- and inter-specific effects have not yet been integrated into a single hazard function in such a way that it allows one to examine their effects on the two fundamental hazards faced by plants.
+Eq.
+\ref{eq:model} allows for additive growth-dependent and growth-independent effects, includes a negative exponential effect of individual's growth rate (our measure of carbon budget), allows the overall mortality rate to vary among censuses via the random effect $\delta_{t}$, while also allowing species-level differences in the shape of $\lambda_i(t)$ via the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$.
+The species-level parameters are in turn modelled as random effects, drawn from  an overall distribution.
 
@@
 \caption{Outline of methodology.
-A) We consider three alternative hazard functions: 1) a growth-independent baseline hazard (i.e.
-average hazard; Panel A; left), 2) a growth-dependent hazard, where growth rate is a proxy for carbon budget (Panel A; middle); and 3) a function that combines both growth-dependent and independent hazards (Panel A; right).
+A) We consider three alternative hazard functions: 1) a growth-independent baseline hazard (A, left), 2) a growth-dependent hazard, where growth rate is a proxy for carbon budget (A, middle); and 3) a function that combines both growth-dependent and independent hazards (A, right).
 The parameters of the models are biologically interpretable: $\alpha$ defines the low growth rate (i.e.
@@
 
-Here, we develop a tree mortality model that not only reconciles the conflicting growth-mortality relationships observed among individuals and between species, but also explicitly accounts for growth-dependent and growth-independent hazards.
-We achieve this by evaluating the following unified hazard function that incorporates individual-, species-, and temporal effects:
-\begin{equation} \label{eq:model}
-\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Growth independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{s,i}(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
-\end{equation}
-Eq.
-\ref{eq:model} allows for the additive, thus independent, effects of growth-dependent and growth-independent hazards, includes a negative exponential effect of individual's growth rate (our measure of carbon budget), and allows for mortality to vary among censuses, via the random effect $\delta_{t}$.
-Further, the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$ vary by species, $s$.
-Here we include the effects of three species traits thought to be major drivers of inter-specific variation in tropical tree mortality -- wood density ($\rho$), light rdemand ($\upsilon$) and maximum dbh ($\psi$).
-We also include species random effects in order to capture species-level differences that are not otherwise accounted for by species traits (for details see Methods).
+\section*{Case study: Tropical forest at Barro Colorado Island}
+
+To illustrate the proposed mortality model, we assess it's ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from containing \Sexpr{n_spp(BCI_training_full)}  species, recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama.
+Mortality rates are known to vary widely among species, and also vary within species in response to light levels and growth rates \cite{Condit:2006, Wright:2010fl, Ruger:2011cv}.
+Applying the additive model, we: i) estimate growth-dependent and growth-independent effects from long-term census data, ii) estimate species-level differences in these effects, and iii) assess how well strength species-level differences is predicted from three key traits, previously shown to correlate with mortality.
+Here we consider three species traits thought to drive inter-specific variation in mortality: wood density, light demand, and maximum dbh \cite{king:2006va, Kraft:2010kq, Wright:2010fl, Ruger:2011cv}.
+Overall, these traits only show weak correlation (r$^2$ \textless 0.13) and thus provide relatively independent measures of a species ecological strategy (SI Appendix, Fig.
+S6).
+
+Our analysis proceeds in a series of stages, in which we compare the predictive power of alternative models, differing in either structure or covariates included.
+We used 10-fold cross-validation to evaluate the skill of different models in their ability to predict outcomes in novel data (i.e.
+not used in model fitting), using a likelihood-based penalty for incorrect prediction (Fig.
+\ref{fig1}C-D; see Methods) \cite{Hooten:2015, Stone:1997tg}.
+n total \Sexpr{n_obs(BCI_training_full)} observations were used to fit these models.
+Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations.
+We therefore fit models across six iterative stages of model development, described below, gradually building up in complexity, comparing their predictive skill on new data.
+We then use the final model to assess how traits align with different effects and quantify the contributions of different effects on overall model performance.
+
+\section*{Results}
 
@@
 
-To validate this model, we also fit two alternative hazard functions: 1) a baseline hazard (i.e.
-a null model) and 2) a growth-dependent only hazard.
-We then compare each functions ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from \Sexpr{n_spp(BCI_training_full)} species recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama (Fig.
-\ref{fig1}).
-The data are repeat censuses of stem diameter and tree status (alive vs.
-dead) taken approximately at 5-year intervals over a 15 yr period (Fig.
+\subsection*{Stage 1: Mortality over time}
+We first assess whether average mortality rates varied among the three censuses, and thus, whether a census effect is required in Eq \ref{eq:model}.
+We compared a model with a constant mortality rate, $\gamma$, to a model where the constant is scaled by a census effect, $\gamma\delta_{t}$.
+Examining the three 5-year census intervals, we found average mortality rates declining over time: with the highest proportion of death occurring from 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%).
+Consequently, we observed a small increase in predictive skill when a census effect was included (Fig.
+\ref{fig2}).
+
+\subsection*{Stage 2: Hazard functions}
+Second, we assessed the predictive performance of three alternative hazard functions (Fig.
 \ref{fig1}A).
-In total \Sexpr{n_obs(BCI_training_full)} observations were used to fit these models.
-We compare the skill of different hazard functions in predicting outcomes in novel data (i.e.
-not used in model fitting) via 10-fold cross-validation (Fig.
-\ref{fig1}D) \cite{Hooten:2015, Stone:1997tg}.
-Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations.
-We therefore fit models across six iterative stages of model development.
-
-In stage 1, we first ask whether average mortality rates vary substantially among censuses, and thus, whether a census effect is required.
-We achieved this by fitting a constant, $\gamma$, that is invariant with respect to species, growth rate, or census period (i.e.
-a null model) and compared it to a constant that is scaled by a census random effect, $\gamma\delta_{t}$.
-This stage revealed that there was significant variation in mortality rates observed among censuses, and as such a random effect for census was included as a scaling factor for all subsequent models.
-
-In stage 2, we assessed the predictive performance of three alternative hazard functions (Fig.
-\ref{fig1}B).
 The simplest form (Fig.
-\ref{fig1}B, left) assumes a constant hazard.
+\ref{fig1}A, left) assumes a constant hazard.
 The second form (Fig.
-\ref{fig1}B, middle) assumes the risk of dying declines towards an asymptote of zero as growth, and thus carbon budget, increases.
-The third form (Fig.
-\ref{fig1}B, right) is the summation of the two previous models; this allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike the second functional form examined here, asymptotes at some growth-independent baseline hazard $\textgreater 0$.
-Combined, these three models capture a variety of functional responses that are both used in empirical studies of plant mortality \cite{Kobe:1995tw,Wyckoff:2000un} and are represented in current vegetation models \cite{Moorcroft:2001ws,McDowell:2011dr}, but which have not previously been systematically compared.
-In doing this, we also investigate which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.
-
-In the remaining stages, we take the most predictive hazard function and growth rate measure from stage 2 and examine how three species traits considered major explanatory variables of inter-specific variation in tropical rainforests \cite{king:2006va,Poorter:2006tq, Chave:2009iy}: wood density, light demand and maximum dbh, affected model predictive performance.
-Specifically, we examined each trait individually on all three $\alpha$, $\beta$ and $\gamma$ (stage 3), as well as 2-trait (stage 4) and 3-trait (stage 5) combinations on these three parameters.
-Lastly, we fit a model that included a species random effect on each parameter (stage 6).
-This allowed us to 1) examine the maximum predictive capacity when accounting for species differences; and 2) provides a relative indication on how much of the total species variation is captured by these three traits.
-Because these models were computationally expensive to run (with some models requiring over 400 computing hours), we were unable to examine interactions among traits or determine the most predictive combination of traits effects and parameters in a cross validated framework.
-
-\section*{Results}
-
-\subsection*{Mortality over time}
-Examining the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates declining over time.
-The highest proportion of tree death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%).
-Consequently, when we allowed hazard rates to be scaled by census (i.e.
-adding the term $\delta_t$) we observed a small increase in predictive skill (Fig.
-\ref{fig2}A).
-Table S1 provides individual census effects.
-
-\subsection*{Hazard functions}
-Comparing the three hazard functions summarised in Fig.
-\ref{fig1}B, we found that the third function --- the function containing both growth-dependent and growth-independent hazard terms, i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the baseline and growth-dependent only hazard functions (Fig.
+\ref{fig1}A, middle) assumes the risk of dying declines towards an asymptote of zero as growth, and thus carbon budget, increases.
+The third "additive" model (Fig.
+\ref{fig1}BA right) is the sum of the previous two.
+In doing this, we also investigated which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.
+Comparing the three hazard functions, the additive function significantly outperformed both the baseline and growth-dependent only hazard functions (Fig.
 \ref{fig2}).
 Moreover, we found that predictive skill was higher when using stem diameter growth relative to stem-area growth (Fig.
-\ref{fig2}A).
-Table S1 gives a summary of mean parameter estimates.
+\ref{fig2}).
+
+\subsection*{Stages 3-5: Species traits}
+
+In the following 3 stages stages, we compare the most predictive hazard function and growth rate measure from stage 2 (additive model using dbh growth rate) and ask how addition of 1, 2 and 3 species-level traits improves model performance.
+Specifically, we introduce effects of the traits maximum dbh ($\psi$), light demand ($\upsilon$) and wood density ($\rho$) on the parameters $\alpha$, $\beta$ and $\gamma$ in Eq.
+\ref{eq:model}, at first one at a time (stage 3), then in 2-trait (stage 4) and 3-trait (stage 5) combinations.
+In general we found that including trait effects on these parameters decreased logarithmic loss, and thus, increased predictive accuracy (Fig.~\ref{fig2}).
+Moreover, we found that incorporating the effect of all three traits on these parameters resulted in higher predictive capacity relative to models that only incorporated one or two of the traits.
+
+\subsection*{Stages 6: Species random effects}
+
+Lastly, we fit a model that included a species random effect on each parameter.
+This allowed us to 1) examine the maximum predictive capacity when accounting for species differences; and 2) provides a relative indication on how much of the total species variation is captured by the three traits.
+We found that the increase in predictive accuracy afforded by traits (stage 5) was small relative to including species random effects on each parameter, which account for unexplained species differences, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig.
+\ref{fig2} Stage 6).
+
+\subsection*{Parameter importance and trait effects}
+
+In order to examine how species differed in hazards functions and to understand the effects of species traits, we used the entire dataset to fit final model containing all three species traits and a species random effect (which captures inter-specific variation that unexplained by traits) on each of $\alpha$, $\beta$ and $\gamma$.
+We found that each trait influenced tree mortality in different ways (Fig.
+\ref{fig3}-\ref{fig4}).
+Specifically, we found that light demanding species were more susceptible to low growth rates (i.e.
+higher $\alpha$), had higher growth-independent hazards (i.e.
+higher $\gamma$) and exhibited shallower declines in growth-dependent hazards with increasing growth rates (i.e.
+lower $\beta$).
+By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards.
+Species' maximum dbh only had significant effects on $\gamma$, whereby species that can achieve larger dbh exhibited lower growth-independent hazards.
 
@@
 \end{figure*}
-
-\subsection*{Species traits}
-Using the hazard function form with highest predictive skill (i.e.
-$(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$; Fig.~\ref{fig1}b right), we examined how effects of a species' maximum dbh ($\psi$), light demand ($\upsilon$) and wood density ($\rho$) placed on $\alpha$, $\beta$ and $\gamma$ affected model predictive accuracy.
-In general we found that including trait effects on these parameters decreased logarithmic loss, and thus, increased predictive accuracy (Fig.~\ref{fig2}).
-Moreover, we found that incorporating the effect of all three traits on these parameters resulted in higher predictive capacity relative to models that only incorporated one or two of the traits.
-However, this increase in predictive accuracy was small relative to including species random effects on each parameter, which account for unexplained species differences, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig.
-\ref{fig2} Stage 6).
-See Tables S1 for trait coefficients and Table S2 or Figs.
-S3-S5 for species-level parameter estimates.
 
@@
 
-In order to examine how species traits -- maximum dbh, light demand and wood density -- affect tree mortality rates, we fitted a final model containing all three species traits and a species random effect (which captures inter-specific variation that unexplained by traits) on each of $\alpha$, $\beta$ and $\gamma$.
-We found that each trait influenced tree mortality in different ways (Fig.
-\ref{fig3}-\ref{fig4}).
-Specifically, we found that light demanding species were more susceptible to low growth rates (i.e.
-higher $\alpha$), had higher growth-independent hazards (i.e.
-higher $\gamma$) and exhibited shallower declines in growth-dependent hazards with increasing growth rates (i.e.
-lower $\beta$).
-By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards.
-Species' maximum dbh only had significant effects on $\gamma$, whereby species that can achieve larger dbh exhibited lower growth-independent hazards.
-Table S1 for trait effects on each parameter.
-\subsection*{Parameter relative importance}
-
-Using the final model, we then examine: 1) its ability distinguish between individuals that died and individuals that survived via AUROC; and 2) the amount of inter-specific variation the model explains (i.e.
+Using the final model, we also examine: 1) its ability distinguish between individuals that died and individuals that survived via AUROC; and 2) the amount of inter-specific variation the model explains (i.e.
 $r^2$) relative to the proportions of individual deaths observed per species between 1995 and 2010 (Table.~\ref{table:1}).
@@
 Moreover, we found that the full model also explained \Sexpr{table1[[1,2]]*100}\% of the total variation observed among species.
+
+We then examined the relative contribution species traits, random effects and growth-dependent and independent hazards (Table.~\ref{table:1}) had on overall hazard rates by individually removing their effects and examining changes in both $r^2$ and AUROC.
+First, we found that the removal of census random effect $\delta_t$ had little impact on either $r^2$ or AUROC.
+The removal of either wood density or light demand resulted in similar magnitude effects on both $r^2$ and AUROC.
+By contrast, the removal of maximum dbh had a slightly smaller effect on both metrics.
+We detected a significant drop in both $r^2$ and AUROC when species random effects (i.e.
+the parameters that capture variation that is otherwise unexplained by traits) were removed, further highlighting that maximum dbh, light demand and wood density explain only a small fraction of the overall inter-specific variation observed among species.
+As expected, the removal of both species traits and species random effects resulted in a model that was unable to capture any inter-specific variation observed among species, and thus had low discriminatory ability.
+Lastly, we found the removal of the growth-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the growth-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation.
+However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the growth-dependent hazard resulted in substantially lower predictive ability.
 
@@
 
-We then examined the relative contribution species traits, random effects and growth-dependent and independent hazards (Table.~\ref{table:1}) had on overall hazard rates by individually removing their effects and examining changes in both $r^2$ and AUROC.
-First, we found that the removal of census random effect $\delta_t$ had little impact on either $r^2$ or AUROC.
-The removal of either wood density or light demand resulted in similar magnitude effects on both $r^2$ and AUROC.
-By contrast, the removal of maximum dbh had a slightly smaller effect on both metrics.
-We detected a significant drop in both $r^2$ and AUROC when species random effects (i.e.
-the parameters that capture variation that is otherwise unexplained by traits) were removed, further highlighting that maximum dbh, light demand and wood density explain only a small fraction of the overall inter-specific variation observed among species.
-As expected, the removal of both species traits and species random effects resulted in a model that was unable to capture any inter-specific variation observed among species, and thus had low discriminatory ability.
-Lastly, we found the removal of the growth-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the growth-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation.
-However, when examining the removal of these hazards in terms of the model's ability to correctly distinguish survivors from those that died (AUROC), we found the opposite -- that the removal of the growth-dependent hazard resulted in substantially lower predictive ability.
+\section*{Discussion}
 
-\section*{Discussion}
 Our Bayesian framework coupled with cross validation revealed that the most predictive hazard model of tropical tree death was one which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) included the effects of species' maximum dbh, wood density and light demand; and 4) incorporated temporal variability.
 Moreover, we found that rates of tropical tree mortality varied substantially between species and that the a species' maximum dbh, wood density and light demand explained only a fraction of the overall inter-specific variation.
+
+There are several distinct advantages of modelling tree mortality as a function of two additive, and thus independent hazards: 1) The model explicitly accounts for the independent effects of the two fundamental types of hazards faced by plants; 2) it allows for the examination of the relative contributions each hazard provides to the overall hazard rate; 3) it allows for examination of how various factors affect each hazard term, and because of this; 4) allows ecological theories of how factors such as species traits affect plant mortality to be empirically tested using commonly collected data.
 
@@
 Our analyses support previous research that advocated that wood density, species maximum size and light demand important explanatory variables of inter-specific variation in tropical rainforests \cite{king:2006va,Poorter:2006tq, Chave:2009iy}.
-Specificially, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates.
-By contrast, both high dbh or wood density species were found to be more tolerant of growth-independent hazards, corrobrating evidence reported elsewhere \cite{Chave:2009iy, Kraft:2010kq}.
+Specifically, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates.
+By contrast, both high dbh or wood density species were found to be more tolerant of growth-independent hazards, corroborating evidence reported elsewhere \cite{Chave:2009iy, Kraft:2010kq}.
 However, our analyses has also supports other work \cite{vanGelder:2006exa, Kunstler:2016km} which highlights that higher wood density species are also more tolerant of low growth rates, and thus, better able to tolerate shade and competiton.
@@
 In total \Sexpr{n_obs(BCI_training_full)} observations were used in this study comprising \Sexpr{n_inds(BCI_training_full)} individual trees and \Sexpr{n_spp(BCI_training_full)} species.
-Because of computational costs, the models fit in this study do not include individual random effects, as this would require estimation of an additional \Sexpr{n_inds(BCI_training_full)} parameters.
+Because of computational costs, the models fit in this study do not include individual random effects.
 Instead, our models assume that repeat measurements of an individual are independent of one another.
@@
 The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m).
-For each subplot, we calculated the number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap).
-As light may penetrate into a subplot from the edge of a subplot, we rescaled this index to account for values in the eight immediate neighbouring subplots.
-Specifically, we used a weighted sum approach whereby the central subplot is assigned a weight of 8 and the eight neighbouring subplots are assigned a weight of 1.
-This meant that the contribution of the central plot was equivalent to the combined effect of all eight neighbouring plots.
-These weighted values were then summed and rescaled between 0 and 1 by dividing by the maximum value estimated across all subplots.
-The light demand for each species was estimated as the mean light index encountered by new saplings appearing in the census (Fig.
+The light demand for each species was estimated as the mean light index encountered by new saplings appearing in the census, as described in SI Appendix S3 (see also Fig.
 S2).
@@
 A species' maximum dbh was determined from directly from the data, by taking the maximum dbh within the 95th percentile observed for each species.
-
-Weak correlation (\textless 0.36) was found between species' wood density, light demand and maximum dbh (See Fig. S6)
 
@@
 In our dataset, \Sexpr{prop_neg_growth(BCI_training_full)}\% of estimated growth rates were negative.
-To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see Supplementary Material S1 for details; Fig.
+To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see  SI Appendix S2 for details; Fig.
 S1).
@@
 We also centered growth rate $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.
-Weak priors on all hyper-parameters were set (see Supplementary Material S2 for details).
-Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{rstan2017} and employing some numerical optimisations (see Supplementary Material S3-S4 for details).
+Weak priors on all hyper-parameters were set (see  SI Appendix S4 for details).
+Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{rstan2017} and employing some numerical optimisations (see  SI Appendix S5-S6 for details).
 We executed three independent chains and in all cases modelled parameters converged within 2000 iterations.
@@
 After discarding the first 1000 iterations as ‘burn in’, a further 1000 iterations were taken from the joint posterior.
-Species parameter estimates from the final model are shown in Figs.
-S3\textendash S5.
+All parameter estimates from the final model are provided in Table S1\textendash S2 andFigs.
+S3\textendash S5 in  SI Appendix.
 
@@
 \showacknow % Display the acknowledgements section
-\vspace{-5em}
 
-\newpage
 % Bibliography
+\pnasbreak
+
 \bibliography{refs.bib}

### 2018-08-08 ms/supplementary.Rnw
Some of the changes suggested by @dfalster
Moderate restructue

@@
-\documentclass[9pt,twocolumn,twoside]{pnas-new}
+\documentclass[9pt,twoside]{pnas-new}
 % Use the lineno option to display guide line numbers if required.
-% Note that the use of elements such as single-column equations
-% may affect the guide line number alignment.
 
-\templatetype{pnasresearcharticle}
+\templatetype{pnassupportinginfo}
 
-\title{
-{\normalfont \fontsize{12pt}{16pt}
-{\color{black50} Supporting information for:}\\} \vskip6pt
-\fontsize{16pt}{20pt} Unifying intra- and inter-specific variation in tropical tree mortality
-}
+% Use the lineno option to display guide line numbers if required.
 
-\author[a,b,1]{James S Camac}
-\author[c]{Richard Condit}
-\author[d]{Richard G FitzJohn}
-\author[e]{Lachlan McCalman}
-\author[e]{Daniel Steinberg}
-\author[a]{Mark Westoby}
-\author[f]{S Joseph Wright}
-\author[a,g]{Daniel S Falster}
+\title{Unifying intra- and inter-specific variation in tropical tree mortality}
+\author{J.S. Camac, R. Condit, R.G. FitzJohn, L. McCalman, D. Steinberg, M. Westoby, S.J. Wright, D.S. Falster}
+\correspondingauthor{Corresponding James S Camac.\\E-mail: james.camac@gmail.com}
 
-\affil[a]{Department of Biological Sciences, Macquarie University, Sydney, New South Wales, Australia}
-\affil[b]{Current Address: Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Melbourne, Victoria, Australia}
-\affil[c]{Field Museum of Natural History and Morton Arboretum, Chicago, Illinois, USA}
-\affil[d]{Imperial College, London, United Kingdom}
-\affil[e]{Data61, CSIRO, Australia}
-\affil[f]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
-\affil[g]{Current Address: Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, Sydney, New South Wales, Australia}
-
-% Please give the surname of the lead author for the running footer
-\leadauthor{Camac}
-
-\dates{This supplement was compiled on \today}
-\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
-
-\hypersetup{
-     linkcolor=pnasbluetext
-}
-
-\RequirePackage[colorlinks=true, allcolors=blue]{hyperref}
+\usepackage{longtable}
 
@@
 
+%% Comment/remove this line before generating final copy for submission
+%%\instructionspage
+
 \maketitle
 
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
-\renewcommand{\theequation}{S\arabic{equation}}
-\renewcommand{\thesection}{S\arabic{section}}
-
-% \setcounter{secnumdepth}{0}
+%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.
+\SItext
 
@@
 
-\section{Estimating True Growth}
+\section*{An overview of methods for estimating mortality}
+
+While a variety of statistical methods have been applied to model tree mortality, these vary in two key ways.
+First, some methods model binary survival outcomes directly, while others seek to estimate the underlying (time-dependent) mortality rate experienced by a plant.
+Estimating hazard rates has the advantage that the model naturally incorporates census interval, and so can be used to make predictions across different time intervals.
+
+Second, models differ in the way the hazard or survival curve responds to covariates.
+Some studies estimate average mortality rates for populations of each species, thereby quantifying differences between species but ignoring variation among individuals within each species.
+Most other studies use a model structure which in one way or another allows mortality or survival to decline to zero.
+
+Other studies, model hazard rates using either a proportional hazards structure, such as a Cox proportional-hazard model or complementary-log-log link binomial regression \cite{Kleinbaum:2012tn}, or otherwise directly as an negative exponential function of growth rate.
+In all these cases, large growth rates cause predicted mortality to decline to zero, such that fast growing individuals never die (Fig 1b).
+The third approach models survival outcomes rather mortality rates per se, using a logistic function (either via a logistic regression, or GLM with logit link).
+
+Two difficulties arise when studying tree mortality in tropical rainforests.
+The first is that large population sizes and long periods of observation are required to make accurate inferences into how various factors affect tree mortality rates \cite{Poorter:2008iu}.
+This requirement arises in part because the observed outcome -- alive vs.
+dead -- is one step removed from the variable we ideally want to measure: $\lambda_i(t)$, the mortality (or `hazard`) rate for individual $i$ at time $t$.
+In classic survival analysis \cite{Kleinbaum:2012tn, Zens:2003jv}, the probability an individual dies between times $t_1$ and $t_2$ is a function of both $\lambda_i(t)$, and the amount of time lapsed:
+\begin{equation} \label{eq:mortality_model}
+p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
+\end{equation}
+Here $\int_{t_1}^{t_2} \lambda_i(t) {\rm d}t$ is the cumulative hazard an individual experiences from  $t_1$ and $t_2$\cite{Kleinbaum:2012tn}.
+The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then a realisation of this cumulative hazard on a probability scale:
+\begin{equation}
+S_{i,t_1\rightarrow t_2} \sim \textrm{Bernoulli}(p_{i,t_1\rightarrow t_2}).
+\end{equation}
+Second, as trees are long-lived and we are estimating the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes collected over long durations are required.
+Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.
+
+outcomes are modelled using a logistic function (either via a logistic regression, or
+
+An individual's total hazard at a given time point is therefore the sum of these two independent hazards.
+However, to date, most statistical models of tree mortality do not explicitly estimate the independent effects of these two fundamental hazards faced by plants.
+Instead, plant mortality is typically modelled using: 1) logistic regression (i.e.
+with logit link), which does not estimate a rate, but rather the log odds of death at a given time point; or 2) via a proportional hazards approach such as a Cox proportional-hazard model or complementary-log-log link binomial regression, both of which assume hazards are multiplicative, and therefore, not independent of one another \cite{Kleinbaum:2012tn}.
+
+While some methods (like logistic regression) model survival outcomes directly, it preferable to estimate a hazard rate, as these can be rescaled across different time intervals \cite{Kleinbaum:2012tn, Zens:2003jv}.
+
+Second, as trees are long-lived and we are estimating the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes collected over long durations are required.
+Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.
+
+Previous approaches for estimating $\lambda_i(t)$ have applied either an average rate per species, or a multiplicative model implying an decline to zero as growth rate 
+Various
+
+Second, as trees are long-lived and we are estimating the shape of a continuous hazard function ($\lambda_i(t)$) from binary data, large sample sizes collected over long durations are required.
+Detailed studies of tree mortality have thus only recently become possible, with the accumulation of growth and survival data from repeat surveys spanning several decades in plots containing thousands of individuals \cite{Anderson-Teixeira-2015a}.
+
+\section*{Estimating True Growth}
 
@@
 
-\section{Parameter distributions}
+\section*{Estimating species light demand}
+
+A species' light demand was calculated using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995.
+The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m).
+The light demand for each species was estimated as the mean light index encountered by new saplings appearing in the census, as described in SI Appendix (see also Fig.
+\ref{figS2}).
+
+For each subplot, we calculated the number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap).
+As light may penetrate into a subplot from the edge of a subplot, we rescaled this index to account for values in the eight immediate neighbouring subplots.
+Specifically, we used a weighted sum approach whereby the central subplot is assigned a weight of 8 and the eight neighbouring subplots are assigned a weight of 1.
+This meant that the contribution of the central plot was equivalent to the combined effect of all eight neighbouring plots.
+These weighted values were then summed and rescaled between 0 and 1 by dividing by the maximum value estimated across all subplots.
+
+\section*{Parameter distributions}
 
@@
 
-\section{Numerical optimisation}
+\section*{Numerical optimisation}
 
@@
 
-\section{Example Stan code}
+\section*{Example Stan code}
 
@@
 
-\section{References}
 \bibliography{refs.bib}
@@
 tableS1(final_model)
+@
+
+\clearpage
+
+<<spp-table,results="asis",center=TRUE, echo= FALSE,>>= 
+tableS2(spp_params)
 @
@@
 
+\clearpage
+\begin{figure*}[!hb]
+\centering
+\includegraphics{figures/supp_fig6}
+\caption{Pairwise comparison between species' wood density, light demand and maximum dbh.}
+\label{figS6}
+\end{figure*}
 \end{document}

### 2018-08-08 ms/supplementary.Rnw
Update to use latest PNAS styles

--- 2018-07-23 ms/supplementary.Rnw
@@
 
-\section{Estimating True Growth}
+\section*{Estimating True Growth}
 
@@
 
-\section{Parameter distributions}
+\section*{Parameter distributions}
 

### 2018-08-09 ms/supplementary.Rnw
Tidy up SuppMatt, including past models

@@
 \author{J.S. Camac, R. Condit, R.G. FitzJohn, L. McCalman, D. Steinberg, M. Westoby, S.J. Wright, D.S. Falster}
-\correspondingauthor{Corresponding James S Camac.\\E-mail: james.camac@gmail.com}
+\correspondingauthor{Corresponding James S Camac. E-mail: james.camac@gmail.com}
 
@@
 Here we provide some further details on methods and additional results, supplementing those presented in the main text. 
-Section ``S1-Estimating True Growth'' describes a probabilistic methods used to estimate individual growth from field measurements, taking into account measurement error and the distribution of growth rate across the community.
-Section ``S2-Parameter distributions'' describes the distributions applied to species-level random effects and priors put over parameters in the fitted model.
-Section ``S3-Numerical optimisation'' outlines some numerical concerns in estimating logloss, while Section ``S4-Example Stan code'' gives an example of the code fitted to the data.
+The section ``Review of methods for estimating mortality'' 
+``Estimating True Growth'' describes a probabilistic methods used to estimate individual growth from field measurements, taking into account measurement error and the distribution of growth rate across the community.
+The section ``Parameter distributions'' describes the distributions applied to species-level random effects and priors put over parameters in the fitted model. 
+The section ``Numerical optimisation'' outlines some numerical concerns in estimating logloss, while 
+the section ``Example Stan code'' gives an example of the code fitted to the data.
 Code and data reproducing the results in this paper are available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.
+
+\section*{Review of methods for estimating mortality}
+
+While a variety of statistical methods have been applied to model tree mortality, these vary in two key ways.
+
+First, is whether the time interval between census intervals is explicitly included in the model.
+Some methods model binary survival outcomes directly, while others seek to estimate the underlying mortality rate experienced by a population per unit time.
+Examples of the first group include logistic regression \cite[e.g.][]{monserud_modeling_1999, russo_interspecific_2008, vanMantgem:2009bi,dietze_tree_2011,}, generalized linear models with logit link \cite[e.g.][]{Visser_functional_2016}, and exponential or weibull models that don't explicitly include time \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un}.
+These methods are suitable for detecting significant correlations with predictors, or predicting survival over a fixed time interval.
+However, the second approach -- estimating hazard rates -- is generally preferable when formulating predictive survival models, as these models naturally incorporate census interval \cite{Kleinbaum:2012tn}.
+Models formulated around a hazard rate can therefore make predictions across different time intervals.
+Moreover, the hazard rate connects directly with the differential equations used to build process-based models of vegetation \cite[e.g.][]{Moorcroft:2001ws}.
+While a number of studies have directly estimated hazard rates, studies differ in the structure of the estimated hazard rate.
+Many studies estimate average hazard rates for populations of each species \cite[e.g.][]{Condit:2006, king:2006va, Kraft:2010kq, Wright:2010fl}.
+These studies are effective at quantifying differences between species but unable to explain variation among individuals within species.
+Others, seek to model variation in the hazard rate via a combination covariates \cite[e.g.][]{Ruger:2011cv} (see also main text).
+
+The other way survival models may differ is in how the predicted survival curve responds to covariates.
+Both the hazard rate and survival models predict a relationship between covariates and the probability of death across a specified time period.
+In the main text, we argue that biologically, it makes sense for this probability to asymptote above zero.
+Yet, aside form the additive hazards model we present, the other methods that we have encountered will all drive probability of death to zero when growth rate is high (assuming growth rate is included as a covariate).
+In other words, the models behave like the centre panel in Fig 1b.
+
+\subsection*{Logistic regression and Generalized linear models}
+
+Studies using logistic regression, or Generalized linear models with a logit link , all seek to predict he probability of death over a specified time interval via a model of the form:
+\begin{equation} \label{eq:logistic_model}
+p_{i,t_1\rightarrow t_2} = \frac {1}{1+e^{ (\beta _{0} + \beta _{1}x_1 + \ldots )}},
+\end{equation}
+where $\beta_{0}, \beta _{1}, \dots$ are coefficients and $x_{1}, x_{2}, \dots$ are covariates.
+As increases in growth tend to decrease mortality risk, the coefficient for growth rate when used as a predictor will be positive.
+High growth rates will cause the exponential term to become large, and $p_{i,t_1\rightarrow t_2} \rightarrow 0$.
+
+\subsection*{Proportional-hazards models}
+
+Cox proportional-hazard models, or complementary-log-log link binomial regression \cite{Kleinbaum:2012tn}, both use a formulation for the hazard rate $\lambda_i(t)$ that takes the form 
+\begin{equation} \label{eq:Cox_model}
+\lambda_i(t)= e^{\beta _{0} + \beta _{1} x_1 + \ldots}.
+\end{equation}
+From eq 1 in the main text, the probability an individual dies is then
+\begin{equation} \label{eq:mortality_model}
+p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right).
+\end{equation}
+As increases in growth rate tend to decrease mortality risk, the coefficient for growth rate when used as a covariate in eq.
+\ref{eq:Cox_model} will be negative.
+High growth rates will therefore cause $\lambda_i(t) \rightarrow 0$ and thus $p_{i,t_1\rightarrow t_2} \rightarrow 0$.
+
+Another different but related approach has been to use a logistic function for $\lambda_i(t)$ \cite[e.g.][]{Ruger:2011cv}.
+This results in a similar outcome to the proportional-hazards models: high growth rates will cause $\lambda_i(t) \rightarrow 0$ and thus $p_{i,t_1\rightarrow t_2} \rightarrow 0$.
+
+\subsection*{Other approaches}
+
+Several papers model survival directly as an exponential function of form:
+\begin{equation} \label{eq:exp_model}
+p_{i,t_1\rightarrow t_2} = e^{\beta _{0} + \beta _{1} x_1 + \ldots},
+\end{equation}
+where $f$ is some function of covariates \cite[e.g.][]{Kobe:1995tw,Pacala:1996vm, Wyckoff:2000un, Hawkes:2000ib, Vieilledent:2010fv}.
+As increases in growth rate tend to decrease mortality risk, the coefficient for growth rate when used as a covariate in eq.
+\ref{eq:exp_model} will be negative.
+High growth rates then drive $p_{i,t_1\rightarrow t_2} \rightarrow 0$..
 
@@
 \ref{figS1} illustrates observed and estimated diameters, and growth rate estimated from raw data including measurement error and from the probabilistic model.
+
+\section*{Estimating species light demand}
+
+A species' light demand was calculated using annual canopy census data collected during 1985\textendash1990 and 1990\textendash1995.
+The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m).
+The light demand for each species was estimated as the mean light index encountered by new saplings appearing in the census, as described in SI Appendix (see also Fig.
+\ref{figS2}).
+
+For each subplot, we calculated the number of strata > 2 m containing vegetation; and then transformed this to a light index ranging from 0 (dense shade) to 1 (gap).
+As light may penetrate into a subplot from the edge of a subplot, we rescaled this index to account for values in the eight immediate neighbouring subplots.
+Specifically, we used a weighted sum approach whereby the central subplot is assigned a weight of 8 and the eight neighbouring subplots are assigned a weight of 1.
+This meant that the contribution of the central plot was equivalent to the combined effect of all eight neighbouring plots.
+These weighted values were then summed and rescaled between 0 and 1 by dividing by the maximum value estimated across all subplots.
 

### 2018-08-10 ms/manuscript.Rnw
Incorporate Mark's and Daniel's changes

@@
 
-\title{Unify intra- and inter-specific drivers of tropical tree mortality via additive hazards approach}
+\title{Partitioning growth-dependent from growth-independent influences on tropical tree mortality}
 
@@
 % 120 words
-\significancestatement{We build a mortality model that not only integrates intra and inter-specific variation but also, simultaneously estimates the two fundamental hazards faced by plants -- Death due to low growth rates and death independent of growth -- and how a species' light demand, maximum dbh and wood density affect each.
-We parameterize this model using over 400,000 observed survival records collected over a 15-year period from more than 180,000 individuals across 203 species.
-We show that growth-dependent hazards are a significant contributor to tree mortality at Barro Colorado Island.
-Moreover, we show that species traits -- wood density, maximum dbh and light demand -- all affect tree mortality rates but ultimately only account for a small fraction of the total variability among species.}
+\significancestatement{We present a model that partitions rates of tropical tree mortality into growth-dependent and growth-independent hazards.
+This creates the opportunity to examine the relative contributions of within-species and across-species variation on tropical tree mortality rates, but also, how species traits affect each hazard.
+We parameterize this model using \textgreater 400,000 observed survival records collected over a 15-year period at Barro Colorado Island from more than 180,000 individuals across 203 species.
+We show that marginal carbon budgets are a major contributor to tree death on Barro Colorado Island.
+Moreover, we found that while species' light demand, maximum dbh and wood density affected tree mortality in different ways, they explained only a small fraction of the total variability observed among species.}
 
@@
 \begin{abstract}
-Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities.
-While past research has identified a variety of factors influencing tree mortality at both individual and species scales, these distinct drivers are yet to be integrated within a unified predictive framework.
-In this study, we use a cross-validated Bayesian framework coupled with classic survival analysis techniques to derive hazard functions for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
-Specifically, we develop an empricial mortality model that not only integrates individual, species, and temporal effects, but is also the first to simultaneously estimate the independent contributions of growth-related hazards and growth-independent hazards on the overall mortality rates.
-We show that functions that separate mortality rates into growth-dependent and growth-independent hazards, use stem diameter growth rather than basal-area growth, and incorporate a species' wood density, maximum dbh and light demand outperform alternative formulations.
-Moreover, we show that while these three traits do affect both growth-dependent and growth-independent hazards, they account for only a small fraction of the total variability observed among species.
-Combined, this study provides a statistical framework for predicting individual-level mortality in highly diverse tropical forests.
-It also highlights how little we know about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
+Tree death drives population dynamics, nutrient cycling, and evolution within plant communities.
+Mortality variation across species is thought to be influenced by different factors relative to variation within species.
+The unified model provided here separates mortality rates into growth-dependent and growth-independent hazards.
+This model creates the opportunity to simultaneously estimate the descriptors of variation both across and within species.
+Moreover, it provides the ability to examine how species traits affect growth-dependent and growth-independent hazards.
+We derive this unified mortality model  using cross-validated Bayesian methods coupled with mortality data collected over three census intervals for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
+We found that growth independent mortality tended to be higher in species with lower wood density, higher light requirements and smaller maximum dbh.
+Mortality due to marginal carbon budget as measured by near-zero growth rate tended to be higher in species with lower wood density and higher light demand.
+The total mortality variation attributable to differences among species was large relative to variation explained by these traits, emphasizing that much remains to be understood.
+This statistical framework strengthens our capacity to parse and understand individual-level mortality in highly diverse tropical forests, and hence to predict its consequences.
 \end{abstract}
@@
 
-Tree death is a key process driving population dynamics, nutrient cycling, and evolution within plant communities.
-While the risk of death varies among forests and from year-to-year \cite{vanMantgem:2009bi}, trees growing within the same forest also vary remarkably in realised risk of death\cite{Condit:2006}.
-In tropical forests, annual mortality can differ 20-60 fold among populations of coexisting species \cite{Condit:2006}.
-These differences reflect differences in the way trees from different species are constructed and function, and arise via a coordinated axis of life-history variation, in which individuals from species with higher potential-growth rates also experience higher mortality \cite{Wright:2010fl}.
-At the same time, the risk of death varies widely within species, with those individuals growing slower or sitting lower in the light spectrum at greater risk \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv}.
-While these inter- and intra-specific patterns of mortality are widely recognised, they show conflicting relationships with tree growth rate (increasing with growth rate across species, decreasing with growth rate within species), making it difficult to integrate both sources of variation into a predictive model.
+Tree death is a fundamental process affecting population dynamics, nutrient cycling, and evolution within plant communities.
+While the risk of death varies substantially among forests and from year to year \cite{vanMantgem:2009bi}, trees growing within the same forest also vary remarkably in realised risk of death\cite{Condit:2006}.
+In tropical forests, annual mortality can differ 20-60 fold among coexisting species \cite{Condit:2006}.
+This presumably reflects differences in the way trees from different species are constructed and function.
+Notably there exists a coordinated axis of life-history variation whereby individuals from species with higher potential growth rates also experience higher mortality \cite{Wright:2010fl}.
+At the same time the risk of death varies widely within species, with individuals growing slower or sitting lower in the light gradient at higher risk \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv}.
+In other words mortality increases with growth rate across species, but decreases with growth rate within species.
+Hence the importance of separating hazards into those connected and unconnected with growth rate, and the need for integrating hazard estimation across and within species.
 
-In this article we outline a new approach for predicting mortality among individuals within a forest, incorporating both inter and intra-specific variation in risk.
-Our approach builds on the notion that the total risk of death at any time is the additive effect of at least two, largely independent risks \cite{Keane:2001db}.
-Although plants die because of many causes, these broadly fall into two main categories \cite{Keane:2001db, McDowell:2011dr}.
-The first are growth-independent hazards, which cause death irrespective of the tree's current growth rate, and may include windfall, fire, and certain types of pathogen or herbivore attack.
+Although plants die from many causes, our approach described here partitions them into two fundamental categories: growth-dependent hazards and growth-independent hazards \cite{Keane:2001db, McDowell:2011dr}.
+Growth-independent hazards cause death irrespective of the tree's current growth rate, and may include windfall, fire, and certain types of pathogen or herbivore attack.
 These hazards are assumed to occur at some average rate (Fig 1a, left).
-The second are growth-dependent hazards, which cause slower growing plants within each species to die at a faster rate, presumably because inferior carbon budgets make them unable to tolerate diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
-Many empirical studies suggest an exponential decline in risk with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv} (Fig 1a, centre).
-An individual's total risk of dying is then the sum of these two (Fig 1a, right).
+Growth-dependent hazards cause slower-growing plants within each species to die at a faster rate, presumably because inferior carbon budgets make them unable to tolerate diverse stresses \cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}.
+Many empirical studies suggest an exponential-like decline in risk with increased growth rate, $X_i(t)$ \cite{Kobe:1995tw, Wyckoff:2000un, Vieilledent:2010fv} (Fig 1a, centre).
+An individual's total risk of dying is then the sum of growth-independent plus growth-dependent hazards (Fig 1a, right).
 Biologically, the additive model implies that even fast-growing individuals experience some risk of death.
 
-While already applied in some vegetation models \cite{Moorcroft:2001ws}, the additive hazards model has not previously been applied in empirical studies.
-Moreover, we note that the underlying mathematical structure of statistical methods used in most previous empirical studies imply responses similar to either the left or centre panels of Fig 1a (see SI Appendix S1).
-In particular, commonly used methods -- including exponential models \cite{Kobe:1995tw, Wyckoff:2000un}, logistic regression \cite{Ruger:2011cv}, generalized linear models \cite{Visser_functional_2016}, and Cox proportional-hazard models \cite{Neumann_climate_2017} -- handle covariates in a manner that allows predicted risk to decline to zero for fast growing individuals (as in Fig 1a centre; SI Appendix S1).
-These methods may therefore fail to capture the full range of variation observed within and among species.
+The additive hazards model has not previously been applied for empirical analysis, though it has been adopted in some vegetation models \cite{Moorcroft:2001ws}.
+In most previous empirical studies the underlying mathematical structure of statistical methods used implies responses similar to either left or centre panels of Fig 1a (see SI Appendix S1).
+In particular, the commonly used exponential models \cite{Kobe:1995tw, Wyckoff:2000un}, logistic regression \cite{Ruger:2011cv}, generalized linear models \cite{Visser_functional_2016}, and Cox proportional-hazard models \cite{Neumann_climate_2017} handle covariates in a manner that allows predicted risk to decline to zero for fast growing individuals (as in Fig 1a centre; SI Appendix S1).
+This makes these methods less likely to capture the full range of variation within and among species.
 
@@
 
-In forests, mortality rates are usually estimated from census data, where the size and fate of individual trees is recorded at successive time points \cite{Anderson-Teixeira-2015a} (Fig 1b).
-The size and status of individuals at each census -- alive vs.
-dead -- can then be used to estimate $\lambda_i(t)$, a continuous, latent, mortality (or `hazard`) rate for all individuals in group $i$ at time $t$.
-Using this approach, the probability an individual dies between times $t_1$ and $t_2$ results from the cumulative hazard they are exposed to between censuses\cite{Kleinbaum:2012tn}:
+Forest mortality rates are usually estimated from census data, where the size and alive vs dead status of individual trees is recorded at successive time points \cite{Anderson-Teixeira-2015a} (Fig 1b).
+From these data one can estimate $\lambda_i(t)$, a continuous, latent mortality rate or hazard rate for each individual, $i$, at time $t$.
+The probability an individual dies between times $t_1$ and $t_2$ results from the cumulative hazard they are exposed to between censuses\cite{Kleinbaum:2012tn}:
 \begin{equation} \label{eq:mortality_model}
@@
 
-Here we propose the following hazard function, incorporating individual-, species-, and temporal effects:
+Here we propose the following hazard function, incorporating individual-, species-, and census-period effects:
 \begin{equation} \label{eq:model}
@@
 Eq.
-\ref{eq:model} allows for additive growth-dependent and growth-independent effects, includes a negative exponential effect of individual's growth rate (our measure of carbon budget), allows the overall mortality rate to vary among censuses via the random effect $\delta_{t}$, while also allowing species-level differences in the shape of $\lambda_i(t)$ via the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$.
-The species-level parameters are in turn modelled as random effects, drawn from  an overall distribution.
+\ref{eq:model} allows for additive growth-dependent and growth-independent effects, includes a negative exponential effect of the individual's growth rate (our measure of carbon budget), allows the overall mortality rate to vary among censuses via the random effect $\delta_{t}$, and also provides for species-level differences in the shape of $\lambda_i(t)$ via the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$.
+The species-level parameters are in turn modelled as a function of both species trait effects and species-level random effects, the latter of which accounts for residual species variation not attributed to traits.
 
@@
 \caption{Outline of methodology.
-A) We consider three alternative hazard functions: 1) a growth-independent baseline hazard (A, left), 2) a growth-dependent hazard, where growth rate is a proxy for carbon budget (A, middle); and 3) a function that combines both growth-dependent and independent hazards (A, right).
-The parameters of the models are biologically interpretable: $\alpha$ defines the low growth rate (i.e.
-low carbon budget) effect on the overall hazard; $\beta$ reflects the sensitivity of mortality rate to changes in growth rate; and $\gamma$ is the asymptote.
+A) We consider three alternative hazard functions: 1) a growth-independent baseline hazard (A, left), 2) a growth-dependent hazard, where growth rate is a proxy for carbon budget (A, middle); and 3) a function that combines growth-dependent with growth-independent hazards (A, right).
+The parameters of the models are biologically interpretable: $\gamma$ is the baseline or growth-independent hazard; $\alpha$ defines the low growth rate (i.e.
+very marginal carbon budget) effect on the overall hazard; $\beta$ captures the sensitivity or curvature of the response to declining growth rate.
 In the most complex model (Panel A; right), $\alpha$ and $\beta$ capture growth-dependent hazards, while $\gamma$ captures growth-independent hazards that kill a plant, irrespective of its growth rate/carbon budget (e.g.
 windfall, fire).
-For each model form, we consider two alternative predictors of growth, $X$ (basal area and stem diameter growth), as well as allowing for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
+For each model form, we consider two alternative measures of growth, $X$ (basal area and stem diameter growth), and we allow for species-level effects on the parameters $\alpha$, $\beta$ and $\gamma$.
 B) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
@@
 Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing (1 fold; orange) and training (9 folds; green) data.
-Predictive capacity was assessed by averaging the log loss's obtained from the 10 test data predictions.}
+Predictive capacity was assessed by averaging the log losses obtained from the 10 test data predictions.}
 \label{fig1}
@@
 
-\section*{Case study: Tropical forest at Barro Colorado Island}
-
-To illustrate the proposed mortality model, we assess it's ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from containing \Sexpr{n_spp(BCI_training_full)}  species, recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama.
-Mortality rates are known to vary widely among species, and also vary within species in response to light levels and growth rates \cite{Condit:2006, Wright:2010fl, Ruger:2011cv}.
-Applying the additive model, we: i) estimate growth-dependent and growth-independent effects from long-term census data, ii) estimate species-level differences in these effects, and iii) assess how well strength species-level differences is predicted from three key traits, previously shown to correlate with mortality.
-Here we consider three species traits thought to drive inter-specific variation in mortality: wood density, light demand, and maximum dbh \cite{king:2006va, Kraft:2010kq, Wright:2010fl, Ruger:2011cv}.
-Overall, these traits only show weak correlation (r$^2$ \textless 0.13) and thus provide relatively independent measures of a species ecological strategy (SI Appendix, Fig.
+To illustrate the proposed mortality model, we assess its ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from \Sexpr{n_spp(BCI_training_full)} species, recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama.
+Applying the additive model, we: i) estimate growth-dependent and growth-independent effects from long-term census data, ii) estimate species-level differences in these effects, and iii) assess how well species-level differences are predicted from three fundamental traits, previously shown to correlate with mortality: wood density, light demand, and maximum dbh \cite{king:2006va, Kraft:2010kq, Wright:2010fl, Ruger:2011cv}.
+These traits show only weak correlation with each other (r$^2$ \textless 0.13) and thus provide relatively independent measures of a species ecological strategy (SI Appendix, Fig.
 S6).
 
-Our analysis proceeds in a series of stages, in which we compare the predictive power of alternative models, differing in either structure or covariates included.
-We used 10-fold cross-validation to evaluate the skill of different models in their ability to predict outcomes in novel data (i.e.
-not used in model fitting), using a likelihood-based penalty for incorrect prediction (Fig.
+Using a likelihood-based penalty for incorrect prediction, we used 10-fold cross-validation to evaluate the skill of different models in their ability to predict outcomes in novel data (i.e.
+not used in model fitting; Fig.
 \ref{fig1}C-D; see Methods) \cite{Hooten:2015, Stone:1997tg}.
-n total \Sexpr{n_obs(BCI_training_full)} observations were used to fit these models.
-Evaluating models in this way is computationally expensive, making it impossible to run all possible model formulations.
-We therefore fit models across six iterative stages of model development, described below, gradually building up in complexity, comparing their predictive skill on new data.
-We then use the final model to assess how traits align with different effects and quantify the contributions of different effects on overall model performance.
+In total \Sexpr{n_obs(BCI_training_full)} observations were used to fit these models.
+Evaluating models in this way is computationally expensive.
+Therefore rather than running all possible model formulations, we fitted models across six levels of increasing complexity, described below, and comparing their predictive skill.
+We then use the final model to assess the effects of species traits on each parameter and quantify their relative contributions on the overall model performance.
 
@@
 \subsection*{Stage 1: Mortality over time}
-We first assess whether average mortality rates varied among the three censuses, and thus, whether a census effect is required in Eq \ref{eq:model}.
-We compared a model with a constant mortality rate, $\gamma$, to a model where the constant is scaled by a census effect, $\gamma\delta_{t}$.
-Examining the three 5-year census intervals, we found average mortality rates declining over time: with the highest proportion of death occurring from 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%).
-Consequently, we observed a small increase in predictive skill when a census effect was included (Fig.
-\ref{fig2}).
+We first assess whether average mortality rates varied among the three censuses, in other words whether a census effect is justified in Eq \ref{eq:model}.
+Comparing a model with a constant mortality rate $\gamma$ to a model where the constant is scaled by a census effect $\gamma\delta_{t}$, and examining the three 5-year census intervals, average mortality rates declined over time.
+The highest proportion of death occurred from 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%).
+Consequently, we observed a small increase in predictive skill when a census effect was included (leftmost two points under stage 1 in Fig.
+\ref{fig2}; see also Table.~\ref{table:1} for quantification of gain in predictive power under successive models).
 
@@
 The second form (Fig.
-\ref{fig1}A, middle) assumes the risk of dying declines towards an asymptote of zero as growth, and thus carbon budget, increases.
+\ref{fig1}A, middle) assumes the risk of dying declines towards an asymptote of zero as growth increases.
 The third "additive" model (Fig.
 \ref{fig1}BA right) is the sum of the previous two.
-In doing this, we also investigated which growth measure (stem diameter, stem area increment) more skilfully predicts growth-dependent mortality.
-Comparing the three hazard functions, the additive function significantly outperformed both the baseline and growth-dependent only hazard functions (Fig.
+In parallel, we investigated also which growth measure (increment of stem diameter or of stem area) predicts growth-dependent mortality better.
+Comparing the three hazard functions, the additive function significantly outperformed both the baseline (the best model in stage 1) and growth-dependent only hazard functions (right vs left of the four symbols under stage 2 in Fig.
 \ref{fig2}).
-Moreover, we found that predictive skill was higher when using stem diameter growth relative to stem-area growth (Fig.
-\ref{fig2}).
+Predictive skill was higher when using stem diameter growth compared to stem-area growth (Fig.
+\ref{fig2} stage 2, open vs closed symbols).
 
 \subsection*{Stages 3-5: Species traits}
-
-In the following 3 stages stages, we compare the most predictive hazard function and growth rate measure from stage 2 (additive model using dbh growth rate) and ask how addition of 1, 2 and 3 species-level traits improves model performance.
+In stages 3-5, we take the additive model using dbh growth rate (the most predictive hazard function and growth rate measure from stage 2) and ask how addition of 1, 2 and 3 species-level traits improves model performance.
 Specifically, we introduce effects of the traits maximum dbh ($\psi$), light demand ($\upsilon$) and wood density ($\rho$) on the parameters $\alpha$, $\beta$ and $\gamma$ in Eq.
@@
 In general we found that including trait effects on these parameters decreased logarithmic loss, and thus, increased predictive accuracy (Fig.~\ref{fig2}).
-Moreover, we found that incorporating the effect of all three traits on these parameters resulted in higher predictive capacity relative to models that only incorporated one or two of the traits.
+Allowing all three traits to have an effect on these parameters resulted in higher predictive capacity relative to models that only incorporated one or two of the traits.
 
-\subsection*{Stages 6: Species random effects}
+\subsection*{Stages 6: Species random effects and trait effects}.
 
-Lastly, we fit a model that included a species random effect on each parameter.
-This allowed us to 1) examine the maximum predictive capacity when accounting for species differences; and 2) provides a relative indication on how much of the total species variation is captured by the three traits.
-We found that the increase in predictive accuracy afforded by traits (stage 5) was small relative to including species random effects on each parameter, which account for unexplained species differences, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig.
+In order to examine how species differed in hazard functions and to understand the effects of species traits, we used the entire dataset to fit a final model containing all three species traits plus a species random effect to capture inter-specific variation not explained by traits.
+The increase in predictive accuracy afforded by traits (stage 5) was fairly small relative to including species random effects on each parameter, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig.
 \ref{fig2} Stage 6).
 
-\subsection*{Parameter importance and trait effects}
-
-In order to examine how species differed in hazards functions and to understand the effects of species traits, we used the entire dataset to fit final model containing all three species traits and a species random effect (which captures inter-specific variation that unexplained by traits) on each of $\alpha$, $\beta$ and $\gamma$.
-We found that each trait influenced tree mortality in different ways (Fig.
+Each trait influenced tree mortality in different ways (Fig.
 \ref{fig3}-\ref{fig4}).
-Specifically, we found that light demanding species were more susceptible to low growth rates (i.e.
+Light demanding species were more susceptible to low growth rates (i.e.
 higher $\alpha$), had higher growth-independent hazards (i.e.
@@
 lower $\beta$).
-By contrast, we found that wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards.
-Species' maximum dbh only had significant effects on $\gamma$, whereby species that can achieve larger dbh exhibited lower growth-independent hazards.
+Wood density had significant impacts on $\alpha$ and $\gamma$, such that species with high wood density were less susceptible to low growth rates and had lower growth-independent hazards.
+Species' maximum dbh only had significant effects on $\gamma$, whereby species capable of achieving larger dbh exhibited lower growth-independent hazards.
 
@@
 
-Using the final model, we also examine: 1) its ability distinguish between individuals that died and individuals that survived via AUROC; and 2) the amount of inter-specific variation the model explains (i.e.
-$r^2$) relative to the proportions of individual deaths observed per species between 1995 and 2010 (Table.~\ref{table:1}).
-We found that the full model had an AUCROC of \Sexpr{table1[[1,3]]}, indicating that the model had a fair ability to correctly discriminate observed deaths from survivals.
-Moreover, we found that the full model also explained \Sexpr{table1[[1,2]]*100}\% of the total variation observed among species.
+\subsection*{Explanatory power}. 
+The full model had an AUROC of \Sexpr{table1[[1,3]]}, indicating a fair ability to correctly discriminate observed deaths from survivals.
+Moreover, we found that the full model also explained \Sexpr{table1[[1,2]]*100}\% of the total variation in mortality observed among species.
 
-We then examined the relative contribution species traits, random effects and growth-dependent and independent hazards (Table.~\ref{table:1}) had on overall hazard rates by individually removing their effects and examining changes in both $r^2$ and AUROC.
-First, we found that the removal of census random effect $\delta_t$ had little impact on either $r^2$ or AUROC.
+Removal of the census random effect $\delta_t$ had little impact on either $r^2$ or AUROC.
 The removal of either wood density or light demand resulted in similar magnitude effects on both $r^2$ and AUROC.
 By contrast, the removal of maximum dbh had a slightly smaller effect on both metrics.
-We detected a significant drop in both $r^2$ and AUROC when species random effects (i.e.
-the parameters that capture variation that is otherwise unexplained by traits) were removed, further highlighting that maximum dbh, light demand and wood density explain only a small fraction of the overall inter-specific variation observed among species.
-As expected, the removal of both species traits and species random effects resulted in a model that was unable to capture any inter-specific variation observed among species, and thus had low discriminatory ability.
+Both $r^2$ and AUROC dropped substantially when species random effects were removed, further highlighting that maximum dbh, light demand and wood density explain only a small fraction of the overall inter-specific variation observed among species.
+When both species traits and species random effects were removed, the resulting a model was by definition unable to capture any variation observed among species, and thus had low discriminatory ability.
 Lastly, we found the removal of the growth-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the growth-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation.
@@
 \section*{Discussion}
+Our Bayesian framework coupled with cross validation revealed that the most predictive hazard model of tropical tree death was one which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) included the effects of species' maximum dbh, wood density and light demand; and 4) incorporated temporal variability.
+Further, rates of tropical tree mortality varied substantially between species and a species' maximum dbh, wood density and light demand explained only a fraction of the overall inter-specific variation.
 
-Our Bayesian framework coupled with cross validation revealed that the most predictive hazard model of tropical tree death was one which: 1) partitioned mortality into growth-dependent and growth-independent hazards; 2) used stem diameter growth rather than basal-area growth; 3) included the effects of species' maximum dbh, wood density and light demand; and 4) incorporated temporal variability.
-Moreover, we found that rates of tropical tree mortality varied substantially between species and that the a species' maximum dbh, wood density and light demand explained only a fraction of the overall inter-specific variation.
-
-There are several distinct advantages of modelling tree mortality as a function of two additive, and thus independent hazards: 1) The model explicitly accounts for the independent effects of the two fundamental types of hazards faced by plants; 2) it allows for the examination of the relative contributions each hazard provides to the overall hazard rate; 3) it allows for examination of how various factors affect each hazard term, and because of this; 4) allows ecological theories of how factors such as species traits affect plant mortality to be empirically tested using commonly collected data.
+There are several distinct advantages of modelling tree mortality as a function of two additive hazards: 1) The model explicitly accounts for the independent effects of the two fundamental types of hazards faced by plants; 2) it allows for the examination of the relative contributions each hazard provides to the overall hazard rate; 3) it allows for examination of how various factors affect each hazard term, and because of this; 4) allows ecological theories of how factors such as species traits affect plant mortality to be empirically tested using commonly collected data.
 
 The findings of this study provide empirical support for dynamic vegetation models that predict tree mortality as a function of both growth-dependent and independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}.
-We show that regardless of growth measure (our proxy for carbon budget), incorporating both hazards significantly improves model predictive skill.
-This is because the growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses (e.g.
-competition, parasites, herbivory).
-By contrast, the growth-independent hazard accounts for deaths caused by events that arise irrespective of an individual's carbon budget (e.g.
-windthrow, lightning strike).
+Regardless of growth measure (our proxy for carbon budget), incorporating both hazards significantly improved model predictive skill.
+This was because the growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses such as competition, parasites and herbivory.
+By contrast, the growth-independent hazard accounts for deaths caused by events that arise irrespective of an individual's carbon budget such as windthrow or lightning strike.
 
-Additionally, the partitioning of mortality rates into both growth-dependent and growth-independent effects allowed us to not only estimate their relative contribution, but also how species traits affect each.
-Like many other studies \cite{King:1994vg,king:2006va, Ruger:2011cv, Kunstler:2016km}, our analyses highlighted the importance of light competition in influencing tropical tree demographic rates.
-Specifically, we found that that the removal of the growth-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table:1}).
-This suggests deficiencies in carbon budgets are a major contributor to tree death on Barro Colorado Island.
+Partitioning mortality rates into growth-dependent versus growth-independent effects allowed us to estimate not only their relative contribution, but also how species traits affect each.
+Like many other studies \cite{King:1994vg, king:2006va, Ruger:2011cv, Kunstler:2016km}, our analyses highlighted the importance of light competition for tropical tree demographic rates.
+Specifically, removal of the growth-dependent hazard term resulted in a substantial reduction in the model's ability to correctly distinguish survivors from those that died (Table~\ref{table:1}).
+This suggests marginal carbon budgets are a major contributor to tree death on Barro Colorado Island.
 
@@
 Specifically, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates.
-By contrast, both high dbh or wood density species were found to be more tolerant of growth-independent hazards, corroborating evidence reported elsewhere \cite{Chave:2009iy, Kraft:2010kq}.
-However, our analyses has also supports other work \cite{vanGelder:2006exa, Kunstler:2016km} which highlights that higher wood density species are also more tolerant of low growth rates, and thus, better able to tolerate shade and competiton.
+By contrast, high maximum-dbh and high wood density species were found to be more tolerant of growth-independent hazards, corroborating evidence reported elsewhere \cite{Chave:2009iy, Kraft:2010kq}.
+Our analyses also support other work \cite{vanGelder:2006exa, Kunstler:2016km} showing that higher wood density species tend also to be more tolerant of low growth rates and better able to tolerate shade and competition.
 
-While the effects of species' wood density, maximum size and light demand are now varyingly being incorporated into mortality algorithms of many vegetation models \cite{Moorcroft:2001ws,Scheiter:2013ed}, our analysis indicate that such effects are likely to only capture a small proportion of the overall inter-specific variation (Fig.
+Effects of species' wood density, maximum size and light demand are now being incorporated into mortality algorithms within many vegetation models in the effort to account for among-species variation \cite{Moorcroft:2001ws,Scheiter:2013ed}.
+However, the present analysis indicates that such traits are likely to capture only a small proportion of the overall inter-specific variation (Fig.
 \ref{fig4}, Table~\ref{table:1}).
-This means that such models are likely to severely underestimate the true variability in tropical rainforest mortality rates.
-As a consequence, this underestimation is likely to manifest in biased estimates of carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.
-
-% Daniel I find these next three paragraphs as ho humm. I'm advocating for their deletion. Instead, we could just end with the above? Thoughts?
-
-%Future research should focus on resolving how climate influences both carbon-dependent and carbon-independent hazards, particularly as climate-driven mortality is increasing \cite{vanMantgem:2009bi,Anderegg:2012ep,Mcdowell:2013wf}.
-Our analyses revealed a marginal decline in mortality at BCI from 1995 to 2010.
-Attributing temporal changes to particular climatic variables is challenging, however, due to low number temporal replication (n=3).
-
-%We should also consider how the interval between censuses affects the estimation of carbon-dependent and carbon-independent hazards.
-Large census intervals may underestimate carbon-dependent mortality, as wide census intervals will not capture deaths due to rapid declines in growth, or events such as drought (although drought might also increase tree growth \cite{Condit2017}).
-Consequently, we may overestimate the relative contribution of carbon-independent hazards.
-
-%Here we showcase a new framework for modelling tropical tree mortality that unifies empirical evidence from within and between species studies.
-This framework also provides an approach for partitioning mortality rates into carbon-dependent and carbon-independent hazards.
-Our findings reveal that while wood density is an important trait affecting mortality rates, we are still only capturing a fraction of the overall species variability in mortality rates.
+This means that dynamic vegetation models incoporating these traits are likely to underestimate the true variability in tropical rainforest mortality rates, and as a consequence underestimate carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.
 
@@
 We discarded species that do not exhibit secondary growth (e.g.
-palms and ferns), contained fewer than 10 individuals or did not contain an estimate of wood density.
+palms and ferns), contained fewer than 10 individuals or where estimated wood density was not available.
 We also excluded individuals that: 1) did not survive at least two censuses (two being required to estimate growth rate); 2) were not consistently measured at 1.3 m above ground; 3) were multi-stemmed; 4) resprouted or seemingly ``returned from the dead''; or 5) were extreme outliers -- stems which grew more than 5 cm yr$^{-1}$ or shrunk more than 25\% of their initial diameter.
@@
 Instead, our models assume that repeat measurements of an individual are independent of one another.
-We believe this is a reasonable assumption given that there is approximately 5-years between censuses.
+We believe this is a reasonable assumption given there was approximately 5-years between censuses.
 
@@
 
-A species' maximum dbh was determined from directly from the data, by taking the maximum dbh within the 95th percentile observed for each species.
+A species' maximum dbh was estimated directly from the data by taking the 95th percentile observed for each species.
 
@@
 Eqs.
-\ref{eq:mortality_model}-\ref{eq:model} were fit to the data using Bayesian inference and with covariates for growth rate in previous census, species traits -- maximum dbh, light demand and wood density, as well as random effects.
+\ref{eq:mortality_model}-\ref{eq:model} were fitted to the data with covariates for growth rate in previous census, species traits -- maximum dbh, light demand and wood density, as well as random effects for both species and census-period.
 Growth rates were estimated from field measurements of diameter, which inevitably include observation error.
@@
 Dolins, \& hundreds of field workers for their contribution; and the National Science Foundation, Smithsonian Tropical Research Institute, \& MacArthur Foundation for funding the design, collection, quality control and management of long-term growth data at BCI.
-We also Fig.
+Fig.
 1a uses images by Tracey Saxby, IAN Image Library.

### 2018-08-10 ms/manuscript.Rnw
More edits

@@
 
-\subsection*{Stages 6: Species random effects and trait effects}.
+\subsection*{Stages 6: Species random effects and trait effects}
 
@@
 Species' maximum dbh only had significant effects on $\gamma$, whereby species capable of achieving larger dbh exhibited lower growth-independent hazards.
+All parameter estimates for the final model can be found in Tables S1\textendash S2.
 
@@
 Note that the intercept is the sum of $\alpha$ and $\gamma$ (i.e.
-the stochastic chance of death plus effect of low growth.}
+the stochastic chance of death plus effect of low growth.
+Species with highest asymptote was \textit{Palicourea guianensis} where \Sexpr{subset(death_proportion(x), species =="Palicourea guianensis")[["n_deaths"]]} out of \Sexpr{subset(death_proportion(x), species =="Palicourea guianensis")[["n_total"]]} individuals died between 1995 and 2010.}
 \label{fig4}
@@
 
-\subsection*{Explanatory power}. 
+\subsection*{Explanatory power} 
 The full model had an AUROC of \Sexpr{table1[[1,3]]}, indicating a fair ability to correctly discriminate observed deaths from survivals.
@@
 After discarding the first 1000 iterations as ‘burn in’, a further 1000 iterations were taken from the joint posterior.
-All parameter estimates from the final model are provided in Table S1\textendash S2 andFigs.
+All parameter estimates from the final model are provided in Table S1\textendash S2 and Figs.
 S3\textendash S5 in SI Appendix.

### 2018-08-10 ms/supplementary.Rnw
More edits

@@
 Here we provide some further details on methods and additional results, supplementing those presented in the main text. 
-The section ``Review of methods for estimating mortality'' 
+The section ``Review of methods for estimating mortality'' briefly describes the statistical approaches commonly used when estimating tree mortality and how they differ to the additive hazards model advocated in this manuscript.
 ``Estimating True Growth'' describes a probabilistic methods used to estimate individual growth from field measurements, taking into account measurement error and the distribution of growth rate across the community.
@@
 Some methods model binary survival outcomes directly, while others seek to estimate the underlying mortality rate experienced by a population per unit time.
-Examples of the first group include logistic regression \cite[e.g.][]{monserud_modeling_1999, russo_interspecific_2008, vanMantgem:2009bi,dietze_tree_2011,}, generalized linear models with logit link \cite[e.g.][]{Visser_functional_2016}, and exponential or weibull models that don't explicitly include time \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un}.
+Examples of the first group include logistic regression \cite[e.g.][]{monserud_modeling_1999, russo_interspecific_2008, vanMantgem:2009bi,dietze_tree_2011}, generalized linear models with logit link \cite[e.g.][]{Visser_functional_2016}, and exponential or weibull models that don't explicitly include time \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un}.
 These methods are suitable for detecting significant correlations with predictors, or predicting survival over a fixed time interval.
-However, the second approach -- estimating hazard rates -- is generally preferable when formulating predictive survival models, as these models naturally incorporate census interval \cite{Kleinbaum:2012tn}.
+However, the second approach -- estimating hazard rates -- is generally preferable when formulating predictive survival models, as these models naturally incorporate census interval \cite{Zens:2003jv,Kleinbaum:2012tn}.
 Models formulated around a hazard rate can therefore make predictions across different time intervals.
@@
 
-Studies using logistic regression, or Generalized linear models with a logit link , all seek to predict he probability of death over a specified time interval via a model of the form:
+Studies using logistic regression, or Generalized linear models with a logit link, all seek to predict the probability of death over a specified time interval via a model of the form:
 \begin{equation} \label{eq:logistic_model}
@@
 \includegraphics{figures/supp_fig6}
-\caption{Pairwise comparison between species' wood density, light demand and maximum dbh.}
+\caption{Pairwise comparison between species' wood density, light demand and maximum dbh. Correlations are pearson correlations}
 \label{figS6}

### 2018-08-19 ms/manuscript.Rnw
Correctly attribute figure panel
Incorporating Rick's comments

@@
 \affil[a]{Department of Biological Sciences, Macquarie University, Sydney, New South Wales, Australia}
-\affil[b]{Field Museum of Natural History and Morton Arboretum, Chicago, Illinois, USA}
+\affil[b]{Field Museum of Natural History, Chicago, IL, USA and Morton Arboretum, Lisle, IL, USA}
 \affil[c]{Data61, CSIRO, Australia}
@@
 Data deposition: Code reproducing the results in this paper is available at \href{https://github.com/traitecoevo/mortality_bci}{github.com/traitecoevo/mortality\_bci}.
-Data from BCI is available at DOIs \href{http://doi.org/10.5479/data.bci.20130603.}{10.5479/data.bci.20130603.}, \href{http://doi.org/10.5479/data.bci20140711}{10.5479/data.bci20140711}.
+Data from BCI is available at DOIs \href{http://doi.org/10.5479/data.bci.20130603}{10.5479/data.bci.20130603}, \href{http://doi.org/10.5479/data.bci20140711}{10.5479/data.bci20140711}.
 }
@@
 the stochastic chance of death plus effect of low growth.
-Species with highest asymptote was \textit{Palicourea guianensis} where \Sexpr{subset(death_proportion(x), species =="Palicourea guianensis")[["n_deaths"]]} out of \Sexpr{subset(death_proportion(x), species =="Palicourea guianensis")[["n_total"]]} individuals died between 1995 and 2010.}
+Species with highest asymptote was \textit{Palicourea guianensis} where \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_deaths"]]} out of \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_total"]]} individuals died between 1995 and 2010.}
 \label{fig4}
@@
 Fig.
-1a uses images by Tracey Saxby, IAN Image Library.
+1b uses images by Tracey Saxby, IAN Image Library.
 J.S.C., R.G.F., L.M.

### 2018-08-19 ms/supplementary.Rnw
Minor typos
Incorporating Rick's comments

@@
 
-\title{Unifying intra- and inter-specific variation in tropical tree mortality}
+\title{Partitioning growth-dependent from growth-independent influences on tropical tree mortality}
 \author{J.S. Camac, R. Condit, R.G. FitzJohn, L. McCalman, D. Steinberg, M. Westoby, S.J. Wright, D.S. Falster}
@@
 Yet, aside form the additive hazards model we present, the other methods that we have encountered will all drive probability of death to zero when growth rate is high (assuming growth rate is included as a covariate).
-In other words, the models behave like the centre panel in Fig 1b.
+In other words, the models behave like the centre panel in Fig 1a.
 

### 2018-08-22 ms/manuscript.Rnw
Incorporate typos found by Daniel

@@
 collected data.
-J.S.C \& D.S.F.
+J.S.C, D.S.F.
+and M.W.
 wrote the manuscript}
@@
 The additive hazards model has not previously been applied for empirical analysis, though it has been adopted in some vegetation models \cite{Moorcroft:2001ws}.
-In most previous empirical studies the underlying mathematical structure of statistical methods used implies responses similar to either left or centre panels of Fig 1a (see SI Appendix S1).
-In particular, the commonly used exponential models \cite{Kobe:1995tw, Wyckoff:2000un}, logistic regression \cite{Ruger:2011cv}, generalized linear models \cite{Visser_functional_2016}, and Cox proportional-hazard models \cite{Neumann_climate_2017} handle covariates in a manner that allows predicted risk to decline to zero for fast growing individuals (as in Fig 1a centre; SI Appendix S1).
+In most previous empirical studies the underlying mathematical structure of statistical methods used implies responses similar to either left or centre panels of Fig 1a (see SI Appendix).
+In particular, the commonly used exponential models \cite{Kobe:1995tw, Wyckoff:2000un}, logistic regression \cite{Ruger:2011cv}, generalized linear models \cite{Visser_functional_2016}, and Cox proportional-hazard models \cite{Neumann_climate_2017} handle covariates in a manner that allows predicted risk to decline to zero for fast growing individuals (as in Fig 1a centre; SI Appendix).
 This makes these methods less likely to capture the full range of variation within and among species.
@@
 \begin{equation} \label{eq:mortality_model}
-p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t\right),
+p_{i,t_1\rightarrow t_2} = 1 - \exp\left(- \int_{t_1}^{t_2} \lambda_i(t) \, {\rm d}t\right).
 \end{equation}
-The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then modelled as a binary draw.
+The observed survival outcome $S_{i, t_1\rightarrow t_2}$ (0 = alive, 1 = died) is then modelled as a binary draw from $p_{i,t_1\rightarrow t_2}$.
 The challenge is then to estimate an appropriate form for the hazard rate, $\lambda_i(t)$.
@@
 \begin{equation} \label{eq:model}
-\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Growth independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{s,i}(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
+\lambda_{i}(t) = \bigg(\overbrace{\gamma_{s}}^{\text{Growth independent}} + \overbrace{\alpha_{s} \, \textrm{e}^{-\beta_{s} \, X_{i}(t)}}^{\text{Growth dependent}}\bigg) \times \overbrace{\delta_{t}}^{\text{Census}}.
 \end{equation}
@@
 The species-level parameters are in turn modelled as a function of both species trait effects and species-level random effects, the latter of which accounts for residual species variation not attributed to traits.
+
+A further advantage of the proposed hazard function (eq.
+{eq:model}) is that it allows us to estimate (at population level), the relative contributions of different mortality events, even if the causes of death for each individual tree remain unknown.
 
@@
 A) Six sequential stages of model selection with each stage increasing in complexity.
-\textbf{Stage 1:} Baseline: constant ($\gamma$) vs Baseline with census effect ($\gamma\delta_t$); \textbf{Stage 2:} Best model of stage 1 compared to hazard functions that also account for individual growth rate (i.e.
+\textbf{Stage 1:} Baseline: constant ($\gamma$) vs Baseline with census effect ($\gamma\,\delta_t$); \textbf{Stage 2:} Best model of stage 1 compared to hazard functions that also account for individual growth rate (i.e.
 carbon budget).
@@
 As such, $r^2$ measures the amount of species-level variation explained by the model or sub-models.
-By contrast, AUROC examines the model's ability to distinguish individuals that died from individuals that survived.")
+By contrast, AUROC examines the model's ability to distinguish individuals that died from individuals that survived, irrespective of species.")
 @
@@
 This suggests marginal carbon budgets are a major contributor to tree death on Barro Colorado Island.
+Specifically, the mortality rate for individuals in most species increased substantially once diameter growth rates dropped below 0.15cm/yr (Fig.
+\ref{fig4}).
 
@@
 \ref{fig4}, Table~\ref{table:1}).
-This means that dynamic vegetation models incoporating these traits are likely to underestimate the true variability in tropical rainforest mortality rates, and as a consequence underestimate carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.
+This means that dynamic vegetation models incorporating these traits are likely to underestimate the true variability in tropical rainforest mortality rates, and as a consequence underestimate carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.
 
@@
 The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m).
-The light demand for each species was estimated as the mean light index encountered by new saplings appearing in the census, as described in SI Appendix S3 (see also Fig.
+The light demand for each species was estimated as the mean light index encountered by new saplings appearing in the census, as described in SI Appendix (see also Fig.
 S2).
@@
 In our dataset, \Sexpr{prop_neg_growth(BCI_training_full)}\% of estimated growth rates were negative.
-To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see  SI Appendix S2 for details; Fig.
+To ensure our mortality model was not biased by these unlikely values we first applied a probabilistic model to estimate \textit{``true growth''}, taking into account measurement error and the distribution of growth rate across the community (see SI Appendix text and Fig.
 S1).
@@
 We also centered growth rate $X_i$ at the lower 5\% quantile for both diameter increment and area growth (0.172 and 0.338, respectively), meaning $\alpha_s$ should be interpreted as the hazard rate when growth rate was very low.
-Weak priors on all hyper-parameters were set (see  SI Appendix S4 for details).
-Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{rstan2017} and employing some numerical optimisations (see  SI Appendix S5-S6 for details).
+Weak priors on all hyper-parameters were set (see  SI Appendix).
+Models were fit in R \Sexpr{paste(version$major, version$minor, sep='.')} using the package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{rstan2017} and employing some numerical optimisations (see  SI Appendix).
 We executed three independent chains and in all cases modelled parameters converged within 2000 iterations.

### 2018-08-22 ms/supplementary.Rnw
Incorporate typos found by Daniel

@@
 The canopy census recorded, in all 5 by 5 m subplots across the 50 ha plot, the presence of leaf in six height intervals (0-2,2-5,5-10,10-20,20-30, \textgreater 30 m).
-The light demand for each species was estimated as the mean light index encountered by new saplings appearing in the census, as described in SI Appendix (see also Fig.
+The light demand for each species was estimated as the mean light index encountered by new saplings appearing in the census (Fig.
 \ref{figS2}).
@@
 To evaluate eq.
-5 from the main text (log loss), we apply the following substitutions, to reduce the number of $\log$ and $\exp$ operations, and thereby maintain numerical accuracy.
-From the main text, equation 5 states that the log loss for a single data point is
+4 from the main text (log loss), we apply the following substitutions, to reduce the number of $\log$ and $\exp$ operations, and thereby maintain numerical accuracy.
+From the main text, eq.
+4 states that the log loss for a single data point is
 \begin{equation} \label{eq:logloss_i}
@@
 where $p_{i,t_2\rightarrow t_3} = 1 - \exp\left(-\Lambda\right)$ is the probability an individual dies between times $t_2$ and $t_3$ (from eq.
-1), $\Lambda$ is the cumulative hazard from $t_2\rightarrow t_3$, and $S_i$ is the observed outcome (0 = alive, 1 = died).
+1), $\Lambda = \int_{t_1}^{t_2} \lambda_i(t) {\rm d}t$ is the cumulative hazard from $t_2\rightarrow t_3$, and $S_i$ is the observed outcome (0 = alive, 1 = died).
 Note that $\textrm{Pr}(S_i =0) = 1- p_{i,t_2\rightarrow t_3} = \exp\left(-\Lambda\right)$.

### 2018-08-22 ms/manuscript.Rnw
Incorporate Joe's comments

@@
 
-\section*{An additive hazards models with species effects}
+\section*{An additive hazards model with species effects}
 
@@
 Eq.
-\ref{eq:model} allows for additive growth-dependent and growth-independent effects, includes a negative exponential effect of the individual's growth rate (our measure of carbon budget), allows the overall mortality rate to vary among censuses via the random effect $\delta_{t}$, and also provides for species-level differences in the shape of $\lambda_i(t)$ via the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$.
+\ref{eq:model} allows for additive growth-dependent and growth-independent effects, includes a negative exponential effect of the individual's growth rate ($X$; our measure of carbon budget), allows the overall mortality rate to vary among censuses via the random effect $\delta_{t}$, and also provides for species-level differences in the shape of $\lambda_i(t)$ via the parameters $\alpha_s$, $\beta_s$ and $\gamma_s$.
 The species-level parameters are in turn modelled as a function of both species trait effects and species-level random effects, the latter of which accounts for residual species variation not attributed to traits.
@@
 This includes two possible hazard functions: a growth-dependent only hazard ($(\alpha e^{-\beta X_{i}})\delta_t$) and a baseline + growth-dependent hazard with census effects ($(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$); \textbf{Stage 3:} The effect of a species' maximum dbh ($\psi$), wood density ($\rho$) and light demand ($\upsilon$) on $\alpha$, $\beta$ and $\gamma$ of most predictive model from stage 2; \textbf{Stage 4:} 2-trait combinations; \textbf{Stage 5:} three-trait combination; \textbf{Stage 6:} Maximum predictive performance by accounting for all species differences via the inclusion of species random effects, $s$ on all three parameters.
-Symbols represent the three hazard functional form: baseline (circle), growth-dependent (triangle), and hazard containing both growth-independent and growth-dependent hazards (square).}
+Symbols represent the three hazard functional form: baseline (circle), growth-dependent (triangle), and hazard containing both growth-independent and growth-dependent hazards (square).
+Colours represent models that don't account for individual growth (white); use individual basal area growth (grey); or use diameter increment growth (black).}
 \label{fig2}
@@
 In order to examine how species differed in hazard functions and to understand the effects of species traits, we used the entire dataset to fit a final model containing all three species traits plus a species random effect to capture inter-specific variation not explained by traits.
-The increase in predictive accuracy afforded by traits (stage 5) was fairly small relative to including species random effects on each parameter, signifying that substantial inter-specific variation in mortality rates remains unexplained (Fig.
-\ref{fig2} Stage 6).
+The increase in predictive accuracy afforded by traits (Fig.
+\ref{fig2}, compare best model at stage 2 with stage 5) was fairly small relative to including species random effects on each parameter (Fig.
+\ref{fig2} compare stages 5 and 6), signifying that substantial inter-specific variation in mortality rates remains unexplained.
 
@@
 A) Estimated average curves for all 203 Barro Colorado Island species used in this study.
-B,C,D) represent average ($\pm$ 95\% credible intervals) curve for a high (red) and low (blue) trait values.
-B) wood density values 0.2 vs 0.8 (g cm$^{-3}$); C) maximum dbh values 3 vs 180 cm; D) light demand values 0.5 vs 0.8.
+B,C,D) represent the average ($\pm$ 95\% credible intervals)  curve for a high (red) and low (blue) trait value.
+B) maximum dbh values 3 vs 180 cm; C) wood density values 0.2 vs 0.8 (g cm$^{-3}$); D) light demand values 0.5 vs 0.8.
 Note that the intercept is the sum of $\alpha$ and $\gamma$ (i.e.
 the stochastic chance of death plus effect of low growth.
-Species with highest asymptote was \textit{Palicourea guianensis} where \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_deaths"]]} out of \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_total"]]} individuals died between 1995 and 2010.}
+Species with highest asymptote in panel A was \textit{Palicourea guianensis} where \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_deaths"]]} out of \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_total"]]} individuals died between 1995 and 2010.}
 \label{fig4}
@@
 \subsection*{Explanatory power} 
-The full model had an AUROC of \Sexpr{table1[[1,3]]}, indicating a fair ability to correctly discriminate observed deaths from survivals.
+The full model had an Area Under the Roc Receiver Curve (AUROC) of \Sexpr{table1[[1,3]]}, indicating a fair ability to correctly discriminate observed deaths from survivals.
 Moreover, we found that the full model also explained \Sexpr{table1[[1,2]]*100}\% of the total variation in mortality observed among species.
@@
 Both $r^2$ and AUROC dropped substantially when species random effects were removed, further highlighting that maximum dbh, light demand and wood density explain only a small fraction of the overall inter-specific variation observed among species.
-When both species traits and species random effects were removed, the resulting a model was by definition unable to capture any variation observed among species, and thus had low discriminatory ability.
+When both species traits and species random effects were removed, the resulting model was by definition unable to capture any variation observed among species, and thus had low discriminatory ability.
 Lastly, we found the removal of the growth-independent hazard term ($\gamma_s$) resulted in a lower $r^2$ relative to the removal of the growth-dependent hazard ($\alpha_s e^{-\beta_s X_{i}}$), indicating that the former hazard explained more inter-specific variation.
@@
 
-Our analyses support previous research that advocated that wood density, species maximum size and light demand important explanatory variables of inter-specific variation in tropical rainforests \cite{king:2006va,Poorter:2006tq, Chave:2009iy}.
+Our analyses support previous research that advocated that wood density, species maximum size and light demand are important explanatory variables of inter-specific variation in tropical rainforests \cite{king:2006va,Poorter:2006tq, Chave:2009iy}.
 Specifically, we show that light demanding species are not only more susceptible to both low growth rates and growth-independent hazards, but also exhibited shallower declines in growth-dependent hazards with increasing growth rates.

### 2018-08-22 ms/manuscript.Rnw
Other small changes including adding a hazards curve plot to the supplementary

@@
 These traits show only weak correlation with each other (r$^2$ \textless 0.13) and thus provide relatively independent measures of a species ecological strategy (SI Appendix, Fig.
-S6).
+S7).
 
@@
 the stochastic chance of death plus effect of low growth.
-Species with highest asymptote in panel A was \textit{Palicourea guianensis} where \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_deaths"]]} out of \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_total"]]} individuals died between 1995 and 2010.}
+Species with highest asymptote in panel A was \textit{Palicourea guianensis} where \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_deaths"]]} out of \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_total"]]} individuals died between 1995 and 2010.
+See Fig S1 for equivalent hazard curves.}
 \label{fig4}
@@
 The light demand for each species was estimated as the mean light index encountered by new saplings appearing in the census, as described in SI Appendix (see also Fig.
-S2).
+S3).
 
@@
 All parameter estimates from the final model are provided in Table S1\textendash S2 and Figs.
-S3\textendash S5 in SI Appendix.
+S4\textendash S6 in SI Appendix.
 

### 2018-08-22 ms/supplementary.Rnw
Other small changes including adding a hazards curve plot to the supplementary

@@
 Fig.
-\ref{figS1} illustrates observed and estimated diameters, and growth rate estimated from raw data including measurement error and from the probabilistic model.
+\ref{figS2} illustrates observed and estimated diameters, and growth rate estimated from raw data including measurement error and from the probabilistic model.
 
@@
 The light demand for each species was estimated as the mean light index encountered by new saplings appearing in the census (Fig.
-\ref{figS2}).
+\ref{figS3}).
 
@@
 
-Models were implemented in the `Stan` framework, which uses its own specific syntax to encode models \cite{StanDevelopmentTeam:2015uz}.
+Models were implemented in the `Stan` framework, which uses its own specific syntax to encode models \cite{StanDevelopmentTeam:2017uz}.
 Code for all models implemented are available in the supplied source code.
@@
 
+\begin{figure}[h!]
+\includegraphics{figures/supp_fig1}
+\caption{Predicted changes in hazard with increasing growth rate.
+A) Estimated average curves for all 203 Barro Colorado Island species used in this study.
+B,C,D) represent the average ($\pm$ 95\% credible intervals) curve for a high (red) and low (blue) trait value.
+B) maximum dbh values 3 vs 180 cm; C) wood density values 0.2 vs 0.8 (g cm$^{-3}$); D) light demand values 0.5 vs 0.8.
+Note that the intercept is the sum of $\alpha$ and $\gamma$ (i.e.
+the stochastic chance of death plus effect of low growth.
+Species with highest asymptote in panel A was \textit{Palicourea guianensis} where \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_deaths"]]} out of \Sexpr{subset(death_proportion(BCI_training_full), species =="Palicourea guianensis")[["n_total"]]} individuals died between 1995 and 2010.}
+\label{figS1}
+\end{figure}
+
 \begin{figure*}[!hb]
 \centering
-\includegraphics[width=0.6\textheight]{figures/supp_fig1.png}
+\includegraphics[width=0.6\textheight]{figures/supp_fig2.png}
 \caption{Observed vs predicted individual growth rates and stem diameters}
-\label{figS1}
-\end{figure*}
-
-\clearpage
-\begin{figure*}[!hb]
-\centering
-\includegraphics[width=\textwidth]{figures/supp_fig2}
-\caption{Map of BCI 50 ha plot showing tree recruits (red crosses) against gap index for censuses 1985\textendash 1990 and 1990\textendash 1995.
-Colour gradient indicate the minimum amount of canopy shading above 2 m in a 5 year census.
-Dark colours indicate dense canopy shading, light colours indicate little or no canopy shading above 2 m.}
 \label{figS2}
@@
 \centering
-\includegraphics[height=0.8\textheight]{figures/supp_fig3}
-\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\alpha$) for each species. Red = Model estimated hyper-distribution species effects were sampled from.}
+\includegraphics[width=\textwidth]{figures/supp_fig3}
+\caption{Map of BCI 50 ha plot showing tree recruits (red crosses) against gap index for censuses 1985\textendash 1990 and 1990\textendash 1995.
+Colour gradient indicate the minimum amount of canopy shading above 2 m in a 5 year census.
+Dark colours indicate dense canopy shading, light colours indicate little or no canopy shading above 2 m.}
 \label{figS3}
@@
 \centering
-\includegraphics{figures/supp_fig4}
-\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\beta$) for each species. Red = Model estimated hyper-distribution species effects were sampled from.}
+\includegraphics[height=0.8\textheight]{figures/supp_fig4}
+\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\alpha$) for each species. Red = Model estimated hyper-distribution species effects were sampled from.}
 \label{figS4}
@@
 \includegraphics{figures/supp_fig5}
-\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\gamma$) for each species. Red = Model estimated hyper-distribution species effects were sampled from.}
+\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\beta$) for each species. Red = Model estimated hyper-distribution species effects were sampled from.}
 \label{figS5}
@@
 \includegraphics{figures/supp_fig6}
-\caption{Pairwise comparison between species' wood density, light demand and maximum dbh. Correlations are pearson correlations}
+\caption{Mean ($\pm$ 95\% Bayesian Credible Interval) log($\gamma$) for each species. Red = Model estimated hyper-distribution species effects were sampled from.}
 \label{figS6}
 \end{figure*}
+
+\clearpage
+\begin{figure*}[!hb]
+\centering
+\includegraphics{figures/supp_fig7}
+\caption{Pairwise comparison between species' wood density, light demand and maximum dbh. Correlations are pearson correlations}
+\label{figS7}
+\end{figure*}
 \end{document}

### 2018-08-22 R/plot_fig4.R

@@ -54,13 +54,19 @@ plot_fig4 <- function(model,
         plot_theme() +
         theme(plot.margin=unit(c(0.35,0.2,0.5,0.5),"cm"))
   
-  # ("1-yr mortality probability")
+  
+  if(isTRUE(hazard_curve)) {
+    title <- "Hazard rate"
+  } else {
+    title <- "Annual mortality probability"
+  }
+  
   cowplot::plot_grid(p1,p2,p3,p4, ncol=1, labels=c("A) Species curves",
                                                    "B) Maximum dbh",
                                                    "C) Wood density",
                                                    "D) Light demand"), 
                      label_size = 7, 
                      label_fontface ="plain") +
-    draw_label("Annual mortality probability", x= 0, y=0.5, vjust= 1.5, angle=90,size = 7) +
+    draw_label(title, x= 0, y=0.5, vjust= 1.5, angle=90,size = 7) +
     draw_label(expression("Annual dbh growth"~("cm yr"^-1)), x= 0.5, y=0.03, angle=0,size = 7)
 }

### 2018-08-22 R/plot_mu_curves.R

@@ -25,7 +25,7 @@ plot_mu_curves <- function(model,
   
   preds <- predict_mu_hazards_curves(model, growth_range, trait_name, trait_values, hazard_curve)
   
-  breaks <- c(0.0001,0.001,0.01,0.1, 1, 10, 100)
+  breaks <- c(0.0001,0.01, 1, 100)
   labels <- sapply(log10(breaks),function(i) as.expression(bquote(10^ .(i))))
   
   p1 <- ggplot2::ggplot(preds, ggplot2::aes(x = dbh_growth,y = mean, group = type, colour = factor(trait_value), fill=factor(trait_value))) + 

### 2018-08-22 R/plot_spp_curves.R

@@ -53,7 +53,7 @@ plot_spp_curves <- function(model,
     col_ramp <- ggplot2::scale_colour_manual(values = "black")
   }
   
-  breaks <- c(0.0001,0.001,0.01,0.1, 1, 10, 100)
+  breaks <- c(0.0001,0.01, 1, 100)
   labels <- sapply(log10(breaks),function(i) as.expression(bquote(10^ .(i))))
   
   if(hazard_curve ==FALSE) {

### 2018-08-22 R/table_1.R

@@ -105,8 +105,8 @@ table_1 <- function(model, data) {
       `Full minus max dbh` = full_minus_dbh,
       `Full minus species error` = full_minus_sppre,
       `Full minus species traits & species error` = full_minus_spp,
-      `Full minus carbon-dependent hazard` = full_minus_growthdep, 
-      `Full minus carbon-independent hazard` = full_minus_growthindep)
+      `Full minus growth-dependent hazard` = full_minus_growthdep, 
+      `Full minus growth-independent hazard` = full_minus_growthindep)
   
   
   # Calculate R2 relative to observed proportion of death per species

### 2018-08-22 ms/manuscript.Rnw
Added sentence just before results providing a little more context to the r2 and auroc statistics

@@
 We then use the final model to assess the effects of species traits on each parameter and quantify their relative contributions on the overall model performance.
+Relative contributions of parameters were assessed using $r^2$ and Area Under the ROC Receiver Curve (AUROC).
+Whereby, $r^2$ estimates were derived from modelled predictions vs proportion of individual deaths observed between 1995 and 2010 for each species, and thus measured the amount of species-level variation explained by the model \cite{Visser_functional_2016}.
+By contrast, AUROC examined the model's ability to distinguish individuals that died from individuals that survived, irrespective of species.
 

### 2018-08-23 ms/manuscript.Rnw
Final set of changes

@@
 The total mortality variation attributable to differences among species was large relative to variation explained by these traits, emphasizing that much remains to be understood.
-This statistical framework strengthens our capacity to parse and understand individual-level mortality in highly diverse tropical forests, and hence to predict its consequences.
+This additive hazards model strengthens our capacity to parse and understand individual-level mortality in highly diverse tropical forests, and hence to predict its consequences.
 \end{abstract}
@@
 \end{equation}
+
 Eq.
@@
 A further advantage of the proposed hazard function (eq.
-{eq:model}) is that it allows us to estimate (at population level), the relative contributions of different mortality events, even if the causes of death for each individual tree remain unknown.
+\ref{eq:model}) is that it allows us to estimate (at population level), the relative contributions of different mortality events, even if the causes of death for each individual tree remain unknown.
 
@@
 
+Basal area growth is thought to be intrinsically linked to changes in total leaf area \cite{Falster2018xx}, and thus, a better predictor of tree mortality.
+However, we found the opposite -- past diameter growth better predicted tropical tree mortality relative to past basal area growth.
+This finding may be due to diameter growth being more variable among smaller, more vulnerable, tree sizes (Fig.
+S8), and thus, better able to capture size-dependent mortality.
+We did not directly account for individual size in our additive hazards model.
+This was because of two main reasons: 1) size is highly correlated with both measures of growth (pearson correlations up to 75\%; Fig S8), and consequently is prone to parameter identifiability issues where the model is unable to discern growth from size effects; 2) the effect of size is highly non-linear \cite{Vieilledent:2010fv,Ruger:2011cv}, and thus, its inclusion and subsequent interpretation would not be straight forward.
+Further work is required in order to determine the independent effects of size and growth effects on tree mortality, but also how to partition sizes effects between growth-dependent and growth-independent hazards.
+
 Effects of species' wood density, maximum size and light demand are now being incorporated into mortality algorithms within many vegetation models in the effort to account for among-species variation \cite{Moorcroft:2001ws,Scheiter:2013ed}.
@@
 \ref{fig4}, Table~\ref{table:1}).
-This means that dynamic vegetation models incorporating these traits are likely to underestimate the true variability in tropical rainforest mortality rates, and as a consequence underestimate carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.
+This means that dynamic vegetation models incorporating these traits are likely to account for a fraction of the true variability in tropical rainforest mortality rates, and as a consequence underestimate carbon, water and nutrient dynamics of ecosystems \cite{Johnson-2016}.
 

### 2018-08-23 ms/manuscript.Rnw
minor typo
Final changes prior to submission

@@
 
-\title{Partitioning growth-dependent from growth-independent influences on tropical tree mortality}
+\title{Partitioning mortality into growth-dependent and growth-independent hazards across 203 tropical tree species}
 
@@
 \ref{fig2}, compare best model at stage 2 with stage 5) was fairly small relative to including species random effects on each parameter (Fig.
-\ref{fig2} compare stages 5 and 6), signifying that substantial inter-specific variation in mortality rates remains unexplained.
+\ref{fig2}, compare stages 5 and 6), signifying that substantial inter-specific variation in mortality rates remains unexplained.
 

### 2018-08-23 ms/supplementary.Rnw
Final set of changes
Final changes prior to submission

@@
 
-\title{Partitioning growth-dependent from growth-independent influences on tropical tree mortality}
+\title{Partitioning mortality into growth-dependent and growth-independent hazards across 203 tropical tree species}
 \author{J.S. Camac, R. Condit, R.G. FitzJohn, L. McCalman, D. Steinberg, M. Westoby, S.J. Wright, D.S. Falster}
@@
 \end{figure*}
+
+\clearpage
+\begin{figure*}[!hb]
+\centering
+\includegraphics[width=0.6\textheight]{figures/supp_fig8.png}
+\caption{Relationship between tree size and both diameter increment growth (A) and basal area growth (B)}
+\label{figS8}
+\end{figure*}
+
 \end{document}

### 2018-11-03 ms/manuscript.Rnw
proof edits

@@
 \author[a,1]{James S Camac}
-\author[b]{Richard Condit}
+\author[b,c]{Richard Condit}
 \author[a,2]{Richard G FitzJohn}
-\author[c]{Lachlan McCalman}
-\author[c]{Daniel Steinberg}
+\author[d]{Lachlan McCalman}
+\author[d]{Daniel Steinberg}
 \author[a]{Mark Westoby}
-\author[d]{S Joseph Wright}
-\author[e]{Daniel S Falster}
+\author[e]{S Joseph Wright}
+\author[f]{Daniel S Falster}
 
-\affil[a]{Department of Biological Sciences, Macquarie University, Sydney, New South Wales, Australia}
-\affil[b]{Field Museum of Natural History, Chicago, IL, USA and Morton Arboretum, Lisle, IL, USA}
-\affil[c]{Data61, CSIRO, Australia}
-\affil[d]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
-\affil[e]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Science, University of New South Wales, Sydney, New South Wales, Australia}
+\affil[a]{Department of Biological Sciences, Macquarie University, Sydney, NSW, 2109, Australia}
+\affil[b]{Field Museum of Natural History, Chicago, IL, 60605, United States of America}
+\affil[c]Morton Arboretum, Lisle, IL ,60532, United States of America}
+\affil[d]{Data61, Commonwealth Scientific and Industrial Research Organisation (CSIRO), Sydney, NSW, 2015, Australia}
+\affil[e]{Smithsonian Tropical Research Institute, Apartado 0843–03092, Balboa, Panama}
+\affil[f]{Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, NSW, 2052, Australia}
 
@@
 We show that marginal carbon budgets are a major contributor to tree death on Barro Colorado Island.
-Moreover, we found that while species' light demand, maximum dbh and wood density affected tree mortality in different ways, they explained only a small fraction of the total variability observed among species.}
+Moreover, we found that while species' light demand, maximum diameter at breast height (dbh) and wood density affected tree mortality in different ways, they explained only a small fraction of the total variability observed among species.}
 
@@
 E-mail: james.camac@gmail.com.
-Present address: Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Melbourne, Victoria, Australia
+Present address: Centre of Excellence for Biosecurity Risk Analysis, School of Biological Sciences, The University of Melbourne, Melbourne, VIC, 3010, Australia
 
@@
 The unified model provided here separates mortality rates into growth-dependent and growth-independent hazards.
-This model creates the opportunity to simultaneously estimate the descriptors of variation both across and within species.
+This model creates the opportunity to simultaneously estimate these hazards both across and within species.
 Moreover, it provides the ability to examine how species traits affect growth-dependent and growth-independent hazards.
 We derive this unified mortality model  using cross-validated Bayesian methods coupled with mortality data collected over three census intervals for 203 tropical rainforest tree species at Barro Colorado Island (BCI) Panama.
-We found that growth independent mortality tended to be higher in species with lower wood density, higher light requirements and smaller maximum dbh.
+We found that growth independent mortality tended to be higher in species with lower wood density, higher light requirements and smaller maximum diameter at breast height (dbh).
 Mortality due to marginal carbon budget as measured by near-zero growth rate tended to be higher in species with lower wood density and higher light demand.
@@
 B) Our data consist of repeat measures of stem diameter ($D$) and status ($S$, 0 = alive or 1 = dead) for individual trees at specific census dates ($t_1$, $t_2$, $t_3$).
+Tree illustration courtesy of Tracey Saxby, Integration and Application Network, University of Maryland Center for Environmental Science (\url{ian.umces.edu/imagelibrary/}).
 C) Each model's skill in predicting observed outcomes ($S$) is quantified via the log-loss function (eq. \ref{eq:logloss}). Lower logloss means higher predictive accuracy.
-D) The predictive skill of alternative models was evaluated via 10-fold cross validation.
+D) The predictive accuracy of alternative models was evaluated via 10-fold cross validation.
 The entire dataset is split into 10 folds ($F_1$, ..., $F_{10}$).
 Alternative models were fit 10 times ($M_1$, ..., $M_{10}$), using different combinations of testing (1 fold; orange) and training (9 folds; green) data.
-Predictive capacity was assessed by averaging the log losses obtained from the 10 test data predictions.}
+Predictive accuracy was assessed by averaging the log loss obtained from the 10 test data predictions.}
 \label{fig1}
@@
 To illustrate the proposed mortality model, we assess its ability to predict the survival status of \Sexpr{n_inds(BCI_training_full)} individual trees from \Sexpr{n_spp(BCI_training_full)} species, recorded over three 5-year census periods at Barro Colorado Island (BCI), Panama.
-Applying the additive model, we: i) estimate growth-dependent and growth-independent effects from long-term census data, ii) estimate species-level differences in these effects, and iii) assess how well species-level differences are predicted from three fundamental traits, previously shown to correlate with mortality: wood density, light demand, and maximum dbh \cite{king:2006va, Kraft:2010kq, Wright:2010fl, Ruger:2011cv}.
+Applying the additive model, we: i) estimate growth-dependent and growth-independent effects from long-term census data, ii) estimate species-level differences in these effects, and iii) assess how well species-level differences are predicted from three fundamental traits, previously shown to correlate with mortality: wood density, light demand, and maximum diameter at breast height (dbh) \cite{king:2006va, Kraft:2010kq, Wright:2010fl, Ruger:2011cv}.
 These traits show only weak correlation with each other (r$^2$ \textless 0.13) and thus provide relatively independent measures of a species ecological strategy (SI Appendix, Fig.
@@
 Evaluating models in this way is computationally expensive.
-Therefore rather than running all possible model formulations, we fitted models across six levels of increasing complexity, described below, and comparing their predictive skill.
-We then use the final model to assess the effects of species traits on each parameter and quantify their relative contributions on the overall model performance.
+Therefore rather than running all possible model formulations, we fitted models across six levels of increasing complexity, described below, and comparing their predictive accuracy We then use the final model to assess the effects of species traits on each parameter and quantify their relative contributions on the overall model performance.
 Relative contributions of parameters were assessed using $r^2$ and Area Under the ROC Receiver Curve (AUROC).
@@
 \includegraphics{figures/fig2.pdf}
-\caption{Predictive skill of alternative hazard functions.
-Values are mean ($\pm$ 95\% confidence intervals) logarithmic loss, with lower values implying greater predictive skill and a logloss $\approx$ 0.693 indicating a model unable to distinguish an observed death from a survival.
+\caption{Predictive accuracy of alternative hazard functions.
+Values are mean ($\pm$ 95\% confidence intervals) logarithmic loss, with lower values implying greater predictive accuracy and a logloss $\approx$ 0.693 indicating a model unable to distinguish an observed death from a survival.
 A) Six sequential stages of model selection with each stage increasing in complexity.
@@
 The highest proportion of death occurred from 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_training_full, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_training_full, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_training_full, 3)}\%).
-Consequently, we observed a small increase in predictive skill when a census effect was included (leftmost two points under stage 1 in Fig.
+Consequently, we observed a small increase in predictive accuracy when a census effect was included (leftmost two points under stage 1 in Fig.
 \ref{fig2}; see also Table.~\ref{table:1} for quantification of gain in predictive power under successive models).
@@
 \subsection*{Stage 2: Hazard functions}
-Second, we assessed the predictive performance of three alternative hazard functions (Fig.
+Second, we assessed the predictive accuracy of three alternative hazard functions (Fig.
 \ref{fig1}A).
@@
 \ref{fig2}).
-Predictive skill was higher when using stem diameter growth compared to stem-area growth (Fig.
+Predictive accuracy was higher when using stem diameter growth compared to stem-area growth (Fig.
 \ref{fig2} stage 2, open vs closed symbols).
@@
 In general we found that including trait effects on these parameters decreased logarithmic loss, and thus, increased predictive accuracy (Fig.~\ref{fig2}).
-Allowing all three traits to have an effect on these parameters resulted in higher predictive capacity relative to models that only incorporated one or two of the traits.
+Allowing all three traits to have an effect on these parameters resulted in higher predictive accuracy relative to models that only incorporated one or two of the traits.
 
@@
 The findings of this study provide empirical support for dynamic vegetation models that predict tree mortality as a function of both growth-dependent and independent hazards \cite{Pacala:1996vm, Moorcroft:2001ws, McDowell:2011dr}.
-Regardless of growth measure (our proxy for carbon budget), incorporating both hazards significantly improved model predictive skill.
+Regardless of growth measure (our proxy for carbon budget), incorporating both hazards significantly improved model predictive accuracy.
 This was because the growth-dependent hazard allows for deaths associated with low carbon budgets, and as a consequence, incorporates intra-specific variability attributed to carbon related stresses such as competition, parasites and herbivory.
@@
 
-\subsection*{Evaluating model skill}
+\subsection*{Evaluating model predictive accuracy}
 Predictive skill was quantified by estimating the average log loss across 10-folds for held-out data, $\bar{\mathcal{L}}$ (Fig.

### 2018-11-03 ms/manuscript.Rnw
minor typo

@@
-\documentclass[9pt,twocolumn,twoside, lineno]{pnas-new}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[9pt,twocolumn,twoside]{pnas-new}\usepackage[]{graphicx}\usepackage[]{color}
 
@@
 \affil[b]{Field Museum of Natural History, Chicago, IL, 60605, United States of America}
-\affil[c]Morton Arboretum, Lisle, IL ,60532, United States of America}
+\affil[c]{Morton Arboretum, Lisle, IL ,60532, United States of America}
 \affil[d]{Data61, Commonwealth Scientific and Industrial Research Organisation (CSIRO), Sydney, NSW, 2015, Australia}
@@
 \acknow{We thank H.
-Muller-Landau and anonymous reviewers for feedback and B.
+Muller-Landau, Peter Vesk and anonymous reviewers for feedback and B.
 Carpenter for technical advice.
@@
 Dolins, \& hundreds of field workers for their contribution; and the National Science Foundation, Smithsonian Tropical Research Institute, \& MacArthur Foundation for funding the design, collection, quality control and management of long-term growth data at BCI.
-Fig.
-1b uses images by Tracey Saxby, IAN Image Library.
 J.S.C., R.G.F., L.M.
